### 2016-02-04 misc/labmtg.rmd
lab mtg stuff

---
title: HIV virulence evolution in structured epidemic models
author: Ben Bolker and Daniel Park
date: "`r format(Sys.time(), '%H:%M %d %B %Y ')`"
bibliography: "../manuscript1/virulence.bib"
csl: apa.csl
output:
  ioslides_presentation
---
<!-- 
apa.csl is a slightly hacked version of APA 
  (modified for "et al" after 2 authors in text)
-->
<!-- .refs is style for reference page (small text) -->
<style>
.refs {
   font-size: 12px;
}
h2 { 
 color: #3399ff;		
}
h3 { 
 color: #3399ff;		
}
.title-slide {
   background-color: #55bbff;
}
div#before-column p.forceBreak {
	break-before: column;
}

div#after-column p.forceBreak {
	break-after: column;
}
</style>
<!--    content: url(https://i.creativecommons.org/l/by-sa/4.0/88x31.png)
>
<!-- Limit image width and height -->
<style type="text/css">
img {     
  max-height: 560px;     
  max-width: 800px; 
}
</style>

```{r setup,echo=FALSE,message=FALSE}
library("ggplot2"); theme_set(theme_classic())
library("MASS")
scale_colour_discrete <- function(...,palette="Set1") scale_colour_brewer(...,palette=palette,guide=guide_legend(reverse=TRUE))
scale_fill_discrete <- function(...,palette="Set1") scale_fill_brewer(...,palette=palette,guide=guide_legend(reverse=TRUE))
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally")
## devtools::install_github("lionel-/ggstance")
library("ggstance")
library("knitr")
opts_chunk$set(echo=FALSE,fig.width=4,fig.height=4) ## ,out.width=400)
```

# Intro to virulence evolution

## Tradeoff theory

- assume that virulence evolution is mediated by parasite-level tradeoff between transmission and "virulence" (clearance/mortality rate)
- need a *saturating* relationship: $\frac{d^2 \, \textrm{transmission}}{d\,\textrm{clearance}^2}<0$
- mediated through pathogen replication rate (?)
- simple form assumes no within-host competition (single clone per host)
- alternatives: short-sighted evolution, within-host competition, effects of genetic drift ... [@LevinBull1994,@EbertBull2003]

## Eco-evolutionary dynamics

- feedback between ecological and evolutionary processes
- typical emphasis on *short-term* dynamics (cf. adaptive dynamics)
- eco-evo-epi: *transient virulence* [@bolker_transient_2010]
- balance between $r$-selection (epidemic phase) and ${\cal R}_0$-selection (endemic phase)
- transient peak observed in phage systems [@berngruber_evolution_2013]

## Virulence evolution in HIV

- set-point viral load (SPVL)
- @Fraser+2007 : plausible tradeoff observed from Rakai data
    - heritability of SPVL
	- correlated with rate of progression to AIDS
	- correlated with probability (rate) of within-couple transmission
- @shirreff_transmission_2011 : transient peak after $\approx 1$ in a *reasonably* realistic model

# Modeling virulence evolution in HIV

## @shirreff_transmission_2011 : evolutionary/within-host

- generally based on Rakai estimates
- Normally distributed mutation
- Normally distributed phenotypic variation
- different infectivity for early, mid, late HIV stages
- Hill functions ($x^d/(a+bx^d)$) for infectivity, progression through mid stage
- Weibull time in mid stage

## @shirreff_transmission_2011 : epidemiological

- **very simple**
- implicit partnership-formation model
$$
\textrm{foi} = \frac{\beta c}{\beta + c + d}
$$

- derived from ${\cal R}_0$ for an instantaneous-partnership-formation model, e.g. @hollingsworth_hiv1_2008, @diekmann_mathematical_2000
- (@herbeck_hiv_2014 is worse [I think])

## Epidemiological models for HIV [@champredon_hiv_2013]

<!-- convert journal.pone.0082906.g001.png -crop 100%x50% output.png-->

![](champ1.png)

## Model modifications

- @champredon_hiv_2013 track $\{X,Y,N,D,P\}$ ($\{S,I,SS,SI,II\}$) $\times$ infected strain
- we need $\{S,SS,I(\alpha),SI(\alpha),II(\alpha,\alpha')\}$
- simplifications:
    - single-stage, exponential infectious period  
($\beta =$ time-weighted average of $\beta_1$, $\beta_2(\alpha)$, $\beta_3$)
	- no phenotypic variation

## Model variations

- 4 models:
    - extra-pair contact (yes/no) 
    - instantaneous partnership formation (yes/no)
- implicit (Shirreff/Hollingsworth) model

## Latin hypercube sampling {.columns-2}

- sample evenly, randomly across (potentially many) uncertain parameters
- following @champredon_hiv_2013, earlier @blower_drugs_1991

<p class="forceBreak"></p>

![](LHScrop.png)
[User:Saittam, Wikipedia](https://commons.wikimedia.org/wiki/File:LHSsampling.png#/media/File:LHSsampling.png)

## Details

- $r$ varies widely across parameters/model sets
     - *scale* parameters to $r=0.04$ (partnership formation, dissolution, *and* transmission probability)
     - scale both uncoupled and extra-coupled rates relative to within-coupled rate
- *response variables*
     - equilibrium virulence
	 - peak virulence time (years)
	 - peak virulence
	 - relative peak virulence (peak/eq)

# Results

```{r load,echo=FALSE,cache=TRUE,message=FALSE}
## get all the batch output and combine it into a convenient (?) format
source("../hivFuns.R")  ## must be loaded before LHSparms3
source("../LHSparms3.R")
ltab <- data.frame(run=seq(nrow(ltab)),ltab)
##  combine them/rename appropriately
run_nums <- c("pairform+epc"=5,"pairform"=6,
               "instswitch"=7,"instswitch+epc"=8,"implicit"=9)
## each entry has
##    I_mat*:
##    vir_mat*:
##    eq_vec*: equilibrium value
##    peak_mat*: peak value (time, height)
##    val_vec*: ???
for (i in run_nums)
    L <- load(paste0("ev_LHS_res",i,".rda"))
## combine
component_names <- c("I_mat","vir_mat","eq_vec","peak_mat","val_vec")
## make one big nested list of components
comp_list <- lapply(component_names,
                    function(c) {
    res <- lapply(paste0(c,run_nums),get)
    setNames(res,names(run_nums))
})
names(comp_list) <- component_names
mat_names <- c("I_mat","vir_mat")
## reduce matrices to melted data frame of mean & quantiles
##  (we don't really need to make giant pictures with all of the
##   individual lines - ribbons with mean and 95% quantiles should
##   be sufficient)
sum_mat_fun <- function(x) {
    data.frame(
        tvec=1:nrow(x),  ## ?? better time index?
        mean=rowMeans(x),
        lwr=apply(x,1,quantile,0.025),
        upr=apply(x,1,quantile,0.975))
}
sum_list <- list()
for (m in mat_names) {
    sum_list[[m]] <- ldply(comp_list[[m]],
                           sum_mat_fun, .id="model")
}
s1 <- ldply(comp_list[["eq_vec"]],
                function(x) data.frame(run=1:length(x),eq_vir=x),
                .id="model")
s2 <- ldply(comp_list[["peak_mat"]],
                function(x) {
                   x <- setNames(data.frame(x),c("peak_time","peak_vir"))
                   data.frame(run=1:nrow(x),x)
                 },
                .id="model")
sum_list[["sum_mat"]] <- full_join(s1,s2)
## unlike merge(), full_join() preserves run order
```

```{r fix_order}
mods <- c("pairform+epc","instswitch+epc","instswitch","pairform","implicit")
fixfac <- function(x) {
    mutate(x,model=factor(model,levels=mods))
}
sum_list[["vir_mat"]] <- fixfac(sum_list[["vir_mat"]])
sum_list[["I_mat"]] <-   fixfac(sum_list[["I_mat"]])
```

## Epidemic trajectories

```{r epi_traj,fig.width=10}
gg_Itraj <- ggplot(sum_list[["I_mat"]],
       aes(tvec,mean,ymin=lwr,ymax=upr))+
    geom_line(aes(colour=model))+
    geom_ribbon(aes(fill=model),alpha=0.4)+
    labs(y="mean prevalence")
grid.arrange(gg_Itraj+ggtitle("linear scale"),
             gg_Itraj+scale_y_log10()+ggtitle("log scale"),nrow=1)
```

## Virulence trajectories {.centered}

```{r vir_traj,fig.width=6}
(gg_virtraj <- ggplot(sum_list[["vir_mat"]],
       aes(tvec,mean,ymin=lwr,ymax=upr))+
    geom_line(aes(colour=model))+
    geom_ribbon(aes(fill=model),alpha=0.4))
```

## Univariate distributions

```{r unidist,fig.width=7}
sL <- fixfac(transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir))
mL <- fixfac(melt(sL,id.vars=c("model","run")))
ggplot(mL,aes(value,model,fill=model))+geom_violinh()+
    facet_wrap(~variable,scale="free")+
    labs(y="",x="")
```

## Pairs plots {.centered}

```{r ggpairs_ex,fig.width=7,fig.height=6.5,dev="png"}
ggp1 <- ggpairs(sL,
        mapping = ggplot2::aes(color = model),
        columns=3:6,
        lower = list(continuous = wrap("points", alpha = 0.5)),
        diag = list(continuous = wrap("densityDiag", alpha=0.5)),
        upper = list(continuous = "blank"))+
    theme(panel.margin=grid::unit(0,"lines"))
## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot
cbscales <- list(scale_colour_discrete(),
                 scale_fill_discrete())
for (row in seq_len(ggp1$nrow))
    for (col in seq_len(ggp1$ncol))
        ggp1[row, col] <- ggp1[row, col] + cbscales[[(row==col)+1]]
ggp1
```

```{r corrsum,message=FALSE,cache=TRUE}
m1 <- rename(melt(sum_list[["sum_mat"]],id.vars=c("model","run")),
             sumvar=variable,sumval=value)
m2 <- rename(melt(ltab,id.vars="run"),
             LHSvar=variable,LHSval=value)
mL2 <- full_join(m1,m2)
```

## Sensitivity

```{r corrsum_plot,dev="png",echo=FALSE,fig.width=10,fig.height=6}
ggplot(mL2,aes(LHSval,sumval,colour=model))+
    geom_point(pch=".",alpha=0.5)+
    facet_grid(sumvar~LHSvar,scales="free")+
    geom_smooth(se=FALSE)+labs(x="",y="")
```

## Conclusions/questions: details

- essentially:
implicit $\approx$ pairform < instswitch < instswitch+epc $\approx$ pairform+epc
- why no low-eq-vir tail for pairform+epc?
- why is pairform, not instswitch, closest to implicit?

## Conclusions/questions: big picture

- how do we decide what to pick on?
- how does scaling $r$ help/hurt?

## References {.refs .columns-2}

### 2016-03-12 R/hivFuns.R
update R code

@@ -1,13 +1,6 @@
 expand <- function(x, ...) {
     UseMethod("expand")
 }
-
-as.HIVvirparlist <- function(x) {
-  res <- append(x,HIVpars.skeleton)
-  class(res) <- c("list","parlist","HIVvirparlist")
-  return(res)
-}
-
 expand.HIVvirparlist <- function(x,...) {
     ## mutate() doesn't work this deeply ...
     x <- within(x,
@@ -15,22 +8,22 @@ expand.HIVvirparlist <- function(x,...) {
           min.alpha <- alphaDist[["min"]]
           max.alpha <- alphaDist[["max"]]
           d.alpha <- alphaDist[["delta"]]
-          alpha <- seq(min.alpha,max.alpha,by=d.alpha)
+          alpha <- seq(alphaDist[["min"]],alphaDist[["max"]],by=d.alpha)
           n.alpha <- length(alpha)
           
           rho <- scale_all * scale_c * rho_base
           c_mean <- scale_all * scale_c * c_mean_base
           
-          v3 <- 10^alpha
-          ## duration <- (Dmax * D50^h)/(D50^h + v3^h)
-          Duration2 <- hill(v3,Dmax,D50,h)
+          v <- 10^alpha
+          ## duration <- (Dmax * D50^h)/(D50^h + v^h)
+          Duration2 <- hill(v,Dmax,D50,h)
           lam = 1/(Duration1 + Duration2 + Duration3)
           lammat = matrix(rep(lam,n.alpha), n.alpha, n.alpha)
           lammat_dis = lammat
           diag(lammat_dis) = 2 * diag(lammat_dis)
           lammat_adj = lammat + t(lammat)
-          ## Beta2 <- Bmax * v3^a / (v3^a + K^a)
-          Beta2 <- hill(K,Bmax,v3,a)
+          ## Beta2 <- Bmax * v^a / (v^a + K^a)
+          Beta2 <- hill(K,Bmax,v,a)
           Beta = scale_all * scale* (Beta1 * Duration1 + Beta2 * Duration2 + Beta3 * Duration3) * lam
           c_e = Beta * c_e_ratio
           c_u = Beta * c_u_ratio
@@ -41,6 +34,7 @@ expand.HIVvirparlist <- function(x,...) {
 hill <- function(x,a,b,p) {
     a*b^p/(b^p+x^p)
 }
+
 
 ##' get r value within specified bounds
 ##'
[…]

### 2016-03-12 R/hivModels.R

@@ -1,97 +1,167 @@
-## compute mutation matrix
 makeMutMat <- function(xpars) {
-    p <- with(xpars, {
-         p <- matrix(NA, n.alpha, n.alpha)
-         for(i in 1:n.alpha){
-             x <- min.alpha + (i-1)*d.alpha
-             denom <- 
-                 diff(pnorm(c(min.alpha-d.alpha/2,max.alpha+d.alpha/2),
-                              mean=x, sd=Vm))
-             for (j in 1:n.alpha){
-                 k <- min.alpha + (j-1)*d.alpha
-                 p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
-             }
-         }
-         p
-         })
-   return(p)
+  p <- with(xpars, {
+    p <- matrix(NA, n.alpha, n.alpha)
+    for(i in 1:n.alpha){
+      x <- min.alpha + (i-1)*d.alpha
+      denom <- 
+        diff(pnorm(c(min.alpha-d.alpha/2,max.alpha+d.alpha/2),
+                   mean=x, sd=Vm))
+      for (j in 1:n.alpha){
+        k <- min.alpha + (j-1)*d.alpha
+        p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
+      }
+    }
+    p
+  })
+  return(p)
 }
 
-
-## multi-strain CD model; non-instantaneous partnership, extra-pair contact
-gfun_epc <- function(parameters,mm=TRUE) {
-    pp <- expand(parameters)
-    attach(pp)
-
-    if (mm) {
-        p <- makeMutMat(pp)
-    } else {
-        p <- matrix(NA, n.alpha, n.alpha)
-        ## compute mutation matrix
-        for(i in 1:n.alpha){
-            x <- alphaDist[["min"]] + (i-1)*d.alpha
-            for (j in 1:n.alpha){
-                k <- alphaDist[["min"]] + (j-1)*d.alpha
-                denom <- with(as.list(alphaDist),
-                              diff(pnorm(c(min-d.alpha/2,max+d.alpha/2), mean=x, sd=Vm)))
-                p[i,j] <- diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
[…]

### 2016-03-12 R/simFuns.R

@@ -54,9 +54,3 @@ Lsub <- function(L,subs) {
         L[[n]] <- subs[[n]]
     L
 }
-expand <- function(x, ...)
-{
-    UseMethod("expand")
-}
-
-

### 2016-03-12 ms/ms.Rnw
manuscript, sim data

\documentclass{article}

%\bibpunct{(}{)}{;}{a}{}{,}
\usepackage{natbib}
\usepackage{palatino}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = blue]{hyperref}
\usepackage{url}
\usepackage{enumerate}
\usepackage[left=50pt,
            textwidth=345pt,
            marginparsep=25pt,
            marginparwidth=124pt]{geometry}
%\usepackage[margin=1in]{geometry}

\newcommand{\shir}{\citeauthor{shirreff_transmission_2011}}
\newcommand{\champr}{\citeauthor{champredon_hiv_2013}}
\newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
\usepackage{changes}
\usepackage[backgroundcolor=lightgray,textsize=small]{todonotes}

<<setup,echo=FALSE>>=
library("knitr")
opts_chunk$set(echo=FALSE,error=FALSE,fig.width=4,fig.height=4) #
@

<<pkgs,warning=FALSE,message=FALSE>>=
library("ggplot2"); theme_set(theme_classic())
library("MASS")
scale_colour_discrete <- function(...,palette="Set1") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1") scale_fill_brewer(...,palette=palette)
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally")
library("knitr")
## devtools::install_github("lionel-/ggstance")
library("ggstance")
@

<<loadstuff,cache=TRUE>>=
load("../simdata/combine_ev_LHS4.rda")
@

\author{Sang Woo Park and Ben Bolker}
\title{Effects of epidemiological structure on the transient evolution of HIV virulence}

\begin{document}
\maketitle

\section{Introduction}

How pathogen virulence evolves is a fundamental question in evolutionary biology, of great theoretical and, potentially, practical importance. While various researchers have offered critiques [CITE Ebert and Bull, Alizon and Michalakis], the trade-off theory --- which postulates that parasite virulence can be explained as the long-term evolutionary outcome of a saturating relationship between parasite clearance rate and transmission rate --- has also been successfully applied in several important cases [CITES?]. One particularly interesting application of these ideas is the work by Fraser et al [CITES] showing that HIV appears to satisfy the prerequisites of the tradeoff theory: in a study of discordant couples (i.e. long-term sexual partnerships with one infected and one uninfected partner), HIV virulence as measured by the rate of progression to AIDS was both heritable and covaried  with the set-point viral load ([DEF]), which in turn predicted the probability of within-couple transmission. Subsequent studies [CITES] used these data to parameterize mechanistic models of HIV virulence evolution, suggesting that HIV invading a novel population would initially evolve increased virulence, peaking after approximately [XXX] years and then declining slightly to a long-stable virulence level.

The work of \shir, and particularly the predicted
transient peak in HIV virulence midway through the epidemic,
highlights the importance of interactions between epidemiological and
evolutionary factors \citep{day_virulence_2004,alizon_price_2009}.
However, despite the attention to mechanistic detail at the individual
or physiological level, the epidemiological structures used in these
models are relatively simple. As we discuss in detail below, the
existing models of HIV eco-evolutionary dynamics either use implicit
models that incorporate the average effects of within-couple sexual
contact --- without representing the explicit dynamics of pair
formation and dissolution or accounting for extra-partnership contact
--- or use an agent-based formulation with parameters that effectively
lead to random mixing among infected and uninfected individuals. Here
we explore the effects of incorporating \emph{explicit}
epidemiological structure in eco-evolutionary models.

We add complexity to the epidemiological model following the general approach of \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships. At the same time, our analysis somewhat simplifies the models of \shir, for computational tractability; we check that our qualitative results are not sensitive to these simplifications. In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity. In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following \champr\ in using a Latin hypercube design. For each model run, we compute a set of metrics (peak virulence, timing of virulence peak, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.

\section{Methods}

\subsection{Model formulation}

As our primary goal is to explore how different epidemiological structures (i.e. partnership dynamics and contact structures) affect our conclusions about the evolution of virulence, our models use a simplified description of within-host dynamics and heritability derived from 
\shir's multi-strain evolutionary model.
Like \shir, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than virulence (i.e. rate of progression to AIDS) itself. 
Instead of accounting for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), we use a single-stage disease model, with a simple exponentially distributed infectious period instead of the Weibull-distributed infectiousness which \shir\ (and many other HIV models) use. However, we do account for varying transmission rates and durations of each disease stage by summing the durations of three stages 
(again based on \shir's model) and taking the duration-weighted average of transmission rates of three stages. Thus the within-couple transmission rate, $\beta$, for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D}
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the \textbf{P}rimary and \textbf{D}isease stages
of infection are independent of the host's SPVL. Following \shir, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the \textbf{A}symptomatic stage are Hill functions of the SPVL:

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{D_{\text{max}} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\beta_{\text{max}} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The \textbf{u}ncoupled and \textbf{e}xtra-couple transmission rates are scaled by multiplying the \textbf{w}ithin-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$.

Like \shir\ but unlike \champr, we simplify the evolutionary model by using a one-to-one genotype-phenotype mapping (i.e. a single strain can only produce a single SPVL). Unlike \champr\ but like \shir, we also use a simple susceptible-infected-susceptible formulation; rather than modeling birth and death (or more specifically, recruitment into the sexually active population and death), we assume that whenever an individual dies from infection, another enters the susceptible compartment.

We developed six multi-strain evolutionary models, each of which is based on different assumptions regarding partnership dynamics. Specifically, we focus on the effect of instantaneous partnership formation and extra couple mixing on the evolution of mean $\log_{10}$ SPVL.

The first four models consider explicit partnership dynamics and are based on Champredon et al's model. Model 1 and 2 assume non-instantaneous partnership formation and consist of five states that are classified by infection status and partnership status. $S$ is the number of single susceptible individuals, and $I$ is the number of single infected individuals. $SS$ is the number of susceptible-susceptible couples, $SI$ is the number of serodiscordant couples, and $II$ is the number of concordant positive couple. Model 1 includes extra couple mixing and uncoupled mixing whereas model 2 only considers within-couple transmission. Model 3 and 4 assume instantaneous partnership formation and consist of 3 states: $SS$, $SI$, and $II$. Like model 1 and 2, model 3 includes extra couple mixing and model 4 only considers within-couple transmission.

In contrast, models 5 and 6 are not explicitly structured.  Model 5 is an implicit serial monogamy model based on \shir's model. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to simulate the effect instantaneous partnership formation, it uses adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model. Finally, model 6 is a simple random-mixing model.

The mutational process in our model is directly taken from \shir. Over the course of infection, mutation occurs within the host. However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection for simplicity (and is not further affected by within-host mutation). Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual. First, the distribution of $\log_{10}$ SPVL is discretized into a vector:

\begin{equation}
\alpha_i =  (\alpha_{max} - \alpha_{min})\frac{(i-1)}{n-1} + \alpha_{min} \qquad i = 1,2,3, \dots n.
\end{equation}
Then, we construct a mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have $\log_{10}$ SPVL of $j$ given that the infector has $\log_{10}$ of $i$. Finally, the probabilities are normalized so that each row sums to 1:

\begin{equation}
M_{ij} = \frac{\Phi(\alpha_j + d/2) - \Phi(\alpha_j - d/2)}{\Phi(\alpha_{max} + d/2) - \Phi(\alpha_{min} - d/2)},
\end{equation}

where $\Phi(j)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\alpha_{\max} - \alpha_{\min})/n$. Transmission rate and disease induced mortality rates are discretized into a vector as well:

\begin{equation}
\begin{aligned}
\beta_i &= \beta(\alpha_i)\\
\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}
\end{aligned}
\end{equation}

\subsection{Latin Hypercube Sampling - we should probably explain model parameters before this...?}

Not only is there discrepancy in parameters regarding partnership dynamics between sources, but it is also difficult to determine the exact value due to many reasons. Furthermore, recent studies suggest that the previously measured transmission rate of the acute phase was over-estimated. Thus, we want to allow for uncertainties in model parameters and perform Latin hyper cube sampling on the following parameters: $\beta_P$, $D_P$, $\beta_D$, $D_D$, $\rho$, $c$, $c_u/c_w$, and $c_e/c_w$. Each parameter is divided  We do not allow for uncertainties in parameters that are directly related to evolutionary process.

\todo{refer to Table 1 and describe where the new parameters come from; e.g. decisions we made about new scaling parameters}

\input{ms_parms.tex}

\subsection{Simulation runs}

\todo{BMB: add to this description}
In disease modeling, one of the most important parameter is the initial exponential growth rate. As it is one of the few parameters that can be directly observed, many modelers try to match the initial exponential growth rate to correctly simulate the epidemic. Furthermore, evolution of virulence is sensitive to the initial growth rate. Thus, we scale a parameter so that every run has an equal exponential rate of growth rate.

In order to allow for all models to have equal initial exponential growth rate, $r$, we need to pick a parameter, $s$, so that $\lim_{s\to 0} r(s) = 0$ and $\lim_{s\to\infty}  r(s) = \infty$. As adjusting either partnership change rate (i.e. partnership formation and dissolution rate) or transmission rate does not fulfill this requirement for certain models, we decided to scale partnership change rate and dissolution rate by the same factor of $\gamma$: $\tsub{beta}{adj} =  \gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$, $\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$. Since transmission rate is adjusted by the scale of $\gamma$, uncoupled and extra-couple transmission rates are adjusted as well. For model 3, 4, and 5, all of which assume instantaneous partnership, only the transmission rate and partnership dissolution rate are adjusted.

We run each model for 1000 times with different parameter sets calculated from latin hypercube sampling with fixed starting condition: mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$ (refer to appendix). After each run, initial exponential growth rate is calculate. Then, paraeters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.

For each model we derive the following summary statistics:

\todo{define each summary statistic and explain what it tells us}
\begin{description}
\item[peak time] ...
\item[peak virulence] ...
\item[relative peak virulence] ...
\item[equilibrium virulence] ...
\end{description}

\section{Results}

<<panel3>>=

@ 
\begin{itemize}
\item Time plot (3-panel plot): baseline dynamics, illustrating summary statistics (peak time, eq virulence, length of time until peak virulence). Full Shirreff model, effects of $I(0)$, $\alpha(0)$, $r$
\item Virulence trajectories under all models
\item Univariate plots
\item Bivariate plots
\item Unscaled sensitivity 
\end{itemize}

\section{Discussion}

\section{To do}
\begin{itemize}
\item clean up code and results
\item incorporate pictures
\item finish writing!
\item hard parts of the results
\begin{itemize}
\item explaining similarities among the models: why does implicit look most like pair formation model? (see R0comparison HTML)
\item explaining sensitivity plot
\end{itemize}
\end{itemize}

\appendix

See \href{http://tinyurl.com/bolker-jrsi-cites}{Bolker JRSI Google scholar cites} (\href{https://scholar.google.ca/scholar?oi=bibs&hl=en&cites=18202064200267310847}{alternative link})

\section{Model details}

Since we use multi-strain models, in which the distribution of $\log_{10}$ SPVL has been discretized into a vector, we use a matrix notation to describe our models. Five states described in the method section is replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$. Subscript indicates a strain that an individual is infected with. For example, $I_i$ is number of infected individuals with $\log_{10}$ SPVL of $\alpha_i$, and $II_{ij}$ is the number of concordant couples, in which two partners have $\log_{10}$ SPVL of $\alpha_i$ and $\alpha_j$. $II_{ij}$ is equivalent to $II_{ji}$. Lastly, we define Kronecker delta as follows:

\begin{equation}
\delta_{ij} = \left.\begin{aligned}
0 & \quad \text{if} \quad i \neq j,\\
1 & \quad \text{if} \quad i = j.
\end{aligned}\right\rbrace
\end{equation}

\subsection*{Model 1 and 2 - Partnership dynamic}

Single individuals acquire partners at rate $\rho$: $S' = - \rho S$ and $I_i' = - \rho I_i$. We follow Champredon et al's results and assume that single individuals are distributed into couple states through binomial distribution:

\begin{equation}
\begin{aligned}
SS' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)}\\
SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
II_{ij}' &= (\frac{1}{2})^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
\end{aligned}
\end{equation}
We introduce Kronecker delta above to differentiate the partnership formation rate for $II_{ij}$ 
when $i = j$ from that of $I \neq j$. When $i = j$ --- like the partnership formation rate of $SS$ --- partnership formation rate becomes $II_{ii}' = \frac{\rho I_i \cdot I_i}{2(S + \sum_k I_k)}$ due to binomial distribution. On the other hand, when $i \neq j$, partnership formation rate becomes $II_{ij}' = \frac{\rho I_i I_j}{S + \sum_k I_k}$.

Partnerships dissolve at rate $c$: $SS' = -c SS$, $SI_i' = -c SI_i$, and $II_{ij} = - c II_{ij}$. Unlike single strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains which the individuals in concordant partnership are infected with (i.e. both partners in $II_{ii}$ enter $I_i$ whereas only one partner in $II_{ij}$ enter $I_i$).

\begin{equation}
\begin{aligned}
S' &= 2 c SS + \sum_k c SI_k \\
I_i' &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}\\
\end{aligned}
\end{equation}
Combining partnership formation and dissolution process yields the following equation:

\begin{equation}
\begin{aligned}
S' &= - \rho S + 2 c SS + \sum_k c SI_k \\
I_i' &= - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}\\
SS' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS\\
SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i\\
II_{ij}' &= (\frac{1}{2})^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k} - c II_{ij}
\end{aligned}
\end{equation}

\subsection*{Model 1 and 2 - Infection}

Within-couple transmission occurs in both models. An infected partner in $SI$ partnership transmits virus to a susceptible partner, and partnership state becomes $II$: $SI_i' = - \beta_i SI_i$. Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $II_{ij}' = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$. On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $II_{ii}' = M_{ii} \beta_i SI_i$. Using the Kronecker delta notation, we obtain following set of equations that describe within-couple transmission dynamics:

\begin{equation}
\begin{aligned}
SI_i' &= - \beta_i SI_i\\
II_{ij} &=  (\frac{1}{2})^{\delta_{ij}} \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)
\end{aligned}
\end{equation}

Champredon et al define the proportion of infectious extra-couple and uncoupled contact through the following term:

\begin{equation}
P = \frac{c_u I + c_e (SI + 2 II)}{c_u (S + I) + 2 c_e(SS + SI + II)}.
\end{equation}
Effective uncoupled, $c_u$, and extra couple, $c_e$, contact rate can be divided into two terms: uncoupled/extra-couple contact rate $\times$ rate of transmission per contact. Therefore, transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above. Using this property, we modify the equation above as follows:

\begin{equation}
P = \frac{r_u I + r_e (SI + 2 II)}{r_u (S + I) + 2 r_e(SS + SI + II)},
\end{equation}
where $r_u = c_u/c_w$ and $r_e = c_e/c_w$ are the relative uncoupled/extra-couple contact rates. This simplification is useful in a multi-strain model since we cannot multiply a vector with a single value (e.g. $c_u S$ in denominator) if we use Champredon et al's equation as it is. Extending the above equation to multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain the following equation:

\begin{equation}
P_i = \frac{r_u I_i + r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_u (S + \sum_k I_k) + r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k (1 + \delta_{lk})II_{lk} )},
\end{equation}
Using the equation above, we can model uncoupled and extra-couple mixing. For convenience, uncoupled and extra-couple transmission rates, $c_u$ and $c_e$, will be replaced with $U_i = r_u \beta_i$ and $E_i = r_e \beta_i$ from now on.

Single susceptible individuals become infected and enter single infected compartment at the total rate of $\sum_k P_k U_k S$. Through mutation, newly infected individuals are distributed into each single infected compartments with different strains: $I_i' = \sum_k M_{ki} P_k U_k S$. Either partners in $SS$ partnership can be infected and the partnership state can become $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$. Formation of $SI_i$ partnership is similar to the process through which single susceptible individuals are distributed into single infected compartments: $SI_i' = \sum_k 2 M_{ki} P_k E_k SS$. Lastly, susceptible partner of an $SI$ partnership can be infected due to uncoupled/extra-couple contacts and partnership can move to an $II$ partnership. Like previous cases, $SI_i$ partnership moves out of the compartment at a rate of $\sum_k P_k E_k SI_i$. Mutation process is similar to that of within-couple transmission. The only difference is that the $\log_10$ SPVL of a newly infected partner is not determined by its original partner but from an extra couple partner (i.e. the term $P_i$): $II_{ij}' = (1 -\frac{\delta_{ij}}{2})(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$. Combining these equations we get the following set of equations that describe all the transmission dynamics

\subsection*{Model 1 and 2 - Disease induced mortality}

Disease induced death rate, $\lambda$, is given by taking the reciprocate of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$. Since we assume SIS formulation, where infected individuals that die from infection enter single susceptible compartment, we obtain the following equation for the single infected individuals:

\begin{equation}
\begin{aligned}
S' &= \sum_k \lambda_k I_k \\
I_i &= - \lambda_i I_i \\
\end{aligned}
\end{equation}

If an infected individual in a partnership dies, partnership dissolves. Thus, an $SI_i$ partnership dissolves at a rate $-\lambda_i$, and the susceptible partner enters single susceptible compartment at rate $\lambda_i SI_i$ (infected partner that dies enter single susceptible compartment at an equal rate as well due to SIS formation):

\begin{equation}
\begin{aligned}
S' &= \sum_k 2 \lambda_k SI_k \\
SI_i &= - \lambda_i SI_i 
\end{aligned}
\end{equation}

Similarly, $II_{ij}$ partnership dissolves at a rate $-(\lambda_i + \lambda_j)$, but two cases, when $i \neq j$ and $i = j$, must be considered separately. When an $II_{ij}$ partnership dissolves due to disease induced mortality, where $i \neq j$, death of partner with strain $i$ causes its partner to enter $I_j$ compartment at rate $\lambda_j II_{ij}$, and vice versa. When an $II_ii$ partnership dissolves, death of either partner causes the other partner to enter $I_i$ compartment at rate $\lambda_i II_{ii}$, which sums up to $2\lambda_i II_{ii}$. Combining these dynamics yield the following equation:

\begin{equation}
\begin{aligned}
S' &= \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk} \\
I_i' &=  \sum_k 2^{\delta_{ik}} \lambda_k II_{ik} \\
II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij}
\end{aligned}
\end{equation}

Finally, combining all these equations give us the full model, which is model 1. We can simply take out the uncoupled and extra couple transmission term to obtain equation 2:

\begin{equation}
\begin{aligned}
S' =& - \rho S + 2 c SS + \sum_k c SI_k - \sum_k P_k U_k S + \sum_k \lambda_k I_k \\
&+ \sum_k 2 \lambda_k SI_k + \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk}\\
I_i' =&  - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c  II_{ik} + \sum_k M_{ki} P_k U_k S- \lambda_i I_i \\
&+ \sum_k 2^{\delta_{ik}} \lambda_k II_{ik} \\
SS' =& \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS - \sum_i 2 P_i E_i SS \\
SI_i' =& \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS - \sum_k P_k E_k SI_i   \\
&- \lambda_i SI_i\\
II_{ij}' =& (\frac{1}{2})^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{(S + \sum_k I_k)} - c II_{ij} + (\frac{1}{2})^{\delta_{ij}} \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
&+ (\frac{1}{2})^{\delta_{ij}} \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
\end{aligned}
\end{equation}

\subsection*{Model 3 and 4 - Partnership dynamic}

Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$. Partnership dissolution is equal to that of model 1 and 2: $SS' = -cSS$, $SI_i' = - cSI_i$, and $II_{ij}' = - II_{ij}$. Once the individuals leave partnership, they enter temporary compartments and are distributed into a partnership through binomial distribution:

\begin{equation}
\begin{aligned}
X &= 2 c SS + \sum_k c SI_k \\
Y_i &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik} \\
SS' &= - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}\\
SI_i' &= - c SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
II_{ij}' &= - c II_{ij} +(\frac{1}{2})^{\delta_{ij}} \frac{Y_i Y_j}{X + \sum_k Y_k}.
\end{aligned}
\end{equation}

\subsection*{Model 3 and 4 - Infection}

Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single state, only extra couple transmission exists:

\begin{equation}
P_i = \frac{r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k (II_{lk} + \delta_{kl} II_{lk}) )}.
\end{equation}
Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equation that is used in model 1 and 2.

\subsection*{Model 3 and 4 - Disease induced mortality}

Disease induced mortality is modeled similar to model 1 and 2. However, as single state does not exist in model 3 and 4, individuals that has left their partnerships due to death of their partners enter temporary compartments and form partners instantly:

\begin{equation}
\begin{aligned}
X &= \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk} \\
Y_i &=  \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik} \\
SS = &= \frac{X^2}{2 (X + \sum_k Y_k)}\\
SI_i &= - \lambda_i SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + (\frac{1}{2})^{\delta_{ij}} \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}
\end{aligned}
\end{equation}

Combining all these dynamics, we have equation 3. If we remove extra-couple transmission, we have equation 4.

\begin{equation}
\begin{aligned}
X =& 2 c SS + \sum_k c SI_k + \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk}\\
Y_i =& c SI_i + \sum_k 2^{\delta_{ik}}  c II_{ik} + \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik} \\
SS  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS\\
SI_i =& - c SI_i + \frac{X Y_i}{X + \sum_k Y_k} - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS\\
&- \sum_k P_k E_k SI_i - \lambda_i SI_i  \\
II_{ij}& - c II_{ij} +(\frac{1}{2})^{\delta_{ij}} \frac{Y_i Y_j}{X + \sum_k Y_k} + (\frac{1}{2})^{\delta_{ij}} \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
&+ (\frac{1}{2})^{\delta_{ij}} \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
\end{aligned}
\end{equation}

\subsection*{Model 5}

Model 5 is an implicit instantaneous partnership formation model, which uses adjusted transmission rate, $\beta'$, that is derived from Hollingsworth et al's approximated basic reproduction number:

\begin{equation}
\beta'_i = \frac{c \beta_i}{c + \beta_i + \lambda_i}.
\end{equation}
Thus, we get the following model:

\begin{equation}
\begin{aligned}
S' & = \sum_k \lambda_k I_k - \sum_k \beta'_k S I_k\\
I_i & = \sum_k M_{ki} \beta'_k S I_k - \lambda_i I_i
\end{aligned}
\end{equation}

\subsection*{Initial distribution of infected individuals}

We follow Champredon et al's result to calculate the initial distribution of infected individuals. For model 1 and 2, we have disease equilibrium state of $S^* = \frac{c}{c + \rho}$ and $SS^* = \frac{1-S^*}{2}$. We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with $\log_{10}$ SPVL of $i$. $Y_i$ is taken from normal distribution with mean 3 and is normalized so that $\sum_i D_i = 1$. Then, we have the following initial distribution of each states:

\begin{equation}
\begin{aligned}
S(0) &= (1 - \epsilon) S^* \\
SS(0) &= (1 - \epsilon)^2 SS^*\\
SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i\\
I_i(0) &=  \epsilon S^* D_i\\
II_{ij}(0) &=  (\frac{1}{2})^{\delta_{ij}} 2\epsilon^2 SS^* D_i D_j.
\end{aligned}
\end{equation}
Since model 3 and 4 do not have single state, $SS^*=1$ at the disease free equilibrium and the initial distribution becomes as follows:

\begin{equation}
\begin{aligned}
SS(0) &= (1 - \epsilon)^2 SS^*\\
SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i\\
II_{ij}(0) &=  (\frac{1}{2})^{\delta_{ij}} 2\epsilon^2 SS^* D_i D_j.
\end{aligned}
\end{equation}
Lastly, as model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution.

\begin{equation}
\begin{aligned}
S(0) &= 1 - \epsilon\\
I_i(0) &=  \epsilon D_i.
\end{aligned}
\end{equation}

\bibliographystyle{chicago}
\bibliography{virulence}
\end{document}

### 2016-03-12 ms/ms_parms.tex
manuscript, sim data

\begin{table}[ht]
\caption{Parameter ranges/values}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate uncoupled & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 & \cite{hollingsworth_hiv1_2008} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 & \cite{hollingsworth_hiv1_2008} \\
$\beta_{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$D_{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\alpha_{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\alpha_{max}$ & Maximum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

### 2016-03-12 ms/ms.Rnw
adding figures

@@
 library("knitr")
-opts_chunk$set(echo=FALSE,error=FALSE,fig.width=4,fig.height=4) #
+opts_chunk$set(echo=FALSE,error=FALSE,fig.width=8,fig.height=4) #
 @
@@
 
-<<panel3>>=
+<<panel3,fig.cap="Time plot (3-panel plot): baseline dynamics, illustrating summary statistics (peak time, eq virulence, length of time until peak virulence).
+Full Shirreff model, effects of $I(0)$, $\\alpha(0)$, $r$",dev="tikz",fig.width=10>>=
+for (i in c("","2","3")) 
+    load(sprintf("../simdata/ev_LHS_resS%s.rda",i))
+Ls <- load("../simdata/shirreff_data.rda")
+tvec <- seq(from=1,by = 0.1,length.out=nrow(vir_matS))
+dimnames(vir_matS) <- list(Time=tvec,
+                           run=paste0("r=",c(0.021,0.042,0.084)))
 
+d_littleR_m <- melt(vir_matS,value.name="Mean.VL")
+shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
+shirreff_df_m$run <- "Shirreff"
+df_tot <- rbind(shirreff_df_m,d_littleR_m)
+g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
+    geom_line()+labs(x="time",y="mean virulence ($\\log_{10}$ SPVL)")
+###
+dimnames(vir_matS2) <- list(Time=tvec,
+               run=c("$10^{-3}$", "$10^{-4}$", "$10^{-5}$"))
+g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL")
+###
+dimnames(vir_matS3) <- list(Time=tvec,
+                            run=c("$2.5$", "$3$", "$3.5$"))
+g3 <- g1 %+% melt(vir_matS3,id.var="Time",value.name="Mean.VL")
+limits <- c(2.5,4.7)
+breaks <- seq(limits[1],limits[2], by = 0.5)
+g1.y <- g1 + scale_y_continuous(limits = limits, breaks = breaks)+
+    scale_colour_discrete(name="run")
+g2.y <- g2 + scale_y_continuous(limits = limits, breaks = breaks)+
+    scale_colour_discrete(name="$I(0)$")
+g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
+        scale_colour_discrete(name="$\\alpha(0)$")
+grid.arrange(g1.y, g2.y, g3.y, nrow=1)
 @
+
+<<virtraj,fig.cap="Envelopes of virulence trajectories under all models">>=
+gg_virtraj <- ggplot(sum_list[["vir_mat"]],
+       aes(tvec,mean,ymin=lwr,ymax=upr))+
+    geom_line(aes(colour=model,linetype=model),lwd=1.5)+
+    geom_ribbon(aes(fill=model),alpha=0.3)
+print(gg_virtraj)
+@
+
+<<unidist,fig.cap="Univariate distributions of summary statistics",fig.width=6,fig.height=6>>=
+sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
+mL <- melt(sL,id.vars=c("model","run"))
+ggplot(mL,aes(value,model,fill=model))+
+    geom_violinh(scale="area")+
+    facet_wrap(~variable,scale="free")+
+    labs(y="")
+@
+
+Figure to-do:
 \begin{itemize}
-\item Time plot (3-panel plot): baseline dynamics, illustrating summary statistics (peak time, eq virulence, length of time until peak virulence).
-Full Shirreff model, effects of $I(0)$, $\alpha(0)$, $r$
-\item Virulence trajectories under all models
+\item fig 1. move legends into panels: tweak size? remove redundant y axes?
+\item fig 3. add random-mixing model
+\item fig 4. re-order, figure out how to tweak random density width
+\end{itemize}
+
+\begin{itemize}
 \item Univariate plots

### 2016-03-12 ms/ms.Rnw
fig tweaking

--- 2016-03-12 ms/ms.Rnw
@@
 
-<<panel3>>=
+<<panel3,fig.cap="Time plot (3-panel plot): baseline dynamics, illustrating summary statistics (peak time, eq virulence, length of time until peak virulence).
+Full Shirreff model, effects of $I(0)$, $\\alpha(0)$, $r$",dev="tikz",fig.width=10>>=
+for (i in c("","2","3")) 
+    load(sprintf("../simdata/ev_LHS_resS%s.rda",i))
+Ls <- load("../simdata/shirreff_data.rda")
+tvec <- seq(from=1,by = 0.1,length.out=nrow(vir_matS))
+dimnames(vir_matS) <- list(Time=tvec,
+                           run=paste0("r=",c(0.021,0.042,0.084)))
 
+d_littleR_m <- melt(vir_matS,value.name="Mean.VL")
+shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
+shirreff_df_m$run <- "Shirreff"
+df_tot <- rbind(shirreff_df_m,d_littleR_m)
+vlab <- "mean virulence ($\\log_{10}$ SPVL)"
+g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
+    geom_line()+labs(x="time",y=vlab)
+###
+dimnames(vir_matS2) <- list(Time=tvec,
+               run=c("$10^{-3}$", "$10^{-4}$", "$10^{-5}$"))
+g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL")
+###
+dimnames(vir_matS3) <- list(Time=tvec,
+                            run=c("$2.5$", "$3$", "$3.5$"))
+g3 <- g1 %+% melt(vir_matS3,id.var="Time",value.name="Mean.VL")
+limits <- c(2.5,4.7)
+breaks <- seq(limits[1],limits[2], by = 0.5)
+g1.y <- g1 + scale_y_continuous(limits = limits, breaks = breaks)+
+    scale_colour_discrete(name="run")
+g2.y <- g2 + scale_y_continuous(limits = limits, breaks = breaks)+
+    scale_colour_discrete(name="$I(0)$")
+g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
+        scale_colour_discrete(name="$\\alpha(0)$")
+grid.arrange(g1.y, g2.y, g3.y, nrow=1)
 @
+
+<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",dev="tikz",fig.width=6,cache=TRUE>>=
+gg_virtraj <- ggplot(sum_list[["vir_mat"]],
+       aes(tvec,mean,ymin=lwr,ymax=upr))+
+    geom_line(aes(colour=model,linetype=model),lwd=1.5)+
+    geom_ribbon(aes(fill=model),alpha=0.3)+
+    labs(x="time",y=vlab)
+print(gg_virtraj)
+@
+
+<<unidist,fig.cap="Univariate distributions of summary statistics",fig.width=8,fig.height=4.8>>=
+sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
+mL <- melt(sL,id.vars=c("model","run"))
+ggplot(mL,aes(value,model,fill=model))+
+    geom_violinh(scale="area")+
+    facet_wrap(~variable,scale="free")+
+    labs(y="")
+@
+
+Figure to-do:
 \begin{itemize}
-\item Time plot (3-panel plot): baseline dynamics, illustrating summary statistics (peak time, eq virulence, length of time until peak virulence).
-Full Shirreff model, effects of $I(0)$, $\alpha(0)$, $r$
-\item Virulence trajectories under all models
+\item fig 1. move legends into panels: tweak size? remove redundant y axes?
+\item fig 3. add random-mixing model
+\item fig 4. re-order, figure out how to tweak random density width
+\end{itemize}
+
+\begin{itemize}
 \item Univariate plots

### 2016-03-16 ms/ms.Rnw
playing with figures

--- 2016-03-12 ms/ms.Rnw
@@
 
-<<panel3>>=
+<<panel3,fig.cap="Baseline dynamics.
+Time series of mean population virulence ($\\log_{10}$ SPVL).
+(a) Shirreff model, effects of varying $r$.
+(b) Effects of varying initial infectious density $I(0)$.
+(c) Effects of varying initial mean virulence $\\alpha(0)$",dev="tikz",fig.height=3,fig.width=8,cache=TRUE>>=
+for (i in c("","2","3")) 
+    load(sprintf("../simdata/ev_LHS_resS%s.rda",i))
+Ls <- load("../simdata/shirreff_data.rda")
+tvec <- seq(from=1,by = 0.1,length.out=nrow(vir_matS))
+dimnames(vir_matS) <- list(Time=tvec,
+                           run=paste0("$r=",c(0.021,0.042,0.084),"$"))
 
+d_littleR_m <- melt(vir_matS,value.name="Mean.VL")
+shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
+shirreff_df_m$run <- "Shirreff"
+df_tot <- rbind(shirreff_df_m,d_littleR_m)
+vlab <- "mean virulence ($\\log_{10}$ SPVL)"
+tlab <- "time (years)"
+g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
+    geom_line()+labs(x=tlab,y=vlab)+
+    theme(legend.position=c(0.8,0.2))
+###
+dimnames(vir_matS2) <- list(Time=tvec,
+               run=c("$10^{-3}$", "$10^{-4}$", "$10^{-5}$"))
+g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL")
+###
+dimnames(vir_matS3) <- list(Time=tvec,
+                            run=c("$2.5$", "$3$", "$3.5$"))
+g3 <- g1 %+% melt(vir_matS3,id.var="Time",value.name="Mean.VL")
+limits <- c(2.5,4.7)
+breaks <- seq(limits[1],limits[2], by = 0.5)
+g1.y <- g1 + scale_y_continuous(limits = limits, breaks = breaks)+
+    scale_colour_discrete(name="run")
+g2.y <- g2 + scale_y_continuous(limits = limits, breaks = breaks)+
+    scale_colour_discrete(name="$I(0)$")
+g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
+        scale_colour_discrete(name="$\\alpha(0)$")
+grid.arrange(g1.y, g2.y, g3.y, nrow=1)
 @
+
+<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",dev="tikz",fig.width=6,cache=TRUE>>=
+gg_virtraj <- ggplot(sum_list[["vir_mat"]],
+       aes(tvec,mean,ymin=lwr,ymax=upr))+
+    geom_line(aes(colour=model,linetype=model),lwd=1.5)+
+    geom_ribbon(aes(fill=model),alpha=0.3)+
+    labs(x="time",y=vlab)
+print(gg_virtraj)
+@
+
+<<unidist,fig.cap="Univariate distributions of summary statistics. Note: point for eqvir/random is a placeholder, will have to be explained",fig.width=8,fig.height=4.8>>=
+sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
+mL <- melt(sL,id.vars=c("model","run"))
+## horrible hack (but doesn't help); subsample
+## w <- which(mL$model=="random")
+## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
+w <- with(mL,which(model=="random" & variable=="eq_vir"))
+rval <- mean(mL$value[w])
+mLw <- droplevels(mL[-w,])
+ggplot(mLw,aes(value,model,fill=model))+
+    geom_violinh(width=1)+
+    facet_wrap(~variable,scale="free_x")+
+    guides(fill=guide_legend(reverse=TRUE))+
+    labs(y="")+
+    geom_point(data=data.frame(model="random",variable="eq_vir",
+                               value=rval),pch=22,size=3,show.legend=FALSE)
+@
+
+<<testrev,echo=FALSE,eval=FALSE>>=
+set.seed(101)
+zz <- data.frame(x=rnorm(3000),f=factor(rep(letters[1:3],each=1000)))
+g0 <-  ggplot(zz,aes(x=f,y=x,fill=f))+geom_violin()
+grid.arrange(
+    g0,
+    g0+ guides(fill=guide_legend(reverse=TRUE)),
+    nrow=2)
+@
+
+Figure to-do:
 \begin{itemize}
-\item Time plot (3-panel plot): baseline dynamics, illustrating summary statistics (peak time, eq virulence, length of time until peak virulence).
-Full Shirreff model, effects of $I(0)$, $\alpha(0)$, $r$
-\item Virulence trajectories under all models
+\item fig 1. add labels? tweak size? remove redundant y axes?
+\item fig 3. add random-mixing model
+\item fig 4. re-order (reverse legend order), figure out how to tweak random density width
+\end{itemize}
+
+\begin{itemize}
 \item Univariate plots
@@
 \end{equation}
+
+\todo{explain non-standard use of Kronecker delta as exponent!}
 

### 2016-03-18 ms/ms.Rnw
still working

@@
 \newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
+\newcommand{\rzero}{{\cal R}_0}
 \usepackage{changes}
@@
 
-\todo{BMB: add to this description}
-In disease modeling, one of the most important parameter is the initial exponential growth rate.
-As it is one of the few parameters that can be directly observed, many modelers try to match the initial exponential growth rate to correctly simulate the epidemic.
-Furthermore, evolution of virulence is sensitive to the initial growth rate.
-Thus, we scale a parameter so that every run has an equal exponential rate of growth rate.
+One of the most difficult parts of our model comparison exercise is coming up with sets of parameters that are commensurate against many different model structures.
+For the most part, our models are too complex to easily derive analytical correspondences among them.
+Given a numerical criterion, such as $r$ (initial exponential growth rate) or $\rzero$ (intrinsic reproductive number), we can adjust one or more parameters by brute force to ensure that all of the models match according to that criterion.
+While $\rzero$ is often considered the most fundamental property of an epidemic, and might thus seem to be a natural goal, here we focus on matching the initial growth rate $r$ for several reasons.
+First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than $\rzero$ (CITE?).
+Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves (CITE Ma et al.), while $\rzero$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease (CITE Wallinga, Teunis, Lipsitch, etc.) Thus, we scale a parameter so that every run has an equal exponential rate of growth rate.
 
@@
 Then, paraeters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.
+
+\todo{mention that baseline parameters are geometric means of ranges}
 
@@
 \section{Results}
+
+Our simplifications of \shir's model reproduce the qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
+Our model produces lower peak virulence ($\approx ?$ vs.
+$\approx ?$) and equilibrium virulence $\approx ?$ vs.
+$\approx?$) than Shirreff's, probably because \ldots \todo{describe more; is this because we can't match both R0 and r?
+Can we match r exactly?}.
+Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases and allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b).
+Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence.
+
+For the baseline set of parameters, with the initial epidemic spread rate scaled to be the
+\todo{just double-checking: Figure 1 shows results from the implicit model, right?}
 
@@
 breaks <- seq(limits[1],limits[2], by = 0.5)
+afun <- function(label,size=8) {
+    annotate(geom="text",label=label,x=-Inf,y=Inf,
+             size=size,vjust=1,hjust=0)
+}
 g1.y <- g1 + scale_y_continuous(limits = limits, breaks = breaks)+
-    scale_colour_discrete(name="run")
+    scale_colour_discrete(name="run")+afun("a")
 g2.y <- g2 + scale_y_continuous(limits = limits, breaks = breaks)+
-    scale_colour_discrete(name="$I(0)$")
+    scale_colour_discrete(name="$I(0)$")+afun("b")
 g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
-        scale_colour_discrete(name="$\\alpha(0)$")
+        scale_colour_discrete(name="$\\alpha(0)$")+afun("c")
 grid.arrange(g1.y, g2.y, g3.y, nrow=1)
@@
 \begin{itemize}
-\item fig 1. add labels? tweak size? remove redundant y axes?
-\item fig 3. add random-mixing model
+\item fig 1. tweak size? remove redundant y axes/labels?
+\item fig 3. add random-mixing model; remove x-axis label; rename factor to make strip labels prettier
 \item fig 4. re-order (reverse legend order), figure out how to tweak random density width

### 2016-03-25 ms/ms.Rnw
tweaks

@@
 
-One of the most difficult parts of our model comparison exercise is coming up with sets of parameters that are commensurate against many different model structures.
+One of the most difficult parts of model comparison is finding parameter sets that are commensurate against many different model structures.
 For the most part, our models are too complex to easily derive analytical correspondences among them.
@@
 As adjusting either partnership change rate (i.e.
-partnership formation and dissolution rate) or transmission rate does not fulfill this requirement for certain models, we decided to scale partnership change rate and dissolution rate by the same factor of $\gamma$: $\tsub{beta}{adj} =  \gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$, $\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$.
+partnership formation and dissolution rate) or transmission rate does not fulfill this requirement for certain models, we decided to scale partnership change rate and dissolution rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =  \gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$, $\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$.
 Since transmission rate is adjusted by the scale of $\gamma$, uncoupled and extra-couple transmission rates are adjusted as well.
@@
 
-We run each model for 1000 times with different parameter sets calculated from latin hypercube sampling with fixed starting condition: mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$ (refer to appendix).
-After each run, initial exponential growth rate is calculate.
-Then, paraeters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.
+We run each model for 1000 times with different parameter sets calculated from Latin hypercube sampling [CITE Blower et al], with fixed starting conditions
+of mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$ (refer to appendix).
+An  initial run of After each run, initial exponential growth rate is calculated.
+Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.
 

### 2016-03-25 ms/ms.Rnw
still working on methods ...

@@
 \usepackage{amssymb}
+\usepackage[usenames,dvipsnames]{xcolor} %% for non-standard cite color
 \usepackage{graphicx}
-\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = blue]{hyperref}
+\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = MidnightBlue]{hyperref}
 \usepackage{url}
@@
 
-\newcommand{\shir}{\citeauthor{shirreff_transmission_2011}}
-\newcommand{\champr}{\citeauthor{champredon_hiv_2013}}
+\newcommand{\shir}{Shirreff et al.}%{\citeauthor{shirreff_transmission_2011}}
+\newcommand{\champr}{Champredon et al.}%{\citeauthor{champredon_hiv_2013}}
 \newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
@@
 
-How pathogen virulence evolves is a fundamental question in evolutionary biology, of great theoretical and, potentially, practical importance.
-While various researchers have offered critiques [CITE Ebert and Bull, Alizon and Michalakis], the trade-off theory --- which postulates that parasite virulence can be explained as the long-term evolutionary outcome of a saturating relationship between parasite clearance rate and transmission rate --- has also been successfully applied in several important cases [CITES?].
-One particularly interesting application of these ideas is the work by Fraser et al [CITES] showing that HIV appears to satisfy the prerequisites of the tradeoff theory: in a study of discordant couples (i.e.
-long-term sexual partnerships with one infected and one uninfected partner), HIV virulence as measured by the rate of progression to AIDS was both heritable and covaried  with the set-point viral load ([DEF]), which in turn predicted the probability of within-couple transmission.
-Subsequent studies [CITES] used these data to parameterize mechanistic models of HIV virulence evolution, suggesting that HIV invading a novel population would initially evolve increased virulence, peaking after approximately [XXX] years and then declining slightly to a long-stable virulence level.
+How pathogen virulence evolves is a fundamental question in evolutionary biology, of both theoretical and (potentially) practical importance.
+The trade-off theory \citep{Ebert1999} --- which postulates that parasite virulence can be explained as the long-term evolutionary outcome of a saturating relationship between parasite clearance rate and transmission rate --- has been criticized \citep{EbertBull2003,alizon_adaptive_2015}, but has also been successfully applied in several important cases \citep{Dwyer+1990} \todo{others?
+de Roode and Altizer?}.
+One particularly interesting application of these ideas is the work by Fraser et al.
+showing that HIV appears to satisfy the prerequisites of the tradeoff theory: in a study of discordant couples (i.e.
+long-term sexual partnerships with one infected and one uninfected partner), HIV virulence as measured by the rate of progression to AIDS was both heritable and covaried with the set-point viral load (i.e., the characteristic virus load measured in blood during the intermediate stage of infection), which in turn predicted the probability of within-couple transmission \citep{Fraser+2007,fraser_virulence_2014}\todo{read/check fraser+ 2014!; check vs.
+\cite{herbeck_is_2012}}.
+Subsequent studies \citep{shirreff_transmission_2011,herbeck_hiv_2014} used these data to parameterize mechanistic models of HIV virulence evolution, suggesting that HIV invading a novel population would initially evolve increased virulence, peaking after approximately [XXX] years and then declining slightly to a long-stable virulence level.
 
@@
 or physiological level, the epidemiological structures used in these
-models are relatively simple. As we discuss in detail below, the
+models are relatively simple. \todo{don't forget to check other recent papers: esp \cite{payne_impact_2014}} As we discuss in detail below, the
 existing models of HIV eco-evolutionary dynamics either use implicit
@@
 \shir's multi-strain evolutionary model.
+Like \shir, we use a simple susceptible-infected-susceptible demographic formulation; rather than modeling birth and death (or more specifically, recruitment into the sexually active population and death), we assume that whenever an individual dies from infection, another enters the susceptible compartment.
+
+\subsubsection{Infection dynamics}
+
+\todo{was going to call this ``within-host'' but it includes transmission}
+
 Like \shir, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than virulence (i.e. rate of progression to AIDS) itself. 
-Instead of accounting for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), we use a single-stage disease model, with a simple exponentially distributed infectious period instead of the Weibull-distributed infectiousness which \shir\ (and many other HIV models) use.
-However, we do account for varying transmission rates and durations of each disease stage by summing the durations of three stages
-(again based on \shir's model) and taking the duration-weighted average of transmission rates of three stages. Thus the within-couple transmission rate, $\beta$, for our models is given by:
+In contrast to \shir, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period.
+We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on \shir's model) and taking the duration-weighted average of transmission rates of three stages.
+Thus the within-couple transmission rate, $\beta$, for our models is given by:
 \begin{equation}
@@
 
-Like \shir\ but unlike \champr, we simplify the evolutionary model by using a one-to-one genotype-phenotype mapping (i.e.
-a single strain can only produce a single SPVL).
-Unlike \champr\ but like \shir, we also use a simple susceptible-infected-susceptible formulation; rather than modeling birth and death (or more specifically, recruitment into the sexually active population and death), we assume that whenever an individual dies from infection, another enters the susceptible compartment.
+\subsubsection{Mutation}
 
-We developed six multi-strain evolutionary models, each of which is based on different assumptions regarding partnership dynamics.
-Specifically, we focus on the effect of instantaneous partnership formation and extra couple mixing on the evolution of mean $\log_{10}$ SPVL.
-
-The first four models consider explicit partnership dynamics and are based on Champredon et al's model.
-Model 1 and 2 assume non-instantaneous partnership formation and consist of five states that are classified by infection status and partnership status.
-$S$ is the number of single susceptible individuals, and $I$ is the number of single infected individuals.
-$SS$ is the number of susceptible-susceptible couples, $SI$ is the number of serodiscordant couples, and $II$ is the number of concordant positive couple.
-Model 1 includes extra couple mixing and uncoupled mixing whereas model 2 only considers within-couple transmission.
-Model 3 and 4 assume instantaneous partnership formation and consist of 3 states: $SS$, $SI$, and $II$.
-Like model 1 and 2, model 3 includes extra couple mixing and model 4 only considers within-couple transmission.
-
-In contrast, models 5 and 6 are not explicitly structured.
-Model 5 is an implicit serial monogamy model based on \shir's model.
-It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
-However, to simulate the effect instantaneous partnership formation, it uses adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model.
-Finally, model 6 is a simple random-mixing model.
-
+Like \shir\ we incorporate a between-host mutation process in the SPVL, but simplify \shir's evolutionary model slightly by using a one-to-one genotype-phenotype mapping. 
 The mutational process in our model is directly taken from \shir.
@@
 \begin{equation}
-\alpha_i =  (\alpha_{max} - \alpha_{min})\frac{(i-1)}{n-1} + \alpha_{min} \qquad i = 1,2,3, \dots n.
+\alpha_i =  (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{(i-1)}{n-1} + \tsub{\alpha}{min} \qquad i = 1,2,3, \dots n.
 \end{equation}
@@
 
-where $\Phi(j)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\alpha_{\max} - \alpha_{\min})/n$.
+where $\Phi(j)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/n$.
 Transmission rate and disease induced mortality rates are discretized into a vector as well:
@@
 
-\subsection{Latin Hypercube Sampling - we should probably explain model parameters before this...?}
+\subsubsection{Contact structure and partnership dynamics}
 
-Not only is there discrepancy in parameters regarding partnership dynamics between sources, but it is also difficult to determine the exact value due to many reasons.
-Furthermore, recent studies suggest that the previously measured transmission rate of the acute phase was over-estimated.
-Thus, we want to allow for uncertainties in model parameters and perform Latin hyper cube sampling on the following parameters: $\beta_P$, $D_P$, $\beta_D$, $D_D$, $\rho$, $c$, $c_u/c_w$, and $c_e/c_w$.
-Each parameter is divided  We do not allow for uncertainties in parameters that are directly related to evolutionary process.
+We developed six multi-strain evolutionary models, designed to cover a gamut between \champr's relatively realistic and \shir's relatively simplistic epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
+non-instantaneous partnership formation and (2) zero vs.
+positive extra-partnership sexual contact and transmission on the evolution of mean $\log_{10}$ SPVL.
+
+Our first four models consider explicit partnership dynamics and are based on \champr's model.
+Model 1 and 2 assume non-instantaneous partnership formation (i.e.
+individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status.
+$S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals.
+$SS$ is the number of susceptible-susceptible couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples.
+Model 1 includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas model 2 only considers within-couple transmission.
+Models 3 and 4 assume instantaneous partnership formation and thus consist of only the three partnered states: $SS$, $SI$, and $II$.
+Parallel to model 1 and 2, model 3 includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and model 4 only considers within-couple transmission.
+
+In contrast, models 5 and 6 are not explicitly structured.
+Model 5 is an implicit serial monogamy model based on the epidemiological model used by \shir.
+It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
+However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008} \todo{is this the right ref?
+where does Shirreff's model originate?}.
+Finally, model 6 is a simple random-mixing model.
+
+\todo{I think we need a section here to describe partnership dynamics/contact structure, i.e. define the $\rho$, $c_u$, $c_e$, $c_w$ parameters - will be similar to \champr}
+\input{ms_parms.tex}
+
+\subsection{Latin hypercube sampling}
+
+Despite considerable effort by many researchers \citep[e.g.][]{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led \champr\ to adopt a Latin hypercube sampling (LHS) strategy \citep{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
+In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
+We do not allow for uncertainties in parameters that are directly related to evolutionary process \todo{i.e., ...
+?}
+
+Latin hypercube sampling is done as in \champr.
+For each parameter, $z$, its range is divided into $N$ \todo{give the number we actually used here} equal intervals on a log scale, where $N$ is the number of total runs:
+
+\begin{equation}
+z_i = \text{exp}(\log(z_{min}) + [\log(z_{max}) - \log(z_{min})] \frac{i-1}{N-1}) \qquad i = 1,...,N.
+\end{equation}
+For simplicity, we assume that all parameters follow logarithmic distribution.
+Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$).
+Then, each column is replaced with a random permutation series of itself.
+Now, each row contains a different parameter set that is used for each simulation run.
+
+Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
+Parameters ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from \champr, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from \cite{hollingsworth_hiv1_2008}.
+Rest of the parameters are taken from \shir.
+
+As we introduce a new parameter, $c_u/c_w$, to model uncoupled and extra-couple transmission dynamics more easily within multi-strain models (refer to appendix), it is important to pick a reasonable range for it.
+\champr assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
+Given \champr's parameter range, it is trivial that the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5.
+Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$.
+Although this adds more uncertainty to the parameter $c_u$ --- \champr's range gives 5 fold difference whereas ours give 25 fold difference --- as there is not much known about the uncoupled transmission rate, it is reasonable to use a wider range.
 
 \todo{refer to Table 1 and describe where the new parameters come from; e.g. decisions we made about new scaling parameters}
-
-\input{ms_parms.tex}
 
@@
 
-For each model we derive the following summary statistics:
+\todo{Do we need to talk about the baseline parameters?
+We never use it...
+We only use it for the first set of graphs but we need to mention that we are using Shirreff's partnership rate just for comparison...So we actually never use baseline parameters...}
+
+For each model we derive the following summary statistics: peak virulence, peak time, equilibrium virulence, and relative peak virulence.
+The transient phase of an epidemic is often characterized by high virulence, and we define peak virulence as the maximum virulence during this phase.
+It is simply calculated by taking the maximum value from the virulence trajectory, and peak time is the time at which the maximum value is reached.
+Once the epidemic enters the endemic phase, evolution of virulence stabilizes and reaches equilibrium.
+Equilibrium virulence is calculated by taking the mean virulence at 4000 years.
+Although most simulations reach equilibrium much earlier, we set our time horizon at a much later date as some simulation runs have slow rate of evolution depending on the parameter set and model assumptions.
+
+We focus on these statistics for the following two reasons.
+First of all, knowing the possible ranges for the peak virulence allows us to estimate the worst-case scenario for the HIV and other sexually transmitted disease epidemics.
+When an epidemic begins, it is often the case that the pathogen has already evolved towards high virulence by the time it is observed [CITE].
+Understanding how virulent a pathogen can evolve before an epidemic begins can be helpful for controlling the disease.
+Furthermore, knowing the initial virulence, peak virulence, timing of the peak virulence, and equilibrium virulence provide sufficient detail to identify the shape of the virulence trajectory.
+During an epidemic outbreak, it is difficult to observe virulence evolution.
+Specifically, in the case of HIV and other sexually transmitted diseases, slow evolutionary time-scale makes observing changes in the mean virulence even more challenging.
+Knowing the ranges of these statistics can help real-time virulence evolution prediction during an epidemic less troublesome.
 
@@
 $\approx ?$) and equilibrium virulence $\approx ?$ vs.
-$\approx?$) than Shirreff's, probably because \ldots \todo{describe more; is this because we can't match both R0 and r?
-Can we match r exactly?}.
+$\approx?$) than Shirreff's, probably because \ldots \todo{describe more; is this because we
+can't match both R0 and r? Can we match r exactly?}. (We're matching epidemic curve, not virulence curve!)
+
   Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases and allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b).
@@
  \todo{just double-checking: Figure 1 shows results from the implicit model, right?}
+
+ Throughout Fig 1 we're using Shirreff's parameters (Beta1 = 2.76, Duration1 = 0.25, Beta3 = 0.76, Duration3 = 0.75, cmeanbase = 1.25, alphaDist = c(min = 2, max = 7,  delta = 0.1))
 
@@
 \item fig 1. tweak size? remove redundant y axes/labels?
-\item fig 3. add random-mixing model; remove x-axis label; rename factor to make strip labels prettier
-\item fig 4. re-order (reverse legend order), figure out how to tweak random density width
-\end{itemize}
+\item fig 2. add random-mixing model; remove x-axis label; rename factor to make strip labels prettier
+\item fig 3. re-order (reverse legend order), figure out how to tweak random density width
+\item fig 4. pairs plot
+\item fig 5.
+sensitivity plot (unscaled plot); take out plots that are not related ...
+implicit model doesn't use rho, cu or ce; random doesn't use rho, cmean?, cu, ce; instswitch doesn't have rho or cu or ce; models without epc don't have ce or cu; instswitch + epc doesn't have rho and cu (no uncoupled compartment)
 
-\begin{itemize}
-\item Univariate plots
-\item Bivariate plots
-\item Unscaled sensitivity 
 \end{itemize}
@@
 \end{equation}
+
+In our models, we introduce a non-standard use of Kronecker delta as exponent, such as $2^{\delta_{ij}}$.
+In a single-strain model, both individuals in a concordant partnership, $II$, enter single-infected compartment, $I$, when a partnership dissolves: $I' = 2 c II'$.
+However, this is not the case in a multi-strain model.
+If infected individuals in an $II$ partnership have different strains, they will enter different compartment.
+On the other hand, if both partners in an $II$ partnership are infected with same strains, both will enter the same compartment.
+Therefore, we use Kronecker delta as exponent to distinguish such difference.
+Further details are explain below.
 
@@
 
+\subsection*{Model 6}
+
+Model 6 is a random mixing model. It is modeled in a same way as model 6 without the adjusted transmission rate:
+
+\begin{equation}
+\begin{aligned}
+S' & = \sum_k \lambda_k I_k - \sum_k \beta_k S I_k\\
+I_i & = \sum_k M_{ki} \beta_k S I_k - \lambda_i I_i
+\end{aligned}
+\end{equation}
+
 \subsection*{Initial distribution of infected individuals}
@@
 \end{equation}
+Model 6 also shares the same distribution of initial infected individuals as model 5.
 

### 2016-03-25 ms/ms_parms.tex
still working on methods ...

@@
 $D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 & \cite{hollingsworth_hiv1_2008} \\
-$\beta_{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
+$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
 $\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 & \cite{shirreff_transmission_2011} \\
 $\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
-$D_{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
+$\tsub{D}{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
 $D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 & \cite{shirreff_transmission_2011} \\
@@
 $\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
-$\alpha_{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
-$\alpha_{max}$ & Maximum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
+$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
+$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
 $n$ & Number of strains & 21 & Assumption\\

### 2016-03-28 ms/ms.Rnw
add figs, tweak

@@
 \usepackage{amssymb}
-\usepackage[usenames,dvipsnames]{xcolor} %% for non-standard cite color
 \usepackage{graphicx}
-\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = MidnightBlue]{hyperref}
 \usepackage{url}
 \usepackage{enumerate}
+\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = MidnightBlue]{hyperref}
 \usepackage[left=50pt,
@@
 opts_chunk$set(echo=FALSE,error=FALSE,fig.width=8,fig.height=4) #
+## https://github.com/yihui/tikzDevice/issues/46 
+local({
+    hook_document <- knitr::knit_hooks$get('document')
+    knitr::knit_hooks$set(document = function(x){
+        res <- hook_document(x)
+        gsub('\\usepackage[]{color}', '\\usepackage[usenames,dvipsnames]{xcolor}', res, fixed = TRUE)
+  })
+})
 @
@@
 library("MASS")
-scale_colour_discrete <- function(...,palette="Set1") scale_colour_brewer(...,palette=palette)
-scale_fill_discrete <- function(...,palette="Set1") scale_fill_brewer(...,palette=palette)
+## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
 library("gridExtra")
@@
 
-\todo{define each summary statistic and explain what it tells us}
-\begin{description}
-\item[peak time] ...
-\item[peak virulence] ...
-\item[relative peak virulence] ...
-\item[equilibrium virulence] ...
-\end{description}
-
 \section{Results}
 
-Our simplifications of \shir's model reproduce the qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
-Our model produces lower peak virulence ($\approx ?$ vs.
+Our simplifications of \shir's model reproduce the qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in \shir's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
+However, our model produces lower peak virulence ($\approx ?$ vs.
 $\approx ?$) and equilibrium virulence $\approx ?$ vs.
-$\approx?$) than Shirreff's, probably because \ldots \todo{describe more; is this because we
-can't match both R0 and r? Can we match r exactly?}. (We're matching epidemic curve, not virulence curve!)
+$\approx?$) than Shirreff's, even for matching initial incidence (i.e., $r=0.042$).
 
 Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases and allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b).
-Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence.
+Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Figure~\ref{fig:panel3}c).
 
- For the baseline set of parameters, with the initial epidemic spread rate scaled to be the
- \todo{just double-checking: Figure 1 shows results from the implicit model, right?}
+\todo{note that in Fig 1 we're using Shirreff's parameters}
+\todo{add Shir. params to table 1: Beta1=2.76, Dur1 = 0.25, Beta3 = 0.76, Dur3 = 0.75, c = 1.25, alpha=\{2,7,0.1\}}
 
- Throughout Fig 1 we're using Shirreff's parameters (Beta1 = 2.76, Duration1 = 0.25, Beta3 = 0.76, Duration3 = 0.75, cmeanbase = 1.25, alphaDist = c(min = 2, max = 7,  delta = 0.1))
+<<setup2,echo=FALSE>>=
+orig_sum_labs <- c("peak_time","peak_vir","eq_vir","rel_peak")
+new_sum_labs <- c("peak time","peak virulence","equil virulence","relative peak")
+fixfac2 <- function(x) factor(x,
+                       levels=orig_sum_labs,
+                       labels=new_sum_labs)
+@
 
@@
 
-<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",dev="tikz",fig.width=6,cache=TRUE>>=
+<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",fig.width=6,cache=TRUE,dev="tikz">>=
 gg_virtraj <- ggplot(sum_list[["vir_mat"]],
@@
 mLw <- droplevels(mL[-w,])
+mLw$variable <- fixfac2(mLw$variable)
 ggplot(mLw,aes(value,model,fill=model))+
@@
     labs(y="")+
-    geom_point(data=data.frame(model="random",variable="eq_vir",
+    geom_point(data=data.frame(model="random",variable="equil virulence",
                                value=rval),pch=22,size=3,show.legend=FALSE)
+@
+
+<<pairplot,echo=FALSE,fig.cap="pairs plot",fig.width=7,fig.height=7>>=
+## names(sL)[na.omit(match(names(sL),orig_sum_labs))] <- paste0("`",new_sum_labs,"`")
+ggp1 <- ggpairs(sL,
+        mapping = ggplot2::aes(color = model),
+        columns=3:6,
+        lower = list(continuous = wrap("points", alpha = 0.5,size=0.5)),
+        diag = list(continuous = "blankDiag"),
+        upper = list(continuous = "blank"))+
+    theme(panel.margin=grid::unit(0,"lines"))
+## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot
+## have to change plot one panel at a time
+cbscales <- list(scale_colour_brewer(palette = 'Dark2'),
+                 scale_fill_brewer(palette = 'Dark2'))  ## DRY
+for (row in seq_len(ggp1$nrow))
+    for (col in seq_len(ggp1$ncol))
+        ggp1[row, col] <- ggp1[row, col] + cbscales[[(row==col)+1]]
+## hack to remove top row/right column
+ggp1$nrow <- ggp1$ncol <- 3
+ggp1$plots <- ggp1$plots[c(5:7,9:11,13:15)]
+print(ggp1)
 @
@@
     nrow=2)
+@
+
+<<plot_sens,dev="png",echo=FALSE,fig.width=10,fig.height=6>>=
+ggplot(mL2,aes(LHSval,sumval,colour=model))+
+    geom_point(pch=".",alpha=0.5)+
+    facet_grid(sumvar~LHSvar,scales="free")+
+    geom_smooth(se=FALSE)+labs(x="",y="")+
+    scale_x_log10()
 @

### 2016-04-01 ms/ms.Rnw
working on ms

@@
 
-Our simplifications of \shir's model reproduce the qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in \shir's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
+Our simplifications of \shir's model reproduce its qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in \shir's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
 However, our model produces lower peak virulence ($\approx ?$ vs.
@@
 
-<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",fig.width=6,cache=TRUE,dev="tikz">>=
+Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (Figure~\ref{fig:virtraj}).
+Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 $\log_{10}$ SPVL), virulence peaks in all models between 200 and 300 years.
+
+<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",fig.width=6,cache=TRUE,message=FALSE,warning=FALSE,dev="tikz">>=
 gg_virtraj <- ggplot(sum_list[["vir_mat"]],
@@
     geom_ribbon(aes(fill=model),alpha=0.3)+
-    labs(x="time",y=vlab)
+    labs(x="time",y=vlab)+
+    scale_x_continuous(limits=c(0,2500))
 print(gg_virtraj)
@@
 \item fig 1. tweak size? remove redundant y axes/labels?
-\item fig 2. add random-mixing model; remove x-axis label; rename factor to make strip labels prettier
-\item fig 3. re-order (reverse legend order), figure out how to tweak random density width
-\item fig 4. pairs plot
+\item fig 2. remove x-axis label; rename factor to make strip labels prettier.  Log scale?
+\item fig 3.
+\item fig 4. pairs plot; legend?
 \item fig 5.

### 2016-05-03 ms/ms.Rnw
more ms updates

@@
 
+\abstract{The evolutionary dynamics of parasite virulence over the
+  course of an emerging epidemic have important implications both for
+  our basic understanding of epidemiological dynamics and,
+  potentially, for the outcomes of public health interventions. In
+  general changes in fitness landscapes over the course of an epidemic
+  will select for higher virulence during the early,
+  exponential-growth phase of the epidemic, but quantitative outcomes
+  can depend sensitively on biological details and the structure of
+  mathematical models used to capture them.  Fraser, Shirreff, and
+  co-workers have proposed a series of models for eco-evolutionary
+  dynamics of HIV that are relatively detailed in their portrayal of
+  the tradeoffs between transmission and virulence (mediated by
+  set-point viral load, SPVL) and their heritability between
+  hosts. However, these models use very simple implicit
+  representations of the transmission process that ignore the
+  partnership dynamics that previous research has found to be critical
+  in predicting epidemics of sexually transmitted diseases.  We
+  explore models that combine HIV virulence tradeoffs with a range of
+  epidemiological structures, modeling partnership formation and
+  dissolution and allowing for individuals to transmit disease outside
+  of partnerships. We assess summary statistics such as the peak value
+  of virulence (SPVL) and the time at which the peak occurs across all
+  models and across a Latin hypercube sample that captures a realistic
+  range of partnership dynamic parameters for sub-Saharan Africa. In
+  order to account for the different interpretations of parameters
+  across model structures, we scale all parameter sets to constrain
+  the simulated epidemic growth rate to be identical, matching a
+  realistic baseline value. Our primary result is that, for this
+  particular model setting, the simplest random-mixing structure is
+  actually the best approximation to the most realistic model; this
+  surprising outcome occurs because the dominance of extra-pair
+  contact in the realistic model tends to mask the effects of
+  partnership structure.
+}
 \section{Introduction}
@@
 
+pairs plot: instswitch +epc is the most different from others
+uncoupled mixing rate 1/5 to 5: rate of uncoupled contact relative
+to within-pair contact.  If it were much lower instswitch+epc would
+look more like instswitch *without* epc
+(uncoupled mixing acts like random mixing)
+
+instswitch is intermediate beween (pairform + implicit)
+
 <<testrev,echo=FALSE,eval=FALSE>>=
@@
 
-Figure to-do:
+Figure to-do: maybe go back to bw theme?
 \begin{itemize}
 \item fig 1. tweak size? remove redundant y axes/labels?
-\item fig 2. remove x-axis label; rename factor to make strip labels prettier.  Log scale?
-\item fig 3.
-\item fig 4. pairs plot; legend?
+\item fig 2. 6-panel figure? log scale? boundary lines instead of ribbons for edges of envelopes? move labels away from axes?
+\item fig 3. change order (Random after pairform)
+\item fig 4. add legend, fix axis labels; maybe transparent
 \item fig 5.
 sensitivity plot (unscaled plot); take out plots that are not related ...
-implicit model doesn't use rho, cu or ce; random doesn't use rho, cmean?, cu, ce; instswitch doesn't have rho or cu or ce; models without epc don't have ce or cu; instswitch + epc doesn't have rho and cu (no uncoupled compartment)
+implicit model doesn't use rho, cu or ce; random doesn't use rho, cmean?, cu, ce; instswitch doesn't have rho or cu or ce; models without epc don't have ce or cu; instswitch + epc doesn't have rho and cu (no uncoupled compartment).
+Maybe squash panels together, higher resolution
 

### 2016-05-03 ms/ms.Rnw
incorporate vir_ms_park{3,4} changes

@@
 \usepackage{amssymb}
+%\usepackage[usenames,dvipsnames]{xcolor} %% for non-standard cite color
 \usepackage{graphicx}
+\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = MidnightBlue]{hyperref}
 \usepackage{url}
 \usepackage{enumerate}
-\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = MidnightBlue]{hyperref}
 \usepackage[left=50pt,
@@
 
-How pathogen virulence evolves is a fundamental question in evolutionary biology, of both theoretical and (potentially) practical importance.
-The trade-off theory \citep{Ebert1999} --- which postulates that parasite virulence can be explained as the long-term evolutionary outcome of a saturating relationship between parasite clearance rate and transmission rate --- has been criticized \citep{EbertBull2003,alizon_adaptive_2015}, but has also been successfully applied in several important cases \citep{Dwyer+1990} \todo{others?
-de Roode and Altizer?}.
-One particularly interesting application of these ideas is the work by Fraser et al.
-showing that HIV appears to satisfy the prerequisites of the tradeoff theory: in a study of discordant couples (i.e.
-long-term sexual partnerships with one infected and one uninfected partner), HIV virulence as measured by the rate of progression to AIDS was both heritable and covaried with the set-point viral load (i.e., the characteristic virus load measured in blood during the intermediate stage of infection), which in turn predicted the probability of within-couple transmission \citep{Fraser+2007,fraser_virulence_2014}\todo{read/check fraser+ 2014!; check vs.
-\cite{herbeck_is_2012}}.
-Subsequent studies \citep{shirreff_transmission_2011,herbeck_hiv_2014} used these data to parameterize mechanistic models of HIV virulence evolution, suggesting that HIV invading a novel population would initially evolve increased virulence, peaking after approximately [XXX] years and then declining slightly to a long-stable virulence level.
+The evolution of pathogen virulence is a fundamental process in
+evolutionary biology, of both theoretical and (potentially) practical
+importance. The trade-off theory \citep{Ebert1999} --- which
+postulates that parasite virulence can be explained as the long-term
+evolutionary outcome of a saturating relationship between parasite
+clearance rate and transmission rate --- has been criticized
+\citep{EbertBull2003,alizon_adaptive_2015}, but has also been
+successfully applied in several important cases \citep{Dwyer+1990}
+\todo{others? de Roode and Altizer? maybe \cite{mackinnon1999genetic},
+  \cite{jensen2006empirical}, and \cite{de2008virulence}?}. One
+particularly interesting application of these ideas is the work by
+Fraser et al. showing that HIV appears to satisfy the prerequisites of
+the tradeoff theory: in a study of discordant couples (i.e. long-term
+sexual partnerships with one infected and one uninfected partner), HIV
+virulence as measured by the rate of progression to AIDS was both
+heritable and covaried with the set-point viral load (i.e., the
+characteristic virus load measured in blood during the intermediate
+stage of infection), which in turn predicted the probability of
+within-couple transmission
+\citep{Fraser+2007,fraser_virulence_2014}\todo{read/check fraser+
+  2014!; check vs. \cite{herbeck_is_2012}}. Subsequent studies
+\citep{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
+parameterize mechanistic models of HIV virulence evolution, suggesting
+that HIV invading a novel population would initially evolve increased
+virulence, peaking after approximately [XXX] years and then declining
+slightly to a long-stable virulence level.
+
+\todo{\cite{gras2009viral} and \cite{muller2009increasing} say that mean virulence has been increasing in Netherlands and Italy.
+Empirical support?
+Worth checking out maybe?
+I think these are similar to what \cite{herbeck_is_2012} says about the evolutionary trend but more specific...}
 
@@
 or physiological level, the epidemiological structures used in these
-models are relatively simple. \todo{don't forget to check other recent papers: esp \cite{payne_impact_2014}} As we discuss in detail below, the
+models are relatively simple. \todo{don't forget to check other recent papers: esp \cite{payne_impact_2014}}
+
+\todo{\cite{lythgoe2013hiv} --- multi-strain model + within host dynamics but assumes random mixing.
+This one considers multi-strains existing within a host.
+Another complication that is important but this paper also ignores the partnership dynamics...}
+
+\todo{\cite{van2014immuno} --- seems even more complicated but doesn't seem like it mentions anything about partnership dynamics... HIV phases are based on Fraser and Hollingsworth et al}
+
+\todo{\cite{herbeck2016evolution} --- seems better than the previous herbeck et al paper but still weird and is not published (bioRxiv)..} As we discuss in detail below, the
 existing models of HIV eco-evolutionary dynamics either use implicit
@@
 \section{Methods}
+
+\begin{itemize}
+\item Since we assume that SPVL is determined by genoyype, we should probably mention papers like this:\cite{alizon2010phylogenetic}
+\item Do we need to mention things like heritability of SPVL or do we not have to mention it since it's already included in Shirreff et al's paper and our models are based on their model?
+It's an important assumption and maybe we need to cite source..?
+\end{itemize}
 
@@
 \end{equation}
-Then, we construct a mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have $\log_{10}$ SPVL of $j$ given that the infector has $\log_{10}$ of $i$.
+Then, we construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have $\log_{10}$ SPVL of $\alpha_j$ given that the infector has $\log_{10}$ SPVL of $\alpha_i$.
 Finally, the probabilities are normalized so that each row sums to 1:
@@
 \begin{equation}
-M_{ij} = \frac{\Phi(\alpha_j + d/2) - \Phi(\alpha_j - d/2)}{\Phi(\alpha_{max} + d/2) - \Phi(\alpha_{min} - d/2)},
+M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\alpha_{max} + d/2;i) - \Phi(\alpha_{min} - d/2;i)},
 \end{equation}
 
-where $\Phi(j)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/n$.
+where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$.
 Transmission rate and disease induced mortality rates are discretized into a vector as well:
@@
 However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008} \todo{is this the right ref?
-where does Shirreff's model originate?}.
+where does Shirreff's model originate?} \todo{Evolutionary dynamic (SPVL distribution, disease progression rate, mutation, and etc) comes from \cite{Fraser+2007} and serial monogamy approximation comes from \cite{hollingsworth_hiv1_2008}}.
 Finally, model 6 is a simple random-mixing model.
+
+The base model (i.e.
+model 1) for the first four models is an extension of \champr's model.
+Individuals in single compartment acquire a partner at a rate, $\rho$, and partnerships dissolve at a rate, $c$.
+Infected individuals in a discordant partnership infect susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
+Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
+Extra-couple and uncoupled transmission are modeled in a same way as \champr's model.
+All the details have been adapted to a multi-strain scenario.
+Model 2, 3, and 4 are derived from the base model by removing epidemiological details (partnership formation and uncoupled/extra-couple contact).
+Model details are explained in the appendix.
 
@@
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-We do not allow for uncertainties in parameters that are directly related to evolutionary process \todo{i.e., ...
+We do not allow for uncertainties in parameters that are directly related to evolutionary process ($\beta_{max}$, $\beta_{50}$, $\beta_k$, $D_{max}$, $D_{50}$, $D_k$, $\sigma_M$) \todo{i.e., ...
 ?}
 
-Latin hypercube sampling is done as in \champr.
-For each parameter, $z$, its range is divided into $N$ \todo{give the number we actually used here} equal intervals on a log scale, where $N$ is the number of total runs:
+Latin hypercube sampling is done as in \champr. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale, where $N$ is the number of total runs:
 
@@
 Since transmission rate is adjusted by the scale of $\gamma$, uncoupled and extra-couple transmission rates are adjusted as well.
-For model 3, 4, and 5, all of which assume instantaneous partnership, only the transmission rate and partnership dissolution rate are adjusted.
+For models 3, 4, and 5, all of which assume instantaneous partnership (and hence do not track single individuals), only the transmission rate and partnership dissolution rate (in this case equivalent to the partnership change rate) are adjusted.
 
@@
 of mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$ (refer to appendix).
-An  initial run of After each run, initial exponential growth rate is calculated.
+After each run, initial exponential growth rate is calculated.
 Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.
-
-\todo{mention that baseline parameters are geometric means of ranges}
-
-\todo{Do we need to talk about the baseline parameters?
-We never use it...
-We only use it for the first set of graphs but we need to mention that we are using Shirreff's partnership rate just for comparison...So we actually never use baseline parameters...}
 
@@
 @
+
+ Throughout Fig 1 we're using Shirreff's parameters (Beta1 = 2.76, Duration1 = 0.25, Beta3 = 0.76, Duration3 = 0.75, cmeanbase = 1.25, alphaDist = c(min = 2, max = 7,  delta = 0.1))
 

### 2016-05-05 ms/ms.Rnw
further tweaks -- still not really doing results/discussion!

@@
 
+\newcommand{\etal}{et al.}
 \newcommand{\shir}{Shirreff et al.}%{\citeauthor{shirreff_transmission_2011}}
@@
 \citep{EbertBull2003,alizon_adaptive_2015}, but has also been
-successfully applied in several important cases \citep{Dwyer+1990}
-\todo{others? de Roode and Altizer? maybe \cite{mackinnon1999genetic},
-  \cite{jensen2006empirical}, and \cite{de2008virulence}?}. One
+successfully applied in a variety of host-pathogen systems \citep{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}. One
 particularly interesting application of these ideas is the work by
-Fraser et al. showing that HIV appears to satisfy the prerequisites of
+Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
 the tradeoff theory: in a study of discordant couples (i.e. long-term
@@
 
-\begin{itemize}
-\item Since we assume that SPVL is determined by genoyype, we should probably mention papers like this:\cite{alizon2010phylogenetic}
-\item Do we need to mention things like heritability of SPVL or do we not have to mention it since it's already included in Shirreff et al's paper and our models are based on their model?
-It's an important assumption and maybe we need to cite source..?
-\end{itemize}
-
-\subsection{Model formulation}
+\todo{more papers on SPVL heritability (\emph{maybe} include): \cite{alizon_phylogenetic_2010,leventhal_potential_2016}}
 
@@
 \subsubsection{Infection dynamics}
-
-\todo{was going to call this ``within-host'' but it includes transmission}
 
@@
 It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
-However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008} \todo{is this the right ref?
-where does Shirreff's model originate?} \todo{Evolutionary dynamic (SPVL distribution, disease progression rate, mutation, and etc) comes from \cite{Fraser+2007} and serial monogamy approximation comes from \cite{hollingsworth_hiv1_2008}}.
+However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \citep{hollingsworth_hiv1_2008}.
 Finally, model 6 is a simple random-mixing model.
@@
 \begin{equation}
-z_i = \text{exp}(\log(z_{min}) + [\log(z_{max}) - \log(z_{min})] \frac{i-1}{N-1}) \qquad i = 1,...,N.
+z_i = \text{exp}(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}) \qquad i = 1,...,N.
 \end{equation}

### 2016-05-06 ms/ms.Rnw
still working

@@
 \usepackage{enumerate}
+%\usepackage[normalem]{ulem}
 \usepackage[left=50pt,
@@
 \begin{split}
-D_A(\alpha) &= \frac{D_{\text{max}} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
-\beta_A(\alpha) &= \frac{\beta_{\text{max}} V_\alpha ^ {\beta_k}}%
+D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
+\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
 {V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
@@
 \begin{equation}
-M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\alpha_{max} + d/2;i) - \Phi(\alpha_{min} - d/2;i)},
+M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{alpha}{min} - d/2;i)},
 \end{equation}
@@
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-We do not allow for uncertainties in parameters that are directly related to evolutionary process ($\beta_{max}$, $\beta_{50}$, $\beta_k$, $D_{max}$, $D_{50}$, $D_k$, $\sigma_M$) \todo{i.e., ...
-?}
+We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using \shir's point estimates throughout.
 
-Latin hypercube sampling is done as in \champr. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale, where $N$ is the number of total runs:
+Latin hypercube sampling is done as in \champr. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
 
 \begin{equation}
-z_i = \text{exp}(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}) \qquad i = 1,...,N.
+z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1,...,N.
 \end{equation}
-For simplicity, we assume that all parameters follow logarithmic distribution.
+For simplicity, we assume that all parameters are uniformly distributed on the log scale.
 Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$).
@@
 Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
-Parameters ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from \champr, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from \cite{hollingsworth_hiv1_2008}.
-Rest of the parameters are taken from \shir.
+Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from \champr, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from \cite{hollingsworth_hiv1_2008}.
+The remaining parameters are taken from \shir \todo{check: redundant with description a couple of paras above?}
 
-As we introduce a new parameter, $c_u/c_w$, to model uncoupled and extra-couple transmission dynamics more easily within multi-strain models (refer to appendix), it is important to pick a reasonable range for it.
-\champr assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
-Given \champr's parameter range, it is trivial that the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5.
+The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix); we need to pick a reasonable range for it.
+\champr\ assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
+Given \champr's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5.
 Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$.
-Although this adds more uncertainty to the parameter $c_u$ --- \champr's range gives 5 fold difference whereas ours give 25 fold difference --- as there is not much known about the uncoupled transmission rate, it is reasonable to use a wider range.
-
-\todo{refer to Table 1 and describe where the new parameters come from; e.g. decisions we made about new scaling parameters}
+Although this adds more uncertainty to the parameter $c_u$ --- \champr's range implies a 5-fold difference whereas ours gives a 25-fold difference --- as there is not much known about the uncoupled transmission rate, we consider the wider range appropriate.
 
@@
 While $\rzero$ is often considered the most fundamental property of an epidemic, and might thus seem to be a natural goal, here we focus on matching the initial growth rate $r$ for several reasons.
-First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than $\rzero$ (CITE?).
-Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves (CITE Ma et al.), while $\rzero$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease (CITE Wallinga, Teunis, Lipsitch, etc.) Thus, we scale a parameter so that every run has an equal exponential rate of growth rate.
+First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than $\rzero$ \todo{CITE?}.
+Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves \todo{CITE Ma et al.}, while $\rzero$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease \todo{CITE Wallinga, Teunis, Lipsitch, etc.} Thus, we scale a parameter so that every run has the same initial exponential growth rate in the incidence.
 
@@
 
-We run each model for 1000 times with different parameter sets calculated from Latin hypercube sampling [CITE Blower et al], with fixed starting conditions
-of mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$ (refer to appendix).
+We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling \todo{CITE Blower et al - or at first mention of LHS}, with fixed starting conditions
+of mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$.
 After each run, initial exponential growth rate is calculated.
@@
 First of all, knowing the possible ranges for the peak virulence allows us to estimate the worst-case scenario for the HIV and other sexually transmitted disease epidemics.
-When an epidemic begins, it is often the case that the pathogen has already evolved towards high virulence by the time it is observed [CITE].
+When an epidemic begins, it is often the case that the pathogen has already evolved towards high virulence by the time it is observed \todo{CITE}.
 Understanding how virulent a pathogen can evolve before an epidemic begins can be helpful for controlling the disease.
@@
 
-Our simplifications of \shir's model reproduce its qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in \shir's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
-However, our model produces lower peak virulence ($\approx ?$ vs.
+Our simplifications of \shir's model reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
+As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in \shir's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
+However, our model produces lower peak virulence \todo{fill in these numbers ...} ($\approx ?$ vs.
 $\approx ?$) and equilibrium virulence $\approx ?$ vs.
-$\approx?$) than Shirreff's, even for matching initial incidence (i.e., $r=0.042$).
+$\approx?$) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042$).
 
-Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases and allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b).
+Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases.
+Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b).
 Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Figure~\ref{fig:panel3}c).
 
-\todo{note that in Fig 1 we're using Shirreff's parameters}
-\todo{add Shir. params to table 1: Beta1=2.76, Dur1 = 0.25, Beta3 = 0.76, Dur3 = 0.75, c = 1.25, alpha=\{2,7,0.1\}}
+\todo{add Shirreff's parameters to table 1: (\sout{Beta1 = 2.76, Duration1 = 0.25}, Beta3 = 0.76, Duration3 = 0.75, \sout{cmeanbase = 1.25, alphaDist = c(min = 2, max = 7,  delta = 0.1)}).}
 
@@
 (b) Effects of varying initial infectious density $I(0)$.
-(c) Effects of varying initial mean virulence $\\alpha(0)$",dev="tikz",fig.height=3,fig.width=8,cache=TRUE>>=
+(c) Effects of varying initial mean virulence $\\alpha(0)$.
+Parameters",dev="tikz",fig.height=3,fig.width=8,cache=TRUE>>=
 for (i in c("","2","3")) 
@@
 mLw$variable <- fixfac2(mLw$variable)
-ggplot(mLw,aes(value,model,fill=model))+
+gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
     geom_violinh(width=1)+
@@
                                value=rval),pch=22,size=3,show.legend=FALSE)
+print(gg_univ)
 @
@@
 <<pairplot,echo=FALSE,fig.cap="pairs plot",fig.width=7,fig.height=7>>=
-## names(sL)[na.omit(match(names(sL),orig_sum_labs))] <- paste0("`",new_sum_labs,"`")
-ggp1 <- ggpairs(sL,
+sL2 <- sL
+names(sL2)[na.omit(match(orig_sum_labs,names(sL)))] <- gsub(" ","_",new_sum_labs)
+    ## paste0("`",new_sum_labs,"`")
+ggp1 <- ggpairs(sL2,
         mapping = ggplot2::aes(color = model),
@@
                  scale_fill_brewer(palette = 'Dark2'))  ## DRY
-for (row in seq_len(ggp1$nrow))
-    for (col in seq_len(ggp1$ncol))
-        ggp1[row, col] <- ggp1[row, col] + cbscales[[(row==col)+1]]
+ggp2 <- ggp1
+for (row in seq_len(ggp2$nrow))
+    for (col in seq_len(ggp2$ncol))
+        ggp2[row, col] <- ggp2[row, col] + cbscales[[(row==col)+1]]
 ## hack to remove top row/right column
-ggp1$nrow <- ggp1$ncol <- 3
-ggp1$plots <- ggp1$plots[c(5:7,9:11,13:15)]
-print(ggp1)
+ggp3 <- ggp2
+ggp3$nrow <- ggp3$ncol <- 3
+ggp3$plots <- ggp3$plots[c(5:7,9:11,13:15)]
+ggp3$xAxisLabels <- ggp3$xAxisLabels[1:3]
+ggp3$yAxisLabels <- ggp3$yAxisLabels[2:4]
+print(ggp3)
 @
 
-pairs plot: instswitch +epc is the most different from others
-uncoupled mixing rate 1/5 to 5: rate of uncoupled contact relative
-to within-pair contact.  If it were much lower instswitch+epc would
-look more like instswitch *without* epc
-(uncoupled mixing acts like random mixing)
+Our chosen summary statistics (peak time, peak virulence, equilibrium
+virulence, and relative peak virulence) all vary considerably across models.
+We first consider the models of intermediate realism (implicit,
+instantaneous-switching with and without extra-pair contact, and
+pair formation without extra-pair contact). Some parameter
+sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10} SPVL$)
+these same sets lead to correspondingly low
+peak virulence ($< 3.5 \log_{10} SPVL$) and early peak times (before 200 years), 
+but high relative peaks ($>1.3$)
+(Figure~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
+At the opposite extreme, parameter sets that produce high equilibrium virulence 
+also produce late peaks ($> 200 \text{years}$), 
+high peak virulence, and low relative peaks ($\approx 1.05$).
+The pair-formation without extra-pair contact and implicit models
+occasionally have parameter sets that select for such low virulence across
+the board that they never exceed their initial virulence, leading to a tail
+of peak times near 0.
 
@@
 @
-
- Throughout Fig 1 we're using Shirreff's parameters (Beta1 = 2.76, Duration1 = 0.25, Beta3 = 0.76, Duration3 = 0.75, cmeanbase = 1.25, alphaDist = c(min = 2, max = 7,  delta = 0.1))
 
@@
 SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
-II_{ij}' &= (\frac{1}{2})^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
+II_{ij}' &= \left(\frac{1}{2}\right)^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
 \end{aligned}
 \end{equation}
-We introduce Kronecker delta above to differentiate the partnership formation rate for $II_{ij}$ 
+We introduce the Kronecker delta above to differentiate the partnership formation rate for $II_{ij}$ 
 when $i = j$ from that of $I \neq j$.
-When $i = j$ --- like the partnership formation rate of $SS$ --- partnership formation rate becomes $II_{ii}' = \frac{\rho I_i \cdot I_i}{2(S + \sum_k I_k)}$ due to binomial distribution.
-On the other hand, when $i \neq j$, partnership formation rate becomes $II_{ij}' = \frac{\rho I_i I_j}{S + \sum_k I_k}$.
+When $i = j$ --- e.g.
+in the $II_{ij}$ term --- the partnership formation rate becomes $II_{ii}' = \frac{\rho I_i \cdot I_i}{2(S + \sum_k I_k)}$ due to binomial distribution \todo{??}.
+On the other hand, when $i \neq j$, the partnership rate is $II_{ij}' = \frac{\rho I_i I_j}{S + \sum_k I_k}$.
 
@@
 Unlike single strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains which the individuals in concordant partnership are infected with (i.e.
-both partners in $II_{ii}$ enter $I_i$ whereas only one partner in $II_{ij}$ enter $I_i$).
+both partners in $II_{ii}$ enter $I_i$ whereas only one partner in $II_{ij}$ enters $I_i$).
 

### 2016-05-06 ms/ms_parms.tex
more ms updates
still working

--- 2016-03-12 ms/ms_parms.tex
@@
 \begin{table}[ht]
-\caption{Parameter ranges/values}
+\caption{Parameter ranges/values.
+Note that $c$ and $\rho$ values are doubled from \champr\  because we keep track of individuals, while they keep track of couples.
+Starred (*) parameters (used in Figure~\ref{fig:panel3}) are from \shir.}
 \centering
@@
 $\rho$ & Partnership formation rate & 1/10-2/5 & \cite{champredon_hiv_2013} \\
-$c$ & Partnership dissolution rate & 1/15-1/5 & \cite{champredon_hiv_2013} \\
-$c_u/c_w$ & Relative contact rate uncoupled & 1/5-5 & Assumption \\
+$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) & \cite{champredon_hiv_2013} \\
+$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
 $c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
-$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 & \cite{hollingsworth_hiv1_2008} \\
+$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) & \cite{hollingsworth_hiv1_2008} \\
 $\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 & \cite{hollingsworth_hiv1_2008} \\
-$D_P$ & Duration of primary infection & 1.23/12-6/12 & \cite{hollingsworth_hiv1_2008} \\
+$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) & \cite{hollingsworth_hiv1_2008,shirreff_transmission_2011} \\
 $D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 & \cite{hollingsworth_hiv1_2008} \\
-$\beta_{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
+$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
 $\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 & \cite{shirreff_transmission_2011} \\
 $\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
-$D_{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
+$\tsub{D}{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
 $D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 & \cite{shirreff_transmission_2011} \\
@@
 $\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
-$\alpha_{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
-$\alpha_{max}$ & Maximum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
-$n$ & Number of strains & 21 & Assumption\\
+$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
+$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
+$n$ & Number of strains & 21 (51*) & Assumption\\
 \hline

### 2016-05-06 ms/ms.Rnw
still working 2

@@
 
-<<unidist,fig.cap="Univariate distributions of summary statistics. Note: point for eqvir/random is a placeholder, will have to be explained",fig.width=8,fig.height=4.8>>=
+<<unidist,fig.cap="Univariate distributions of summary statistics.
+The distribution of equilibrium virulence for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.",fig.width=8,fig.height=4.8>>=
 sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
@@
 Our chosen summary statistics (peak time, peak virulence, equilibrium
-virulence, and relative peak virulence) all vary considerably across models.
+virulence, and relative peak virulence) all vary considerably across models
+\ref{fig:unidist}.
 We first consider the models of intermediate realism (implicit,
@@
 
-instswitch is intermediate beween (pairform + implicit)
+The most striking aspect of the univariate comparisons in Figure~\ref{fig:unidist}
+is the similarity of the simplest (random mixing) and the most complex
+(pair formation with extra-pair contact) models,
 

### 2016-05-06 ms/ms.Rnw
more bib updates

@@
 
-\todo{I think we need a section here to describe partnership dynamics/contact structure, i.e. define the $\rho$, $c_u$, $c_e$, $c_w$ parameters - will be similar to \champr}
 \input{ms_parms.tex}
@@
 Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from \champr, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from \cite{hollingsworth_hiv1_2008}.
-The remaining parameters are taken from \shir \todo{check: redundant with description a couple of paras above?}
+The remaining parameters are taken from \shir.
 
@@
 Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Figure~\ref{fig:panel3}c).
-
-\todo{add Shirreff's parameters to table 1: (\sout{Beta1 = 2.76, Duration1 = 0.25}, Beta3 = 0.76, Duration3 = 0.75, \sout{cmeanbase = 1.25, alphaDist = c(min = 2, max = 7,  delta = 0.1)}).}
 
@@
 pair formation without extra-pair contact). Some parameter
-sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10} SPVL$)
+sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL)
 these same sets lead to correspondingly low
-peak virulence ($< 3.5 \log_{10} SPVL$) and early peak times (before 200 years), 
+peak virulence ($< 3.5 \log_{10}$ SPVL) and early peak times (before 200 years), 
 but high relative peaks ($>1.3$)
@@
 At the opposite extreme, parameter sets that produce high equilibrium virulence 
-also produce late peaks ($> 200 \text{years}$), 
+also produce late peaks ($> 200 \text{ years}$), 
 high peak virulence, and low relative peaks ($\approx 1.05$).
@@
 (pair formation with extra-pair contact) models,
+
+\emph{why implicit = pair formation? when we increased the unadjusted partnership
+mixing rates, the two models started to look different; so they may just be
+at some lower bound ...}
 
@@
 <<plot_sens,dev="png",echo=FALSE,fig.width=10,fig.height=6>>=
+subset(mL2,
+       !(model=="implicit" &
+           LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")))
 ggplot(mL2,aes(LHSval,sumval,colour=model))+
@@
 \item fig 2. 6-panel figure? log scale? boundary lines instead of ribbons for edges of envelopes? move labels away from axes?
-\item fig 3. change order (Random after pairform)
-\item fig 4. add legend, fix axis labels; maybe transparent
+\item fig 3. change order (flip instswitch and instswitch + epd)
+\item fig 4. add legend, maybe transparent
 \item fig 5.

### 2016-05-06 ms/ms_parms.tex
more bib updates

@@
 $\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) & \cite{hollingsworth_hiv1_2008} \\
-$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 & \cite{hollingsworth_hiv1_2008} \\
+$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) & \cite{hollingsworth_hiv1_2008} \\
 $D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) & \cite{hollingsworth_hiv1_2008,shirreff_transmission_2011} \\
-$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 & \cite{hollingsworth_hiv1_2008} \\
+$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) & \cite{hollingsworth_hiv1_2008} \\
 $\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\

### 2016-05-10 ms/ms.Rnw
considerable work on ms

@@
 \newcommand{\champr}{Champredon et al.}%{\citeauthor{champredon_hiv_2013}}
-\newcommand{\tsub}[2]{#1_{{\small \textrm{#2}}}}
+\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
+\newcommand{\khalf}{\left(\frac{1}{2}\right)^{\delta_{ij}}}  % (1/2)^kronecker
+\newcommand{\Lspvl}{$\log_{10}$ SPVL}
 \newcommand{\rzero}{{\cal R}_0}
@@
 <<pkgs,warning=FALSE,message=FALSE>>=
-library("ggplot2"); theme_set(theme_classic())
+library("ggplot2"); theme_set(theme_bw())
+zero_margin <- theme(panel.margin=grid::unit(0,"lines"))
 library("MASS")
@@
 Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual.
-First, the distribution of $\log_{10}$ SPVL is discretized into a vector:
+First, the distribution of \Lspvl\ is discretized into a vector:
 
@@
 \end{equation}
-Then, we construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have $\log_{10}$ SPVL of $\alpha_j$ given that the infector has $\log_{10}$ SPVL of $\alpha_i$.
+Then, we construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
 Finally, the probabilities are normalized so that each row sums to 1:
@@
 \begin{equation}
-M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{alpha}{min} - d/2;i)},
+M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
 \end{equation}
@@
 non-instantaneous partnership formation and (2) zero vs.
-positive extra-partnership sexual contact and transmission on the evolution of mean $\log_{10}$ SPVL.
+positive extra-partnership sexual contact and transmission on the evolution of mean \Lspvl.
 
 Our first four models consider explicit partnership dynamics and are based on \champr's model.
-Model 1 and 2 assume non-instantaneous partnership formation (i.e.
+Models 1 and 2 assume non-instantaneous partnership formation (i.e.
 individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status.
@@
 First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than $\rzero$ \todo{CITE?}.
-Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves \todo{CITE Ma et al.}, while $\rzero$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease \todo{CITE Wallinga, Teunis, Lipsitch, etc.} Thus, we scale a parameter so that every run has the same initial exponential growth rate in the incidence.
+Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves \citep{ma_estimating_2014}, while $\rzero$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease \citep{wallinga_how_2007}.
+Thus, we scale a parameter so that every run has the same initial exponential growth rate in the incidence.
 
@@
 We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling \todo{CITE Blower et al - or at first mention of LHS}, with fixed starting conditions
-of mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$.
+of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$.
 After each run, initial exponential growth rate is calculated.
@@
 First of all, knowing the possible ranges for the peak virulence allows us to estimate the worst-case scenario for the HIV and other sexually transmitted disease epidemics.
-When an epidemic begins, it is often the case that the pathogen has already evolved towards high virulence by the time it is observed \todo{CITE}.
+Pathogens may already have evolved towards high virulence during the early stages of an epidemic, by the time it is observed by public health authorities.
 Understanding how virulent a pathogen can evolve before an epidemic begins can be helpful for controlling the disease.
@@
 Specifically, in the case of HIV and other sexually transmitted diseases, slow evolutionary time-scale makes observing changes in the mean virulence even more challenging.
-Knowing the ranges of these statistics can help real-time virulence evolution prediction during an epidemic less troublesome.
+Knowing the ranges of these statistics can help real-time virulence evolution prediction during an epidemic less troublesome.\todo{needs a bit of work}
 
@@
 As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in \shir's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
-However, our model produces lower peak virulence \todo{fill in these numbers ...} ($\approx ?$ vs.
-$\approx ?$) and equilibrium virulence $\approx ?$ vs.
-$\approx?$) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042$).
+However, our model produces lower peak virulence ($\approx 4.3$ vs.
+$\approx 4.6$ \Lspvl)
+and equilibrium virulence $\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
 
@@
     annotate(geom="text",label=label,x=-Inf,y=Inf,
-             size=size,vjust=1,hjust=0)
+             size=size,vjust=0.98,hjust=0.02)
 }
@@
 Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (Figure~\ref{fig:virtraj}).
-Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 $\log_{10}$ SPVL), virulence peaks in all models between 200 and 300 years.
+Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 \Lspvl), virulence peaks in all models between 200 and 300 years.
 
-<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",fig.width=6,cache=TRUE,message=FALSE,warning=FALSE,dev="tikz">>=
-gg_virtraj <- ggplot(sum_list[["vir_mat"]],
+<<virtraj,fig.cap="Envelopes of virulence trajectories under all models.
+All models were run until $t=4000~\\textrm{years}$; truncated series are shown here.",fig.width=6,cache=TRUE,message=FALSE,warning=FALSE,dev="tikz">>=
+ss <- subset(sum_list[["vir_mat"]],tvec<750)
+gg_virtraj <- ggplot(ss,
                      aes(tvec,mean,ymin=lwr,ymax=upr))+
@@
     geom_ribbon(aes(fill=model),alpha=0.3)+
-    labs(x="time",y=vlab)+
-    scale_x_continuous(limits=c(0,2500))
+    labs(x="time (years)",y=vlab)
+## scale_x_continuous(limits=c(0,2500))
 print(gg_virtraj)
@@
 
-<<pairplot,echo=FALSE,fig.cap="pairs plot",fig.width=7,fig.height=7>>=
+<<pairplot_funs,echo=FALSE>>=
+trim_gg <- function(gg) {
+    n <- gg$nrow
+    gg$nrow <- gg$ncol <- n-1
+    v <- 1:n^2
+    gg$plots <- gg$plots[v>n & v%%n!=0]
+    gg$xAxisLabels <- gg$xAxisLabels[-n]
+    gg$yAxisLabels <- gg$yAxisLabels[-1]
+    return(gg)
+}
+
+tweak_legends_gg <- function(gg,pos=1.5) {
+    n <- gg$nrow
+    for (i in 1:n) {
+        for (j in 1:n) {
+            inner <- getPlot(gg,i,j)
+            if (i==1 & j==1) {
+                inner <- inner + ggplot2::theme(legend.position=c(pos,0.5)) +
+                    guides(colour = guide_legend(override.aes = list(size=6)))
+            } else {
+                inner <- inner + ggplot2::theme(legend.position="none")
+            }
+            gg <- putPlot(gg,inner,i,j)
+        }
+    }
+    return(gg)
+}
+tweak_colours_gg <- function(gg) {
+    n <- gg$nrow
+    for (i in 1:n) {
+        for (j in 1:n) {
+            inner <- getPlot(gg,i,j)
+            inner <- inner+scale_colour_brewer(palette = 'Dark2')
+            gg <- putPlot(gg,inner,i,j)
+        }
+    }
+    return(gg)
+}
+@ 
+<<pairplot,echo=FALSE,fig.cap="Pairs plot: bivariate relationships among summary statistics for each model structure.
+Dashed line in equilibrium vs.
+peak virulence plot shows 1:1 line.",fig.width=7,fig.height=7>>=
 sL2 <- sL
@@
         columns=3:6,
-        lower = list(continuous = wrap("points", alpha = 0.5,size=0.5)),
+        legends=TRUE,
+        lower = list(continuous = wrap("points",size=0.1)), ## alpha = 0.3,size=0.5)),
         diag = list(continuous = "blankDiag"),
-        upper = list(continuous = "blank"))+
-    theme(panel.margin=grid::unit(0,"lines"))
+        upper = list(continuous = "blank"))
 ## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot
 ## have to change plot one panel at a time
-cbscales <- list(scale_colour_brewer(palette = 'Dark2'),
-                 scale_fill_brewer(palette = 'Dark2'))  ## DRY
-ggp2 <- ggp1
-for (row in seq_len(ggp2$nrow))
-    for (col in seq_len(ggp2$ncol))
-        ggp2[row, col] <- ggp2[row, col] + cbscales[[(row==col)+1]]
-## hack to remove top row/right column
-ggp3 <- ggp2
-ggp3$nrow <- ggp3$ncol <- 3
-ggp3$plots <- ggp3$plots[c(5:7,9:11,13:15)]
-ggp3$xAxisLabels <- ggp3$xAxisLabels[1:3]
-ggp3$yAxisLabels <- ggp3$yAxisLabels[2:4]
-print(ggp3)
+ggp2 <- tweak_colours_gg(tweak_legends_gg(trim_gg(ggp1)))
+## add 1-to-1 line
+inner <- getPlot(ggp2,2,1)
+inner <- inner+geom_abline(intercept=0,slope=1,lty=2)
+ggp2 <- putPlot(ggp2,inner,2,1)
+## tweak scale
+inner <- getPlot(ggp2,3,1)
+inner <- inner+scale_x_continuous(breaks=seq(2.5,4.5,by=0.5))
+ggp2 <- putPlot(ggp2,inner,3,1)
+print(ggp2)
 @
@@
 the board that they never exceed their initial virulence, leading to a tail
-of peak times near 0.
+of peak times near zero.
 
-The most striking aspect of the univariate comparisons in Figure~\ref{fig:unidist}
-is the similarity of the simplest (random mixing) and the most complex
-(pair formation with extra-pair contact) models,
+% \end{document}
 
-\emph{why implicit = pair formation? when we increased the unadjusted partnership
-mixing rates, the two models started to look different; so they may just be
-at some lower bound ...}
+The most striking aspect of the univariate comparisons in
+Figure~\ref{fig:unidist}, (and the bivariate comparisons in
+Figure~\ref{fig:pairplot}) is the match between the output of the
+least (random mixing) and the most complex (pair formation with
+extra-pair contact) models. The random-mixing model has lower
+variability, because it is unaffected by uncertainty in pair formation
+and extra-pair contact parameters, but otherwise the virulence
+dynamics of these two extreme models are remarkably similar.
+This phenomenon is driven by the strength of extra-pair contact in the
+pairform+epc model. When individuals spend time uncoupled between
+partnerships, and when these single individuals can transmit disease
+to coupled individuals, the resulting unstructured mixing overwhelms
+the effect of structured mixing within couples, leading to mixing
+that is effectively close to random.
+
+Plotting the bivariate result distributions (Figure~\ref{fig:pairplot})
+shows that most of the relationships among the summary statistics
+are monotonic, except those involving the relative peak virulence
+(bottom row). Changes in parameters that increase the equilibrium
+virulence initially increase the peak virulence even more, so
+that the relative peak virulence increases as well, but beyond
+an equilibrium virulence of about 2.5 \Lspvl\ the peak virulence
+increases slower than the equilibrium virulence, leading to a
+decrease in the relative peak virulence.
+
+The bivariate relationships also help distinguish the results of 
+different models with similar univariate distributions. While the
+relationship between equilibrium virulence and peak time is
+similar for all model structures (top left panel), the other
+relationships are more separate. In particular, the implicit
+and pair-formation (without extra-pair contact) are very similar
+to each other, but distinct from the other models. We still do
+not have a convincing explanation for this distinction; we
+would have expected the implicit model to be most similar to the
+the instantaneous-switching model without extra-pair contact,
+which most closely matches its derivation. However, we note
+that the implicit model derivation is based on defining
+the force of infection to match a scaled version of $\rzero$,
+and as such would be expected to match the equilibrium behaviour
+but not necessarily the epidemic-phase behaviour of a model
+with explicit partnership dynamics.
 
@@
 
-<<plot_sens,dev="png",echo=FALSE,fig.width=10,fig.height=6>>=
-subset(mL2,
-       !(model=="implicit" &
-           LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")))
-ggplot(mL2,aes(LHSval,sumval,colour=model))+
+<<plot_sens,dev="png",echo=FALSE,out.width="\\textwidth",fig.width=8,fig.height=3.5,dpi=240,fig.cap="Sensitivity plot.
+For each parameter in the Latin hypercube sample and each summary statistic, shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \\emph{unscaled} parameter value, i.e.
+prior to adjusting the parameters to achieve the standard initial epidemic growth rate.">>=
+mL3 <- subset(mL2,
+       !((model=="implicit" &
+          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
+         (model=="random" &
+          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio","c_mean_base")) |
+         (model=="instswitch" &
+          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
+         (model=="instswitch+epc" &
+          LHSvar %in% c("rho_base","c_u_ratio")) |
+         (model=="pairform" &
+          LHSvar %in% c("c_u_ratio","c_e_ratio"))))
+## reorder factors
+mL3$LHSvar <- factor(mL3$LHSvar,
+                     levels=c("Beta1","Beta3",
+                              "Duration1","Duration3",
+                              "c_mean_base","c_e_ratio","rho_base",
+                              "c_u_ratio"))
+## FIXME: could also use labels argument to rename levels, labeller=label_parsed
+## in facet_grid to get pretty math
+ggplot(mL3,aes(LHSval,sumval,colour=model))+
     geom_point(pch=".",alpha=0.5)+
@@
     geom_smooth(se=FALSE)+labs(x="",y="")+
-    scale_x_log10()
+    scale_x_log10(expand=c(0,0.08))+  ## leave a little extra room?
+    zero_margin
 @
 
-Figure to-do: maybe go back to bw theme?
-\begin{itemize}
-\item fig 1. tweak size? remove redundant y axes/labels?
-\item fig 2. 6-panel figure? log scale? boundary lines instead of ribbons for edges of envelopes? move labels away from axes?
-\item fig 3. change order (flip instswitch and instswitch + epd)
-\item fig 4. add legend, maybe transparent
-\item fig 5.
-sensitivity plot (unscaled plot); take out plots that are not related ...
-implicit model doesn't use rho, cu or ce; random doesn't use rho, cmean?, cu, ce; instswitch doesn't have rho or cu or ce; models without epc don't have ce or cu; instswitch + epc doesn't have rho and cu (no uncoupled compartment).
-Maybe squash panels together, higher resolution
-
-\end{itemize}
+Finally, the sensitivity plots (Figure~\ref{fig:plot_sens}) show the effects 
+of each parameter on the summary statistics. In almost every case the
+effects of the parameters are monotonic; note that
+the plot shows the effects of the \emph{unscaled} parameters, i.e.
+before they have been adjusted to achieve a standard initial epidemic
+growth rate.
+The transmission rates ($\beta_P$, $\beta_D$)
+and durations ($D_P$, $D_D$) in the primary and disease stages generally
+decrease the equilibrium virulence, peak virulence, and peak time,
+although the random and pair-formation+epc models have high, relatively
+constant values with respect to these parameters. 
+\todo{I know you explained to me before why beta and Duration go in
+the same direction but I forget ...}
+The base contact rate increases virulence and peak time in almost all
+cases, although the pair-formation+epc model is again relatively insensitive.
+The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
+virulence in the instantaneous-switching+epc model, but not the pair-formation+epc
+model (probably because the uncoupled individuals present in the pair-formation+epc
+model make extra-pair contact by coupled individuals less important).
+Surprisingly, neither of the pair-formation models is particularly sensitive to the 
+rate of partnership formation ($\rho$); the rate of uncoupled contact
+increases virulence and peak time in the pair-formation+epc model, 
+which is the only model to which it applies\todo{is that right??}.
 
 \section{Discussion}
+
+All models must simplify the world.  Many constraints --- such as data
+availability, computation time, or code complexity --- drive the need
+for parsimony, with different constraints applying in different
+contexts. The critical question that modelers must ask is whether the
+simplified model gives adequate answers, or whether the
+simplifications have led to qualitatively or quantitatively incorrect
+conclusions.  This question is especially important for modelers who
+are hoping that their conclusions will guide management decisions.
+
+In the particular example of HIV virulence eco-evolutionary dynamics
+and epidemiological, we reach the slightly ironic conclusion that the
+effort put into building a more realistic model essentially cancels
+out, leaving us in the same position as if we had ignored the problem
+of epidemiological complexity entirely and used a naive random-mixing
+contact model.  \cite{herbeck_hiv_2014} build a
+network model of partnerships, but they set the ``partnership''
+duration to 1 day --- very unrealistic in epidemic terms, but perhaps
+actually more true to real-world HIV epidemiological dynamics than a
+model with realistic partnership durations that neglects extra-pair
+contact \citep{herbeck2016evolution}. Adding still further realism
+might push the results farther away from the random-mixing model. For
+example, our model forms partnerships randomly, and assumes that
+extra-pair contact is randomly mixing across the population;
+one could instead implement extra-pair contact as arising from
+multiple concurrent partnerships (some, such as contact with sex
+workers, of very short duration), or more structured partnership
+formation (by age, ethnicity, or behaviour group). The effects of
+other realistic complications such as explicit modeling of two
+sexes (both in contact structure and differential transmission
+probabilities), temporal and spatial variation in epidemic processes,
+or presence of genetic variation in hosts are harder to predict.
+
+Parameterization is one of the biggest challenges of epidemiological
+modeling. In addition to following \champr\ by doing Latin hypercube
+sampling across a wide range of epidemiological parameters, we 
+calibrated each set of parameters to the same initial epidemic
+growth rate, chosen to match the results of previous models
+\citep{shirreff_transmission_2011}.  The previous models drew their
+parameters from previous cohort studies from the 1980s
+\todo{check dates!} \citep{hollingsworth_hiv1_2008}
+rather than doing any explicit calibration to epidemic curves,
+but they give reasonable order-of-magnitude
+growth rates ($\approx 0.4~\textrm{year}^{-1}$) 
+for the early stages of the HIV epidemic (although considerably
+lower than estimates of $\approx 0.07-0.1~\textrm{year}^{-1}$
+based on population genetic reconstructions \citep{faria_early_2014}).
+\todo{do we know what $r$ Champredon et al. get??}
+However, our reason for calibrating was not to match any
+specific observed epidemic, but rather to make sure that
+we were making meaningful comparisons across a range of
+models with radically different epidemiological structures, and
+hence involving different interpretations of the same quantitative
+parameters.  For example, in models with instantaneous switching the
+partnership dissolution rate $c$ is identical to the partnership
+formation rate; in models with explicit partnership formation,
+the partnership formation rate is also $c$ at equilibrium,
+but might vary over the course of an epidemic.
+It is not obvious whether models with equal parameters but
+different structures should be directly compared; calibration
+solves this problem.
+
+More generally, any model that wants to be
+taken seriously for management and forecasting purposes should
+be calibrated to \emph{all} available data, using informative
+priors to incorporate both realistic distributions of uncertainty
+for all parameters from independent measurements \citep{elderd_uncertainty_2006}
+and calibration from population-level observations of epidemic
+trajectories. For practical (rather than exploratory) purposes,
+such a procedure would also be an improvement on the common --- although not universal --- %
+practice of assessing uncertainty over uniform ranges, rather than
+distributions that allow more continuous variation in support over
+the range of a parameter.
+
+Researchers have documented that HIV virulence and set-point viral
+load are changing, on time scales comparable to those portrayed here
+(e.g., compare Figure~\ref{fig:virtraj} to \cite{herbeck_is_2012}'s
+estimated rate of change of 1.3 \Lspvl\ per century [95\% CI -0.1 to
+  3]), and have begun to build relatively realistic models that
+attempt to describe how interventions such as mass antiretroviral
+therapy (ART) can be expected to change the trajectory of virulence
+evolution \citep{payne_impact_2014,herbeck2016evolution}.  While these
+efforts are well-intentioned, we caution that epidemiological and
+other structural details that are currently omitted from these models
+could significantly change their conclusions.
 
@@
 \item clean up code and results
-\item incorporate pictures
 \item finish writing!
@@
 \end{itemize}
+\item cite \cite{LipsitchNowak1995}
+\end{itemize}
+
+\begin{itemize}
+\item fig 1. tweak size? remove redundant y axes/labels?
+\item fig 2. 6-panel figure? log scale? boundary lines instead of ribbons for edges of envelopes? move labels away from axes?
+\item fig 3. change order (flip instswitch and instswitch + epd)
+\item fig 4. squash panels together?
+\item fig 5.
+sensitivity plot (unscaled plot); check we havetake out plots that are not related ...
+implicit model doesn't use rho, cu or ce; random doesn't use rho, cmean?, cu, ce; instswitch doesn't have rho or cu or ce; models without epc don't have ce or cu; instswitch + epc doesn't have rho and cu (no uncoupled compartment).
+Maybe squash panels together, higher resolution
 \end{itemize}
@@
 
-Since we use multi-strain models, in which the distribution of $\log_{10}$ SPVL has been discretized into a vector, we use a matrix notation to describe our models.
-Five states described in the method section is replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$.
-Subscript indicates a strain that an individual is infected with.
-For example, $I_i$ is number of infected individuals with $\log_{10}$ SPVL of $\alpha_i$, and $II_{ij}$ is the number of concordant couples, in which two partners have $\log_{10}$ SPVL of $\alpha_i$ and $\alpha_j$.
-$II_{ij}$ is equivalent to $II_{ji}$.
-Lastly, we define Kronecker delta as follows:
+Since we use multi-strain models, in which the distribution of \Lspvl\ has been discretized into a vector, we use a matrix notation to describe our models.
+The five states described in the Methods section are replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$, where the subscripts denote the strain that an individual is infected with.
+For example, $I_i$ is number of infected individuals with \Lspvl\ of $\alpha_i$, and $II_{ij}$ is the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$ (independent of order; $II_{ij}$ is synonymous with $II_{ji}$).
+Below, we use the Kronecker delta (i.e.
+$\delta_{ij}=1$ if $i=j$ and 1 otherwise) in a slightly non-standard fashion as an exponent, e.g.
+$2^{\delta_{ij}}$, to set a value to 2 when $i=j$ and 1 otherwise.
 
-\begin{equation}
-\delta_{ij} = \left.\begin{aligned}
-0 & \quad \text{if} \quad i \neq j,\\
-1 & \quad \text{if} \quad i = j.
-\end{aligned}\right\rbrace
-\end{equation}
-
-In our models, we introduce a non-standard use of Kronecker delta as exponent, such as $2^{\delta_{ij}}$.
-In a single-strain model, both individuals in a concordant partnership, $II$, enter single-infected compartment, $I$, when a partnership dissolves: $I' = 2 c II'$.
-However, this is not the case in a multi-strain model.
-If infected individuals in an $II$ partnership have different strains, they will enter different compartment.
-On the other hand, if both partners in an $II$ partnership are infected with same strains, both will enter the same compartment.
-Therefore, we use Kronecker delta as exponent to distinguish such difference.
-Further details are explain below.
-
-\todo{explain non-standard use of Kronecker delta as exponent!}
-
-\subsection*{Model 1 and 2 - Partnership dynamic}
+\subsection*{Models 1 and 2: partnership dynamics}
 
 Single individuals acquire partners at rate $\rho$: $S' = - \rho S$ and $I_i' = - \rho I_i$.
-We follow Champredon et al's results and assume that single individuals are distributed into couple states through binomial distribution:
+We follow Champredon et al's results and assume that single individuals are distributed into coupled states through binomial distribution\todo{?
+don't understand this wording (ditto below)}:
+\todo{I think we need different notation for partnership formation (i.e. LHS is wrong)}
 
@@
 \begin{aligned}
-SS' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)}\\
-SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
-II_{ij}' &= \left(\frac{1}{2}\right)^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
+(SS)' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)}\\
+(SI_i)' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
+(II_{ij})' &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
 \end{aligned}
 \end{equation}
-We introduce the Kronecker delta above to differentiate the partnership formation rate for $II_{ij}$ 
-when $i = j$ from that of $I \neq j$.
-When $i = j$ --- e.g.
-in the $II_{ij}$ term --- the partnership formation rate becomes $II_{ii}' = \frac{\rho I_i \cdot I_i}{2(S + \sum_k I_k)}$ due to binomial distribution \todo{??}.
-On the other hand, when $i \neq j$, the partnership rate is $II_{ij}' = \frac{\rho I_i I_j}{S + \sum_k I_k}$.
+In a single-strain model, both individuals in a concordant partnership, $II$, enter the single-infected compartment, $I$, when a partnership dissolves: $I' = 2 c II'$.
+However, this is not the case in a multi-strain model.
+If infected individuals in an $II$ partnership have different strains, they will enter different compartments; if they are both infected with the same strain, they both enter the same compartment.
+As described above, we use the Kronecker delta notation to implement this difference between the $i=j$ and $i\neq j$ cases.
 
@@
 SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i\\
-II_{ij}' &= (\frac{1}{2})^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k} - c II_{ij}
+II_{ij}' &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k} - c II_{ij}
 \end{aligned}
@@
 
-\subsection*{Model 1 and 2 - Infection}
+\subsection*{Models 1 and 2: infection}
 
@@
 SI_i' &= - \beta_i SI_i\\
-II_{ij} &=  (\frac{1}{2})^{\delta_{ij}} \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)
+II_{ij} &=  \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)
 \end{aligned}
@@
 \end{equation}
-Effective uncoupled, $c_u$, and extra couple, $c_e$, contact rate can be divided into two terms: uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
-Therefore, transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above.
+Effective uncoupled, $c_u$, and extra couple, $c_e$, contact rate is the product of two terms: uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
+Therefore, the transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above.
 Using this property, we modify the equation above as follows:
@@
 This simplification is useful in a multi-strain model since we cannot multiply a vector with a single value (e.g.
-$c_u S$ in denominator) if we use Champredon et al's equation as it is.
-Extending the above equation to multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain the following equation:
+$c_u S$ in denominator) if we use Champredon et al's equation in its original form.
+Extending the above equation to multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain:
 
@@
 Using the equation above, we can model uncoupled and extra-couple mixing.
-For convenience, uncoupled and extra-couple transmission rates, $c_u$ and $c_e$, will be replaced with $U_i = r_u \beta_i$ and $E_i = r_e \beta_i$ from now on.
+For convenience, uncoupled and extra-couple transmission rates, $c_u$ and $c_e$, will be replaced with $U_i = r_u \beta_i$ and $E_i = r_e \beta_i$ hereafter.
 
 Single susceptible individuals become infected and enter single infected compartment at the total rate of $\sum_k P_k U_k S$.
-Through mutation, newly infected individuals are distributed into each single infected compartments with different strains: $I_i' = \sum_k M_{ki} P_k U_k S$.
-Either partners in $SS$ partnership can be infected and the partnership state can become $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$.
-Formation of $SI_i$ partnership is similar to the process through which single susceptible individuals are distributed into single infected compartments: $SI_i' = \sum_k 2 M_{ki} P_k E_k SS$.
-Lastly, susceptible partner of an $SI$ partnership can be infected due to uncoupled/extra-couple contacts and partnership can move to an $II$ partnership.
-Like previous cases, $SI_i$ partnership moves out of the compartment at a rate of $\sum_k P_k E_k SI_i$.
-Mutation process is similar to that of within-couple transmission.
-The only difference is that the $\log_10$ SPVL of a newly infected partner is not determined by its original partner but from an extra couple partner (i.e.
+Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $I_i' = \sum_k M_{ki} P_k U_k S$.
+Either partner in an $SS$ partnership can be infected, with the partnership state becoming $SI$, at the total rate of $\sum_i 2 P_i E_i SS$.
+The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $SI_i' = \sum_k 2 M_{ki} P_k E_k SS$.
+Lastly, the susceptible partner in an $SI$ partnership can be infected due to uncoupled/extra-couple contacts and partnership can change to an $II$ partnership.
+As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$.
+The mutation process is similar to that of within-couple transmission.
+The only difference is that the $\log_10$ SPVL of a newly infected partner is not determined by its social partner but from an extra-couple partner (i.e.
 the term $P_i$): $II_{ij}' = (1 -\frac{\delta_{ij}}{2})(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$.
-Combining these equations we get the following set of equations that describe all the transmission dynamics
+Combining these equations we get the following set of equations that describe all the transmission dynamics \todo{something missing here?}
 
@@
 
-Disease induced death rate, $\lambda$, is given by taking the reciprocate of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
-Since we assume SIS formulation, where infected individuals that die from infection enter single susceptible compartment, we obtain the following equation for the single infected individuals:
+The disease induced death rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
+Since we assume an SIS formulation, where infected individuals that die from infection are immediately replaced by an individual in the single susceptible compartment, we obtain the following equation for single infected individuals:
 
@@
 
-If an infected individual in a partnership dies, partnership dissolves.
-Thus, an $SI_i$ partnership dissolves at a rate $-\lambda_i$, and the susceptible partner enters single susceptible compartment at rate $\lambda_i SI_i$ (infected partner that dies enter single susceptible compartment at an equal rate as well due to SIS formation):
+If an infected individual in a partnership dies, the partnership dissolves.
+Thus, an $SI_i$ partnership dissolves at a rate $-\lambda_i$, and the susceptible partner enters the single susceptible compartment at rate $\lambda_i SI_i$ (due to the SIS formulation the the infected partner that dies also gives rise, at an equal rate, to an individual entering the single susceptible compartment):
 
@@
 
-Similarly, $II_{ij}$ partnership dissolves at a rate $-(\lambda_i + \lambda_j)$, but two cases, when $i \neq j$ and $i = j$, must be considered separately.
-When an $II_{ij}$ partnership dissolves due to disease induced mortality, where $i \neq j$, death of partner with strain $i$ causes its partner to enter $I_j$ compartment at rate $\lambda_j II_{ij}$, and vice versa.
-When an $II_ii$ partnership dissolves, death of either partner causes the other partner to enter $I_i$ compartment at rate $\lambda_i II_{ii}$, which sums up to $2\lambda_i II_{ii}$.
-Combining these dynamics yield the following equation:
+Similarly, $II_{ij}$ partnerships dissolve at a rate $-(\lambda_i + \lambda_j)$, but two cases, when $i \neq j$ and $i = j$, must be considered separately.
+When an $II_{ij}$ partnership dissolves due to disease-induced mortality, where $i \neq j$, the death of the partner with strain $i$ causes its partner to enter $I_j$ compartment at rate $\lambda_j II_{ij}$, and vice versa.
+When an $II_ii$ partnership dissolves, the death of either partner causes the other partner to enter the $I_i$ compartment at rate $\lambda_i II_{ii}$, which sums up to $2\lambda_i II_{ii}$.
+Combining these dynamics yields:
 
@@
 
-Finally, combining all these equations give us the full model, which is model 1. We can simply take out the uncoupled and extra couple transmission term to obtain equation 2:
+Finally, combining all these equations give us the full model, which is Model 1. We can simply take out the uncoupled and extra couple transmission term to obtain equation 2\todo{?}:
 
@@
 &- \lambda_i SI_i\\
-II_{ij}' =& (\frac{1}{2})^{\delta_{ij}} \cdot \frac{\rho I_i \cdot I_j}{(S + \sum_k I_k)} - c II_{ij} + (\frac{1}{2})^{\delta_{ij}} \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
-&+ (\frac{1}{2})^{\delta_{ij}} \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
+II_{ij}' =& \khalf \cdot \frac{\rho I_i \cdot I_j}{(S + \sum_k I_k)} - c II_{ij} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
+&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
 \end{aligned}
@@
 
-\subsection*{Model 3 and 4 - Partnership dynamic}
+\subsection*{Models 3 and 4: Partnership dynamics}
 
 Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$.
-Partnership dissolution is equal to that of model 1 and 2: $SS' = -cSS$, $SI_i' = - cSI_i$, and $II_{ij}' = - II_{ij}$.
-Once the individuals leave partnership, they enter temporary compartments and are distributed into a partnership through binomial distribution:
+Partnership dissolution rates are equal to those of model 1 and 2: $SS' = -cSS$, $SI_i' = - cSI_i$, and $II_{ij}' = - II_{ij}$.
+Once individuals leave a partnership, they enter temporary compartments and are distributed into a partnership through binomial distribution:
 
@@
 SI_i' &= - c SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
-II_{ij}' &= - c II_{ij} +(\frac{1}{2})^{\delta_{ij}} \frac{Y_i Y_j}{X + \sum_k Y_k}.
+II_{ij}' &= - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k}.
 \end{aligned}
@@
 SI_i &= - \lambda_i SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
-II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + (\frac{1}{2})^{\delta_{ij}} \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}
+II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + \khalf \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}
 \end{aligned}
@@
 &- \sum_k P_k E_k SI_i - \lambda_i SI_i  \\
-II_{ij}& - c II_{ij} +(\frac{1}{2})^{\delta_{ij}} \frac{Y_i Y_j}{X + \sum_k Y_k} + (\frac{1}{2})^{\delta_{ij}} \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
-&+ (\frac{1}{2})^{\delta_{ij}} \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
+II_{ij}& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
+&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
 \end{aligned}
@@
 
-Model 5 is an implicit instantaneous partnership formation model, which uses adjusted transmission rate, $\beta'$, that is derived from Hollingsworth et al's approximated basic reproduction number:
+Following \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta'$, that is derived from \cite{hollingsworth_hiv1_2008}'s approximate basic reproduction number:
 
@@
 For model 1 and 2, we have disease equilibrium state of $S^* = \frac{c}{c + \rho}$ and $SS^* = \frac{1-S^*}{2}$.
-We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with $\log_{10}$ SPVL of $i$.
+We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with \Lspvl\ of $i$.
 $Y_i$ is taken from normal distribution with mean 3 and is normalized so that $\sum_i D_i = 1$.
@@
 I_i(0) &=  \epsilon S^* D_i\\
-II_{ij}(0) &=  (\frac{1}{2})^{\delta_{ij}} 2\epsilon^2 SS^* D_i D_j.
+II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
 \end{aligned}
@@
 SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i\\
-II_{ij}(0) &=  (\frac{1}{2})^{\delta_{ij}} 2\epsilon^2 SS^* D_i D_j.
+II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
 \end{aligned}

### 2016-05-12 ms/ms.Rnw
vir_ms_park5

@@
 
+\todo{shireff says that the interval between two strains is 0.05 but uses 0.1 in the code...
+Do we want to do a simulation with 101 strains or is it OK to keep it as 51?
+It wouldn't make any but I guess it's better to match what's written in the paper?
+Shirreff simulation that was used in figure 1 was done by using 51 strains.}
+
 \begin{equation}
@@
 In contrast, models 5 and 6 are not explicitly structured.
-Model 5 is an implicit serial monogamy model based on the epidemiological model used by \shir.
+Model 5 is an implicit serial monogamy model based on the epidemiological model used by \shir \todo{cite?}.
 It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
@@
 pair formation without extra-pair contact). Some parameter
-sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL)
+sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL) \todo{period? semicolon?}
 these same sets lead to correspondingly low
@@
 This phenomenon is driven by the strength of extra-pair contact in the
-pairform+epc model. When individuals spend time uncoupled between
+pairform+epc model \todo{might be better to call it something else?}. When individuals spend time uncoupled between
 partnerships, and when these single individuals can transmit disease
@@
          (model=="random" &
-          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio","c_mean_base")) |
+          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
          (model=="instswitch" &
@@
                               "c_mean_base","c_e_ratio","rho_base",
-                              "c_u_ratio"))
+                              "c_u_ratio"),
+										 labels = c("beta[P]", "beta[D]", "D[P]", "D[D]", "c", "c[e]/c[w]", "rho", "c[u]/c[w]"))
 ## FIXME: could also use labels argument to rename levels, labeller=label_parsed
@@
     geom_point(pch=".",alpha=0.5)+
-    facet_grid(sumvar~LHSvar,scales="free")+
+    facet_grid(sumvar~LHSvar,scales="free", labeller = label_parsed)+
     geom_smooth(se=FALSE)+labs(x="",y="")+
@@
 increases virulence and peak time in the pair-formation+epc model, 
-which is the only model to which it applies\todo{is that right??}.
+which is the only model to which it applies\todo{is that right?? Yes}.
 
@@
 Parameterization is one of the biggest challenges of epidemiological
-modeling. In addition to following \champr\ by doing Latin hypercube
+modeling. In addition to following \champr\ \todo{cite?} by doing Latin hypercube
 sampling across a wide range of epidemiological parameters, we 
@@
 parameters from previous cohort studies from the 1980s
-\todo{check dates!} \citep{hollingsworth_hiv1_2008}
+\todo{check dates!} \citep{hollingsworth_hiv1_2008} \todo{That is the correct date.
+We need to cite \citep{Fraser+2007} as well.
+This is where the evolutionary parameters come from.
+\citep{wawer2005rates} might have to be cited as well since Hollingsworth et al base their study on this paper, and Shirreff get their partnership change rate from this paper.}
 rather than doing any explicit calibration to epidemic curves,
 but they give reasonable order-of-magnitude
-growth rates ($\approx 0.4~\textrm{year}^{-1}$) 
+growth rates ($\approx 0.4~\textrm{year}^{-1}$) \todo{0.04 not 0.4?}
 for the early stages of the HIV epidemic (although considerably
@@
 based on population genetic reconstructions \citep{faria_early_2014}).
-\todo{do we know what $r$ Champredon et al. get??}
+\todo{do we know what $r$ Champredon et al. get?? They give a really wide range of values so I'm not sure...}
 However, our reason for calibrating was not to match any
@@
 the partnership formation rate is also $c$ at equilibrium,
-but might vary over the course of an epidemic.
+but might vary over the course of an epidemic. \todo{Not really sure what this example means...}
 It is not obvious whether models with equal parameters but

### 2016-05-12 ms/ms.Rnw
working on figures

@@
                run=c("$10^{-3}$", "$10^{-4}$", "$10^{-5}$"))
-g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL")
+g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL") +
+	labs(y = "")
 ###
@@
                             run=c("$2.5$", "$3$", "$3.5$"))
-g3 <- g1 %+% melt(vir_matS3,id.var="Time",value.name="Mean.VL")
+g3 <- g1 %+% melt(vir_matS3,id.var="Time",value.name="Mean.VL") +
+	labs(y = "")
 limits <- c(2.5,4.7)
@@
 ss <- subset(sum_list[["vir_mat"]],tvec<750)
+
+ss$model <- factor(ss$model, c("random","pairform+epc", "pairform","instswitch+epc","instswitch","implicit"))
+
 gg_virtraj <- ggplot(ss,
@@
 	geom_line(aes(colour=model,linetype=model),lwd=1.5)+
-    geom_ribbon(aes(fill=model),alpha=0.3)+
-    labs(x="time (years)",y=vlab)
+	geom_ribbon(aes(fill = model), alpha=0.3)+
+	labs(x="time (years)",y=vlab) +
+	theme(axis.title.y = element_text(margin = margin(0,20,0,0)),
+		axis.title.x = element_text(margin = margin(10,0,0,0))) +
+	facet_wrap(~model, nrow = 2)
+
 ## scale_x_continuous(limits=c(0,2500))
@@
 sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
+sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
 mL <- melt(sL,id.vars=c("model","run"))
@@
 										 labels = c("beta[P]", "beta[D]", "D[P]", "D[D]", "c", "c[e]/c[w]", "rho", "c[u]/c[w]"))
+
+mL3$model <- factor(mL3$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
+
 ## FIXME: could also use labels argument to rename levels, labeller=label_parsed

### 2016-05-12 ms/ms.Rnw
small updates

@@
 or physiological level, the epidemiological structures used in these
-models are relatively simple. \todo{don't forget to check other recent papers: esp \cite{payne_impact_2014}}
+models are relatively simple.
 
@@
 It wouldn't make any but I guess it's better to match what's written in the paper?
-Shirreff simulation that was used in figure 1 was done by using 51 strains.}
+Shirreff simulation that was used in figure 1 was done by using 51 strains.
+\textbf{51 is OK, I think; maybe we should comment}}
 
@@
 While $\rzero$ is often considered the most fundamental property of an epidemic, and might thus seem to be a natural goal, here we focus on matching the initial growth rate $r$ for several reasons.
-First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than $\rzero$ \todo{CITE?}.
+First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than $\rzero$.
 Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves \citep{ma_estimating_2014}, while $\rzero$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease \citep{wallinga_how_2007}.
@@
 pair formation without extra-pair contact). Some parameter
-sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL) \todo{period? semicolon?}
+sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL);
 these same sets lead to correspondingly low
@@
 This phenomenon is driven by the strength of extra-pair contact in the
-pairform+epc model \todo{might be better to call it something else?}. When individuals spend time uncoupled between
+pairform+epc model \todo{might be better to call it something else? \textbf{agreed. ideas?}}. When individuals spend time uncoupled between
 partnerships, and when these single individuals can transmit disease
@@
 increases virulence and peak time in the pair-formation+epc model, 
-which is the only model to which it applies\todo{is that right?? Yes}.
+which is the only model to which it applies.
 
@@
 Parameterization is one of the biggest challenges of epidemiological
-modeling. In addition to following \champr\ \todo{cite?} by doing Latin hypercube
+modeling. In addition to following \cite{champredon_hiv_2013} 
+by doing Latin hypercube
 sampling across a wide range of epidemiological parameters, we 
@@
 but they give reasonable order-of-magnitude
-growth rates ($\approx 0.4~\textrm{year}^{-1}$) \todo{0.04 not 0.4?}
+growth rates ($\approx 0.04~\textrm{year}^{-1}$)
 for the early stages of the HIV epidemic (although considerably

### 2016-05-13 ms/ms.Rnw
small changes to the figures

@@
     geom_line()+labs(x=tlab,y=vlab)+
-    theme(legend.position=c(0.8,0.2))
+    theme(legend.position=c(0.8,0.2),
+    axis.title.y = element_text(margin = margin(0,15,0,0)))
 ###
@@
 	labs(x="time (years)",y=vlab) +
-	theme(axis.title.y = element_text(margin = margin(0,20,0,0)),
-		axis.title.x = element_text(margin = margin(10,0,0,0))) +
+	theme(axis.title.y = element_text(margin = margin(0,15,0,0)),
+		axis.title.x = element_text(margin = margin(15,0,0,0))) +
 	facet_wrap(~model, nrow = 2)

### 2016-05-13 ms/ms.Rnw
zero margins on plos 2&3; TODO tweaks

--- 2016-05-12 ms/ms.Rnw
@@
 zero_margin <- theme(panel.margin=grid::unit(0,"lines"))
+zero_x_margin <-
+    theme(panel.margin.x=grid::unit(0, "lines"))
 library("MASS")
@@
 First, the distribution of \Lspvl\ is discretized into a vector:
-
-\todo{shireff says that the interval between two strains is 0.05 but uses 0.1 in the code...
-Do we want to do a simulation with 101 strains or is it OK to keep it as 51?
-It wouldn't make any but I guess it's better to match what's written in the paper?
-Shirreff simulation that was used in figure 1 was done by using 51 strains.
-\textbf{51 is OK, I think; maybe we should comment}}
-
 \begin{equation}
@@
 \end{equation}
-Then, we construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
+We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with \cite{shirreff_transmission_2011} (Figure~\ref{fig:panel3}) we use $n=51$ (i.e.
+a bin width of 0.05 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 \Lspvl), which we use for all other simulations.
+
+We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
 Finally, the probabilities are normalized so that each row sums to 1:
-
 \begin{equation}
@@
 \end{equation}
-
 where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$.
 Transmission rate and disease induced mortality rates are discretized into a vector as well:
-
 \begin{equation}
@@
 		axis.title.x = element_text(margin = margin(10,0,0,0))) +
-	facet_wrap(~model, nrow = 2)
+	facet_wrap(~model, nrow = 2)+
+    zero_margin
 
@@
     geom_point(data=data.frame(model="random",variable="equil virulence",
-                               value=rval),pch=22,size=3,show.legend=FALSE)
+                               value=rval),pch=22,size=3,show.legend=FALSE)+
+    zero_x_margin
 print(gg_univ)
@@
 \citep{shirreff_transmission_2011}.  The previous models drew their
-parameters from previous cohort studies from the 1980s
-\todo{check dates!} \citep{hollingsworth_hiv1_2008} \todo{That is the correct date.
-We need to cite \citep{Fraser+2007} as well.
-This is where the evolutionary parameters come from.
-\citep{wawer2005rates} might have to be cited as well since Hollingsworth et al base their study on this paper, and Shirreff get their partnership change rate from this paper.}
+parameters from previous cohort studies from the 1990s
+\citep{wawer2005rates,hollingsworth_hiv1_2008}
 rather than doing any explicit calibration to epidemic curves,

### 2016-05-13 ms/ms.Rnw

@@
 This phenomenon is driven by the strength of extra-pair contact in the
-pairform+epc model \todo{might be better to call it something else? \textbf{agreed. ideas?}}. When individuals spend time uncoupled between
+model with explicit pair formation and extra-pair contact 
+(``pairform+epc'' in Figures~\ref{fig:virtraj}-\ref{fig:plot_sens}). When individuals spend time uncoupled between
 partnerships, and when these single individuals can transmit disease
@@
 growth rate.
-The transmission rates ($\beta_P$, $\beta_D$)
+Increases in the transmission rates ($\beta_P$, $\beta_D$)
 and durations ($D_P$, $D_D$) in the primary and disease stages generally
 decrease the equilibrium virulence, peak virulence, and peak time,
-although the random and pair-formation+epc models have high, relatively
+although the random and pair-formation+extra-pair contact
+models have high, relatively
 constant values with respect to these parameters. 
-\todo{I know you explained to me before why beta and Duration go in
-the same direction but I forget ...}
+\todo{better explanation of why increasing $\beta$/$D$ and
+then scaling $r$ should lead to lower $\rzero$?}
+
 The base contact rate increases virulence and peak time in almost all
-cases, although the pair-formation+epc model is again relatively insensitive.
+cases, although the pair-formation+extra-pair contact
+model is again relatively insensitive.
 The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
@@
 \item fig 1. tweak size? remove redundant y axes/labels?
-\item fig 2. 6-panel figure? log scale? boundary lines instead of ribbons for edges of envelopes? move labels away from axes?
-\item fig 3. change order (flip instswitch and instswitch + epd)
-\item fig 4. squash panels together?
-\item fig 5.
-sensitivity plot (unscaled plot); check we havetake out plots that are not related ...
-implicit model doesn't use rho, cu or ce; random doesn't use rho, cmean?, cu, ce; instswitch doesn't have rho or cu or ce; models without epc don't have ce or cu; instswitch + epc doesn't have rho and cu (no uncoupled compartment).
-Maybe squash panels together, higher resolution
+\item fig 3. add units to strip labels?
+\item fig 4. fix axis labels; squash panels together
+\item fig 5. sensitivity plot (unscaled plot); fix strip labels for summary statistics; fix x-axis labels
 \end{itemize}

### 2016-05-13 ms/ms.Rnw
updates to TODO; fig 1 legend tweak

@@
     geom_line()+labs(x=tlab,y=vlab)+
-    theme(legend.position=c(0.8,0.2))
+    ## fixme: adjust legend position
+    theme(legend.position=c(0.75,0.25))
 ###
@@
 breaks <- seq(limits[1],limits[2], by = 0.5)
+## FIXME: adjust vjust/hjust
 afun <- function(label,size=8) {
@@
 
-<<testrev,echo=FALSE,eval=FALSE>>=
-set.seed(101)
-zz <- data.frame(x=rnorm(3000),f=factor(rep(letters[1:3],each=1000)))
-g0 <-  ggplot(zz,aes(x=f,y=x,fill=f))+geom_violin()
-grid.arrange(
-    g0,
-    g0+ guides(fill=guide_legend(reverse=TRUE)),
-    nrow=2)
-@
-
 <<plot_sens,dev="png",echo=FALSE,out.width="\\textwidth",fig.width=8,fig.height=3.5,dpi=240,fig.cap="Sensitivity plot.
@@
 \begin{itemize}
-\item clean up code and results
-\item finish writing!
-\item hard parts of the results
+\item use PLoS template for ms
+\item check on code/documentation on repo
+\item hard parts of the results:
 \begin{itemize}
@@
 \item cite \cite{LipsitchNowak1995}
-\end{itemize}
-
+\item figures:
   \begin{itemize}
-\item fig 1. tweak size? remove redundant y axes/labels?
+  \item fig 1. tweak label position in (a)
   \item fig 3. add units to strip labels?
   \item fig 4. fix axis labels; squash panels together
-\item fig 5. sensitivity plot (unscaled plot); fix strip labels for summary statistics; fix x-axis labels
+  \item fig 5. fix strip labels for summary statistics; fix x-axis labels
+  \item create fig A1 (scaled sensitivity plot) for supp materials
+  \end{itemize}
 \end{itemize}

### 2016-05-17 ms/plos_latex.tex
started working moving ms to plos temp

% Template for PLoS
% Version 3.2 March 2016
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Effects of Epidemiological Structure on the Transient Evolution of HIV Virulence} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Name1 Surname\textsuperscript{1,2\Yinyang},
Name2 Surname\textsuperscript{2\Yinyang},
Name3 Surname\textsuperscript{2,3\textcurrency},
Name4 Surname\textsuperscript{2},
Name5 Surname\textsuperscript{2\ddag},
Name6 Surname\textsuperscript{2\ddag},
Name7 Surname\textsuperscript{1,2,3*},
with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* correspondingauthor@institute.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions. In
  general changes in fitness landscapes over the course of an epidemic
  will select for higher virulence during the early,
  exponential-growth phase of the epidemic, but quantitative outcomes
  can depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser, Shirreff, and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use very simple implicit
  representations of the transmission process that ignore the
  partnership dynamics that previous research has found to be critical
  in predicting epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the peak value
  of virulence (SPVL) and the time at which the peak occurs across all
  models and across a Latin hypercube sample that captures a realistic
  range of partnership dynamic parameters for sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. Our primary result is that, for this
  particular model setting, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}

The evolution of pathogen virulence is a fundamental process in
evolutionary biology, of both theoretical and (potentially) practical
importance. The trade-off theory \cite{Ebert1999} --- which
postulates that parasite virulence can be explained as the long-term
evolutionary outcome of a saturating relationship between parasite
clearance rate and transmission rate --- has been criticized
\cite{EbertBull2003,alizon_adaptive_2015}, but has also been
successfully applied in a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}. One
particularly interesting application of these ideas is the work by
Fraser \textit{et al}. showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in a study of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
within-couple transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately [XXX] years and then declining
slightly to a long-stable virulence level.

The work of Shirreff \textit{et al.} \cite{shirreff_transmission_2011}, and particularly the predicted
transient peak in HIV virulence midway through the epidemic,
highlights the importance of interactions between epidemiological and
evolutionary factors \cite{day_virulence_2004,alizon_price_2009}.
However, despite the attention to mechanistic detail at the individual
or physiological level, the epidemiological structures used in these
models are relatively simple.

As we discuss in detail below, the
existing models of HIV eco-evolutionary dynamics either use implicit
models that incorporate the average effects of within-couple sexual
contact --- without representing the explicit dynamics of pair
formation and dissolution or accounting for extra-partnership contact
--- or use an agent-based formulation with parameters that effectively
lead to random mixing among infected and uninfected individuals. Here
we explore the effects of incorporating \emph{explicit}
epidemiological structure in eco-evolutionary models.

We add complexity to the epidemiological model following the general approach of Champredon \textit{et al.} \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships. At the same time, our analysis somewhat simplifies the models of Shirreff \textit{et al.} \cite{shirreff_transmission_2011}, for computational tractability; we check that our qualitative results are not sensitive to these simplifications. In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity. In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following Champredon \textit{et al.} \cite{champredon_hiv_2013} in using a Latin hypercube design. For each model run, we compute a set of metrics (peak virulence, timing of virulence peak, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.

\section*{Materials and Methods}

As our primary goal is to explore how different epidemiological structures (i.e. partnership dynamics and contact structures) affect our conclusions about the evolution of virulence, our models use a simplified description of within-host dynamics and heritability derived from 
Shirreff \textit{et al.}'s multi-strain evolutionary model \cite{shirreff_transmission_2011}. Like Shirreff \textit{et al.}, we use a simple susceptible-infected-susceptible demographic formulation; rather than modeling birth and death (or more specifically, recruitment into the sexually active population and death), we assume that whenever an individual dies from infection, another enters the susceptible compartment.

\subsection*{Infection dynamics}

Like Shirreff \textit{et al.} \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than virulence (i.e. rate of progression to AIDS) itself. 
In contrast to Shirreff \textit{et al.}, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period. We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \textit{et al.}'s model) and taking the duration-weighted average of transmission rates of three stages. Thus the within-couple transmission rate, $\beta$, for our models is given by:
\begin{equation}
\label{eq:1}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D}
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the \textbf{P}rimary and \textbf{D}isease stages
of infection are independent of the host's SPVL. Following Shirreff \textit{et al.}, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the \textbf{A}symptomatic stage are Hill functions of the SPVL:

\begin{equation}
\label{eq:2}
\begin{split}
D_A(\alpha) &= \frac{D_{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\beta_{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The \textbf{u}ncoupled and \textbf{e}xtra-couple transmission rates are scaled by multiplying the \textbf{w}ithin-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$.

\subsection*{Mutation}

Like Shirreff \textit{et al.} \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \textit{et al.}'s evolutionary model slightly by using a one-to-one genotype-phenotype mapping. 
The mutational process in our model is directly taken from Shirreff \textit{et al.}. Over the course of infection, mutation occurs within the host. However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection for simplicity (and is not further affected by within-host mutation). Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual. First, the distribution of $\log_{10}$ SPVL is discretized into a vector:
\begin{equation}
\label{eq:3}
\alpha_i =  (\alpha_{max} - \alpha_{min})\frac{(i-1)}{n-1} + \alpha_{min} \qquad i = 1,2,3, \dots n.
\end{equation}
We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \textit{et al.} \cite{shirreff_transmission_2011} (Figure~\ref{fig:panel3}) we use $n=51$ (i.e. a bin width of 0.05 $\log_{10}$ SPVL for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 $\log_{10}$ SPVL), which we use for all other simulations.

We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have $\log_{10}$ SPVL of $\alpha_j$ given that the infector has $\log_{10}$ SPVL of $\alpha_i$. Finally, the probabilities are normalized so that each row sums to 1:
\begin{equation}
M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\alpha_{max} + d/2;i) - \Phi(\alpha_{min} - d/2;i)},
\end{equation}
where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\alpha_{max} - \alpha_{min})/(n-1)$. Transmission rate and disease induced mortality rates are discretized into a vector as well:
\begin{equation}
\begin{aligned}
\beta_i &= \beta(\alpha_i)\\
\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}
\end{aligned}
\end{equation}

\subsection*{Contact structure and partnership dynamics}

We developed six multi-strain evolutionary models, designed to cover a gamut between Champredon text{et al.}'s relatively realistic \cite{champredon_hiv_2013} and Shirreff \textit{et al.}'s relatively simplistic \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation and (2) zero vs. positive extra-partnership sexual contact and transmission on the evolution of mean $\log_{10}$ SPVL.

Our first four models consider explicit partnership dynamics and are based on Champredon text{et al.}'s model \cite{champredon_hiv_2013}. Models 1 and 2 assume non-instantaneous partnership formation (i.e. individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status. $S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals. $SS$ is the number of susceptible-susceptible couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples. Model 1 includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas model 2 only considers within-couple transmission. Models 3 and 4 assume instantaneous partnership formation and thus consist of only the three partnered states: $SS$, $SI$, and $II$. Parallel to model 1 and 2, model 3 includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and model 4 only considers within-couple transmission.

In contrast, models 5 and 6 are not explicitly structured.  Model 5 is an implicit serial monogamy model based on the epidemiological model used by Shirreff \textit{et al.} \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}. Finally, model 6 is a simple random-mixing model.

The base model (i.e. model 1) for the first four models is an extension of Champredon text{et al.}'s model \cite{champredon_hiv_2013}. Individuals in single compartment acquire a partner at a rate, $\rho$, and partnerships dissolve at a rate, $c$. Infected individuals in a discordant partnership infect susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate). Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing. Extra-couple and uncoupled transmission are modeled in a same way as Champredon text{et al.}'s model. All the details have been adapted to a multi-strain scenario. Model 2, 3, and 4 are derived from the base model by removing epidemiological details (partnership formation and uncoupled/extra-couple contact). Model details are explained in the appendix.

% Results and Discussion can be combined.
\section*{Results}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

% Place tables after the first paragraph in which they are cited.
\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{table}

%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}
\subsubsection*{3rd Level Heading.} 
Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit. Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat.

\begin{enumerate}
	\item{react}
	\item{diffuse free particles}
	\item{increment time by dt and go to 1}
\end{enumerate}

\subsection*{Sed ac quam id nisi malesuada congue.}

Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.

\begin{itemize}
	\item First bulleted item.
	\item Second bulleted item.
	\item Third bulleted item.
\end{itemize}

\section*{Discussion}
Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero~\cite{bib3}.

\section*{Conclusion}

CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.

Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed suscipit iaculis mi at feugiat. Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna. Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed. Nam condimentum sem eget mollis euismod. Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex. Nunc est dui, sodales sed mauris nec, auctor sagittis leo. Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor. For more information, see \nameref{S1_Appendix}.

\section*{Supporting Information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 File.}
\label{S1_File}
{\bf Lorem Ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Video.}
\label{S1_Video}
{\bf Lorem Ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Appendix.}
\label{S1_Appendix}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\section*{Acknowledgments}
Cras egestas velit mauris, eu mollis turpis pellentesque sit amet. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here.
%

\bibliographystyle{plos2015}
\bibliography{virulence}

\end{document}

### 2016-05-18 ms/plos_latex.tex
done for now

@@
 \begin{equation}
-\label{eq:1}
 \beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D}
@@
 \begin{equation}
-\label{eq:2}
 \begin{split}
@@
 \begin{equation}
-\label{eq:3}
 \alpha_i =  (\alpha_{max} - \alpha_{min})\frac{(i-1)}{n-1} + \alpha_{min} \qquad i = 1,2,3, \dots n.
@@
 a bin width of 0.05 $\log_{10}$ SPVL for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 $\log_{10}$ SPVL), which we use for all other simulations.
+
+\begin{figure}[!h]
+\caption{{\bf Baseline dynamics.}
+Time series of mean population virulence ($\log_{10}$ SPVL).
+(a) Shirreff model, effects of varying $r$.
+(b) Effects of varying initial infectious density $I(0)$.
+(c) Effects of varying initial mean virulence $\alpha(0)$.}
+\label{fig:panel3}
+\end{figure}
 
@@
 
+\subsection*{Latin hypercube sampling}
+
+Despite considerable effort by many researchers \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \textit{et al.} \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
+In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
+We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\beta_{max}$, $\beta_{50}$, $\beta_k$, $D_{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \textit{et al.}'s point estimates throughout \cite{shirreff_transmission_2011}.
+
+Latin hypercube sampling is done as in Champredon \textit{et al.} \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
+
+\begin{equation}
+z_i = \exp\left(\log(z_{min}) + [\log(z_{max}) - \log(z_{min})] \frac{i-1}{N-1}\right) \qquad i = 1,...,N.
+\end{equation}
+For simplicity, we assume that all parameters are uniformly distributed on the log scale.
+Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$).
+Then, each column is replaced with a random permutation series of itself.
+Now, each row contains a different parameter set that is used for each simulation run.
+
+Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
+Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \textit{et al.} \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \textit{et al.} \cite{hollingsworth_hiv1_2008}.
+The remaining parameters are taken from Shirreff \textit{et al.} \cite{shirreff_transmission_2011}.
+
+The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix); we need to pick a reasonable range for it.
+Champredon \textit{et al.} \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
+Given Champredon \textit{et al.}'s parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5.
+Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$.
+Although this adds more uncertainty to the parameter $c_u$ --- Champredon \textit{et al.}'s range implies a 5-fold difference whereas ours gives a 25-fold difference --- as there is not much known about the uncoupled transmission rate, we consider the wider range appropriate.
+
+\subsection*{Simulation runs}
+
+One of the most difficult parts of model comparison is finding parameter sets that are commensurate against many different model structures.
+For the most part, our models are too complex to easily derive analytical correspondences among them.
+Given a numerical criterion, such as $r$ (initial exponential growth rate) or ${\cal R}_0$ (intrinsic reproductive number), we can adjust one or more parameters by brute force to ensure that all of the models match according to that criterion.
+While ${\cal R}_0$ is often considered the most fundamental property of an epidemic, and might thus seem to be a natural goal, here we focus on matching the initial growth rate $r$ for several reasons.
+First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than ${\cal R}_0$.
+Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves \cite{ma_estimating_2014}, while ${\cal R}_0$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease \cite{wallinga_how_2007}.
+Thus, we scale a parameter so that every run has the same initial exponential growth rate in the incidence.
+
+In order to allow for all models to have equal initial exponential growth rate, $r$, we need to pick a parameter, $s$, so that $\lim_{s\to 0} r(s) = 0$ and $\lim_{s\to\infty}  r(s) = \infty$.
+As adjusting either partnership change rate (i.e.
+partnership formation and dissolution rate) or transmission rate does not fulfill this requirement for certain models, we decided to scale partnership change rate and dissolution rate by the same factor of $\gamma$: $\beta_{adj} =  \gamma \beta_{base}$, $c_{adj} = \gamma c_{base}$, $\rho_{adj} = \gamma \rho_{base}$.
+Since transmission rate is adjusted by the scale of $\gamma$, uncoupled and extra-couple transmission rates are adjusted as well.
+For models 3, 4, and 5, all of which assume instantaneous partnership (and hence do not track single individuals), only the transmission rate and partnership dissolution rate (in this case equivalent to the partnership change rate) are adjusted.
+
+We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
+of mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$.
+After each run, initial exponential growth rate is calculated.
+Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.
+
+For each model we derive the following summary statistics: peak virulence, peak time, equilibrium virulence, and relative peak virulence.
+The transient phase of an epidemic is often characterized by high virulence, and we define peak virulence as the maximum virulence during this phase.
+It is simply calculated by taking the maximum value from the virulence trajectory, and peak time is the time at which the maximum value is reached.
+Once the epidemic enters the endemic phase, evolution of virulence stabilizes and reaches equilibrium.
+Equilibrium virulence is calculated by taking the mean virulence at 4000 years.
+Although most simulations reach equilibrium much earlier, we set our time horizon at a much later date as some simulation runs have slow rate of evolution depending on the parameter set and model assumptions.
+
+We focus on these statistics for the following two reasons.
+First of all, knowing the possible ranges for the peak virulence allows us to estimate the worst-case scenario for the HIV and other sexually transmitted disease epidemics.
+Pathogens may already have evolved towards high virulence during the early stages of an epidemic, by the time it is observed by public health authorities.
+Understanding how virulent a pathogen can evolve before an epidemic begins can be helpful for controlling the disease.
+Furthermore, knowing the initial virulence, peak virulence, timing of the peak virulence, and equilibrium virulence provide sufficient detail to identify the shape of the virulence trajectory.
+During an epidemic outbreak, it is difficult to observe virulence evolution.
+Specifically, in the case of HIV and other sexually transmitted diseases, slow evolutionary time-scale makes observing changes in the mean virulence even more challenging.
+Knowing the ranges of these statistics can help real-time virulence evolution prediction during an epidemic less troublesome.
+
 % Results and Discussion can be combined.
 \section*{Results}
-Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
 
-% Place tables after the first paragraph in which they are cited.
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-\begin{tabular}{|l|l|l|l|l|l|l|l|}
-\hline
-\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
-$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-\end{tabular}
-\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-\end{flushleft}
-\label{table1}
-\end{adjustwidth}
-\end{table}
+Our simplifications of Shirreff \textit{et al.}'s model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
+As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \textit{et al.}'s full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
+However, our model produces lower peak virulence ($\approx 4.3$ vs.
+$\approx 4.6$ $\log_{10}$ SPVL)
+and equilibrium virulence $\approx 4.25$ vs. $\approx 4.5$ $\log_{10}$ SPVL) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
 
-%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
-\subsection*{\lorem\ and \ipsum\ Nunc blandit a tortor.}
-\subsubsection*{3rd Level Heading.} 
-Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
+Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases.
+Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b).
+Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Figure~\ref{fig:panel3}c).
 
-\begin{enumerate}
-	\item{react}
-	\item{diffuse free particles}
-	\item{increment time by dt and go to 1}
-\end{enumerate}
+Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (Figure~\ref{fig:virtraj}).
+Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 $\log_{10}$ SPVL), virulence peaks in all models between 200 and 300 years.
 
-\subsection*{Sed ac quam id nisi malesuada congue.}
+\begin{figure}[!h]
+\caption{{\bf Envelopes of virulence trajectories under all models.}
+All models were run until $t=4000~\textrm{years}$; truncated series are shown here.}
+\label{fig:virtraj}
+\end{figure}
 
-Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero.
+Our chosen summary statistics (peak time, peak virulence, equilibrium
+virulence, and relative peak virulence) all vary considerably across models
+\ref{fig:unidist}.
+We first consider the models of intermediate realism (implicit,
+instantaneous-switching with and without extra-pair contact, and
+pair formation without extra-pair contact). Some parameter
+sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL);
+these same sets lead to correspondingly low
+peak virulence ($< 3.5 \log_{10}$ SPVL) and early peak times (before 200 years), 
+but high relative peaks ($>1.3$)
+(Figure~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
+At the opposite extreme, parameter sets that produce high equilibrium virulence 
+also produce late peaks ($> 200 \text{ years}$), 
+high peak virulence, and low relative peaks ($\approx 1.05$).
+The pair-formation without extra-pair contact and implicit models
+occasionally have parameter sets that select for such low virulence across
+the board that they never exceed their initial virulence, leading to a tail
+of peak times near zero.
 
-\begin{itemize}
-	\item First bulleted item.
-	\item Second bulleted item.
-	\item Third bulleted item.
-\end{itemize}
+\begin{figure}[!h]
+\caption{{\bf Univariate distributions of summary statistics.}
+The distribution of equilibrium virulence for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
+\label{fig:unidist}
+\end{figure}
+
+\begin{figure}[!h]
+\caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
+Dashed line in equilibrium vs. peak virulence plot shows 1:1 line.}
+\label{fig:pairplot}
+\end{figure}
+
+The most striking aspect of the univariate comparisons in
+Figure~\ref{fig:unidist}, (and the bivariate comparisons in
+Figure~\ref{fig:pairplot}) is the match between the output of the
+least (random mixing) and the most complex (pair formation with
+extra-pair contact) models. The random-mixing model has lower
+variability, because it is unaffected by uncertainty in pair formation
+and extra-pair contact parameters, but otherwise the virulence
+dynamics of these two extreme models are remarkably similar.
+This phenomenon is driven by the strength of extra-pair contact in the
+model with explicit pair formation and extra-pair contact 
+(``pairform+epc'' in Figures~\ref{fig:virtraj}-\ref{fig:plot_sens}). When individuals spend time uncoupled between
+partnerships, and when these single individuals can transmit disease
+to coupled individuals, the resulting unstructured mixing overwhelms
+the effect of structured mixing within couples, leading to mixing
+that is effectively close to random.
+
+\begin{figure}[!h]
+\caption{{\bf Sensitivity plot.}
+For each parameter in the Latin hypercube sample and each summary statistic, shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
+prior to adjusting the parameters to achieve the standard initial epidemic growth rate.}
+\label{fig:plot_sens}
+\end{figure}
+
+Plotting the bivariate result distributions (Figure~\ref{fig:pairplot})
+shows that most of the relationships among the summary statistics
+are monotonic, except those involving the relative peak virulence
+(bottom row). Changes in parameters that increase the equilibrium
+virulence initially increase the peak virulence even more, so
+that the relative peak virulence increases as well, but beyond
+an equilibrium virulence of about 2.5 $\log_{10}$ SPVL the peak virulence
+increases slower than the equilibrium virulence, leading to a
+decrease in the relative peak virulence.
+
+The bivariate relationships also help distinguish the results of 
+different models with similar univariate distributions. While the
+relationship between equilibrium virulence and peak time is
+similar for all model structures (top left panel), the other
+relationships are more separate. In particular, the implicit
+and pair-formation (without extra-pair contact) are very similar
+to each other, but distinct from the other models. We still do
+not have a convincing explanation for this distinction; we
+would have expected the implicit model to be most similar to the
+the instantaneous-switching model without extra-pair contact,
+which most closely matches its derivation. However, we note
+that the implicit model derivation is based on defining
+the force of infection to match a scaled version of ${\cal R}_0$,
+and as such would be expected to match the equilibrium behaviour
+but not necessarily the epidemic-phase behaviour of a model
+with explicit partnership dynamics.
+
+Finally, the sensitivity plots (Figure~\ref{fig:plot_sens}) show the effects 
+of each parameter on the summary statistics. In almost every case the
+effects of the parameters are monotonic; note that
+the plot shows the effects of the \emph{unscaled} parameters, i.e.
+before they have been adjusted to achieve a standard initial epidemic
+growth rate.
+Increases in the transmission rates ($\beta_P$, $\beta_D$)
+and durations ($D_P$, $D_D$) in the primary and disease stages generally
+decrease the equilibrium virulence, peak virulence, and peak time,
+although the random and pair-formation+extra-pair contact
+models have high, relatively
+constant values with respect to these parameters.
+
+The base contact rate increases virulence and peak time in almost all
+cases, although the pair-formation+extra-pair contact
+model is again relatively insensitive.
+The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
+virulence in the instantaneous-switching+epc model, but not the pair-formation+epc
+model (probably because the uncoupled individuals present in the pair-formation+epc
+model make extra-pair contact by coupled individuals less important).
+Surprisingly, neither of the pair-formation models is particularly sensitive to the 
+rate of partnership formation ($\rho$); the rate of uncoupled contact
+increases virulence and peak time in the pair-formation+epc model, 
+which is the only model to which it applies.
 
 \section*{Discussion}
-Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat euismod diam.
-Proin rutrum vel massa non gravida.
-Quisque tempor sem et dignissim rutrum.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi at justo vitae nulla elementum commodo eu id massa.
-In vitae diam ac augue semper tincidunt eu ut eros.
-Fusce fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
-Aliquam in enim semper, aliquam massa id, cursus neque.
-Praesent faucibus semper libero~\cite{bib3}.
+
+All models must simplify the world.  Many constraints --- such as data
+availability, computation time, or code complexity --- drive the need
+for parsimony, with different constraints applying in different
+contexts. The critical question that modelers must ask is whether the
+simplified model gives adequate answers, or whether the
+simplifications have led to qualitatively or quantitatively incorrect
+conclusions.  This question is especially important for modelers who
+are hoping that their conclusions will guide management decisions.
+
+In the particular example of HIV virulence eco-evolutionary dynamics
+and epidemiological, we reach the slightly ironic conclusion that the
+effort put into building a more realistic model essentially cancels
+out, leaving us in the same position as if we had ignored the problem
+of epidemiological complexity entirely and used a naive random-mixing
+contact model. Herbeck \textit{et al.} \cite{herbeck_hiv_2014} build a
+network model of partnerships, but they set the ``partnership''
+duration to 1 day --- very unrealistic in epidemic terms, but perhaps
+actually more true to real-world HIV epidemiological dynamics than a
+model with realistic partnership durations that neglects extra-pair
+contact \cite{herbeck2016evolution}. Adding still further realism
+might push the results farther away from the random-mixing model. For
+example, our model forms partnerships randomly, and assumes that
+extra-pair contact is randomly mixing across the population;
+one could instead implement extra-pair contact as arising from
+multiple concurrent partnerships (some, such as contact with sex
+workers, of very short duration), or more structured partnership
+formation (by age, ethnicity, or behaviour group). The effects of
+other realistic complications such as explicit modeling of two
+sexes (both in contact structure and differential transmission
+probabilities), temporal and spatial variation in epidemic processes,
+or presence of genetic variation in hosts are harder to predict.
+
+Parameterization is one of the biggest challenges of epidemiological
+modeling. In addition to following Champredon \textit{et al.} \cite{champredon_hiv_2013} 
+by doing Latin hypercube
+sampling across a wide range of epidemiological parameters, we 
+calibrated each set of parameters to the same initial epidemic
+growth rate, chosen to match the results of previous models
+\cite{shirreff_transmission_2011}.  The previous models drew their
+parameters from previous cohort studies from the 1990s
+\cite{wawer2005rates,hollingsworth_hiv1_2008}
+rather than doing any explicit calibration to epidemic curves,
+but they give reasonable order-of-magnitude
+growth rates ($\approx 0.04~\textrm{year}^{-1}$)
+for the early stages of the HIV epidemic (although considerably
+lower than estimates of $\approx 0.07-0.1~\textrm{year}^{-1}$
+based on population genetic reconstructions \cite{faria_early_2014}).
+However, our reason for calibrating was not to match any
+specific observed epidemic, but rather to make sure that
+we were making meaningful comparisons across a range of
+models with radically different epidemiological structures, and
+hence involving different interpretations of the same quantitative
+parameters.  For example, in models with instantaneous switching the
+partnership dissolution rate $c$ is identical to the partnership
+formation rate; in models with explicit partnership formation,
+the partnership formation rate is also $c$ at equilibrium,
+but might vary over the course of an epidemic.
+It is not obvious whether models with equal parameters but
+different structures should be directly compared; calibration
+solves this problem.
+
+More generally, any model that wants to be
+taken seriously for management and forecasting purposes should
+be calibrated to \emph{all} available data, using informative
+priors to incorporate both realistic distributions of uncertainty
+for all parameters from independent measurements \cite{elderd_uncertainty_2006}
+and calibration from population-level observations of epidemic
+trajectories. For practical (rather than exploratory) purposes,
+such a procedure would also be an improvement on the common --- although not universal --- %
+practice of assessing uncertainty over uniform ranges, rather than
+distributions that allow more continuous variation in support over
+the range of a parameter.
+
+Researchers have documented that HIV virulence and set-point viral
+load are changing, on time scales comparable to those portrayed here
+(e.g., compare Figure~\ref{fig:virtraj} to Herbeck \text{et al}'s
+estimated rate of change of 1.3 $\log_{10}$ SPVL per century [95\% CI -0.1 to
+  3] \cite{herbeck_is_2012}), and have begun to build relatively realistic models that
+attempt to describe how interventions such as mass antiretroviral
+therapy (ART) can be expected to change the trajectory of virulence
+evolution \cite{payne_impact_2014,herbeck2016evolution}.  While these
+efforts are well-intentioned, we caution that epidemiological and
+other structural details that are currently omitted from these models
+could significantly change their conclusions.
 

### 2016-05-18 ms/plos_latex.tex
updates to PLoS version, doi glitch in bib

@@
 %% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
+\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
+\newcommand{\khalf}{\left(\frac{1}{2}\right)^{\delta_{ij}}}  % (1/2)^kronecker
+\newcommand{\Lspvl}{$\log_{10}$ SPVL}
+\newcommand{\rzero}{{\mathcal R}_0}
+\newcommand{\etal}{\textit{et al.}}
 
@@
 \\
-Name1 Surname\textsuperscript{1,2\Yinyang},
-Name2 Surname\textsuperscript{2\Yinyang},
-Name3 Surname\textsuperscript{2,3\textcurrency},
-Name4 Surname\textsuperscript{2},
-Name5 Surname\textsuperscript{2\ddag},
-Name6 Surname\textsuperscript{2\ddag},
-Name7 Surname\textsuperscript{1,2,3*},
-with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
+Sang Woo Park\textsuperscript{1}
+Benjamin M. Bolker\textsuperscript{2,3,4,*}
 \\
 \bigskip
-\textbf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{1} Program in Health Sciences, McMaster University, Hamilton, Ontario, Canada
 \\
-\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{2} Department of Mathematics \& Statistics,  McMaster University, Hamilton, Ontario, Canada
 \\
-\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{3} Department of Biology,  McMaster University, Hamilton, Ontario, Canada
+\\
+\textbf{4} Institute for Infectious Disease Research,  McMaster University, Hamilton, Ontario, Canada
 \\
@@
 % Primary Equal Contribution Note
-\Yinyang These authors contributed equally to this work.
+%\Yinyang These authors contributed equally to this work.
 
@@
 % Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-\ddag These authors also contributed equally to this work.
+%\ddag These authors also contributed equally to this work.
 
 % Current address notes
-\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
 % \textcurrency b Insert second current address 
@@
 % Deceased author note
-\dag Deceased
+%\dag Deceased
 
 % Group/Consortium Author Note
-\textpilcrow Membership list can be found in the Acknowledgments section.
+%\textpilcrow Membership list can be found in the Acknowledgments section.
 
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* correspondingauthor@institute.edu
+* bolker@mcmaster.ca
 
@@
 % Please keep the abstract below 300 words
+% current is 298!
 \section*{Abstract}
@@
 particularly interesting application of these ideas is the work by
-Fraser \textit{et al}. showing that HIV appears to satisfy the prerequisites of
+Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
 the tradeoff theory: in a study of discordant couples (i.e. long-term
@@
 
-The work of Shirreff \textit{et al.} \cite{shirreff_transmission_2011}, and particularly the predicted
+The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted
 transient peak in HIV virulence midway through the epidemic,
@@
 
-We add complexity to the epidemiological model following the general approach of Champredon \textit{et al.} \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships.
-At the same time, our analysis somewhat simplifies the models of Shirreff \textit{et al.} \cite{shirreff_transmission_2011}, for computational tractability; we check that our qualitative results are not sensitive to these simplifications.
+We add complexity to the epidemiological model following the general approach of Champredon \etal\ \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships.
+At the same time, our analysis somewhat simplifies the models of Shirreff \etal\ \cite{shirreff_transmission_2011}, for computational tractability; we check that our qualitative results are not sensitive to these simplifications.
 In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity.
-In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following Champredon \textit{et al.} \cite{champredon_hiv_2013} in using a Latin hypercube design.
+In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following Champredon \etal\ \cite{champredon_hiv_2013} in using a Latin hypercube design.
 For each model run, we compute a set of metrics (peak virulence, timing of virulence peak, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.
-
-\section*{Materials and Methods}
 
@@
 partnership dynamics and contact structures) affect our conclusions about the evolution of virulence, our models use a simplified description of within-host dynamics and heritability derived from
-Shirreff \textit{et al.}'s multi-strain evolutionary model \cite{shirreff_transmission_2011}.
-Like Shirreff \textit{et al.}, we use a simple susceptible-infected-susceptible demographic formulation; rather than modeling birth and death (or more specifically, recruitment into the sexually active population and death), we assume that whenever an individual dies from infection, another enters the susceptible compartment.
+Shirreff \etal's multi-strain evolutionary model \cite{shirreff_transmission_2011}.
+Like Shirreff \etal, we use a simple susceptible-infected-susceptible demographic formulation; rather than modeling birth and death (or more specifically, recruitment into the sexually active population and death), we assume that whenever an individual dies from infection, another enters the susceptible compartment.
 
@@
 
-Like Shirreff \textit{et al.} \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than virulence (i.e.
+Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than virulence (i.e.
 rate of progression to AIDS) itself.
-In contrast to Shirreff \textit{et al.}, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period.
-We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \textit{et al.}'s model) and taking the duration-weighted average of transmission rates of three stages.
+In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period.
+We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \etal's model) and taking the duration-weighted average of transmission rates of three stages.
 Thus the within-couple transmission rate, $\beta$, for our models is given by:
@@
 
-Like Shirreff \textit{et al.} \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \textit{et al.}'s evolutionary model slightly by using a one-to-one genotype-phenotype mapping.
-The mutational process in our model is directly taken from Shirreff \textit{et al.}.
+Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \etal's evolutionary model slightly by using a one-to-one genotype-phenotype mapping.
+The mutational process in our model is directly taken from Shirreff \etal.
 Over the course of infection, mutation occurs within the host.
@@
 Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual.
-First, the distribution of $\log_{10}$ SPVL is discretized into a vector:
+First, the distribution of \Lspvl\ is discretized into a vector:
 \begin{equation}
@@
 \end{equation}
-We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \textit{et al.} \cite{shirreff_transmission_2011} (Figure~\ref{fig:panel3}) we use $n=51$ (i.e.
-a bin width of 0.05 $\log_{10}$ SPVL for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 $\log_{10}$ SPVL), which we use for all other simulations.
+We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (Fig.~\ref{fig:panel3}) we use $n=51$ (i.e.
+a bin width of 0.05 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 \Lspvl), which we use for all other simulations.
 
@@
 \caption{{\bf Baseline dynamics.}
-Time series of mean population virulence ($\log_{10}$ SPVL).
+Time series of mean population virulence (\Lspvl).
 (a) Shirreff model, effects of varying $r$.
@@
 
-We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have $\log_{10}$ SPVL of $\alpha_j$ given that the infector has $\log_{10}$ SPVL of $\alpha_i$.
+We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
 Finally, the probabilities are normalized so that each row sums to 1:
@@
 
-We developed six multi-strain evolutionary models, designed to cover a gamut between Champredon text{et al.}'s relatively realistic \cite{champredon_hiv_2013} and Shirreff \textit{et al.}'s relatively simplistic \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+We developed six multi-strain evolutionary models, designed to cover a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simplistic \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
 Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
 non-instantaneous partnership formation and (2) zero vs.
-positive extra-partnership sexual contact and transmission on the evolution of mean $\log_{10}$ SPVL.
+positive extra-partnership sexual contact and transmission on the evolution of mean \Lspvl.
 
@@
 In contrast, models 5 and 6 are not explicitly structured.
-Model 5 is an implicit serial monogamy model based on the epidemiological model used by Shirreff \textit{et al.} \cite{shirreff_transmission_2011}.
+Model 5 is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
 It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
@@
 
+\begin{table}[h!]
+\caption{Parameter ranges/values.
+Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples.
+Starred (*) parameters (used in Figure~\ref{fig:panel3}) are from \cite{shirreff_transmission_2011}.}
+\centering
+\begin{tabular}{c p{2in} c l}
+\hline 
+Notation & Description & Range/Value & Source\\
+\hline % inserts single horizontal line
+$\rho$ & Partnership formation rate & 1/10-2/5 & \cite{champredon_hiv_2013} \\
+$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) & \cite{champredon_hiv_2013} \\
+$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
+$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
+$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) & \cite{hollingsworth_hiv1_2008} \\
+$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) & \cite{hollingsworth_hiv1_2008} \\
+$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) & \cite{hollingsworth_hiv1_2008,shirreff_transmission_2011} \\
+$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) & \cite{hollingsworth_hiv1_2008} \\
+$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
+$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 & \cite{shirreff_transmission_2011} \\
+$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
+$\tsub{D}{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
+$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 & \cite{shirreff_transmission_2011} \\
+$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
+$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
+$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
+$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
+$n$ & Number of strains & 21 (51*) & Assumption\\
+\hline
+\end{tabular}
+\label{table:parmsTable}
+\end{table}
+
 \subsection*{Latin hypercube sampling}
 
-Despite considerable effort by many researchers \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \textit{et al.} \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
+Despite considerable effort by many researchers \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\beta_{max}$, $\beta_{50}$, $\beta_k$, $D_{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \textit{et al.}'s point estimates throughout \cite{shirreff_transmission_2011}.
+We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\beta_{max}$, $\beta_{50}$, $\beta_k$, $D_{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
 
-Latin hypercube sampling is done as in Champredon \textit{et al.} \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
+Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
 
@@
 Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
-Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \textit{et al.} \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \textit{et al.} \cite{hollingsworth_hiv1_2008}.
-The remaining parameters are taken from Shirreff \textit{et al.} \cite{shirreff_transmission_2011}.
+Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
+The remaining parameters are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
 
 The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix); we need to pick a reasonable range for it.
-Champredon \textit{et al.} \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
-Given Champredon \textit{et al.}'s parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5.
+Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
+Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5.
 Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$.
-Although this adds more uncertainty to the parameter $c_u$ --- Champredon \textit{et al.}'s range implies a 5-fold difference whereas ours gives a 25-fold difference --- as there is not much known about the uncoupled transmission rate, we consider the wider range appropriate.
+Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- as there is not much known about the uncoupled transmission rate, we consider the wider range appropriate.
 
@@
 We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
-of mean $\log_{10}$ SPVL of 3 and epidemic size of $10^{-4}$.
+of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$.
 After each run, initial exponential growth rate is calculated.
@@
 
-Our simplifications of Shirreff \textit{et al.}'s model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \textit{et al.}'s full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a).
+Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
+As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Fig.~\ref{fig:panel3}a).
 However, our model produces lower peak virulence ($\approx 4.3$ vs.
-$\approx 4.6$ $\log_{10}$ SPVL)
-and equilibrium virulence $\approx 4.25$ vs. $\approx 4.5$ $\log_{10}$ SPVL) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
+$\approx 4.6$ \Lspvl)
+and equilibrium virulence $\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
 
 Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases.
-Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b).
-Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Figure~\ref{fig:panel3}c).
+Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (Fig.~\ref{fig:panel3}b).
+Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Fig.~\ref{fig:panel3}c).
 
-Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (Figure~\ref{fig:virtraj}).
-Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 $\log_{10}$ SPVL), virulence peaks in all models between 200 and 300 years.
+Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (Fig.~\ref{fig:virtraj}).
+Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 \Lspvl), virulence peaks in all models between 200 and 300 years.
 
@@
 but high relative peaks ($>1.3$)
-(Figure~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
+(Fig.~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
 At the opposite extreme, parameter sets that produce high equilibrium virulence 
@@
 The most striking aspect of the univariate comparisons in
-Figure~\ref{fig:unidist}, (and the bivariate comparisons in
-Figure~\ref{fig:pairplot}) is the match between the output of the
+Fig.~\ref{fig:unidist}, (and the bivariate comparisons in
+Fig.~\ref{fig:pairplot}) is the match between the output of the
 least (random mixing) and the most complex (pair formation with
@@
 
-Plotting the bivariate result distributions (Figure~\ref{fig:pairplot})
+Plotting the bivariate result distributions (Fig.~\ref{fig:pairplot})
 shows that most of the relationships among the summary statistics
@@
 that the relative peak virulence increases as well, but beyond
-an equilibrium virulence of about 2.5 $\log_{10}$ SPVL the peak virulence
+an equilibrium virulence of about 2.5 \Lspvl\ the peak virulence
 increases slower than the equilibrium virulence, leading to a
@@
 
-Finally, the sensitivity plots (Figure~\ref{fig:plot_sens}) show the effects 
+Finally, the sensitivity plots (Fig.~\ref{fig:plot_sens}) show the effects 
 of each parameter on the summary statistics. In almost every case the
@@
 of epidemiological complexity entirely and used a naive random-mixing
-contact model. Herbeck \textit{et al.} \cite{herbeck_hiv_2014} build a
+contact model. Herbeck \etal\ \cite{herbeck_hiv_2014} build a
 network model of partnerships, but they set the ``partnership''
@@
 Parameterization is one of the biggest challenges of epidemiological
-modeling. In addition to following Champredon \textit{et al.} \cite{champredon_hiv_2013} 
+modeling. In addition to following Champredon \etal\ \cite{champredon_hiv_2013} 
 by doing Latin hypercube
@@
 load are changing, on time scales comparable to those portrayed here
-(e.g., compare Figure~\ref{fig:virtraj} to Herbeck \text{et al}'s
-estimated rate of change of 1.3 $\log_{10}$ SPVL per century [95\% CI -0.1 to
+(e.g., compare Fig.~\ref{fig:virtraj} to Herbeck \text{et al}'s
+estimated rate of change of 1.3 \Lspvl\ per century [95\% CI -0.1 to
   3] \cite{herbeck_is_2012}), and have begun to build relatively realistic models that
@@
 \section*{Acknowledgments}
-Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
-Interdum et malesuada fames ac ante ipsum primis in faucibus.
-Nam id pretium nisi.
-Sed ac quam id nisi malesuada congue.
-Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
+NSERC; C. Fraser
 
@@
 
-\bibliographystyle{plos2015}
 \bibliography{virulence}

### 2016-05-20 ms/plos_latex.tex
moved tabled
moved table
fixed stuff

@@
 \newcommand{\etal}{\textit{et al.}}
+\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
 
@@
 \begin{split}
-D_A(\alpha) &= \frac{D_{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
-\beta_A(\alpha) &= \frac{\beta_{max} V_\alpha ^ {\beta_k}}%
+D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
+\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
 {V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
@@
 \begin{equation}
-\alpha_i =  (\alpha_{max} - \alpha_{min})\frac{(i-1)}{n-1} + \alpha_{min} \qquad i = 1,2,3, \dots n.
+\alpha_i =  (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{(i-1)}{n-1} + \tsub{\alpha}{min} \qquad i = 1,2,3, \dots n.
 \end{equation}
 We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (Fig.~\ref{fig:panel3}) we use $n=51$ (i.e.
-a bin width of 0.05 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 \Lspvl), which we use for all other simulations.
+a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 \Lspvl), which we use for all other simulations.
 
@@
 \begin{equation}
-M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\alpha_{max} + d/2;i) - \Phi(\alpha_{min} - d/2;i)},
+M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
 \end{equation}
-where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\alpha_{max} - \alpha_{min})/(n-1)$.
+where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$.
 Transmission rate and disease induced mortality rates are discretized into a vector as well:
@@
 
+\subsection*{Latin hypercube sampling}
+
+Despite considerable effort by many researchers \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
+In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
+We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
+
+Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
+
+\begin{equation}
+z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1,...,N.
+\end{equation}
+For simplicity, we assume that all parameters are uniformly distributed on the log scale.
+Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$).
+Then, each column is replaced with a random permutation series of itself.
+Now, each row contains a different parameter set that is used for each simulation run.
+
+Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
+Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
+The remaining parameters are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
+
+%Tables need to be placed after the first paragraph in which they are cited.
+
 \begin{table}[h!]
@@
 
-\subsection*{Latin hypercube sampling}
-
-Despite considerable effort by many researchers \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
-In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\beta_{max}$, $\beta_{50}$, $\beta_k$, $D_{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
-
-Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
-
-\begin{equation}
-z_i = \exp\left(\log(z_{min}) + [\log(z_{max}) - \log(z_{min})] \frac{i-1}{N-1}\right) \qquad i = 1,...,N.
-\end{equation}
-For simplicity, we assume that all parameters are uniformly distributed on the log scale.
-Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$).
-Then, each column is replaced with a random permutation series of itself.
-Now, each row contains a different parameter set that is used for each simulation run.
-
-Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
-Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
-The remaining parameters are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
-
 The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix); we need to pick a reasonable range for it.
@@
 As adjusting either partnership change rate (i.e.
-partnership formation and dissolution rate) or transmission rate does not fulfill this requirement for certain models, we decided to scale partnership change rate and dissolution rate by the same factor of $\gamma$: $\beta_{adj} =  \gamma \beta_{base}$, $c_{adj} = \gamma c_{base}$, $\rho_{adj} = \gamma \rho_{base}$.
+partnership formation and dissolution rate) or transmission rate does not fulfill this requirement for certain models, we decided to scale partnership change rate and dissolution rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =  \gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$, $\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$.
 Since transmission rate is adjusted by the scale of $\gamma$, uncoupled and extra-couple transmission rates are adjusted as well.
@@
 Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.
+
+In order to make the computation faster, we simplify the model simulations to calculate $\gamma$ for each run.
+While scaling the parameters to match the exponential growth rate to a target value, models are ran for only 500 years, unlike the real simulations (ran for 4000 years) which are run after the parameters are scaled to obtain the summary statistics.
+Furthermore, we use Runge-Kutta method ode45 to find $\gamma$ whereas lsoda is used for the actual simulations.
 
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
-
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 File.}
-\label{S1_File}
-{\bf Lorem Ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Video.}
-\label{S1_Video}
-{\bf Lorem Ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
 \paragraph*{S1 Appendix.}
 \label{S1_Appendix}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
+{\bf Model details.} Since we use multi-strain models, in which the distribution of \Lspvl\ has been discretized into a vector, we use a matrix notation to describe our models.
+The five states described in the Methods section are replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$, where the subscripts denote the strain that an individual is infected with.
+For example, $I_i$ is number of infected individuals with \Lspvl\ of $\alpha_i$, and $II_{ij}$ is the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$ (independent of order; $II_{ij}$ is synonymous with $II_{ji}$).
+Below, we use the Kronecker delta (i.e.
+$\delta_{ij}=1$ if $i=j$ and 1 otherwise) in a slightly non-standard fashion as an exponent, e.g.
+$2^{\delta_{ij}}$, to set a value to 2 when $i=j$ and 1 otherwise.
 
-\paragraph*{S1 Table.}
-\label{S1_Table}
-{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
+\subsection*{Models 1 and 2: partnership dynamics}
+
+Single individuals acquire partners at rate $\rho$: $S' = - \rho S$ and $I_i' = - \rho I_i$.
+We follow Champredon et al's results and assume that single individuals are distributed into coupled states through binomial distribution\todo{?
+don't understand this wording (ditto below)}:
+\todo{I think we need different notation for partnership formation (i.e. LHS is wrong)}
+
+\begin{equation}
+\begin{aligned}
+(SS)' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)}\\
+(SI_i)' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
+(II_{ij})' &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
+\end{aligned}
+\end{equation}
+In a single-strain model, both individuals in a concordant partnership, $II$, enter the single-infected compartment, $I$, when a partnership dissolves: $I' = 2 c II'$.
+However, this is not the case in a multi-strain model.
+If infected individuals in an $II$ partnership have different strains, they will enter different compartments; if they are both infected with the same strain, they both enter the same compartment.
+As described above, we use the Kronecker delta notation to implement this difference between the $i=j$ and $i\neq j$ cases.
+
+Partnerships dissolve at rate $c$: $SS' = -c SS$, $SI_i' = -c SI_i$, and $II_{ij} = - c II_{ij}$.
+Unlike single strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains which the individuals in concordant partnership are infected with (i.e.
+both partners in $II_{ii}$ enter $I_i$ whereas only one partner in $II_{ij}$ enters $I_i$).
+
+\begin{equation}
+\begin{aligned}
+S' &= 2 c SS + \sum_k c SI_k \\
+I_i' &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}\\
+\end{aligned}
+\end{equation}
+Combining partnership formation and dissolution process yields the following equation:
+
+\begin{equation}
+\begin{aligned}
+S' &= - \rho S + 2 c SS + \sum_k c SI_k \\
+I_i' &= - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}\\
+SS' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS\\
+SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i\\
+II_{ij}' &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k} - c II_{ij}
+\end{aligned}
+\end{equation}
+
+\subsection*{Models 1 and 2: infection}
+
+Within-couple transmission occurs in both models.
+An infected partner in $SI$ partnership transmits virus to a susceptible partner, and partnership state becomes $II$: $SI_i' = - \beta_i SI_i$.
+Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $II_{ij}' = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$.
+On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $II_{ii}' = M_{ii} \beta_i SI_i$.
+Using the Kronecker delta notation, we obtain following set of equations that describe within-couple transmission dynamics:
+
+\begin{equation}
+\begin{aligned}
+SI_i' &= - \beta_i SI_i\\
+II_{ij} &=  \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)
+\end{aligned}
+\end{equation}
+
+Champredon et al define the proportion of infectious extra-couple and uncoupled contact through the following term:
+
+\begin{equation}
+P = \frac{c_u I + c_e (SI + 2 II)}{c_u (S + I) + 2 c_e(SS + SI + II)}.
+\end{equation}
+Effective uncoupled, $c_u$, and extra couple, $c_e$, contact rate is the product of two terms: uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
+Therefore, the transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above.
+Using this property, we modify the equation above as follows:
+
+\begin{equation}
+P = \frac{r_u I + r_e (SI + 2 II)}{r_u (S + I) + 2 r_e(SS + SI + II)},
+\end{equation}
+where $r_u = c_u/c_w$ and $r_e = c_e/c_w$ are the relative uncoupled/extra-couple contact rates.
+This simplification is useful in a multi-strain model since we cannot multiply a vector with a single value (e.g.
+$c_u S$ in denominator) if we use Champredon et al's equation in its original form.
+Extending the above equation to multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain:
+
+\begin{equation}
+P_i = \frac{r_u I_i + r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_u (S + \sum_k I_k) + r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k 2^\delta_{lk} II_{lk} )},
+\end{equation}
+Using the equation above, we can model uncoupled and extra-couple mixing.
+For convenience, uncoupled and extra-couple transmission rates, $c_u$ and $c_e$, will be replaced with $U_i = r_u \beta_i$ and $E_i = r_e \beta_i$ hereafter.
+
+Single susceptible individuals become infected and enter single infected compartment at the total rate of $\sum_k P_k U_k S$.
+Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $I_i' = \sum_k M_{ki} P_k U_k S$.
+Either partner in an $SS$ partnership can be infected, with the partnership state becoming $SI$, at the total rate of $\sum_i 2 P_i E_i SS$.
+The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $SI_i' = \sum_k 2 M_{ki} P_k E_k SS$.
+Lastly, the susceptible partner in an $SI$ partnership can be infected due to uncoupled/extra-couple contacts and partnership can change to an $II$ partnership.
+As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$.
+The mutation process is similar to that of within-couple transmission.
+The only difference is that the \Lspvl\ of a newly infected partner is not determined by its social partner but from an extra-couple partner (i.e.
+the term $P_i$): $II_{ij}' = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$.
+Combining these equations we get the following set of equations that describe all the transmission dynamics:
+
+\begin{equation}
+\begin{aligned}
+S' =& - \sum_k P_k U_k S\\
+I_i' =& \sum_k M_{ki} P_k U_k S\\
+SS' =&  - \sum_i 2 P_i E_i SS \\
+SI_i' =& \sum_k 2 M_{ki} P_k E_k SS - \beta_i SI_i - \sum_k P_k E_k SI_i\\
+II_{ij}' =& \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) + \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i\\
+&+ M_{ki} P_k E_k SI_j))
+\end{aligned}
+\end{equation}
+
+\subsection*{Model 1 and 2 - Disease induced mortality}
+
+The disease induced death rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
+Since we assume an SIS formulation, where infected individuals that die from infection are immediately replaced by an individual in the single susceptible compartment, we obtain the following equation for single infected individuals:
+
+\begin{equation}
+\begin{aligned}
+S' &= \sum_k \lambda_k I_k \\
+I_i &= - \lambda_i I_i \\
+\end{aligned}
+\end{equation}
+
+If an infected individual in a partnership dies, the partnership dissolves.
+Thus, an $SI_i$ partnership dissolves at a rate $-\lambda_i$, and the susceptible partner enters the single susceptible compartment at rate $\lambda_i SI_i$ (due to the SIS formulation the the infected partner that dies also gives rise, at an equal rate, to an individual entering the single susceptible compartment):
+
+\begin{equation}
+\begin{aligned}
+S' &= \sum_k 2 \lambda_k SI_k \\
+SI_i &= - \lambda_i SI_i 
+\end{aligned}
+\end{equation}
+
+Similarly, $II_{ij}$ partnerships dissolve at a rate $-(\lambda_i + \lambda_j)$, but two cases, when $i \neq j$ and $i = j$, must be considered separately.
+When an $II_{ij}$ partnership dissolves due to disease-induced mortality, where $i \neq j$, the death of the partner with strain $i$ causes its partner to enter $I_j$ compartment at rate $\lambda_j II_{ij}$, and vice versa.
+When an $II_ii$ partnership dissolves, the death of either partner causes the other partner to enter the $I_i$ compartment at rate $\lambda_i II_{ii}$, which sums up to $2\lambda_i II_{ii}$.
+Combining these dynamics yields:
+
+\begin{equation}
+\begin{aligned}
+S' &= \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk} \\
+I_i' &=  \sum_k 2^{\delta_{ik}} \lambda_k II_{ik} \\
+II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij}
+\end{aligned}
+\end{equation}
+
+Finally, combining all these equations give us the full model, which is Model 1. We can simply take out the uncoupled and extra couple transmission term to obtain equation 2\todo{?}:
+
+\begin{equation}
+\begin{aligned}
+S' =& - \rho S + 2 c SS + \sum_k c SI_k - \sum_k P_k U_k S + \sum_k \lambda_k I_k \\
+&+ \sum_k 2 \lambda_k SI_k + \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk}\\
+I_i' =&  - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c  II_{ik} + \sum_k M_{ki} P_k U_k S- \lambda_i I_i \\
+&+ \sum_k 2^{\delta_{ik}} \lambda_k II_{ik} \\
+SS' =& \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS - \sum_i 2 P_i E_i SS \\
+SI_i' =& \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS - \sum_k P_k E_k SI_i   \\
+&- \lambda_i SI_i\\
+II_{ij}' =& \khalf \cdot \frac{\rho I_i \cdot I_j}{(S + \sum_k I_k)} - c II_{ij} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
+&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
+\end{aligned}
+\end{equation}
+
+\subsection*{Models 3 and 4: Partnership dynamics}
+
+Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$.
+Partnership dissolution rates are equal to those of model 1 and 2: $SS' = -cSS$, $SI_i' = - cSI_i$, and $II_{ij}' = - II_{ij}$.
+Once individuals leave a partnership, they enter temporary compartments and are distributed into a partnership through binomial distribution:
+
+\begin{equation}
+\begin{aligned}
+X &= 2 c SS + \sum_k c SI_k \\
+Y_i &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik} \\
+SS' &= - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}\\
+SI_i' &= - c SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
+II_{ij}' &= - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k}.
+\end{aligned}
+\end{equation}
+
+\subsection*{Model 3 and 4 - Infection}
+
+Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single state, only extra couple transmission exists:
+
+\begin{equation}
+P_i = \frac{r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k (2^\delta_{kl} II_{lk}) )}.
+\end{equation}
+Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equation that is used in model 1 and 2.
+
+\subsection*{Model 3 and 4 - Disease induced mortality}
+
+Disease induced mortality is modeled similar to model 1 and 2.
+However, as single state does not exist in model 3 and 4, individuals that has left their partnerships due to death of their partners enter temporary compartments and form partners instantly:
+
+\begin{equation}
+\begin{aligned}
+X &= \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk} \\
+Y_i &=  \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik} \\
+SS = &= \frac{X^2}{2 (X + \sum_k Y_k)}\\
+SI_i &= - \lambda_i SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
+II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + \khalf \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}
+\end{aligned}
+\end{equation}
+
+Combining all these dynamics, we have equation 3. If we remove extra-couple transmission, we have equation 4.
+
+\begin{equation}
+\begin{aligned}
+X =& 2 c SS + \sum_k c SI_k + \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk}\\
+Y_i =& c SI_i + \sum_k 2^{\delta_{ik}}  c II_{ik} + \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik} \\
+SS  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS\\
+SI_i =& - c SI_i + \frac{X Y_i}{X + \sum_k Y_k} - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS\\
+&- \sum_k P_k E_k SI_i - \lambda_i SI_i  \\
+II_{ij}& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
+&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
+\end{aligned}
+\end{equation}
+
+\subsection*{Model 5}
+
+Following \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta'$, that is derived from \cite{hollingsworth_hiv1_2008}'s approximate basic reproduction number:
+
+\begin{equation}
+\beta'_i = \frac{c \beta_i}{c + \beta_i + \lambda_i}.
+\end{equation}
+Thus, we get the following model:
+
+\begin{equation}
+\begin{aligned}
+S' & = \sum_k \lambda_k I_k - \sum_k \beta'_k S I_k\\
+I_i & = \sum_k M_{ki} \beta'_k S I_k - \lambda_i I_i
+\end{aligned}
+\end{equation}
+
+\subsection*{Model 6}
+
+Model 6 is a random mixing model. It is modeled in a same way as model 6 without the adjusted transmission rate:
+
+\begin{equation}
+\begin{aligned}
+S' & = \sum_k \lambda_k I_k - \sum_k \beta_k S I_k\\
+I_i & = \sum_k M_{ki} \beta_k S I_k - \lambda_i I_i
+\end{aligned}
+\end{equation}
+
+\subsection*{Initial distribution of infected individuals}
+
+We follow Champredon et al's result to calculate the initial distribution of infected individuals.
+For model 1 and 2, we have disease equilibrium state of $S^* = \frac{c}{c + \rho}$ and $SS^* = \frac{1-S^*}{2}$.
+We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with \Lspvl\ of $i$.
+$Y_i$ is taken from normal distribution with mean 3 and is normalized so that $\sum_i D_i = 1$.
+Then, we have the following initial distribution of each states:
+
+\begin{equation}
+\begin{aligned}
+S(0) &= (1 - \epsilon) S^* \\
+SS(0) &= (1 - \epsilon)^2 SS^*\\
+SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i\\
+I_i(0) &=  \epsilon S^* D_i\\
+II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
+\end{aligned}
+\end{equation}
+Since model 3 and 4 do not have single state, $SS^*=1$ at the disease free equilibrium and the initial distribution becomes as follows:
+
+\begin{equation}
+\begin{aligned}
+SS(0) &= (1 - \epsilon)^2 SS^*\\
+SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i\\
+II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
+\end{aligned}
+\end{equation}
+Lastly, as model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution.
+
+\begin{equation}
+\begin{aligned}
+S(0) &= 1 - \epsilon\\
+I_i(0) &=  \epsilon D_i.
+\end{aligned}
+\end{equation}
+Model 6 also shares the same distribution of initial infected individuals as model 5.
 

### 2016-05-23 figures/figures.rmd
working on figures

---
title: "Figures"
output: html_document
---

```{r pkgs,warning=FALSE,message=FALSE, echo = FALSE}
library("knitr")
opts_chunk$set(echo=FALSE,error=FALSE)
windowsFonts(Times=windowsFont("Times"))
library("ggplot2"); theme_set(theme_bw(base_size = 12, base_family = "Times"))
zero_margin <- theme(panel.margin=grid::unit(0,"lines"))
zero_x_margin <-
    theme(panel.margin.x=grid::unit(0, "lines"))
library("MASS")
## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally")
library("knitr")
## devtools::install_github("lionel-/ggstance")
library("ggstance")

```

```{r load, echo = FALSE}

load("../simdata/combine_ev_LHS4.rda")

```

```{r setup2,echo=FALSE}
orig_sum_labs <- c("peak_time","peak_vir","eq_vir","rel_peak")
new_sum_labs <- c("peak time","peak virulence","equilibrium virulence","relative peak")
fixfac2 <- function(x) factor(x,
                       levels=orig_sum_labs,
                       labels=new_sum_labs)
```

### Figure 1

```{r fig1,fig.height=3.5,fig.width=10, echo = FALSE, cache = TRUE,dpi = 600}

for (i in c("","2","3")) 
    load(sprintf("../simdata/ev_LHS_resS%s.rda",i))
Ls <- load("../simdata/shirreff_data.rda")
tvec <- seq(from=1,by = 0.1,length.out=nrow(vir_matS))
dimnames(vir_matS) <- list(Time=tvec,
                           run=paste0("r=",c(0.021,0.042,0.084)))

d_littleR_m <- melt(vir_matS,value.name="Mean.VL")
shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
shirreff_df_m$run <- "Shirreff"
df_tot <- rbind(shirreff_df_m,d_littleR_m)
tlab <- "time (years)"
g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
    geom_line()+labs(x=tlab,y=expression(mean~virulence~(log[10]~SPVL)))+
    theme(legend.position=c(0.75,0.25))
###
dimnames(vir_matS2) <- list(Time=tvec,
               run=c("10^{-3}", "10^{-4}", "10^{-5}"))
g2 <- g1 %+% melt(vir_matS2,id.var="Time",value.name="Mean.VL") +
	labs(y = "")
###
dimnames(vir_matS3) <- list(Time=tvec,
                            run=c("2.5", "3", "3.5"))

vir_matM3 <- melt(vir_matS3,id.var="Time",value.name="Mean.VL")
vir_matM3$run = factor(vir_matM3$run, levels = c("2.5", "3", "3.5"))

g3 <- g1 %+% vir_matM3 +
	labs(y = "")
limits <- c(2.5,4.7)
breaks <- seq(limits[1],limits[2], by = 0.5)
afun <- function(label,size=8) {
    annotate(geom="text",label=label,x=-Inf,y=Inf,
             size=size,vjust=0.98,hjust=0.02)
}
g1.y <- g1 + scale_y_continuous(limits = limits, breaks = breaks)+
    scale_colour_discrete(name="run")+afun("a")

g2.labs <- list(bquote(10^{-3}), bquote(10^{-4}), bquote(10^{-5}))

g2.y <- g2 + scale_y_continuous(limits = limits, breaks = breaks)+
    scale_colour_discrete(name="I(0)", labels = g2.labs)+afun("b")
g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
        scale_colour_discrete(name=bquote(alpha(0)))+afun("c")
grid.arrange(g1.y, g2.y, g3.y, nrow=1)
```

### Figure 2

```{r fig2, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400}

ss <- subset(sum_list[["vir_mat"]],tvec<750)

ss$model <- factor(ss$model, c("random","pairform+epc", "pairform","instswitch+epc","instswitch","implicit"))

gg_virtraj <- ggplot(ss,
                     aes(tvec,mean,ymin=lwr,ymax=upr))+
	geom_line(aes(colour=model,linetype=model),lwd=1)+
	geom_ribbon(aes(fill = model), alpha=0.3)+
	labs(x="time (years)",y=expression(mean~virulence~(log[10]~SPVL))) +
	theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
		axis.title.x = element_text(margin = margin(10,0,0,0))) +
	facet_wrap(~model, nrow = 2) +
	zero_margin

## scale_x_continuous(limits=c(0,2500))
print(gg_virtraj)

```

### Figure 3

```{r fig3, fig.width=8,fig.height=4.8, echo = FALSE, cache = TRUE,dpi = 600}

sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
mL <- melt(sL,id.vars=c("model","run"))
## horrible hack (but doesn't help); subsample
## w <- which(mL$model=="random")
## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
w <- with(mL,which(model=="random" & variable=="eq_vir"))
rval <- mean(mL$value[w])
mLw <- droplevels(mL[-w,])
mLw$variable <- fixfac2(mLw$variable)

gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
	geom_violinh(width=1)+
	facet_wrap(~variable,scale="free_x")+
	guides(fill=guide_legend(reverse=TRUE))+
	labs(y="")+
	geom_point(data=data.frame(model="random",variable="equilibrium virulence",
                               value=rval),pch=22,size=3,show.legend=FALSE) +
	zero_x_margin
print(gg_univ)

```

```{r fig4_func, echo = FALSE}

trim_gg <- function(gg) {
    n <- gg$nrow
    gg$nrow <- gg$ncol <- n-1
    v <- 1:n^2
    gg$plots <- gg$plots[v>n & v%%n!=0]
    gg$xAxisLabels <- gg$xAxisLabels[-n]
    gg$yAxisLabels <- gg$yAxisLabels[-1]
    return(gg)
}

tweak_legends_gg <- function(gg,pos=1.5) {
    n <- gg$nrow
    for (i in 1:n) {
        for (j in 1:n) {
            inner <- getPlot(gg,i,j)
            if (i==1 & j==1) {
                inner <- inner + ggplot2::theme(legend.position=c(pos,0.5)) +
                    guides(colour = guide_legend(override.aes = list(size=6)))
            } else {
                inner <- inner + ggplot2::theme(legend.position="none")
            }
            gg <- putPlot(gg,inner,i,j)
        }
    }
    return(gg)
}
tweak_colours_gg <- function(gg) {
    n <- gg$nrow
    for (i in 1:n) {
        for (j in 1:n) {
            inner <- getPlot(gg,i,j)
            inner <- inner+scale_colour_brewer(palette = 'Dark2')
            gg <- putPlot(gg,inner,i,j)
        }
    }
    return(gg)
}

```

### Figure 4

```{r fig4,fig.width=7,fig.height=7, echo = FALSE, cache = TRUE,dpi = 600}
sL2 <- sL
names(sL2)[na.omit(match(orig_sum_labs,names(sL)))] <- gsub(" ","_",new_sum_labs)
    ## paste0("`",new_sum_labs,"`")
ggp1 <- ggpairs(sL2,
        mapping = ggplot2::aes(color = model),
        columns=3:6,
        legends=TRUE,
        lower = list(continuous = wrap("points",size=0.1)), ## alpha = 0.3,size=0.5)),
        diag = list(continuous = "blankDiag"),
        upper = list(continuous = "blank"))
## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot
## have to change plot one panel at a time
ggp2 <- tweak_colours_gg(tweak_legends_gg(trim_gg(ggp1)))
## add 1-to-1 line
inner <- getPlot(ggp2,2,1)
inner <- inner+geom_abline(intercept=0,slope=1,lty=2)
ggp2 <- putPlot(ggp2,inner,2,1)
## tweak scale
inner <- getPlot(ggp2,3,1)
inner <- inner+scale_x_continuous(breaks=seq(2.5,4.5,by=0.5))
ggp2 <- putPlot(ggp2,inner,3,1)
print(ggp2)

```

### Figure 5

```{r fig5, echo = FALSE, fig.width = 10, cache = TRUE,dpi = 600}
mL3 <- subset(mL2,
       !((model=="implicit" &
          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
         (model=="random" &
          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
         (model=="instswitch" &
          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
         (model=="instswitch+epc" &
          LHSvar %in% c("rho_base","c_u_ratio")) |
         (model=="pairform" &
          LHSvar %in% c("c_u_ratio","c_e_ratio"))))
## reorder factors
mL3$LHSvar <- factor(mL3$LHSvar,
                     levels=c("Beta1","Beta3",
                              "Duration1","Duration3",
                              "c_mean_base","c_e_ratio","rho_base",
                              "c_u_ratio"),
										 labels = c("beta[P]", "beta[D]", "D[P]", "D[D]", "c", "c[e]/c[w]", "rho", "c[u]/c[w]"))
## FIXME: could also use labels argument to rename levels, labeller=label_parsed
## in facet_grid to get pretty math

mL3$model <- factor(mL3$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))

ggplot(mL3,aes(LHSval,sumval,colour=model))+
    geom_point(pch=".",alpha=0.5)+
    facet_grid(sumvar~LHSvar,scales="free", labeller = label_parsed)+
    geom_smooth(se=FALSE)+labs(x="",y="")+
    scale_x_log10(expand=c(0,0.08))+  ## leave a little extra room?
    zero_margin

```

### 2016-05-23 ms/plos_latex.tex
add author summary; tweaks

@@
 %% Include all macros below
-\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
 \newcommand{\khalf}{\left(\frac{1}{2}\right)^{\delta_{ij}}}  % (1/2)^kronecker
@@
 }
+%% short title: Epidemiological Structure and HIV Virulence Evolution
+%% (53/70 characters)
 \newline
@@
 \section*{Author Summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+
+%% 162 words
+
+Pathogens such as HIV can evolve rapidly in response to changing
+conditions; such changes include both the increases in disease
+incidence over the course of the epidemic and treatment
+interventions. While researchers have successfully used computational
+models to explore these evolutionary dynamics, these models often
+neglect epidemiological processes, such as the formation and
+dissolution of sexual partnerships; other research has shown that
+these processes can strongly affect epidemic outcomes. We built and
+compared models with a range of realism and complexity in the way they
+modeled both partnership dynamics and the possibility of sexual
+contact outside of partnerships. In contrast to models of intermediate
+complexity, simple random-mixing models with no partnership structure
+at all actually gave the closest match to the most realistic and
+complex models; extra-pair contact tended to wash out the effects
+of epidemiological structure. Nevertheless, the large differences
+in evolutionary dynamics among different epidemiological models
+suggests that researchers trying to predict the evolution of
+pathogens should proceed with caution.
 
@@
 of infection are independent of the host's SPVL.
-Following Shirreff \textit{et al.}, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the \textbf{A}symptomatic stage are Hill functions of the SPVL:
+Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the \textbf{A}symptomatic stage are Hill functions of the SPVL:
 
@@
 
-Our first four models consider explicit partnership dynamics and are based on Champredon text{et al.}'s model \cite{champredon_hiv_2013}.
+Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}.
 Models 1 and 2 assume non-instantaneous partnership formation (i.e.
@@
 The base model (i.e.
-model 1) for the first four models is an extension of Champredon text{et al.}'s model \cite{champredon_hiv_2013}.
+model 1) for the first four models is an extension of Champredon \etal's model \cite{champredon_hiv_2013}.
 Individuals in single compartment acquire a partner at a rate, $\rho$, and partnerships dissolve at a rate, $c$.
@@
 Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
-Extra-couple and uncoupled transmission are modeled in a same way as Champredon text{et al.}'s model.
+Extra-couple and uncoupled transmission are modeled in a same way as Champredon \etal's model.
 All the details have been adapted to a multi-strain scenario.
@@
 After each run, initial exponential growth rate is calculated.
-Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff et al's model.
+Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff \etal's model.
 
@@
 We first consider the models of intermediate realism (implicit,
-instantaneous-switching with and without extra-pair contact, and
+instantaneous-switching with and without extra-pair contact (``epc'' hereafter), and
 pair formation without extra-pair contact). Some parameter
@@
 load are changing, on time scales comparable to those portrayed here
-(e.g., compare Fig.~\ref{fig:virtraj} to Herbeck \text{et al}'s
+(e.g., compare Fig.~\ref{fig:virtraj} to Herbeck \etal's
 estimated rate of change of 1.3 \Lspvl\ per century [95\% CI -0.1 to
@@
 could significantly change their conclusions.
-
-\section*{Conclusion}
-
-CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.
-
-Sed non aliquet felis.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris commodo justo ac dui pretium imperdiet.
-Sed suscipit iaculis mi at feugiat.
-Ut neque ipsum, luctus id lacus ut, laoreet scelerisque urna.
-Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-Nam condimentum sem eget mollis euismod.
-Nullam dui urna, gravida venenatis dui et, tincidunt sodales ex.
-Nunc est dui, sodales sed mauris nec, auctor sagittis leo.
-Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor.
-For more information, see \nameref{S1_Appendix}.
 
@@
 Single individuals acquire partners at rate $\rho$: $S' = - \rho S$ and $I_i' = - \rho I_i$.
-We follow Champredon et al's results and assume that single individuals are distributed into coupled states through binomial distribution\todo{?
-don't understand this wording (ditto below)}:
-\todo{I think we need different notation for partnership formation (i.e. LHS is wrong)}
-
-\begin{equation}
-\begin{aligned}
-(SS)' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)}\\
-(SI_i)' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
-(II_{ij})' &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
+We follow Champredon \etal's results and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
+
+\newcommand{\PF}{\textrm{PF}}
+\begin{equation}
+\begin{aligned}
+\PF(SS) &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)}\\
+\PF(SI_i) &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
+\PF(II_{ij}) &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
 \end{aligned}
@@
 
-Partnerships dissolve at rate $c$: $SS' = -c SS$, $SI_i' = -c SI_i$, and $II_{ij} = - c II_{ij}$.
+Partnerships dissolve at per-partnership rate $c$: the dissolution rates for $SS$, $SI_i$, and $II_{ij}$ pairs are $c SS$, $c SI_i$, and $c II_{ij}$ respectively.
 Unlike single strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains which the individuals in concordant partnership are infected with (i.e.
-both partners in $II_{ii}$ enter $I_i$ whereas only one partner in $II_{ij}$ enters $I_i$).
+both partners in $II_{ii}$ enter $I_i$ whereas one partner in $II_{ij}$ enters the $I_i$ compartment while the other enters $I_j$).
 
@@
 \end{equation}
-Combining partnership formation and dissolution process yields the following equation:
+
+Combining the partnership formation and dissolution processes yields the following equation:
 
@@
 On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $II_{ii}' = M_{ii} \beta_i SI_i$.
-Using the Kronecker delta notation, we obtain following set of equations that describe within-couple transmission dynamics:
+Using the Kronecker delta notation, we obtain the following set of equations that describes within-couple transmission dynamics:
 
@@
 
-Champredon et al define the proportion of infectious extra-couple and uncoupled contact through the following term:
+Champredon \etal\ define the proportion of infectious extra-couple and uncoupled contact through the following term:
 
@@
 \end{equation}
-Effective uncoupled, $c_u$, and extra couple, $c_e$, contact rate is the product of two terms: uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
+The effective uncoupled, $c_u$, and extra couple, $c_e$, contact rates are the product of two terms: uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
 Therefore, the transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above.
@@
 This simplification is useful in a multi-strain model since we cannot multiply a vector with a single value (e.g.
-$c_u S$ in denominator) if we use Champredon et al's equation in its original form.
+$c_u S$ in denominator) if we use Champredon \etal's equation in its original form.
 Extending the above equation to multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain:
@@
 
-Finally, combining all these equations give us the full model, which is Model 1. We can simply take out the uncoupled and extra couple transmission term to obtain equation 2\todo{?}:
+Finally, combining all these equations give us the full model, which is Model 1. We can simply drop the uncoupled and extra-couple transmission terms to obtain equation 2:
 
@@
 Partnership dissolution rates are equal to those of model 1 and 2: $SS' = -cSS$, $SI_i' = - cSI_i$, and $II_{ij}' = - II_{ij}$.
-Once individuals leave a partnership, they enter temporary compartments and are distributed into a partnership through binomial distribution:
+Once individuals leave a partnership, they enter temporary compartments and are distributed into a partnership as follows:
 
@@
 
-We follow Champredon et al's result to calculate the initial distribution of infected individuals.
+We follow Champredon \etal's result to calculate the initial distribution of infected individuals.
 For model 1 and 2, we have disease equilibrium state of $S^* = \frac{c}{c + \rho}$ and $SS^* = \frac{1-S^*}{2}$.

### 2016-05-23 figures/figures.rmd
lots of figure tweaking

@@
 opts_chunk$set(echo=FALSE,error=FALSE)
+if (.Platform$OS.type=="windows") {
     windowsFonts(Times=windowsFont("Times"))
+} else {
+    theme_update(text=element_text(family="Times"))
+}
 library("ggplot2"); theme_set(theme_bw(base_size = 12, base_family = "Times"))
@@
 new_sum_labs <- c("peak time","peak virulence","equilibrium virulence","relative peak")
-fixfac2 <- function(x) factor(x,
+fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
+    if (atop) new_sum_labs <-
+                  gsub("(.*) (.*)","atop(\\1,\\2)",new_sum_labs)
+    factor(x,
            levels=orig_sum_labs,
            labels=new_sum_labs)
+}
 ```
@@
 
-```{r fig1,fig.height=3.5,fig.width=10, echo = FALSE, cache = TRUE,dpi = 600}
-
+```{r fig1,fig.height=3.5,fig.width=10, echo = FALSE, dpi = 600}
 for (i in c("","2","3")) 
@@
 
-trim_gg <- function(gg) {
+trim_gg <- function(gg,hack_spaces=TRUE) {
     n <- gg$nrow
@@
     gg$yAxisLabels <- gg$yAxisLabels[-1]
+    ## ggpairs tries to parse labels, so this fails if they
+    ## have spaces in them ... maybe there's a better way?
+    if (hack_spaces) {
+        gg$xAxisLabels <- gsub("_"," ",gg$xAxisLabels)
+        gg$yAxisLabels <- gsub("_"," ",gg$yAxisLabels)
+    }
     return(gg)
@@
                               "c_u_ratio"),
-										 labels = c("beta[P]", "beta[D]", "D[P]", "D[D]", "c", "c[e]/c[w]", "rho", "c[u]/c[w]"))
+                     labels = c("beta[P]", "beta[D]", "D[P]", "D[D]",
+                                "c", "c[e]/c[w]", "rho", "c[u]/c[w]"))
+
 ## FIXME: could also use labels argument to rename levels, labeller=label_parsed
@@
 
+mL3$sumvar <- fixfac2(mL3$sumvar)
+
+L <- function(labels,multi_line=TRUE) {
+    r <- if (all(grepl(" ",labels[[1]]))) {
+             list(as.character(gsub(" ","\n",labels[[1]])))
+    } else {
+        label_parsed(labels,multi_line=multi_line)
+    }
+    return(r)
+}
+
+brkfun <- function(x) {
+    nx <- 1
+    n <- 2
+    brks <- axisTicks(log(x), log=TRUE,  nint = n)
+    ## browser()
+    brks <- brks[brks>x[1] & brks<x[2]]
+    nx <- length(brks)
+    ## cat("brk1",brks,"\n")
+    while (nx<2 && n<10) {
+        n <- n+1
+        ## cat(n,"\n")
+        brks <- axisTicks(log10(x), log=TRUE,  nint = n)
+        brks <- brks[brks>x[1] & brks<x[2]]
+        ## cat("brk2",brks,"\n")
+        nx <- length(brks)
+    }
+    if (n==10) browser()
+    r <- brks[c(1,length(brks))]
+    ## hacks
+    if (all(r==c(0.01,1))) r <- c(0.02,0.5)
+    if (all(r==c(0.2,5))) r <- c(0.5,5)
+    return(r)
+}
+
 ggplot(mL3,aes(LHSval,sumval,colour=model))+
     geom_point(pch=".",alpha=0.5)+
-    facet_grid(sumvar~LHSvar,scales="free", labeller = label_parsed)+
+    facet_grid(sumvar~LHSvar,scales="free",
+               labeller = L)+
     geom_smooth(se=FALSE)+labs(x="",y="")+
-    scale_x_log10(expand=c(0,0.08))+  ## leave a little extra room?
+    ## leave a little extra room?
+    scale_x_log10(expand=c(0,0.08),breaks=brkfun)+
     zero_margin
-
 ```

### 2016-05-23 figures/figures.rmd
more figure tweaks

@@
 library("knitr")
+library("ggplot2"); theme_set(theme_bw(base_size = 12,
+                                       base_family = "Times"))
 opts_chunk$set(echo=FALSE,error=FALSE)
@@
     windowsFonts(Times=windowsFont("Times"))
-} else {
-    theme_update(text=element_text(family="Times"))
 } 
-library("ggplot2"); theme_set(theme_bw(base_size = 12, base_family = "Times"))
 zero_margin <- theme(panel.margin=grid::unit(0,"lines"))
@@
 library("reshape2")  ## for melt()
-library("GGally")
-library("knitr")
+library("GGally") ## need BMB version
+## devtools::install_github("bbolker/GGally")
 ## devtools::install_github("lionel-/ggstance")
 library("ggstance")
-
 ```
@@
 ```{r load, echo = FALSE}
-
 load("../simdata/combine_ev_LHS4.rda")
-
 ```
@@
 afun <- function(label,size=8) {
-    annotate(geom="text",label=label,x=-Inf,y=Inf,
-             size=size,vjust=0.98,hjust=0.02)
+    annotate(geom="text",label=label,x=10,y=Inf,
+             ## http://stackoverflow.com/questions/20083700/how-to-have-annotated-text-style-to-inherit-from-theme-set-options
+             family= theme_get()$text[["family"]],
+             size=size,
+             vjust=1.5,hjust=0.1)
+             ## vjust=0.98,hjust=0.02)
 }
@@
 ```{r fig4_func, echo = FALSE}
-
 trim_gg <- function(gg,hack_spaces=TRUE) {
@@
 ggp2 <- putPlot(ggp2,inner,3,1)
-print(ggp2)
-
+print(ggp2,spacingProportion=0)
 ```

### 2016-05-25 ms/plos_latex.tex
small tweaks

@@
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
+\newcommand{\PF}{\textrm{PF}}
+\newcommand{\DS}{\textrm{DS}}
+\newcommand{\WT}{\textrm{WT}}
+\newcommand{\ET}{\textrm{ET}}
+\newcommand{\DM}{\textrm{DM}}
 
@@
 \begin{equation}
-\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D}
+\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
 \end{equation}
@@
 \begin{aligned}
-\beta_i &= \beta(\alpha_i)\\
-\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}
+\beta_i &= \beta(\alpha_i),\\
+\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
 \end{aligned}
@@
 
-Single individuals acquire partners at rate $\rho$: $S' = - \rho S$ and $I_i' = - \rho I_i$.
+Single individuals acquire partners at per-person rate $\rho$.
+Partnership formation rate for $S$, $I_i$ are $- \rho S$ and $-\rho I_i$, respectively.
 We follow Champredon \etal's results and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
 
-\newcommand{\PF}{\textrm{PF}}
-\begin{equation}
-\begin{aligned}
-\PF(SS) &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)}\\
-\PF(SI_i) &= \frac{\rho S \cdot I_i}{S + \sum_k I_k}\\
-\PF(II_{ij}) &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}
-\end{aligned}
-\end{equation}
-In a single-strain model, both individuals in a concordant partnership, $II$, enter the single-infected compartment, $I$, when a partnership dissolves: $I' = 2 c II'$.
-However, this is not the case in a multi-strain model.
-If infected individuals in an $II$ partnership have different strains, they will enter different compartments; if they are both infected with the same strain, they both enter the same compartment.
-As described above, we use the Kronecker delta notation to implement this difference between the $i=j$ and $i\neq j$ cases.
+\begin{equation}
+\begin{aligned}
+\PF(SS) &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)},\\
+\PF(SI_i) &= \frac{\rho S \cdot I_i}{S + \sum_k I_k},\\
+\PF(II_{ij}) &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}.
+\end{aligned}
+\end{equation}
 
@@
 both partners in $II_{ii}$ enter $I_i$ whereas one partner in $II_{ij}$ enters the $I_i$ compartment while the other enters $I_j$).
-
-\begin{equation}
-\begin{aligned}
-S' &= 2 c SS + \sum_k c SI_k \\
-I_i' &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}\\
+Thus, coupled individuals are distributed into single states through partnership dissolution (DS):
+
+\begin{equation}
+\begin{aligned}
+\DS(S) &= 2 c SS + \sum_k c SI_k, \\
+\DS(I_i) &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}.\\
 \end{aligned}
@@
 
-Within-couple transmission occurs in both models.
-An infected partner in $SI$ partnership transmits virus to a susceptible partner, and partnership state becomes $II$: $SI_i' = - \beta_i SI_i$.
-Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $II_{ij}' = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$.
-On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $II_{ii}' = M_{ii} \beta_i SI_i$.
+Within-couple transmission (WT) occurs in both models.
+An infected partner in $SI$ partnership transmits virus to a susceptible partner, and partnership state becomes $II$: $\WT(SI_i) = - \beta_i SI_i$.
+Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$.
+On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $\WT(II_{ii}) = M_{ii} \beta_i SI_i$.
 Using the Kronecker delta notation, we obtain the following set of equations that describes within-couple transmission dynamics:
@@
 \begin{aligned}
-SI_i' &= - \beta_i SI_i\\
-II_{ij} &=  \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)
+\WT(SI_i) &= - \beta_i SI_i,\\
+\WT(II_{ij}) &=  \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji}. \beta_j SI_j)
 \end{aligned}
@@
 \begin{equation}
-P_i = \frac{r_u I_i + r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_u (S + \sum_k I_k) + r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k 2^\delta_{lk} II_{lk} )},
-\end{equation}
-Using the equation above, we can model uncoupled and extra-couple mixing.
+P_i = \frac{r_u I_i + r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_u (S + \sum_k I_k) + r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k 2^\delta_{lk} II_{lk} )}.
+\end{equation}
+Using the equation above, we can model extra-pair transmission (ET).
 For convenience, uncoupled and extra-couple transmission rates, $c_u$ and $c_e$, will be replaced with $U_i = r_u \beta_i$ and $E_i = r_e \beta_i$ hereafter.
@@
 Single susceptible individuals become infected and enter single infected compartment at the total rate of $\sum_k P_k U_k S$.
-Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $I_i' = \sum_k M_{ki} P_k U_k S$.
+Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $\ET(I_i) = \sum_k M_{ki} P_k U_k S$.
 Either partner in an $SS$ partnership can be infected, with the partnership state becoming $SI$, at the total rate of $\sum_i 2 P_i E_i SS$.
-The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $SI_i' = \sum_k 2 M_{ki} P_k E_k SS$.
+The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $\ET(SI_i) = \sum_k 2 M_{ki} P_k E_k SS$.
 Lastly, the susceptible partner in an $SI$ partnership can be infected due to uncoupled/extra-couple contacts and partnership can change to an $II$ partnership.
@@
 The only difference is that the \Lspvl\ of a newly infected partner is not determined by its social partner but from an extra-couple partner (i.e.
-the term $P_i$): $II_{ij}' = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$.
+the term $P_i$): $\ET(II_{ij}) = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$.
 Combining these equations we get the following set of equations that describe all the transmission dynamics:
@@
 \begin{aligned}
-S' =& - \sum_k P_k U_k S\\
-I_i' =& \sum_k M_{ki} P_k U_k S\\
-SS' =&  - \sum_i 2 P_i E_i SS \\
-SI_i' =& \sum_k 2 M_{ki} P_k E_k SS - \beta_i SI_i - \sum_k P_k E_k SI_i\\
-II_{ij}' =& \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) + \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i\\
-&+ M_{ki} P_k E_k SI_j))
+S' =& - \sum_k P_k U_k S,\\
+I_i' =& \sum_k M_{ki} P_k U_k S,\\
+SS' =&  - \sum_i 2 P_i E_i SS, \\
+SI_i' =& \sum_k 2 M_{ki} P_k E_k SS - \beta_i SI_i - \sum_k P_k E_k SI_i,\\
+II_{ij}' =& \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji}, \beta_j SI_j) + \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i\\
+&+ M_{ki} P_k E_k SI_j)).
 \end{aligned}
@@
 
-The disease induced death rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
+The disease induced mortality (DM) rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
 Since we assume an SIS formulation, where infected individuals that die from infection are immediately replaced by an individual in the single susceptible compartment, we obtain the following equation for single infected individuals:
@@
 \begin{aligned}
-S' &= \sum_k \lambda_k I_k \\
-I_i &= - \lambda_i I_i \\
+\DM(S) &= \sum_k \lambda_k I_k, \\
+\DM(I_i) &= - \lambda_i I_i. \\
 \end{aligned}
@@
 \begin{aligned}
-S' &= \sum_k 2 \lambda_k SI_k \\
-SI_i &= - \lambda_i SI_i 
+\DM(S) &= \sum_k 2 \lambda_k SI_k, \\
+\DM(SI_i) &= - \lambda_i SI_i.
 \end{aligned}
@@
 \begin{aligned}
-S' &= \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk} \\
-I_i' &=  \sum_k 2^{\delta_{ik}} \lambda_k II_{ik} \\
-II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij}
+\DM(S) &= \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk}, \\
+\DM(I_i) &=  \sum_k 2^{\delta_{ik}} \lambda_k II_{ik}, \\
+\DM(II_{ij}) &= -(\lambda_i + \lambda_j) II_{ij}.
 \end{aligned}
@@
 Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$.
-Partnership dissolution rates are equal to those of model 1 and 2: $SS' = -cSS$, $SI_i' = - cSI_i$, and $II_{ij}' = - II_{ij}$.
+Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - II_{ij}$.
 Once individuals leave a partnership, they enter temporary compartments and are distributed into a partnership as follows:
@@
 \begin{aligned}
-X &= \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk} \\
-Y_i &=  \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik} \\
-SS = &= \frac{X^2}{2 (X + \sum_k Y_k)}\\
-SI_i &= - \lambda_i SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
-II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + \khalf \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}
+X &= \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk}, \\
+Y_i &=  \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik}, \\
+SS' = &= \frac{X^2}{2 (X + \sum_k Y_k)},\\
+SI_i' &= - \lambda_i SI_i + \frac{X Y_i}{X + \sum_k Y_k},\\
+II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + \khalf \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}.
 \end{aligned}
@@
 \begin{aligned}
-X =& 2 c SS + \sum_k c SI_k + \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk}\\
-Y_i =& c SI_i + \sum_k 2^{\delta_{ik}}  c II_{ik} + \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik} \\
-SS  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS\\
+X =& 2 c SS + \sum_k c SI_k + \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk},\\
+Y_i =& c SI_i + \sum_k 2^{\delta_{ik}}  c II_{ik} + \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik}, \\
+SS  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS,\\
 SI_i =& - c SI_i + \frac{X Y_i}{X + \sum_k Y_k} - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS\\
-&- \sum_k P_k E_k SI_i - \lambda_i SI_i  \\
-II_{ij}& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
-&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
+&- \sum_k P_k E_k SI_i - \lambda_i SI_i,\\
+II_{ij}& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)\\
+&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}.
 \end{aligned}
@@
 \begin{aligned}
-S' & = \sum_k \lambda_k I_k - \sum_k \beta'_k S I_k\\
-I_i & = \sum_k M_{ki} \beta'_k S I_k - \lambda_i I_i
+S' & = \sum_k \lambda_k I_k - \sum_k \beta'_k S I_k.\\
+I_i & = \sum_k M_{ki} \beta'_k S I_k - \lambda_i I_i.
 \end{aligned}
@@
 
-Model 6 is a random mixing model. It is modeled in a same way as model 6 without the adjusted transmission rate:
-
-\begin{equation}
-\begin{aligned}
-S' & = \sum_k \lambda_k I_k - \sum_k \beta_k S I_k\\
-I_i & = \sum_k M_{ki} \beta_k S I_k - \lambda_i I_i
+Model 6 is a random mixing model. It is modeled in a same way as model 5 without the adjusted transmission rate:
+
+\begin{equation}
+\begin{aligned}
+S' & = \sum_k \lambda_k I_k - \sum_k \beta_k S I_k,\\
+I_i & = \sum_k M_{ki} \beta_k S I_k - \lambda_i I_i.
 \end{aligned}
@@
 \begin{aligned}
-S(0) &= (1 - \epsilon) S^* \\
-SS(0) &= (1 - \epsilon)^2 SS^*\\
-SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i\\
-I_i(0) &=  \epsilon S^* D_i\\
+S(0) &= (1 - \epsilon) S^*, \\
+SS(0) &= (1 - \epsilon)^2 SS^*,\\
+SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i,\\
+I_i(0) &=  \epsilon S^* D_i,\\
 II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
@@
 \begin{aligned}
-SS(0) &= (1 - \epsilon)^2 SS^*\\
-SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i\\
+SS(0) &= (1 - \epsilon)^2 SS^*,\\
+SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i,\\
 II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
@@
 \begin{aligned}
-S(0) &= 1 - \epsilon\\
+S(0) &= 1 - \epsilon,\\
 I_i(0) &=  \epsilon D_i.

### 2016-05-28 figures/figures.rmd
worked on figures

@@
 }
+
 ```
@@
 
+### Figure 2.1 - Transmission rate
+
+```{r fig2.1, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400}
+
+source("../R/hivFuns.R")
+source("../R/Param.R")
+
+returnBeta <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(K,Bmax,v,a))
+	})
+}
+
+ss2 <- transform(ss, mean = returnBeta(mean,HIVpars.skeleton),
+								 lwr = returnBeta(lwr,HIVpars.skeleton),
+								 upr = returnBeta(upr,HIVpars.skeleton))
+
+gg_betatraj <- ggplot(ss2,
+                     aes(tvec,mean,ymin=lwr,ymax=upr))+
+	geom_line(aes(colour=model,linetype=model),lwd=1)+
+	geom_ribbon(aes(fill = model), alpha=0.3)+
+	labs(x="time (years)",y=expression(mean~transmission~rate~during~asymptomatic~stage)) +
+	theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
+		axis.title.x = element_text(margin = margin(10,0,0,0))) +
+	facet_wrap(~model, nrow = 2) +
+	zero_margin
+
+## scale_x_continuous(limits=c(0,2500))
+print(gg_betatraj)
+
+```
+
+### Figure 2.2 - Duration
+
+```{r fig2.2, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400} 
+returnDur <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(v,Dmax,D50,h))
+	})
+}
+
+ss3 <- transform(ss, mean = returnDur(mean,HIVpars.skeleton),
+								 lwr = returnDur(lwr,HIVpars.skeleton),
+								 upr = returnDur(upr,HIVpars.skeleton))
+
+gg_durtraj <- ggplot(ss3,
+                     aes(tvec,mean,ymin=lwr,ymax=upr))+
+	geom_line(aes(colour=model,linetype=model),lwd=1)+
+	geom_ribbon(aes(fill = model), alpha=0.3)+
+	labs(x="time (years)",y=expression(mean~duration~of~asymptomatic~stage)) +
+	theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
+		axis.title.x = element_text(margin = margin(10,0,0,0))) +
+	facet_wrap(~model, nrow = 2) +
+	zero_margin
+
+## scale_x_continuous(limits=c(0,2500))
+print(gg_durtraj)
+
+```
+
 ### Figure 3
@@
 print(gg_univ)
+
+```
+
+### Figure 3.1
+
+```{r fig3.1, fig.width=8,fig.height=4.8, echo = FALSE, cache = TRUE,dpi = 600}
+
+sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
+sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
+
+sL.epi <- transform(sL, eq_t = returnBeta(eq_vir,HIVpars.skeleton),
+										peak_t = returnBeta(peak_vir,HIVpars.skeleton),
+										eq_dur = returnDur(eq_vir,HIVpars.skeleton),
+										peak_dur = returnDur(peak_vir,HIVpars.skeleton))
+
+sL.epi <- sL.epi[,-c(3:6)]
+
+mL.epi <- melt(sL.epi,id.vars=c("model","run"))
+## horrible hack (but doesn't help); subsample
+## w <- which(mL$model=="random")
+## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
+w.epi.t <- with(mL.epi,which(model=="random" & variable=="eq_t"))
+w.epi.d <- with(mL.epi,which(model=="random" & variable=="eq_dur"))
+rval.epi.t <- mean(mL.epi$value[w.epi.t])
+rval.epi.d <- mean(mL.epi$value[w.epi.d])
+mLw.epi <- droplevels(mL.epi[-c(w.epi.t,w.epi.d),])
+mLw.epi$variable <- factor(mLw.epi$variable, levels = c("eq_t", "peak_t", "eq_dur", "peak_dur"),
+													 labels = c("Equilbrium asymptomatic transmission rate",
+													 					 "Maximum transmission rate",
+													 					 "Equilibrium duration of asymptomatic stage",
+													 					 "Minimum duration of asymptomatic stage"))
+
+gg_univ.epi <- ggplot(mLw.epi,aes(value,model,fill=model))+
+	geom_violinh(width=1)+
+	facet_wrap(~variable,scale="free_x")+
+	guides(fill=guide_legend(reverse=TRUE))+
+	labs(y="")+
+	geom_point(data=data.frame(model="random",variable=c("Equilbrium asymptomatic transmission rate",
+																											 "Equilibrium duration of asymptomatic stage"),
+                               value=c(rval.epi.t, rval.epi.d)),pch=22,size=3,show.legend=FALSE) +
+	zero_x_margin
+print(gg_univ.epi)
 

### 2016-05-28 R/LHSparms2.R

@@ -1,45 +1,0 @@
-source("hivFuns.R")
-n.trial <- 1000  ## total number of sims
-
-## baseline parameters, not including parameters that will vary
-## in the LHS process
-HIVpars.skeleton <- list(
-  alphaDist=c(min=2,max=7,delta=0.5),
-  scale = 1,
-  scale_c = 1,
-  scale_all = 1,
-  Bmax = 0.317,
-  ## Beta1 = 2.76,
-  ## Beta3 = 0.76,
-  ## Duration1 = 1/4,
-  ## Duration3 = 0.75,
-  K = 13938,
-  a = 1.02,
-  Dmax = 25.4,
-  h = 0.41,
-  D50 = 3058,
-  Vm = 0.12,
-  inisd = 0.2,
-  Ini_I = 1e-4
-)
-
-## parameter ranges
-## also see
-## https://twitter.com/noamross/status/665212230581948417
-HIVpars_range <-
-    data.frame(min=c(1.31,1.23/12,0.413,4.81/12,1/10,1/15,1/5,0.01),
-               max=c(5.09,6/12,1.28,14/12,2/5,1/5,5,1),
-               row.names=c("Beta1","Duration1","Beta3","Duration3",
-               "rho_base","c_mean_base","c_u_ratio","c_e_ratio"))
-
-## full (unscrambled) sequences
-ltab <- as.data.frame(apply(HIVpars_range,1,
-                            function(x) exp(seq(log(x[1]),log(x[2]),
-                                                length=n.trial))))
-
-## create LHS sample
-set.seed(101)
-ltab[] <- lapply(ltab,sample)
-
-ltab_mean <- as.HIVvirparlist(apply(HIVpars_range, 1, geom_mean))
-

### 2016-05-28 R/LHSparms3.R

@@ -1,46 +1,0 @@
-n.trial <- 1000  ## total number of sims
-
-## FIXME: describe all of these!
-HIVpars.skeleton <- list(
-  alphaDist=c(min=2,max=7,delta=0.25),
-  scale = 1,
-  scale_c = 1,
-  scale_all = 1,
-  Bmax = 0.317,
-  K = 13938,
-  a = 1.02,
-  Dmax = 25.4,
-  h = 0.41,
-  D50 = 3058,
-  Vm = 0.12,
-  inisd = 0.2,
-  Ini_I = 1e-4,
-  ini_V = 3.5
-)
-
-
-## FIXME: document all of these!
-HIVpars_range <- data.frame(min=c(1.31,1.23/12,0.413,
-                                  4.81/12,1/10,1/15,1/5,0.01),
-                            max=c(5.09,6/12,1.28,14/12,2/5,1/5,5,1),
-                            row.names=c("Beta1","Duration1",
-                                        "Beta3","Duration3",
-                                        "rho_base","c_mean_base",
-                                        "c_u_ratio","c_e_ratio"))
-
-
-ltab <- as.data.frame(apply(HIVpars_range,1,
-                            function(x) exp(seq(log(x[1]),log(x[2]),
-                                                length=n.trial))))
-
-set.seed(101)
-ltab[] <- lapply(ltab,sample)
-
-as.HIVvirparlist <- function(x) {
-  res <- append(x,HIVpars.skeleton)
-  class(res) <- c("list","parlist","HIVvirparlist")
-  return(res)
-}
-
-ltab_mean <- as.HIVvirparlist(apply(HIVpars_range, 1, geom_mean))
-

### 2016-05-28 R/hivModels.R

@@ -21,20 +21,7 @@ gfun <- function(parameters, experimental=FALSE) {
   pp <- expand(parameters)
   attach(pp)
   
-  p <- matrix(NA, n.alpha, n.alpha)
-  
-  
-  for(i in 1:n.alpha){
-    x = alphaDist[["min"]] + (i-1)*d.alpha
-    
-    for(j in 1:n.alpha){
-      k = alphaDist[["min"]] + (j-1)*d.alpha
-      denom <- with(as.list(alphaDist),
-                    diff(pnorm(c(min-d.alpha/2,max+d.alpha/2), mean=x, sd=Vm)))
-      p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
-    }
-  }
-  
+  p <- makeMutMat(pp)
   
   detach(pp)
   
@@ -165,20 +152,7 @@ gfun2 <- function(parameters) {
   pp <- expand(parameters)
   attach(pp)
   
-  p <- matrix(NA, n.alpha, n.alpha)
-  
-  
-  for(i in 1:n.alpha){
-    x = alphaDist[["min"]] + (i-1)*d.alpha
-    
-    for(j in 1:n.alpha){
-      k = alphaDist[["min"]] + (j-1)*d.alpha
-      denom <- with(as.list(alphaDist),
-                    diff(pnorm(c(min-d.alpha/2,max+d.alpha/2), mean=x, sd=Vm)))
-      p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
-    }
-  }
-  
+  p <- makeMutMat(pp)
   
   detach(pp)
   
@@ -224,19 +198,7 @@ gfun3 <- function(parameters) {
   pp <- expand(parameters)
   attach(pp)
   
-  p <- matrix(NA, n.alpha, n.alpha)
-  
-  for(i in 1:n.alpha){
-    x = alphaDist[["min"]] + (i-1)*d.alpha
-    
-    for(j in 1:n.alpha){
-      k = alphaDist[["min"]] + (j-1)*d.alpha
-      denom <- with(as.list(alphaDist),
-                    diff(pnorm(c(min-d.alpha/2,max+d.alpha/2), mean=x, sd=Vm)))
-      p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
-    }
-  }
-  
[…]

### 2016-05-28 R/SSQ.R
SSQ

@@ -1,0 +1,26 @@
+library("lme4")
+load("../simdata/combine_ev_LHS4.rda")
+sL <- sum_list[["sum_mat"]]
+
+model = sL$model
+eqV = sL$eq_vir
+peakV = sL$peak_vir
+
+lmer(eqV ~ 1 + (1|model), data = sL, REML=FALSE)
+lmer(peakV ~ 1 + (1|model), data = sL, REML=FALSE)
+
+findSSQ <- function(var){
+	wSSQ = sum((var-ave(var, model))^2) ##within
+	group_mean = tapply(var,list(model),mean)
+	bSSQ = sum(((group_mean)-mean(var))^2)*1000 ##between
+	total = wSSQ + bSSQ
+	wR = wSSQ/total
+	bR = bSSQ/total
+	cat("within:", wR,"\n")
+	cat("between:", bR,"\n")
+}
+
+findSSQ(eqV)
+findSSQ(peakV)
+
+##why is this different from the lmer results?

### 2016-05-29 poster/hivlhs_eeid2016.Rnw
poster stuff

%\documentclass[final,hyperref={pdfpagelabels=false}]{beamer}
\documentclass{beamer}
\usepackage{grffile}
\mode<presentation>{\usetheme{drexel}}

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{amsmath,amsthm, amssymb, latexsym}
\usefonttheme[onlymath]{serif}

%% You can change the poster orientation (portrait vs. landscape) and the paper size (a0, a1, &c.) here:
\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}

\usepackage{array,booktabs,tabularx}
\newcolumntype{Z}{>{\centering\arraybackslash}X} % centered tabularx columns
\newcommand{\pphantom}{\textcolor{ta3aluminium}} % phantom introduces a vertical space in p formatted table columns??!!

%% MY MACROS %%
\newcommand{\R}{{\mathcal R}}
\newcommand{\Ro}{\mathcal{R}_0}
\newcommand{\Ihat}{{\hat{I}}}
\newcommand{\Imin}{I_{\rm min}}
\newcommand{\etal}{\emph{et al.}}
\newcommand{\fref}[1]{Figure~\ref{fig:#1}}

<<pkgs,warning=FALSE,message=FALSE, echo = FALSE>>=
library("knitr")
library("ggplot2"); theme_set(theme_bw(base_size = 32,
                                       base_family = "Times"))
opts_chunk$set(echo=FALSE,error=FALSE)
if (.Platform$OS.type=="windows") {
    windowsFonts(Times=windowsFont("Times"))
} 
zero_margin <- theme(panel.margin=grid::unit(0,"lines"))
zero_x_margin <-
    theme(panel.margin.x=grid::unit(0, "lines"))
library("MASS")
## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally") ## need BMB version
## devtools::install_github("bbolker/GGally")
## devtools::install_github("lionel-/ggstance")
library("ggstance")
@

<<load, echo = FALSE>>=
load("../simdata/combine_ev_LHS4.rda")
@

<<setup2,echo=FALSE>>=
orig_sum_labs <- c("peak_time","peak_vir","eq_vir","rel_peak")
new_sum_labs <- c("peak time","peak virulence","equilibrium virulence","relative peak")
fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
    if (atop) new_sum_labs <-
                  gsub("(.*) (.*)","atop(\\1,\\2)",new_sum_labs)
    factor(x,
           levels=orig_sum_labs,
           labels=new_sum_labs)
}
@

<<setup3, echo=FALSE,dev="tikz",cache=TRUE>>=
source("../R/hivFuns.R")
source("../R/LHSparms3.R")
Lspvl <- "$\\log_{10}~\\textrm{SPVL}$"
pars <- expand(ltab_mean)
dd <-with(pars,
     melt(data.frame(alpha=alpha,"transmission probability"=Beta2,
                     "duration (years)"=Duration2,check.names=FALSE),
          id.vars="alpha"))
gg_tradeoff <- ggplot(dd,aes(alpha,value))+facet_wrap(~variable,scale="free")+
    geom_line()+
    labs(x= Lspvl)
@

%%%\title{\huge On the Persistence of Invading Pathogens}
\title{\LARGE HIV virulence evolution in structured epidemic models}

\author{\href{http://www.math.mcmaster.ca/bolker/}{Ben Bolker and Sang Woo Park}}
\email{\href{mailto:bolker@mcmaster.ca}{\texttt{bolker@mcmaster.ca}}}
\institute[McMaster University]{McMaster University, Hamilton, Ontario, Canada %%\\\vspace{5mm}\texttt{earn@math.mcmaster.ca}
}
\date[June 2016]{June 2016}

%% Uncomment the following line to define a custom logo (or to tweak the size of the logo)
\DrexelLogo{\includegraphics[width=\WhenPortrait{1.75}{1}\linewidth]{images/Maclogo}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newlength{\columnheight}
\setlength{\columnheight}{105cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{frame}
%\includegraphics[width=\textwidth]{images/photobanner.pdf}
\vspace{0.5cm}
  \noindent\begin{columns}
    % ---------------------------------------------------------%
    % Set up a column 
    \begin{column}{.33\textwidth}
      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
        \begin{minipage}[T]{.95\textwidth}  % tweaks the width, makes a new \textwidth
          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
            % ---------------------------------------------------------%
            % fill each column with content

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{block}{Summary}
\noindent
\hspace{-1.6cm}
Pathogens can evolve rapidly in response to changing
conditions (e.g., epidemic stage or public health interventions). 
Models of \textbf{eco-evolutionary dynamics} often
neglect important epidemiological processes, such as the formation and
dissolution of sexual partnerships. We
compared \textbf{models with a range of complexity} in their
representation of partnership dynamics and extra-partnership contact.
Simple random-mixing actually gave the closest match to the most realistic models; extra-pair contact washed out the effects
of epidemiological structure. Nevertheless, the large differences
in evolutionary dynamics among different epidemiological models
suggests that researchers trying to predict the evolution of
pathogens should proceed with caution.

\end{block}

\vspace{0.5cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{block}{Tradeoff theory}

<<tradeoff_fig,echo=FALSE,fig.width=10>>=
print(gg_tradeoff)
@ 
  \begin{itemize}
    \item \textbf{tradeoff theory}: virulence evolution
      mediated by transmission-vs-clearance tradeoff
    \item still debated \cite{EbertBull2003,alizon_adaptive_2015}
    \item HIV \cite{fraser_virulence_2014}: \textbf{set-point viral load}
      correlated with transmission probability, rate of progression to AIDS
    \item data from Rakai, Uganda
    \item eco-evolutionary virulence dynamics: \cite{shirreff_transmission_2011}
  \end{itemize}

\end{block}

%%\vfill
\vspace{0.5cm}

\begin{block}{Epidemiological structure}

  %\vspace{-3cm}
  \begin{center}
    \includegraphics[width=\textwidth]{images/champ1.png}
  \end{center}

Spectrum of model complexity: 
\begin{itemize}
\item \textbf{pair formation}: instantaneous or delayed?
\item \textbf{extra-pair contact} (\textbf{epc}): present or absent?
\item \textbf{implicit model}: no explicit partnerships,
  force of infection expression derived from $\Ro$ of pair-formation
  model (without epc)
\item \textbf{random-mixing model}: standard SIR model
\end{itemize}

Slightly simplified disease model (single stage only;
genetic but not phenotypic variation)

\end{block}

}
\end{minipage}
\end{beamercolorbox}
\end{column}
    % ---------------------------------------------------------%
    % end the column

    % ---------------------------------------------------------%
    % Set up a column 
    \begin{column}{.33\textwidth}
      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
            % ---------------------------------------------------------%
            % fill each column with content

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{block}{Basic evolutionary dynamics}

\vspace{-1cm}

<<virdyn,cache=TRUE,dev="tikz",fig.width=25,fig.height=30>>=
ss <- subset(sum_list[["vir_mat"]],tvec<750)
ss$model <- factor(ss$model, c("random",
         "pairform+epc", "pairform","instswitch+epc","instswitch","implicit"))

gg_virtraj <- ggplot(ss,
                     aes(tvec,mean,ymin=lwr,ymax=upr))+
    geom_line(aes(colour=model,linetype=model),lwd=1)+
    geom_ribbon(aes(fill = model), alpha=0.3)+
    labs(x="time (years)",y=Lspvl) +
    ## theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
    ##  axis.title.x = element_text(margin = margin(10,0,0,0))) +
    facet_wrap(~model, nrow = 3) +
    zero_margin+
    theme(legend.text=element_text(size=16))
## scale_x_continuous(limits=c(0,2500))
print(gg_virtraj)
@

\vspace{0.5cm}

\end{block}            
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{block}{Univariate summary}

<<univ,cache=TRUE,dev="tikz",fig.width=45,fig.height=30>>=
sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
mL <- melt(sL,id.vars=c("model","run"))
## horrible hack (but doesn't help); subsample
## w <- which(mL$model=="random")
## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
w <- with(mL,which(model=="random" & variable=="eq_vir"))
rval <- mean(mL$value[w])
mLw <- droplevels(mL[-w,])
mLw$variable <- fixfac2(mLw$variable)

gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
	geom_violinh(width=1)+
	facet_wrap(~variable,scale="free_x")+
	guides(fill=guide_legend(reverse=TRUE))+
	labs(y="")+
	geom_point(data=data.frame(model="random",variable="equilibrium virulence",
                               value=rval),pch=22,size=3,show.legend=FALSE) +
	zero_x_margin
print(gg_univ)
@

\end{block}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\vspace{0.5cm}

\begin{block}{Bivariate summary}

<<ggfuns,echo=FALSE>>=
trim_gg <- function(gg,hack_spaces=TRUE) {
    n <- gg$nrow
    gg$nrow <- gg$ncol <- n-1
    v <- 1:n^2
    gg$plots <- gg$plots[v>n & v%%n!=0]
    gg$xAxisLabels <- gg$xAxisLabels[-n]
    gg$yAxisLabels <- gg$yAxisLabels[-1]
    ## ggpairs tries to parse labels, so this fails if they
    ## have spaces in them ... maybe there's a better way?
    if (hack_spaces) {
        gg$xAxisLabels <- gsub("_"," ",gg$xAxisLabels)
        gg$yAxisLabels <- gsub("_"," ",gg$yAxisLabels)
    }
    return(gg)
}

tweak_legends_gg <- function(gg,pos=1.5) {
    n <- gg$nrow
    for (i in 1:n) {
        for (j in 1:n) {
            inner <- getPlot(gg,i,j)
            if (i==1 & j==1) {
                inner <- inner + ggplot2::theme(legend.position=c(pos,0.5)) +
                    guides(colour = guide_legend(override.aes = list(size=6)))
            } else {
                inner <- inner + ggplot2::theme(legend.position="none")
            }
            gg <- putPlot(gg,inner,i,j)
        }
    }
    return(gg)
}

tweak_colours_gg <- function(gg) {
    n <- gg$nrow
    for (i in 1:n) {
        for (j in 1:n) {
            inner <- getPlot(gg,i,j)
            inner <- inner+scale_colour_brewer(palette = 'Dark2')
            gg <- putPlot(gg,inner,i,j)
        }
    }
    return(gg)
}
@

<<ggpairs,echo=FALSE,fig.width=25,fig.height=25>>=
sL2 <- sL
names(sL2)[na.omit(match(orig_sum_labs,names(sL)))] <- gsub(" ","_",new_sum_labs)
    ## paste0("`",new_sum_labs,"`")
ggp1 <- ggpairs(sL2,
        mapping = ggplot2::aes(color = model),
        columns=3:6,
        legends=TRUE,
        lower = list(continuous = wrap("points",size=0.1)), ## alpha = 0.3,size=0.5)),
        diag = list(continuous = "blankDiag"),
        upper = list(continuous = "blank"))
## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot
## have to change plot one panel at a time
ggp2 <- tweak_colours_gg(tweak_legends_gg(trim_gg(ggp1)))
## add 1-to-1 line
inner <- getPlot(ggp2,2,1)
inner <- inner+geom_abline(intercept=0,slope=1,lty=2)
ggp2 <- putPlot(ggp2,inner,2,1)
## tweak scale
inner <- getPlot(ggp2,3,1)
inner <- inner+scale_x_continuous(breaks=seq(2.5,4.5,by=0.5))
ggp2 <- putPlot(ggp2,inner,3,1)
print(ggp2,spacingProportion=0)
@   
\end{block}

}
          % ---------------------------------------------------------%
          % end the column
        \end{minipage}
      \end{beamercolorbox}
   \end{column}

    % ---------------------------------------------------------%
    % end the column
    \begin{column}{.33\textwidth}
      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
            % ---------------------------------------------------------%
            % fill each column with content

            \begin{block}{Sensivity}
<<sens,dev="png",echo=FALSE>>=
mL3 <- subset(mL2,
       !((model=="implicit" &
          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
         (model=="random" &
          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
         (model=="instswitch" &
          LHSvar %in% c("rho_base","c_u_ratio","c_e_ratio")) |
         (model=="instswitch+epc" &
          LHSvar %in% c("rho_base","c_u_ratio")) |
         (model=="pairform" &
          LHSvar %in% c("c_u_ratio","c_e_ratio"))))
## reorder factors
mL3$LHSvar <- factor(mL3$LHSvar,
                     levels=c("Beta1","Beta3",
                              "Duration1","Duration3",
                              "c_mean_base","c_e_ratio","rho_base",
                              "c_u_ratio"),
                     labels = c("beta[P]", "beta[D]", "D[P]", "D[D]",
                                "c", "c[e]/c[w]", "rho", "c[u]/c[w]"))

## FIXME: could also use labels argument to rename levels, labeller=label_parsed
## in facet_grid to get pretty math

mL3$model <- factor(mL3$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))

mL3$sumvar <- fixfac2(mL3$sumvar)

L <- function(labels,multi_line=TRUE) {
    r <- if (all(grepl(" ",labels[[1]]))) {
             list(as.character(gsub(" ","\n",labels[[1]])))
    } else {
        label_parsed(labels,multi_line=multi_line)
    }
    return(r)
}

brkfun <- function(x) {
    nx <- 1
    n <- 2
    brks <- axisTicks(log(x), log=TRUE,  nint = n)
    ## browser()
    brks <- brks[brks>x[1] & brks<x[2]]
    nx <- length(brks)
    ## cat("brk1",brks,"\n")
    while (nx<2 && n<10) {
        n <- n+1
        ## cat(n,"\n")
        brks <- axisTicks(log10(x), log=TRUE,  nint = n)
        brks <- brks[brks>x[1] & brks<x[2]]
        ## cat("brk2",brks,"\n")
        nx <- length(brks)
    }
    if (n==10) browser()
    r <- brks[c(1,length(brks))]
    ## hacks
    if (all(r==c(0.01,1))) r <- c(0.02,0.5)
    if (all(r==c(0.2,5))) r <- c(0.5,5)
    return(r)
}

ggplot(mL3,aes(LHSval,sumval,colour=model))+
    geom_point(pch=".",alpha=0.5)+
    facet_grid(sumvar~LHSvar,scales="free",
               labeller = L)+
    geom_smooth(se=FALSE)+labs(x="",y="")+
    ## leave a little extra room?
    scale_x_log10(expand=c(0,0.08),breaks=brkfun)+
    zero_margin
@               
              \end{block}
            \begin{block}{Conclusions}
\end{block}
\begin{block}{References}

{\tiny
%% Add numbered references
\setbeamertemplate{bibliography item}[text]
%% Stop line breaks
\setbeamertemplate{bibliography entry title}{}
\setbeamertemplate{bibliography entry location}{}
\setbeamertemplate{bibliography entry note}{}
%% Stop font color changes
% \setbeamercolor*{bibliography item}{fg=black}
% \setbeamercolor*{bibliography entry title}{fg=black}
% \setbeamercolor*{bibliography entry author}{fg=black}
% \setbeamercolor*{bibliography entry location}{fg=black}
% \setbeamercolor*{bibliography entry note}{fg=black} 
%% Generate bibliography
\bibliographystyle{vancouver}
\bibliography{../ms/virulence}
}

\end{block}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\vspace{0.5cm}

\begin{block}{Acknowledgements}

We thank Christophe Fraser for providing access to code, and 
the Natural Sciences and Engineering Research Council of Canada (NSERC)
for funding.

\end{block}

          }
        \end{minipage}
      \end{beamercolorbox}
   \end{column}

 \end{columns}
\end{frame}
\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Local Variables: 
%%% mode: latex
%%% TeX-PDF-mode: t
%%% End:

### 2016-05-29 poster/hivlhs_eeid2016.Rnw
more makefile/ms tweaks

@@
 
-<<setup3, echo=FALSE,dev="tikz",cache=TRUE>>=
+<<setup3,echo=FALSE,dev="tikz">>=
 source("../R/hivFuns.R")
-source("../R/LHSparms3.R")
+source("../R/Param.R")
+ltab_mean <- apply(HIVpars_range, 1, geom_mean)
+HIVpars.base <- as.HIVvirparlist(ltab_mean)
 Lspvl <- "$\\log_{10}~\\textrm{SPVL}$"
-pars <- expand(ltab_mean)
+pars <- expand(HIVpars.base)
 dd <-with(pars,
@@
 Models of \textbf{eco-evolutionary dynamics} often
-neglect important epidemiological processes, such as the formation and
-dissolution of sexual partnerships. We
-compared \textbf{models with a range of complexity} in their
-representation of partnership dynamics and extra-partnership contact.
-Simple random-mixing actually gave the closest match to the most realistic models; extra-pair contact washed out the effects
-of epidemiological structure. Nevertheless, the large differences
-in evolutionary dynamics among different epidemiological models
-suggests that researchers trying to predict the evolution of
-pathogens should proceed with caution.
-
+neglect important epidemiological processes, such as the 
+dynamics of sexual partnerships. We
+compared \textbf{models with a range of complexity} 
+of partnership dynamics and extra-partnership contact.
 \end{block}
@@
     \item HIV \cite{fraser_virulence_2014}: \textbf{set-point viral load}
-      correlated with transmission probability, rate of progression to AIDS
-    \item data from Rakai, Uganda
+      correlated with transmission probability, rate of progression to AIDS (data from Rakai, Uganda)
     \item eco-evolutionary virulence dynamics: \cite{shirreff_transmission_2011}
@@
 
-\vspace{-1cm}
-
 <<virdyn,cache=TRUE,dev="tikz",fig.width=25,fig.height=30>>=
@@
     labs(x="time (years)",y=Lspvl) +
-    ## theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
-    ##  axis.title.x = element_text(margin = margin(10,0,0,0))) +
+    theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
+          axis.title.x = element_text(margin = margin(10,0,0,0))) +
     facet_wrap(~model, nrow = 3) +
     zero_margin+
-    theme(legend.text=element_text(size=16))
-## scale_x_continuous(limits=c(0,2500))
+    theme(legend.position="none")+
+    scale_x_continuous(breaks=c(0,600))
 print(gg_virtraj)
@@
 \vspace{0.5cm}
+
+\begin{itemize}
+
+\end{itemize}
 
@@
               \end{block}
+
             \begin{block}{Conclusions}
+
+\begin{itemize}
+\item Random-mixing models gave the closest match to the most realistic models; extra-pair contact washed out the effects
+of epidemiological structure
+\item Large differences
+in evolutionary dynamics among different epidemiological models
+suggest caution in predicting evolutionary responses
+\end{itemize}
+
 \end{block}

### 2016-05-29 poster/hivlhs_eeid2016.Rnw
working on poster

@@
 \usepackage{grffile}
-\mode<presentation>{\usetheme{drexel}}
+\mode<presentation>{\usetheme{mcmaster}}
 
+\usepackage{wrapfig}
+% \usepackage{picins} %% ??? screws up tikz ????
 \usepackage[english]{babel}
@@
 
-<<setup3,echo=FALSE,dev="tikz">>=
+<<setup3,echo=FALSE>>=
 source("../R/hivFuns.R")
@@
 \setlength{\columnheight}{105cm}
+
+\newlength{\shortcolumnheight}
+\setlength{\shortcolumnheight}{75cm}
 
@@
 
-<<tradeoff_fig,echo=FALSE,fig.width=10>>=
+<<tradeoff_fig,echo=FALSE,fig.height=14,fig.width=30,out.width="0.95\\textwidth",dev="tikz",cache=TRUE>>=
 print(gg_tradeoff)
 @
+
 \begin{itemize}
-    \item \textbf{tradeoff theory}: virulence evolution
-      mediated by transmission-vs-clearance tradeoff
+\item \textbf{tradeoff theory}: virulence evolution mediated by transmission-vs-clearance tradeoff
 \item still debated \cite{EbertBull2003,alizon_adaptive_2015}
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{images/champ1.png}
+    \includegraphics[width=\textwidth]{images/champ1.png} \\
+    Champredon \etal\ 2013 \cite{champredon_hiv_2013}
   \end{center}
@@
 \item \textbf{pair formation}: instantaneous or delayed?
-\item \textbf{extra-pair contact} (\textbf{epc}): present or absent?
+\item \textbf{extra-pair contact} (epc): present or absent?
 \item \textbf{implicit model}: no explicit partnerships,
@@
 
-Slightly simplified disease model (single stage only;
-genetic but not phenotypic variation)
+Simplified disease model (single stage only)
 
@@
 
+\begin{block}{Sampling strategy}
+% https://www.sharelatex.com/learn/Wrapping_text_around_figures
+  \setlength{\abovecaptionskip}{0pt}
+  \setlength{\belowcaptionskip}{0pt}
+  \setlength{\intextsep}{0pt}
+  \setlength{\textfloatsep}{0pt}
+% \begin{wrapfigure}{l}{0.6\textwidth}
+%%\begin{columns}
+%%  \begin{column{0.5
+%%\begin{center}
+
+%\pichskip{15pt}% Horizontal gap between picture and text
+%\parpic[l][t]{%
+%  \begin{minipage}{0.6\textwidth}
+    \includegraphics[width=0.55\textwidth]{images/LHScrop.png}
+%  \end{minipage}
+%}
+%%  \end{center}
+%\caption{}
+%\end{wrapfigure}
+\begin{itemize}
+  \item Latin hypercube sampling: parameters from
+    \cite{champredon_hiv_2013}
+  \item parameters calibrated across models to the same \textbf{initial epidemic growth rate} ($r$)
+\end{itemize}
+\end{block}
 }
@@
     % Set up a column 
-    \begin{column}{.33\textwidth}
+\begin{column}{.65\textwidth}
+\begin{columns}
+    \begin{column}{.5\textwidth}
       \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
         \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
-          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
+          \parbox[t][\shortcolumnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
             % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
@@
 
-\begin{block}{Basic evolutionary dynamics}
+\begin{block}{Basic eco-evolutionary dynamics}
 
-<<virdyn,cache=TRUE,dev="tikz",fig.width=25,fig.height=30>>=
+<<virdyn,dev="tikz",fig.width=25,fig.height=30,cache=TRUE,out.width="0.95\\textwidth">>=
 ss <- subset(sum_list[["vir_mat"]],tvec<750)
@@
 \begin{itemize}
-
+\item significant variation across model structures
+\item least (random) and most (pairform+epc) models \textbf{most similar}:
+  single individuals and extra-pair contact wash out effects of structure
+\item implicit model is \textbf{most different}
+\item random-mixing model underestimates (?) variability
 \end{itemize}
@@
 
-<<univ,cache=TRUE,dev="tikz",fig.width=45,fig.height=30>>=
+<<univ,cache=TRUE,dev="tikz",fig.width=45,fig.height=40,out.width="0.95\\textwidth">>=
 sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
@@
     labs(y="")+
-	geom_point(data=data.frame(model="random",variable="equilibrium virulence",
-                               value=rval),pch=22,size=3,show.legend=FALSE) +
+    geom_point(data=data.frame(model="random",
+                               variable="equilibrium virulence",
+                               value=rval),
+               pch=22,size=8,show.legend=FALSE) +
+    theme(legend.position="none")+
     zero_x_margin
@@
 
+}
+         % ---------------------------------------------------------%
+          % end the column
+        \end{minipage}
+      \end{beamercolorbox}
+   \end{column}
+
+    % ---------------------------------------------------------%
+    % end the column
+    \begin{column}{.5\textwidth}
+      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
+        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
+          \parbox[t][\shortcolumnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
+            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
+            % ---------------------------------------------------------%
+            % fill each column with content
+
 \begin{block}{Bivariate summary}
-
 <<ggfuns,echo=FALSE>>=
@@
 
-<<ggpairs,echo=FALSE,fig.width=25,fig.height=25>>=
+<<ggpairs,echo=FALSE,fig.width=10,fig.height=10>>=
 sL2 <- sL
@@
 @
+
+\end{block}
+
+\begin{block}{Conclusions and open questions}
+
+\begin{itemize}
+\item Random-mixing models gave the closest match to the most realistic models; extra-pair contact washed out the effects
+of epidemiological structure
+\item Variation among models (model structure) $\approx$ variation within models (parameter uncertainty)
+\item Large differences
+in evolutionary dynamics among different epidemiological models
+suggest caution in predicting evolutionary responses
+\end{itemize}
+
+\end{block}
+
+\begin{block}{References}
+
+{\tiny
+%% Add numbered references
+\setbeamertemplate{bibliography item}[text]
+%% Stop line breaks
+\setbeamertemplate{bibliography entry title}{}
+\setbeamertemplate{bibliography entry location}{}
+\setbeamertemplate{bibliography entry note}{}
+%% Stop font color changes
+% \setbeamercolor*{bibliography item}{fg=black}
+% \setbeamercolor*{bibliography entry title}{fg=black}
+% \setbeamercolor*{bibliography entry author}{fg=black}
+% \setbeamercolor*{bibliography entry location}{fg=black}
+% \setbeamercolor*{bibliography entry note}{fg=black} 
+%% Generate bibliography
+\bibliographystyle{vancouver}
+\bibliography{../ms/virulence}
+}
+
+\end{block}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% \vspace{0.5cm}
+
+\begin{block}{Acknowledgements}
+
+We thank Christophe Fraser for providing access to code, and 
+the Natural Sciences and Engineering Research Council of Canada (NSERC)
+for funding.
+
 \end{block}
@@
 }
-          % ---------------------------------------------------------%
-          % end the column
         \end{minipage}
@@
     \end{column}
-
-    % ---------------------------------------------------------%
-    % end the column
-    \begin{column}{.33\textwidth}
+    \end{columns}
      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
-          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
-            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
-            % ---------------------------------------------------------%
-            % fill each column with content
+         \parbox[t][30cm]{\textwidth}{ % must be some better way to set
 
-            \begin{block}{Sensivity}
-<<sens,dev="png",echo=FALSE>>=
+\begin{block}{Sensitivity}
+<<sens,dev="png",echo=FALSE,fig.width=20,fig.height=10,cache=TRUE>>=
 mL3 <- subset(mL2,
@@
 \end{block}
-
-            \begin{block}{Conclusions}
-
-\begin{itemize}
-\item Random-mixing models gave the closest match to the most realistic models; extra-pair contact washed out the effects
-of epidemiological structure
-\item Large differences
-in evolutionary dynamics among different epidemiological models
-suggest caution in predicting evolutionary responses
-\end{itemize}
-
-\end{block}
-\begin{block}{References}
-
-{\tiny
-%% Add numbered references
-\setbeamertemplate{bibliography item}[text]
-%% Stop line breaks
-\setbeamertemplate{bibliography entry title}{}
-\setbeamertemplate{bibliography entry location}{}
-\setbeamertemplate{bibliography entry note}{}
-%% Stop font color changes
-% \setbeamercolor*{bibliography item}{fg=black}
-% \setbeamercolor*{bibliography entry title}{fg=black}
-% \setbeamercolor*{bibliography entry author}{fg=black}
-% \setbeamercolor*{bibliography entry location}{fg=black}
-% \setbeamercolor*{bibliography entry note}{fg=black} 
-%% Generate bibliography
-\bibliographystyle{vancouver}
-\bibliography{../ms/virulence}
-}
-
-\end{block}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\vspace{0.5cm}
-
-\begin{block}{Acknowledgements}
-
-We thank Christophe Fraser for providing access to code, and 
-the Natural Sciences and Engineering Research Council of Canada (NSERC)
-for funding.
-
-\end{block}
-
 }
@@
     \end{beamercolorbox}
-   \end{column}
-
- \end{columns}
+ \end{column}  % end right-hand 2/3 column
+ \end{columns}  % end overall columns
 \end{frame}

### 2016-05-30 poster/hivlhs_eeid2016.Rnw
working on poster

@@
 \documentclass{beamer}
+\usepackage{ebgaramond}
+%% reset default
+\renewcommand{\rmdefault}{cmss}
 \usepackage{grffile}
@@
 \newcommand{\pphantom}{\textcolor{ta3aluminium}} % phantom introduces a vertical space in p formatted table columns??!!
+
+% http://tex.stackexchange.com/questions/13423/how-to-change-the-color-of-href-links-for-real
+\hypersetup{colorlinks,linkcolor=}
 
@@
 
-<<setup3,echo=FALSE>>=
+<<setup3,echo=FALSE,cache=TRUE>>=
 source("../R/hivFuns.R")
@@
 gg_tradeoff <- ggplot(dd,aes(alpha,value))+facet_wrap(~variable,scale="free")+
-    geom_line()+
-    labs(x= Lspvl)
+    geom_line(size=10)+
+    labs(x= Lspvl,y="")
 @
@@
 
-<<tradeoff_fig,echo=FALSE,fig.height=14,fig.width=30,out.width="0.95\\textwidth",dev="tikz",cache=TRUE>>=
+<<tradeoff_fig,echo=FALSE,fig.height=14,fig.width=30,out.width="0.95\\textwidth",dev="tikz",cache=TRUE,dependson="setup3">>=
 print(gg_tradeoff)
@@
 \begin{itemize}
-\item \textbf{tradeoff theory}: virulence evolution mediated by transmission-vs-clearance tradeoff
+\item virulence evolution mediated by transmission-vs-clearance tradeoff
 \item still debated \cite{EbertBull2003,alizon_adaptive_2015}
@@
 
-\begin{block}{Basic eco-evolutionary dynamics}
+\begin{block}{Eco-evolutionary dynamics}
 
@@
 
-<<univ,cache=TRUE,dev="tikz",fig.width=45,fig.height=40,out.width="0.95\\textwidth">>=
+<<univ,cache=TRUE,dev="tikz",fig.width=30,fig.height=30,out.width="0.95\\textwidth">>=
 sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
@@
     guides(fill=guide_legend(reverse=TRUE))+
-    labs(y="")+
+    labs(y="",x="")+
     geom_point(data=data.frame(model="random",

### 2016-05-30 poster/hivlhs_eeid2016.Rnw
restructure poster; drop URLs from bib

@@
 
-\newlength{\shortcolumnheight}
-\setlength{\shortcolumnheight}{75cm}
+\newlength{\topcolumnheight}
+\setlength{\topcolumnheight}{50cm}
+
+\newlength{\bottomcolumnheight}
+\setlength{\bottomcolumnheight}{30cm}
 
@@
   \vspace{0.5cm}
-  \noindent\begin{columns}
+  \noindent
+\begin{columns}
     % ---------------------------------------------------------%
-    % Set up a column 
   \begin{column}{.33\textwidth}
     \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
-        \begin{minipage}[T]{.95\textwidth}  % tweaks the width, makes a new \textwidth
-          \parbox[t][\columnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
-            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
-            % ---------------------------------------------------------%
-            % fill each column with content
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+      \begin{minipage}[T]{.95\textwidth}
+        \parbox[t][\columnheight]{\textwidth}{
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{block}{Tradeoff theory}
 
-<<tradeoff_fig,echo=FALSE,fig.height=14,fig.width=30,out.width="0.95\\textwidth",dev="tikz",cache=TRUE,dependson="setup3">>=
-print(gg_tradeoff)
-@
+\includegraphics[width=0.95\textwidth]{figure/tradeoff.pdf}
 
@@
 
-%%\vfill
 \vspace{0.5cm}
@@
 \end{column}
-    % ---------------------------------------------------------%
-    % end the column
 
-    % ---------------------------------------------------------%
-    % Set up a column 
 \begin{column}{.65\textwidth}
+
 \begin{columns}
@@
     \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
-        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
-          \parbox[t][\shortcolumnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
-            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
-            % ---------------------------------------------------------%
-            % fill each column with content
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+      \begin{minipage}[T]{.95\textwidth}
+        \parbox[t][\topcolumnheight]{\textwidth}{
 
@@
 
-<<virdyn,dev="tikz",fig.width=25,fig.height=30,cache=TRUE,out.width="0.95\\textwidth">>=
-ss <- subset(sum_list[["vir_mat"]],tvec<750)
-ss$model <- factor(ss$model, c("random",
-         "pairform+epc", "pairform","instswitch+epc","instswitch","implicit"))
-
-gg_virtraj <- ggplot(ss,
-                     aes(tvec,mean,ymin=lwr,ymax=upr))+
-    geom_line(aes(colour=model,linetype=model),lwd=1)+
-    geom_ribbon(aes(fill = model), alpha=0.3)+
-    labs(x="time (years)",y=Lspvl) +
-    theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
-          axis.title.x = element_text(margin = margin(10,0,0,0))) +
-    facet_wrap(~model, nrow = 3) +
-    zero_margin+
-    theme(legend.position="none")+
-    scale_x_continuous(breaks=c(0,600))
-print(gg_virtraj)
-@
+\includegraphics[width=0.95\textwidth]{figure/virtraj.pdf}
 
@@
 \end{block}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{block}{Univariate summary}
-
-<<univ,cache=TRUE,dev="tikz",fig.width=30,fig.height=30,out.width="0.95\\textwidth">>=
-sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
-sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
-mL <- melt(sL,id.vars=c("model","run"))
-## horrible hack (but doesn't help); subsample
-## w <- which(mL$model=="random")
-## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
-w <- with(mL,which(model=="random" & variable=="eq_vir"))
-rval <- mean(mL$value[w])
-mLw <- droplevels(mL[-w,])
-mLw$variable <- fixfac2(mLw$variable)
-
-gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
-    geom_violinh(width=1)+
-    facet_wrap(~variable,scale="free_x")+
-    guides(fill=guide_legend(reverse=TRUE))+
-    labs(y="",x="")+
-    geom_point(data=data.frame(model="random",
-                               variable="equilibrium virulence",
-                               value=rval),
-               pch=22,size=8,show.legend=FALSE) +
-    theme(legend.position="none")+
-    zero_x_margin
-print(gg_univ)
-@
-
-\end{block}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\vspace{0.5cm}
-
-}
-         % ---------------------------------------------------------%
-          % end the column
+        } % end parbox
         \end{minipage}
@@
 
-    % ---------------------------------------------------------%
-    % end the column
   \begin{column}{.5\textwidth}
@@
       \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
-          \parbox[t][\shortcolumnheight]{\textwidth}{ % must be some better way to set the the height, width and textwidth simultaneously
-            % Since all columns are the same length, it is all nice and tidy.  You have to get the height empirically
-            % ---------------------------------------------------------%
-            % fill each column with content
+        \parbox[t][\topcolumnheight]{\textwidth}{
 
-\begin{block}{Bivariate summary}
-<<ggfuns,echo=FALSE>>=
-trim_gg <- function(gg,hack_spaces=TRUE) {
-    n <- gg$nrow
-    gg$nrow <- gg$ncol <- n-1
-    v <- 1:n^2
-    gg$plots <- gg$plots[v>n & v%%n!=0]
-    gg$xAxisLabels <- gg$xAxisLabels[-n]
-    gg$yAxisLabels <- gg$yAxisLabels[-1]
-    ## ggpairs tries to parse labels, so this fails if they
-    ## have spaces in them ... maybe there's a better way?
-    if (hack_spaces) {
-        gg$xAxisLabels <- gsub("_"," ",gg$xAxisLabels)
-        gg$yAxisLabels <- gsub("_"," ",gg$yAxisLabels)
-    }
-    return(gg)
-}
+            \begin{block}{Univariate summary}
+              \includegraphics[width=0.95\textwidth]{figure/univ.pdf}
+            \end{block}
+            \vspace{0.5cm}
 
-tweak_legends_gg <- function(gg,pos=1.5) {
-    n <- gg$nrow
-    for (i in 1:n) {
-        for (j in 1:n) {
-            inner <- getPlot(gg,i,j)
-            if (i==1 & j==1) {
-                inner <- inner + ggplot2::theme(legend.position=c(pos,0.5)) +
-                    guides(colour = guide_legend(override.aes = list(size=6)))
-            } else {
-                inner <- inner + ggplot2::theme(legend.position="none")
-            }
-            gg <- putPlot(gg,inner,i,j)
-        }
-    }
-    return(gg)
-}
-
-tweak_colours_gg <- function(gg) {
-    n <- gg$nrow
-    for (i in 1:n) {
-        for (j in 1:n) {
-            inner <- getPlot(gg,i,j)
-            inner <- inner+scale_colour_brewer(palette = 'Dark2')
-            gg <- putPlot(gg,inner,i,j)
-        }
-    }
-    return(gg)
-}
-@
-
-<<ggpairs,echo=FALSE,fig.width=10,fig.height=10>>=
-sL2 <- sL
-names(sL2)[na.omit(match(orig_sum_labs,names(sL)))] <- gsub(" ","_",new_sum_labs)
-    ## paste0("`",new_sum_labs,"`")
-ggp1 <- ggpairs(sL2,
-        mapping = ggplot2::aes(color = model),
-        columns=3:6,
-        legends=TRUE,
-        lower = list(continuous = wrap("points",size=0.1)), ## alpha = 0.3,size=0.5)),
-        diag = list(continuous = "blankDiag"),
-        upper = list(continuous = "blank"))
-## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot
-## have to change plot one panel at a time
-ggp2 <- tweak_colours_gg(tweak_legends_gg(trim_gg(ggp1)))
-## add 1-to-1 line
-inner <- getPlot(ggp2,2,1)
-inner <- inner+geom_abline(intercept=0,slope=1,lty=2)
-ggp2 <- putPlot(ggp2,inner,2,1)
-## tweak scale
-inner <- getPlot(ggp2,3,1)
-inner <- inner+scale_x_continuous(breaks=seq(2.5,4.5,by=0.5))
-ggp2 <- putPlot(ggp2,inner,3,1)
-print(ggp2,spacingProportion=0)
-@
-
-\end{block}
-
-\begin{block}{Conclusions and open questions}
-
-\begin{itemize}
-\item Random-mixing models gave the closest match to the most realistic models; extra-pair contact washed out the effects
-of epidemiological structure
-\item Variation among models (model structure) $\approx$ variation within models (parameter uncertainty)
-\item Large differences
-in evolutionary dynamics among different epidemiological models
-suggest caution in predicting evolutionary responses
-\end{itemize}
-
-\end{block}
-
-\begin{block}{References}
-
-{\tiny
-%% Add numbered references
-\setbeamertemplate{bibliography item}[text]
-%% Stop line breaks
-\setbeamertemplate{bibliography entry title}{}
-\setbeamertemplate{bibliography entry location}{}
-\setbeamertemplate{bibliography entry note}{}
-%% Stop font color changes
-% \setbeamercolor*{bibliography item}{fg=black}
-% \setbeamercolor*{bibliography entry title}{fg=black}
-% \setbeamercolor*{bibliography entry author}{fg=black}
-% \setbeamercolor*{bibliography entry location}{fg=black}
-% \setbeamercolor*{bibliography entry note}{fg=black} 
-%% Generate bibliography
-\bibliographystyle{vancouver}
-\bibliography{../ms/virulence}
-}
-
-\end{block}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \vspace{0.5cm}
-
-\begin{block}{Acknowledgements}
-
-We thank Christophe Fraser for providing access to code, and 
-the Natural Sciences and Engineering Research Council of Canada (NSERC)
-for funding.
-
-\end{block}
-
-}
+        } % end parbox
         \end{minipage}
@@
     \end{column}
+
 \end{columns}
+
 \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
@@
     \end{beamercolorbox}
+
+\begin{columns}
+    \begin{column}{.5\textwidth}
+      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
+        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
+          \parbox[t][\bottomcolumnheight]{\textwidth}{
+
+\begin{block}{Conclusions and open questions}
+
+\begin{itemize}
+\item Random-mixing models gave the closest match to the most realistic models; extra-pair contact washed out the effects
+of epidemiological structure
+\item Variation among models (model structure) $\approx$ variation within models (parameter uncertainty)
+\item Large differences
+in evolutionary dynamics among different epidemiological models
+suggest caution in predicting evolutionary responses
+\end{itemize}
+
+\end{block}
+}
+        \end{minipage}
+      \end{beamercolorbox}
+    \end{column}
+    \begin{column}{.5\textwidth}
+      \begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
+        \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
+          \parbox[t][\bottomcolumnheight]{\textwidth}{
+
+\begin{block}{References}
+
+{\tiny
+%% Add numbered references
+\setbeamertemplate{bibliography item}[text]
+%% Stop line breaks
+\setbeamertemplate{bibliography entry title}{}
+\setbeamertemplate{bibliography entry location}{}
+\setbeamertemplate{bibliography entry note}{}
+%% Stop font color changes
+% \setbeamercolor*{bibliography item}{fg=black}
+% \setbeamercolor*{bibliography entry title}{fg=black}
+% \setbeamercolor*{bibliography entry author}{fg=black}
+% \setbeamercolor*{bibliography entry location}{fg=black}
+% \setbeamercolor*{bibliography entry note}{fg=black} 
+%% Generate bibliography
+\bibliographystyle{vancouver}
+\bibliography{../ms/virulence}
+}
+\end{block}
+
+\begin{block}{Acknowledgements}
+
+We thank Christophe Fraser for providing access to code, and 
+the Natural Sciences and Engineering Research Council of Canada (NSERC)
+for funding.
+
+\end{block}
+
+}
+        \end{minipage}
+      \end{beamercolorbox}
+    \end{column}
+    \end{columns}
  \end{column}  % end right-hand 2/3 column
- \end{columns}  % end overall columns
+ \end{columns}  % end 50/50 columns
 \end{frame}

### 2016-05-30 poster/figure/tradeoff.R

@@ -1,0 +1,13 @@
+source("fig_setup.R")
+
+dd <-with(pars,
+     melt(data.frame(alpha=alpha,"transmission probability"=Beta2,
+                     "duration (years)"=Duration2,check.names=FALSE),
+          id.vars="alpha"))
+gg_tradeoff <- ggplot(dd,aes(alpha,value))+facet_wrap(~variable,scale="free")+
+    geom_line(size=10)+
+    labs(x= Lspvl,y="")
+
+tikz(file="tradeoff.tikz",width=12,height=8,standAlone=TRUE)
+print(gg_tradeoff)
+dev.off()

### 2016-05-30 poster/figure/univ.R

@@ -1,0 +1,28 @@
+source("fig_setup.R")
+
+sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
+sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
+mL <- melt(sL,id.vars=c("model","run"))
+## horrible hack (but doesn't help); subsample
+## w <- which(mL$model=="random")
+## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
+w <- with(mL,which(model=="random" & variable=="eq_vir"))
+rval <- mean(mL$value[w])
+mLw <- droplevels(mL[-w,])
+mLw$variable <- fixfac2(mLw$variable)
+
+gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
+    geom_violinh(width=1)+
+    facet_wrap(~variable,scale="free_x")+
+    guides(fill=guide_legend(reverse=TRUE))+
+    labs(y="",x="")+
+    geom_point(data=data.frame(model="random",
+                               variable="equilibrium virulence",
+                               value=rval),
+               pch=22,size=8,show.legend=FALSE) +
+    theme(legend.position="none")+
+    zero_x_margin
+
+tikz(file="univ.tikz",width=12,height=10,standAlone=TRUE)
+print(gg_univ)
+dev.off()

### 2016-05-30 poster/figure/virtraj.R

@@ -1,0 +1,21 @@
+source("fig_setup.R")
+
+ss <- subset(sum_list[["vir_mat"]],tvec<750)
+ss$model <- factor(ss$model, c("random",
+         "pairform+epc", "pairform","instswitch+epc","instswitch","implicit"))
+
+gg_virtraj <- ggplot(ss,
+                     aes(tvec,mean,ymin=lwr,ymax=upr))+
+    geom_line(aes(colour=model,linetype=model),lwd=1)+
+    geom_ribbon(aes(fill = model), alpha=0.3)+
+    labs(x="time (years)",y=Lspvl) +
+    theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
+          axis.title.x = element_text(margin = margin(10,0,0,0))) +
+    facet_wrap(~model, nrow = 3) +
+    zero_margin+
+    theme(legend.position="none")+
+    scale_x_continuous(breaks=c(0,600))
+
+tikz(file="virtraj.tikz",width=8,height=11,standAlone=TRUE)
+print(gg_virtraj)
+dev.off()

### 2016-05-30 poster/hivlhs_eeid2016.Rnw
still tweaking

@@
 \usepackage{ebgaramond}
+\newcommand{\fakebf}{\fontfamily{mdugm}\fontseries{b}\selectfont}
 %% reset default
@@
 %%%\title{\huge On the Persistence of Invading Pathogens}
-\title{\LARGE HIV virulence evolution in structured epidemic models}
+\title{\fakebf{HIV virulence evolution in structured epidemic models}}
 
@@
 \email{\href{mailto:bolker@mcmaster.ca}{\texttt{bolker@mcmaster.ca}}}
-\institute[McMaster University]{McMaster University, Hamilton, Ontario, Canada %%\\\vspace{5mm}\texttt{earn@math.mcmaster.ca}
+\institute[McMaster University]{McMaster University, Hamilton, Ontario, Canada 
+%%\\\vspace{5mm}\texttt{earn@math.mcmaster.ca}
 }
+%\institute[]{}
 \date[June 2016]{June 2016}
@@
 
+\vspace{-2.4 cm} % shouldn't need this??
 \begin{block}{Summary}
@@
 
-Spectrum of model complexity: 
+\textbf{infection} from (1) inf. partner (discordant couples);
+(2) other coupled inf.; (3) uncoupled inf.
+
+\vspace{0.2cm}
+Model choices:
 \begin{itemize}
@@
 %  \begin{minipage}{0.6\textwidth}
-    \includegraphics[width=0.55\textwidth]{images/LHScrop.png}
+    \includegraphics[width=0.25\textwidth]{images/LHScrop.png}
 %  \end{minipage}
@@
 
-            \begin{block}{Univariate summary}
+            \begin{block}{Univariate summaries}
+              \includegraphics[width=0.7\textwidth]{figure/univ_scheme.pdf}
+
+              \vspace{1cm}
               \includegraphics[width=0.95\textwidth]{figure/univ.pdf}

### 2016-05-30 poster/figure/tradeoff.R

@@ -5,7 +5,7 @@ dd <-with(pars,
                      "duration (years)"=Duration2,check.names=FALSE),
           id.vars="alpha"))
 gg_tradeoff <- ggplot(dd,aes(alpha,value))+facet_wrap(~variable,scale="free")+
-    geom_line(size=10)+
+    geom_line(size=8)+
     labs(x= Lspvl,y="")
 
 tikz(file="tradeoff.tikz",width=12,height=8,standAlone=TRUE)

### 2016-05-30 poster/figure/univ.R

@@ -23,6 +23,10 @@ gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
     theme(legend.position="none")+
     zero_x_margin
 
-tikz(file="univ.tikz",width=12,height=10,standAlone=TRUE)
+tikz(file="univ.tikz",w
+     ## width=12,height=10,
+     ## width=10,height=8,
+     width=9,height=9*5/6,
+     standAlone=TRUE)
 print(gg_univ)
 dev.off()

### 2016-05-31 poster/hivlhs_eeid2016.Rnw
further poster fussing

@@
-%\documentclass[final,hyperref={pdfpagelabels=false}]{beamer}
+%\documentclass[final,hyperref={hidelinks,pdfpagelabels=false}]{beamer}
 \documentclass{beamer}
@@
 % http://tex.stackexchange.com/questions/13423/how-to-change-the-color-of-href-links-for-real
-\hypersetup{colorlinks,linkcolor=}
+\hypersetup{hidelinks}
 
@@
 
-\author{\href{http://www.math.mcmaster.ca/bolker/}{Ben Bolker and Sang Woo Park}}
+\author{\fakebf\href{http://www.math.mcmaster.ca/bolker/}{Ben Bolker and Sang Woo Park}}
 \email{\href{mailto:bolker@mcmaster.ca}{\texttt{bolker@mcmaster.ca}}}
+%% URL for project (footer)
+\projurl{\texttt{https://github.com/bbolker/HIV\_LHS/}}
 \institute[McMaster University]{McMaster University, Hamilton, Ontario, Canada 
@@
 %% Uncomment the following line to define a custom logo (or to tweak the size of the logo)
-\DrexelLogo{\includegraphics[width=\WhenPortrait{1.75}{1}\linewidth]{images/Maclogo}}
+\McMasterLogo{\includegraphics[width=\WhenPortrait{1.75}{1}\linewidth]{images/Maclogo}}
 
@@
 \item still debated \cite{EbertBull2003,alizon_adaptive_2015}
-\item HIV \cite{fraser_virulence_2014}: \textbf{set-point viral load}
+\item HIV \cite{fraser_virulence_2014}: \textbf{set-point viral load} ($\approx$ ``virulence'')
       correlated with transmission probability, rate of progression to AIDS (data from Rakai, Uganda)
@@
 
-\textbf{infection} from (1) inf. partner (discordant couples);
+\textbf{infection} from (1) infected partner ($SI$ couples);
 (2) other coupled inf.; (3) uncoupled inf.
 
-\vspace{0.2cm}
-Model choices:
+\vspace{0.4cm}
+
 \begin{itemize}
@@
 
-\begin{block}{Sampling strategy}
+\begin{block}{Parameter uncertainty/exploration}
 % https://www.sharelatex.com/learn/Wrapping_text_around_figures
@@
 \begin{itemize}
-\item significant variation across model structures
+
 \item least (random) and most (pairform+epc) models \textbf{most similar}:
@@
 \item implicit model is \textbf{most different}
-\item random-mixing model underestimates (?) variability
+\item random-mixing model underestimates variability
 \end{itemize}
@@
               \includegraphics[width=0.95\textwidth]{figure/univ.pdf}
+\begin{itemize}
+\item peak timing: epc $>$ finite pair-formation effects
+\item equilibrium virulence: interaction
+\item low-equilibrium outcomes for intermediate-complexity models
+\end{itemize}
+
             \end{block}
@@
 \begin{itemize}
-\item Random-mixing models gave the closest match to the most realistic models; extra-pair contact washed out the effects
+\item Random-mixing models best matched the most realistic models; extra-pair and uncoupled individuals washed out the effects
 of epidemiological structure
+\item Implicit models did worst
 \item Variation among models (model structure) $\approx$ variation within models (parameter uncertainty)
@@
 in evolutionary dynamics among different epidemiological models
-suggest caution in predicting evolutionary responses
+$\to$ caution in predicting evolutionary responses
+\item \textbf{neglected:} disease life history details, sex workers, age-structured mixing \dots
+agent-based models?
 \end{itemize}
@@
 
-We thank Christophe Fraser for providing access to code, and 
+We thank C. Fraser and D. Champredon for code access and 
 the Natural Sciences and Engineering Research Council of Canada (NSERC)

### 2016-05-31 poster/figure/virtraj.R

@@ -8,7 +8,7 @@ gg_virtraj <- ggplot(ss,
                      aes(tvec,mean,ymin=lwr,ymax=upr))+
     geom_line(aes(colour=model,linetype=model),lwd=1)+
     geom_ribbon(aes(fill = model), alpha=0.3)+
-    labs(x="time (years)",y=Lspvl) +
+    labs(x="time (years)",y="population mean set-point viral load ($\\log_{10}$)") +
     theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
           axis.title.x = element_text(margin = margin(10,0,0,0))) +
     facet_wrap(~model, nrow = 3) +

### 2016-05-31 poster/hivlhs_eeid2016.Rnw
final?? poster

@@
 
+Simplified disease model (single stage only)
+
+\vspace{0.4cm}
+
 \begin{itemize}
@@
 \end{itemize}
-
-Simplified disease model (single stage only)
 
@@
 \begin{itemize}
-
+\item \textbf{80\% of total variability} in peak SPVL is among- vs. within-model
 \item least (random) and most (pairform+epc) models \textbf{most similar}:
@@
 \begin{block}{Sensitivity}
-<<sens,dev="png",echo=FALSE,fig.width=20,fig.height=10,cache=TRUE>>=
+<<sens,dev="png",echo=FALSE,fig.width=20,fig.height=10,cache=TRUE,out.width="0.95\\textwidth">>=
 mL3 <- subset(mL2,

### 2016-05-31 R/SSQ.R

@@ -6,8 +6,14 @@ model = sL$model
 eqV = sL$eq_vir
 peakV = sL$peak_vir
 
-lmer(eqV ~ 1 + (1|model), data = sL, REML=FALSE)
-lmer(peakV ~ 1 + (1|model), data = sL, REML=FALSE) 
+m_peakV <- lmer(peakV ~ 1 + (1|model), data = sL, REML=FALSE)
+v_among <- VarCorr(m_peakV)[[1]]
+v_within <- sigma(m_peakV)^2
+v_among/(v_among+v_within)
+
+
+library(nlme)
+lme(eqV~1,random=~1|model,data=sL,method="ML")
 
 findSSQ <- function(var){
 	wSSQ = sum((var-ave(var, model))^2) ##within

### 2016-06-01 poster/hivlhs_eeid2016.Rnw
working on poster

@@
 \begin{block}{Sensitivity}
-<<sens,dev="png",echo=FALSE,fig.width=20,fig.height=10,cache=TRUE,out.width="0.95\\textwidth">>=
+\includegraphics[width=0.9\textwidth]{figure/sens-1.png}
+<<sens,dev="png",echo=FALSE,fig.width=20,fig.height=10,cache=TRUE,out.width="0.95\\textwidth",eval=FALSE>>=
 mL3 <- subset(mL2,
@@
 @
+\begin{itemize}
+\item {\small $\beta_{P,D}$: early/late-stage transmission probability; $D_{P,D}$: early/late-stage duration; $c$: partnership dissolution rate; $c_e/c_w$: extra- vs within-pair contact ratio; $\rho$: partnership formation rate; $c_u/c_w$: uncoupled contact rate ratio}
+\end{itemize}
 \end{block}
@@
 \item Implicit models did worst
-\item Variation among models (model structure) $\approx$ variation within models (parameter uncertainty)
+\item Variation among models (model structure) $\gtrsim$ variation within models (parameter uncertainty)
 \item Large differences

### 2016-06-01 poster/figure/tradeoff.R

@@ -1,11 +1,14 @@
 source("fig_setup.R")
 
-dd <-with(pars,
+pars2 <- HIVpars.base
+pars2$alphaDist <- c(min=0,max=9,delta=0.25)
+pars2 <- expand(pars2)
+dd <-with(pars2,
      melt(data.frame(alpha=alpha,"transmission probability"=Beta2,
                      "duration (years)"=Duration2,check.names=FALSE),
           id.vars="alpha"))
 gg_tradeoff <- ggplot(dd,aes(alpha,value))+facet_wrap(~variable,scale="free")+
-    geom_line(size=8)+
+    geom_line(size=7)+
     labs(x= Lspvl,y="")
 
 tikz(file="tradeoff.tikz",width=12,height=8,standAlone=TRUE)

### 2016-06-01 poster/hivlhs_eeid2016.Rnw
fix size (R)

@@
 %% You can change the poster orientation (portrait vs. landscape) and the paper size (a0, a1, &c.) here:
-\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}
+
+\usepackage[orientation=portrait,size=custom,width=71,height=102,scale=1.2,debug]{beamerposter}
+%## size=a0,
 
@@
 \newlength{\columnheight}
-\setlength{\columnheight}{105cm}
+\setlength{\columnheight}{100cm}
 
 \newlength{\topcolumnheight}
-\setlength{\topcolumnheight}{50cm}
+\setlength{\topcolumnheight}{37cm}
 
@@
 
-\vspace{-2.4 cm} % shouldn't need this??
+\vspace{-1 cm} % shouldn't need this??
 \begin{block}{Summary}
@@
 
+%\newcommand{\topfudge}{-5cm}
+
 \begin{column}{.65\textwidth}
-
 \begin{columns}
@@
         \parbox[t][\topcolumnheight]{\textwidth}{ 
+          \vspace{-4.5cm} % shouldn't need this??
 
@@
   single individuals and extra-pair contact wash out effects of structure
-\item implicit model is \textbf{most different}
+%\item implicit model is \textbf{most different}
 \item random-mixing model underestimates variability
@@
 
+\vspace{-4.5cm} % shouldn't need this??
             \begin{block}{Univariate summaries}
@@
 \item equilibrium virulence: interaction
-\item low-equilibrium outcomes for intermediate-complexity models
+%\item low-equilibrium outcomes for intermediate-complexity models
 \end{itemize}
@@
   \begin{minipage}[T]{.95\textwidth} % tweaks the width, makes a new \textwidth
-    \parbox[t][30cm]{\textwidth}{ % must be some better way to set
+    \parbox[t][26cm]{\textwidth}{ % must be some better way to set
 

### 2016-06-11 R/SSQ.R
transmission/duration quantile

@@ -1,4 +1,6 @@
 library("lme4")
+source("hivFuns.R")
+source("Param.R")
 load("../simdata/combine_ev_LHS4.rda")
 sL <- sum_list[["sum_mat"]]
 
@@ -27,8 +29,6 @@ findSSQ <- function(var){
 findSSQ(eqV)
 findSSQ(peakV)
 
-##why is this different from the lmer results?
-
 ## do some calculations on the scales of transmission prob/duration as
 ## well as on the log10 SPVL scale, for example.
 ##
@@ -38,3 +38,49 @@ findSSQ(peakV)
 ##    ditto for the implicit model
 
 ##    ditto for duration
+
+
+sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
+sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
+
+returnBeta <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(K,Bmax,v,a))
+	})
+}
+
+returnDur <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(v,Dmax,D50,h))
+	})
+}
+
+sum_fun <- function(x) {
+	data.frame(
+		var = names(x),
+		mean=colMeans(x),
+		lwr=apply(x,2,quantile,0.025),
+		upr=apply(x,2,quantile,0.975))
+}
+
+sL.epi <- transform(sL, eq_t = returnBeta(eq_vir,HIVpars.skeleton),
+										peak_t = returnBeta(peak_vir,HIVpars.skeleton),
+										eq_dur = returnDur(eq_vir,HIVpars.skeleton),
+										peak_dur = returnDur(peak_vir,HIVpars.skeleton))
+sL.epi <- sL.epi[,-c(3:6)]
+
+sL.epi
+
+model_name <- c("pairform+epc", "pairform", "instswitch", "instswitch+epc", "implicit", "random")
+
+epiList <- list()
+
+for(i in 1:6){
+	epiList[[model_name[i]]] = sL.epi[c((1000*(i-1)+1):(1000*i)),3:6]
+}
+
+library("plyr")
+
+ldply(epiList, sum_fun)

### 2016-06-24 figures/figures.rmd
transmission/duration quantile
working on paper

@@
                                        base_family = "Times"))
+library("dplyr")
+library("tidyr")
 opts_chunk$set(echo=FALSE,error=FALSE)
@@
 
+```{r sumtab,as.is=TRUE}
+rq <- function(x,q,digits=3) {
+    round(quantile(x,q),digits)
+}
+mLtab <- mL.epi %>% group_by(model,variable) %>%
+    summarise(lwr=rq(value,0.025),
+              med=rq(value,0.5),
+              upr=rq(value,0.975)) %>%
+    mutate(variable=abbreviate(variable,30)) %>%
+    ungroup() %>%
+    arrange(variable,model)
+kable(mLtab)
+```
+```{r sumvals,echo=FALSE}
+ff <- function(model,variable,r=3) {
+    mm <- mLtab[mLtab$model==model & mLtab$variable==variable,3:5]
+    round(unlist(mm),r)
+}
+mindur_random <- ff("random","peak_dur",1)
+mindur_pfepc <-  ff("pairform+epc","peak_dur",1)
+mindur_implicit  <- ff("implicit","peak_dur",1)
+maxt_random <- ff("random","peak_t",2)
+maxt_pfepc <- ff("pairform+epc","peak_t",2)
+maxt_implicit  <- ff("implicit","peak_t",2)
+save(list=ls(pattern="^(maxt|mindur)"),file="../ms/maxminstats.RData")
+```
 ```{r fig4_func, echo = FALSE}

### 2016-06-24 ms/plos_latex.tex
added methods section back
more makefile/ms tweaks
working on paper

@@
 % Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
+\usepackage{setspace} 
+\doublespacing
 
@@
   our basic understanding of epidemiological dynamics and,
-  potentially, for the outcomes of public health interventions. In
-  general changes in fitness landscapes over the course of an epidemic
-  will select for higher virulence during the early,
-  exponential-growth phase of the epidemic, but quantitative outcomes
+  potentially, for the outcomes of public health interventions.
+  Changes in the fitness landscape
+  will generally select for higher virulence during the early
+  phase of an epidemic, but quantitative outcomes
   can depend sensitively on biological details and the structure of
@@
   set-point viral load, SPVL) and their heritability between
-  hosts. However, these models use very simple implicit
-  representations of the transmission process that ignore the
+  hosts. However, these models use implicit
+  representations of the transmission process that drastically simplify the
   partnership dynamics that previous research has found to be critical
-  in predicting epidemics of sexually transmitted diseases.  We
+  in driving epidemics of sexually transmitted diseases.  We
   explore models that combine HIV virulence tradeoffs with a range of
@@
   the simulated epidemic growth rate to be identical, matching a
-  realistic baseline value. Our primary result is that, for this
+  realistic baseline value. For this
   particular model setting, the simplest random-mixing structure is
@@
 
-Pathogens such as HIV can evolve rapidly in response to changing
-conditions; such changes include both the increases in disease
-incidence over the course of the epidemic and treatment
+Pathogens such as HIV can evolve rapidly in response to changes
+in their environments; such changes include both increases in disease
+prevalence over the course of the epidemic and treatment
 interventions. While researchers have successfully used computational
 models to explore these evolutionary dynamics, these models often
-neglect epidemiological processes, such as the formation and
+neglect details such as the formation and
 dissolution of sexual partnerships; other research has shown that
 these processes can strongly affect epidemic outcomes. We built and
-compared models with a range of realism and complexity in the way they
-modeled both partnership dynamics and the possibility of sexual
-contact outside of partnerships. In contrast to models of intermediate
-complexity, simple random-mixing models with no partnership structure
-at all actually gave the closest match to the most realistic and
-complex models; extra-pair contact tended to wash out the effects
-of epidemiological structure. Nevertheless, the large differences
+compared models that used different methods to
+model both partnership dynamics and sexual
+contact outside of partnerships. Models of intermediate
+complexity predicted much lower peak virulence (15 years to
+progress to AIDS) compared to both more realistic models
+and simple random-mixing models with no partnership structure
+at all (both approx. 7.25 years to progress to AIDS); 
+extra-pair contact tended to wash out the effects
+of epidemiological structure. The large differences
 in evolutionary dynamics among different epidemiological models
@@
 Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
-the tradeoff theory: in a study of discordant couples (i.e. long-term
+the tradeoff theory: in studies of discordant couples (i.e. long-term
 sexual partnerships with one infected and one uninfected partner), HIV
@@
 stage of infection), which in turn predicted the probability of
-within-couple transmission
+transmission
 \cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
@@
 that HIV invading a novel population would initially evolve increased
-virulence, peaking after approximately [XXX] years and then declining
+virulence, peaking after approximately 100-200 years and then declining
 slightly to a long-stable virulence level.
@@
 evolutionary factors \cite{day_virulence_2004,alizon_price_2009}.
-However, despite the attention to mechanistic detail at the individual
+However, despite these studies' attention to detail at the individual
 or physiological level, the epidemiological structures used in these
@@
 We add complexity to the epidemiological model following the general approach of Champredon \etal\ \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships.
-At the same time, our analysis somewhat simplifies the models of Shirreff \etal\ \cite{shirreff_transmission_2011}, for computational tractability; we check that our qualitative results are not sensitive to these simplifications.
 In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity.
@@
 
+\section*{Materials and Methods}
+
 \subsection*{Infection dynamics}
 
-Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than virulence (i.e.
-rate of progression to AIDS) itself.
+Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than th rate of progression to AIDS itself
+(we refer to SPVL as ``virulence'' hereafter).
 In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period.
@@
 \end{equation}
-where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the \textbf{P}rimary and \textbf{D}isease stages
+where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
 of infection are independent of the host's SPVL.
-Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the \textbf{A}symptomatic stage are Hill functions of the SPVL:
+Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:
 
@@
 where $V_{\alpha} = 10^\alpha$. 
-The \textbf{u}ncoupled and \textbf{e}xtra-couple transmission rates are scaled by multiplying the \textbf{w}ithin-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$.
+The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$.
 
@@
 
-Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \etal's evolutionary model slightly by using a one-to-one genotype-phenotype mapping.
+Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping.
 The mutational process in our model is directly taken from Shirreff \etal.
 Over the course of infection, mutation occurs within the host.
-However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection for simplicity (and is not further affected by within-host mutation).
+However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection (and is not further affected by within-host mutation).
 Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual.
@@
 We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (Fig.~\ref{fig:panel3}) we use $n=51$ (i.e.
-a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width 0.25 \Lspvl), which we use for all other simulations.
+a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width = 0.25 \Lspvl), which we use for all other simulations.
 
@@
 
-We developed six multi-strain evolutionary models, designed to cover a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simplistic \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+We developed six multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
 Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
@@
 Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}.
-Models 1 and 2 assume non-instantaneous partnership formation (i.e.
+The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e.
 individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status.
 $S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals.
-$SS$ is the number of susceptible-susceptible couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples.
-Model 1 includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas model 2 only considers within-couple transmission.
-Models 3 and 4 assume instantaneous partnership formation and thus consist of only the three partnered states: $SS$, $SI$, and $II$.
-Parallel to model 1 and 2, model 3 includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and model 4 only considers within-couple transmission.
-
-In contrast, models 5 and 6 are not explicitly structured.
-Model 5 is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
+$SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples.
+The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission.
+The next two models assume instantaneous partnership formation (``instswitch'') and thus consist of only the three partnered states: $SS$, $SI$, and $II$.
+Like the first two models, these models differ in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and the fourth (``instswitch'') only considers within-couple transmission.
+
+The last two models do not explicitly track sexual partnerships.
+One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
 It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
-However, to simulate the effect of instantaneous partnership formation, it uses an adjusted transmission rate that is derived from approximated basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}.
-Finally, model 6 is a simple random-mixing model.
+However, to reflect the effect of (instantaneously formed) partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}.
+Finally, the last model (``random'') is a simple random-mixing model.
 
 The base model (i.e.
-model 1) for the first four models is an extension of Champredon \etal's model \cite{champredon_hiv_2013}.
-Individuals in single compartment acquire a partner at a rate, $\rho$, and partnerships dissolve at a rate, $c$.
-Infected individuals in a discordant partnership infect susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
+pairform+epc) for the first four models is an extension of Champredon \etal's model \cite{champredon_hiv_2013}.
+Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$.
+Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
 Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
-Extra-couple and uncoupled transmission are modeled in a same way as Champredon \etal's model.
+Extra-couple and uncoupled transmission are modeled in the same way as Champredon \etal's model.
 All the details have been adapted to a multi-strain scenario.
-Model 2, 3, and 4 are derived from the base model by removing epidemiological details (partnership formation and uncoupled/extra-couple contact).
+The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact).
 Model details are explained in the appendix.
@@
 
-Despite considerable effort by many researchers \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
-In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
+Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
+In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
 We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
@@
 Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
-
 \begin{equation}
@@
 \end{equation}
-For simplicity, we assume that all parameters are uniformly distributed on the log scale.
-Following the vectorization of a parameter range, a matrix is constructed so that each column contains a vector of a parameter series which it represents ($z_1$,...$z_N$).
-Then, each column is replaced with a random permutation series of itself.
-Now, each row contains a different parameter set that is used for each simulation run.
+Random permutations of these vectors form columns in a sample parameter matrix; each row contains a different parameter set that is used for one simulation run.
 
@@
 Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples.
-Starred (*) parameters (used in Figure~\ref{fig:panel3}) are from \cite{shirreff_transmission_2011}.}
+Starred (*) parameters (used in Figure~\ref{fig:panel3}), and Descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
 \centering
@@
 Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$.
-Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- as there is not much known about the uncoupled transmission rate, we consider the wider range appropriate.
+Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is not much known about the uncoupled transmission rate.
 
@@
 
-One of the most difficult parts of model comparison is finding parameter sets that are commensurate against many different model structures.
-For the most part, our models are too complex to easily derive analytical correspondences among them.
-Given a numerical criterion, such as $r$ (initial exponential growth rate) or ${\cal R}_0$ (intrinsic reproductive number), we can adjust one or more parameters by brute force to ensure that all of the models match according to that criterion.
-While ${\cal R}_0$ is often considered the most fundamental property of an epidemic, and might thus seem to be a natural goal, here we focus on matching the initial growth rate $r$ for several reasons.
-First, our primary interest is in the transient evolutionary dynamics of virulence, which are more strongly affected by $r$ than ${\cal R}_0$.
-Second, $r$ is in general more directly observable in real epidemics; $r$ can be estimated by simply fitting an exponential curve to the initial incidence or prevalence curves \cite{ma_estimating_2014}, while ${\cal R}_0$ typically requires either (1) knowledge of \emph{all} epidemic parameters or (2) relatively sophisticated back-calculation based on $r$ and knowledge of the serial interval or generation interval of the disease \cite{wallinga_how_2007}.
-Thus, we scale a parameter so that every run has the same initial exponential growth rate in the incidence.
-
-In order to allow for all models to have equal initial exponential growth rate, $r$, we need to pick a parameter, $s$, so that $\lim_{s\to 0} r(s) = 0$ and $\lim_{s\to\infty}  r(s) = \infty$.
-As adjusting either partnership change rate (i.e.
-partnership formation and dissolution rate) or transmission rate does not fulfill this requirement for certain models, we decided to scale partnership change rate and dissolution rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =  \gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$, $\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$.
-Since transmission rate is adjusted by the scale of $\gamma$, uncoupled and extra-couple transmission rates are adjusted as well.
-For models 3, 4, and 5, all of which assume instantaneous partnership (and hence do not track single individuals), only the transmission rate and partnership dissolution rate (in this case equivalent to the partnership change rate) are adjusted.
+One of the most difficult parts of model comparison is finding
+parameter sets that are commensurate with many different model
+structures. For the most part, our models are too complex to easily
+derive analytical correspondences among them. Given a numerical
+criterion, such as $r$ (initial exponential growth rate) or ${\cal
+  R}_0$ (intrinsic reproductive number), we can adjust one or more
+parameters by brute force to ensure that all of the models match
+according to that criterion. While ${\cal R}_0$ is often considered
+the most fundamental property of an epidemic, and might thus seem to
+be a natural matching criterion, here we focus on matching the initial
+growth rate $r$ for several reasons. First, our primary interest is in
+the transient evolutionary dynamics of virulence, which are more
+strongly affected by $r$ than ${\cal R}_0$. Second, $r$ is 
+more directly observable in real epidemics; $r$ can be estimated by
+fitting an exponential curve to the initial incidence or
+prevalence curves \cite{ma_estimating_2014}, while ${\cal R}_0$
+typically requires either (1) knowledge of \emph{all} epidemic
+parameters or (2) calculations based on
+$r$ and knowledge of the serial interval or generation interval of the
+disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
+every run has the same initial exponential growth rate of incidence.
+
+In order to allow for all models to have equal initial exponential
+growth rate, $r$, we need to pick a parameter, $s$, such that
+$\lim_{s\to 0} r(s) = 0$ and $\lim_{s\to\infty} r(s) = \infty$. As
+adjusting either partnership change rate (i.e. partnership formation
+and dissolution rate) or transmission rate fails this requirement for
+some of our models, we scaled partnership change rate and
+dissolution rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =
+\gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$,
+$\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$. Since transmission rate
+is adjusted by the scale of $\gamma$, uncoupled and extra-couple
+transmission rates are adjusted as well. For the instantaneous-switching
+and implicit models, none of which track single individuals, 
+only the transmission rate and partnership
+dissolution rate (in this case equivalent to the partnership change
+rate) are adjusted.
 
@@
 After each run, initial exponential growth rate is calculated.
-Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that of Shirreff \etal's model.
-
-In order to make the computation faster, we simplify the model simulations to calculate $\gamma$ for each run.
-While scaling the parameters to match the exponential growth rate to a target value, models are ran for only 500 years, unlike the real simulations (ran for 4000 years) which are run after the parameters are scaled to obtain the summary statistics.
-Furthermore, we use Runge-Kutta method ode45 to find $\gamma$ whereas lsoda is used for the actual simulations.
+Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that implied by Shirreff \etal's original models.
+
+The calibration runs for each parameter set are slightly
+simplified. We run each model for only 500 years
+(full simulations are run for 4000 years), and we use a 4/5 order 
+Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
+\cite{soetaert_solving_2010}), whereas we use
+a stiff solver (LSODA) for the full simulations.
 
@@
 $\approx 4.6$ \Lspvl)
-and equilibrium virulence $\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
+and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
 
@@
 \section*{Acknowledgments}
-NSERC; C. Fraser
+We would like to thank Christophe Fraser for access to simulation code;
+this work was funded by an NSERC Discovery Grant.
 

### 2016-06-24 R/SSQ.R

@@ -1,5 +1,8 @@
 library("lme4")
 load("../simdata/combine_ev_LHS4.rda")
+source("hivFuns.R")
+source("Param.R")
+
 sL <- sum_list[["sum_mat"]]
 
 model = sL$model
@@ -7,6 +10,7 @@ eqV = sL$eq_vir
 peakV = sL$peak_vir
 
 m_peakV <- lmer(peakV ~ 1 + (1|model), data = sL, REML=FALSE)
+
 v_among <- c(VarCorr(m_peakV)[[1]])  ## c() dumps extra information
 v_within <- sigma(m_peakV)^2
 v_among/(v_among+v_within)

### 2016-06-24 ms/plos_latex.tex
more paper

@@
 \ref{fig:unidist}.
-We first consider the models of intermediate realism (implicit,
-instantaneous-switching with and without extra-pair contact (``epc'' hereafter), and
-pair formation without extra-pair contact). Some parameter
+We first consider the models of intermediate realism: implicit,
+instantaneous-switching with and without extra-pair contact, and
+pair formation without extra-pair contact. Some parameter
 sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL);
@@
 Fig.~\ref{fig:unidist}, (and the bivariate comparisons in
-Fig.~\ref{fig:pairplot}) is the match between the output of the
+Fig.~\ref{fig:pairplot}) is the similarity between the results of the
 least (random mixing) and the most complex (pair formation with
@@
 dynamics of these two extreme models are remarkably similar.
-This phenomenon is driven by the strength of extra-pair contact in the
+This phenomenon is driven by the strong effects of extra-pair contact in the
 model with explicit pair formation and extra-pair contact 
@@
 
+Expressing these results in terms of more directly interpretable epidemic
+parameters, i.e. using the Hill functions to translate \Lspvl\ to
+within-couple transmission probabilities and average time of progression
+to AIDS, shows that these differences are practically as well as 
+scientifically important. The random-mixing and pairform+epc models
+predict minimimum times to progression (at the virulence peak) of
+5.7 (95\% CI 6.1-6.3) and 6.0 (5.0-7.7) years respectively, while
+the implicit model gives progression times about twice as long:
+12.5 (9.5-15.6) years. The corresponding differences in 
+within-couple transmission
+probability are even more extreme, about a fourfold difference:
+0.249 (0.24-0.26) and 0.252 (0.19-0.28) per year for the 
+random and pairform+epc models vs. 0.059 (0.02-0.13) per year
+for the implicit model.
+
 \begin{figure}[!h]
@@
 Plotting the bivariate result distributions (Fig.~\ref{fig:pairplot})
-shows that most of the relationships among the summary statistics
-are monotonic, except those involving the relative peak virulence
+shows that most of the summary statistics
+are monotonically related, except those involving the relative peak virulence
 (bottom row). Changes in parameters that increase the equilibrium
@@
 The bivariate relationships also help distinguish the results of 
-different models with similar univariate distributions. While the
+different models with similar univariate dynamical summaries. While the
 relationship between equilibrium virulence and peak time is
@@
 the plot shows the effects of the \emph{unscaled} parameters, i.e.
-before they have been adjusted to achieve a standard initial epidemic
+before they have been calibrated to achieve a standard initial epidemic
 growth rate.
@@
 decrease the equilibrium virulence, peak virulence, and peak time,
-although the random and pair-formation+extra-pair contact
+although the random and pair-formation+epc 
 models have high, relatively
@@
 
-The base contact rate increases virulence and peak time in almost all
-cases, although the pair-formation+extra-pair contact
+The partnership dissolution rate ($c$), which essentially
+acts as a contact rate in the model,
+increases virulence and peak time in almost all
+cases, although the pair-formation+epc
 model is again relatively insensitive.
@@
 simplified model gives adequate answers, or whether the
-simplifications have led to qualitatively or quantitatively incorrect
-conclusions.  This question is especially important for modelers who
+simplifications lead to qualitative or quantitative errors.
+This question is especially important for modelers who
 are hoping that their conclusions will guide management decisions.

### 2016-06-24 ms/plos_latex.tex
more paper ...

@@
 of epidemiological complexity entirely and used a naive random-mixing
-contact model. Herbeck \etal\ \cite{herbeck_hiv_2014} build a
-network model of partnerships, but they set the ``partnership''
-duration to 1 day --- very unrealistic in epidemic terms, but perhaps
+contact model. In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
 actually more true to real-world HIV epidemiological dynamics than a
 model with realistic partnership durations that neglects extra-pair
-contact \cite{herbeck2016evolution}. Adding still further realism
-might push the results farther away from the random-mixing model. For
+contact \cite{herbeck2016evolution}. Making the model even more realistic
+might make the random-mixing model less appropriate. For
 example, our model forms partnerships randomly, and assumes that
 extra-pair contact is randomly mixing across the population;
-one could instead implement extra-pair contact as arising from
+one could instead model extra-pair contact as arising from
 multiple concurrent partnerships (some, such as contact with sex
-workers, of very short duration), or more structured partnership
+workers, of very short duration) and/or more structured partnership
 formation (by age, ethnicity, or behaviour group). The effects of
@@
 growth rate, chosen to match the results of previous models
-\cite{shirreff_transmission_2011}.  The previous models drew their
-parameters from previous cohort studies from the 1990s
+\cite{shirreff_transmission_2011}.  Previous models 
+in this area have drawn their
+parameters from cohort studies from the 1990s
 \cite{wawer2005rates,hollingsworth_hiv1_2008}
@@
 and calibration from population-level observations of epidemic
-trajectories. For practical (rather than exploratory) purposes,
-such a procedure would also be an improvement on the common --- although not universal --- %
-practice of assessing uncertainty over uniform ranges, rather than
-distributions that allow more continuous variation in support over
+trajectories. Such a procedure would also be an improvement on the common --- although not universal --- %
+practice, which we have followed here,
+of assessing uncertainty over uniform ranges rather than
+using distributions that allow more continuous variation in support over
 the range of a parameter.

### 2016-06-25 figures/figures.rmd
figures

@@
 g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
-    geom_line()+labs(x=tlab,y=expression(mean~virulence~(log[10]~SPVL)))+
+    geom_line()+labs(x=tlab,y=expression(population~mean~set-point~viral~load~(log[10])))+
     theme(legend.position=c(0.75,0.25))
@@
 	geom_ribbon(aes(fill = model), alpha=0.3)+
-	labs(x="time (years)",y=expression(mean~virulence~(log[10]~SPVL))) +
+	labs(x="time (years)",y=expression(population~mean~set-point~viral~load~(log[10]))) +
 	theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
-		axis.title.x = element_text(margin = margin(10,0,0,0))) +
+		axis.title.x = element_text(margin = margin(10,0,0,0)),
+		legend.position = "none") +
 	facet_wrap(~model, nrow = 2) +
@@
 	geom_ribbon(aes(fill = model), alpha=0.3)+
-	labs(x="time (years)",y=expression(mean~transmission~rate~during~asymptomatic~stage)) +
+	labs(x="time (years)",y=expression(mean~transmission~rate~during~asymptomatic~stage~(year^-1))) +
 	theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
-		axis.title.x = element_text(margin = margin(10,0,0,0))) +
+		axis.title.x = element_text(margin = margin(10,0,0,0)),
+		legend.position = "none") +
 	facet_wrap(~model, nrow = 2) +
@@
 	geom_ribbon(aes(fill = model), alpha=0.3)+
-	labs(x="time (years)",y=expression(mean~duration~of~asymptomatic~stage)) +
+	labs(x="time (years)",y=expression(mean~duration~of~asymptomatic~stage~(years))) +
 	theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
-		axis.title.x = element_text(margin = margin(10,0,0,0))) +
+		axis.title.x = element_text(margin = margin(10,0,0,0)),
+		legend.position = "none") +
 	facet_wrap(~model, nrow = 2) +
@@
 	geom_violinh(width=1)+
+	theme(legend.position = "none") +
 	facet_wrap(~variable,scale="free_x")+
@@
 
-sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
 sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
@@
 
-```{r sumtab,as.is=TRUE}
+```{r sumtab,as.is=TRUE,eval=FALSE}
 rq <- function(x,q,digits=3) {
@@
 ```
-```{r sumvals,echo=FALSE}
+
+```{r sumvals,echo=FALSE,eval=FALSE}
 ff <- function(model,variable,r=3) {
@@
 ```
+
 ```{r fig4_func, echo = FALSE}

### 2016-06-25 ms/plos_latex.tex
cover letter, tweak bib

@@
 Sang Woo Park\textsuperscript{1}
-Benjamin M. Bolker\textsuperscript{2,3,4,*}
+Benjamin M. Bolker\textsuperscript{1,2,3,*}
 \\
 \bigskip
-\textbf{1} Program in Health Sciences, McMaster University, Hamilton, Ontario, Canada
+\textbf{1} Department of Mathematics \& Statistics,  McMaster University, Hamilton, Ontario, Canada
 \\
-\textbf{2} Department of Mathematics \& Statistics,  McMaster University, Hamilton, Ontario, Canada
+\textbf{2} Department of Biology,  McMaster University, Hamilton, Ontario, Canada
 \\
-\textbf{3} Department of Biology,  McMaster University, Hamilton, Ontario, Canada
-\\
-\textbf{4} Institute for Infectious Disease Research,  McMaster University, Hamilton, Ontario, Canada
+\textbf{3} Institute for Infectious Disease Research,  McMaster University, Hamilton, Ontario, Canada
 \\
@@
 
-Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than th rate of progression to AIDS itself
+Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than the rate of progression to AIDS itself
 (we refer to SPVL as ``virulence'' hereafter).
@@
 \end{equation}
-We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (Fig.~\ref{fig:panel3}) we use $n=51$ (i.e.
+We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e.
 a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width = 0.25 \Lspvl), which we use for all other simulations.
@@
 Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples.
-Starred (*) parameters (used in Figure~\ref{fig:panel3}), and Descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
+Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and Descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
 \centering
@@
 Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Fig.~\ref{fig:panel3}a).
+As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
 However, our model produces lower peak virulence ($\approx 4.3$ vs.
@@
 Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases.
-Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (Fig.~\ref{fig:panel3}b).
-Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Fig.~\ref{fig:panel3}c).
-
-Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (Fig.~\ref{fig:virtraj}).
+Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
+Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).
+
+Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (\figurename~\ref{fig:virtraj}).
 Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 \Lspvl), virulence peaks in all models between 200 and 300 years.
@@
 but high relative peaks ($>1.3$)
-(Fig.~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
+(\figurename~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
 At the opposite extreme, parameter sets that produce high equilibrium virulence 
@@
 The most striking aspect of the univariate comparisons in
-Fig.~\ref{fig:unidist}, (and the bivariate comparisons in
-Fig.~\ref{fig:pairplot}) is the similarity between the results of the
+\figurename~\ref{fig:unidist}, (and the bivariate comparisons in
+\figurename~\ref{fig:pairplot}) is the similarity between the results of the
 least (random mixing) and the most complex (pair formation with
@@
 model with explicit pair formation and extra-pair contact 
-(``pairform+epc'' in Figures~\ref{fig:virtraj}-\ref{fig:plot_sens}). When individuals spend time uncoupled between
+(``pairform+epc'' in \figurename{}s~\ref{fig:virtraj}-\ref{fig:plot_sens}). When individuals spend time uncoupled between
 partnerships, and when these single individuals can transmit disease
@@
 scientifically important. The random-mixing and pairform+epc models
-predict minimimum times to progression (at the virulence peak) of
+predict minimum times to progression (at the virulence peak) of
 5.7 (95\% CI 6.1-6.3) and 6.0 (5.0-7.7) years respectively, while
@@
 random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model.
+for the implicit model (Figs~???? show univariate distributions
+on the epidemiological scales
+of progression time and transmission probability,
+for all summary measures and all models).
 
@@
 
-Plotting the bivariate result distributions (Fig.~\ref{fig:pairplot})
+Plotting the bivariate result distributions (\figurename~\ref{fig:pairplot})
 shows that most of the summary statistics
@@
 
-Finally, the sensitivity plots (Fig.~\ref{fig:plot_sens}) show the effects 
+Finally, the sensitivity plots (\figurename~\ref{fig:plot_sens}) show the effects 
 of each parameter on the summary statistics. In almost every case the
@@
 load are changing, on time scales comparable to those portrayed here
-(e.g., compare Fig.~\ref{fig:virtraj} to Herbeck \etal's
+(e.g., compare \figurename~\ref{fig:virtraj} to Herbeck \etal's
 estimated rate of change of 1.3 \Lspvl\ per century [95\% CI -0.1 to
@@
 therapy (ART) can be expected to change the trajectory of virulence
-evolution \cite{payne_impact_2014,herbeck2016evolution}.  While these
+evolution \cite{payne_impact_2014,roberts2015impact,herbeck2016evolution}.  While these
 efforts are well-intentioned, we caution that epidemiological and
@@
 
+\section*{Acknowledgements}
+We would like to thank Christophe Fraser for access to simulation code;
+this work was funded by an NSERC Discovery Grant.
+
 \section*{Supporting Information}
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Appendix.}
+\paragraph*{Appendix S1}
 \label{S1_Appendix}
@@
 \end{equation}
-Lastly, as model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution.
+Lastly, as model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution:
 
@@
 \end{equation}
-Model 6 also shares the same distribution of initial infected individuals as model 5.
-
-\section*{Acknowledgments}
-We would like to thank Christophe Fraser for access to simulation code;
-this work was funded by an NSERC Discovery Grant.
+Model 6 has the same distribution of initial infected individuals as model 5.
+
+\paragraph*{Appendix S2}
+\label{S2_Appendix}
 
@@
 
-\bibliography{virulence}
+% \bibliography{virulence}
+\begin{thebibliography}{10}
+
+\bibitem{Ebert1999}
+Ebert D.
+\newblock The evolution and expression of parasite virulence.
+\newblock In: Stearns SC, editor. Evolution in Health \& Disease. New York:
+  Oxford University Press, Oxford, UK; 1999. p. 161--172.
+
+\bibitem{EbertBull2003}
+Ebert D, Bull JJ.
+\newblock Challenging the trade-off model for the evolution of virulence: is
+  virulence management feasible?
+\newblock Trends Microbiol. 2003;11(1):15--20.
+
+\bibitem{alizon_adaptive_2015}
+Alizon S, Michalakis Y.
+\newblock Adaptive virulence evolution: the good old fitness-based approach.
+\newblock Trends in Ecology \& Evolution. 2015;30(5):248--254.
+\newblock doi:{10.1016/j.tree.2015.02.009}.
+
+\bibitem{Dwyer+1990}
+Dwyer G, Levin SA, Buttel L.
+\newblock A simulation model of the population dynamics and evolution of
+  myxomatosis.
+\newblock Ecol Monog. 1990;60:423--447.
+
+\bibitem{mackinnon1999genetic}
+Mackinnon MJ, Read AF.
+\newblock Genetic relationships between parasite virulence and transmission in
+  the rodent malaria {{\em Plasmodium chabaudi}}.
+\newblock Evolution. 1999; p. 689--703.
+
+\bibitem{jensen2006empirical}
+Jensen KH, Little T, Skorping A, Ebert D.
+\newblock Empirical support for optimal virulence in a castrating parasite.
+\newblock PLoS Biol. 2006;4(7):e197.
+
+\bibitem{deroode2008virulence}
+De~Roode JC, Yates AJ, Altizer S.
+\newblock Virulence-transmission trade-offs and population divergence in
+  virulence in a naturally occurring butterfly parasite.
+\newblock Proceedings of the National Academy of Sciences.
+  2008;105(21):7489--7494.
+
+\bibitem{Fraser+2007}
+Fraser C, Hollingsworth TD, Chapman R, de~Wolf F, Hanage WP.
+\newblock Variation in {HIV}-1 set-point viral load: Epidemiological analysis
+  and an evolutionary hypothesis.
+\newblock PNAS. 2007;104:17441--17446.
+
+\bibitem{fraser_virulence_2014}
+Fraser C, Lythgoe K, Leventhal GE, Shirreff G, Hollingsworth TD, Alizon S,
+  et~al.
+\newblock Virulence and Pathogenesis of {HIV}-1 Infection: An Evolutionary
+  Perspective.
+\newblock Science. 2014;343(6177):1243727.
+\newblock doi:{10.1126/science.1243727}.
+
+\bibitem{shirreff_transmission_2011}
+Shirreff G, Pellis L, Laeyendecker O, Fraser C.
+\newblock Transmission Selects for {HIV-1} Strains of Intermediate Virulence: A
+  Modelling Approach.
+\newblock PLoS Computational Biology. 2011;7(10):e1002185.
+\newblock doi:{10.1371/journal.pcbi.1002185}.
+
+\bibitem{herbeck_hiv_2014}
+Herbeck JT, Mittler JE, Gottlieb GS, Mullins JI.
+\newblock An {HIV} Epidemic Model Based on Viral Load Dynamics: Value in
+  Assessing Empirical Trends in {HIV} Virulence and Community Viral Load.
+\newblock {PLoS} Comput Biol. 2014;10(6):e1003673.
+
+\bibitem{day_virulence_2004}
+Day T, Proulx SR.
+\newblock A General Theory for the Evolutionary Dynamics of Virulence.
+\newblock The American Naturalist. 2004;163(4):E40--E63.
+\newblock doi:{10.1086/382548}.
+
+\bibitem{alizon_price_2009}
+Alizon S.
+\newblock The {Price} equation framework to study disease within-host
+  evolution.
+\newblock Journal of Evolutionary Biology. 2009;22(5):1123--1132.
+\newblock doi:{10.1111/j.1420-9101.2009.01726.x}.
+
+\bibitem{champredon_hiv_2013}
+Champredon D, Bellan S, Dushoff J.
+\newblock {HIV} Sexual Transmission Is Predominantly Driven by Single
+  Individuals Rather than Discordant Couples: A Model-Based Approach.
+\newblock PLoS ONE. 2013;8(12):e82906.
+\newblock doi:{10.1371/journal.pone.0082906}.
+
+\bibitem{hollingsworth_hiv1_2008}
+Hollingsworth TD, Anderson RM, Fraser C.
+\newblock {HIV}-1 Transmission, by Stage of Infection.
+\newblock Journal of Infectious Diseases. 2008;198(5):687--693.
+\newblock doi:{10.1086/590501}.
+
+\bibitem{blower_drugs_1991}
+Blower SM, Hartel D, Dowlatabadi H, Anderson RM, May RM.
+\newblock Drugs, Sex and {HIV}: A Mathematical Model for {New York City}.
+\newblock Philosophical Transactions of the Royal Society of London B:
+  Biological Sciences. 1991;331(1260):171--187.
+\newblock doi:{10.1098/rstb.1991.0006}.
+
+\bibitem{ma_estimating_2014}
+Ma J, Dushoff J, Bolker BM, Earn DJD.
+\newblock Estimating Initial Epidemic Growth Rates.
+\newblock Bulletin of Mathematical Biology. 2014;76(1):245--260.
+\newblock doi:{10.1007/s11538-013-9918-2}.
+
+\bibitem{wallinga_how_2007}
+Wallinga J, Lipsitch M.
+\newblock How generation intervals shape the relationship between growth rates
+  and reproductive numbers.
+\newblock Proceedings of the Royal Society of London B: Biological Sciences.
+  2007;274(1609):599--604.
+\newblock doi:{10.1098/rspb.2006.3754}.
+
+\bibitem{soetaert_solving_2010}
+Soetaert K, Petzoldt T, Setzer RW.
+\newblock Solving Differential Equations in {R}: Package {deSolve}.
+\newblock Journal of Statistical Software. 2010;33(9):1--25.
+\newblock doi:{10.18637/jss.v033.i09}.
+
+\bibitem{herbeck2016evolution}
+Herbeck J, Mittler J, Gottlieb G, Goodreau S, Murphy J, Cori A, et~al.
+\newblock Evolution of {HIV} virulence in response to widespread scale up of
+  antiretroviral therapy: a modeling study.
+\newblock bioRxiv. 2016; p. 039560.
+
+\bibitem{wawer2005rates}
+Wawer MJ, Gray RH, Sewankambo NK, Serwadda D, Li X, Laeyendecker O, et~al.
+\newblock Rates of {HIV}-1 transmission per coital act, by stage of {HIV}-1
+  infection, in {Rakai}, {Uganda}.
+\newblock Journal of Infectious Diseases. 2005;191(9):1403--1409.
+
+\bibitem{faria_early_2014}
+Faria NR, Rambaut A, Suchard MA, Baele G, Bedford T, Ward MJ, et~al.
+\newblock The early spread and epidemic ignition of {HIV}-1 in human
+  populations.
+\newblock Science (New York, NY). 2014;346(6205):56--61.
+\newblock doi:{10.1126/science.1256739}.
+
+\bibitem{elderd_uncertainty_2006}
+Elderd BD, Dukic VM, Dwyer G.
+\newblock Uncertainty in predictions of disease spread and public health
+  responses to bioterrorism and emerging diseases.
+\newblock Proceedings of the National Academy of Sciences. 2006;103(42):15693
+  --15697.
+\newblock doi:{10.1073/pnas.0600816103}.
+
+\bibitem{herbeck_is_2012}
+Herbeck JT, Müller V, Maust BS, Ledergerber B, Torti C, Di~Giambenedetto S,
+  et~al.
+\newblock Is the virulence of {HIV} changing? {A} meta-analysis of trends in
+  prognostic markers of {HIV} disease progression and transmission.
+\newblock AIDS (London, England). 2012;26(2):193--205.
+\newblock doi:{10.1097/QAD.0b013e32834db418}.
+
+\bibitem{payne_impact_2014}
+Payne R, Muenchhoff M, Mann J, Roberts HE, Matthews P, Adland E, et~al.
+\newblock Impact of {HLA}-driven {HIV} adaptation on virulence in populations
+  of high {HIV} seroprevalence.
+\newblock Proceedings of the National Academy of Sciences.
+  2014;111(50):E5393--E5400.
+\newblock doi:{10.1073/pnas.1413339111}.
+
+\bibitem{roberts2015impact}
+Roberts HE, Goulder PJ, McLean AR.
+\newblock The impact of antiretroviral therapy on population-level virulence
+  evolution of {HIV}-1.
+\newblock Journal of The Royal Society Interface. 2015;12(113):20150888.
+
+\end{thebibliography}
 

### 2016-06-26 ms/plos_latex.tex
mess with figure pipeline; figures allowed in ms. on submission

@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig1.pdf}
 \caption{{\bf Baseline dynamics.}
@@
 
-The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix); we need to pick a reasonable range for it.
+The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix S1); we need to pick a reasonable range for it.
 Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig2.pdf}
 \caption{{\bf Envelopes of virulence trajectories under all models.}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig3.pdf}
 \caption{{\bf Univariate distributions of summary statistics.}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig4.pdf}
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig5.pdf}
 \caption{{\bf Sensitivity plot.}
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{Appendix S1}
+\paragraph*{Appendix S1: model details}
 \label{S1_Appendix}
-{\bf Model details.} Since we use multi-strain models, in which the distribution of \Lspvl\ has been discretized into a vector, we use a matrix notation to describe our models.
-The five states described in the Methods section are replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$, where the subscripts denote the strain that an individual is infected with.
+
+Since we use multi-strain models in which the distribution of \Lspvl\ has been discretized into a vector, we use a matrix notation to describe our models.
+The five states described in the \emph{Methods} section are replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$, where the subscripts denote the strain with which an individual is infected.
 For example, $I_i$ is number of infected individuals with \Lspvl\ of $\alpha_i$, and $II_{ij}$ is the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$ (independent of order; $II_{ij}$ is synonymous with $II_{ji}$).
@@
 
-\subsection*{Models 1 and 2: partnership dynamics}
+\subsection*{Models 1 (``pairform+epc'') and 2 (``pairform'')}
+
+\subsubsection*{Partnership dynamics}
 
@@
 
-\subsection*{Models 1 and 2: infection}
+\subsubsection*{Pair-formation models: infection dynamics}
 
@@
 
-\subsection*{Model 1 and 2 - Disease induced mortality}
+\subsubsection*{Pair formation models: Disease induced mortality}
 
@@
 
-\subsection*{Models 3 and 4: Partnership dynamics}
+\subsection*{Models 3 (``instswitch'') and 4 (``instswitch'')}
+\subsubsection*{Partnership dynamics}
 
@@
 
-\subsection*{Model 3 and 4 - Infection}
-
-Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single state, only extra couple transmission exists:
+\subsubsection*{Instantaneous-switching models: Infection dynamics}
+
+Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single (uncoupled) state, only extra-couple transmission exists:
 
@@
 
-\subsection*{Model 3 and 4 - Disease induced mortality}
+\subsubsection*{Instantaneous-switching models: Disease induced mortality}
 
@@
 
-\subsection*{Model 5}
+\subsection*{Implicit model}
 
@@
 
-\subsection*{Model 6}
+\subsection*{Random-mixing model}
 
@@
 
-\paragraph*{Appendix S2}
+\paragraph*{Appendix S2: dynamics of transmission and virulence}
+
 \label{S2_Appendix}
+
+This section presents
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig_S2_1.pdf}
+\caption{{\bf Envelopes of transmission trajectories under all models.}
+This figure matches \figurename~\ref{fig:virtraj}, but displays the
+envelope of population-mean transmission probabilities rather than \Lspvl\ over time
+for each model.
+}
+\label{fig:transtraj}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig_S2_2.pdf}
+\caption{{\bf Envelopes of progression trajectories under all models.}
+This figure matches \figurename~\ref{fig:virtraj}, but displays the
+envelope of population-mean expected time of progression to AIDS (i.e., length of
+intermediate HIV phase) rather than \Lspvl\ over time
+for each model.
+}
+\label{fig:durtraj}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig_S2_3.pdf}
+\caption{{\bf Univariate distributions of transmission probabilities and progression.}
+This figure matches \figurename~\ref{fig:unidist}, but displays the
+univariate distributions for the transmission probability and 
+progression time at the virulence
+peak and at the epidemiological equilibrium,
+rather than the distributions of \Lspvl.
+}
+\label{fig:tranprogsum}
+\end{figure}
 

### 2016-09-07 R/Param.R
update R code

@@ -17,7 +17,10 @@ HIVpars.skeleton <- list(
   Vm = 0.12,                            ##Mutational standard deviation of log10SPVL (Shirreff)
   inisd = 0.2,                          ##Standard deviation of the initial SPVL distribution
   Ini_I = 1e-4,                         ##Initial number of infected individuals
-  ini_V = 3.5                           ##Initial mean SPVL
+  ini_V = 3.5,                          ##Initial mean SPVL
+  n.risk = 3,                           ##number of risk groups
+  r.risk = c(0.9,1,1.1),                ##risk group mixing ratio
+  p.risk = c(0.3, 0.4, 0.3)             ##risk group proportion
 )
 
 

### 2016-09-07 R/hivModels.R

@@ -146,7 +146,6 @@ gfun <- function(parameters, experimental=FALSE) {
   if (experimental) return(gX) else return(g)
 }
 
-gfunX <- 
 #serial
 gfun2 <- function(parameters) {
   pp <- expand(parameters)

### 2016-09-07 R/simulations/ev_LHS16.R

@@ -1,0 +1,45 @@
+library("deSolve")
+library("reshape2")
+library("ggplot2"); theme_set(theme_bw())
+source("../simFuns.R")  ## for transform.list
+source("../hivFuns.R")
+source("../Param.R")
+source("../hivModels.R")
+
+HIVpars.skeleton <- transform(HIVpars.skeleton, ini_V = 3)
+
+I_mat16 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+vir_mat16 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+eq_vec16 <- rep(0, n.trial)
+peak_mat16 <- matrix(NA, nrow = n.trial, ncol = 2)
+val_vec16 <- rep(0, n.trial)
+
+if (file.exists("ev_LHS_res16.rda")) {
+  load("ev_LHS_res16.rda")
+} else {
+  for (i in 1:n.trial) {
+    cat(i)
+    HIVpars <- as.HIVvirparlist(ltab[i,])
+    yini <- calc_yini(HIVpars)
+    little_r <- get_rval(gfun2(HIVpars),yini, HIVpars)
+
+    if(little_r < 0.04){
+      
+      interval = c(1,10)
+    }else{
+      interval = c(0.01,1)
+    }
+    val_vec16[i] = find_scale(0.04, gfun2, yini, HIVpars, interval=interval, adjpar = "scale_all")
+    HIVpars_adj <- transform(HIVpars, scale_all = val_vec16[i])
+    r <- lsoda(unlist(yini), func = gfun2(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+
+    I_mat16[,i] = r[,(ncol(r) - 1)]      
+    vir_mat16[,i] = r[,ncol(r)]
+    eq_vec16[i] = vir_mat16[length(tvec),i]
+    
+    peak_mat16[i,1] = which.max(vir_mat16[,i])
+    peak_mat16[i,2] = max(vir_mat16[,i])
+    
+  }
+  save("I_mat16","vir_mat16", "eq_vec16","peak_mat16", "val_vec16",file="ev_LHS_res16.rda")
+}

### 2016-09-07 R/simulations/ev_LHS17.R

@@ -1,0 +1,45 @@
+library("deSolve")
+library("reshape2")
+library("ggplot2"); theme_set(theme_bw())
+source("../simFuns.R")  ## for transform.list
+source("../hivFuns.R")
+source("../Param.R")
+source("../hivModels.R")
+
+HIVpars.skeleton <- transform(HIVpars.skeleton, ini_V = 3)
+
+I_mat17 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+vir_mat17 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+eq_vec17 <- rep(0, n.trial)
+peak_mat17 <- matrix(NA, nrow = n.trial, ncol = 2)
+val_vec17 <- rep(0, n.trial)
+
+if (file.exists("ev_LHS_res17.rda")) {
+  load("ev_LHS_res17.rda")
+} else {
+  for (i in 1:n.trial) {
+    cat(i)
+    HIVpars <- as.HIVvirparlist(ltab[i,])
+    yini <- calc_yini2(HIVpars)
+    little_r <- get_rval(gfun3(HIVpars),yini, HIVpars)
+    
+    if(little_r < 0.04){
+      
+      interval = c(1,10)
+    }else{
+      interval = c(0.01,1)
+    }
+    val_vec17[i] = find_scale(0.04, gfun3, yini, HIVpars, interval=interval, adjpar = "scale_all")
+    HIVpars_adj <- transform(HIVpars, scale_all = val_vec17[i])
+    r <- lsoda(unlist(yini), func = gfun3(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    
+    I_mat17[,i] = r[,(ncol(r) - 1)]
+    vir_mat17[,i] = r[,ncol(r)]
+    eq_vec17[i] = vir_mat17[length(tvec),i]
+    
+    peak_mat17[i,1] = which.max(vir_mat17[,i])
+    peak_mat17[i,2] = max(vir_mat17[,i])
+    
+  }
+  save("I_mat17","vir_mat17", "eq_vec17","peak_mat17", "val_vec17",file="ev_LHS_res17.rda")
+}

### 2016-09-07 R/simulations/ev_LHS_full.R

@@ -1,0 +1,45 @@
+library("deSolve")
+library("reshape2")
+library("ggplot2"); theme_set(theme_bw())
+source("../simFuns.R")  ## for transform.list
+source("../hivFuns.R")
+source("../Param.R")
+source("../hivModels.R")
+
+HIVpars.skeleton <- transform(HIVpars.skeleton, ini_V = 3)
+
+I_mat15 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+vir_mat15 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+eq_vec15 <- rep(0, n.trial)
+peak_mat15 <- matrix(NA, nrow = n.trial, ncol = 2)
+val_vec15 <- rep(0, n.trial)
+
+if (file.exists("ev_LHS_res15.rda")) {
+  load("ev_LHS_res15.rda")
+} else {
+  for (i in 1:n.trial) {
+    cat(i)
+    HIVpars <- as.HIVvirparlist(ltab[i,])
+    yini <- calc_yini(HIVpars)
+    little_r <- get_rval(gfun(HIVpars),yini, HIVpars)
+    
+    if(little_r < 0.04){
+      
+      interval = c(1,10)
+    }else{
+      interval = c(0.01,1)
+    }
+    val_vec15[i] = find_scale(0.04, gfun, yini, HIVpars, interval=interval, adjpar = "scale_all")
+    HIVpars_adj <- transform(HIVpars, scale_all = val_vec15[i])
+    r <- lsoda(unlist(yini), func = gfun(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    
+    I_mat15[,i] = r[,(ncol(r) - 1)]
+    vir_mat15[,i] = r[,ncol(r)]
+    eq_vec15[i] = vir_mat15[length(tvec),i]
+    
+    peak_mat15[i,1] = which.max(vir_mat15[,i])
+    peak_mat15[i,2] = max(vir_mat15[,i])
+    
+  }
+  save("I_mat15", "vir_mat15", "eq_vec15","peak_mat15", "val_vec15",file="ev_LHS_res15.rda")
+}

### 2016-09-07 R/simulations/figure1.R

@@ -1,0 +1,139 @@
+library("deSolve")
+library("reshape2")
+library("ggplot2"); theme_set(theme_classic())
+source("../simFuns.R")  ## for transform.list
+source("../hivFuns.R")
+source("../Param.R")
+source("../hivModels.R")
+library("gridExtra")
+
+get_rval <- function(g, yini, pp, plot.it=FALSE,
+                     tvec = c(1:500),
+                     lims=c(0.01,0.05),
+                     verbose=FALSE) {
+  
+    start <- unlist(yini)
+    
+    r <- rk(y=start,
+                 times=tvec,
+                 func=g,
+                 parms=pp, method = "ode45")
+        
+    Itot <- r[,(ncol(r)-1)]
+    
+    if(max(Itot) < 5e-2){
+      return(0)
+    }else{
+      
+      dd <- data.frame(tvec=tvec,Itot=Itot)
+      
+      dsub <- subset(dd, Itot>lims[1] & Itot<lims[2])
+      mm <- try(lm(log(Itot)~tvec,data=dsub))
+      if (plot.it) {
+        plot(log(Itot)~tvec,data=dsub)
+        abline(mm,col=2)
+      }
+      cc <- coef(mm)[2]
+      return(cc)
+    }
+}
+
+HIVpars.shirreff <- transform(HIVpars.shirreff, ini_V = 3)
+
+tvec = seq(from =1, to =600, by = 0.1)
+
+I_matS <- matrix(NA, nrow = length(tvec), ncol = 3)
+vir_matS <- matrix(NA, nrow = length(tvec), ncol = 3)
+val_vecS <- rep(NA, 3)
+
+ini_r <- c(0.021, 0.042, 0.084)
+
+if (file.exists("ev_LHS_resS.rda")) {
+  load("ev_LHS_resS.rda")
+} else {
+  for(i in 1:3){
+    cat(i)
+    HIVpars <- transform(HIVpars.shirreff, Ini_I = 0.001)
+    yini <- calc_yini3(HIVpars)
+    
+    little_r <- get_rval(gfun5(HIVpars),yini, HIVpars)
+    
+    if(little_r < ini_r[i]){
+      interval = c(1,2)
+    }else{
+      interval = c(0.01,1)
+    }
+    
[…]

### 2016-09-07 ms/plos_latex.tex

@@
+% ## http://journals.plos.org/ploscompbiol/s/submit-now
 % Template for PLoS
@@
 \section*{Acknowledgements}
-We would like to thank Christophe Fraser for access to simulation code;
-this work was funded by an NSERC Discovery Grant.
+We would like to thank Christophe Fraser and
+David Champredon for access to simulation code;
+this work was funded by NSERC Discovery Grant 386590-2010.
 

### 2016-09-12 misc/fullModel.rmd
add full model with risk het stuff

---
title: "Attempting to write a full model"
output: html_document
---

I want to write a full model. Let's start by considering partnership dynamics:

$$
\begin{aligned}
N_i &= - \rho_i N_i + \sum_k 2^{\delta_{ik}} NN_{ik} \frac{c_i + c_k}{2}\\
NN_{ij} &= \frac{1}{2^{\delta_{ij}}} \frac{\rho_i N_i \rho_j N_j}{\sum_k \rho_k N_k} - \frac{c_i + c_j}{2} NN_{ij}
\end{aligned}
$$

Let's put that into a code:
```{r load, echo = FALSE, message=FALSE}
library("deSolve")
library("reshape2")
library("ggplot2"); theme_set(theme_bw())
source("../simFuns.R")  ## for transform.list
source("../hivFuns.R")
source("../Param.R")
source("../hivModels.R")
```
```{r}
calc_yini_partner <- function(parameters){
  with(c(expand(parameters)),{
    yini <- list(
      N  = p.risk,
      NN = matrix(0, n.risk, n.risk))
    return(yini)
  })
}

model1 <- function(parameters) {
  pp <- expand(parameters)
  attach(pp)

  cc_mat <- outer(r.risk * c_mean, r.risk * c_mean, "+")/2

  detach(pp)

  g <- function(t,yini,parameters) {
      with(as.list(c(yini,pp)), 
      {
          N <- yini[1:n.risk]
          NN <- matrix(yini[-(1:n.risk)], n.risk, n.risk)
          NN_adj <- NN
          diag(NN_adj) <- 2 * diag(NN_adj)

          frate <- r.risk * rho * N
          f.sum <- sum(frate)

          dN <- - frate + colSums(cc_mat * NN_adj)

          f.NN <- outer(frate, frate, "*")/f.sum 
          diag(f.NN) <- diag(f.NN)/2

          dNN <- f.NN - cc_mat * NN

          list(c(dN, dNN))
      })
  }
}
```

This model seems to be running fine. Now, I would like to calculate the single/couple ratios at equilibrium.... It seems too complicated especially with infected individuals... I can probably get away with it by deriving it numerically:

```{r}

calc_yini_full <- function(parameters){
    with(c(expand(parameters)),{
        c = Ini_I
        I_vec = c * I_initialize(parameters)
        yini <- list(
            S  = (1 - c) * p.risk,
            I = c(outer(I_vec, p.risk)),
            SS = matrix(0, n.risk, n.risk),
            SI = matrix(0, n.risk, n.risk * n.alpha),
            II = matrix(0, n.risk * n.alpha, n.risk *  n.alpha))
        return(yini)
    })
}

colSums2 <- function(data, by){
    if(is.vector(data)){
        data <- matrix(data, nrow = 1)
    }
    length <- ncol(data)
    n <- length/by
    z <- colSums(data)
    z.mat <- matrix(z, ncol = n)
    return(colSums(z.mat))
}

makeMutMat <- function(xpars) {
    p <- with(xpars, {
        p <- matrix(NA, n.alpha, n.alpha)
        for(i in 1:n.alpha){
            x <- min.alpha + (i-1)*d.alpha
            denom <- 
                diff(pnorm(c(min.alpha-d.alpha/2,max.alpha+d.alpha/2),
                    mean=x, sd=Vm))
            for (j in 1:n.alpha){
                k <- min.alpha + (j-1)*d.alpha
                p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
            }
        }
    p
    })
    return(p)
}

model2 <- function(parameters,
                   infection = FALSE,
                   mortality = FALSE) {
    pp <- expand(parameters)
    attach(pp)

    S.ind <- 1:n.risk
    I.ind <- (n.risk+1):(n.risk+n.risk*n.alpha)
    SS.ind <- (n.risk+n.risk*n.alpha+1):(n.risk+n.risk*n.alpha+n.risk*n.risk)
    SI.ind <- (n.risk+n.risk*n.alpha+n.risk*n.risk+1):(n.risk+n.risk*n.alpha+n.risk*n.risk+n.alpha*n.risk*n.risk)
    II.ind <- (n.risk+n.risk*n.alpha+n.risk*n.risk+n.alpha*n.risk*n.risk+1):(n.risk+n.risk*n.alpha+n.risk*n.risk+n.alpha*n.risk*n.risk+n.alpha*n.alpha*n.risk*n.risk)

    rho2 <- r.risk * rho
    c2 <- r.risk * c_mean

    rho2.I <- rep(rho2, each = n.alpha)

    cc_mat <- outer(c2, c2, "+")/2

    c2.I <- rep(c2, each = n.alpha)
    cc_mat.I <- outer(c2, c2.I, "+")/2

    cc_mat.II <- outer(c2.I, c2.I, "+")/2

    alpha2 <- rep(alpha, n.risk)

    #infection
    if(infection){
        p <- makeMutMat(parameters)

        beta.SI <- matrix(rep(Beta, n.risk*n.risk), nrow = n.risk, byrow = TRUE)

        distribute.SIinf <- function(SIinf){
            attach(pp)
            dist.SI <- matrix(0, n.risk * n.alpha, n.risk * n.alpha)
            for(i in 1:n.risk){
                vec.i <- c(1:n.alpha) + n.alpha * (i-1)
                for(j in 1:n.risk){
                    vec.j <- c(1:n.alpha) + n.alpha * (j-1)
                    dist.SI[vec.i,vec.j] <- SIinf[i,vec.j] * p
                }
            }
            detach(pp)
            return(dist.SI)
        }

        c_u_ratio2 <- r.risk * c_u_ratio
        c_e_ratio2 <- r.risk * c_e_ratio

        c_u_ratio2.I <- rep(c_u_ratio2, each = n.alpha)
        c_e_ratio2.I <- rep(c_e_ratio2, each = n.alpha)

        c_u2 <- outer(r.risk, c_u)

        r.risk.couple <- outer(r.risk, r.risk, "+")

    }

    if(mortality){
        lam2 <- rep(lam, n.risk)
        lammat_dis2 <- matrix(rep(lam2, n.alpha*n.risk), n.alpha * n.risk, n.alpha * n.risk)
        diag(lammat_dis2) <- 2 * diag(lammat_dis2)
        lammat_adj2 <- outer(lam2, lam2, "+")

    }

    inf.SI.from <- matrix(0, n.risk, n.risk * n.alpha)
    inf.SI.to <- matrix(0, n.risk * n.alpha, n.risk *  n.alpha)
    inf.S.from <- rep(0, n.risk)
    inf.S.to <- rep(0, n.risk * n.alpha)
    inf.SS.from <- matrix(0, n.risk, n.risk)
    inf.SS.to <- matrix(0, n.risk, n.risk * n.alpha)
    inf.SI.from2 <- matrix(0, n.risk, n.risk * n.alpha)
    inf.SI.to2 <- matrix(0, n.risk * n.alpha, n.risk *  n.alpha)

    mort.I.out <- rep(0, n.risk * n.alpha)
    mort.I.sus <- rep(0, n.risk)
    mort.SI.out <- matrix(0, n.risk, n.risk * n.alpha)
    mort.SI.sus <- rep(0, n.risk)
    mort.II.out <- matrix(0, n.risk * n.alpha, n.risk *  n.alpha)
    mort.II.I <- rep(0, n.risk * n.alpha)
    mort.II.sus <- rep(0, n.risk)

    detach(pp)

    g <- function(t,yini,parameters) {
        with(as.list(c(yini,pp)), 
        {

            S <- yini[S.ind]
            I <- yini[I.ind]
            SS <- matrix(yini[SS.ind], n.risk, n.risk)
            SI <- matrix(yini[SI.ind], n.risk, n.risk * n.alpha)
            II <- matrix(yini[II.ind], n.risk * n.alpha, n.risk * n.alpha)

            ##Partnership formation rate
            N <- S + colSums2(I, n.alpha)

            frate <- rho2 * N
            f.sum <- sum(frate)

            SS_adj <- SS
            diag(SS_adj) <- 2 * diag(SS_adj)

            f.SS <- outer(rho2 * S, rho2 * S)/f.sum
            diag(f.SS) <- diag(f.SS)/2

            f.SI <- outer(rho2 * S, rho2.I * I)/f.sum

            II_adj <- II
            diag(II_adj) <- 2 * diag(II_adj)

            f.II <- outer(rho2.I * I, rho2.I * I)/f.sum
            diag(f.II) <- diag(f.II)/2

            NN <- colSums(SS_adj) + rowSums(SI) + colSums2(SI, n.alpha) + colSums2(II_adj, n.alpha)

            ##Partnership dissolution rate
            d.SS <- cc_mat * SS

            d.SI <- cc_mat.I * SI

            ##trying to model disease dynamics

            if(infection){
                ##within couple infection
                inf.SI.from <- beta.SI * SI
                inf.SI.tmp <- distribute.SIinf(inf.SI.from)
                inf.SI.to <- inf.SI.tmp + t(inf.SI.tmp)
                diag(inf.SI.to) <- diag(inf.SI.to)/2

                ##extra couple infection
                #prop <- (c_u_ratio * I + c_e_ratio * (SI + colSums(II_adj)))/
                #(c_u_ratio * (S + sum(I)) +
                #c_e_ratio * (2 * SS + 2 * sum(SI) + sum(II_adj)))

                prop.mat <- matrix(c(c_u_ratio2.I * I + c_e_ratio2.I * (colSums(SI) + colSums(II_adj))), ncol = n.risk)
                prop.num <- rowSums(prop.mat)
                prop.denom <- sum(c_u_ratio2 * N + c_e_ratio2 * NN)
                prop <- prop.num/prop.denom

                inf.S.rate <- S * sweep(c_u2, 2, prop, "*")
                inf.S.from <- rowSums(inf.S.rate)
                inf.S.to <- c(apply(inf.S.rate, 1, function(x) x %*% p))

                inf.ce.tmp <- c_e * prop
                risk.SS <- r.risk.couple * SS
                inf.SS.from <- risk.SS * sum(inf.ce.tmp)

                ce.mut <- inf.ce.tmp %*% p
                inf.SS.tmp2 <- outer(c(t(r.risk * SS_adj)), c(ce.mut))
                inf.SS.to <- matrix(c(t(inf.SS.tmp2)), nrow = n.risk, byrow = TRUE)

                SI2 <- r.risk * SI
                inf.SI.from2 <- sum(inf.ce.tmp) * SI2

                inf.SI.to2 <- matrix(0, n.risk * n.alpha, n.risk * n.alpha)

                for(i in 1:n.risk){
                    tmp.vec <- c(1:n.alpha) + (i-1) * n.alpha
                    inf.SI.to2[,tmp.vec] <- outer(SI2[i,], ce.mut)
                }

                inf.SI.to2 <- inf.SI.to2 + t(inf.SI.to2)
                diag(inf.SI.to2) <- diag(inf.SI.to2)/2
            }

            if(mortality){
                mort.I.out <- lam2 * I
                mort.I.sus <- colSums2(mort.I.out, n.alpha)

                mort.SI.out <- sweep(SI, 2, lam2, "*")
                mort.SI.sus <- rowSums(mort.SI.out) + colSums2(mort.SI.out, n.alpha)

                mort.II.out <- lammat_adj2 * II
                mort.II.I <- colSums(lammat_dis2 * II)
                mort.II.sus <- colSums2(t(lammat_dis2 * II), n.alpha)
            }

            dS <- - rho2 * S + colSums(cc_mat * SS_adj) + rowSums(d.SI) - inf.S.from + mort.I.sus +
                mort.SI.sus + mort.II.sus
            dI <- - rho2.I * I + colSums(d.SI) + colSums(cc_mat.II * II_adj) + inf.S.to - mort.I.out + mort.II.I
            dSS <- f.SS - d.SS - inf.SS.from
            dSI <- f.SI - d.SI - inf.SI.from + inf.SS.to - inf.SI.from2 - mort.SI.out
            dII <- f.II - cc_mat.II * II + inf.SI.to + inf.SI.to2 - mort.II.out

            ##total ratio of people in the risk group
            #tot <- N + NN

            ##total number of couples
            #coup <- sum(SS[lower.tri(SS, diag = TRUE)]) + sum(SI) + sum(II[lower.tri(II, diag = TRUE)])

            ##total susceptible
            #sus <- S + colSums(SS_adj) + rowSums(SI)
            inf <- I + colSums(SI) + colSums(II_adj)

            tot_I <- sum(inf)
            tot_V <- sum(alpha *  inf)

            test <- inf.S.to

            list(c(dS, dI, dSS, dSI, dII), tot_I = tot_I, mean_V = tot_V/tot_I)
        })
    }
}

```

OK I think I have the full model... Now I need to test this somehow...

```{r comparison}
tvec <- seq(1,100, 0.1)
yini1 <- calc_yini_partner(HIVpars.mean)
yini2 <- calc_yini_full(HIVpars.mean)
r1 <- lsoda(unlist(yini1), func = model1(HIVpars.mean), parms = HIVpars.mean, times = tvec)
r2 <- lsoda(unlist(yini2), func = model2(HIVpars.mean), parms = HIVpars.mean, times = tvec)

N.risk1 <- with(HIVpars.mean, return(r1[,2:(1+n.risk)]))
N.risk2 <- with(expand(HIVpars.mean),{
        S <- r2[,2:(1+n.risk)]
        I <- apply(r2[,(2+n.risk):(1+n.risk+n.risk*n.alpha)], 1, function(x) colSums2(x, n.alpha))
        return(S + t(I))
})

all.equal(unname(N.risk1), unname(N.risk2))

matplot(N.risk1, type = "l")
matplot(N.risk2, type = "l")
```

It seems like partnership dynamics work fine. Can we check disease dynamics?

```{r disease}
pars <- transform(HIVpars.mean, n.risk = 1, r.risk = 1, p.risk = 1, scale_all = 5)

calc_yini_tmp <- function(parameters){
  with(c(expand(parameters)),{
    c = Ini_I
    I_vec = c * I_initialize(parameters)
    yini <- list(
      S  = (1 - c),
      SS = 0,
      SI = rep(0, n.alpha),
      I = I_vec,
      II = matrix(0, n.alpha, n.alpha))
    return(yini)
  })
}

yini_norisk <- calc_yini_tmp(pars)
yini_full <- calc_yini_full(pars)

tvec <- seq(1, 30, 0.1)
r_norisk <- lsoda(unlist(yini_norisk), func = gfun(pars), parms = pars, times = tvec)
r_full <- lsoda(unlist(yini_full), func = model2(pars, infection = TRUE, mortality = TRUE), parms = pars, times = tvec)

all.equal(r_norisk[,488],r_full[,488])

matplot(cbind(r_norisk[,488], r_full[,488]), type = "l")
matplot(cbind(r_norisk[,487], r_full[,487]), type = "l")
```

Introducing heterogeneity:

```{r}
pars2 <- transform(HIVpars.mean, n.risk = 2, r.risk = c(0.8,4.8), p.risk = c(0.95, 0.05), scale_all = 5)
yini_het <- calc_yini_full(pars2)
r_het <- lsoda(unlist(yini_het), func = model2(pars2, infection = TRUE, mortality = TRUE), parms = pars2, times = tvec)

matplot(cbind(r_full[,488], r_het[,1899]), type = "l")
matplot(cbind(r_full[,487], r_het[,1898]), type = "l")
```

One more test. This should be identical to homogeneous model

```{r}
pars3 <- transform(HIVpars.mean, n.risk = 2, r.risk = c(1,1), p.risk = c(0.5, 0.5), scale_all = 5)
yini_test <- calc_yini_full(pars3)
r_test <- lsoda(unlist(yini_test), func = model2(pars3, infection = TRUE, mortality = TRUE), parms = pars3, times = tvec)

matplot(cbind(r_full[,488], r_test[,1899]), type = "l")
matplot(cbind(r_full[,487], r_test[,1898]), type = "l")
```

Seems like this model works?

### 2016-09-12 misc/fullModel.cpp

@@ -1,0 +1,100 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// note both . and _ have potential semantic meaning in C++
+//  (I think _ is safer, though ...)
+
+//[[Rcpp::export]]
+CharacterVector getListElement(List stuff) {
+    return as<CharacterVector>(stuff["a"]);
+}
+
+// getListElement(list(a="abc"))
+
+List g(double t, NumericVector yini, List parameters,
+		NumericMatrix cc_mat) {
+
+    int i,j,k; // counters
+
+    // http://gallery.rcpp.org/articles/subsetting/    
+
+    // should we (1) pass 1 long vector and split it?
+    //           (2) pass 1 long vector and use vector indexing (maybe 
+    // .. #define n(i) N[i]
+    // .. #define nn(j,k) N[nrisk+k*nrisk+j]
+
+    // FIXME: is there a more compact but still reasonably clear
+    //    way to do this?  macro def?
+    int nrisk = as<int>(parameters["n.risk"]);
+    double rrisk = as<double>(parameters["r.risk"]);
+    double rho = as<double>(parameters["rho"]);
+
+    // N <- yini[1:n.risk]
+    NumericVector N(nrisk);
+    for (i=0; i<nrisk; i++) 
+	N(i) = yini(i);
+
+    // copy last (nrisk^2) elements into NN matrix
+    // NN <- matrix(yini[-(1:n.risk)], n.risk, n.risk)
+    // NN_adj <- NN
+    // should be able to replace this with a single incantation
+    // something like std::fill(NN.begin(), NN.end(), N[nrisk,N.end()]);
+
+    NumericMatrix NN(nrisk,nrisk);
+    NumericMatrix NN_adj(nrisk,nrisk);
+
+    for (i=0; i<nrisk; i++) 
+	for (j=0; j<nrisk; j++)  {
+	    NN_adj = NN(i,j) = N[nrisk+k];
+	    k++;
+	}
+    // double diagonal
+    // diag(NN_adj) <- 2 * diag(NN_adj)
+    for (i=0; i<nrisk; i++) 
+	NN_adj(i,i) *= 2;
+
+    //           frate <- r.risk * rho * N
+    //           f.sum <- sum(frate)
+
+    NumericVector frate(nrisk);
[…]

### 2016-09-12 R/misc/fullModel.cpp
fixed fullModel.cpp

@@ -11,90 +11,103 @@ CharacterVector getListElement(List stuff) {
 
 // getListElement(list(a="abc"))
 
+//[[Rcpp::export]]
 List g(double t, NumericVector yini, List parameters,
-		NumericMatrix cc_mat) {
+       NumericMatrix cc_mat) {
+	
+	int i,j,k; // counters
 
-    int i,j,k; // counters
+  // http://gallery.rcpp.org/articles/subsetting/    
 
-    // http://gallery.rcpp.org/articles/subsetting/    
+  // should we (1) pass 1 long vector and split it?
+  //           (2) pass 1 long vector and use vector indexing (maybe 
+  // .. #define n(i) N[i]
+  // .. #define nn(j,k) N[nrisk+k*nrisk+j]
 
-    // should we (1) pass 1 long vector and split it?
-    //           (2) pass 1 long vector and use vector indexing (maybe 
-    // .. #define n(i) N[i]
-    // .. #define nn(j,k) N[nrisk+k*nrisk+j]
+  // FIXME: is there a more compact but still reasonably clear
+  //    way to do this?  macro def?
+  int nrisk = as<int>(parameters["n.risk"]);
+  NumericVector rrisk(nrisk);
+  rrisk = parameters["r.risk"];
+  
+  double rho = as<double>(parameters["rho"]);
+	
+  // N <- yini[1:n.risk]
+  NumericVector N(nrisk);
+  
+  for (i=0; i<nrisk; i++){
+  	
+  	N[i] = yini[i];
+  }
+  
 
-    // FIXME: is there a more compact but still reasonably clear
-    //    way to do this?  macro def?
-    int nrisk = as<int>(parameters["n.risk"]);
-    double rrisk = as<double>(parameters["r.risk"]);
-    double rho = as<double>(parameters["rho"]);
+  // copy last (nrisk^2) elements into NN matrix
+  // NN <- matrix(yini[-(1:n.risk)], n.risk, n.risk)
+  // NN_adj <- NN
+  // should be able to replace this with a single incantation
+  // something like std::fill(NN.begin(), NN.end(), N[nrisk,N.end()]);
 
[…]

### 2016-09-13 R/misc/fullModel.rmd
fixed fullModel.cpp
backup

--- 2016-09-12 misc/fullModel.rmd
@@
 
-Seems like this model works?
+Seems like this model works? Now, let's test rcpp function...
+
+```{r rcpp}
+library(Rcpp)
+sourceCpp("fullModel.cpp")
+
+r1 <- lsoda(unlist(yini1), func = model1(HIVpars.mean), parms = HIVpars.mean, times = tvec)
+r1rcpp <- lsoda(unlist(yini1), func = g, parms = expand(HIVpars.mean), times = tvec)
+yini2 <- list(N = c(0.1, 0.2, 0.3), NN = matrix(c(0.1, 0.1, 0, 0,0,0,0,0,0), 3, 3))
+yini3 <- calc_yini_full(HIVpars.mean)
+g2(0, unlist(yini3), expand(HIVpars.mean))
+model1(expand(HIVpars.mean))(0, unlist(yini2), expand(HIVpars.mean))
+```

### 2016-09-13 R/hivFuns.R

@@ -27,6 +27,11 @@ expand.HIVvirparlist <- function(x,...) {
           Beta = scale_all * scale* (Beta1 * Duration1 + Beta2 * Duration2 + Beta3 * Duration3) * lam
           c_e = Beta * c_e_ratio
           c_u = Beta * c_u_ratio
+          
+          ##Heterogeneity
+          cc_mat <- outer(r.risk * c_mean, r.risk * c_mean, "+")/2
+          rho2 <- r.risk * rho
+          rho2.I <- rep(rho2, each = n.alpha)
        })
     return(x)
 }

### 2016-09-13 R/misc/fullModel.cpp

@@ -11,9 +11,23 @@ CharacterVector getListElement(List stuff) {
 
 // getListElement(list(a="abc"))
 
+NumericMatrix outer2 (NumericVector x,
+                  NumericVector y){
+	unsigned int xn = x.length();
+	unsigned int yn = y.length();
+	
+	NumericMatrix res(xn,yn);
+
+	for(unsigned int i = 0; i < xn; i ++){
+		for(unsigned int j = 0; j < yn; j ++){
+			res(i, j) = x[i] * y[j];
+		}
+	}
+	return res;
+}
+
 //[[Rcpp::export]]
-List g(double t, NumericVector yini, List parameters,
-       NumericMatrix cc_mat) {
+List g(double t, NumericVector yini, List parameters) {
 	
 	int i,j,k; // counters
 
@@ -29,14 +43,13 @@ List g(double t, NumericVector yini, List parameters,
   int nrisk = as<int>(parameters["n.risk"]);
   NumericVector rrisk(nrisk);
   rrisk = parameters["r.risk"];
-  
   double rho = as<double>(parameters["rho"]);
+  NumericMatrix cc_mat(nrisk, nrisk);
+  cc_mat = as<NumericMatrix>(parameters["cc_mat"]);
 	
   // N <- yini[1:n.risk]
   NumericVector N(nrisk);
-  
   for (i=0; i<nrisk; i++){
-  	
   	N[i] = yini[i];
   }
   
@@ -49,21 +62,23 @@ List g(double t, NumericVector yini, List parameters,
 
   NumericMatrix NN(nrisk,nrisk);
   NumericMatrix NN_adj(nrisk,nrisk);
-	
+  
  	for (i=0; i<nrisk; i++){
  		k = i * nrisk;
  		
  		for (j=0; j<nrisk; j++)  {
- 			NN_adj = NN(i,j) = N[nrisk+k];
+ 			NN_adj(j,i) = NN(j,i) = yini[nrisk+k];
  			k++;
  		}
  	}
-		
+ 	
   // double diagonal
   // diag(NN_adj) <- 2 * diag(NN_adj)
   for (i=0; i<nrisk; i++){
   	NN_adj(i,i) *= 2;
   }
+  
+  
 	
[…]

### 2016-09-13 R/misc/fullModel.rmd
done partnership

--- 2016-09-12 misc/fullModel.rmd
@@
 
-Seems like this model works?
+Seems like this model works? Now, let's test rcpp function...
+
+```{r rcpp}
+library(Rcpp)
+sourceCpp("fullModel.cpp")
+
+r1 <- lsoda(unlist(yini1), func = model1(HIVpars.mean), parms = HIVpars.mean, times = tvec)
+r1rcpp <- lsoda(unlist(yini1), func = g, parms = expand(HIVpars.mean), times = tvec)
+yini2 <- list(N = c(0.1, 0.2, 0.3), NN = matrix(c(0.1, 0.1, 0, 0,0,0,0,0,0), 3, 3))
+yini3 <- calc_yini_full(HIVpars.mean)
+tmpa <- unname(g2(0, unlist(yini3), expand(HIVpars.mean))$dSS)
+tmpb <- unname(model2(expand(HIVpars.mean))(0, unlist(yini3), expand(HIVpars.mean))[[1]][67:75])
+all.equal(tmpa, tmpb)
+```

### 2016-09-13 R/hivFuns.R

@@ -31,7 +31,11 @@ expand.HIVvirparlist <- function(x,...) {
           ##Heterogeneity
           cc_mat <- outer(r.risk * c_mean, r.risk * c_mean, "+")/2
           rho2 <- r.risk * rho
+          c2 <- r.risk * c_mean
           rho2.I <- rep(rho2, each = n.alpha)
+          c2.I <- rep(c2, each = n.alpha)
+          cc_mat.I <- outer(c2, c2.I, "+")/2
+          cc_mat.II <- outer(c2.I, c2.I, "+")/2
        })
     return(x)
 }

### 2016-09-13 R/misc/fullModel.cpp

@@ -140,6 +140,10 @@ List g2(double t, NumericVector yini, List parameters) {
 	cc_mat = as<NumericMatrix>(parameters["cc_mat"]);
 	NumericVector rho2 = parameters["rho2"];
 	NumericVector rho2I = parameters["rho2.I"];
+	NumericMatrix cc_matI(nrisk, nrisk * nalpha);
+	cc_matI = as<NumericMatrix>(parameters["cc_mat.I"]);
+	NumericMatrix cc_matII(nrisk * nalpha, nrisk * nalpha);
+	cc_matII = as<NumericMatrix>(parameters["cc_mat.II"]);
 	
 	//S <- yini[S.ind]
 	NumericVector S(nrisk);
@@ -253,33 +257,68 @@ List g2(double t, NumericVector yini, List parameters) {
 	}
 	
 	//d.SS <- cc_mat * SS
+	NumericMatrix d_SS(nrisk, nrisk);
+	for(i = 0; i < nrisk; i ++){
+		for(j = 0; j < nrisk; j ++){
+			d_SS(i,j) = cc_mat(i,j) * SS(i,j);
+		}
+	}
 	
 	//d.SI <- cc_mat.I * SI
-	
-	
+	NumericMatrix d_SI(nrisk, nrisk*nalpha);
+	for(i = 0; i < nrisk; i ++){
+		for(j = 0; j < nrisk * nalpha; j ++){
+			d_SI(i,j) = cc_matI(i,j) * SI(i,j);
+		}
+	}
 	
 	//dS <- - rho2 * S + colSums(cc_mat * SS_adj) + rowSums(d.SI) - inf.S.from + mort.I.sus +
 	//	mort.SI.sus + mort.II.sus
-	
 	NumericVector dS(nrisk);
+	dS = - rho2 * S;
+	for(i = 0; i < nrisk; i ++){
+		dS[i] += sum(cc_mat(_,i) * SS_adj(_,i));
+		dS[i] += sum(d_SI(i,_));
+	}
 	
 	//dI <- - rho2.I * I + colSums(d.SI) + colSums(cc_mat.II * II_adj) + inf.S.to - mort.I.out + mort.II.I
-	
 	NumericVector dI(nrisk * nalpha);
+	dI = - rho2I * I;
+	for(i = 0; i < nrisk * nalpha; i ++){
+		dI[i] += sum(d_SI(_,i));
+		dI[i] += sum(cc_matII(_,i) * II_adj(_,i));
+	}
 	
[…]

### 2016-09-14 R/misc/fullModel.rmd
almost done

--- 2016-09-12 misc/fullModel.rmd
@@
 
-makeMutMat <- function(xpars) {
-    p <- with(xpars, {
-        p <- matrix(NA, n.alpha, n.alpha)
-        for(i in 1:n.alpha){
-            x <- min.alpha + (i-1)*d.alpha
-            denom <- 
-                diff(pnorm(c(min.alpha-d.alpha/2,max.alpha+d.alpha/2),
-                    mean=x, sd=Vm))
-            for (j in 1:n.alpha){
-                k <- min.alpha + (j-1)*d.alpha
-                p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
-            }
-        }
-    p
-    })
-    return(p)
-}
-
 model2 <- function(parameters,
@@
     if(infection){
-        p <- makeMutMat(parameters)
-
         beta.SI <- matrix(rep(Beta, n.risk*n.risk), nrow = n.risk, byrow = TRUE)
@@
                 inf.SI.from <- beta.SI * SI
+
                 inf.SI.tmp <- distribute.SIinf(inf.SI.from)
@@
                 ce.mut <- inf.ce.tmp %*% p
-                inf.SS.tmp2 <- outer(c(t(r.risk * SS_adj)), c(ce.mut))
+                inf.SS.tmp2 <- outer(c((r.risk * SS_adj)), c(ce.mut))
                 inf.SS.to <- matrix(c(t(inf.SS.tmp2)), nrow = n.risk, byrow = TRUE)
@@
                 mort.SI.sus + mort.II.sus
+
             dI <- - rho2.I * I + colSums(d.SI) + colSums(cc_mat.II * II_adj) + inf.S.to - mort.I.out + mort.II.I
@@
 
-            list(c(dS, dI, dSS, dSI, dII), tot_I = tot_I, mean_V = tot_V/tot_I)
+            list(c(dS, dI, dSS, dSI, dII), 
+            		 ##tot_I = tot_I, mean_V = tot_V/tot_I
+            		 inf.SI.to2)
         })
@@
 
-Seems like this model works?
+Seems like this model works? Now, let's test rcpp function...
+
+```{r rcpp}
+library(Rcpp)
+sourceCpp("fullModel.cpp")
+
+r1 <- lsoda(unlist(yini1), func = model1(HIVpars.mean), parms = HIVpars.mean, times = tvec)
+r1rcpp <- lsoda(unlist(yini1), func = g, parms = expand(HIVpars.mean), times = tvec)
+yini2 <- list(N = c(0.1, 0.2, 0.3), NN = matrix(c(0.1, 0.1, 0, 0,0,0,0,0,0), 3, 3))
+yini3 <- calc_yini_full(HIVpars.mean)
+library(microbenchmark)
+##length(unlist(yini3)) = 4233
+yini4 <- rnorm(4233)
+
+microbenchmark(
+tmpa = c(unname(g2(0, unlist(yini4), expand(HIVpars.mean))$dI)),
+tmpb = c(unname(model2(expand(HIVpars.mean), infection = TRUE, mortality = TRUE)(0, unlist(yini4), expand(HIVpars.mean))[[1]][4:66]))
+)
+##1:3, 4:66. 67:75, 76:264
+##-c(1:264)
+all.equal(tmpa, tmpb)
+plot(tmpa)
+plot(tmpb)
+```
+
+InfSIto

### 2016-09-14 R/hivFuns.R

@@ -36,6 +36,35 @@ expand.HIVvirparlist <- function(x,...) {
           c2.I <- rep(c2, each = n.alpha)
           cc_mat.I <- outer(c2, c2.I, "+")/2
           cc_mat.II <- outer(c2.I, c2.I, "+")/2
+          
+          #Infection
+          p <- matrix(NA, n.alpha, n.alpha)
+          for(i in 1:n.alpha){
+          	x <- min.alpha + (i-1)*d.alpha
+          	denom <- 
+          		diff(pnorm(c(min.alpha-d.alpha/2,max.alpha+d.alpha/2),
+          							 mean=x, sd=Vm))
+          	for (j in 1:n.alpha){
+          		k <- min.alpha + (j-1)*d.alpha
+          		p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
+          	}
+          }
+          
+          beta.SI <- matrix(rep(Beta, n.risk*n.risk), nrow = n.risk, byrow = TRUE)
+          
+          #extracouple
+          c_u_ratio2 <- r.risk * c_u_ratio
+          c_e_ratio2 <- r.risk * c_e_ratio
+          c_u_ratio2.I <- rep(c_u_ratio2, each = n.alpha)
+          c_e_ratio2.I <- rep(c_e_ratio2, each = n.alpha)
+          c_u2 <- outer(r.risk, c_u)
+          r.risk.couple <- outer(r.risk, r.risk, "+")
+          
+          lam2 <- rep(lam, n.risk)
+          lammat_dis2 <- matrix(rep(lam2, n.alpha*n.risk), n.alpha * n.risk, n.alpha * n.risk)
+          diag(lammat_dis2) <- 2 * diag(lammat_dis2)
+          lammat_adj2 <- outer(lam2, lam2, "+")
+          
        })
     return(x)
 }
@@ -43,7 +72,6 @@ expand.HIVvirparlist <- function(x,...) {
 hill <- function(x,a,b,p) {
     a*b^p/(b^p+x^p)
 }
-
 
 ##' get r value within specified bounds
 ##'

### 2016-09-14 R/hivModels.R

@@ -1,29 +1,5 @@
-makeMutMat <- function(xpars) {
-  p <- with(xpars, {
-    p <- matrix(NA, n.alpha, n.alpha)
-    for(i in 1:n.alpha){
-      x <- min.alpha + (i-1)*d.alpha
-      denom <- 
-        diff(pnorm(c(min.alpha-d.alpha/2,max.alpha+d.alpha/2),
-                   mean=x, sd=Vm))
-      for (j in 1:n.alpha){
-        k <- min.alpha + (j-1)*d.alpha
-        p[i,j] = diff(pnorm(k + c(-1,1)*d.alpha/2, mean = x, sd = Vm))/denom
-      }
-    }
-    p
-  })
-  return(p)
-}
-
 #extra couple
 gfun <- function(parameters, experimental=FALSE) {
-  pp <- expand(parameters)
-  attach(pp)
-  
-  p <- makeMutMat(pp)
-  
-  detach(pp)
   
   g <- function(t,yini,parameters) {
       with(as.list(c(yini,expand(parameters))), 
@@ -148,12 +124,6 @@ gfun <- function(parameters, experimental=FALSE) {
 
 #serial
 gfun2 <- function(parameters) {
-  pp <- expand(parameters)
-  attach(pp)
-  
-  p <- makeMutMat(pp)
-  
-  detach(pp)
   
   g <- function(t,yini,parameters) {
     with(as.list(c(yini,expand(parameters))), 
@@ -194,11 +164,6 @@ return(g)
 
 #serial + inst
 gfun3 <- function(parameters) {
-  pp <- expand(parameters)
-  attach(pp)
-  
-  p <- makeMutMat(pp)
-  detach(pp)
   
   g <- function(t,yini,parameters) {
     with(as.list(c(yini,expand(parameters))), 
@@ -247,12 +212,6 @@ return(g)
 
 #serial + inst + extra
 gfun4 <- function(parameters) {
-  pp <- expand(parameters)
-  attach(pp)
-  
-  p <- makeMutMat(pp)
-  
-  detach(pp)
   
   g <- function(t,yini,parameters) {
     with(as.list(c(yini,expand(parameters))), 
@@ -305,12 +264,6 @@ return(g)
 
 #shirreff
 gfun5 <- function(parameters) {
-  pp <- expand(parameters)
-  attach(pp)
-  
-  p <- makeMutMat(pp)
-  
-  detach(pp)
   
   g <- function(t,yini,parameters) {
[…]

### 2016-09-14 R/misc/fullModel.cpp

@@ -130,8 +130,9 @@ List g(double t, NumericVector yini, List parameters) {
 //[[Rcpp::export]]
 List g2(double t, NumericVector yini, List parameters) {
 	
-	int i,j,k;
+	int i,j,k,l;
 	
+	//partnership
 	int nrisk = as<int>(parameters["n.risk"]);
 	int nalpha = as<int>(parameters["n.alpha"]);
 	NumericVector rrisk(nrisk);
@@ -144,6 +145,36 @@ List g2(double t, NumericVector yini, List parameters) {
 	cc_matI = as<NumericMatrix>(parameters["cc_mat.I"]);
 	NumericMatrix cc_matII(nrisk * nalpha, nrisk * nalpha);
 	cc_matII = as<NumericMatrix>(parameters["cc_mat.II"]);
+	
+	//infection
+	NumericMatrix betaSI(nrisk, nrisk * nalpha);
+	betaSI = as<NumericMatrix>(parameters["beta.SI"]);
+	NumericMatrix p(nalpha, nalpha);
+	p = as<NumericMatrix>(parameters["p"]);
+	
+	//epc infection
+	NumericVector c_e(nalpha);
+	c_e = parameters["c_e"];
+	NumericVector c_u_ratio2(nrisk);
+	c_u_ratio2 = parameters["c_u_ratio2"];
+	NumericVector c_e_ratio2(nrisk);
+	c_e_ratio2 = parameters["c_e_ratio2"];
+	NumericVector c_u_ratio2I(nalpha * nrisk);
+	c_u_ratio2I = parameters["c_u_ratio2.I"];
+	NumericVector c_e_ratio2I(nalpha * nrisk);
+	c_e_ratio2I = parameters["c_e_ratio2.I"];
+	NumericMatrix c_u2(nrisk, nalpha);
+	c_u2 = as<NumericMatrix>(parameters["c_u2"]);
+	NumericMatrix rriskcouple(nrisk, nrisk);
+	rriskcouple = as<NumericMatrix>(parameters["r.risk.couple"]);
+	
+	//mortality
+	NumericVector lam2(nalpha * nrisk);
+	lam2 = parameters["lam2"];
+	NumericMatrix lammat_dis2(nalpha * nrisk, nalpha * nrisk);
[…]

### 2016-09-14 R/misc/fullModel.rmd
finished moving to c++

@@
             		 ##tot_I = tot_I, mean_V = tot_V/tot_I
-            		 inf.SI.to2)
+            		 mort.SI.out)
         })
@@
 microbenchmark(
-tmpa = c(unname(g2(0, unlist(yini4), expand(HIVpars.mean))$dI)),
-tmpb = c(unname(model2(expand(HIVpars.mean), infection = TRUE, mortality = TRUE)(0, unlist(yini4), expand(HIVpars.mean))[[1]][4:66]))
+tmpa = c(unname(g2(0, unlist(yini4), expand(HIVpars.mean))$dSI)),
+tmpb = c(unname(model2(expand(HIVpars.mean), infection = TRUE, mortality = TRUE)(0, unlist(yini4), expand(HIVpars.mean))[[1]][76:264]))
 )
@@
 ```
-
-InfSIto

### 2016-09-14 R/misc/fullModel.cpp

@@ -403,12 +403,22 @@ List g2(double t, NumericVector yini, List parameters) {
 		cemut[i] = sum(infcetmp * p(_,i));
 	}
 	//FIXME:
+	
+	NumericMatrix infSStmp2(nrisk * nrisk, nalpha);
+	for(i = 0; i < nrisk; i ++){
+		k = nrisk * i;
+		for(j = 0; j < nrisk; j ++){
+			infSStmp2(k,_) = rrisk[j] * SS_adj(j,i) * cemut;
+			k++;
+		}
+	}
+	
 	NumericMatrix infSSto(nrisk, nrisk * nalpha);
 	for(i= 0; i < nrisk; i ++){
 		for(j = 0; j < nrisk * nalpha; j ++){
-			k = floor(j/nalpha);
-			l = j%nalpha;
-			infSSto(i,j) =  rrisk[k] * cemut[l] * SS_adj(i,k);
+			k = j % nalpha;
+			l = i * nrisk + floor(j/nalpha);
+			infSSto(i,j) =  infSStmp2(l,k);
 		}
 	}
 	
@@ -446,52 +456,52 @@ List g2(double t, NumericVector yini, List parameters) {
 	mortIout = lam2 * I;
 	NumericVector mortIsus(nrisk);
 	for(i = 0; i < nrisk * nalpha; i ++){
-		j = i % nalpha;
+		j = floor(i/nalpha);
 		mortIsus[j] += mortIout[i];
 	}
 	
 	//mort.SI.out <- sweep(SI, 2, lam2, "*")
 	//mort.SI.sus <- rowSums(mort.SI.out) + colSums2(mort.SI.out, n.alpha)
-	//NumericMatrix mortSIout(nrisk, nrisk * nalpha);
-	//NumericVector mortSIsus(nrisk);
-	//for(i = 0; i < nrisk; i ++){
-  //	mortSIout(i,_) = SI(i,_) * lam2;
-	//	mortSIsus[i] = sum(mortSIout(i,_));
-	//}
+	NumericMatrix mortSIout(nrisk, nrisk * nalpha);
+	NumericVector mortSIsus(nrisk);
+	for(i = 0; i < nrisk; i ++){
+  	mortSIout(i,_) = SI(i,_) * lam2;
+		mortSIsus[i] = sum(mortSIout(i,_));
+	}
 	
-	//for(i = 0; i < nrisk * nalpha; i ++){
-  //	j = i % nalpha;
[…]

### 2016-09-14 R/misc/fullModel.rmd
test

@@
 
+full_model_rcpp <- function(t, yini, parameters){
+	return(list(unlist(g2(t, yini, parameters))))
+}
+
+pp <- expand(HIVpars.mean)
+
+tvec <- seq(1, 30, 0.5)
+
+r_rcpp <- lsoda(unlist(yini2), func = full_model_rcpp, parms = pp, times = tvec)
+N_rcpp <- with(expand(HIVpars.mean),{
+        S <- r_rcpp[,2:(1+n.risk)]
+        I <- apply(r_rcpp[,(2+n.risk):(1+n.risk+n.risk*n.alpha)], 1, function(x) colSums2(x, n.alpha))
+        return(S + t(I))
+})
+
+r_full <- lsoda(unlist(yini2), func = model2(HIVpars.mean, infection = TRUE, mortality = TRUE), parms = HIVpars.mean, times = tvec)
+N_R <- with(expand(HIVpars.mean),{
+        S <- r_full[,2:(1+n.risk)]
+        I <- apply(r_full[,(2+n.risk):(1+n.risk+n.risk*n.alpha)], 1, function(x) colSums2(x, n.alpha))
+        return(S + t(I))
+})
+
+all.equal(N_rcpp, N_R)
+
+library(microbenchmark)
+microbenchmark(
+	rcpp = lsoda(unlist(yini2), func = full_model_rcpp, parms = pp, times = tvec),
+	R= lsoda(unlist(yini2), func = model2(HIVpars.mean, infection = TRUE, mortality = TRUE), parms = HIVpars.mean, times = tvec)
+)
+
+```
+
+Junk:
+
+```{r junk, eval = FALSE}
+
 r1 <- lsoda(unlist(yini1), func = model1(HIVpars.mean), parms = HIVpars.mean, times = tvec)

### 2016-09-14 R/misc/fullModel.cpp

@@ -533,8 +533,7 @@ List g2(double t, NumericVector yini, List parameters) {
                            Rcpp::Named("dI") = dI,
                            Rcpp::Named("dSS") = dSS,
                            Rcpp::Named("dSI") = dSI,
-                           Rcpp::Named("dII") = dII,
-                           Rcpp::Named("test") = mortSIout);
+                           Rcpp::Named("dII") = dII);
 	
 }
 

### 2016-09-17 R/misc/fullModel.rmd
tweais

@@
 
-            list(c(dS, dI, dSS, dSI, dII), 
-            		 ##tot_I = tot_I, mean_V = tot_V/tot_I
-            		 mort.SI.out)
+            list(c(dS, dI, dSS, dSI, dII), tot_I = tot_I, mean_V = tot_V/tot_I)
         })
@@
 full_model_rcpp <- function(t, yini, parameters){
-	return(list(unlist(g2(t, yini, parameters))))
+	return(list(unlist(rcppFull(t, yini, parameters))))
 }
 
-pp <- expand(HIVpars.mean)
+parstmp <- transform(HIVpars.mean, n.risk = 6, r.risk = c(0.8, 0.9, 1, 1, 1.1, 1.2), p.risk = c(0.15, 0.15, 0.2 , 0.2, 0.15, 0.15))
 
-tvec <- seq(1, 30, 0.5)
+yini2 <- calc_yini_full(parstmp)
 
-r_rcpp <- lsoda(unlist(yini2), func = full_model_rcpp, parms = pp, times = tvec)
+pp <- expand(parstmp)
+
+tvec <- seq(1, 4000, 1)
+
+system.time(r_rcpp <- rk(unlist(yini2), func = full_model_rcpp, parms = pp, times = tvec))
 N_rcpp <- with(expand(HIVpars.mean),{
@@
 
-microbenchmark(
-tmpa = c(unname(g2(0, unlist(yini4), expand(HIVpars.mean))$dSI)),
-tmpb = c(unname(model2(expand(HIVpars.mean), infection = TRUE, mortality = TRUE)(0, unlist(yini4), expand(HIVpars.mean))[[1]][76:264]))
-)
+tmpa = c(unname(rcppPartner(0, unlist(yini4), expand(HIVpars.mean))$dII))
+
+tmpb = c(unname(model2(expand(HIVpars.mean))(0, unlist(yini4), expand(HIVpars.mean))[[1]][-c(1:264)]))
+
 ##1:3, 4:66. 67:75, 76:264

### 2016-09-17 R/Param.R

@@ -18,9 +18,8 @@ HIVpars.skeleton <- list(
   inisd = 0.2,                          ##Standard deviation of the initial SPVL distribution
   Ini_I = 1e-4,                         ##Initial number of infected individuals
   ini_V = 3.5,                          ##Initial mean SPVL
-  n.risk = 3,                           ##number of risk groups
-  r.risk = c(0.9,1,1.1),                ##risk group mixing ratio
-  p.risk = c(0.3, 0.4, 0.3)             ##risk group proportion
+  n.risk = 10,                          ##number of risk groups
+  r.risk = c(1:10)                      ##risk group mixing ratio
 )
 
 
@@ -31,8 +30,16 @@ ltab <- as.data.frame(apply(HIVpars_range,1,
                             function(x) exp(seq(log(x[1]),log(x[2]),
                                                 length=n.trial))))
 
+Het_range <- data.frame(min=c(1, 1),
+												max=c(100, 5),
+												row.names=c("size", "mu"))
+ltab.het <- as.data.frame(apply(Het_range,1,
+																function(x) exp(seq(log(x[1]),log(x[2]),
+																										length=n.trial))))
+
 set.seed(101)
 ltab[] <- lapply(ltab,sample)
+ltab.het[] <- lapply(ltab.het,sample)
 
 as.HIVvirparlist <- function(x) {
   res <- append(x,HIVpars.skeleton)

### 2016-09-17 R/fullModel.R

@@ -1,0 +1,57 @@
+calc_yini_full_ini <- function(parameters){
+	with(c(parameters),{
+		c = Ini_I
+		I_vec = c * I_initialize(parameters)
+		yini <- list(
+			S  = (1 - c) * p.risk,
+			I = c(outer(I_vec, p.risk)),
+			SS = matrix(0, n.risk, n.risk),
+			SI = matrix(0, n.risk, n.risk * n.alpha),
+			II = matrix(0, n.risk * n.alpha, n.risk *  n.alpha))
+		return(yini)
+	})
+}
+
+full_model_rcpp <- function(t, yini, parameters){
+	tmp <- unlist(rcppFull(t, yini, parameters))
+	n <- length(tmp)
+	grad <- tmp[-c(n-1,n)]
+	
+	return(list(grad,  tot_I = tmp[n-1], mean_V = tmp[n]))
+}
+
+partner_model_rcpp <- function(t, yini, parameters){
+	tmp <- unlist(rcppPartner(t, yini, parameters))
+	n <- length(tmp)
+	grad <- tmp[-c(n-1,n)]
+	
+	return(list(grad,  tot_I = tmp[n-1], mean_V = tmp[n]))
+}
+
+calc_yini_full <- function(parameters){
+	yini0 <- calc_yini_full_ini(parameters)
+	
+	p2 <- transform(parameters, scale_all = 5)
+	
+	r <- rk(unlist(yini0), func = partner_model_rcpp, parms = p2, times = c(1:50))
+	
+	r2 <- r[50,-1]
+	
+	with(parameters,{
+		S.ind <- 1:n.risk
+		I.ind <- (n.risk+1):(n.risk+n.risk*n.alpha)
+		SS.ind <- (n.risk+n.risk*n.alpha+1):(n.risk+n.risk*n.alpha+n.risk*n.risk)
+		SI.ind <- (n.risk+n.risk*n.alpha+n.risk*n.risk+1):(n.risk+n.risk*n.alpha+n.risk*n.risk+n.alpha*n.risk*n.risk)
+		II.ind <- (n.risk+n.risk*n.alpha+n.risk*n.risk+n.alpha*n.risk*n.risk+1):
+			(n.risk+n.risk*n.alpha+n.risk*n.risk+n.alpha*n.risk*n.risk+n.alpha*n.alpha*n.risk*n.risk)
+		
+		yini <- list(
+			S = r2[S.ind],
[…]

### 2016-09-17 R/fullModel.cpp

@@ -128,7 +128,7 @@ List g(double t, NumericVector yini, List parameters) {
 }
 
 //[[Rcpp::export]]
-List g2(double t, NumericVector yini, List parameters) {
+List rcppFull(double t, NumericVector yini, List parameters) {
 	
 	int i,j,k,l;
 	
@@ -175,6 +175,10 @@ List g2(double t, NumericVector yini, List parameters) {
 	lammat_dis2 = as<NumericMatrix>(parameters["lammat_dis2"]);
 	NumericMatrix lammat_adj2(nalpha * nrisk, nalpha * nrisk);
 	lammat_adj2 = as<NumericMatrix>(parameters["lammat_adj2"]);
+	
+	//alpha
+	NumericVector alpha2(nalpha * nrisk);
+	alpha2 = parameters["alpha2"];
 	
 	//S <- yini[S.ind]
 	NumericVector S(nrisk);
@@ -529,12 +533,234 @@ List g2(double t, NumericVector yini, List parameters) {
 		}
 	}
 	
+	//inf <- I + colSums(SI) + colSums(II_adj)
+	
+	//tot_I <- sum(inf)
+	//tot_V <- sum(alpha *  inf)
+	
+	NumericVector inf(nrisk * nalpha);
+	
+	for(i = 0; i < nrisk * nalpha; i ++){
+		inf[i] = I[i] + sum(SI(_,i)) + sum(II_adj(_,i));
+	}
+	
+	double tot_I = sum(inf);
+	double mean_V = sum(alpha2 * inf)/tot_I;
+	
 	return Rcpp::List::create(Rcpp::Named("dS") = dS,
                            Rcpp::Named("dI") = dI,
                            Rcpp::Named("dSS") = dSS,
                            Rcpp::Named("dSI") = dSI,
-                           Rcpp::Named("dII") = dII);
+                           Rcpp::Named("dII") = dII,
+                           Rcpp::Named("tot_I") = tot_I,
+                           Rcpp::Named("mean_V") = mean_V);
 	
 }
 
-
+//[[Rcpp::export]]
[…]

### 2016-09-17 R/hivFuns.R

@@ -17,18 +17,24 @@ expand.HIVvirparlist <- function(x,...) {
           v <- 10^alpha
           ## duration <- (Dmax * D50^h)/(D50^h + v^h)
           Duration2 <- hill(v,Dmax,D50,h)
-          lam = 1/(Duration1 + Duration2 + Duration3)
-          lammat = matrix(rep(lam,n.alpha), n.alpha, n.alpha)
-          lammat_dis = lammat
-          diag(lammat_dis) = 2 * diag(lammat_dis)
-          lammat_adj = lammat + t(lammat)
+          lam <- 1/(Duration1 + Duration2 + Duration3)
+          lammat <- matrix(rep(lam,n.alpha), n.alpha, n.alpha)
+          lammat_dis <- lammat
+          diag(lammat_dis) <- 2 * diag(lammat_dis)
+          lammat_adj <- lammat + t(lammat)
           ## Beta2 <- Bmax * v^a / (v^a + K^a)
           Beta2 <- hill(K,Bmax,v,a)
-          Beta = scale_all * scale* (Beta1 * Duration1 + Beta2 * Duration2 + Beta3 * Duration3) * lam
-          c_e = Beta * c_e_ratio
-          c_u = Beta * c_u_ratio
+          Beta <- scale_all * scale* (Beta1 * Duration1 + Beta2 * Duration2 + Beta3 * Duration3) * lam
+          c_e <- Beta * c_e_ratio
+          c_u <- Beta * c_u_ratio
+          
+          ##Risk distribution
+          p.risk_base <- dnbinom(x = r.risk, size = size, mu = mu)
+          p.risk <- p.risk_base/sum(p.risk_base)
           
           ##Heterogeneity
+          alpha2 <- rep(alpha, n.risk)
+          
           cc_mat <- outer(r.risk * c_mean, r.risk * c_mean, "+")/2
           rho2 <- r.risk * rho
           c2 <- r.risk * c_mean
@@ -116,10 +122,16 @@ get_rval <- function(g, yini, pp, plot.it=FALSE,
 
 get_rval2 <- function(val, g, yini, basepar,
                       adjpar="scale",
-                      verbose=FALSE,...) {
+                      verbose=FALSE,
+											full = FALSE, ...) {
     pp <- basepar
     pp[[adjpar]] <- val
-    r <- get_rval(g(pp), yini, pp,...)
+    if(full){
+    	pp2 <- expand(pp)
[…]

### 2016-09-17 R/simulations/ev_LHS_full.R

@@ -1,45 +1,0 @@
-library("deSolve")
-library("reshape2")
-library("ggplot2"); theme_set(theme_bw())
-source("../simFuns.R")  ## for transform.list
-source("../hivFuns.R")
-source("../Param.R")
-source("../hivModels.R")
-
-HIVpars.skeleton <- transform(HIVpars.skeleton, ini_V = 3)
-
-I_mat15 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
-vir_mat15 <- matrix(NA, nrow = length(tvec), ncol = n.trial)
-eq_vec15 <- rep(0, n.trial)
-peak_mat15 <- matrix(NA, nrow = n.trial, ncol = 2)
-val_vec15 <- rep(0, n.trial)
-
-if (file.exists("ev_LHS_res15.rda")) {
-  load("ev_LHS_res15.rda")
-} else {
-  for (i in 1:n.trial) {
-    cat(i)
-    HIVpars <- as.HIVvirparlist(ltab[i,])
-    yini <- calc_yini(HIVpars)
-    little_r <- get_rval(gfun(HIVpars),yini, HIVpars)
-    
-    if(little_r < 0.04){
-      
-      interval = c(1,10)
-    }else{
-      interval = c(0.01,1)
-    }
-    val_vec15[i] = find_scale(0.04, gfun, yini, HIVpars, interval=interval, adjpar = "scale_all")
-    HIVpars_adj <- transform(HIVpars, scale_all = val_vec15[i])
-    r <- lsoda(unlist(yini), func = gfun(HIVpars_adj), parms = HIVpars_adj, times = tvec)
-    
-    I_mat15[,i] = r[,(ncol(r) - 1)]
-    vir_mat15[,i] = r[,ncol(r)]
-    eq_vec15[i] = vir_mat15[length(tvec),i]
-    
-    peak_mat15[i,1] = which.max(vir_mat15[,i])
-    peak_mat15[i,2] = max(vir_mat15[,i])
-    
-  }
-  save("I_mat15", "vir_mat15", "eq_vec15","peak_mat15", "val_vec15",file="ev_LHS_res15.rda")
-}

### 2016-09-17 R/simulations/ev_LHSfull.R

@@ -1,0 +1,54 @@
+library("Rcpp")
+library("deSolve")
+library("reshape2")
+library("ggplot2"); theme_set(theme_bw())
+sourceCpp("../fullModel.cpp")
+source("../fullModel.R")
+source("../simFuns.R")  ## for transform.list
+source("../hivFuns.R")
+source("../Param.R")
+source("../hivModels.R")
+
+ltab.full <- cbind(ltab, ltab.het)
+
+HIVpars.skeleton <- transform(HIVpars.skeleton, ini_V = 3)
+
+tvec <- c(1:1000)
+
+I_matFull <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+vir_matFull <- matrix(NA, nrow = length(tvec), ncol = n.trial)
+eq_vecFull <- rep(0, n.trial)
+peak_matFull <- matrix(NA, nrow = n.trial, ncol = 2)
+val_vecFull <- rep(0, n.trial)
+
+if (file.exists("ev_LHS_resFull.rda")) {
+	load("ev_LHS_resFull.rda")
+} else {
+	for (i in 1:100) {
+		cat(i)
+		HIVpars <- as.HIVvirparlist(ltab.full[i,])
+		pp <- expand(HIVpars)
+		yini <- calc_yini_full(pp)
+		little_r <- get_rval(full_model_rcpp, yini, pp, tvec = c(1:30))
+		
+		if(little_r < 0.04){
+			interval = c(1,10)
+		}else{
+			interval = c(0.01,1)
+		}
+		val_vecFull[i] = find_scale(0.04, full_model_rcpp, yini, HIVpars, interval=interval,
+															adjpar = "scale_all", full = TRUE, tvec = c(1:300))
+		HIVpars_adj <- transform(HIVpars, scale_all = val_vecFull[i])
+		pp_adj <- expand(HIVpars_adj)
+		r <- rk(unlist(yini), func = full_model_rcpp, parms = pp_adj, times = tvec)
+		
+		I_matFull[,i] = r[,(ncol(r) - 1)]
+		vir_matFull[,i] = r[,ncol(r)]
+		eq_vecFull[i] = vir_matFull[length(tvec),i]
+		
+		peak_matFull[i,1] = which.max(vir_matFull[,i])
+		peak_matFull[i,2] = max(vir_matFull[,i])
+		
+	}
+	save("I_matFull", "vir_matFull", "eq_vecFull","peak_matFull", "val_vecFull",file="ev_LHS_resFull.rda")
+}

### 2016-09-21 ms/plos_latex.tex
tweaks to the manuscript

@@
 Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples.
-Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and Descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
+Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
 \centering
@@
 \hline % inserts single horizontal line
-$\rho$ & Partnership formation rate & 1/10-2/5 & \cite{champredon_hiv_2013} \\
-$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) & \cite{champredon_hiv_2013} \\
+$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
+$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
 $c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
 $c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
-$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) & \cite{hollingsworth_hiv1_2008} \\
-$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) & \cite{hollingsworth_hiv1_2008} \\
-$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) & \cite{hollingsworth_hiv1_2008,shirreff_transmission_2011} \\
-$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) & \cite{hollingsworth_hiv1_2008} \\
-$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 & \cite{shirreff_transmission_2011} \\
-$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 & \cite{shirreff_transmission_2011} \\
+$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
+$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
+$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
+$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
+$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
+$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
 $\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
-$\tsub{D}{max}$ & Duration of primary infection & 25.4 & \cite{shirreff_transmission_2011} \\
-$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 & \cite{shirreff_transmission_2011} \\
+$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
+$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
 $D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
@@
 Partnership formation rate for $S$, $I_i$ are $- \rho S$ and $-\rho I_i$, respectively.
-We follow Champredon \etal's results and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
+We follow Champredon \etal's results \cite{champredon_hiv_2013} and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
 
@@
 Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - II_{ij}$.
-Once individuals leave a partnership, they enter temporary compartments and are distributed into a partnership as follows:
+Once individuals leave a partnership, they are instantaneously distributed into coupled states.
+In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the number of susceptible individuals that leave the partnership at a given time, and $Y_i$ the number of infected individuals with \Lspvl\ of $\alpha_i$ who leave partnership at a given time.
+These temporarily single individuals then form couples through the same partnership formation rule described in the previous section:
 
@@
 Y_i =& c SI_i + \sum_k 2^{\delta_{ik}}  c II_{ik} + \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik}, \\
-SS  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS,\\
-SI_i =& - c SI_i + \frac{X Y_i}{X + \sum_k Y_k} - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS\\
+SS'  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS,\\
+SI_i' =& - c SI_i + \frac{X Y_i}{X + \sum_k Y_k} - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS\\
 &- \sum_k P_k E_k SI_i - \lambda_i SI_i,\\
-II_{ij}& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)\\
+II_{ij}'& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)\\
 &+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}.
@@
 
-Following \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta'$, that is derived from \cite{hollingsworth_hiv1_2008}'s approximate basic reproduction number:
-
-\begin{equation}
-\beta'_i = \frac{c \beta_i}{c + \beta_i + \lambda_i}.
+Following \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta^\ast$, that is derived from \cite{hollingsworth_hiv1_2008}'s approximate basic reproduction number:
+
+\begin{equation}
+\beta^\ast_i = \frac{c \beta_i}{c + \beta_i + \lambda_i}.
 \end{equation}
@@
 \begin{aligned}
-S' & = \sum_k \lambda_k I_k - \sum_k \beta'_k S I_k.\\
-I_i & = \sum_k M_{ki} \beta'_k S I_k - \lambda_i I_i.
+S' & = \sum_k \lambda_k I_k - \sum_k \beta^\ast_k S I_k.\\
+I_i' & = \sum_k M_{ki} \beta^\ast_k S I_k - \lambda_i I_i.
 \end{aligned}
@@
 S' & = \sum_k \lambda_k I_k - \sum_k \beta_k S I_k,\\
-I_i & = \sum_k M_{ki} \beta_k S I_k - \lambda_i I_i.
+I_i' & = \sum_k M_{ki} \beta_k S I_k - \lambda_i I_i.
 \end{aligned}

### 2016-09-21 R/Param.R

@@ -41,6 +41,8 @@ set.seed(101)
 ltab[] <- lapply(ltab,sample)
 ltab.het[] <- lapply(ltab.het,sample)
 
+ltab <- cbind(ltab, ltab.het)
+
 as.HIVvirparlist <- function(x) {
   res <- append(x,HIVpars.skeleton)
   class(res) <- c("list","parlist","HIVvirparlist")

### 2016-09-21 R/hivFuns.R

@@ -98,10 +98,9 @@ get_rval <- function(g, yini, pp, plot.it=FALSE,
     r <- rk(y=start,
                  times=tvec,
                  func=g,
-                 parms=pp, method = "ode45")
+                 parms=pp)
         
     Itot <- r[,(ncol(r)-1)]
-    
     if(max(Itot) < 5e-2){
       return(0)
     }else{

### 2016-09-21 R/hivModels.R

@@ -1,8 +1,8 @@
 #extra couple
 gfun <- function(parameters, experimental=FALSE) {
-  
+	px <- expand(parameters)
   g <- function(t,yini,parameters) {
-      with(as.list(c(yini,expand(parameters))), 
+      with(as.list(c(yini,px)), 
       { 
           SI <- yini[3:(2+n.alpha)]
           I <- yini[(3+n.alpha):(2+2*n.alpha)]
@@ -56,9 +56,7 @@ gfun <- function(parameters, experimental=FALSE) {
   }
 
   ## tweak for performance
-
-  px <- expand(parameters)
-  gX <- function(t,yini,parameters) {
+	gX <- function(t,yini,parameters) {
       with(as.list(c(yini,px)),
       { 
           SI <- yini[3:(2+n.alpha)]
@@ -124,9 +122,9 @@ gfun <- function(parameters, experimental=FALSE) {
 
 #serial
 gfun2 <- function(parameters) {
-  
+	px <- expand(parameters)
   g <- function(t,yini,parameters) {
-    with(as.list(c(yini,expand(parameters))), 
+    with(as.list(c(yini,px)), 
 {      
   
   SI = yini[3:(2+n.alpha)]
@@ -164,9 +162,9 @@ return(g)
 
 #serial + inst
 gfun3 <- function(parameters) {
-  
+	px <- expand(parameters)
   g <- function(t,yini,parameters) {
-    with(as.list(c(yini,expand(parameters))), 
+    with(as.list(c(yini,px)), 
 {      
   
   SI = yini[2:(1+n.alpha)]
@@ -212,9 +210,9 @@ return(g)
 
 #serial + inst + extra
 gfun4 <- function(parameters) {
-  
+	px <- expand(parameters)
   g <- function(t,yini,parameters) {
-    with(as.list(c(yini,expand(parameters))), 
+    with(as.list(c(yini,px)), 
 {      
   
   SI = yini[2:(1+n.alpha)]
@@ -264,9 +262,9 @@ return(g)
 
 #shirreff
 gfun5 <- function(parameters) {
-  
+	px <- expand(parameters)
   g <- function(t,yini,parameters) {
-    with(as.list(c(yini,expand(parameters))), 
+    with(as.list(c(yini,px)), 
 {      
   
   I = yini[-1]
@@ -290,9 +288,9 @@ return(g)
 
 #random mixing
 gfun_random <- function(parameters) {
-  
[…]

### 2016-09-21 R/simulations/ev_LHS15.R

@@ -31,7 +31,7 @@ if (file.exists("ev_LHS_res15.rda")) {
     }
     val_vec15[i] = find_scale(0.04, gfun, yini, HIVpars, interval=interval, adjpar = "scale_all")
     HIVpars_adj <- transform(HIVpars, scale_all = val_vec15[i])
-    r <- lsoda(unlist(yini), func = gfun(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    r <- rk(unlist(yini), func = gfun(HIVpars_adj), parms = HIVpars_adj, times = tvec)
     
     I_mat15[,i] = r[,(ncol(r) - 1)]
     vir_mat15[,i] = r[,ncol(r)]

### 2016-09-21 R/simulations/ev_LHS16.R

@@ -31,7 +31,7 @@ if (file.exists("ev_LHS_res16.rda")) {
     }
     val_vec16[i] = find_scale(0.04, gfun2, yini, HIVpars, interval=interval, adjpar = "scale_all")
     HIVpars_adj <- transform(HIVpars, scale_all = val_vec16[i])
-    r <- lsoda(unlist(yini), func = gfun2(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    r <- rk(unlist(yini), func = gfun2(HIVpars_adj), parms = HIVpars_adj, times = tvec)
 
     I_mat16[,i] = r[,(ncol(r) - 1)]      
     vir_mat16[,i] = r[,ncol(r)]

### 2016-09-21 R/simulations/ev_LHS17.R

@@ -31,7 +31,7 @@ if (file.exists("ev_LHS_res17.rda")) {
     }
     val_vec17[i] = find_scale(0.04, gfun3, yini, HIVpars, interval=interval, adjpar = "scale_all")
     HIVpars_adj <- transform(HIVpars, scale_all = val_vec17[i])
-    r <- lsoda(unlist(yini), func = gfun3(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    r <- rk(unlist(yini), func = gfun3(HIVpars_adj), parms = HIVpars_adj, times = tvec)
     
     I_mat17[,i] = r[,(ncol(r) - 1)]
     vir_mat17[,i] = r[,ncol(r)]

### 2016-09-21 R/simulations/ev_LHS18.R

@@ -31,7 +31,7 @@ if (file.exists("ev_LHS_res18.rda")) {
     }
     val_vec18[i] = find_scale(0.04, gfun4, yini, HIVpars, interval=interval, adjpar = "scale_all")
     HIVpars_adj <- transform(HIVpars, scale_all = val_vec18[i])
-    r <- lsoda(unlist(yini), func = gfun4(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    r <- rk(unlist(yini), func = gfun4(HIVpars_adj), parms = HIVpars_adj, times = tvec)
     
     I_mat18[,i] = r[,(ncol(r) - 1)]
     vir_mat18[,i] = r[,ncol(r)]

### 2016-09-21 R/simulations/ev_LHS19.R

@@ -31,7 +31,7 @@ if (file.exists("ev_LHS_res19.rda")) {
     }
     val_vec19[i] = find_scale(0.04, gfun5, yini, HIVpars, interval=interval, adjpar = "scale_all")
     HIVpars_adj <- transform(HIVpars, scale_all = val_vec19[i])
-    r <- lsoda(unlist(yini), func = gfun5(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    r <- rk(unlist(yini), func = gfun5(HIVpars_adj), parms = HIVpars_adj, times = tvec)
     
     I_mat19[,i] = r[,(ncol(r) - 1)]
     vir_mat19[,i] = r[,ncol(r)]

### 2016-09-21 R/simulations/ev_LHS20.R

@@ -31,7 +31,7 @@ if (file.exists("ev_LHS_res20.rda")) {
     }
     val_vec20[i] = find_scale(0.04, gfun_random, yini, HIVpars, interval=interval, adjpar = "scale_all")
     HIVpars_adj <- transform(HIVpars, scale_all = val_vec20[i])
-    r <- lsoda(unlist(yini), func = gfun_random(HIVpars_adj), parms = HIVpars_adj, times = tvec)
+    r <- rk(unlist(yini), func = gfun_random(HIVpars_adj), parms = HIVpars_adj, times = tvec)
     
     I_mat20[,i] = r[,(ncol(r) - 1)]
     vir_mat20[,i] = r[,ncol(r)]

### 2016-09-21 R/simulations/ev_LHSfull.R

@@ -9,11 +9,9 @@ source("../hivFuns.R")
 source("../Param.R")
 source("../hivModels.R")
 
-ltab.full <- cbind(ltab, ltab.het)
-
 HIVpars.skeleton <- transform(HIVpars.skeleton, ini_V = 3)
 
-tvec <- c(1:1000)
+tvec <- c(1:2000)
 
 I_matFull <- matrix(NA, nrow = length(tvec), ncol = n.trial)
 vir_matFull <- matrix(NA, nrow = length(tvec), ncol = n.trial)
@@ -26,7 +24,7 @@ if (file.exists("ev_LHS_resFull.rda")) {
 } else {
 	for (i in 1:100) {
 		cat(i)
-		HIVpars <- as.HIVvirparlist(ltab.full[i,])
+		HIVpars <- as.HIVvirparlist(ltab[i,])
 		pp <- expand(HIVpars)
 		yini <- calc_yini_full(pp)
 		little_r <- get_rval(full_model_rcpp, yini, pp, tvec = c(1:30))

### 2016-09-25 figures/figures.rmd
new simulations

@@
 ```{r load, echo = FALSE}
-load("../simdata/combine_ev_LHS4.rda")
+load("../simdata/combineResults.rda")
 ```
@@
 ```{r fig5, echo = FALSE, fig.width = 10, cache = TRUE,dpi = 600}
+mL2 <- subset(mL2, !(LHSvar %in% c("kappa", "mu")))
 mL3 <- subset(mL2,
@@
           LHSvar %in% c("c_u_ratio","c_e_ratio"))))
+
 ## reorder factors

### 2016-10-03 ms/plos_latex.tex
update to the manuscript (still need more updates)

@@
 \newcommand{\khalf}{\left(\frac{1}{2}\right)^{\delta_{ij}}}  % (1/2)^kronecker
+\newcommand{\kkhalf}{\left(\frac{1}{2}\right)^{\delta_{ij} \delta_{kl}}}  % (1/2)^(kronecker * kronecker)
 \newcommand{\Lspvl}{$\log_{10}$ SPVL}
@@
 
-We developed six multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+We developed seven multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
 Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
-non-instantaneous partnership formation and (2) zero vs.
-positive extra-partnership sexual contact and transmission on the evolution of mean \Lspvl.
+non-instantaneous partnership formation, (2) zero vs.
+positive extra-partnership sexual contact and transmission and (3) homogeneous vs.
+heterogeneous mixing on the evolution of mean \Lspvl.
 
@@
 
-The last two models do not explicitly track sexual partnerships.
+The next two models do not explicitly track sexual partnerships.
 One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
@@
 
+Lastly, we add heterogeneous mixing to pairform+epc model.
+Individuals are divided into different risk groups based on the sexual activity level, and we assume that sexual activity level is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year [CITE].
+Model details can be found in the appendix.
+
 The base model (i.e.
-pairform+epc) for the first four models is an extension of Champredon \etal's model \cite{champredon_hiv_2013}.
+pairform+epc) for the first four models and the heterogeneous model is an extension of Champredon \etal's model \cite{champredon_hiv_2013}.
 Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$.
@@
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
+For the heterogeneity model, mean and coefficient of variance (CV) for number of non-cohabiting partners are sampled as well.
 We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
@@
 simplified. We run each model for only 500 years
-(full simulations are run for 4000 years), and we use a 4/5 order 
+(full simulations are run for 4000 years). We use a 4/5 order 
 Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
-\cite{soetaert_solving_2010}), whereas we use
-a stiff solver (LSODA) for the full simulations.
+\cite{soetaert_solving_2010}) for all simulations.
 
@@
 Single individuals acquire partners at per-person rate $\rho$.
-Partnership formation rate for $S$, $I_i$ are $- \rho S$ and $-\rho I_i$, respectively.
+Partnership formation rate for $S$, $I_i$ are $\rho S$ and $\rho I_i$, respectively.
 We follow Champredon \etal's results \cite{champredon_hiv_2013} and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
@@
 Within-couple transmission (WT) occurs in both models.
-An infected partner in $SI$ partnership transmits virus to a susceptible partner, and partnership state becomes $II$: $\WT(SI_i) = - \beta_i SI_i$.
+An infected partner in $SI$ partnership transmits virus to a susceptible partner at per-partnership rate $\beta$: $\WT(SI_i) = - \beta_i SI_i$.
 Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$.
@@
 
-Single susceptible individuals become infected and enter single infected compartment at the total rate of $\sum_k P_k U_k S$.
+Single susceptible individuals become infected through uncoupled contact at per-person rate $\sum_k P_k U_k$ and enter single infected state.
 Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $\ET(I_i) = \sum_k M_{ki} P_k U_k S$.
-Either partner in an $SS$ partnership can be infected, with the partnership state becoming $SI$, at the total rate of $\sum_i 2 P_i E_i SS$.
+Either partner in an $SS$ partnership becomes infected at per-person rate $\sum_k P_k E_k$, and partnership state changes to an $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$.
 The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $\ET(SI_i) = \sum_k 2 M_{ki} P_k E_k SS$.
-Lastly, the susceptible partner in an $SI$ partnership can be infected due to uncoupled/extra-couple contacts and partnership can change to an $II$ partnership.
+Lastly, the susceptible partner in an $SI$ partnership becomes infected from extra-couple contacts at a per-person rate of $\sum_k P_k E_k$, and partnership state changes to an $II$ partnership.
 As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$.
@@
 \end{equation}
+
+\subsection*{Heteregenous model}
+
+We extend "pairform+epc model" by allowing for heterogeneity in sexual behaviour.
+Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model the mentioned behaviours are scaled by the same factor based on the risk group.
+In other words, an individual in a higher risk is more likely to form a stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing.
+We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group.
+For simplicity, we assume that the transmission rate per partnership is not affected by sexual behaviour.
+
+\subsubsection*{Partnership dynamics}
+
+Individuals in a risk group $i$ leave single state at a rate of $\varphi_i \rho$.
+Let $XY_{ij,kl}$ be a coupled state where $X$ and $Y$ are the infection status (susceptible or infected) of each partner, $k$ and $l$ are the risk groups $X$ and $Y$ belong to respectively, and $i$ and $j$ are the strains of an infected partner.
+If a partner is susceptible, strain index is replaced by $\cdot$.
+For example, $SI_{\cdot j,kl}$ is the number of partners where the susceptible partner is in risk group $k$ and infected partner is in risk group $l$ and has \Lspvl\ of $\alpha_j$.
+For simplicity, we assume that people mix randomly.
+Then, we can write the partnership formation process as follows:
+
+\begin{equation}
+\PF(XY_{ij, kl}) = \kkhalf \frac{\varphi_k \rho X_{i,k} \varphi_l \rho Y_{j, l}}{\sum\limits_m \varphi_m \rho (S_{\cdot, m} + \sum\limits_n I_{n,m})}
+\end{equation}
+
+For dissolution process, individual in risk group $i$ leave partnership at a rate $\varphi_i c$.
+If partnership is formed between two individuals from a different risk group, rate at which they leave the partnership differs.
+We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners.
+Therefore, $XY_{ij, kl}$ dissolve at a rate $\frac{\varphi_k + \varphi_l}{2} c XY_{ij, kl}$, and both $X_{i,k}$ and $Y_{j,l}$ return to single state at a rate of $\frac{\varphi_k + \varphi_l}{2} c XY_{ij, kl}$.
+
+\subsubsection*{Heterogeneous models: Infection dynamics}
+
+Since we assume that the rate of transmission per partnership stays constant across different risk groups, within-couple infection process is similar to other models:
+
+\begin{equation}
+\begin{aligned}
+WT(SI_{\cdot j, kl} &= - \beta_j SI_{\cdot j, kl}\\
+WT(II_{ij, kl}) &= \kkhalf \cdot (M_{ji} \beta_j SI_{\cdot j, kl} + M_{ij} \beta_i SI_{\cdot i, lk})
+\end{aligned}
+\end{equation}
+
+Note that $II_{ij,kl}$ can be formed from two types of partnerships: 1) Infected partner with \Lspvl\ of $\alpha_j$ and risk group of $l$ infects a susceptible partner in risk group $k$, yielding \Lspvl\ of $\alpha_i$ through mutation.
+2) Infected partner with \Lspvl\ of $\alpha_i$ and risk group of $k$ infected a susceptible partner in risk group $l$, yielding \Lspvl\ of $\alpha_j$ through mutation.
+On the other hand, if $i = j$ and $k = l$, $II_{ii,kk}$ can only be formed from $SI_{\cdot i, kk}$ partnership, which is resolved by $\kkhalf$.
+
+Heterogeneous extra-couple and uncoupled contact process is similar to partnership formation process.
+Relative uncoupled/extra-couple contact rates are scaled by the factor of $\varphi_i$, where $i$ is the risk group.
+First, we define $Q_{i}$ as the total rate of uncoupled/extra couple contact by individuals in risk group $k$:
+
+\begin{equation}
+\begin{aligned}
+Q_i = &\varphi_i r_u (S_{\cdot,i} +  \sum_j I_{j,i}) + \varphi_i r_e \bigg( \sum_k 2^{\delta_{ik}} SS_{\cdot, ik} +\\
+&\sum_l \sum_j (SI_{\cdot j,il} + SI_{\cdot j, li}) + \sum_j \sum_l \sum_k 2^{\delta_{kl} \delta_{ij}} II_{kl,ij} \bigg)
+\end{aligned}
+\end{equation}
+We now define $P_{k,i}$ as the proportion of the extra-couple and uncoupled contact that arises from an infected individual from risk group $i$ with \Lspvl\ of $\alpha_k$:
+\begin{equation}
+P_{k,i} = \frac{\varphi_i r_u I_{k,i} + \varphi_i r_e (SI_{k,i} + \sum_j \sum_l 2^{\delta_{kl} \delta_{ij} } II_{kl,ij} )}{\sum_j Q_j}
+\end{equation}
+Since the relative uncoupled/extra couple contact ratio are scaled by the factor of $\varphi_i$, uncoupled and extra-couple transmission rates are scaled by the same factor as well: $U_{k,i} = \varphi_i r_u \beta_k$ and $E_{k,i} = \varphi_i r_e \beta_k$.
+Once again, we assume random mixing between individuals.
+Then, a susceptible individual in risk group $i$ becomes infected through extra-couple and uncoupled contact at a per capita rate of $\sum_j \sum_k P_{k,j} X_{k,i}$.
+Once infected, individuals are distributed into each $\Lspvl$ categories through mutation.
+
+\subsubsection*{Heterogeneous model: Disease induced mortality}
+
+Disease induced mortality is not affected by the sexual behaviour of an individual.
 

### 2016-10-03 R/Param.R
tweaks to the code

@@ -29,8 +29,8 @@ ltab <- as.data.frame(apply(HIVpars_range,1,
                             function(x) exp(seq(log(x[1]),log(x[2]),
                                                 length=n.trial))))
 
-Het_range <- data.frame(min=c(0.1, 0.103),
-												max=c(20.324, 1.206),
+Het_range <- data.frame(min=c(0.01, 0.103),
+												max=c(100, 1.206),
 												row.names=c("kappa", "mu"))
 ltab.het <- as.data.frame(apply(Het_range,1,
 																function(x) exp(seq(log(x[1]),log(x[2]),

### 2016-10-03 R/hivFuns.R

@@ -94,7 +94,7 @@ hill <- function(x,a,b,p) {
 ##' @param verbose verbose (debugging) output?
 get_rval <- function(g, yini, pp, plot.it=FALSE,
                      tvec = c(1:500),
-                     lims=c(1e-3,5e-2),
+                     lims=c(1e-3,1e-2),
                      verbose=FALSE) {
   
     start <- unlist(yini)
@@ -102,11 +102,11 @@ get_rval <- function(g, yini, pp, plot.it=FALSE,
     r <- rk(y=start,
                  times=tvec,
                  func=g,
-                 parms=pp, hmax = 0.5)
+                 parms=pp, hmax = 0.3)
         
     Itot <- r[,(ncol(r)-1)]
     
-    if(max(Itot) < 5e-2){
+    if(max(Itot) < 1e-2){
       return(0)
     }else{
       

### 2016-10-03 R/simulations/ev_LHSfull_assemble.R

@@ -74,13 +74,16 @@ if (FALSE) {
     
     min(all_comb$I_matFull[2000,], na.rm = TRUE)
     
-    little_r <- unlist(apply(all_comb$I_matFull, 2, function(x){
+    Imat <- all_comb$I_matFull[,!is.na(all_comb$I_matFull[1,])]
+    
+    little_r <- unlist(apply(Imat, 2, function(x){
     	df <- data.frame(t = 1:2000, I = x)
-    	l <- try(lm(log(I)~t, data = df, subset = I > 1e-3 & I < 3e-2))
+    	l <- try(lm(log(I)~t, data = df, subset = I > 1e-3 & I < 1e-2))
     	ifelse(!inherits(l,"try-error"), coef(l)[[2]], NA) 
     }))
+    
     plot(density(little_r, na.rm = TRUE))
-    matplot(all_comb$I_matFull[,which(little_r < 0.05)], type = "l", lty = 1, col = pcol, log = "y")
+    matplot(Imat[,which(little_r < 0.05)], type = "l", lty = 1, col = pcol, log = "y")
     length(which(little_r < 0.05))
     length(which(little_r > 0.05))
 }

### 2016-10-03 R/simulations/ev_LHSfull_chunk.R

@@ -53,7 +53,7 @@ for (i in start_sim:end_sim) {
             HIVpars_adj <- transform(HIVpars, scale_all = val_vecFull[mrow])
             pp_adj <- expand(HIVpars_adj)
             rk(unlist(yini), func = full_model_rcpp, parms = pp_adj,
-               times = tvec, hmax = 0.5)
+               times = tvec, hmax = 0.3)
             })
     if (!inherits(r,"try-error")) {
         I_matFull[,mrow] = r[,(ncol(r) - 1)]

### 2016-10-03 R/simulations/ev_LHSfull_chunk.R
setup for big batch run 3

@@ -9,6 +9,8 @@ source("../hivFuns.R")
 source("../Param.R")
 source("../hivModels.R")
 
+BATCHTAG <- "_v3"
+
 t0 <- proc.time()
 argvals <- commandArgs(trailingOnly=TRUE) ## chunk number: full run is (20)x50
 ## indexed from 0
@@ -18,7 +20,7 @@ cat(batch_num,batch_size,"\n")
 
 start_sim <- batch_num*batch_size+1  
 end_sim <- (batch_num+1)*batch_size
-fn <- paste0("ev_LHSfull_",batch_num,"_v2.rda")
+fn <- paste0("ev_LHSfull_",batch_num,BATCHTAG,".rda")
 cat(start_sim,end_sim,fn,"\n")
 
 n.trial <- end_sim-start_sim+1 ## overwrite source()d value

### 2016-10-04 ms/plos_latex.tex
tweaks to ms

@@
 
+Although such simplification affects evolution of virulence, we show that our simplified model is able to produce result that is qualitatively similar to that of Shirreff \etal\'s \cite{shirreff_transmission_2011}.
+
 \subsection*{Mutation}
@@
 \begin{equation}
-\alpha_i =  (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{(i-1)}{n-1} + \tsub{\alpha}{min} \qquad i = 1,2,3, \dots n.
+\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
 \end{equation}
@@
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-For the heterogeneity model, mean and coefficient of variance (CV) for number of non-cohabiting partners are sampled as well.
+For the heterogeneity model, mean and squared coefficient of variation (CV) for number of non-cohabiting partners are sampled as well.
 We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
@@
 \begin{equation}
-z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1,...,N.
+z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1, 2, 3, \dots,N.
 \end{equation}
@@
 
-The disease induced mortality (DM) rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
+The per-person disease induced mortality (DM) rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
 Since we assume an SIS formulation, where infected individuals that die from infection are immediately replaced by an individual in the single susceptible compartment, we obtain the following equation for single infected individuals:
@@
 If an infected individual in a partnership dies, the partnership dissolves.
-Thus, an $SI_i$ partnership dissolves at a rate $-\lambda_i$, and the susceptible partner enters the single susceptible compartment at rate $\lambda_i SI_i$ (due to the SIS formulation the the infected partner that dies also gives rise, at an equal rate, to an individual entering the single susceptible compartment):
+Thus, an $SI_i$ partnership dissolves at per-partnership rate $\lambda_i$, and the susceptible partner enters the single susceptible compartment at rate $\lambda_i SI_i$ (due to the SIS formulation, the infected partner that dies also gives rise, at an equal rate, to an individual entering the single susceptible compartment):
 
@@
 
-Similarly, $II_{ij}$ partnerships dissolve at a rate $-(\lambda_i + \lambda_j)$, but two cases, when $i \neq j$ and $i = j$, must be considered separately.
+Similarly, since $II_{ij}$ partnerships are composed of two infected partners, they dissolve at a per-partnership rate $(\lambda_i + \lambda_j)$.
+However, two cases, when $i \neq j$ and $i = j$, must be considered separately.
 When an $II_{ij}$ partnership dissolves due to disease-induced mortality, where $i \neq j$, the death of the partner with strain $i$ causes its partner to enter $I_j$ compartment at rate $\lambda_j II_{ij}$, and vice versa.
@@
 Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$.
-Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - II_{ij}$.
+Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - c II_{ij}$.
 Once individuals leave a partnership, they are instantaneously distributed into coupled states.
@@
 
-Individuals in a risk group $i$ leave single state at a rate of $\varphi_i \rho$.
+Individuals in a risk group $i$ leave single state at per-person rate $\varphi_i \rho$.
 Let $XY_{ij,kl}$ be a coupled state where $X$ and $Y$ are the infection status (susceptible or infected) of each partner, $k$ and $l$ are the risk groups $X$ and $Y$ belong to respectively, and $i$ and $j$ are the strains of an infected partner.
@@
 
-For dissolution process, individual in risk group $i$ leave partnership at a rate $\varphi_i c$.
+For dissolution process, individual in risk group $i$ expects to leave partnership at a rate $\varphi_i c$.
 If partnership is formed between two individuals from a different risk group, rate at which they leave the partnership differs.
 We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners.
-Therefore, $XY_{ij, kl}$ dissolve at a rate $\frac{\varphi_k + \varphi_l}{2} c XY_{ij, kl}$, and both $X_{i,k}$ and $Y_{j,l}$ return to single state at a rate of $\frac{\varphi_k + \varphi_l}{2} c XY_{ij, kl}$.
+Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $\frac{\varphi_k + \varphi_l}{2} c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate.
 
@@
 Then, a susceptible individual in risk group $i$ becomes infected through extra-couple and uncoupled contact at a per capita rate of $\sum_j \sum_k P_{k,j} X_{k,i}$.
-Once infected, individuals are distributed into each $\Lspvl$ categories through mutation.
+Once infected, individuals are distributed into each Strain categories through mutation.
 
@@
 \end{equation}
-Lastly, as model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution:
+As model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution:
 
@@
 Model 6 has the same distribution of initial infected individuals as model 5.
+
+Lastly, for heterogeneity model, we assume that the risk distribution of the population follows gamma distribution and calculate the shape and scale parameters given the mean and squared coefficient of variation.
+Using the shape and scale parameterws, we define gamma quantile function $Q(p)$ and $p_j =  \tsub{p}{min} + (\tsub{p}{max} - \tsub{p}{min}) \frac{j-1}{n_r + 1}$, where $n_r$ is number of risk groups and $j = 1, 2, 3, \dots, n_r + 1$.
+Since $Q(1) = \infty$, we set $\tsub{p}{max} = 0.99$ and $\tsub{p}{min} = 0.01$.
+Then, we define $\varphi_i = \frac{Q(p_j) + Q(p_{j+1})}{2}$.
+We define $R_i$ as the proportion of individuals in risk group $i$ in a disease free equilibrium and assume $R_i$ is equal for all $i$, i.e.
+$R_i = \frac{1}{n_r}$.
+In order to start the simulation in a quasi-equilibrium state, we first run the model with the following initial state:
+
+\begin{equation}
+\begin{aligned}
+S_{\cdot,i}(0) &= (1 - \epsilon) R_i,\\
+I_{k,i}(0) &= \epsilon D_k R_i,\\
+SS_{\cdot,ij}(0) &= SI_{\cdot k, ij}(0) = II_{kl,ij} (0) = 0.\\
+\end{aligned}
+\end{equation}
+For this particular simulation, we disregard infection process as well as disease induced mortality in order to preserve the strain distribution of infected individuals.
+Furthermore, since the scaling parameter, $\gamma$, does not affect the risk group distribution in the absence of disease transmission, we increase the scaling parameter to 5 ($\gamma = 5$) to speed up the simulation and run the model for 50 years.
+After the model has reached its quasi-equilibrium state, we take this distribution of susceptible and infected individuals as the initial state of the actual simulation.
 

### 2016-10-04 R/simulations/ev_LHSfull_assemble.R
first assembly of batch v3

@@ -1,7 +1,7 @@
 ## go to directory with individual batchfile outputs ...
 ## setwd("../../simdata/ev_LHSfull")
 
-tag <- "_v2"  ## batch version identifier
+tag <- "_v3"  ## batch version identifier
 nrun <- function(x) {
    r <- readLines(x)
    sum(grepl("([0-9.]+ ){5}",r))
@@ -50,12 +50,12 @@ if (length(rdafiles)==0) {
     c_names <- names(all_list[[1]])
     all_comb <- lapply(c_names,comb_fun)
     names(all_comb) <- c_names
-    save("all_comb",file="../ev_LHSfull_comb.rda")
+    save("all_comb",file=paste0("../ev_LHSfull_comb",tag,".rda"))
 }
 
 if (FALSE) {
     ## setwd("..") ## back up to main simdata dir
-    load("ev_LHSfull_comb.rda")
+    load(paste0("ev_LHSfull_comb",tag,".rda"))
     ## quick look at results ...
     pcol <- adjustcolor("black",alpha=0.1)
     matplot(all_comb$I_matFull,type="l",lty=1,col=pcol)
@@ -69,21 +69,4 @@ if (FALSE) {
     plot(density(na.omit(all_comb$peak_matFull[,1])))
     plot(density(na.omit(all_comb$peak_matFull[,2])))
     plot(density(all_comb$val_vecFull))
-    
-    ##little_r
-    
-    min(all_comb$I_matFull[2000,], na.rm = TRUE)
-    
-    Imat <- all_comb$I_matFull[,!is.na(all_comb$I_matFull[1,])]
-    
-    little_r <- unlist(apply(Imat, 2, function(x){
-    	df <- data.frame(t = 1:2000, I = x)
-    	l <- try(lm(log(I)~t, data = df, subset = I > 1e-3 & I < 1e-2))
-    	ifelse(!inherits(l,"try-error"), coef(l)[[2]], NA) 
-    }))
-    
-    plot(density(little_r, na.rm = TRUE))
-    matplot(Imat[,which(little_r < 0.05)], type = "l", lty = 1, col = pcol, log = "y")
-    length(which(little_r < 0.05))
-    length(which(little_r > 0.05))
 }

### 2016-10-07 ms/plos_latex.tex
working on abstract/intro

@@
   of virulence (SPVL) and the time at which the peak occurs across all
-  models and across a Latin hypercube sample that captures a realistic
-  range of partnership dynamic parameters for sub-Saharan Africa. In
+  models for a range of 
+  partnership dynamic parameters 
+  applicable to the HIV epidemic in sub-Saharan Africa. In
   order to account for the different interpretations of parameters
@@
   the simulated epidemic growth rate to be identical, matching a
-  realistic baseline value. For this
-  particular model setting, the simplest random-mixing structure is
+  realistic baseline value. 
+  Although SPVL trajectories are broadly similar
+  across model structures --- an initial increase in SPVL, 
+  followed by a peak and a decline to an equilibrium SPVL slightly
+  lower than a peak --- the timing and height of the peak vary
+  considerably.
+  Models of intermediate
+  complexity predicted much lower peak virulence (15 years to
+  progress to AIDS) compared to both more realistic models
+  and simple random-mixing models with no partnership structure
+  at all (both approx. 7.25 years to progress to AIDS).
+  In this range of models, the simplest random-mixing structure is
   actually the best approximation to the most realistic model; this
@@
 evolutionary biology, of both theoretical and (potentially) practical
-importance. The trade-off theory \cite{Ebert1999} --- which
-postulates that parasite virulence can be explained as the long-term
-evolutionary outcome of a saturating relationship between parasite
-clearance rate and transmission rate --- has been criticized
+importance. In general, evolutionary suggests that
+disease strains that can reproduce more, where reproduction
+means the amount of \emph{between-host} transmission, or
+the number of new hosts infected, will increase in prevalence.
+Pathogens can increase their between-host 
+fitness, or net reproduction rate, either
+by increasing their birth rate, the rate at which they
+infect new hosts, or by decreasing their death rate, the rate
+at which hosts recover or die from disease.
+The trade-off theory \cite{Ebert1999} --- postulates that
+the birth and death rate are both linked to the rate at which
+the pathogen reproduces the host, and that the pathogen
+evolution will thus strike a balance between the
+pathogen's rate of transmission
+to new hosts and its rate of killing its host (or of provoking
+the host's immune system to eliminate it).
+The tradeoff theory has been criticized
 \cite{EbertBull2003,alizon_adaptive_2015}, but has also been
@@
 derive analytical correspondences among them. Given a numerical
-criterion, such as $r$ (initial exponential growth rate) or ${\cal
-  R}_0$ (intrinsic reproductive number), we can adjust one or more
+criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
+(intrinsic reproductive number), we can adjust one or more
 parameters by brute force to ensure that all of the models match
-according to that criterion. While ${\cal R}_0$ is often considered
+according to that criterion. While $\rzero$ is often considered
 the most fundamental property of an epidemic, and might thus seem to
@@
 the transient evolutionary dynamics of virulence, which are more
-strongly affected by $r$ than ${\cal R}_0$. Second, $r$ is 
+strongly affected by $r$ than $\rzero$. Second, $r$ is 
 more directly observable in real epidemics; $r$ can be estimated by
 fitting an exponential curve to the initial incidence or
-prevalence curves \cite{ma_estimating_2014}, while ${\cal R}_0$
+prevalence curves \cite{ma_estimating_2014}, while $\rzero$
 typically requires either (1) knowledge of \emph{all} epidemic
@@
 that the implicit model derivation is based on defining
-the force of infection to match a scaled version of ${\cal R}_0$,
+the force of infection to match a scaled version of $\rzero$,
 and as such would be expected to match the equilibrium behaviour

### 2016-10-09 figures/figures.R
new figures

@@ -4,7 +4,6 @@ library("ggplot2"); theme_set(theme_bw(base_size = 12,
                                        base_family = "Times"))
 library("dplyr")
 library("tidyr")
-opts_chunk$set(echo=FALSE,error=FALSE)
 if (.Platform$OS.type=="windows") {
     windowsFonts(Times=windowsFont("Times"))
 } 
@@ -23,13 +22,16 @@ library("GGally") ## need BMB version
 ## devtools::install_github("bbolker/GGally")
 ## devtools::install_github("lionel-/ggstance")
 library("ggstance")
+source("../R/hivFuns.R")
+source("../R/Param.R")
 
 ## load data
-load("../simdata/combine_ev_LHS4.rda")
+load("../simdata/combineResults.rda")
 
 ## setup2
-orig_sum_labs <- c("peak_time","peak_vir","eq_vir","rel_peak")
-new_sum_labs <- c("peak time","peak SPVL","equilibrium SPVL","relative peak")
+orig_sum_labs <- c("peak_time","peak_dur","eq_dur","rel_dur")
+new_sum_labs <- c("time at which progression time is minimal (years)","minimum mean progression time (years)",
+									"equilibrium mean progression time (years)","relative ratio: minimum/equilibrium progression time")
 fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
     if (atop) new_sum_labs <-
                   gsub("(.*) (.*)","atop(\\1,\\2)",new_sum_labs)
@@ -128,9 +130,6 @@ fig_objects <- c(fig_objects,"gg_virtraj")
 
 ### Figure 2.1 - Transmission rate
 
-source("../R/hivFuns.R")
-source("../R/Param.R")
-
 returnBeta <- function(lSpvl, parameters){
 	with(as.list(c(parameters)),{
 		v = 10^lSpvl
@@ -195,13 +194,31 @@ ggsave(gg_durtraj,file="fig_S2_2.png",width=6,height=4,dpi=400)
 
 ### Figure 3
 
-sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
-sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
-mL <- melt(sL,id.vars=c("model","run"))
[…]

### 2016-10-09 ms/plos_latex.tex
cite
tweak to ms

@@
 where $V_{\alpha} = 10^\alpha$. 
-The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$.
-
-Although such simplification affects evolution of virulence, we show that our simplified model is able to produce result that is qualitatively similar to that of Shirreff \etal\'s \cite{shirreff_transmission_2011}.
+The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix).
+Although such simplification affects evolution of virulence, we show that our simplified model is able to produce result that is qualitatively similar to that of Shirreff \etal's \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}).
+As our focus is not making accurate predictions, we believe that being able to produce qualitatively similar dynamics is sufficient for this study.
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig1.pdf}
+\caption{{\bf Baseline dynamics.}
+Time series of mean population \Lspvl\.
+(a) Shirreff model, effects of varying exponential growth rate, $r$.
+(b) Effects of varying initial infectious density $I(0)$.
+(c) Effects of varying initial mean virulence $\alpha(0)$.
+Orange curve in panel (a) shows that our simplified model is able to produce similar trajectory compared to Shirreff \etal's model using the same parameter set \cite{shirreff_transmission_2011}.
+In order to compare the models under a similar environment, we match the initial exponential growth rate (see main text "Simulation runs").
+Furthermore, we experiment with wide ranges of initial values to show that evolutionary dynamics are sensitive to starting parameters but maintains qualitative features.}
+\label{fig:panel3}
+\end{figure}
 
@@
 a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width = 0.25 \Lspvl), which we use for all other simulations.
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig1.pdf}
-\caption{{\bf Baseline dynamics.}
-Time series of mean population virulence (\Lspvl).
-(a) Shirreff model, effects of varying $r$.
-(b) Effects of varying initial infectious density $I(0)$.
-(c) Effects of varying initial mean virulence $\alpha(0)$.}
-\label{fig:panel3}
-\end{figure}
 
@@
 However, to reflect the effect of (instantaneously formed) partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}.
-Finally, the last model (``random'') is a simple random-mixing model.
-
-Lastly, we add heterogeneous mixing to pairform+epc model.
+Another model (``random'') is a simple random-mixing model.
+
+Lastly, we add heterogeneous mixing to pairform+epc model (``hetero'').
 Individuals are divided into different risk groups based on the sexual activity level, and we assume that sexual activity level is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year [CITE].
@@
 $n$ & Number of strains & 21 (51*) & Assumption\\
+$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & CITE\\
+$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & CITE\\
 \hline
@@
 
-Champredon \etal\ define the proportion of infectious extra-couple and uncoupled contact through the following term:
+Champredon \etal\ \cite{champredon_hiv_2013} define the proportion of infectious extra-couple and uncoupled contact through the following term:
 

### 2016-10-11 ms/plos_latex.tex
equation > model

@@
 
-Finally, combining all these equations give us the full model, which is Model 1. We can simply drop the uncoupled and extra-couple transmission terms to obtain equation 2:
+Finally, combining all these equations give us the full model, which is Model 1. We can simply drop the uncoupled and extra-couple transmission terms to obtain model 2:
 
@@
 
-Combining all these dynamics, we have equation 3. If we remove extra-couple transmission, we have equation 4.
+Combining all these dynamics, we have model 3. If we remove extra-couple transmission, we have model 4.
 

### 2016-10-11 R/assemble.R
working on figs

@@ -1,0 +1,62 @@
+simdatapath <- "../simdata"
+source("hivFuns.R")  ## must be loaded before LHSparms3
+source("LHSparms3.R")
+ltab <- data.frame(run=seq(nrow(ltab)),ltab)
+##  combine them/rename appropriately
+run_nums <- c("pairform+epc"=5,"pairform"=6,
+              "instswitch"=7,"instswitch+epc"=8,"implicit"=9)
+####              "heterogeneous"=10)
+## each entry has
+##    I_mat*:
+##    vir_mat*:
+##    eq_vec*: equilibrium value
+##    peak_mat*: peak value (time, height)
+##    val_vec*: ???
+for (i in run_nums)
+    L <- load(file.path(simdatapath,paste0("ev_LHS_res",i,".rda")))
+## combine
+component_names <- c("I_mat","vir_mat","eq_vec","peak_mat","val_vec")
+## make one big nested list of components
+comp_list <- lapply(component_names,
+                    function(c) {
+    res <- lapply(paste0(c,run_nums),get)
+    setNames(res,names(run_nums))
+})
+names(comp_list) <- component_names
+mat_names <- c("I_mat","vir_mat")
+## reduce matrices to melted data frame of mean & quantiles
+##  (we don't really need to make giant pictures with all of the
+##   individual lines - ribbons with mean and 95% quantiles should
+##   be sufficient)
+sum_mat_fun <- function(x) {
+    data.frame(
+        tvec=1:nrow(x),  ## ?? better time index?
+        mean=rowMeans(x),
+        lwr=apply(x,1,quantile,0.025),
+        upr=apply(x,1,quantile,0.975))
+}
+sum_list <- list()
+for (m in mat_names) {
+    sum_list[[m]] <- ldply(comp_list[[m]],
+                           sum_mat_fun, .id="model")
+}
+s1 <- ldply(comp_list[["eq_vec"]],
+                function(x) data.frame(run=1:length(x),eq_vir=x),
+                .id="model")
+s2 <- ldply(comp_list[["peak_mat"]],
+                function(x) {
+                   x <- setNames(data.frame(x),c("peak_time","peak_vir"))
+                   data.frame(run=1:nrow(x),x)
+                 },
+                .id="model")
+sum_list[["sum_mat"]] <- full_join(s1,s2)
+## unlike merge(), full_join() preserves run order
+```
+
[…]

### 2016-10-11 figures/figures.R

@@ -28,6 +28,16 @@ library("ggstance")
 load("../simdata/combine_ev_LHS4.rda")
 load("../simdata/combineResults.rda")
 
+source("../R/hivFuns.R")
+source("../R/Param.R")
+
+returnBeta <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(K,Bmax,v,a))
+	})
+}
+
 ## setup2
 orig_sum_labs <- c("peak_time","peak_vir","eq_vir","rel_peak")
 new_sum_labs <- c("peak time","peak SPVL","equilibrium SPVL","relative peak")
@@ -108,6 +118,15 @@ ss <- subset(sum_list[["vir_mat"]],tvec<750)
 
 ss$model <- factor(ss$model, c("random","pairform+epc", "pairform","instswitch+epc","instswitch","implicit","heterogeneous"))
 
+
+ss <- transform(ss,
+                cmodel=droplevels(factor(gsub("+epc","",model,fixed=TRUE),
+                                         levels=levels(model))))
+
+ss2 <- transform(ss, mean = returnBeta(mean,HIVpars.skeleton),
+                 lwr = returnBeta(lwr,HIVpars.skeleton),
+                 upr = returnBeta(upr,HIVpars.skeleton))
+                                 
 gg_virtraj <- ggplot(ss,
                      aes(tvec,mean,ymin=lwr,ymax=upr))+
 	geom_line(aes(colour=model,linetype=model),lwd=1)+
@@ -117,13 +136,18 @@ gg_virtraj <- ggplot(ss,
 	theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
 		axis.title.x = element_text(margin = margin(10,0,0,0)),
 		legend.position = "none") +
-	facet_wrap(~model, nrow = 2) +
+	facet_wrap(~cmodel, nrow = 2) +
 	zero_margin
 
-## experiment: collapse pairs, + direct labeling???
 
 library(directlabels)
-direct.label(gg_virtraj+expand_limits(x=1000),method="top.bumpup")
+## like top.points, but with vjust=0.5
+top.points2 <- gapply.fun(transform(d[which.max(d$y), ],
+                               ## *smaller* numbers move labels right/up
+                                    hjust = 0.1, vjust = -1))
+
[…]

### 2016-10-11 figures/figures.R
updated fig 2; no strips, y=progression time

@@ -31,6 +31,13 @@ load("../simdata/combineResults.rda")
 source("../R/hivFuns.R")
 source("../R/Param.R")
 
+returnDur <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(v,Dmax,D50,h))
+	})
+}
+
 returnBeta <- function(lSpvl, parameters){
 	with(as.list(c(parameters)),{
 		v = 10^lSpvl
@@ -123,9 +130,9 @@ ss <- transform(ss,
                 cmodel=droplevels(factor(gsub("+epc","",model,fixed=TRUE),
                                          levels=levels(model))))
 
-ss2 <- transform(ss, mean = returnBeta(mean,HIVpars.skeleton),
-                 lwr = returnBeta(lwr,HIVpars.skeleton),
-                 upr = returnBeta(upr,HIVpars.skeleton))
+ss2 <- transform(ss, mean = returnDur(mean,HIVpars.skeleton),
+                 lwr = returnDur(lwr,HIVpars.skeleton),
+                 upr = returnDur(upr,HIVpars.skeleton))
                                  
 gg_virtraj <- ggplot(ss,
                      aes(tvec,mean,ymin=lwr,ymax=upr))+
@@ -146,7 +153,11 @@ top.points2 <- gapply.fun(transform(d[which.max(d$y), ],
                                ## *smaller* numbers move labels right/up
                                     hjust = 0.1, vjust = -1))
 
-direct.label(gg_virtraj %+% ss2,
+nostrips <- theme(strip.background = element_blank(),
+                  strip.text.x = element_blank())
+## http://stackoverflow.com/questions/10547487/r-removing-facet-wrap-labels-completely-in-ggplot2
+direct.label(gg_virtraj %+% ss2 +
+             labs(y="progression to AIDS (years)")+nostrips,
              method=list("top.points2","bumpup"))
 ## r fig2, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400}
 
@@ -180,12 +191,6 @@ ggsave(gg_betatraj,file="fig_S2_1.png",width=6,height=4,dpi=400)
 
 ### Figure 2.2 - Duration
 
-returnDur <- function(lSpvl, parameters){
-	with(as.list(c(parameters)),{
-		v = 10^lSpvl
-		return(hill(v,Dmax,D50,h))
-	})
-}
[…]

### 2016-10-14 ms/plos_latex.tex
bib update

@@
 Lastly, we add heterogeneous mixing to pairform+epc model (``hetero'').
-Individuals are divided into different risk groups based on the sexual activity level, and we assume that sexual activity level is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year [CITE].
+Individuals are divided into different risk groups based on the sexual activity level, and we assume that sexual activity level is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics}.
 Model details can be found in the appendix.
@@
 $n$ & Number of strains & 21 (51*) & Assumption\\
-$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & CITE\\
-$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & CITE\\
+$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
+$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
 \hline
@@
 
+\bibitem{omori2015dynamics}
+Omori R, Chemaitelly H, Abu-Raddad LJ.
+\newblock Dynamics of non-cohabiting sex partnering in sub-Saharan Africa: a
+  modelling study with implications for HIV transmission.
+\newblock Sexually transmitted infections. 2015; p. sextrans--2014.
+
 \bibitem{blower_drugs_1991}

### 2016-10-14 ms/plos_latex.tex
working ...

@@
   phase of an epidemic, but quantitative outcomes
-  can depend sensitively on biological details and the structure of
-  mathematical models used to capture them.  Fraser, Shirreff, and
+  may depend sensitively on biological details and the structure of
+  mathematical models used to capture them.  Fraser and
   co-workers have proposed a series of models for eco-evolutionary
@@
   dissolution and allowing for individuals to transmit disease outside
-  of partnerships. We assess summary statistics such as the peak value
-  of virulence (SPVL) and the time at which the peak occurs across all
+  of partnerships. We assess summary statistics such as the minimum
+  time of progression to AIDS and the time at which this 
+  extreme value (peak virulence) occurs across all
   models for a range of 
@@
   realistic baseline value. 
-  Although SPVL trajectories are broadly similar
-  across model structures --- an initial increase in SPVL, 
-  followed by a peak and a decline to an equilibrium SPVL slightly
-  lower than a peak --- the timing and height of the peak vary
+  Although virulence trajectories are broadly similar
+  across model structures --- an initial increase in virulence, 
+  followed by a peak and a decline to an equilibrium virulence slightly
+  lower than a peak --- the timing and magnitude of the peak vary
   considerably.
@@
 
-The evolution of pathogen virulence is a fundamental process in
-evolutionary biology, of both theoretical and (potentially) practical
-importance. In general, evolutionary suggests that
-disease strains that can reproduce more, where reproduction
-means the amount of \emph{between-host} transmission, or
-the number of new hosts infected, will increase in prevalence.
-Pathogens can increase their between-host 
-fitness, or net reproduction rate, either
-by increasing their birth rate, the rate at which they
+The evolution of pathogen virulence has both 
+theoretical and, potentially, practical
+importance. In general, evolutionary theory suggests that
+disease strains that can reproduce more --- where reproduction
+is defined the amount of \emph{between-host} transmission, or
+the number of new hosts infected --- will increase in prevalence.
+Pathogens can increase their net reproduction rate either
+by increasing their birth rate, the rate (per infected host) at which they
 infect new hosts, or by decreasing their death rate, the rate
 at which hosts recover or die from disease.
-The trade-off theory \cite{Ebert1999} --- postulates that
+The trade-off theory \cite{anderson_coevolution_1982,Ebert1999} --- postulates that
 the birth and death rate are both linked to the rate at which
-the pathogen reproduces the host, and that the pathogen
+the pathogen reproduces within the host, and that the pathogen
 evolution will thus strike a balance between the
@@
 the host's immune system to eliminate it).
-The tradeoff theory has been criticized
-\cite{EbertBull2003,alizon_adaptive_2015}, but has also been
-successfully applied in a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}. One
-particularly interesting application of these ideas is the work by
-Fraser \etal\ showing that HIV appears to satisfy the prerequisites of
+Some biologists have criticized the tradeoff theory
+\cite{EbertBull2003,alizon_adaptive_2015}, but others have
+successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
+Fraser \etal\ have applied these ideas in a particularly
+interesting way 
+by showing that HIV appears to satisfy the prerequisites of
 the tradeoff theory: in studies of discordant couples (i.e. long-term
@@
 
-The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted
-transient peak in HIV virulence midway through the epidemic,
+The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted transient peak in HIV virulence midway through the epidemic,
 highlights the importance of interactions between epidemiological and
@@
 
-We add complexity to the epidemiological model following the general approach of Champredon \etal\ \cite{champredon_hiv_2013}; individuals join and leave partnerships at a specified rate, and can have sexual contact both within and outside of established partnerships.
-In order to explore how virulence evolution depends on epidemiological structure, we consider a series of models with increasing levels of complexity.
-In order to avoid dependence of the results on a particular set of parameters --- as we explain below, finding matching sets of parameters across models with widely differing epidemiological structures is challenging --- we evaluate our models across a wide range of parameters, again following Champredon \etal\ \cite{champredon_hiv_2013} in using a Latin hypercube design.
-For each model run, we compute a set of metrics (peak virulence, timing of virulence peak, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.
-
-As our primary goal is to explore how different epidemiological structures (i.e.
-partnership dynamics and contact structures) affect our conclusions about the evolution of virulence, our models use a simplified description of within-host dynamics and heritability derived from
-Shirreff \etal's multi-strain evolutionary model \cite{shirreff_transmission_2011}.
-Like Shirreff \etal, we use a simple susceptible-infected-susceptible demographic formulation; rather than modeling birth and death (or more specifically, recruitment into the sexually active population and death), we assume that whenever an individual dies from infection, another enters the susceptible compartment.
+We add complexity to the epidemiological model following the general
+approach of Champredon \etal\ \cite{champredon_hiv_2013},
+which is in turn based on work of Dietz and Hadeler
+\cite{dietz_epidemiological_1988}; individuals
+join and leave partnerships at a specified rate, and can have sexual
+contact both within and outside of established partnerships. In order
+to explore how virulence evolution depends on epidemiological
+structure, we consider a series of models with increasing levels of
+complexity. In order to avoid dependence of the results on a
+particular set of parameters --- as we explain below, finding matching
+sets of parameters across models with widely differing epidemiological
+structures is challenging --- we evaluate our models across a wide
+range of parameters, again following Champredon
+\etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
+design. For each model run, we compute a set of metrics (peak
+virulence, timing of virulence peak, equilibrium virulence) that
+summarize the evolutionary trajectory of a simulated HIV epidemic.
+
+As our primary goal is to explore how different epidemiological
+structures (i.e. partnership dynamics and contact structures) affect
+our conclusions about the evolution of virulence, our models use a
+simplified description of within-host dynamics and heritability
+derived from Shirreff \etal's multi-strain evolutionary model
+\cite{shirreff_transmission_2011}. Like Shirreff \etal, we use a
+simple susceptible-infected-susceptible demographic formulation;
+rather than modeling birth and death (or more specifically,
+recruitment into the sexually active population and death), we assume
+that whenever an individual dies from infection, another enters the
+susceptible compartment.
 
@@
 The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix).
-Although such simplification affects evolution of virulence, we show that our simplified model is able to produce result that is qualitatively similar to that of Shirreff \etal's \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}).
+Although such simplification could certainly affect the evolution of virulence (Kret\cite{kretzschmar_effect_1998} , our simplified model is able to produce qualitatively similar results that is qualitatively similar to that of Shirreff \etal's \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}).
 As our focus is not making accurate predictions, we believe that being able to produce qualitatively similar dynamics is sufficient for this study.

### 2016-10-17 ms/plos_latex.tex
still working ...

@@
 \newcommand{\DM}{\textrm{DM}}
+
+\newcommand{\todo}[1]{\textbf{#1}}
 
@@
   Although virulence trajectories are broadly similar
-  across model structures --- an initial increase in virulence, 
-  followed by a peak and a decline to an equilibrium virulence slightly
-  lower than a peak --- the timing and magnitude of the peak vary
+  across model structures --- an initial increase in virulence/decrease
+  in progression time, followed by a minimum and an increase to an equilibrium 
+  progression time slightly higher than the minimum --- the timing and magnitude of the minimum vary
   considerably.
@@
 
-As we discuss in detail below, the
+As we discuss in detail below,
 existing models of HIV eco-evolutionary dynamics either use implicit
@@
 \etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
-design. For each model run, we compute a set of metrics (peak
-virulence, timing of virulence peak, equilibrium virulence) that
-summarize the evolutionary trajectory of a simulated HIV epidemic.
+design. For each model run, we compute a set of metrics (minimum
+progression time/peak virulence, timing of maximum virulence, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.
 
@@
 \end{split}
+\label{eq:hillfuns}
 \end{equation}
@@
 The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix).
-Although such simplification could certainly affect the evolution of virulence (Kret\cite{kretzschmar_effect_1998} , our simplified model is able to produce qualitatively similar results that is qualitatively similar to that of Shirreff \etal's \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}).
-As our focus is not making accurate predictions, we believe that being able to produce qualitatively similar dynamics is sufficient for this study.
+Simplifying the model of HIV pathogenesis from three stages to a single stage could affect our conclusions about the evolution of virulence (e.g.
+Kretzschmar and Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics and multiple stages of infectivity have interactive effects on $\rzero$).
+However, our simplified model produces results that are qualitatively similar to those of Shirreff \etal's \cite{shirreff_transmission_2011} model; when our model is calibrated to have a similar initial epidemic growth rate $r$, the peak \Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly higher (4.7 vs.
+4.3); the relative height of the peak is \todo{compute difference and fill this in}
+(\figurename~\ref{fig:panel3}).
 
@@
 \caption{{\bf Baseline dynamics.}
-Time series of mean population \Lspvl\.
-(a) Shirreff model, effects of varying exponential growth rate, $r$.
+Time series of mean population \Lspvl.
+(a) Contrast with three-stage Shirreff model and effects of varying exponential growth rate, $r$.
 (b) Effects of varying initial infectious density $I(0)$.
 (c) Effects of varying initial mean virulence $\alpha(0)$.
-Orange curve in panel (a) shows that our simplified model is able to produce similar trajectory compared to Shirreff \etal's model using the same parameter set \cite{shirreff_transmission_2011}.
-In order to compare the models under a similar environment, we match the initial exponential growth rate (see main text "Simulation runs").
-Furthermore, we experiment with wide ranges of initial values to show that evolutionary dynamics are sensitive to starting parameters but maintains qualitative features.}
+The orange curve in panel (a) shows that our simplified model can produce similar trajectories to those Shirreff \etal's model \cite{shirreff_transmission_2011}.
+In order to compare the models, we match the initial exponential growth rate (see main text ``Simulation runs").
+Furthermore, we experiment with wide ranges of initial values (panels b and c) to show that evolutionary dynamics are not overly sensitive to starting parameter values, which we hold constant in our simulations.}
 \label{fig:panel3}
@@
 
-Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping.
+Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL.
+We simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping rather
+than allowing for variation in phenotypes of a single genotype.
 The mutational process in our model is directly taken from Shirreff \etal.
 Over the course of infection, mutation occurs within the host.
-However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection (and is not further affected by within-host mutation).
-Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual.
-First, the distribution of \Lspvl\ is discretized into a vector:
+However, it is assumed that SPVL of the strain transmitted by an infected individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation.
+Instead, the mutational effect takes place in a single
+step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
 \begin{equation}
@@
 We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e.
-a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width = 0.25 \Lspvl), which we use for all other simulations.
+a bin width of 0.1 \Lspvl\ for $\alpha$), but reducing $n$ to 21 (bin width = 0.25 \Lspvl) makes little difference; we use this coarser grid for all other simulations reported.
 
@@
 positive extra-partnership sexual contact and transmission and (3) homogeneous vs.
-heterogeneous mixing on the evolution of mean \Lspvl.
+heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
 
@@
 The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission.
-The next two models assume instantaneous partnership formation (``instswitch'') and thus consist of only the three partnered states: $SS$, $SI$, and $II$.
-Like the first two models, these models differ in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and the fourth (``instswitch'') only considers within-couple transmission.
-
-The next two models do not explicitly track sexual partnerships.
+
+The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and simpler, implicit models, assume instantaneous partnership formation (``instswitch'').
+The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
+Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.
+
+The next two models represent extreme simplifications of sexual partnership dynamics.
 One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
 It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
-However, to reflect the effect of (instantaneously formed) partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}.
-Another model (``random'') is a simple random-mixing model.
-
-Lastly, we add heterogeneous mixing to pairform+epc model (``hetero'').
-Individuals are divided into different risk groups based on the sexual activity level, and we assume that sexual activity level is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics}.
-Model details can be found in the appendix.
+However, to reflect the effect of partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}.
+The second model of this pair (``random'') is a simple random-mixing model.
+
+Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero'').
+Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix for model details).
+We assume random activity-weighted mixing between risk groups \todo{cite old Anderson-May HIV paper?}.
+While this model lacks some
+important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, we see it as a first step toward assessing the effects of epidemiological complexity.
+As even these models are pushing the limits of compartmental-based models, adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
 
@@
 \caption{Parameter ranges/values.
-Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples.
+Values of $c$ and $\rho$ are doubled from those given by Champredon \etal; we keep track of individuals in the model, while they keep track of couples.
 Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
@@
 
+\todo{add something about choice of heterogeneity}
+
 \subsection*{Simulation runs}
@@
 
-For each model we derive the following summary statistics: peak virulence, peak time, equilibrium virulence, and relative peak virulence.
-The transient phase of an epidemic is often characterized by high virulence, and we define peak virulence as the maximum virulence during this phase.
-It is simply calculated by taking the maximum value from the virulence trajectory, and peak time is the time at which the maximum value is reached.
-Once the epidemic enters the endemic phase, evolution of virulence stabilizes and reaches equilibrium.
-Equilibrium virulence is calculated by taking the mean virulence at 4000 years.
+Although each disease strain's core characteristic is its SPVL, the
+SPVL has one-to-one correspondences (based on eq.~\ref{eq:hillfuns})
+with both the expected time to progression to AIDS and with the rate
+(probability per unit time) of HIV transmission. Because the time to
+progression (measured in years) is easier to interpret than
+SPVL (measured in \Lspvl\ units), we summarize the virulence
+trajectories for each model run in terms of time to progression
+rather than SPVL. Because the  time to progression is inversely
+related to SPVL (increasing SPVL decreases the time to progression),
+the time to progression is technically measuring inverse
+virulence rather than virulence (we did not think that 
+reporting virulence as the rate of progression to AIDS, in units
+of $\textrm{years}^{-1}$, would help interpretability).
+For each model we derive the following summary statistics:
+minimum expected time to progression;
+time at which this minimum occurs
+(corresponding to peak virulence --- this is also the time at which the
+maximum rate of progression, maximum SPVL, and maximum transmission rate
+occur); equilibrium time to progression; 
+and the ratio of progression time at its minimum to the equilibrium
+value. Equilibrium progression time is calculated after 4000 years of simulated
+time.
 Although most simulations reach equilibrium much earlier, we set our time horizon at a much later date as some simulation runs have slow rate of evolution depending on the parameter set and model assumptions.
 
-We focus on these statistics for the following two reasons.
-First of all, knowing the possible ranges for the peak virulence allows us to estimate the worst-case scenario for the HIV and other sexually transmitted disease epidemics.
-Pathogens may already have evolved towards high virulence during the early stages of an epidemic, by the time it is observed by public health authorities.
-Understanding how virulent a pathogen can evolve before an epidemic begins can be helpful for controlling the disease.
-Furthermore, knowing the initial virulence, peak virulence, timing of the peak virulence, and equilibrium virulence provide sufficient detail to identify the shape of the virulence trajectory.
-During an epidemic outbreak, it is difficult to observe virulence evolution.
-Specifically, in the case of HIV and other sexually transmitted diseases, slow evolutionary time-scale makes observing changes in the mean virulence even more challenging.
-Knowing the ranges of these statistics can help real-time virulence evolution prediction during an epidemic less troublesome.
+Knowing the minimum progression time, timing of the minimum progression time/peak virulence, and equilibrium progression time provide sufficient detail to identify the overall shape of the virulence trajectory.
+In particular, knowing the timing of the peak virulence (how many years
+into the epidemic the virulence peaks) can help epidemiologists
+guess whether the virulence of an emerging pathogen is likely (1)
+to have peaked early, possibly even before the pathogen is detected
+spreading in the population, and decline over the remaining course
+of the epidemic; (2) to increase, peak, and decline over the
+foreseeable future; or (3) to increase very slowly, peaking only
+in the far future. To the extent that our simplistic model for HIV
+reflects reality, we would take the peak time of 150-300 years 
+(Figure~\ref{fig:panel3}c) to mean that, in the absence of
+treatment, the epidemic would probably still be increasing in virulence.
 
@@
 Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (\figurename~\ref{fig:virtraj}).
-Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 \Lspvl), virulence peaks in all models between 200 and 300 years.
+Although the speed of virulence evolution varies, leading to wide variation in the minimum progression time (means ranging from approximately \todo{fill in}), virulence peaks in all models between 200 and 300 years.
 
@@
 \includegraphics[width=\textwidth]{../figures/fig2.pdf}
-\caption{{\bf Envelopes of virulence trajectories under all models.}
+\caption{{\bf Envelopes of virulence trajectories (expected
+time of progression to AIDS) under all models.}
 All models were run until $t=4000~\textrm{years}$; truncated series are shown here.}
@@
 
-Our chosen summary statistics (peak time, peak virulence, equilibrium
-virulence, and relative peak virulence) all vary considerably across models
+Our chosen summary statistics (peak time, minimum progression time, equilibrium
+progression time, and relative progression time) all vary considerably across models
 \ref{fig:unidist}.
@@
 pair formation without extra-pair contact. Some parameter
-sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL);
+sets for these models lead to low equilibrium virulence ($\approx 18$ years
+to progression);
 these same sets lead to correspondingly low
-peak virulence ($< 3.5 \log_{10}$ SPVL) and early peak times (before 200 years), 
-but high relative peaks ($>1.3$)
-(\figurename~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
-At the opposite extreme, parameter sets that produce high equilibrium virulence 
-also produce late peaks ($> 200 \text{ years}$), 
-high peak virulence, and low relative peaks ($\approx 1.05$).
+peak virulence (16 years to progression) and early peak times (before 200 years).
+At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
+also produce late peaks ($> 200~\text{years}$) and
+high peak virulence (4 years to progression).
 The pair-formation without extra-pair contact and implicit models
@@
 least (random mixing) and the most complex (pair formation with
-extra-pair contact) models. The random-mixing model has lower
-variability, because it is unaffected by uncertainty in pair formation
-and extra-pair contact parameters, but otherwise the virulence
-dynamics of these two extreme models are remarkably similar.
+extra-pair contact, pairform+epc with heterogeneity) models.
+The random-mixing model has the lowest variability, because it is unaffected by uncertainty in pair formation and extra-pair contact parameters, but otherwise the virulence
+dynamics of these three extreme models are remarkably similar.
 This phenomenon is driven by the strong effects of extra-pair contact in the
@@
 that is effectively close to random.
-
-Expressing these results in terms of more directly interpretable epidemic
-parameters, i.e. using the Hill functions to translate \Lspvl\ to
-within-couple transmission probabilities and average time of progression
-to AIDS, shows that these differences are practically as well as 
+Once unstructured mixing is strong, adding realistic heterogeneity
+of mixing to the model has little effect other than increasing
+the variability in the outcomes.
+
+These differences are practically as well as 
 scientifically important. The random-mixing and pairform+epc models
@@
 random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model (Figs~???? show univariate distributions
+for the implicit model (Figs~???? \todo{fill in!} show univariate distributions
 on the epidemiological scales
@@
 
-Plotting the bivariate result distributions (\figurename~\ref{fig:pairplot})
-shows that most of the summary statistics
-are monotonically related, except those involving the relative peak virulence
-(bottom row). Changes in parameters that increase the equilibrium
-virulence initially increase the peak virulence even more, so
-that the relative peak virulence increases as well, but beyond
-an equilibrium virulence of about 2.5 \Lspvl\ the peak virulence
-increases slower than the equilibrium virulence, leading to a
-decrease in the relative peak virulence.
-
-The bivariate relationships also help distinguish the results of 
+The bivariate relationships (\figurename~\ref{fig:pairplot}) help distinguish the results of 
 different models with similar univariate dynamical summaries. While the
@@
 
-% \bibliography{virulence}
-\begin{thebibliography}{10}
-
-\bibitem{Ebert1999}
-Ebert D.
-\newblock The evolution and expression of parasite virulence.
-\newblock In: Stearns SC, editor. Evolution in Health \& Disease. New York:
-  Oxford University Press, Oxford, UK; 1999. p. 161--172.
-
-\bibitem{EbertBull2003}
-Ebert D, Bull JJ.
-\newblock Challenging the trade-off model for the evolution of virulence: is
-  virulence management feasible?
-\newblock Trends Microbiol. 2003;11(1):15--20.
-
-\bibitem{alizon_adaptive_2015}
-Alizon S, Michalakis Y.
-\newblock Adaptive virulence evolution: the good old fitness-based approach.
-\newblock Trends in Ecology \& Evolution. 2015;30(5):248--254.
-\newblock doi:{10.1016/j.tree.2015.02.009}.
-
-\bibitem{Dwyer+1990}
-Dwyer G, Levin SA, Buttel L.
-\newblock A simulation model of the population dynamics and evolution of
-  myxomatosis.
-\newblock Ecol Monog. 1990;60:423--447.
-
-\bibitem{mackinnon1999genetic}
-Mackinnon MJ, Read AF.
-\newblock Genetic relationships between parasite virulence and transmission in
-  the rodent malaria {{\em Plasmodium chabaudi}}.
-\newblock Evolution. 1999; p. 689--703.
-
-\bibitem{jensen2006empirical}
-Jensen KH, Little T, Skorping A, Ebert D.
-\newblock Empirical support for optimal virulence in a castrating parasite.
-\newblock PLoS Biol. 2006;4(7):e197.
-
-\bibitem{deroode2008virulence}
-De~Roode JC, Yates AJ, Altizer S.
-\newblock Virulence-transmission trade-offs and population divergence in
-  virulence in a naturally occurring butterfly parasite.
-\newblock Proceedings of the National Academy of Sciences.
-  2008;105(21):7489--7494.
-
-\bibitem{Fraser+2007}
-Fraser C, Hollingsworth TD, Chapman R, de~Wolf F, Hanage WP.
-\newblock Variation in {HIV}-1 set-point viral load: Epidemiological analysis
-  and an evolutionary hypothesis.
-\newblock PNAS. 2007;104:17441--17446.
-
-\bibitem{fraser_virulence_2014}
-Fraser C, Lythgoe K, Leventhal GE, Shirreff G, Hollingsworth TD, Alizon S,
-  et~al.
-\newblock Virulence and Pathogenesis of {HIV}-1 Infection: An Evolutionary
-  Perspective.
-\newblock Science. 2014;343(6177):1243727.
-\newblock doi:{10.1126/science.1243727}.
-
-\bibitem{shirreff_transmission_2011}
-Shirreff G, Pellis L, Laeyendecker O, Fraser C.
-\newblock Transmission Selects for {HIV-1} Strains of Intermediate Virulence: A
-  Modelling Approach.
-\newblock PLoS Computational Biology. 2011;7(10):e1002185.
-\newblock doi:{10.1371/journal.pcbi.1002185}.
-
-\bibitem{herbeck_hiv_2014}
-Herbeck JT, Mittler JE, Gottlieb GS, Mullins JI.
-\newblock An {HIV} Epidemic Model Based on Viral Load Dynamics: Value in
-  Assessing Empirical Trends in {HIV} Virulence and Community Viral Load.
-\newblock {PLoS} Comput Biol. 2014;10(6):e1003673.
-
-\bibitem{day_virulence_2004}
-Day T, Proulx SR.
-\newblock A General Theory for the Evolutionary Dynamics of Virulence.
-\newblock The American Naturalist. 2004;163(4):E40--E63.
-\newblock doi:{10.1086/382548}.
-
-\bibitem{alizon_price_2009}
-Alizon S.
-\newblock The {Price} equation framework to study disease within-host
-  evolution.
-\newblock Journal of Evolutionary Biology. 2009;22(5):1123--1132.
-\newblock doi:{10.1111/j.1420-9101.2009.01726.x}.
-
-\bibitem{champredon_hiv_2013}
-Champredon D, Bellan S, Dushoff J.
-\newblock {HIV} Sexual Transmission Is Predominantly Driven by Single
-  Individuals Rather than Discordant Couples: A Model-Based Approach.
-\newblock PLoS ONE. 2013;8(12):e82906.
-\newblock doi:{10.1371/journal.pone.0082906}.
-
-\bibitem{hollingsworth_hiv1_2008}
-Hollingsworth TD, Anderson RM, Fraser C.
-\newblock {HIV}-1 Transmission, by Stage of Infection.
-\newblock Journal of Infectious Diseases. 2008;198(5):687--693.
-\newblock doi:{10.1086/590501}.
-
-\bibitem{omori2015dynamics}
-Omori R, Chemaitelly H, Abu-Raddad LJ.
-\newblock Dynamics of non-cohabiting sex partnering in sub-Saharan Africa: a
-  modelling study with implications for HIV transmission.
-\newblock Sexually transmitted infections. 2015; p. sextrans--2014.
-
-\bibitem{blower_drugs_1991}
-Blower SM, Hartel D, Dowlatabadi H, Anderson RM, May RM.
-\newblock Drugs, Sex and {HIV}: A Mathematical Model for {New York City}.
-\newblock Philosophical Transactions of the Royal Society of London B:
-  Biological Sciences. 1991;331(1260):171--187.
-\newblock doi:{10.1098/rstb.1991.0006}.
-
-\bibitem{ma_estimating_2014}
-Ma J, Dushoff J, Bolker BM, Earn DJD.
-\newblock Estimating Initial Epidemic Growth Rates.
-\newblock Bulletin of Mathematical Biology. 2014;76(1):245--260.
-\newblock doi:{10.1007/s11538-013-9918-2}.
-
-\bibitem{wallinga_how_2007}
-Wallinga J, Lipsitch M.
-\newblock How generation intervals shape the relationship between growth rates
-  and reproductive numbers.
-\newblock Proceedings of the Royal Society of London B: Biological Sciences.
-  2007;274(1609):599--604.
-\newblock doi:{10.1098/rspb.2006.3754}.
-
-\bibitem{soetaert_solving_2010}
-Soetaert K, Petzoldt T, Setzer RW.
-\newblock Solving Differential Equations in {R}: Package {deSolve}.
-\newblock Journal of Statistical Software. 2010;33(9):1--25.
-\newblock doi:{10.18637/jss.v033.i09}.
-
-\bibitem{herbeck2016evolution}
-Herbeck J, Mittler J, Gottlieb G, Goodreau S, Murphy J, Cori A, et~al.
-\newblock Evolution of {HIV} virulence in response to widespread scale up of
-  antiretroviral therapy: a modeling study.
-\newblock bioRxiv. 2016; p. 039560.
-
-\bibitem{wawer2005rates}
-Wawer MJ, Gray RH, Sewankambo NK, Serwadda D, Li X, Laeyendecker O, et~al.
-\newblock Rates of {HIV}-1 transmission per coital act, by stage of {HIV}-1
-  infection, in {Rakai}, {Uganda}.
-\newblock Journal of Infectious Diseases. 2005;191(9):1403--1409.
-
-\bibitem{faria_early_2014}
-Faria NR, Rambaut A, Suchard MA, Baele G, Bedford T, Ward MJ, et~al.
-\newblock The early spread and epidemic ignition of {HIV}-1 in human
-  populations.
-\newblock Science (New York, NY). 2014;346(6205):56--61.
-\newblock doi:{10.1126/science.1256739}.
-
-\bibitem{elderd_uncertainty_2006}
-Elderd BD, Dukic VM, Dwyer G.
-\newblock Uncertainty in predictions of disease spread and public health
-  responses to bioterrorism and emerging diseases.
-\newblock Proceedings of the National Academy of Sciences. 2006;103(42):15693
-  --15697.
-\newblock doi:{10.1073/pnas.0600816103}.
-
-\bibitem{herbeck_is_2012}
-Herbeck JT, Müller V, Maust BS, Ledergerber B, Torti C, Di~Giambenedetto S,
-  et~al.
-\newblock Is the virulence of {HIV} changing? {A} meta-analysis of trends in
-  prognostic markers of {HIV} disease progression and transmission.
-\newblock AIDS (London, England). 2012;26(2):193--205.
-\newblock doi:{10.1097/QAD.0b013e32834db418}.
-
-\bibitem{payne_impact_2014}
-Payne R, Muenchhoff M, Mann J, Roberts HE, Matthews P, Adland E, et~al.
-\newblock Impact of {HLA}-driven {HIV} adaptation on virulence in populations
-  of high {HIV} seroprevalence.
-\newblock Proceedings of the National Academy of Sciences.
-  2014;111(50):E5393--E5400.
-\newblock doi:{10.1073/pnas.1413339111}.
-
-\bibitem{roberts2015impact}
-Roberts HE, Goulder PJ, McLean AR.
-\newblock The impact of antiretroviral therapy on population-level virulence
-  evolution of {HIV}-1.
-\newblock Journal of The Royal Society Interface. 2015;12(113):20150888.
-
-\end{thebibliography}
-
+\bibliography{virulence}
 \end{document}

### 2016-10-17 figures/figures.R

@@ -2,6 +2,7 @@
 library("knitr")
 library("ggplot2"); theme_set(theme_bw(base_size = 12,
                                        base_family = "Times"))
+library("MASS")  ## must be before dplyr (select)?
 library("dplyr")
 library("tidyr")
 if (.Platform$OS.type=="windows") {
@@ -10,13 +11,13 @@ if (.Platform$OS.type=="windows") {
 zero_margin <- theme(panel.margin=grid::unit(0,"lines"))
 zero_x_margin <-
     theme(panel.margin.x=grid::unit(0, "lines"))
-library("MASS")
 ## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
 scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
 library("gridExtra")
 library("plyr")  ## for ldply()
 library("dplyr") ## for full_join()
+library("magrittr") ## for full_join()
 library("reshape2")  ## for melt()
 library("GGally") ## need BMB version
 ## devtools::install_github("bbolker/GGally")
@@ -55,7 +56,7 @@ shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
 shirreff_df_m$run <- "Shirreff"
 df_tot <- rbind(shirreff_df_m,d_littleR_m)
 tlab <- "time (years)"
-g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
+g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run,linetype=run))+
     geom_line()+labs(x=tlab,y=expression(population~mean~set-point~viral~load~(log[10])))+
     theme(legend.position=c(0.75,0.25))
 ###
@@ -213,7 +214,7 @@ sL$model <- factor(sL$model, c("random", "hetero","pairform+epc", "pairform", "i
 sL.mod <- transform(sL, peak_dur = returnDur(peak_vir,HIVpars.skeleton),
 										eq_dur = returnDur(eq_vir,HIVpars.skeleton))
 sL.mod <- transform(sL.mod, rel_dur = peak_dur/eq_dur)
-sL.mod <- sL.mod[,-c(3, 5, 6)]
[…]

### 2016-10-18 figures/figures.R
tweaking ms

@@ -352,7 +352,7 @@ ggp1 <- ggpairs(sL2,
         mapping = ggplot2::aes(color = model,pch=model),
         columns=3:5,
         legends=TRUE,
-        lower = list(continuous = wrap("points",size=0.2)), ## alpha = 0.3,size=0.5)),
+        lower = list(continuous = wrap("points",alpha=0.6,size=1)), ## alpha = 0.3,size=0.5)),
         diag = list(continuous = "blankDiag"),
         upper = list(continuous = "blank"))
 ## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot

### 2016-10-18 ms/plos_latex.tex
tweaking ms
add latexdiff comparison

@@
 The evolutionary dynamics of parasite virulence over the
-  course of an emerging epidemic have important implications both for
-  our basic understanding of epidemiological dynamics and,
-  potentially, for the outcomes of public health interventions.
-  Changes in the fitness landscape
-  will generally select for higher virulence during the early
-  phase of an epidemic, but quantitative outcomes
+course of an emerging epidemic have important implications for
+our understanding of epidemiological dynamics.
+While changes in the fitness landscape
+generally select for higher virulence early in an
+epidemic, quantitative outcomes
 may depend sensitively on biological details and the structure of
-  mathematical models used to capture them.  Fraser and
-  co-workers have proposed a series of models for eco-evolutionary
-  dynamics of HIV that are relatively detailed in their portrayal of
+mathematical models used to portray them.  Fraser \emph{et al.} 
+have proposed a model for the eco-evolutionary
+dynamics of HIV that captures
 the tradeoffs between transmission and virulence (mediated by
@@
 partnership dynamics that previous research has found to be critical
-  in driving epidemics of sexually transmitted diseases.  We
-  explore models that combine HIV virulence tradeoffs with a range of
+in driving epidemics of sexually transmitted diseases.  
+Our models combine HIV virulence tradeoffs with a range of
 epidemiological structures, modeling partnership formation and
 dissolution and allowing for individuals to transmit disease outside
-  of partnerships. We assess summary statistics such as the minimum
-  time of progression to AIDS and the time at which this 
-  extreme value (peak virulence) occurs across all
+of partnerships. We assess summary statistics such as the peak virulence
+(corresponding to the minimum expected time of progression to AIDS) across all
 models for a range of 
 partnership dynamic parameters 
-  applicable to the HIV epidemic in sub-Saharan Africa. In
-  order to account for the different interpretations of parameters
-  across model structures, we scale all parameter sets to constrain
-  the simulated epidemic growth rate to be identical, matching a
-  realistic baseline value. 
+applicable to the HIV epidemic in sub-Saharan Africa.
 Although virulence trajectories are broadly similar
-  across model structures --- an initial increase in virulence/decrease
-  in progression time, followed by a minimum and an increase to an equilibrium 
-  progression time slightly higher than the minimum --- the timing and magnitude of the minimum vary
+across model structures, the timing and magnitude of the 
+minimum expected time to progression vary
 considerably.
 Models of intermediate
-  complexity predicted much lower peak virulence (15 years to
+complexity predicted slower progression (a minimum of 15 years to
 progress to AIDS) compared to both more realistic models
 and simple random-mixing models with no partnership structure
-  at all (both approx. 7.25 years to progress to AIDS).
-  In this range of models, the simplest random-mixing structure is
-  actually the best approximation to the most realistic model; this
+at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
+In this range of models, the simplest random-mixing structure best
+approximates the most realistic model; this
 surprising outcome occurs because the dominance of extra-pair
@@
 Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix for model details).
-We assume random activity-weighted mixing between risk groups \todo{cite old Anderson-May HIV paper?}.
+We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
 While this model lacks some

### 2016-10-19 ms/plosHIVcover_resubm.tex

\documentclass[10pt]{letter}

\usepackage{graphicx}
\usepackage{myltr2}
\usepackage{times}
%\usepackage[sort&compress]{natbib}

\bibliographystyle{abbrv}

\makeatletter
% http://tex.stackexchange.com/questions/18033/using-bibtex-with-letter-class
\newenvironment{thebibliography}[1]
     {\list{\@biblabel{\@arabic\c@enumiv}}%
           {\settowidth\labelwidth{\@biblabel{#1}}%
            \leftmargin\labelwidth
            \advance\leftmargin\labelsep
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\newcommand\newblock{\hskip .11em\@plus.33em\@minus.07em}
\makeatother

\newcommand{\revcomment}[1]{\emph{#1}}
\newcommand{\response}[1]{#1}

\begin{document}

\date{\today}

\signature{\includegraphics[height=0.35in]{bbsig3.png}\\Benjamin Bolker}

\begin{letter}{
}

\opening{To the editor,}

\revcomment{The abstract really needs to be improved. It’s missing the main points
of the paper (that different contact patterns change the predicted
trajectories and optimums of virulence), and has unnecessary minor
details (like the fact that parameters were sampled using a Latin
hypercube).}

\response{We've followed the reviewer's suggestions in adding
important points and removing unimportant details from the abstract}

\revcomment{The introduction is really lacking a cogent description of the study of
the evolution of virulence, and the motivation of the paper would be
totally unclear to someone not already very familiar with this field.
The main points that are needed to spell out this issue are:

[... detailed suggestions omitted]
}

\response{We thank the reviewer for pointing this out --- we've 
been over this ground so often that we've come to take this basic
framework for granted!  We've 
}
The paper claims to add considerations of population structure to the
study of virulence but only focuses on the rules of partnership
formation, while assuming all individuals have identical behavior
patterns, and not on another major aspect of population structure, which
is the heterogeneity in the number of partners (or rates of partnership
change) among individuals. In the case of a static network, this would
be considered the degree heterogeneity. Many studies have shown that
sexual contact networks tend to me more on the scale-free spectrum, and
so the assumption of uniformity of behavior seems weak. This study would
be most impactful if on top of examining these switch patterns it
examined the influence of contact heterogeneity.

In Figure 3 it would be helpful to also include time to progression to
AIDS as a figure panel

Reviewer #2: Modeling studies have shown that HIV could evolve towards
immediate virulence where the trade-off between host survival and
transmission is optimized. The aim of this study is to investigate how
different assumptions about sexual partnership dynamics affect the
transient evolution of HIV virulence in a population. While this is a
clear objective, the study is rather technical and, in my opinion, does
not provide major insights or conclusions. My major concerns regarding
the approach of the study is given below. Furthermore, there are several
issues in terms of structure and the model description that should be
considered by the authors.

Approach:
The authors argue that they compare more “realistic” models (based on
ref. 14) to the relatively simple model by Shireff et al. (ref. 10). In
my view, all these models are quite simplistic in that they do not take
into account age-structure and heterogeneity in sexual behavior.
Essentially, all these models are random mixing models where every
individual can make contact with every other individual. The only
difference to the random mixing model by Shireff et al. is that
transmission can occur in ongoing partnerships and that some models
allow for concurrency. Arguably much more important aspects that affect
HIV epidemics and virulence evolution are the above-mentioned age- and
risk-structure. Furthermore, the authors use a single-stage disease
model and do not distinguish between differences in transmissibility
between the acute (primary) and chronic (asymptomatic) stage of
infections, which might influence HIV virulence evolution more than
differences in the sexual partnership dynamics. Hence, I am not
convinced that the authors use a “more realistic” model than the one by
vShireff et al.

Structure:
- The references to the equations are not always correct.
- Figure 1 is described in Materials and Methods before the model is
fully explained. Consequently, the parameter r cannot be understood.
- Tables and Figures often lack dimensions, i.e., is is unclear what the
axes represent. Table 1 is more or less useless without providing
dimensions for the various parameters.
- The author summary seems to provide more information about the results
than the abstract. However, the authors use SPVL and progression to AIDS
as two different proxies for virulence which can be confusing.

Model:
The authors make use of a previously published model from ref. 14. This
model is based on the pair-formation formalism initially developed by
Dietz & Hadeler (1988, J Math Biol, PMID: 3351391) and later used by
Kretzschmar et al. (1994 & 1998, Math Biosci, PMID: 7833594 & 9597826).
I feel that the authors should refer to some of these original
publications as well. Unfortunately, the description of the models in
appendix S1 seems to be partially erroneous or incorrect. For example,
the primes for the derivatives are often missing (e.g., equation 22, 24
and 25) or it is unclear whether equations represent state variables or
derivates thereof. Also, the reasoning behind the “instswitch” models is
unclear to me. The authors say that “Once individuals leave a
partnership, they enter temporary compartments…”. The variables (or
derivatives?) X and Y_i in equation 19 consist of positive terms only,
so how do individuals leave these compartments? In my understanding the
standard pair model (equation 9) can easily account for instantaneous
partnership formation by simply setting the pair formation rate rho to a
high value.
 ------------------------------------------------------------------------

*Reviewers who wish to reveal their identities to the authors and other
reviewers should include their name here (optional). These names will
not be published with the manuscript, should it be accepted.*

Reviewer #1: (No Response)

Reviewer #2: (No Response)

\closing{Sincerely,}

\bibliography{virulence}

\end{letter}
\end{document}

### 2016-10-19 ms/plos_latex.tex
added 1 todo in ms and worked on other todo's

@@
 However, our simplified model produces results that are qualitatively similar to those of Shirreff \etal's \cite{shirreff_transmission_2011} model; when our model is calibrated to have a similar initial epidemic growth rate $r$, the peak \Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly higher (4.7 vs.
-4.3); the relative height of the peak is \todo{compute difference and fill this in}
-(\figurename~\ref{fig:panel3}).
+4.3); the relative height of the peak is 1.07 (\figurename~\ref{fig:panel3}).
 
@@
 
-\todo{add something about choice of heterogeneity}
+Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity.
+We decided to use squared CV instead of variance as variance measures is often unreliable \todo{CITE?
+You mentioned that variance is unreliable and we should use CV but we didn't talk about it very much...}.
+To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men as that was the group with the largest variability \cite{omori2015dynamics}.
+However, we decided to use 0.01 - 100 as the sampling range for squared CV despite the wider CV range given by Omori \etal \cite{omori2015dynamics} (0 - 20.324).
+Although married female have largest variability for CV, the range of the mean is narrow and small, which might explain for such large variability.
+Assessing ranges of CV of other groups, we concluded that 0.01 - 100 is both reasonable and wide enough to cover uncertainties in parameters.
+For simplicity, we assume that the number of non-cohabiting partner follows a gamma distribution.
 
@@
 Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (\figurename~\ref{fig:virtraj}).
-Although the speed of virulence evolution varies, leading to wide variation in the minimum progression time (means ranging from approximately \todo{fill in}), virulence peaks in all models between 200 and 300 years.
+Although the speed of virulence evolution varies, leading to wide variation in the minimum expected progression time (means ranging from approximately 6 to 12 years), virulence peaks in all models between 200 and 300 years.
 
@@
 random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model (Figs~???? \todo{fill in!} show univariate distributions
+for the implicit model (\figurename~\ref{fig:unidist} show univariate distributions
 on the epidemiological scales

### 2016-10-19 figures/figures.R

@@ -59,6 +59,19 @@ tlab <- "time (years)"
 g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run,linetype=run))+
     geom_line()+labs(x=tlab,y=expression(population~mean~set-point~viral~load~(log[10])))+
     theme(legend.position=c(0.75,0.25))
+
+## Compute difference
+
+df_tot %>%
+	subset(run %in% c("Shirreff", "r=0.042")) %>%
+	group_by(run) %>%
+	## summarise(max(Mean.VL)) ## Not sure why this only returns one value
+	summarise_at("Mean.VL", max) %>%
+	select(Mean.VL) %>%
+	unlist -> TODO.values
+
+TODO.values[2]/TODO.values[1]
+
 ###
 dimnames(vir_matS2) <- list(Time=tvec,
                run=c("10^{-3}", "10^{-4}", "10^{-5}"))
@@ -223,6 +236,13 @@ w <- with(mL,which(model=="random" & variable=="eq_dur"))
 rval <- mean(mL$value[w])
 mLw <- droplevels(mL[-w,])
 mLw$variable <- fixfac2(mLw$variable)
+
+## TODO: get mean minimum expected time for each model:
+
+mL %>%
+	group_by(model) %>%
+	filter(variable == "peak_dur") %>%
+	summarise_at("value", mean)
 
 gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
 	geom_violinh(width=1)+

### 2016-10-19 ms/plos_latex.tex
changing figure names - saving fig 4 as png crashes r studio - fig 5 still doesn't look very nice - still need to work on fig 4

@@
 
-Our chosen summary statistics (peak time, minimum progression time, equilibrium
+Our chosen summary statistics (peak time, minimum expected progression time, equilibrium expected
 progression time, and relative progression time) all vary considerably across models

### 2016-10-19 figures/figures.R

@@ -31,8 +31,8 @@ load("../simdata/combineResults.rda")
 
 ## setup2
 orig_sum_labs <- c("peak_time","peak_dur","eq_dur","rel_dur")
-new_sum_labs <- c("time at which progression time is minimal (years)","minimum mean progression time (years)",
-									"equilibrium mean progression time (years)","relative ratio: minimum/equilibrium progression time")
+new_sum_labs <- c("peak time (years)","minimum expected progression time (years)",
+									"equilibrium expected progression time (years)","relative minimum ratio")
 fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
     if (atop) new_sum_labs <-
                   gsub("(.*) (.*)","atop(\\1,\\2)",new_sum_labs)
@@ -250,7 +250,7 @@ gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
 	facet_wrap(~variable,scale="free_x")+
 	guides(fill=guide_legend(reverse=TRUE))+
 	labs(y="")+
-	geom_point(data=data.frame(model="random",variable="equilibrium mean progression time (years)",
+	geom_point(data=data.frame(model="random",variable="equilibrium expected progression time (years)",
                                value=rval),pch=22,size=3,show.legend=FALSE) +
 	zero_x_margin
 

### 2016-10-19 figures/figures.R
tried sampling for fig 4

@@ -364,6 +364,11 @@ tweak_colours_gg <- function(gg) {
 
 sL2 <- sL.mod[,c("model", "run", "eq_dur", "peak_time", "peak_dur")]
 
+set.seed(101)
+sL2 <- sL2 %>%
+	group_by(model) %>%
+	sample_n(100)
+
 new_sum_labs2 <- gsub("\\s\\(years\\)?", "" ,new_sum_labs[1:3])
 
 names(sL2)[na.omit(match(orig_sum_labs,names(sL2)))] <- gsub(" ","_",new_sum_labs2)
@@ -372,7 +377,7 @@ ggp1 <- ggpairs(sL2,
         mapping = ggplot2::aes(color = model,pch=model),
         columns=3:5,
         legends=TRUE,
-        lower = list(continuous = wrap("points",alpha=0.6,size=1)), ## alpha = 0.3,size=0.5)),
+        lower = list(continuous = wrap("points",alpha=0.6,size=2)), ## alpha = 0.3,size=0.5)),
         diag = list(continuous = "blankDiag"),
         upper = list(continuous = "blank"))
 ## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot

### 2016-10-21 R/combineResults.R
figure merging and tweaking

@@ -1,16 +1,16 @@
 simdatapath <- "../simdata"
 
-library("ggplot2"); theme_set(theme_classic())
+library(ggplot2); theme_set(theme_classic())
 scale_colour_discrete <- function(...,palette="Set1") scale_colour_brewer(...,palette=palette)
 scale_fill_discrete <- function(...,palette="Set1") scale_fill_brewer(...,palette=palette)
-library("gridExtra")
-library("plyr")  ## for ldply()
-library("dplyr") ## for full_join()
-library("reshape2")  ## for melt()
-library("GGally")
-library("knitr")
+library(gridExtra)
+library(plyr)  ## for ldply()
+library(dplyr) ## for full_join()
+library(reshape2)  ## for melt()
+library(GGally)
+library(knitr)
 ## devtools::install_github("lionel-/ggstance")
-library("ggstance")
+library(ggstance)
 
 source("hivFuns.R")
 source("Param.R")
@@ -101,19 +101,30 @@ sum_list[["I_mat"]] <-   fixfac(sum_list[["I_mat"]])
 sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
 mL <- melt(sL,id.vars=c("model","run"))
 
-m1 <- rename(melt(sum_list[["sum_mat"]],id.vars=c("model","run")),
+sumX <- transform(sum_list[["sum_mat"]],
+                  eq_dur=returnDur(eq_vir,HIVpars.skeleton),
+                  peak_dur=returnDur(peak_vir,HIVpars.skeleton),
+                  eq_beta=returnBeta(eq_vir,HIVpars.skeleton),
+                  peak_beta=returnBeta(peak_vir,HIVpars.skeleton))
+
+m1 <- rename(melt(sumX,id.vars=c("model","run")),
              sumvar=variable,sumval=value)
+
 m2 <- rename(melt(ltab,id.vars="run"),
              LHSvar=variable,LHSval=value)
 mL2 <- full_join(m1,m2,by="run")
 
 ## what are we doing here? why exclude those rows?
 ## need to add columns for het runs, for sensitivity plot ...
-ltab2 <- ltab[,c(2,4,5,6,7)]
-names(ltab2) <- c("Beta1_adj", "Beta3_adj", "rho_ad", "c_mean_adj")
+## was scaling
+ltab2 <- ltab[,c(2,4,6,7)]
[…]

### 2016-10-21 R/hivFuns.R

@@ -169,3 +169,18 @@ find_scale <- function(target,
 geom_mean <- function(a){
   exp(mean(log(a)))
 }
+
+## convert from SPVL to duration
+returnDur <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(v,Dmax,D50,h))
+	})
+}
+
+returnBeta <- function(lSpvl, parameters){
+	with(as.list(c(parameters)),{
+		v = 10^lSpvl
+		return(hill(K,Bmax,v,a))
+	})
+}

### 2016-10-21 figures/figures.R

@@ -33,19 +33,6 @@ source("../R/Param.R")
 load("../simdata/combineResults.rda")
 
 ## FIXME: maybe should go in hivFuns.R ?
-returnDur <- function(lSpvl, parameters){
-	with(as.list(c(parameters)),{
-		v = 10^lSpvl
-		return(hill(v,Dmax,D50,h))
-	})
-}
-
-returnBeta <- function(lSpvl, parameters){
-	with(as.list(c(parameters)),{
-		v = 10^lSpvl
-		return(hill(K,Bmax,v,a))
-	})
-}
 
 ## setup2
 orig_sum_labs <- c("peak_time","peak_dur","eq_dur","rel_dur")
@@ -153,6 +140,7 @@ ss <- transform(ss,
                 cmodel=droplevels(factor(gsub("+epc","",model,fixed=TRUE),
                                          levels=levels(model))))
 
+## base plot for Figure 2 and related (duration, SPVL, beta trajectories)
 gg_basetraj <- ggplot(ss,aes(tvec,mean,ymin=lwr,ymax=upr))+
     geom_line(aes(colour=model,linetype=model),lwd=1)+
     geom_ribbon(aes(fill = model), alpha=0.3)+
@@ -170,7 +158,7 @@ top.points2 <- gapply.fun(transform(d[which.max(d$y), ],
 
 bottom.points2 <- gapply.fun(transform(d[which.min(d$y), ],
                                ## *smaller* numbers move labels right/up
-                                    hjust = 0.1, vjust = 2))
+                                    hjust = 0.1, vjust = 1.5))
 
 ## http://stackoverflow.com/questions/10547487/r-removing-facet-wrap-labels-completely-in-ggplot2
 
@@ -198,11 +186,11 @@ fig_objects <- c(fig_objects,"gg_durtraj")
 
 ### Figure 2.1 - Transmission rate
 
-ss_trans <- transform(ss, mean = returnBeta(mean,HIVpars.skeleton),
+ss_beta <- transform(ss, mean = returnBeta(mean,HIVpars.skeleton),
                       lwr = returnBeta(lwr,HIVpars.skeleton),
                       upr = returnBeta(upr,HIVpars.skeleton))
 
-gg_betatraj <- direct.label(gg_basetraj %+% ss_trans + nostrips +
+gg_betatraj <- direct.label(gg_basetraj %+% ss_beta + nostrips +
[…]

### 2016-10-23 ms/plos_latex.tex
working on ms; figures for other talks

@@
 % Please keep the abstract below 300 words
-% current is 298!
+% current is 282
 \section*{Abstract}
-The evolutionary dynamics of parasite virulence over the
-course of an emerging epidemic have important implications for
-our understanding of epidemiological dynamics.
-While changes in the fitness landscape
+The evolutionary dynamics of parasite virulence change
+in important ways over the
+course of an emerging epidemic.
+Changes in the fitness landscape
 generally select for higher virulence early in an
-epidemic, quantitative outcomes
-may depend sensitively on biological details and the structure of
+epidemic; however, quantitative outcomes
+may depend sensitively on epidemiological details and the structure of
 mathematical models used to portray them.  Fraser \emph{et al.} 
@@
 Models of intermediate
-complexity predicted slower progression (a minimum of 15 years to
+complexity as used by Fraser \emph{et al.} 
+predicted lower slower progression/lower virulence (a minimum of 15 years to
 progress to AIDS) compared to both more realistic models
@@
 surprising outcome occurs because the dominance of extra-pair
-contact in the realistic model tends to mask the effects of
+contact in the realistic model tends to swamp the effects of
 partnership structure.
@@
 
-Pathogens such as HIV can evolve rapidly in response to changes
-in their environments; such changes include both increases in disease
-prevalence over the course of the epidemic and treatment
-interventions. While researchers have successfully used computational
-models to explore these evolutionary dynamics, these models often
-neglect details such as the formation and
-dissolution of sexual partnerships; other research has shown that
-these processes can strongly affect epidemic outcomes. We built and
-compared models that used different methods to
-model both partnership dynamics and sexual
-contact outside of partnerships. Models of intermediate
-complexity predicted much lower peak virulence (15 years to
-progress to AIDS) compared to both more realistic models
-and simple random-mixing models with no partnership structure
-at all (both approx. 7.25 years to progress to AIDS); 
-extra-pair contact tended to wash out the effects
-of epidemiological structure. The large differences
-in evolutionary dynamics among different epidemiological models
-suggests that researchers trying to predict the evolution of
-pathogens should proceed with caution.
+Pathogens such as HIV can evolve rapidly in response to changes in
+their environments; such changes include both increases in disease
+prevalence and disease virulence over the course of an epidemic, or
+decreases in both after treatment interventions. While researchers
+have successfully used computational models to explore these
+evolutionary dynamics, these models often neglect details such as the
+formation and dissolution of sexual partnerships; other research has
+shown that these processes can strongly affect epidemic outcomes. We
+built and compared models that used different methods to model both
+partnership dynamics and sexual contact outside of stable
+partnerships. Models of intermediate complexity predicted much lower
+virulence over the course of the epidemic (a minimum of 15 years to
+progress to AIDS) compared to both more realistic models and simple
+random-mixing models with no partnership structure at all (both
+approx. 7.25 years to progress to AIDS); sexual contact outside of
+stable partnerships tended to wash out the effects of epidemiological
+structure. The large differences in evolutionary dynamics among
+different epidemiological models suggests that researchers trying to
+predict the evolution of pathogens should proceed with caution.
 
@@
 disease strains that can reproduce more --- where reproduction
-is defined the amount of \emph{between-host} transmission, or
+is defined here as the amount of \emph{between-host} transmission, or
 the number of new hosts infected --- will increase in prevalence.
@@
 at which hosts recover or die from disease.
-The trade-off theory \cite{anderson_coevolution_1982,Ebert1999} --- postulates that
-the birth and death rate are both linked to the rate at which
-the pathogen reproduces within the host, and that the pathogen
+The \emph{trade-off theory} \cite{anderson_coevolution_1982,Ebert1999} --- postulates that
+the between-host birth and death rate are both linked to the rate at which
+the pathogen exploits host resources for within-host reproduction, 
+and that pathogen
 evolution will thus strike a balance between the
@@
 Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than the rate of progression to AIDS itself
-(we refer to SPVL as ``virulence'' hereafter).
-In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period.
+(hereafter ``virulence'' will refer either to the SPVL, or to 
+the rate of progression to AIDS; these two quantities are
+deterministically linked in the model).
+In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period; we show below that our results are not overly
+sensitive to this simplification.
 We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \etal's model) and taking the duration-weighted average of transmission rates of three stages.
@@
 where $V_{\alpha} = 10^\alpha$.
-The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix).
-Simplifying the model of HIV pathogenesis from three stages to a single stage could affect our conclusions about the evolution of virulence (e.g.
-Kretzschmar and Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics and multiple stages of infectivity have interactive effects on $\rzero$).
-However, our simplified model produces results that are qualitatively similar to those of Shirreff \etal's \cite{shirreff_transmission_2011} model; when our model is calibrated to have a similar initial epidemic growth rate $r$, the peak \Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly higher (4.7 vs.
-4.3); the relative height of the peak is 1.07 (\figurename~\ref{fig:panel3}).
+
+The uncoupled and extra-couple transmission rates (i.e., the rates of
+transmission among people outside of a stable partnership, or between
+people inside of a stable partnership and people other than their
+partner) are scaled by
+multiplying the within-couple transmission rate $\beta$ by the contact
+ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix). Simplifying the model
+of HIV pathogenesis from three stages to a single stage could affect
+our conclusions about the evolution of virulence (e.g. Kretzschmar and
+Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
+and multiple stages of infectivity have interactive effects on
+$\rzero$). However, our simplified model produces results that are
+qualitatively similar to those of Shirreff \etal's
+\cite{shirreff_transmission_2011} model; when our model is calibrated
+to have a similar initial epidemic growth rate $r$, the peak
+\Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly
+higher (4.6 \Lspvl\ vs. 4.3 \Lspvl, or 7\% higher: \figurename~\ref{fig:panel3}).
 
@@
 Time series of mean population \Lspvl.
-(a) Contrast with three-stage Shirreff model and effects of varying exponential growth rate, $r$.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying exponential growth rates, $r$.
 (b) Effects of varying initial infectious density $I(0)$.

### 2016-10-23 figures/figures.R

@@ -182,7 +182,7 @@ gg_durtraj <- direct.label(gg_basetraj %+% ss_dur + nostrips +
 ggsave(gg_durtraj,file="fig2.pdf",width=6,height=4)
 ggsave(gg_durtraj,file="fig2.png",width=6,height=4,dpi=400)
 
-fig_objects <- c(fig_objects,"gg_durtraj")
+fig_objects <- c(fig_objects,"gg_durtraj","bottom.points2","top.points2")
 
 ### Figure 2.1 - Transmission rate
 

### 2016-10-24 ms/plos_latex.tex
still working on ms ...

@@
 Pathogens can increase their net reproduction rate either
-by increasing their birth rate, the rate (per infected host) at which they
-infect new hosts, or by decreasing their death rate, the rate
+by increasing their transmission rate, 
+the rate (per infected host) at which they
+infect new hosts, or by decreasing their clearance or disease-induced
+mortality rate, the rate
 at which hosts recover or die from disease.
-The \emph{trade-off theory} \cite{anderson_coevolution_1982,Ebert1999} --- postulates that
-the between-host birth and death rate are both linked to the rate at which
+The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
+the transmission and disease-induced mortality rate are both linked to the rate at which
 the pathogen exploits host resources for within-host reproduction, 
@@
 Time series of mean population \Lspvl.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying exponential growth rates, $r$.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
 (b) Effects of varying initial infectious density $I(0)$.
 (c) Effects of varying initial mean virulence $\alpha(0)$.
-The orange curve in panel (a) shows that our simplified model can produce similar trajectories to those Shirreff \etal's model \cite{shirreff_transmission_2011}.
-In order to compare the models, we match the initial exponential growth rate (see main text ``Simulation runs").
-Furthermore, we experiment with wide ranges of initial values (panels b and c) to show that evolutionary dynamics are not overly sensitive to starting parameter values, which we hold constant in our simulations.}
+The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories.
+Panels b and c illustrate the sensitivity of virulence trajectories to initial conditions $I(0)$ and $\alpha(0)$, which we hold constant in our simulations.}
 \label{fig:panel3}
@@
 where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$.
-Transmission rate and disease induced mortality rates are discretized into a vector as well:
+Transmission rate and disease induced mortality rates are discretized as well:
 \begin{equation}
@@
 
-We developed seven multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
 Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
-non-instantaneous partnership formation, (2) zero vs.
-positive extra-partnership sexual contact and transmission and (3) homogeneous vs.
+non-instantaneous partnership formation; (2) zero vs.
+positive extra-partnership sexual contact and transmission; and (3) homogeneous vs.
 heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
@@
 
-The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and simpler, implicit models, assume instantaneous partnership formation (``instswitch'').
+The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch'').
 The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
@@
 
-The next two models represent extreme simplifications of sexual partnership dynamics.
+The fifth and sixth models represent extreme simplifications of sexual partnership dynamics.
 One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
-It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics.
-However, to reflect the effect of partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}.
+It is actually a random mixing model that explicitly tracks only the total number of susceptible and infected individuals.
+However, to reflect the effect of partnership structure, it uses an adjusted transmission rate derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}.
 The second model of this pair (``random'') is a simple random-mixing model.
@@
 Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero'').
-Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix for model details).
+Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix).
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
 While this model lacks some
-important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, we see it as a first step toward assessing the effects of epidemiological complexity.
-As even these models are pushing the limits of compartmental-based models, adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
-
-The base model (i.e.
-pairform+epc) for the first four models and the heterogeneous model is an extension of Champredon \etal's model \cite{champredon_hiv_2013}.
+important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
+As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises \todo{how many?} coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
+
+The pairform+epc and heterogeneous models use the basic epidemiological framework of Champredon \etal \cite{champredon_hiv_2013}.
 Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$.
@@
 Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
-Extra-couple and uncoupled transmission are modeled in the same way as Champredon \etal's model.
-All the details have been adapted to a multi-strain scenario.
-The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact).
-Model details are explained in the appendix.
+Extra-couple and uncoupled transmission are modeled in the same way as in Champredon \etal's model.
+All the details have been adapted to a multi-strain scenario, so that we track (for example) a matrix
+$II_{ij}$ that records the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$. 
+The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix).
 
@@
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-For the heterogeneity model, mean and squared coefficient of variation (CV) for number of non-cohabiting partners are sampled as well.
-We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
+For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
+We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
 
@@
 \caption{Parameter ranges/values.
-Values of $c$ and $\rho$ are doubled from those given by Champredon \etal; we keep track of individuals in the model, while they keep track of couples.
+Values of $c$ and $\rho$ are doubled from those given by Champredon \etal\ because we keep track of individuals in the model, while they keep track of couples.
 Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
@@
 
-The one completely new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix S1); we need to pick a reasonable range for it.
+One parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix S1);
+it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  so we need to pick a reasonable range for it.
 Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
@@
 Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity.
-We decided to use squared CV instead of variance as variance measures is often unreliable \todo{CITE?
-You mentioned that variance is unreliable and we should use CV but we didn't talk about it very much...}.
-To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men as that was the group with the largest variability \cite{omori2015dynamics}.
-However, we decided to use 0.01 - 100 as the sampling range for squared CV despite the wider CV range given by Omori \etal \cite{omori2015dynamics} (0 - 20.324).
-Although married female have largest variability for CV, the range of the mean is narrow and small, which might explain for such large variability.
-Assessing ranges of CV of other groups, we concluded that 0.01 - 100 is both reasonable and wide enough to cover uncertainties in parameters.
-For simplicity, we assume that the number of non-cohabiting partner follows a gamma distribution.
+%%  BMB: I don't think we need to go into detail about variance vs CV
+To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men, as that was the group with the largest variability \cite{omori2015dynamics}.
+Omori \etal \cite{omori2015dynamics} give a very wide
+range for the coefficient of variation ($\approx$ 0 - 20, corresponding
+to squared CV range of 0-400):
+we narrowed this range for $\textrm{CV}^2$ to 0.01-100.
+At the bottom end of the range, estimating that a group behaves
+perfectly homogeneously ($CV=0$) is likely to be a sampling artifact;
+at the upper end, the estimate is also likely to be noisy because
+of the low mean value among married females (who have the largest
+range of CV). 
+We assume that the number of non-cohabiting partners follows a Gamma distribution.
 
@@
 disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
-every run has the same initial exponential growth rate of incidence.
+every run has the same initial exponential growth rate in 
+disease incidence.
 
@@
 After each run, initial exponential growth rate is calculated.
-Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, which is approximately equal to that implied by Shirreff \etal's original models.
-
-The calibration runs for each parameter set are slightly
-simplified. We run each model for only 500 years
-(full simulations are run for 4000 years). We use a 4/5 order 
+Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
+For calibration purposes, we run each model for only 500 years
+(full simulations are run for 4000 years), which is always long
+enough to capture the exponential growth phase of the model. 
+We use a 4/5 order 
 Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
@@
 reflects reality, we would take the peak time of 150-300 years 
-(Figure~\ref{fig:panel3}c) to mean that, in the absence of
+(\figurename~\ref{fig:panel3}c) to mean that, in the absence of
 treatment, the epidemic would probably still be increasing in virulence.
@@
 Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
-Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).
-
-Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (\figurename~\ref{fig:virtraj}).
+Decreasing the initial virulence
+also leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).
+
+Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of the expected time of progression
+to AIDS (\figurename~\ref{fig:virtraj}: lower progression time corresponds
+to higher virulence).
 Although the speed of virulence evolution varies, leading to wide variation in the minimum expected progression time (means ranging from approximately 6 to 12 years), virulence peaks in all models between 200 and 300 years.
@@
 progression time, and relative progression time) all vary considerably across models
-\ref{fig:unidist}.
+(\figurename~\ref{fig:unidist}).
 We first consider the models of intermediate realism: implicit,
@@
 these same sets lead to correspondingly low
-peak virulence (16 years to progression) and early peak times (before 200 years).
+peak virulence (16 years to progression) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
 At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
@@
 \caption{{\bf Univariate distributions of summary statistics.}
-The distribution of equilibrium virulence for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
+The distribution of equilibrium expected progression
+time (lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
 \label{fig:unidist}
@@
 These differences are practically as well as 
-scientifically important. The random-mixing and pairform+epc models
-predict minimum times to progression (at the virulence peak) of
-5.7 (95\% CI 6.1-6.3) and 6.0 (5.0-7.7) years respectively, while
-the implicit model gives progression times about twice as long:
-12.5 (9.5-15.6) years. The corresponding differences in 
+scientifically important. The random-mixing, pairform+epc,
+and heterogeneous models all predict rapid
+progression to AIDS at the virulence peak
+(median/95\% CI = 6.1 (5.7-6.3), 
+6.02 (5.04-7.7), 6.03  (4.8-9.2)). 
+In contrast, 
+the implicit model predicts minimum progression times about twice as long:
+12.5 (9.6-15.6) years. The corresponding differences in 
 within-couple transmission
 probability are even more extreme, about a fourfold difference:
-0.249 (0.24-0.26) and 0.252 (0.19-0.28) per year for the 
+0.249 (0.24-0.26), 0.252 (0.19-0.28), and 0.252 (0.15-0.28) per year for the 
 random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model (\figurename~\ref{fig:unidist} show univariate distributions
-on the epidemiological scales
-of progression time and transmission probability,
-for all summary measures and all models).
+for the implicit model (see Appendix for 
+plots showing univariate summaries
+of \Lspvl\ and transmission probability).
 
@@
 \caption{{\bf Sensitivity plot.}
-For each parameter in the Latin hypercube sample and each summary statistic, shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
-prior to adjusting the parameters to achieve the standard initial epidemic growth rate.}
+For most parameters in the Latin hypercube sample and each summary statistic, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
+prior to adjusting the parameters to achieve the standard initial epidemic growth rate.
+\todo{mention subsampling}}
 \label{fig:plot_sens}
@@
 different models with similar univariate dynamical summaries. While the
-relationship between equilibrium virulence and peak time is
+relationship between equilibrium progression time and peak time is
 similar for all model structures (top left panel), the other
-relationships are more separate. In particular, the implicit
+relationships show more variation. In particular, the implicit
 and pair-formation (without extra-pair contact) are very similar
@@
 models have high, relatively
-constant values with respect to these parameters.
+constant values with respect to these parameters
+(because the
 
@@
 
-\subsection*{Heteregenous model}
-
-We extend "pairform+epc model" by allowing for heterogeneity in sexual behaviour.
+\subsection*{Heterogenous model}
+
+We extend the ``pairform+epc'' model by allowing for heterogeneity in sexual behaviour.
 Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model the mentioned behaviours are scaled by the same factor based on the risk group.
@@
 Lastly, for heterogeneity model, we assume that the risk distribution of the population follows gamma distribution and calculate the shape and scale parameters given the mean and squared coefficient of variation.
-Using the shape and scale parameterws, we define gamma quantile function $Q(p)$ and $p_j =  \tsub{p}{min} + (\tsub{p}{max} - \tsub{p}{min}) \frac{j-1}{n_r + 1}$, where $n_r$ is number of risk groups and $j = 1, 2, 3, \dots, n_r + 1$.
+Using the shape and scale parameters, we define gamma quantile function $Q(p)$ and $p_j =  \tsub{p}{min} + (\tsub{p}{max} - \tsub{p}{min}) \frac{j-1}{n_r + 1}$, where $n_r$ is number of risk groups and $j = 1, 2, 3, \dots, n_r + 1$.
 Since $Q(1) = \infty$, we set $\tsub{p}{max} = 0.99$ and $\tsub{p}{min} = 0.01$.

### 2016-10-24 figures/figures.R

@@ -230,10 +230,17 @@ rval <- mean(mL$value[w])
 mLw <- droplevels(mL[-w,])
 mLw$variable <- fixfac2(mLw$variable)
 
-(minprogtime <- mL %>%
+minprogtime <- mL %>%
 	group_by(model) %>%
 	filter(variable == "peak_dur") %>%
-	summarise(value=mean(value)))
+	summarise(med=median(value),
+                  lwr=quantile(value,0.025),
+                  upr=quantile(value,0.975))
+
+options(digits=3)
+minprogtime
+
+
 
 gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
 	geom_violinh(width=1)+
@@ -275,6 +282,12 @@ sL.epi <- transform(sL, eq_t = returnBeta(eq_vir,HIVpars.skeleton),
                     peak_dur = returnDur(peak_vir,HIVpars.skeleton))
 
 sL.epi <- sL.epi[,-c(3:6)]
+
+maxtrans <- sL.epi %>%
+	group_by(model) %>%
+	summarise(med=median(peak_t),
+                  lwr=quantile(peak_t,0.025),
+                  upr=quantile(peak_t,0.975))
 
 mL.epi <- melt(sL.epi,id.vars=c("model","run"))
 ## horrible hack (but doesn't help); subsample

### 2016-10-24 ms/plos_latex.tex
a little more

@@
 
-Finally, the sensitivity plots (\figurename~\ref{fig:plot_sens}) show the effects 
+Finally, the sensitivity plot (\figurename~\ref{fig:plot_sens}) shows the effects 
 of each parameter on the summary statistics. In almost every case the
@@
 constant values with respect to these parameters
-(because the
+(because the patterns are so similar across this set of parameters,
+\figurename~\ref{fig:plot_sens} shows only $D_P$).
 
@@
 model make extra-pair contact by coupled individuals less important).
-Surprisingly, neither of the pair-formation models is particularly sensitive to the 
-rate of partnership formation ($\rho$); the rate of uncoupled contact
-increases virulence and peak time in the pair-formation+epc model, 
-which is the only model to which it applies.
+Surprisingly, once calibration
+is taken into account, the remaining parameters have little effect overall.
+The rate of partnership formation ($\rho$) 
+has little impact on the models with finite pair-formation times.
+The relative rate of uncoupled contact ($c_u/c_w$) slightly decreases the
+minimum and equilibrium progression time and delays the peak time in the
+pair-formation+epc model, but neither the uncoupled contact rate nor
+the mean ($\mu$) or CV$^2$ of the number of non-cohabiting sexual
+partners has much systematic effect in the heterogeneous model.
 
@@
 In the particular example of HIV virulence eco-evolutionary dynamics
-and epidemiological, we reach the slightly ironic conclusion that the
+and the complexity of epidemiological structures
+we reach the slightly ironic conclusion that the
 effort put into building a more realistic model essentially cancels
-out, leaving us in the same position as if we had ignored the problem
-of epidemiological complexity entirely and used a naive random-mixing
+out, putting us back where we started when used a naive random-mixing
 contact model. In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps

### 2016-10-24 ms/plos_latex.tex
24530 differential equations!

@@
 important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
-As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises \todo{how many?} coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
+As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises $24530$ coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
 

### 2016-10-24 ms/plos_latex.tex
still working on ms

--- 2016-10-24 ms/plos_latex.tex
@@
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
-Dashed line in equilibrium vs. peak virulence plot shows 1:1 line.}
+Dashed line in equilibrium vs. peak virulence plot shows 1:1 line.
+ \todo{mention subsampling}}
 \label{fig:pairplot}
@@
 For most parameters in the Latin hypercube sample and each summary statistic, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
-prior to adjusting the parameters to achieve the standard initial epidemic growth rate.
-\todo{mention subsampling}}
+prior to adjusting the parameters to achieve the standard initial epidemic growth rate.}
 \label{fig:plot_sens}
@@
 
-All models must simplify the world.  Many constraints --- such as data
-availability, computation time, or code complexity --- drive the need
+All models must simplify the world.  Many constraints --- among them data
+availability, computation time, and code complexity --- drive the need
 for parsimony, with different constraints applying in different
@@
 out, putting us back where we started when used a naive random-mixing
-contact model. In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
-actually more true to real-world HIV epidemiological dynamics than a
-model with realistic partnership durations that neglects extra-pair
-contact \cite{herbeck2016evolution}. Making the model even more realistic
-might make the random-mixing model less appropriate. For
+contact model.
+However, we are not quite back where we started, as the
+complex models lead to wider, presumably more realistic
+confidence intervals on the predictions.
+In general, unstructured mixing --- whether occurring through 
+purely random mixing, or through extra-pair contact and contact
+among people outside of stable partnerships --- tends to drive
+faster virulence evolution, leading to higher peak virulence and 
+lower times to progression at the peak time.
+
+%In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
+%actually more true to real-world HIV epidemiological dynamics than a
+%model with realistic partnership durations that neglects extra-pair
+%contact \cite{herbeck2016evolution}. 
+Taking further steps to make the model even more realistic
+might add further structure,
+making the random-mixing model predictions less accurate. For
 example, our model forms partnerships randomly, and assumes that
@@
 Single individuals acquire partners at per-person rate $\rho$.
-Partnership formation rate for $S$, $I_i$ are $\rho S$ and $\rho I_i$, respectively.
-We follow Champredon \etal's results \cite{champredon_hiv_2013} and assume that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
+Partnership formation rates for $S$ and $I_i$ are $\rho S$ and $\rho I_i$, respectively.
+We follow Champredon \etal \cite{champredon_hiv_2013} in assuming that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
 
@@
 Partnerships dissolve at per-partnership rate $c$: the dissolution rates for $SS$, $SI_i$, and $II_{ij}$ pairs are $c SS$, $c SI_i$, and $c II_{ij}$ respectively.
-Unlike single strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains which the individuals in concordant partnership are infected with (i.e.
+Unlike a single-strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains with which the individuals in concordant partnership are infected (i.e.
 both partners in $II_{ii}$ enter $I_i$ whereas one partner in $II_{ij}$ enters the $I_i$ compartment while the other enters $I_j$).
-Thus, coupled individuals are distributed into single states through partnership dissolution (DS):
+Thus, coupled individuals are distributed into single states through partnership dissolution (DS) rates:
 
@@
 An infected partner in $SI$ partnership transmits virus to a susceptible partner at per-partnership rate $\beta$: $\WT(SI_i) = - \beta_i SI_i$.
-Since we assume that mutation occurs, $II_{ij}$, where $i \neq j$, can be formed from both $SI_i$ and $SI_j$ partnership: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$.
+Since we assume that mutation occurs, $II_{ij}$ pairs, where $i \neq j$, can be formed from either $SI_i$ or $SI_j$ partnerships: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$.
 On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $\WT(II_{ii}) = M_{ii} \beta_i SI_i$.
-Using the Kronecker delta notation, we obtain the following set of equations that describes within-couple transmission dynamics:
+Using the Kronecker delta notation, we obtain the following set of equations for within-couple transmission dynamics:
 
@@
 \end{equation}
-The effective uncoupled, $c_u$, and extra couple, $c_e$, contact rates are the product of two terms: uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
+The effective uncoupled, $c_u$, and extra couple, $c_e$, contact rates are the product of uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
 Therefore, the transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above.

### 2016-10-24 ms/plos_latex.tex
added one more to do

--- 2016-10-24 ms/plos_latex.tex
@@
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
-Dashed line in equilibrium vs. peak virulence plot shows 1:1 line.}
+Dashed line in equilibrium vs. peak virulence plot shows 1:1 line. 100 values were sampled from each model to allow for clearer distinction between different models}
 \label{fig:pairplot}
@@
 For most parameters in the Latin hypercube sample and each summary statistic, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
-prior to adjusting the parameters to achieve the standard initial epidemic growth rate.
-\todo{mention subsampling}}
+prior to adjusting the parameters to achieve the standard initial epidemic growth rate}
 \label{fig:plot_sens}
@@
 
-This section presents
+This section presents \todo{You forgot to finish writing this section last time. Maybe we can actually get rid of these plots unless you want to change progression figures to SPVL figures}.
 

### 2016-10-24 ms/plos_latex.tex
a few more tweaks

--- 2016-10-24 ms/plos_latex.tex
@@
 important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
-As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises $24530$ coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
+As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
 
@@
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
-Dashed line in equilibrium vs. peak virulence plot shows 1:1 line.
- \todo{mention subsampling}}
+Dashed line in equilibrium vs. peak virulence plot shows 1:1 line. 100 values were sampled from each model to allow for clearer distinction between the models}
 \label{fig:pairplot}
@@
 where $r_u = c_u/c_w$ and $r_e = c_e/c_w$ are the relative uncoupled/extra-couple contact rates.
-This simplification is useful in a multi-strain model since we cannot multiply a vector with a single value (e.g.
+This simplification is useful in a multi-strain model since we cannot multiply a vector by a scalar value (e.g.
 $c_u S$ in denominator) if we use Champredon \etal's equation in its original form.
-Extending the above equation to multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain:
+Extending the above equation to the multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain:
 
@@
 
-Single susceptible individuals become infected through uncoupled contact at per-person rate $\sum_k P_k U_k$ and enter single infected state.
+Single susceptible individuals become infected through uncoupled contact at per-person rate $\sum_k P_k U_k$ and enter the single infected state.
 Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $\ET(I_i) = \sum_k M_{ki} P_k U_k S$.
-Either partner in an $SS$ partnership becomes infected at per-person rate $\sum_k P_k E_k$, and partnership state changes to an $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$.
+Either partner in an $SS$ partnership can become infected at per-person rate $\sum_k P_k E_k$, and partnership state changes to an $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$.
 The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $\ET(SI_i) = \sum_k 2 M_{ki} P_k E_k SS$.
-Lastly, the susceptible partner in an $SI$ partnership becomes infected from extra-couple contacts at a per-person rate of $\sum_k P_k E_k$, and partnership state changes to an $II$ partnership.
+Lastly, the susceptible partner in an $SI$ partnership can become infected from extra-couple contacts at a per-person rate of $\sum_k P_k E_k$, so that the partnership changes to an $II$ partnership.
 As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$.
@@
 the term $P_i$): $\ET(II_{ij}) = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$.
-Combining these equations we get the following set of equations that describe all the transmission dynamics:
+Combining these equations we get the following set of equations that describe the complete transmission dynamics:
 
@@
 
-Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single (uncoupled) state, only extra-couple transmission exists:
+Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single (uncoupled) state, only extra-couple transmission occurs:
 
@@
 \end{equation}
-Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equation that is used in model 1 and 2.
+Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equation that is used in models 1 and 2.
 
@@
 
-Disease induced mortality is modeled similar to model 1 and 2.
-However, as single state does not exist in model 3 and 4, individuals that has left their partnerships due to death of their partners enter temporary compartments and form partners instantly:
+Disease-induced mortality is modeled similarly to model 1 and 2.
+However, as the single state does not exist in model 3 and 4, individuals that leave partnerships due to death of their partners enter temporary compartments and form partnerships instantly:
 
@@
 
-Following \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta^\ast$, that is derived from \cite{hollingsworth_hiv1_2008}'s approximate basic reproduction number:
+Following \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta^\ast$, that is derived from Hollingsworth \etal's approximate basic reproduction number \cite{hollingsworth_hiv1_2008}:
 
@@
 
-Model 6 is a random mixing model. It is modeled in a same way as model 5 without the adjusted transmission rate:
+Model 6 is a random-mixing model. It is modeled in a same way as model 5 without the adjusted transmission rate:
 
@@
 We extend the ``pairform+epc'' model by allowing for heterogeneity in sexual behaviour.
-Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model the mentioned behaviours are scaled by the same factor based on the risk group.
-In other words, an individual in a higher risk is more likely to form a stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing.
+Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group.
+In other words, an individual in a higher risk is also more likely to form a stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing.
 We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group.
-For simplicity, we assume that the transmission rate per partnership is not affected by sexual behaviour.
+For simplicity, we assume that the transmission rate per partnership is unaffected by sexual behaviour.
 
@@
 For example, $SI_{\cdot j,kl}$ is the number of partners where the susceptible partner is in risk group $k$ and infected partner is in risk group $l$ and has \Lspvl\ of $\alpha_j$.
-For simplicity, we assume that people mix randomly.
-Then, we can write the partnership formation process as follows:
+For simplicity, we assume that people undergo random \emph{activity-weighted} mixing \cite{may_transmission_1988}.
+Then we can write the partnership formation process as follows:
 
@@
 
-For dissolution process, individual in risk group $i$ expects to leave partnership at a rate $\varphi_i c$.
-If partnership is formed between two individuals from a different risk group, rate at which they leave the partnership differs.
+For dissolution process, an individual in risk group $i$ leaves its partnership at a rate $\varphi_i c$.
+If a partnership is formed between two individuals from a different risk group, the rates at which they leave the partnership differ.
 We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners.
-Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $\frac{\varphi_k + \varphi_l}{2} c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate.
+Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate.
 
@@
 
-Since we assume that the rate of transmission per partnership stays constant across different risk groups, within-couple infection process is similar to other models:
+Since we assume that the rate of transmission per partnership stays constant across different risk groups, the within-couple infection process is similar to other models:
 
@@
 
-Heterogeneous extra-couple and uncoupled contact process is similar to partnership formation process.
+The heterogeneous extra-couple and uncoupled contact process is similar to the partnership formation process (activity-weighted random mixing).
 Relative uncoupled/extra-couple contact rates are scaled by the factor of $\varphi_i$, where $i$ is the risk group.
@@
 \end{equation}
-Since the relative uncoupled/extra couple contact ratio are scaled by the factor of $\varphi_i$, uncoupled and extra-couple transmission rates are scaled by the same factor as well: $U_{k,i} = \varphi_i r_u \beta_k$ and $E_{k,i} = \varphi_i r_e \beta_k$.
+Since the relative uncoupled/extra couple contact ratios are scaled by the factor of $\varphi_i$, uncoupled and extra-couple transmission rates are scaled by the same factor as well: $U_{k,i} = \varphi_i r_u \beta_k$ and $E_{k,i} = \varphi_i r_e \beta_k$.
 Once again, we assume random mixing between individuals.
 Then, a susceptible individual in risk group $i$ becomes infected through extra-couple and uncoupled contact at a per capita rate of $\sum_j \sum_k P_{k,j} X_{k,i}$.
-Once infected, individuals are distributed into each Strain categories through mutation.
+Once infected, individuals are distributed into strain categories through mutation.
 
@@
 We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with \Lspvl\ of $i$.
-$Y_i$ is taken from normal distribution with mean 3 and is normalized so that $\sum_i D_i = 1$.
+$Y_i$ is taken from the Normal distribution with mean 3 and is normalized so that $\sum_i D_i = 1$.
 Then, we have the following initial distribution of each states:
@@
 \end{equation}
-Since model 3 and 4 do not have single state, $SS^*=1$ at the disease free equilibrium and the initial distribution becomes as follows:
+Since model 3 and 4 do not have single states, $SS^*=1$ at the disease free equilibrium and the initial distribution becomes:
 
@@
 
-Lastly, for heterogeneity model, we assume that the risk distribution of the population follows gamma distribution and calculate the shape and scale parameters given the mean and squared coefficient of variation.
-Using the shape and scale parameters, we define gamma quantile function $Q(p)$ and $p_j =  \tsub{p}{min} + (\tsub{p}{max} - \tsub{p}{min}) \frac{j-1}{n_r + 1}$, where $n_r$ is number of risk groups and $j = 1, 2, 3, \dots, n_r + 1$.
+Lastly, for the heterogeneity model, we assume that the risk distribution of the population follows a Gamma distribution and calculate the shape and scale parameters from the mean and squared coefficient of variation.
+Using the shape and scale parameters, we define the Gamma quantile function $Q(p)$ and $p_j =  \tsub{p}{min} + (\tsub{p}{max} - \tsub{p}{min}) \frac{j-1}{n_r + 1}$, where $n_r$ is number of risk groups and $j = 1, 2, 3, \dots, n_r + 1$.
 Since $Q(1) = \infty$, we set $\tsub{p}{max} = 0.99$ and $\tsub{p}{min} = 0.01$.
 Then, we define $\varphi_i = \frac{Q(p_j) + Q(p_{j+1})}{2}$.
-We define $R_i$ as the proportion of individuals in risk group $i$ in a disease free equilibrium and assume $R_i$ is equal for all $i$, i.e.
+We define $R_i$ as the proportion of individuals in risk group $i$ at the disease-free equilibrium and assume $R_i$ is equal for all $i$, i.e.
 $R_i = \frac{1}{n_r}$.
@@
 \end{equation}
-For this particular simulation, we disregard infection process as well as disease induced mortality in order to preserve the strain distribution of infected individuals.
+For this particular simulation, we disregard the infection process as well as disease-induced mortality in order to preserve the strain distribution of infected individuals.
 Furthermore, since the scaling parameter, $\gamma$, does not affect the risk group distribution in the absence of disease transmission, we increase the scaling parameter to 5 ($\gamma = 5$) to speed up the simulation and run the model for 50 years.
@@
 
-This section presents
+This section presents alternate versions of
+the figures from the main text showing time dynamics and summary
+statistics in terms of \Lspvl\ and per-year transmission probability
+rather than expected progression time to AIDS.
 
@@
 
+\clearpage
+
 \nolinenumbers

### 2016-10-24 ms/plos_latex.tex
small tweak (both comments and changes are inside \todo)

@@
 the rate of progression to AIDS; these two quantities are
-deterministically linked in the model).
+deterministically linked in the model) \todo{Doesn't this somewhat contradict with what we say later on?
+Here, we say we focus on SPVL but later we say we summarize the models using progression time rather than SPVL...}.
 In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period; we show below that our results are not overly
@@
 Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.
+\todo{Although these models can also be achieved by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.}
 

### 2016-10-25 ms/plos_latex.tex
small tweak

@@
 Models of intermediate
-complexity as used by Fraser \emph{et al.} 
+complexity as used by Fraser \emph{et al.} \todo{why not shirreff et al?} 
 predicted lower slower progression/lower virulence (a minimum of 15 years to
@@
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
+\todo{does this sentence regarding heterogeneity feel redundant?} For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
 We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
@@
 Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
-\cite{soetaert_solving_2010}) for all simulations.
+\cite{soetaert_solving_2010}) for all simulations. \todo{mention that we got rid of error runs for hetero model?}
 

### 2016-10-25 ms/plos_latex.tex
couple more tweaks to ms

@@
 Once individuals leave a partnership, they are instantaneously distributed into coupled states.
-In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the number of susceptible individuals that leave the partnership at a given time, and $Y_i$ the number of infected individuals with \Lspvl\ of $\alpha_i$ who leave partnership at a given time.
+In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the number of susceptible individuals that leave partnership at a given time, and $Y_i$ the number of infected individuals with \Lspvl\ of $\alpha_i$ who leave partnership at a given time.
 These temporarily single individuals then form couples through the same partnership formation rule described in the previous section:
@@
 \end{equation}
-Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equation that is used in models 1 and 2.
+Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equations that are used in models 1 and 2.
 
@@
 Disease-induced mortality is modeled similarly to model 1 and 2.
-However, as the single state does not exist in model 3 and 4, individuals that leave partnerships due to death of their partners enter temporary compartments and form partnerships instantly:
+However, as the single state does not exist in model 3 and 4, individuals that leave partnerships due to death of their partners form partnerships instantly:
 
@@
 
-Following \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta^\ast$, that is derived from Hollingsworth \etal's approximate basic reproduction number \cite{hollingsworth_hiv1_2008}:
+Following Shirreff \etal \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta^\ast$, that is derived from Hollingsworth \etal's approximate basic reproduction number \cite{hollingsworth_hiv1_2008}:
 
@@
 \begin{aligned}
-S' & = \sum_k \lambda_k I_k - \sum_k \beta_k S I_k,\\
-I_i' & = \sum_k M_{ki} \beta_k S I_k - \lambda_i I_i.
+S' & = \sum_k \lambda_k I_k - \sum_k c \beta_k S I_k,\\
+I_i' & = \sum_k M_{ki} c \beta_k S I_k - \lambda_i I_i.
 \end{aligned}
@@
 Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group.
-In other words, an individual in a higher risk is also more likely to form a stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing.
+In other words, an individual in a higher risk is also more likely to form a \todo{long-term sexual partnership instead of stable?} stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing.
 We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group.
@@
 Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate.
+\todo{Herbeck 2016 does this.
+We can cite them?}
 

### 2016-10-25 ms/plos_latex.tex
small tweaks

@@
 Since the relative uncoupled/extra couple contact ratios are scaled by the factor of $\varphi_i$, uncoupled and extra-couple transmission rates are scaled by the same factor as well: $U_{k,i} = \varphi_i r_u \beta_k$ and $E_{k,i} = \varphi_i r_e \beta_k$.
-Once again, we assume random mixing between individuals.
+Once again, we assume activity-weighted mixing between individuals.
 Then, a susceptible individual in risk group $i$ becomes infected through extra-couple and uncoupled contact at a per capita rate of $\sum_j \sum_k P_{k,j} X_{k,i}$.
@@
 We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with \Lspvl\ of $i$.
-$Y_i$ is taken from the Normal distribution with mean 3 and is normalized so that $\sum_i D_i = 1$.
+$D_i$ is taken from the Gaussian distribution with mean 3 and standard deviation 0.2 and is normalized so that $\sum_i D_i = 1$.
 Then, we have the following initial distribution of each states:

### 2016-10-25 ms/plosHIVcover_resubm.tex
cover letter; ms tweaks; fig tweaks

@@
 
+We appreciate the opportunity to revise our manuscript.
+In addition to extensive clarification and correction following
+the reviewers' comments, we implemented a more complex model
+that includes variation in contact rates among individuals.
+
+\hrule
+
+\textbf{Reviewer \#1}
 \revcomment{The abstract really needs to be improved. It’s missing the main points
@@
 \response{We've followed the reviewer's suggestions in adding
-important points and removing unimportant details from the abstract}
+important points and removing unimportant details from the abstract.}
 
@@
 The main points that are needed to spell out this issue are:
-
-[... detailed suggestions omitted]
+[... detailed suggestions omitted ...]
 }
@@
 been over this ground so often that we've come to take this basic
-framework for granted!  We've 
-}
+framework for granted!  We've  extensively revised the introduction
+along the lines suggested by the reviewer.}
+
+\revcomment{
 The paper claims to add considerations of population structure to the
@@
 be most impactful if on top of examining these switch patterns it
-examined the influence of contact heterogeneity.
+examined the influence of contact heterogeneity.}
 
-In Figure 3 it would be helpful to also include time to progression to
-AIDS as a figure panel
+\response{We added a new model to our study that allows for 
+discrete classes of heterogeneity 
+in contact rate, with activity-weighted random mixing; this
+is a standard approach for implementing contact heterogeneity
+in a compartmental model. We discuss other avenues for
+increasing epidemiological realism (e.g., an explicitly
+heterosexual model with sex-specific transmission and
+progression characteristics, as well as age-structured
+mixing patterns) and comment that they will require a
+switch from our current compartmental approach to an agent-based
+framework.}
 
-Reviewer #2: Modeling studies have shown that HIV could evolve towards
+\revcomment{In Figure 3 it would be helpful to also include time to progression to
+AIDS as a figure panel}
+
+\response{Our figures 2-5 are now framed in terms of expected
+progression time rather than SPVL, as we thought this scale
+would generally be more interpretable (as before, the supplementary
+material includes summary figures based on the other metrics, in this
+case SPVL and transmission probability).}
+
+\textbf{Reviewer \#2}
+\revcomment{Modeling studies have shown that HIV could evolve towards
 immediate virulence where the trade-off between host survival and
@@
 clear objective, the study is rather technical and, in my opinion, does
-not provide major insights or conclusions. My major concerns regarding
-the approach of the study is given below. Furthermore, there are several
-issues in terms of structure and the model description that should be
-considered by the authors.
+not provide major insights or conclusions.
+}
 
-Approach:
+\response{Importance is in the eye of the beholder, of course; we
+have tried to clarify the take-home points of our study
+(epidemiological structures modify quantitative conclusions about
+virulence evolution; realistic models that account for both
+pair formation and extra-pair contact are best approximated by
+random-mixing models and worst approximated by models with
+intermediate complexity; extra-pair and other forms of unstructured
+contact generally act to increase the rate of virulence evolution).
+}
+
+\revcomment{
 The authors argue that they compare more “realistic” models (based on
@@
 transmission can occur in ongoing partnerships and that some models
-allow for concurrency. Arguably much more important aspects that affect
+allow for concurrency. 
+Arguably much more important aspects that affect
 HIV epidemics and virulence evolution are the above-mentioned age- and
-risk-structure. Furthermore, the authors use a single-stage disease
+risk-structure.}
+
+\response{As stated above in the response to reviewer \#1, we have
+added a model with contact heterogeneity to our study; while
+important, age-structured mixing is too complex for our current
+modeling framework and will have to be left for future studies.}
+
+\revcomment{
+ Furthermore, the authors use a single-stage disease
 model and do not distinguish between differences in transmissibility
@@
 convinced that the authors use a “more realistic” model than the one by
-vShireff et al.
+Shireff et al.}
 
-Structure:
-- The references to the equations are not always correct.
-- Figure 1 is described in Materials and Methods before the model is
-fully explained. Consequently, the parameter r cannot be understood.
-- Tables and Figures often lack dimensions, i.e., is is unclear what the
-axes represent. Table 1 is more or less useless without providing
-dimensions for the various parameters.
-- The author summary seems to provide more information about the results
-than the abstract. However, the authors use SPVL and progression to AIDS
-as two different proxies for virulence which can be confusing.
+\response{We agree that we have compromised the complexity of
+the disease life-history model somewhat in order to explore 
+the effects of epidemiological complexity.
+Figure 1 represents our effort to show that
+this simplification does not have a large impact on the results
+(although it's always possible that there could be an interaction
+between disease life-history complexity and epidemiological complexity
+\ldots); we have tried to clarify this point (see ll. 84ff.).
 
-Model:
-The authors make use of a previously published model from ref. 14. This
+\textbf{Structure}
+
+\begin{itemize}
+\item \revcomment{The references to the equations are not always correct.}
+\response{We have corrected all the errors we could find.}
+\item \revcomment{Figure 1 is described in Materials and Methods before the model is fully explained. Consequently, the parameter r cannot be understood.}
+\response{We have added some detail to the caption.}
+\item \revcomment{Tables and Figures often lack dimensions, i.e., is is unclear what the axes represent. Table 1 is more or less useless without providing
+dimensions for the various parameters.}
+\response{We have added dimensions throughout.}
+\item \revcomment{The author summary seems to provide more information about the results than the abstract.
+However, the authors use SPVL and progression to AIDS as two different proxies for virulence which can be confusing.}
+\response{We have thoroughly revised the abstract in response to reviewer \#1's comments.
+Commenting on both SPVL and progression time is admittedly tricky; in our revised version we have focused on SPVL only where making explicit comparisons with Shirreff \emph{et al.}'s models, and have otherwise stated results in terms of progression time.}
+\end{itemize}
+
+\textbf{Model}
+
+\begin{itemize}
+\item \revcomment{The authors make use of a previously published model from ref. 14. This
 model is based on the pair-formation formalism initially developed by
-Dietz & Hadeler (1988, J Math Biol, PMID: 3351391) and later used by
-Kretzschmar et al. (1994 & 1998, Math Biosci, PMID: 7833594 & 9597826).
+Dietz \& Hadeler (1988, J Math Biol, PMID: 3351391) and later used by
+Kretzschmar et al. (1994 \& 1998, Math Biosci, PMID: 7833594 \& 9597826).
 I feel that the authors should refer to some of these original
-publications as well. Unfortunately, the description of the models in
+publications as well.}
+\response{We have added citations to these references.}
+
+\item \revcomment{Unfortunately, the description of the models in
 appendix S1 seems to be partially erroneous or incorrect. For example,
@@
 and 25) or it is unclear whether equations represent state variables or
-derivates thereof. Also, the reasoning behind the “instswitch” models is
+derivates thereof.}
+\response{We have fixed these errors.}
+}
+\item \revcomment{ Also, the reasoning behind the “instswitch” models is
 unclear to me. The authors say that “Once individuals leave a
 partnership, they enter temporary compartments…”. The variables (or
-derivatives?) X and Y_i in equation 19 consist of positive terms only,
-so how do individuals leave these compartments? In my understanding the
+derivatives?) $X$ and $Y_i$ in equation 19 consist of positive terms only,
+so how do individuals leave these compartments?}
+\response{We have clarified that these terms are partnership
+leaving rates, and have added several lines of text clarifying
+the meaning of the term.}
+\item \revcomment{
+ In my understanding the
 standard pair model (equation 9) can easily account for instantaneous
 partnership formation by simply setting the pair formation rate rho to a
-high value.
- ------------------------------------------------------------------------
+high value}
+\response{We have added the following sentence (ll. 141-145):
+``Although these models can also be implemented
+by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.''}
 
-*Reviewers who wish to reveal their identities to the authors and other
-reviewers should include their name here (optional). These names will
-not be published with the manuscript, should it be accepted.*
-
-Reviewer #1: (No Response)
-
-Reviewer #2: (No Response)
+\end{itemize}
 

### 2016-10-25 ms/plos_latex.tex
cover letter; ms tweaks; fig tweaks

@@
 Models of intermediate
-complexity as used by Fraser \emph{et al.} \todo{why not shirreff et al?} 
+complexity as used by Shirreff \emph{et al.}
 predicted lower slower progression/lower virulence (a minimum of 15 years to
@@
 sexual partnerships with one infected and one uninfected partner), HIV
-virulence as measured by the rate of progression to AIDS was both
-heritable and covaried with the set-point viral load (i.e., the
+virulence as measured by the rate of progression to AIDS was 
+both heritable and covaried with the set-point viral load (SPVL: i.e., the
 characteristic virus load measured in blood during the intermediate
-stage of infection), which in turn predicted the probability of
-transmission
+stage of infection) and the probability of transmission.
+Higher viral loads led to shorter progression times (higher virulence)
+and faster transmission, with a decelerating relationship
+between virulence and transmission as required by the tradeoff theory
 \cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
-\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
+\cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
 parameterize mechanistic models of HIV virulence evolution, suggesting
@@
 
-Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than the rate of progression to AIDS itself
-(hereafter ``virulence'' will refer either to the SPVL, or to 
-the rate of progression to AIDS; these two quantities are
-deterministically linked in the model) \todo{Doesn't this somewhat contradict with what we say later on?
-Here, we say we focus on SPVL but later we say we summarize the models using progression time rather than SPVL...}.
-In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period; we show below that our results are not overly
-sensitive to this simplification.
-We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \etal's model) and taking the duration-weighted average of transmission rates of three stages.
-Thus the within-couple transmission rate, $\beta$, for our models is given by:
+As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
+explicitly track the evolution of mean $\log_{10}$ SPVL
+(which we denote as $\alpha$), rather than the rate of progression to
+AIDS itself (hereafter ``virulence'' will refer either to the SPVL, or
+to the rate of progression to AIDS; these two quantities are
+deterministically linked in the model).  
+In contrast to Shirreff \etal, we use a
+single-stage disease model instead of accounting explicitly for
+progression through the three main stages of HIV infection (primary,
+asymptomatic, and disease), and we use a simple exponentially
+distributed infectious period instead of a more realistic
+Weibull-distributed infectious period; we show below that our results
+are not overly sensitive to this simplification. We account for
+varying transmission rates and durations of each disease stage by
+summing the durations of three stages (again based on Shirreff \etal's
+model) and taking the duration-weighted average of transmission rates
+of three stages. Thus the within-couple transmission rate, $\beta$,
+for our models is given by:
 \begin{equation}
@@
 Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.
-\todo{Although these models can also be achieved by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.}
+Although these models can also be implemented
+by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.
 
@@
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-\todo{does this sentence regarding heterogeneity feel redundant?} For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
+For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
 We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
@@
 Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity.
-%%  BMB: I don't think we need to go into detail about variance vs CV
 To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men, as that was the group with the largest variability \cite{omori2015dynamics}.
@@
 Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
-\cite{soetaert_solving_2010}) for all simulations. \todo{mention that we got rid of error runs for hetero model?}
+\cite{soetaert_solving_2010}) for all simulations. 
+(For the heterogeneous model, approximately
+10\% of the samples failed due to numerical instability.)
 
@@
 Once individuals leave a partnership, they are instantaneously distributed into coupled states.
-In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the number of susceptible individuals that leave partnership at a given time, and $Y_i$ the number of infected individuals with \Lspvl\ of $\alpha_i$ who leave partnership at a given time.
-These temporarily single individuals then form couples through the same partnership formation rule described in the previous section:
+In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the total rate at which susceptible individuals leave partnership at a given time, and $Y_i$ is the total partnership leaving rate of infected individuals with \Lspvl\ of $\alpha_i$ at a given time.
+These temporarily single individuals then form couples through the same partnership formation rules described in the previous section:
 
@@
 Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group.
-In other words, an individual in a higher risk is also more likely to form a \todo{long-term sexual partnership instead of stable?} stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing.
+In other words, an individual in a higher risk is also more likely to form a long-term sexual partnership, leave a long-term partnership, and engage in a extra-couple/uncoupled mixing.
 We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group.
@@
 We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners.
-Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate.
-\todo{Herbeck 2016 does this.
-We can cite them?}
+Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate \cite{herbeck_evolution_2016}.
 

### 2016-10-25 ms/plosHIVcover_resubm.tex
update cover letter

@@
 In addition to extensive clarification and correction following
-the reviewers' comments, we implemented a more complex model
-that includes variation in contact rates among individuals.
+the reviewers' comments (see detailed response below), 
+we have implemented an additional, more complex model
+that includes heterogeneity in contact rates among individuals,
+in order to address the reviewers' concern that we
+did not look at a sufficiently realistic set of 
+epidemiological models.
 
+\vskip10pt
 \hrule

### 2016-10-26 ms/plos_latex.tex
moving supporting information to different files + small tweaks

@@
 multiplying the within-couple transmission rate $\beta$ by the contact
-ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix). Simplifying the model
+ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1). Simplifying the model
 of HIV pathogenesis from three stages to a single stage could affect
@@
 Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero'').
-Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix).
+Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
@@
 $II_{ij}$ that records the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$. 
-The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix).
+The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix S1).
 
@@
 
-One parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (Appendix S1);
+One parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see Appendix S1);
 it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  so we need to pick a reasonable range for it.
@@
 random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model (see Appendix for 
+for the implicit model (see Appendix S2 for 
 plots showing univariate summaries
@@
 
-Since we use multi-strain models in which the distribution of \Lspvl\ has been discretized into a vector, we use a matrix notation to describe our models.
-The five states described in the \emph{Methods} section are replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$, where the subscripts denote the strain with which an individual is infected.
-For example, $I_i$ is number of infected individuals with \Lspvl\ of $\alpha_i$, and $II_{ij}$ is the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$ (independent of order; $II_{ij}$ is synonymous with $II_{ji}$).
-Below, we use the Kronecker delta (i.e.
-$\delta_{ij}=1$ if $i=j$ and 1 otherwise) in a slightly non-standard fashion as an exponent, e.g.
-$2^{\delta_{ij}}$, to set a value to 2 when $i=j$ and 1 otherwise.
-
-\subsection*{Models 1 (``pairform+epc'') and 2 (``pairform'')}
-
-\subsubsection*{Partnership dynamics}
-
-Single individuals acquire partners at per-person rate $\rho$.
-Partnership formation rates for $S$ and $I_i$ are $\rho S$ and $\rho I_i$, respectively.
-We follow Champredon \etal \cite{champredon_hiv_2013} in assuming that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:
-
-\begin{equation}
-\begin{aligned}
-\PF(SS) &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)},\\
-\PF(SI_i) &= \frac{\rho S \cdot I_i}{S + \sum_k I_k},\\
-\PF(II_{ij}) &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}.
-\end{aligned}
-\end{equation}
-
-Partnerships dissolve at per-partnership rate $c$: the dissolution rates for $SS$, $SI_i$, and $II_{ij}$ pairs are $c SS$, $c SI_i$, and $c II_{ij}$ respectively.
-Unlike a single-strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains with which the individuals in concordant partnership are infected (i.e.
-both partners in $II_{ii}$ enter $I_i$ whereas one partner in $II_{ij}$ enters the $I_i$ compartment while the other enters $I_j$).
-Thus, coupled individuals are distributed into single states through partnership dissolution (DS) rates:
-
-\begin{equation}
-\begin{aligned}
-\DS(S) &= 2 c SS + \sum_k c SI_k, \\
-\DS(I_i) &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}.\\
-\end{aligned}
-\end{equation}
-
-Combining the partnership formation and dissolution processes yields the following equation:
-
-\begin{equation}
-\begin{aligned}
-S' &= - \rho S + 2 c SS + \sum_k c SI_k \\
-I_i' &= - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}\\
-SS' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS\\
-SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i\\
-II_{ij}' &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k} - c II_{ij}
-\end{aligned}
-\end{equation}
-
-\subsubsection*{Pair-formation models: infection dynamics}
-
-Within-couple transmission (WT) occurs in both models.
-An infected partner in $SI$ partnership transmits virus to a susceptible partner at per-partnership rate $\beta$: $\WT(SI_i) = - \beta_i SI_i$.
-Since we assume that mutation occurs, $II_{ij}$ pairs, where $i \neq j$, can be formed from either $SI_i$ or $SI_j$ partnerships: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$.
-On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $\WT(II_{ii}) = M_{ii} \beta_i SI_i$.
-Using the Kronecker delta notation, we obtain the following set of equations for within-couple transmission dynamics:
-
-\begin{equation}
-\begin{aligned}
-\WT(SI_i) &= - \beta_i SI_i,\\
-\WT(II_{ij}) &=  \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji}. \beta_j SI_j)
-\end{aligned}
-\end{equation}
-
-Champredon \etal\ \cite{champredon_hiv_2013} define the proportion of infectious extra-couple and uncoupled contact through the following term:
-
-\begin{equation}
-P = \frac{c_u I + c_e (SI + 2 II)}{c_u (S + I) + 2 c_e(SS + SI + II)}.
-\end{equation}
-The effective uncoupled, $c_u$, and extra couple, $c_e$, contact rates are the product of uncoupled/extra-couple contact rate $\times$ rate of transmission per contact.
-Therefore, the transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above.
-Using this property, we modify the equation above as follows:
-
-\begin{equation}
-P = \frac{r_u I + r_e (SI + 2 II)}{r_u (S + I) + 2 r_e(SS + SI + II)},
-\end{equation}
-where $r_u = c_u/c_w$ and $r_e = c_e/c_w$ are the relative uncoupled/extra-couple contact rates.
-This simplification is useful in a multi-strain model since we cannot multiply a vector by a scalar value (e.g.
-$c_u S$ in denominator) if we use Champredon \etal's equation in its original form.
-Extending the above equation to the multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain:
-
-\begin{equation}
-P_i = \frac{r_u I_i + r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_u (S + \sum_k I_k) + r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k 2^\delta_{lk} II_{lk} )}.
-\end{equation}
-Using the equation above, we can model extra-pair transmission (ET).
-For convenience, uncoupled and extra-couple transmission rates, $c_u$ and $c_e$, will be replaced with $U_i = r_u \beta_i$ and $E_i = r_e \beta_i$ hereafter.
-
-Single susceptible individuals become infected through uncoupled contact at per-person rate $\sum_k P_k U_k$ and enter the single infected state.
-Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $\ET(I_i) = \sum_k M_{ki} P_k U_k S$.
-Either partner in an $SS$ partnership can become infected at per-person rate $\sum_k P_k E_k$, and partnership state changes to an $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$.
-The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $\ET(SI_i) = \sum_k 2 M_{ki} P_k E_k SS$.
-Lastly, the susceptible partner in an $SI$ partnership can become infected from extra-couple contacts at a per-person rate of $\sum_k P_k E_k$, so that the partnership changes to an $II$ partnership.
-As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$.
-The mutation process is similar to that of within-couple transmission.
-The only difference is that the \Lspvl\ of a newly infected partner is not determined by its social partner but from an extra-couple partner (i.e.
-the term $P_i$): $\ET(II_{ij}) = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$.
-Combining these equations we get the following set of equations that describe the complete transmission dynamics:
-
-\begin{equation}
-\begin{aligned}
-S' =& - \sum_k P_k U_k S,\\
-I_i' =& \sum_k M_{ki} P_k U_k S,\\
-SS' =&  - \sum_i 2 P_i E_i SS, \\
-SI_i' =& \sum_k 2 M_{ki} P_k E_k SS - \beta_i SI_i - \sum_k P_k E_k SI_i,\\
-II_{ij}' =& \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji}, \beta_j SI_j) + \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i\\
-&+ M_{ki} P_k E_k SI_j)).
-\end{aligned}
-\end{equation}
-
-\subsubsection*{Pair formation models: Disease induced mortality}
-
-The per-person disease induced mortality (DM) rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$.
-Since we assume an SIS formulation, where infected individuals that die from infection are immediately replaced by an individual in the single susceptible compartment, we obtain the following equation for single infected individuals:
-
-\begin{equation}
-\begin{aligned}
-\DM(S) &= \sum_k \lambda_k I_k, \\
-\DM(I_i) &= - \lambda_i I_i. \\
-\end{aligned}
-\end{equation}
-
-If an infected individual in a partnership dies, the partnership dissolves.
-Thus, an $SI_i$ partnership dissolves at per-partnership rate $\lambda_i$, and the susceptible partner enters the single susceptible compartment at rate $\lambda_i SI_i$ (due to the SIS formulation, the infected partner that dies also gives rise, at an equal rate, to an individual entering the single susceptible compartment):
-
-\begin{equation}
-\begin{aligned}
-\DM(S) &= \sum_k 2 \lambda_k SI_k, \\
-\DM(SI_i) &= - \lambda_i SI_i.
-\end{aligned}
-\end{equation}
-
-Similarly, since $II_{ij}$ partnerships are composed of two infected partners, they dissolve at a per-partnership rate $(\lambda_i + \lambda_j)$.
-However, two cases, when $i \neq j$ and $i = j$, must be considered separately.
-When an $II_{ij}$ partnership dissolves due to disease-induced mortality, where $i \neq j$, the death of the partner with strain $i$ causes its partner to enter $I_j$ compartment at rate $\lambda_j II_{ij}$, and vice versa.
-When an $II_ii$ partnership dissolves, the death of either partner causes the other partner to enter the $I_i$ compartment at rate $\lambda_i II_{ii}$, which sums up to $2\lambda_i II_{ii}$.
-Combining these dynamics yields:
-
-\begin{equation}
-\begin{aligned}
-\DM(S) &= \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk}, \\
-\DM(I_i) &=  \sum_k 2^{\delta_{ik}} \lambda_k II_{ik}, \\
-\DM(II_{ij}) &= -(\lambda_i + \lambda_j) II_{ij}.
-\end{aligned}
-\end{equation}
-
-Finally, combining all these equations give us the full model, which is Model 1. We can simply drop the uncoupled and extra-couple transmission terms to obtain model 2:
-
-\begin{equation}
-\begin{aligned}
-S' =& - \rho S + 2 c SS + \sum_k c SI_k - \sum_k P_k U_k S + \sum_k \lambda_k I_k \\
-&+ \sum_k 2 \lambda_k SI_k + \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk}\\
-I_i' =&  - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c  II_{ik} + \sum_k M_{ki} P_k U_k S- \lambda_i I_i \\
-&+ \sum_k 2^{\delta_{ik}} \lambda_k II_{ik} \\
-SS' =& \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS - \sum_i 2 P_i E_i SS \\
-SI_i' =& \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS - \sum_k P_k E_k SI_i   \\
-&- \lambda_i SI_i\\
-II_{ij}' =& \khalf \cdot \frac{\rho I_i \cdot I_j}{(S + \sum_k I_k)} - c II_{ij} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
-&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
-\end{aligned}
-\end{equation}
-
-\subsection*{Models 3 (``instswitch'') and 4 (``instswitch'')}
-\subsubsection*{Partnership dynamics}
-
-Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$.
-Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - c II_{ij}$.
-Once individuals leave a partnership, they are instantaneously distributed into coupled states.
-In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the total rate at which susceptible individuals leave partnership at a given time, and $Y_i$ is the total partnership leaving rate of infected individuals with \Lspvl\ of $\alpha_i$ at a given time.
-These temporarily single individuals then form couples through the same partnership formation rules described in the previous section:
-
-\begin{equation}
-\begin{aligned}
-X &= 2 c SS + \sum_k c SI_k \\
-Y_i &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik} \\
-SS' &= - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}\\
-SI_i' &= - c SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
-II_{ij}' &= - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k}.
-\end{aligned}
-\end{equation}
-
-\subsubsection*{Instantaneous-switching models: Infection dynamics}
-
-Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single (uncoupled) state, only extra-couple transmission occurs:
-
-\begin{equation}
-P_i = \frac{r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k (2^\delta_{kl} II_{lk}) )}.
-\end{equation}
-Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equations that are used in models 1 and 2.
-
-\subsubsection*{Instantaneous-switching models: Disease induced mortality}
-
-Disease-induced mortality is modeled similarly to model 1 and 2.
-However, as the single state does not exist in model 3 and 4, individuals that leave partnerships due to death of their partners form partnerships instantly:
-
-\begin{equation}
-\begin{aligned}
-X &= \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk}, \\
-Y_i &=  \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik}, \\
-SS' = &= \frac{X^2}{2 (X + \sum_k Y_k)},\\
-SI_i' &= - \lambda_i SI_i + \frac{X Y_i}{X + \sum_k Y_k},\\
-II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + \khalf \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}.
-\end{aligned}
-\end{equation}
-
-Combining all these dynamics, we have model 3. If we remove extra-couple transmission, we have model 4.
-
-\begin{equation}
-\begin{aligned}
-X =& 2 c SS + \sum_k c SI_k + \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk},\\
-Y_i =& c SI_i + \sum_k 2^{\delta_{ik}}  c II_{ik} + \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik}, \\
-SS'  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS,\\
-SI_i' =& - c SI_i + \frac{X Y_i}{X + \sum_k Y_k} - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS\\
-&- \sum_k P_k E_k SI_i - \lambda_i SI_i,\\
-II_{ij}'& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)\\
-&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}.
-\end{aligned}
-\end{equation}
-
-\subsection*{Implicit model}
-
-Following Shirreff \etal \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta^\ast$, that is derived from Hollingsworth \etal's approximate basic reproduction number \cite{hollingsworth_hiv1_2008}:
-
-\begin{equation}
-\beta^\ast_i = \frac{c \beta_i}{c + \beta_i + \lambda_i}.
-\end{equation}
-Thus, we get the following model:
-
-\begin{equation}
-\begin{aligned}
-S' & = \sum_k \lambda_k I_k - \sum_k \beta^\ast_k S I_k.\\
-I_i' & = \sum_k M_{ki} \beta^\ast_k S I_k - \lambda_i I_i.
-\end{aligned}
-\end{equation}
-
-\subsection*{Random-mixing model}
-
-Model 6 is a random-mixing model. It is modeled in a same way as model 5 without the adjusted transmission rate:
-
-\begin{equation}
-\begin{aligned}
-S' & = \sum_k \lambda_k I_k - \sum_k c \beta_k S I_k,\\
-I_i' & = \sum_k M_{ki} c \beta_k S I_k - \lambda_i I_i.
-\end{aligned}
-\end{equation}
-
-\subsection*{Heterogenous model}
-
-We extend the ``pairform+epc'' model by allowing for heterogeneity in sexual behaviour.
-Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group.
-In other words, an individual in a higher risk is also more likely to form a long-term sexual partnership, leave a long-term partnership, and engage in a extra-couple/uncoupled mixing.
-We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group.
-For simplicity, we assume that the transmission rate per partnership is unaffected by sexual behaviour.
-
-\subsubsection*{Partnership dynamics}
-
-Individuals in a risk group $i$ leave single state at per-person rate $\varphi_i \rho$.
-Let $XY_{ij,kl}$ be a coupled state where $X$ and $Y$ are the infection status (susceptible or infected) of each partner, $k$ and $l$ are the risk groups $X$ and $Y$ belong to respectively, and $i$ and $j$ are the strains of an infected partner.
-If a partner is susceptible, strain index is replaced by $\cdot$.
-For example, $SI_{\cdot j,kl}$ is the number of partners where the susceptible partner is in risk group $k$ and infected partner is in risk group $l$ and has \Lspvl\ of $\alpha_j$.
-For simplicity, we assume that people undergo random \emph{activity-weighted} mixing \cite{may_transmission_1988}.
-Then we can write the partnership formation process as follows:
-
-\begin{equation}
-\PF(XY_{ij, kl}) = \kkhalf \frac{\varphi_k \rho X_{i,k} \varphi_l \rho Y_{j, l}}{\sum\limits_m \varphi_m \rho (S_{\cdot, m} + \sum\limits_n I_{n,m})}
-\end{equation}
-
-For dissolution process, an individual in risk group $i$ leaves its partnership at a rate $\varphi_i c$.
-If a partnership is formed between two individuals from a different risk group, the rates at which they leave the partnership differ.
-We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners.
-Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate \cite{herbeck_evolution_2016}.
-
-\subsubsection*{Heterogeneous models: Infection dynamics}
-
-Since we assume that the rate of transmission per partnership stays constant across different risk groups, the within-couple infection process is similar to other models:
-
-\begin{equation}
-\begin{aligned}
-WT(SI_{\cdot j, kl} &= - \beta_j SI_{\cdot j, kl}\\
-WT(II_{ij, kl}) &= \kkhalf \cdot (M_{ji} \beta_j SI_{\cdot j, kl} + M_{ij} \beta_i SI_{\cdot i, lk})
-\end{aligned}
-\end{equation}
-
-Note that $II_{ij,kl}$ can be formed from two types of partnerships: 1) Infected partner with \Lspvl\ of $\alpha_j$ and risk group of $l$ infects a susceptible partner in risk group $k$, yielding \Lspvl\ of $\alpha_i$ through mutation.
-2) Infected partner with \Lspvl\ of $\alpha_i$ and risk group of $k$ infected a susceptible partner in risk group $l$, yielding \Lspvl\ of $\alpha_j$ through mutation.
-On the other hand, if $i = j$ and $k = l$, $II_{ii,kk}$ can only be formed from $SI_{\cdot i, kk}$ partnership, which is resolved by $\kkhalf$.
-
-The heterogeneous extra-couple and uncoupled contact process is similar to the partnership formation process (activity-weighted random mixing).
-Relative uncoupled/extra-couple contact rates are scaled by the factor of $\varphi_i$, where $i$ is the risk group.
-First, we define $Q_{i}$ as the total rate of uncoupled/extra couple contact by individuals in risk group $k$:
-
-\begin{equation}
-\begin{aligned}
-Q_i = &\varphi_i r_u (S_{\cdot,i} +  \sum_j I_{j,i}) + \varphi_i r_e \bigg( \sum_k 2^{\delta_{ik}} SS_{\cdot, ik} +\\
-&\sum_l \sum_j (SI_{\cdot j,il} + SI_{\cdot j, li}) + \sum_j \sum_l \sum_k 2^{\delta_{kl} \delta_{ij}} II_{kl,ij} \bigg)
-\end{aligned}
-\end{equation}
-We now define $P_{k,i}$ as the proportion of the extra-couple and uncoupled contact that arises from an infected individual from risk group $i$ with \Lspvl\ of $\alpha_k$:
-\begin{equation}
-P_{k,i} = \frac{\varphi_i r_u I_{k,i} + \varphi_i r_e (SI_{k,i} + \sum_j \sum_l 2^{\delta_{kl} \delta_{ij} } II_{kl,ij} )}{\sum_j Q_j}
-\end{equation}
-Since the relative uncoupled/extra couple contact ratios are scaled by the factor of $\varphi_i$, uncoupled and extra-couple transmission rates are scaled by the same factor as well: $U_{k,i} = \varphi_i r_u \beta_k$ and $E_{k,i} = \varphi_i r_e \beta_k$.
-Once again, we assume activity-weighted mixing between individuals.
-Then, a susceptible individual in risk group $i$ becomes infected through extra-couple and uncoupled contact at a per capita rate of $\sum_j \sum_k P_{k,j} X_{k,i}$.
-Once infected, individuals are distributed into strain categories through mutation.
-
-\subsubsection*{Heterogeneous model: Disease induced mortality}
-
-Disease induced mortality is not affected by the sexual behaviour of an individual.
-
-\subsection*{Initial distribution of infected individuals}
-
-We follow Champredon \etal's result to calculate the initial distribution of infected individuals.
-For model 1 and 2, we have disease equilibrium state of $S^* = \frac{c}{c + \rho}$ and $SS^* = \frac{1-S^*}{2}$.
-We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with \Lspvl\ of $i$.
-$D_i$ is taken from the Gaussian distribution with mean 3 and standard deviation 0.2 and is normalized so that $\sum_i D_i = 1$.
-Then, we have the following initial distribution of each states:
-
-\begin{equation}
-\begin{aligned}
-S(0) &= (1 - \epsilon) S^*, \\
-SS(0) &= (1 - \epsilon)^2 SS^*,\\
-SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i,\\
-I_i(0) &=  \epsilon S^* D_i,\\
-II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
-\end{aligned}
-\end{equation}
-Since model 3 and 4 do not have single states, $SS^*=1$ at the disease free equilibrium and the initial distribution becomes:
-
-\begin{equation}
-\begin{aligned}
-SS(0) &= (1 - \epsilon)^2 SS^*,\\
-SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i,\\
-II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
-\end{aligned}
-\end{equation}
-As model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution:
-
-\begin{equation}
-\begin{aligned}
-S(0) &= 1 - \epsilon,\\
-I_i(0) &=  \epsilon D_i.
-\end{aligned}
-\end{equation}
-Model 6 has the same distribution of initial infected individuals as model 5.
-
-Lastly, for the heterogeneity model, we assume that the risk distribution of the population follows a Gamma distribution and calculate the shape and scale parameters from the mean and squared coefficient of variation.
-Using the shape and scale parameters, we define the Gamma quantile function $Q(p)$ and $p_j =  \tsub{p}{min} + (\tsub{p}{max} - \tsub{p}{min}) \frac{j-1}{n_r + 1}$, where $n_r$ is number of risk groups and $j = 1, 2, 3, \dots, n_r + 1$.
-Since $Q(1) = \infty$, we set $\tsub{p}{max} = 0.99$ and $\tsub{p}{min} = 0.01$.
-Then, we define $\varphi_i = \frac{Q(p_j) + Q(p_{j+1})}{2}$.
-We define $R_i$ as the proportion of individuals in risk group $i$ at the disease-free equilibrium and assume $R_i$ is equal for all $i$, i.e.
-$R_i = \frac{1}{n_r}$.
-In order to start the simulation in a quasi-equilibrium state, we first run the model with the following initial state:
-
-\begin{equation}
-\begin{aligned}
-S_{\cdot,i}(0) &= (1 - \epsilon) R_i,\\
-I_{k,i}(0) &= \epsilon D_k R_i,\\
-SS_{\cdot,ij}(0) &= SI_{\cdot k, ij}(0) = II_{kl,ij} (0) = 0.\\
-\end{aligned}
-\end{equation}
-For this particular simulation, we disregard the infection process as well as disease-induced mortality in order to preserve the strain distribution of infected individuals.
-Furthermore, since the scaling parameter, $\gamma$, does not affect the risk group distribution in the absence of disease transmission, we increase the scaling parameter to 5 ($\gamma = 5$) to speed up the simulation and run the model for 50 years.
-After the model has reached its quasi-equilibrium state, we take this distribution of susceptible and infected individuals as the initial state of the actual simulation.
-
 \paragraph*{Appendix S2: dynamics of transmission and virulence}
@@
 \label{S2_Appendix}
-
-This section presents alternate versions of
-the figures from the main text showing time dynamics and summary
-statistics in terms of \Lspvl\ and per-year transmission probability
-rather than expected progression time to AIDS.
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig_S2_1.pdf}
-\caption{{\bf Envelopes of transmission trajectories under all models.}
-This figure matches \figurename~\ref{fig:virtraj}, but displays the
-envelope of population-mean transmission probabilities rather than \Lspvl\ over time
-for each model.
-}
-\label{fig:transtraj}
-\end{figure}
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig_S2_2.pdf}
-\caption{{\bf Envelopes of progression trajectories under all models.}
-This figure matches \figurename~\ref{fig:virtraj}, but displays the
-envelope of population-mean expected time of progression to AIDS (i.e., length of
-intermediate HIV phase) rather than \Lspvl\ over time
-for each model.
-}
-\label{fig:durtraj}
-\end{figure}
-
-\begin{figure}[!ht]
-  \includegraphics[width=\textwidth]{../figures/fig_S2_3.pdf}
-\caption{{\bf Univariate distributions of transmission probabilities and progression.}
-This figure matches \figurename~\ref{fig:unidist}, but displays the
-univariate distributions for the transmission probability and 
-progression time at the virulence
-peak and at the epidemiological equilibrium,
-rather than the distributions of \Lspvl.
-}
-\label{fig:tranprogsum}
-\end{figure}
 

### 2016-10-26 ms/plos_supp1.tex
moving supporting information to different files + small tweaks

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=1.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% Remove comment for double spacing
\usepackage{setspace} 
\doublespacing

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

%% Include all macros below
\newcommand{\khalf}{\left(\frac{1}{2}\right)^{\delta_{ij}}}  % (1/2)^kronecker
\newcommand{\kkhalf}{\left(\frac{1}{2}\right)^{\delta_{ij} \delta_{kl}}}  % (1/2)^(kronecker * kronecker)
\newcommand{\Lspvl}{$\log_{10}$ SPVL}
\newcommand{\rzero}{{\mathcal R}_0}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\PF}{\textrm{PF}}
\newcommand{\DS}{\textrm{DS}}
\newcommand{\WT}{\textrm{WT}}
\newcommand{\ET}{\textrm{ET}}
\newcommand{\DM}{\textrm{DM}}

%% END MACROS SECTION

\begin{document}

\section*{Supporting Information S1: model details}
\label{S1_Appendix}

Since we use multi-strain models in which the distribution of \Lspvl\ has been discretized into a vector, we use a matrix notation to describe our models. The five states described in the \emph{Methods} section are replaced with the following notations: $S$, $I_i$, $SS$, $SI_i$, $II_{ij}$, where the subscripts denote the strain with which an individual is infected. For example, $I_i$ is number of infected individuals with \Lspvl\ of $\alpha_i$, and $II_{ij}$ is the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$ (independent of order; $II_{ij}$ is synonymous with $II_{ji}$). 
Below, we use the Kronecker delta (i.e. $\delta_{ij}=1$ if $i=j$ and 1 otherwise) in a slightly non-standard fashion as an exponent, e.g. $2^{\delta_{ij}}$, to set a value to 2 when $i=j$ and 1 otherwise.

\subsection*{Models 1 (``pairform+epc'') and 2 (``pairform'')}

\subsubsection*{Partnership dynamics}

Single individuals acquire partners at per-person rate $\rho$. Partnership formation rates for $S$ and $I_i$ are $\rho S$ and $\rho I_i$, respectively. We follow Champredon \etal \cite{champredon_hiv_2013} in assuming that single individuals are distributed into coupled states with pair-formation (PF) rates as follows:

\begin{equation}
\begin{aligned}
\PF(SS) &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)},\\
\PF(SI_i) &= \frac{\rho S \cdot I_i}{S + \sum_k I_k},\\
\PF(II_{ij}) &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k}.
\end{aligned}
\end{equation}

Partnerships dissolve at per-partnership rate $c$: the dissolution rates for $SS$, $SI_i$, and $II_{ij}$ pairs are $c SS$, $c SI_i$, and $c II_{ij}$ respectively. Unlike a single-strain model, where both individuals leaving the $II$ partnership would enter $I$, we have to account for strains with which the individuals in concordant partnership are infected (i.e. both partners in $II_{ii}$ enter $I_i$ whereas one partner in $II_{ij}$ enters the $I_i$ compartment while the other enters $I_j$). Thus, coupled individuals are distributed into single states through partnership dissolution (DS) rates:

\begin{equation}
\begin{aligned}
\DS(S) &= 2 c SS + \sum_k c SI_k, \\
\DS(I_i) &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}.\\
\end{aligned}
\end{equation}

Combining the partnership formation and dissolution processes yields the following equation:

\begin{equation}
\begin{aligned}
S' &= - \rho S + 2 c SS + \sum_k c SI_k \\
I_i' &= - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik}\\
SS' &= \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS\\
SI_i' &= \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i\\
II_{ij}' &= \khalf \cdot \frac{\rho I_i \cdot I_j}{S + \sum_k I_k} - c II_{ij}
\end{aligned}
\end{equation}

\subsubsection*{Pair-formation models: infection dynamics}

Within-couple transmission (WT) occurs in both models. An infected partner in $SI$ partnership transmits virus to a susceptible partner at per-partnership rate $\beta$: $\WT(SI_i) = - \beta_i SI_i$. Since we assume that mutation occurs, $II_{ij}$ pairs, where $i \neq j$, can be formed from either $SI_i$ or $SI_j$ partnerships: $\WT(II_{ij}) = M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j$. On the other hand, $II_{ii}$ can only be formed from an $SI_i$ partnership: $\WT(II_{ii}) = M_{ii} \beta_i SI_i$. Using the Kronecker delta notation, we obtain the following set of equations for within-couple transmission dynamics:

\begin{equation}
\begin{aligned}
\WT(SI_i) &= - \beta_i SI_i,\\
\WT(II_{ij}) &=  \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji}. \beta_j SI_j)
\end{aligned}
\end{equation}

Champredon \etal\ \cite{champredon_hiv_2013} define the proportion of infectious extra-couple and uncoupled contact through the following term:

\begin{equation}
P = \frac{c_u I + c_e (SI + 2 II)}{c_u (S + I) + 2 c_e(SS + SI + II)}.
\end{equation}
The effective uncoupled, $c_u$, and extra couple, $c_e$, contact rates are the product of uncoupled/extra-couple contact rate $\times$ rate of transmission per contact. Therefore, the transmission rate per contact term in $c_u$ and $c_e$ is canceled out in the equation above. Using this property, we modify the equation above as follows:

\begin{equation}
P = \frac{r_u I + r_e (SI + 2 II)}{r_u (S + I) + 2 r_e(SS + SI + II)},
\end{equation}
where $r_u = c_u/c_w$ and $r_e = c_e/c_w$ are the relative uncoupled/extra-couple contact rates. This simplification is useful in a multi-strain model since we cannot multiply a vector by a scalar value (e.g. $c_u S$ in denominator) if we use Champredon \etal's equation in its original form. Extending the above equation to the multi-strain model so that $P_i$ represents the proportion of the extra-couple and uncoupled contact of an infected individual with strain $i$, we obtain:

\begin{equation}
P_i = \frac{r_u I_i + r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_u (S + \sum_k I_k) + r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k 2^\delta_{lk} II_{lk} )}.
\end{equation}
Using the equation above, we can model extra-pair transmission (ET). For convenience, uncoupled and extra-couple transmission rates, $c_u$ and $c_e$, will be replaced with $U_i = r_u \beta_i$ and $E_i = r_e \beta_i$ hereafter.

Single susceptible individuals become infected through uncoupled contact at per-person rate $\sum_k P_k U_k$ and enter the single infected state. Through mutation, newly infected individuals are distributed into single infected compartments with different strains: $\ET(I_i) = \sum_k M_{ki} P_k U_k S$. Either partner in an $SS$ partnership can become infected at per-person rate $\sum_k P_k E_k$, and partnership state changes to an $SI$ partnership at the total rate of $\sum_i 2 P_i E_i SS$. The formation of $SI_i$ partnerships is similar to the process through which single susceptible individuals are distributed into single infected compartments: $\ET(SI_i) = \sum_k 2 M_{ki} P_k E_k SS$. Lastly, the susceptible partner in an $SI$ partnership can become infected from extra-couple contacts at a per-person rate of $\sum_k P_k E_k$, so that the partnership changes to an $II$ partnership. As in the previous cases, $SI_i$ partnerships are lost at a rate of $\sum_k P_k E_k SI_i$. The mutation process is similar to that of within-couple transmission. The only difference is that the \Lspvl\ of a newly infected partner is not determined by its social partner but from an extra-couple partner (i.e. the term $P_i$): $\ET(II_{ij}) = (\frac{1}{2})^{\delta_{ij}}(\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j))$. Combining these equations we get the following set of equations that describe the complete transmission dynamics:

\begin{equation}
\begin{aligned}
S' =& - \sum_k P_k U_k S,\\
I_i' =& \sum_k M_{ki} P_k U_k S,\\
SS' =&  - \sum_i 2 P_i E_i SS, \\
SI_i' =& \sum_k 2 M_{ki} P_k E_k SS - \beta_i SI_i - \sum_k P_k E_k SI_i,\\
II_{ij}' =& \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji}, \beta_j SI_j) + \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i\\
&+ M_{ki} P_k E_k SI_j)).
\end{aligned}
\end{equation}

\subsubsection*{Pair formation models: Disease induced mortality}

The per-person disease induced mortality (DM) rate, $\lambda$, is given by taking the reciprocal of the total duration of the infection: $\lambda_i = 1/(D_A + D_P(\alpha_i) + D_D)$. Since we assume an SIS formulation, where infected individuals that die from infection are immediately replaced by an individual in the single susceptible compartment, we obtain the following equation for single infected individuals:

\begin{equation}
\begin{aligned}
\DM(S) &= \sum_k \lambda_k I_k, \\
\DM(I_i) &= - \lambda_i I_i. \\
\end{aligned}
\end{equation}

If an infected individual in a partnership dies, the partnership dissolves. Thus, an $SI_i$ partnership dissolves at per-partnership rate $\lambda_i$, and the susceptible partner enters the single susceptible compartment at rate $\lambda_i SI_i$ (due to the SIS formulation, the infected partner that dies also gives rise, at an equal rate, to an individual entering the single susceptible compartment):

\begin{equation}
\begin{aligned}
\DM(S) &= \sum_k 2 \lambda_k SI_k, \\
\DM(SI_i) &= - \lambda_i SI_i.
\end{aligned}
\end{equation}

Similarly, since $II_{ij}$ partnerships are composed of two infected partners, they dissolve at a per-partnership rate $(\lambda_i + \lambda_j)$. However, two cases, when $i \neq j$ and $i = j$, must be considered separately. When an $II_{ij}$ partnership dissolves due to disease-induced mortality, where $i \neq j$, the death of the partner with strain $i$ causes its partner to enter $I_j$ compartment at rate $\lambda_j II_{ij}$, and vice versa. When an $II_ii$ partnership dissolves, the death of either partner causes the other partner to enter the $I_i$ compartment at rate $\lambda_i II_{ii}$, which sums up to $2\lambda_i II_{ii}$. Combining these dynamics yields:

\begin{equation}
\begin{aligned}
\DM(S) &= \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk}, \\
\DM(I_i) &=  \sum_k 2^{\delta_{ik}} \lambda_k II_{ik}, \\
\DM(II_{ij}) &= -(\lambda_i + \lambda_j) II_{ij}.
\end{aligned}
\end{equation}

Finally, combining all these equations give us the full model, which is Model 1. We can simply drop the uncoupled and extra-couple transmission terms to obtain model 2:

\begin{equation}
\begin{aligned}
S' =& - \rho S + 2 c SS + \sum_k c SI_k - \sum_k P_k U_k S + \sum_k \lambda_k I_k \\
&+ \sum_k 2 \lambda_k SI_k + \sum_l \sum_k  2^{\delta_{lk}} \lambda_k II_{lk}\\
I_i' =&  - \rho I_i + c SI_i + \sum_k 2^{\delta_{ik}} c  II_{ik} + \sum_k M_{ki} P_k U_k S- \lambda_i I_i \\
&+ \sum_k 2^{\delta_{ik}} \lambda_k II_{ik} \\
SS' =& \frac{\rho S \cdot S}{2 (S + \sum_k I_k)} - c SS - \sum_i 2 P_i E_i SS \\
SI_i' =& \frac{\rho S \cdot I_i}{S + \sum_k I_k} - c SI_i - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS - \sum_k P_k E_k SI_i   \\
&- \lambda_i SI_i\\
II_{ij}' =& \khalf \cdot \frac{\rho I_i \cdot I_j}{(S + \sum_k I_k)} - c II_{ij} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j) \\
&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}
\end{aligned}
\end{equation}

\subsection*{Models 3 (``instswitch'') and 4 (``instswitch'')}
\subsubsection*{Partnership dynamics}

Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$. Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - c II_{ij}$. Once individuals leave a partnership, they are instantaneously distributed into coupled states. In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the total rate at which susceptible individuals leave partnership at a given time, and $Y_i$ is the total partnership leaving rate of infected individuals with \Lspvl\ of $\alpha_i$ at a given time. These temporarily single individuals then form couples through the same partnership formation rules described in the previous section:

\begin{equation}
\begin{aligned}
X &= 2 c SS + \sum_k c SI_k \\
Y_i &= c SI_i + \sum_k 2^{\delta_{ik}} c II_{ik} \\
SS' &= - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}\\
SI_i' &= - c SI_i + \frac{X Y_i}{X + \sum_k Y_k}\\
II_{ij}' &= - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k}.
\end{aligned}
\end{equation}

\subsubsection*{Instantaneous-switching models: Infection dynamics}

Model 3 and 4 share the within-couple transmission term with model 1 and 2. Since there is no single (uncoupled) state, only extra-couple transmission occurs:

\begin{equation}
P_i = \frac{r_e (SI_i + \sum_k (II_{ik} + \delta_{ik} II_{ik}))}{r_e(2 SS + \sum_k 2 SI_k + \sum_l \sum_k (2^\delta_{kl} II_{lk}) )}.
\end{equation}
Movement from $SS$ state to $SI$ state and $SI$ to $SS$ is modeled through the same equations that are used in models 1 and 2.

\subsubsection*{Instantaneous-switching models: Disease induced mortality}

Disease-induced mortality is modeled similarly to model 1 and 2. However, as the single state does not exist in model 3 and 4, individuals that leave partnerships due to death of their partners form partnerships instantly:

\begin{equation}
\begin{aligned}
X &= \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk}, \\
Y_i &=  \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik}, \\
SS' = &= \frac{X^2}{2 (X + \sum_k Y_k)},\\
SI_i' &= - \lambda_i SI_i + \frac{X Y_i}{X + \sum_k Y_k},\\
II_{ij}' &= -(\lambda_i + \lambda_j) II_{ij} + \khalf \cdot \frac{Y_i Y_j}{X + \sum_k Y_k}.
\end{aligned}
\end{equation}

Combining all these dynamics, we have model 3. If we remove extra-couple transmission, we have model 4.

\begin{equation}
\begin{aligned}
X =& 2 c SS + \sum_k c SI_k + \sum_k 2 \lambda_k SI_k + \sum_l \sum_k 2^{\delta_{lk}}  \lambda_k II_{lk},\\
Y_i =& c SI_i + \sum_k 2^{\delta_{ik}}  c II_{ik} + \sum_k  2^{\delta_{ik}}  \lambda_k II_{ik}, \\
SS'  =& - c SS + \frac{X^2}{2 (X + \sum_k Y_k)}  - \sum_i 2 P_i E_i SS,\\
SI_i' =& - c SI_i + \frac{X Y_i}{X + \sum_k Y_k} - \beta_i SI_i + \sum_k 2 M_{ki} P_k E_k SS\\
&- \sum_k P_k E_k SI_i - \lambda_i SI_i,\\
II_{ij}'& - c II_{ij} +\khalf \frac{Y_i Y_j}{X + \sum_k Y_k} + \khalf \cdot (M_{ij} \beta_i SI_i + M_{ji} \beta_j SI_j)\\
&+ \khalf \cdot (\sum_k (M_{kj} P_k E_k SI_i + M_{ki} P_k E_k SI_j)) -(\lambda_i + \lambda_j) II_{ij}.
\end{aligned}
\end{equation}

\subsection*{Implicit model}

Following Shirreff \etal \cite{shirreff_transmission_2011}, Model 5 is an implicit instantaneous partnership formation model that uses an adjusted transmission rate, $\beta^\ast$, that is derived from Hollingsworth \etal's approximate basic reproduction number \cite{hollingsworth_hiv1_2008}:

\begin{equation}
\beta^\ast_i = \frac{c \beta_i}{c + \beta_i + \lambda_i}.
\end{equation}
Thus, we get the following model:

\begin{equation}
\begin{aligned}
S' & = \sum_k \lambda_k I_k - \sum_k \beta^\ast_k S I_k.\\
I_i' & = \sum_k M_{ki} \beta^\ast_k S I_k - \lambda_i I_i.
\end{aligned}
\end{equation}

\subsection*{Random-mixing model}

Model 6 is a random-mixing model. It is modeled in a same way as model 5 without the adjusted transmission rate:

\begin{equation}
\begin{aligned}
S' & = \sum_k \lambda_k I_k - \sum_k c \beta_k S I_k,\\
I_i' & = \sum_k M_{ki} c \beta_k S I_k - \lambda_i I_i.
\end{aligned}
\end{equation}

\subsection*{Heterogenous model}

We extend the ``pairform+epc'' model by allowing for heterogeneity in sexual behaviour. Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group. In other words, an individual in a higher risk is also more likely to form a long-term sexual partnership, leave a long-term partnership, and engage in a extra-couple/uncoupled mixing. We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group. For simplicity, we assume that the transmission rate per partnership is unaffected by sexual behaviour.

\subsubsection*{Partnership dynamics}

Individuals in a risk group $i$ leave single state at per-person rate $\varphi_i \rho$. Let $XY_{ij,kl}$ be a coupled state where $X$ and $Y$ are the infection status (susceptible or infected) of each partner, $k$ and $l$ are the risk groups $X$ and $Y$ belong to respectively, and $i$ and $j$ are the strains of an infected partner. If a partner is susceptible, strain index is replaced by $\cdot$. For example, $SI_{\cdot j,kl}$ is the number of partners where the susceptible partner is in risk group $k$ and infected partner is in risk group $l$ and has \Lspvl\ of $\alpha_j$. For simplicity, we assume that people undergo random \emph{activity-weighted} mixing \cite{may_transmission_1988}. Then we can write the partnership formation process as follows:

\begin{equation}
\PF(XY_{ij, kl}) = \kkhalf \frac{\varphi_k \rho X_{i,k} \varphi_l \rho Y_{j, l}}{\sum\limits_m \varphi_m \rho (S_{\cdot, m} + \sum\limits_n I_{n,m})}
\end{equation}

For dissolution process, an individual in risk group $i$ leaves its partnership at a rate $\varphi_i c$. If a partnership is formed between two individuals from a different risk group, the rates at which they leave the partnership differ. We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners. Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate \cite{herbeck_evolution_2016}.

\subsubsection*{Heterogeneous models: Infection dynamics}

Since we assume that the rate of transmission per partnership stays constant across different risk groups, the within-couple infection process is similar to other models:

\begin{equation}
\begin{aligned}
WT(SI_{\cdot j, kl} &= - \beta_j SI_{\cdot j, kl}\\
WT(II_{ij, kl}) &= \kkhalf \cdot (M_{ji} \beta_j SI_{\cdot j, kl} + M_{ij} \beta_i SI_{\cdot i, lk})
\end{aligned}
\end{equation}

Note that $II_{ij,kl}$ can be formed from two types of partnerships: 1) Infected partner with \Lspvl\ of $\alpha_j$ and risk group of $l$ infects a susceptible partner in risk group $k$, yielding \Lspvl\ of $\alpha_i$ through mutation. 2) Infected partner with \Lspvl\ of $\alpha_i$ and risk group of $k$ infected a susceptible partner in risk group $l$, yielding \Lspvl\ of $\alpha_j$ through mutation. On the other hand, if $i = j$ and $k = l$, $II_{ii,kk}$ can only be formed from $SI_{\cdot i, kk}$ partnership, which is resolved by $\kkhalf$.

The heterogeneous extra-couple and uncoupled contact process is similar to the partnership formation process (activity-weighted random mixing). Relative uncoupled/extra-couple contact rates are scaled by the factor of $\varphi_i$, where $i$ is the risk group. First, we define $Q_{i}$ as the total rate of uncoupled/extra couple contact by individuals in risk group $k$:

\begin{equation}
\begin{aligned}
Q_i = &\varphi_i r_u (S_{\cdot,i} +  \sum_j I_{j,i}) + \varphi_i r_e \bigg( \sum_k 2^{\delta_{ik}} SS_{\cdot, ik} +\\
&\sum_l \sum_j (SI_{\cdot j,il} + SI_{\cdot j, li}) + \sum_j \sum_l \sum_k 2^{\delta_{kl} \delta_{ij}} II_{kl,ij} \bigg)
\end{aligned}
\end{equation}
We now define $P_{k,i}$ as the proportion of the extra-couple and uncoupled contact that arises from an infected individual from risk group $i$ with \Lspvl\ of $\alpha_k$:
\begin{equation}
P_{k,i} = \frac{\varphi_i r_u I_{k,i} + \varphi_i r_e (SI_{k,i} + \sum_j \sum_l 2^{\delta_{kl} \delta_{ij} } II_{kl,ij} )}{\sum_j Q_j}
\end{equation}
Since the relative uncoupled/extra couple contact ratios are scaled by the factor of $\varphi_i$, uncoupled and extra-couple transmission rates are scaled by the same factor as well: $U_{k,i} = \varphi_i r_u \beta_k$ and $E_{k,i} = \varphi_i r_e \beta_k$. Once again, we assume activity-weighted mixing between individuals. Then, a susceptible individual in risk group $i$ becomes infected through extra-couple and uncoupled contact at a per capita rate of $\sum_j \sum_k P_{k,j} X_{k,i}$. Once infected, individuals are distributed into strain categories through mutation.

\subsubsection*{Heterogeneous model: Disease induced mortality}

Disease induced mortality is not affected by the sexual behaviour of an individual.

\subsection*{Initial distribution of infected individuals}

We follow Champredon \etal's result to calculate the initial distribution of infected individuals. For model 1 and 2, we have disease equilibrium state of $S^* = \frac{c}{c + \rho}$ and $SS^* = \frac{1-S^*}{2}$. We let $\epsilon = 10^{-4}$, which is the total number of infected individuals in the beginning of simulation and $D$ be the vector such that $D_i$ represent the proportion of individuals with \Lspvl\ of $i$. $D_i$ is taken from the Gaussian distribution with mean 3 and standard deviation 0.2 and is normalized so that $\sum_i D_i = 1$. Then, we have the following initial distribution of each states:

\begin{equation}
\begin{aligned}
S(0) &= (1 - \epsilon) S^*, \\
SS(0) &= (1 - \epsilon)^2 SS^*,\\
SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i,\\
I_i(0) &=  \epsilon S^* D_i,\\
II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
\end{aligned}
\end{equation}
Since model 3 and 4 do not have single states, $SS^*=1$ at the disease free equilibrium and the initial distribution becomes:

\begin{equation}
\begin{aligned}
SS(0) &= (1 - \epsilon)^2 SS^*,\\
SI_i(0) &= 2 \epsilon (1-\epsilon) SS^* D_i,\\
II_{ij}(0) &=  \khalf 2\epsilon^2 SS^* D_i D_j.
\end{aligned}
\end{equation}
As model 5 is an implicit model, which does not consider different stages of partnership, we have the following initial distribution:

\begin{equation}
\begin{aligned}
S(0) &= 1 - \epsilon,\\
I_i(0) &=  \epsilon D_i.
\end{aligned}
\end{equation}
Model 6 has the same distribution of initial infected individuals as model 5.

Lastly, for the heterogeneity model, we assume that the risk distribution of the population follows a Gamma distribution and calculate the shape and scale parameters from the mean and squared coefficient of variation. Using the shape and scale parameters, we define the Gamma quantile function $Q(p)$ and $p_j =  \tsub{p}{min} + (\tsub{p}{max} - \tsub{p}{min}) \frac{j-1}{n_r + 1}$, where $n_r$ is number of risk groups and $j = 1, 2, 3, \dots, n_r + 1$. Since $Q(1) = \infty$, we set $\tsub{p}{max} = 0.99$ and $\tsub{p}{min} = 0.01$. Then, we define $\varphi_i = \frac{Q(p_j) + Q(p_{j+1})}{2}$. We define $R_i$ as the proportion of individuals in risk group $i$ at the disease-free equilibrium and assume $R_i$ is equal for all $i$, i.e. $R_i = \frac{1}{n_r}$. In order to start the simulation in a quasi-equilibrium state, we first run the model with the following initial state:

\begin{equation}
\begin{aligned}
S_{\cdot,i}(0) &= (1 - \epsilon) R_i,\\
I_{k,i}(0) &= \epsilon D_k R_i,\\
SS_{\cdot,ij}(0) &= SI_{\cdot k, ij}(0) = II_{kl,ij} (0) = 0.\\
\end{aligned}
\end{equation}
For this particular simulation, we disregard the infection process as well as disease-induced mortality in order to preserve the strain distribution of infected individuals. Furthermore, since the scaling parameter, $\gamma$, does not affect the risk group distribution in the absence of disease transmission, we increase the scaling parameter to 5 ($\gamma = 5$) to speed up the simulation and run the model for 50 years. After the model has reached its quasi-equilibrium state, we take this distribution of susceptible and infected individuals as the initial state of the actual simulation.

\bibliography{virulence}
\end{document}

### 2016-10-26 ms/plos_supp2.tex
moving supporting information to different files + small tweaks

--- 2016-10-26 ms/plos_latex.tex
@@
-% ## http://journals.plos.org/ploscompbiol/s/submit-now
-% Template for PLoS
-% Version 3.2 March 2016
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - tabs/spacing/line breaks within cells to alter layout or alignment
-% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
-% - colors, shading, or graphic objects
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
 \documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\usepackage[top=0.85in,left=1.75in,footskip=0.75in]{geometry}
 
@@
 \usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
 
@@
 \begin{document}
-\vspace*{0.2in}
 
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Effects of Epidemiological Structure on the Transient Evolution of HIV Virulence} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
-}
-%% short title: Epidemiological Structure and HIV Virulence Evolution
-%% (53/70 characters)
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Sang Woo Park\textsuperscript{1}
-Benjamin M. Bolker\textsuperscript{1,2,3,*}
-\\
-\bigskip
-\textbf{1} Department of Mathematics \& Statistics,  McMaster University, Hamilton, Ontario, Canada
-\\
-\textbf{2} Department of Biology,  McMaster University, Hamilton, Ontario, Canada
-\\
-\textbf{3} Institute for Infectious Disease Research,  McMaster University, Hamilton, Ontario, Canada
-\\
-\bigskip
+\paragraph*{Supporting Information S2: dynamics of transmission and virulence}
 
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-%\ddag These authors also contributed equally to this work.
-
-% Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-%\dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* bolker@mcmaster.ca
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-% current is 282
-\section*{Abstract}
-The evolutionary dynamics of parasite virulence change
-in important ways over the
-course of an emerging epidemic.
-Changes in the fitness landscape
-generally select for higher virulence early in an
-epidemic; however, quantitative outcomes
-may depend sensitively on epidemiological details and the structure of
-mathematical models used to portray them.  Fraser \emph{et al.} 
-have proposed a model for the eco-evolutionary
-dynamics of HIV that captures
-the tradeoffs between transmission and virulence (mediated by
-set-point viral load, SPVL) and their heritability between
-hosts. However, these models use implicit
-representations of the transmission process that drastically simplify the
-partnership dynamics that previous research has found to be critical
-in driving epidemics of sexually transmitted diseases.  
-Our models combine HIV virulence tradeoffs with a range of
-epidemiological structures, modeling partnership formation and
-dissolution and allowing for individuals to transmit disease outside
-of partnerships. We assess summary statistics such as the peak virulence
-(corresponding to the minimum expected time of progression to AIDS) across all
-models for a range of 
-partnership dynamic parameters 
-applicable to the HIV epidemic in sub-Saharan Africa.
-Although virulence trajectories are broadly similar
-across model structures, the timing and magnitude of the 
-minimum expected time to progression vary
-considerably.
-Models of intermediate
-complexity as used by Shirreff \emph{et al.}
-predicted lower slower progression/lower virulence (a minimum of 15 years to
-progress to AIDS) compared to both more realistic models
-and simple random-mixing models with no partnership structure
-at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
-In this range of models, the simplest random-mixing structure best
-approximates the most realistic model; this
-surprising outcome occurs because the dominance of extra-pair
-contact in the realistic model tends to swamp the effects of
-partnership structure.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author Summary}
-
-%% 162 words
-
-Pathogens such as HIV can evolve rapidly in response to changes in
-their environments; such changes include both increases in disease
-prevalence and disease virulence over the course of an epidemic, or
-decreases in both after treatment interventions. While researchers
-have successfully used computational models to explore these
-evolutionary dynamics, these models often neglect details such as the
-formation and dissolution of sexual partnerships; other research has
-shown that these processes can strongly affect epidemic outcomes. We
-built and compared models that used different methods to model both
-partnership dynamics and sexual contact outside of stable
-partnerships. Models of intermediate complexity predicted much lower
-virulence over the course of the epidemic (a minimum of 15 years to
-progress to AIDS) compared to both more realistic models and simple
-random-mixing models with no partnership structure at all (both
-approx. 7.25 years to progress to AIDS); sexual contact outside of
-stable partnerships tended to wash out the effects of epidemiological
-structure. The large differences in evolutionary dynamics among
-different epidemiological models suggests that researchers trying to
-predict the evolution of pathogens should proceed with caution.
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
-
-The evolution of pathogen virulence has both 
-theoretical and, potentially, practical
-importance. In general, evolutionary theory suggests that
-disease strains that can reproduce more --- where reproduction
-is defined here as the amount of \emph{between-host} transmission, or
-the number of new hosts infected --- will increase in prevalence.
-Pathogens can increase their net reproduction rate either
-by increasing their transmission rate, 
-the rate (per infected host) at which they
-infect new hosts, or by decreasing their clearance or disease-induced
-mortality rate, the rate
-at which hosts recover or die from disease.
-The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
-the transmission and disease-induced mortality rate are both linked to the rate at which
-the pathogen exploits host resources for within-host reproduction, 
-and that pathogen
-evolution will thus strike a balance between the
-pathogen's rate of transmission
-to new hosts and its rate of killing its host (or of provoking
-the host's immune system to eliminate it).
-Some biologists have criticized the tradeoff theory
-\cite{EbertBull2003,alizon_adaptive_2015}, but others have
-successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
-Fraser \etal\ have applied these ideas in a particularly
-interesting way 
-by showing that HIV appears to satisfy the prerequisites of
-the tradeoff theory: in studies of discordant couples (i.e. long-term
-sexual partnerships with one infected and one uninfected partner), HIV
-virulence as measured by the rate of progression to AIDS was 
-both heritable and covaried with the set-point viral load (SPVL: i.e., the
-characteristic virus load measured in blood during the intermediate
-stage of infection) and the probability of transmission.
-Higher viral loads led to shorter progression times (higher virulence)
-and faster transmission, with a decelerating relationship
-between virulence and transmission as required by the tradeoff theory
-\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
-\cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
-parameterize mechanistic models of HIV virulence evolution, suggesting
-that HIV invading a novel population would initially evolve increased
-virulence, peaking after approximately 100-200 years and then declining
-slightly to a long-stable virulence level.
-
-The work of Shirreff \etal\ \cite{shirreff_transmission_2011}, and particularly the predicted transient peak in HIV virulence midway through the epidemic,
-highlights the importance of interactions between epidemiological and
-evolutionary factors \cite{day_virulence_2004,alizon_price_2009}.
-However, despite these studies' attention to detail at the individual
-or physiological level, the epidemiological structures used in these
-models are relatively simple.
-
-As we discuss in detail below,
-existing models of HIV eco-evolutionary dynamics either use implicit
-models that incorporate the average effects of within-couple sexual
-contact --- without representing the explicit dynamics of pair
-formation and dissolution or accounting for extra-partnership contact
---- or use an agent-based formulation with parameters that effectively
-lead to random mixing among infected and uninfected individuals. Here
-we explore the effects of incorporating \emph{explicit}
-epidemiological structure in eco-evolutionary models.
-
-We add complexity to the epidemiological model following the general
-approach of Champredon \etal\ \cite{champredon_hiv_2013},
-which is in turn based on work of Dietz and Hadeler
-\cite{dietz_epidemiological_1988}; individuals
-join and leave partnerships at a specified rate, and can have sexual
-contact both within and outside of established partnerships. In order
-to explore how virulence evolution depends on epidemiological
-structure, we consider a series of models with increasing levels of
-complexity. In order to avoid dependence of the results on a
-particular set of parameters --- as we explain below, finding matching
-sets of parameters across models with widely differing epidemiological
-structures is challenging --- we evaluate our models across a wide
-range of parameters, again following Champredon
-\etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
-design. For each model run, we compute a set of metrics (minimum
-progression time/peak virulence, timing of maximum virulence, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.
-
-As our primary goal is to explore how different epidemiological
-structures (i.e. partnership dynamics and contact structures) affect
-our conclusions about the evolution of virulence, our models use a
-simplified description of within-host dynamics and heritability
-derived from Shirreff \etal's multi-strain evolutionary model
-\cite{shirreff_transmission_2011}. Like Shirreff \etal, we use a
-simple susceptible-infected-susceptible demographic formulation;
-rather than modeling birth and death (or more specifically,
-recruitment into the sexually active population and death), we assume
-that whenever an individual dies from infection, another enters the
-susceptible compartment.
-
-\section*{Materials and Methods}
-
-\subsection*{Infection dynamics}
-
-As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
-explicitly track the evolution of mean $\log_{10}$ SPVL
-(which we denote as $\alpha$), rather than the rate of progression to
-AIDS itself (hereafter ``virulence'' will refer either to the SPVL, or
-to the rate of progression to AIDS; these two quantities are
-deterministically linked in the model).  
-In contrast to Shirreff \etal, we use a
-single-stage disease model instead of accounting explicitly for
-progression through the three main stages of HIV infection (primary,
-asymptomatic, and disease), and we use a simple exponentially
-distributed infectious period instead of a more realistic
-Weibull-distributed infectious period; we show below that our results
-are not overly sensitive to this simplification. We account for
-varying transmission rates and durations of each disease stage by
-summing the durations of three stages (again based on Shirreff \etal's
-model) and taking the duration-weighted average of transmission rates
-of three stages. Thus the within-couple transmission rate, $\beta$,
-for our models is given by:
-\begin{equation}
-\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
-\end{equation}
-where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
-of infection are independent of the host's SPVL.
-Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:
-
-\begin{equation}
-\begin{split}
-D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
-\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
-{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
-\end{split}
-\label{eq:hillfuns}
-\end{equation}
-where $V_{\alpha} = 10^\alpha$.
-
-The uncoupled and extra-couple transmission rates (i.e., the rates of
-transmission among people outside of a stable partnership, or between
-people inside of a stable partnership and people other than their
-partner) are scaled by
-multiplying the within-couple transmission rate $\beta$ by the contact
-ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1). Simplifying the model
-of HIV pathogenesis from three stages to a single stage could affect
-our conclusions about the evolution of virulence (e.g. Kretzschmar and
-Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
-and multiple stages of infectivity have interactive effects on
-$\rzero$). However, our simplified model produces results that are
-qualitatively similar to those of Shirreff \etal's
-\cite{shirreff_transmission_2011} model; when our model is calibrated
-to have a similar initial epidemic growth rate $r$, the peak
-\Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly
-higher (4.6 \Lspvl\ vs. 4.3 \Lspvl, or 7\% higher: \figurename~\ref{fig:panel3}).
+This section presents alternate versions of
+the figures from the main text showing time dynamics and summary
+statistics in terms of \Lspvl\ and per-year transmission probability
+rather than expected progression time to AIDS.
 
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig1.pdf}
-\caption{{\bf Baseline dynamics.}
-Time series of mean population \Lspvl.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
-(b) Effects of varying initial infectious density $I(0)$.
-(c) Effects of varying initial mean virulence $\alpha(0)$.
-The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories.
-Panels b and c illustrate the sensitivity of virulence trajectories to initial conditions $I(0)$ and $\alpha(0)$, which we hold constant in our simulations.}
-\label{fig:panel3}
-\end{figure}
-
-\subsection*{Mutation}
-
-Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL.
-We simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping rather
-than allowing for variation in phenotypes of a single genotype.
-The mutational process in our model is directly taken from Shirreff \etal.
-Over the course of infection, mutation occurs within the host.
-However, it is assumed that SPVL of the strain transmitted by an infected individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation.
-Instead, the mutational effect takes place in a single
-step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
-\begin{equation}
-\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
-\end{equation}
-We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e.
-a bin width of 0.1 \Lspvl\ for $\alpha$), but reducing $n$ to 21 (bin width = 0.25 \Lspvl) makes little difference; we use this coarser grid for all other simulations reported.
-
-We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
-Finally, the probabilities are normalized so that each row sums to 1:
-\begin{equation}
-M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
-\end{equation}
-where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$.
-Transmission rate and disease induced mortality rates are discretized as well:
-\begin{equation}
-\begin{aligned}
-\beta_i &= \beta(\alpha_i),\\
-\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
-\end{aligned}
-\end{equation}
-
-\subsection*{Contact structure and partnership dynamics}
-
-We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
-Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
-non-instantaneous partnership formation; (2) zero vs.
-positive extra-partnership sexual contact and transmission; and (3) homogeneous vs.
-heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
-
-Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}.
-The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e.
-individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status.
-$S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals.
-$SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples.
-The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission.
-
-The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch'').
-The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
-Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.
-Although these models can also be implemented
-by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.
-
-The fifth and sixth models represent extreme simplifications of sexual partnership dynamics.
-One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
-It is actually a random mixing model that explicitly tracks only the total number of susceptible and infected individuals.
-However, to reflect the effect of partnership structure, it uses an adjusted transmission rate derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}.
-The second model of this pair (``random'') is a simple random-mixing model.
-
-Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero'').
-Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
-We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
-While this model lacks some
-important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
-As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
-
-The pairform+epc and heterogeneous models use the basic epidemiological framework of Champredon \etal \cite{champredon_hiv_2013}.
-Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$.
-Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
-Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
-Extra-couple and uncoupled transmission are modeled in the same way as in Champredon \etal's model.
-All the details have been adapted to a multi-strain scenario, so that we track (for example) a matrix
-$II_{ij}$ that records the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$. 
-The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix S1).
-
-\subsection*{Latin hypercube sampling}
-
-Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
-In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
-We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
-
-Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
-\begin{equation}
-z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1, 2, 3, \dots,N.
-\end{equation}
-Random permutations of these vectors form columns in a sample parameter matrix; each row contains a different parameter set that is used for one simulation run.
-
-Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
-Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
-The remaining parameters are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
-
-%Tables need to be placed after the first paragraph in which they are cited.
-
-\begin{table}[h!]
-\caption{Parameter ranges/values.
-Values of $c$ and $\rho$ are doubled from those given by Champredon \etal\ because we keep track of individuals in the model, while they keep track of couples.
-Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
-\centering
-\begin{tabular}{c p{2in} c l}
-\hline 
-Notation & Description & Range/Value & Source\\
-\hline % inserts single horizontal line
-$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
-$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
-$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
-$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
-$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
-$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
-$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
-$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
-$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
-$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
-$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
-$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
-$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
-$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
-$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
-$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
-$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
-$n$ & Number of strains & 21 (51*) & Assumption\\
-$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
-$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
-\hline
-\end{tabular}
-\label{table:parmsTable}
-\end{table}
-
-One parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see Appendix S1);
-it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  so we need to pick a reasonable range for it.
-Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
-Given Champredon \etal's parameter range, the possible maximum and minimum values of $c_u/c_w$ are 5 and 1/5.
-Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$.
-Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is not much known about the uncoupled transmission rate.
-
-Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity.
-To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men, as that was the group with the largest variability \cite{omori2015dynamics}.
-Omori \etal \cite{omori2015dynamics} give a very wide
-range for the coefficient of variation ($\approx$ 0 - 20, corresponding
-to squared CV range of 0-400):
-we narrowed this range for $\textrm{CV}^2$ to 0.01-100.
-At the bottom end of the range, estimating that a group behaves
-perfectly homogeneously ($CV=0$) is likely to be a sampling artifact;
-at the upper end, the estimate is also likely to be noisy because
-of the low mean value among married females (who have the largest
-range of CV). 
-We assume that the number of non-cohabiting partners follows a Gamma distribution.
-
-\subsection*{Simulation runs}
-
-One of the most difficult parts of model comparison is finding
-parameter sets that are commensurate with many different model
-structures. For the most part, our models are too complex to easily
-derive analytical correspondences among them. Given a numerical
-criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
-(intrinsic reproductive number), we can adjust one or more
-parameters by brute force to ensure that all of the models match
-according to that criterion. While $\rzero$ is often considered
-the most fundamental property of an epidemic, and might thus seem to
-be a natural matching criterion, here we focus on matching the initial
-growth rate $r$ for several reasons. First, our primary interest is in
-the transient evolutionary dynamics of virulence, which are more
-strongly affected by $r$ than $\rzero$. Second, $r$ is 
-more directly observable in real epidemics; $r$ can be estimated by
-fitting an exponential curve to the initial incidence or
-prevalence curves \cite{ma_estimating_2014}, while $\rzero$
-typically requires either (1) knowledge of \emph{all} epidemic
-parameters or (2) calculations based on
-$r$ and knowledge of the serial interval or generation interval of the
-disease \cite{wallinga_how_2007}. Thus, we scale parameters so that
-every run has the same initial exponential growth rate in 
-disease incidence.
-
-In order to allow for all models to have equal initial exponential
-growth rate, $r$, we need to pick a parameter, $s$, such that
-$\lim_{s\to 0} r(s) = 0$ and $\lim_{s\to\infty} r(s) = \infty$. As
-adjusting either partnership change rate (i.e. partnership formation
-and dissolution rate) or transmission rate fails this requirement for
-some of our models, we scaled partnership change rate and
-dissolution rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =
-\gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$,
-$\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$. Since transmission rate
-is adjusted by the scale of $\gamma$, uncoupled and extra-couple
-transmission rates are adjusted as well. For the instantaneous-switching
-and implicit models, none of which track single individuals, 
-only the transmission rate and partnership
-dissolution rate (in this case equivalent to the partnership change
-rate) are adjusted.
-
-We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
-of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$.
-After each run, initial exponential growth rate is calculated.
-Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
-For calibration purposes, we run each model for only 500 years
-(full simulations are run for 4000 years), which is always long
-enough to capture the exponential growth phase of the model. 
-We use a 4/5 order 
-Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
-\cite{soetaert_solving_2010}) for all simulations. 
-(For the heterogeneous model, approximately
-10\% of the samples failed due to numerical instability.)
-
-Although each disease strain's core characteristic is its SPVL, the
-SPVL has one-to-one correspondences (based on eq.~\ref{eq:hillfuns})
-with both the expected time to progression to AIDS and with the rate
-(probability per unit time) of HIV transmission. Because the time to
-progression (measured in years) is easier to interpret than
-SPVL (measured in \Lspvl\ units), we summarize the virulence
-trajectories for each model run in terms of time to progression
-rather than SPVL. Because the  time to progression is inversely
-related to SPVL (increasing SPVL decreases the time to progression),
-the time to progression is technically measuring inverse
-virulence rather than virulence (we did not think that 
-reporting virulence as the rate of progression to AIDS, in units
-of $\textrm{years}^{-1}$, would help interpretability).
-For each model we derive the following summary statistics:
-minimum expected time to progression;
-time at which this minimum occurs
-(corresponding to peak virulence --- this is also the time at which the
-maximum rate of progression, maximum SPVL, and maximum transmission rate
-occur); equilibrium time to progression; 
-and the ratio of progression time at its minimum to the equilibrium
-value. Equilibrium progression time is calculated after 4000 years of simulated
-time.
-Although most simulations reach equilibrium much earlier, we set our time horizon at a much later date as some simulation runs have slow rate of evolution depending on the parameter set and model assumptions.
-
-Knowing the minimum progression time, timing of the minimum progression time/peak virulence, and equilibrium progression time provide sufficient detail to identify the overall shape of the virulence trajectory.
-In particular, knowing the timing of the peak virulence (how many years
-into the epidemic the virulence peaks) can help epidemiologists
-guess whether the virulence of an emerging pathogen is likely (1)
-to have peaked early, possibly even before the pathogen is detected
-spreading in the population, and decline over the remaining course
-of the epidemic; (2) to increase, peak, and decline over the
-foreseeable future; or (3) to increase very slowly, peaking only
-in the far future. To the extent that our simplistic model for HIV
-reflects reality, we would take the peak time of 150-300 years 
-(\figurename~\ref{fig:panel3}c) to mean that, in the absence of
-treatment, the epidemic would probably still be increasing in virulence.
-
-% Results and Discussion can be combined.
-\section*{Results}
-
-Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
-However, our model produces lower peak virulence ($\approx 4.3$ vs.
-$\approx 4.6$ \Lspvl)
-and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
-
-Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases.
-Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
-Decreasing the initial virulence
-also leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).
-
-Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of the expected time of progression
-to AIDS (\figurename~\ref{fig:virtraj}: lower progression time corresponds
-to higher virulence).
-Although the speed of virulence evolution varies, leading to wide variation in the minimum expected progression time (means ranging from approximately 6 to 12 years), virulence peaks in all models between 200 and 300 years.
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig2.pdf}
-\caption{{\bf Envelopes of virulence trajectories (expected
-time of progression to AIDS) under all models.}
-All models were run until $t=4000~\textrm{years}$; truncated series are shown here.}
-\label{fig:virtraj}
-\end{figure}
-
-Our chosen summary statistics (peak time, minimum expected progression time, equilibrium expected
-progression time, and relative progression time) all vary considerably across models
-(\figurename~\ref{fig:unidist}).
-We first consider the models of intermediate realism: implicit,
-instantaneous-switching with and without extra-pair contact, and
-pair formation without extra-pair contact. Some parameter
-sets for these models lead to low equilibrium virulence ($\approx 18$ years
-to progression);
-these same sets lead to correspondingly low
-peak virulence (16 years to progression) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
-At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
-also produce late peaks ($> 200~\text{years}$) and
-high peak virulence (4 years to progression).
-The pair-formation without extra-pair contact and implicit models
-occasionally have parameter sets that select for such low virulence across
-the board that they never exceed their initial virulence, leading to a tail
-of peak times near zero.
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig3.pdf}
-\caption{{\bf Univariate distributions of summary statistics.}
-The distribution of equilibrium expected progression
-time (lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
-\label{fig:unidist}
+\includegraphics[width=\textwidth]{../figures/fig_S2_1.pdf}
+\caption{{\bf Envelopes of transmission trajectories under all models.}
+This figure matches Fig 2 in main text, but displays the
+envelope of population-mean transmission probabilities rather than \Lspvl\ over time
+for each model.
+}
+\label{fig:transtraj}
 \end{figure}
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig4.pdf}
-\caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
-Dashed line in equilibrium vs. peak virulence plot shows 1:1 line. 100 values were sampled from each model to allow for clearer distinction between the models}
-\label{fig:pairplot}
+\includegraphics[width=\textwidth]{../figures/fig_S2_2.pdf}
+\caption{{\bf Envelopes of progression trajectories under all models.}
+This figure matches Fig 2 in main text, but displays the
+envelope of population-mean expected time of progression to AIDS (i.e., length of
+intermediate HIV phase) rather than \Lspvl\ over time
+for each model.
+}
+\label{fig:durtraj}
 \end{figure}
 
-The most striking aspect of the univariate comparisons in
-\figurename~\ref{fig:unidist}, (and the bivariate comparisons in
-\figurename~\ref{fig:pairplot}) is the similarity between the results of the
-least (random mixing) and the most complex (pair formation with
-extra-pair contact, pairform+epc with heterogeneity) models.
-The random-mixing model has the lowest variability, because it is unaffected by uncertainty in pair formation and extra-pair contact parameters, but otherwise the virulence
-dynamics of these three extreme models are remarkably similar.
-This phenomenon is driven by the strong effects of extra-pair contact in the
-model with explicit pair formation and extra-pair contact 
-(``pairform+epc'' in \figurename{}s~\ref{fig:virtraj}-\ref{fig:plot_sens}). When individuals spend time uncoupled between
-partnerships, and when these single individuals can transmit disease
-to coupled individuals, the resulting unstructured mixing overwhelms
-the effect of structured mixing within couples, leading to mixing
-that is effectively close to random.
-Once unstructured mixing is strong, adding realistic heterogeneity
-of mixing to the model has little effect other than increasing
-the variability in the outcomes.
-
-These differences are practically as well as 
-scientifically important. The random-mixing, pairform+epc,
-and heterogeneous models all predict rapid
-progression to AIDS at the virulence peak
-(median/95\% CI = 6.1 (5.7-6.3), 
-6.02 (5.04-7.7), 6.03  (4.8-9.2)). 
-In contrast, 
-the implicit model predicts minimum progression times about twice as long:
-12.5 (9.6-15.6) years. The corresponding differences in 
-within-couple transmission
-probability are even more extreme, about a fourfold difference:
-0.249 (0.24-0.26), 0.252 (0.19-0.28), and 0.252 (0.15-0.28) per year for the 
-random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model (see Appendix S2 for 
-plots showing univariate summaries
-of \Lspvl\ and transmission probability).
-
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig5.pdf}
-\caption{{\bf Sensitivity plot.}
-For most parameters in the Latin hypercube sample and each summary statistic, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
-prior to adjusting the parameters to achieve the standard initial epidemic growth rate.}
-\label{fig:plot_sens}
+  \includegraphics[width=\textwidth]{../figures/fig_S2_3.pdf}
+\caption{{\bf Univariate distributions of transmission probabilities and progression.}
+This figure matches Fig 2 in main text, but displays the
+univariate distributions for the transmission probability and 
+progression time at the virulence
+peak and at the epidemiological equilibrium,
+rather than the distributions of \Lspvl.
+}
+\label{fig:tranprogsum}
 \end{figure}
 
-The bivariate relationships (\figurename~\ref{fig:pairplot}) help distinguish the results of 
-different models with similar univariate dynamical summaries. While the
-relationship between equilibrium progression time and peak time is
-similar for all model structures (top left panel), the other
-relationships show more variation. In particular, the implicit
-and pair-formation (without extra-pair contact) are very similar
-to each other, but distinct from the other models. We still do
-not have a convincing explanation for this distinction; we
-would have expected the implicit model to be most similar to the
-the instantaneous-switching model without extra-pair contact,
-which most closely matches its derivation. However, we note
-that the implicit model derivation is based on defining
-the force of infection to match a scaled version of $\rzero$,
-and as such would be expected to match the equilibrium behaviour
-but not necessarily the epidemic-phase behaviour of a model
-with explicit partnership dynamics.
-
-Finally, the sensitivity plot (\figurename~\ref{fig:plot_sens}) shows the effects 
-of each parameter on the summary statistics. In almost every case the
-effects of the parameters are monotonic; note that
-the plot shows the effects of the \emph{unscaled} parameters, i.e.
-before they have been calibrated to achieve a standard initial epidemic
-growth rate.
-Increases in the transmission rates ($\beta_P$, $\beta_D$)
-and durations ($D_P$, $D_D$) in the primary and disease stages generally
-decrease the equilibrium virulence, peak virulence, and peak time,
-although the random and pair-formation+epc 
-models have high, relatively
-constant values with respect to these parameters
-(because the patterns are so similar across this set of parameters,
-\figurename~\ref{fig:plot_sens} shows only $D_P$).
-
-The partnership dissolution rate ($c$), which essentially
-acts as a contact rate in the model,
-increases virulence and peak time in almost all
-cases, although the pair-formation+epc
-model is again relatively insensitive.
-The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
-virulence in the instantaneous-switching+epc model, but not the pair-formation+epc
-model (probably because the uncoupled individuals present in the pair-formation+epc
-model make extra-pair contact by coupled individuals less important).
-Surprisingly, once calibration
-is taken into account, the remaining parameters have little effect overall.
-The rate of partnership formation ($\rho$) 
-has little impact on the models with finite pair-formation times.
-The relative rate of uncoupled contact ($c_u/c_w$) slightly decreases the
-minimum and equilibrium progression time and delays the peak time in the
-pair-formation+epc model, but neither the uncoupled contact rate nor
-the mean ($\mu$) or CV$^2$ of the number of non-cohabiting sexual
-partners has much systematic effect in the heterogeneous model.
-
-\section*{Discussion}
-
-All models must simplify the world.  Many constraints --- among them data
-availability, computation time, and code complexity --- drive the need
-for parsimony, with different constraints applying in different
-contexts. The critical question that modelers must ask is whether the
-simplified model gives adequate answers, or whether the
-simplifications lead to qualitative or quantitative errors.
-This question is especially important for modelers who
-are hoping that their conclusions will guide management decisions.
-
-In the particular example of HIV virulence eco-evolutionary dynamics
-and the complexity of epidemiological structures
-we reach the slightly ironic conclusion that the
-effort put into building a more realistic model essentially cancels
-out, putting us back where we started when used a naive random-mixing
-contact model.
-However, we are not quite back where we started, as the
-complex models lead to wider, presumably more realistic
-confidence intervals on the predictions.
-In general, unstructured mixing --- whether occurring through 
-purely random mixing, or through extra-pair contact and contact
-among people outside of stable partnerships --- tends to drive
-faster virulence evolution, leading to higher peak virulence and 
-lower times to progression at the peak time.
-
-%In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
-%actually more true to real-world HIV epidemiological dynamics than a
-%model with realistic partnership durations that neglects extra-pair
-%contact \cite{herbeck2016evolution}. 
-Taking further steps to make the model even more realistic
-might add further structure,
-making the random-mixing model predictions less accurate. For
-example, our model forms partnerships randomly, and assumes that
-extra-pair contact is randomly mixing across the population;
-one could instead model extra-pair contact as arising from
-multiple concurrent partnerships (some, such as contact with sex
-workers, of very short duration) and/or more structured partnership
-formation (by age, ethnicity, or behaviour group). The effects of
-other realistic complications such as explicit modeling of two
-sexes (both in contact structure and differential transmission
-probabilities), temporal and spatial variation in epidemic processes,
-or presence of genetic variation in hosts are harder to predict.
-
-Parameterization is one of the biggest challenges of epidemiological
-modeling. In addition to following Champredon \etal\ \cite{champredon_hiv_2013} 
-by doing Latin hypercube
-sampling across a wide range of epidemiological parameters, we 
-calibrated each set of parameters to the same initial epidemic
-growth rate, chosen to match the results of previous models
-\cite{shirreff_transmission_2011}.  Previous models 
-in this area have drawn their
-parameters from cohort studies from the 1990s
-\cite{wawer2005rates,hollingsworth_hiv1_2008}
-rather than doing any explicit calibration to epidemic curves,
-but they give reasonable order-of-magnitude
-growth rates ($\approx 0.04~\textrm{year}^{-1}$)
-for the early stages of the HIV epidemic (although considerably
-lower than estimates of $\approx 0.07-0.1~\textrm{year}^{-1}$
-based on population genetic reconstructions \cite{faria_early_2014}).
-However, our reason for calibrating was not to match any
-specific observed epidemic, but rather to make sure that
-we were making meaningful comparisons across a range of
-models with radically different epidemiological structures, and
-hence involving different interpretations of the same quantitative
-parameters.  For example, in models with instantaneous switching the
-partnership dissolution rate $c$ is identical to the partnership
-formation rate; in models with explicit partnership formation,
-the partnership formation rate is also $c$ at equilibrium,
-but might vary over the course of an epidemic.
-It is not obvious whether models with equal parameters but
-different structures should be directly compared; calibration
-solves this problem.
-
-More generally, any model that wants to be
-taken seriously for management and forecasting purposes should
-be calibrated to \emph{all} available data, using informative
-priors to incorporate both realistic distributions of uncertainty
-for all parameters from independent measurements \cite{elderd_uncertainty_2006}
-and calibration from population-level observations of epidemic
-trajectories. Such a procedure would also be an improvement on the common --- although not universal --- %
-practice, which we have followed here,
-of assessing uncertainty over uniform ranges rather than
-using distributions that allow more continuous variation in support over
-the range of a parameter.
-
-Researchers have documented that HIV virulence and set-point viral
-load are changing, on time scales comparable to those portrayed here
-(e.g., compare \figurename~\ref{fig:virtraj} to Herbeck \etal's
-estimated rate of change of 1.3 \Lspvl\ per century [95\% CI -0.1 to
-  3] \cite{herbeck_is_2012}), and have begun to build relatively realistic models that
-attempt to describe how interventions such as mass antiretroviral
-therapy (ART) can be expected to change the trajectory of virulence
-evolution \cite{payne_impact_2014,roberts2015impact,herbeck2016evolution}.  While these
-efforts are well-intentioned, we caution that epidemiological and
-other structural details that are currently omitted from these models
-could significantly change their conclusions.
-
-\section*{Acknowledgements}
-We would like to thank Christophe Fraser and
-David Champredon for access to simulation code;
-this work was funded by NSERC Discovery Grant 386590-2010.
-
-\section*{Supporting Information}
-
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{Appendix S1: model details}
-\label{S1_Appendix}
-
-\paragraph*{Appendix S2: dynamics of transmission and virulence}
-
-\label{S2_Appendix}
-
-\clearpage
-
-\nolinenumbers
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here.
-%
-
-\bibliography{virulence}
 \end{document}

### 2016-10-26 ms/plos_supp2.tex
updated supporting figures

@@
 This figure matches Fig 2 in main text, but displays the
-envelope of population-mean transmission probabilities rather than \Lspvl\ over time
+envelope of population-mean transmission probabilities rather than expected progression time to AIDS over time
 for each model.
@@
 
+\clearpage
+
 \begin{figure}[!ht]
@@
 This figure matches Fig 2 in main text, but displays the
-envelope of population-mean expected time of progression to AIDS (i.e., length of
-intermediate HIV phase) rather than \Lspvl\ over time
+envelope of \Lspvl\ rather than expected progression time to AIDS over time
 for each model.
 }
-\label{fig:durtraj}
+\label{fig:virtraj}
 \end{figure}
+
+\clearpage
 
@@
   \includegraphics[width=\textwidth]{../figures/fig_S2_3.pdf}
-\caption{{\bf Univariate distributions of transmission probabilities and progression.}
-This figure matches Fig 2 in main text, but displays the
+\caption{{\bf Univariate distributions of transmission probabilities and \Lspvl\.}
+This figure matches Fig 3 in main text, but displays the
 univariate distributions for the transmission probability and 
-progression time at the virulence
+\Lspvl\ time at the virulence
 peak and at the epidemiological equilibrium,
-rather than the distributions of \Lspvl.
+rather than the distributions of progression time to AIDS.
 }
-\label{fig:tranprogsum}
+\label{fig:tranvirsum}
 \end{figure}

### 2016-10-26 figures/figures.R

@@ -197,16 +197,15 @@ gg_betatraj <- direct.label(gg_basetraj %+% ss_beta + nostrips +
 ## scale_x_continuous(limits=c(0,2500))
 ## r fig2.1, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400} 
 
-ggsave(gg_betatraj,file="fig_S2_1.pdf",width=6,height=4)
-ggsave(gg_betatraj,file="fig_S2_1.png",width=6,height=4,dpi=400)
-
+ggsave(gg_betatraj,file="fig_S2_1.pdf",width=8,height=6)
+ggsave(gg_betatraj,file="fig_S2_1.png",width=8,height=6,dpi=400)
 
 ## scale_x_continuous(limits=c(0,2500))
 
 ## r fig2.2, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400} 
 
-ggsave(gg_durtraj,file="fig_S2_2.pdf",width=6,height=4)
-ggsave(gg_durtraj,file="fig_S2_2.png",width=6,height=4,dpi=400)
+ggsave(gg_virtraj,file="fig_S2_2.pdf",width=8,height=6)
+ggsave(gg_virtraj,file="fig_S2_2.png",width=8,height=6,dpi=400)
 
 ### Figure 3
 
@@ -277,11 +276,8 @@ fig_objects <- c(fig_objects,"gg_univ")
 
 sL$model <- factor(sL$model, m_order)
 sL.epi <- transform(sL, eq_t = returnBeta(eq_vir,HIVpars.skeleton),
-                    peak_t = returnBeta(peak_vir,HIVpars.skeleton),
-                    eq_dur = returnDur(eq_vir,HIVpars.skeleton),
-                    peak_dur = returnDur(peak_vir,HIVpars.skeleton))
-
-sL.epi <- sL.epi[,-c(3:6)]
+                    peak_t = returnBeta(peak_vir,HIVpars.skeleton)) %>%
+	select(model, run, eq_vir, peak_vir, eq_t, peak_t)
 
 maxtrans <- sL.epi %>%
 	group_by(model) %>%
@@ -294,28 +290,32 @@ mL.epi <- melt(sL.epi,id.vars=c("model","run"))
 ## w <- which(mL$model=="random")
 ## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
[…]

### 2016-10-26 ms/plos_latex.tex
update bib

@@
 
-\bibliography{virulence}
+\begin{thebibliography}{10}
+
+\bibitem{alizon_virulence_2009}
+Alizon S, Hurford A, Mideo N, van Baalen M.
+\newblock Virulence evolution and the trade-off hypothesis: history, current
+  state of affairs and the future.
+\newblock J Evol Biol. 2009;22:245--259.
+\newblock doi:{10.1111/j.1420-9101.2008.01658.x}.
+
+\bibitem{EbertBull2003}
+Ebert D, Bull JJ.
+\newblock Challenging the trade-off model for the evolution of virulence: is
+  virulence management feasible?
+\newblock Trends Microbiol. 2003;11(1):15--20.
+
+\bibitem{alizon_adaptive_2015}
+Alizon S, Michalakis Y.
+\newblock Adaptive virulence evolution: the good old fitness-based approach.
+\newblock Trends in Ecology \& Evolution. 2015;30(5):248--254.
+\newblock doi:{10.1016/j.tree.2015.02.009}.
+
+\bibitem{Dwyer+1990}
+Dwyer G, Levin SA, Buttel L.
+\newblock A simulation model of the population dynamics and evolution of
+  myxomatosis.
+\newblock Ecol Monog. 1990;60:423--447.
+
+\bibitem{mackinnon1999genetic}
+Mackinnon MJ, Read AF.
+\newblock Genetic relationships between parasite virulence and transmission in
+  the rodent malaria {{\em Plasmodium chabaudi}}.
+\newblock Evolution. 1999; p. 689--703.
+
+\bibitem{jensen2006empirical}
+Jensen KH, Little T, Skorping A, Ebert D.
+\newblock Empirical support for optimal virulence in a castrating parasite.
+\newblock PLoS Biol. 2006;4(7):e197.
+
+\bibitem{deroode2008virulence}
+De~Roode JC, Yates AJ, Altizer S.
+\newblock Virulence-transmission trade-offs and population divergence in
+  virulence in a naturally occurring butterfly parasite.
+\newblock Proceedings of the National Academy of Sciences.
+  2008;105(21):7489--7494.
+
+\bibitem{Fraser+2007}
+Fraser C, Hollingsworth TD, Chapman R, de~Wolf F, Hanage WP.
+\newblock Variation in {HIV}-1 set-point viral load: Epidemiological analysis
+  and an evolutionary hypothesis.
+\newblock PNAS. 2007;104:17441--17446.
+
+\bibitem{fraser_virulence_2014}
+Fraser C, Lythgoe K, Leventhal GE, Shirreff G, Hollingsworth TD, Alizon S,
+  et~al.
+\newblock Virulence and Pathogenesis of {HIV}-1 Infection: An Evolutionary
+  Perspective.
+\newblock Science. 2014;343(6177):1243727.
+\newblock doi:{10.1126/science.1243727}.
+
+\bibitem{shirreff_transmission_2011}
+Shirreff G, Pellis L, Laeyendecker O, Fraser C.
+\newblock Transmission Selects for {HIV-1} Strains of Intermediate Virulence: A
+  Modelling Approach.
+\newblock PLoS Computational Biology. 2011;7(10):e1002185.
+\newblock doi:{10.1371/journal.pcbi.1002185}.
+
+\bibitem{herbeck_hiv_2014}
+Herbeck JT, Mittler JE, Gottlieb GS, Mullins JI.
+\newblock An {HIV} Epidemic Model Based on Viral Load Dynamics: Value in
+  Assessing Empirical Trends in {HIV} Virulence and Community Viral Load.
+\newblock {PLoS} Comput Biol. 2014;10(6):e1003673.
+
+\bibitem{herbeck_evolution_2016}
+Herbeck JT, Mittler JE, Gottlieb GS, Goodreau SM, Murphy JT, Cori A, et~al.
+\newblock Evolution of {HIV} virulence in response to widespread scale up of
+  antiretroviral therapy: a modeling study.
+\newblock Virus Evolution. 2016;2(2):vew028.
+\newblock doi:{10.1093/ve/vew028}.
+
+\bibitem{day_virulence_2004}
+Day T, Proulx SR.
+\newblock A General Theory for the Evolutionary Dynamics of Virulence.
+\newblock The American Naturalist. 2004;163(4):E40--E63.
+\newblock doi:{10.1086/382548}.
+
+\bibitem{alizon_price_2009}
+Alizon S.
+\newblock The {Price} equation framework to study disease within-host
+  evolution.
+\newblock Journal of Evolutionary Biology. 2009;22(5):1123--1132.
+\newblock doi:{10.1111/j.1420-9101.2009.01726.x}.
+
+\bibitem{champredon_hiv_2013}
+Champredon D, Bellan S, Dushoff J.
+\newblock {HIV} Sexual Transmission Is Predominantly Driven by Single
+  Individuals Rather than Discordant Couples: A Model-Based Approach.
+\newblock PLoS ONE. 2013;8(12):e82906.
+\newblock doi:{10.1371/journal.pone.0082906}.
+
+\bibitem{dietz_epidemiological_1988}
+Dietz K, Hadeler KP.
+\newblock Epidemiological models for sexually transmitted diseases.
+\newblock Journal of Mathematical Biology. 1988;26(1):1--25.
+
+\bibitem{kretzschmar_effect_1998}
+Kretzschmar M, Dietz K.
+\newblock The effect of pair formation and variable infectivity on the spread
+  of an infection without recovery.
+\newblock Mathematical Biosciences. 1998;148(1):83--113.
+
+\bibitem{hollingsworth_hiv1_2008}
+Hollingsworth TD, Anderson RM, Fraser C.
+\newblock {HIV}-1 Transmission, by Stage of Infection.
+\newblock Journal of Infectious Diseases. 2008;198(5):687--693.
+\newblock doi:{10.1086/590501}.
+
+\bibitem{omori2015dynamics}
+Omori R, Chemaitelly H, Abu-Raddad LJ.
+\newblock Dynamics of non-cohabiting sex partnering in sub-{Saharan} {Africa}:
+  a modelling study with implications for {HIV} transmission.
+\newblock Sexually Transmitted Infections.
+  2015;doi:{10.1136/sextrans-2014-051925}.
+
+\bibitem{may_transmission_1988}
+May RM, Anderson RM.
+\newblock The Transmission Dynamics of Human Immunodeficiency Virus (HIV) [and
+  Discussion].
+\newblock Philosophical Transactions of the Royal Society of London Series B,
+  Biological Sciences. 1988;321(1207):565--607.
+
+\bibitem{delva_connecting_2016}
+Delva W, Leventhal GE, Helleringer S.
+\newblock Connecting the dots: network data and models in {HIV} epidemiology.
+\newblock AIDS. 2016;30(13):2009--2020.
+\newblock doi:{10.1097/QAD.0000000000001184}.
+
+\bibitem{blower_drugs_1991}
+Blower SM, Hartel D, Dowlatabadi H, Anderson RM, May RM.
+\newblock Drugs, Sex and {HIV}: A Mathematical Model for {New York City}.
+\newblock Philosophical Transactions of the Royal Society of London B:
+  Biological Sciences. 1991;331(1260):171--187.
+\newblock doi:{10.1098/rstb.1991.0006}.
+
+\bibitem{ma_estimating_2014}
+Ma J, Dushoff J, Bolker BM, Earn DJD.
+\newblock Estimating Initial Epidemic Growth Rates.
+\newblock Bulletin of Mathematical Biology. 2014;76(1):245--260.
+\newblock doi:{10.1007/s11538-013-9918-2}.
+
+\bibitem{wallinga_how_2007}
+Wallinga J, Lipsitch M.
+\newblock How generation intervals shape the relationship between growth rates
+  and reproductive numbers.
+\newblock Proceedings of the Royal Society of London B: Biological Sciences.
+  2007;274(1609):599--604.
+\newblock doi:{10.1098/rspb.2006.3754}.
+
+\bibitem{soetaert_solving_2010}
+Soetaert K, Petzoldt T, Setzer RW.
+\newblock Solving Differential Equations in {R}: Package {deSolve}.
+\newblock Journal of Statistical Software. 2010;33(9):1--25.
+\newblock doi:{10.18637/jss.v033.i09}.
+
+\bibitem{wawer2005rates}
+Wawer MJ, Gray RH, Sewankambo NK, Serwadda D, Li X, Laeyendecker O, et~al.
+\newblock Rates of {HIV}-1 transmission per coital act, by stage of {HIV}-1
+  infection, in {Rakai}, {Uganda}.
+\newblock Journal of Infectious Diseases. 2005;191(9):1403--1409.
+
+\bibitem{faria_early_2014}
+Faria NR, Rambaut A, Suchard MA, Baele G, Bedford T, Ward MJ, et~al.
+\newblock The early spread and epidemic ignition of {HIV}-1 in human
+  populations.
+\newblock Science (New York, NY). 2014;346(6205):56--61.
+\newblock doi:{10.1126/science.1256739}.
+
+\bibitem{elderd_uncertainty_2006}
+Elderd BD, Dukic VM, Dwyer G.
+\newblock Uncertainty in predictions of disease spread and public health
+  responses to bioterrorism and emerging diseases.
+\newblock Proceedings of the National Academy of Sciences. 2006;103(42):15693
+  --15697.
+\newblock doi:{10.1073/pnas.0600816103}.
+
+\bibitem{herbeck_is_2012}
+Herbeck JT, Müller V, Maust BS, Ledergerber B, Torti C, Di~Giambenedetto S,
+  et~al.
+\newblock Is the virulence of {HIV} changing? {A} meta-analysis of trends in
+  prognostic markers of {HIV} disease progression and transmission.
+\newblock AIDS (London, England). 2012;26(2):193--205.
+\newblock doi:{10.1097/QAD.0b013e32834db418}.
+
+\bibitem{payne_impact_2014}
+Payne R, Muenchhoff M, Mann J, Roberts HE, Matthews P, Adland E, et~al.
+\newblock Impact of {HLA}-driven {HIV} adaptation on virulence in populations
+  of high {HIV} seroprevalence.
+\newblock Proceedings of the National Academy of Sciences.
+  2014;111(50):E5393--E5400.
+\newblock doi:{10.1073/pnas.1413339111}.
+
+\bibitem{roberts2015impact}
+Roberts HE, Goulder PJ, McLean AR.
+\newblock The impact of antiretroviral therapy on population-level virulence
+  evolution of {HIV}-1.
+\newblock Journal of The Royal Society Interface. 2015;12(113):20150888.
+
+\bibitem{herbeck2016evolution}
+Herbeck J, Mittler J, Gottlieb G, Goodreau S, Murphy J, Cori A, et~al.
+\newblock Evolution of {HIV} virulence in response to widespread scale up of
+  antiretroviral therapy: a modeling study.
+\newblock bioRxiv. 2016; p. 039560.
+
+\end{thebibliography}
 \end{document}

### 2016-10-26 ms/S1_appendix.tex
tweaks

--- 2016-10-26 ms/plos_supp1.tex
@@
 
-\bibliography{virulence}
+\begin{thebibliography}{1}
+
+\bibitem{champredon_hiv_2013}
+Champredon D, Bellan S, Dushoff J.
+\newblock {HIV} Sexual Transmission Is Predominantly Driven by Single
+  Individuals Rather than Discordant Couples: A Model-Based Approach.
+\newblock PLoS ONE. 2013;8(12):e82906.
+\newblock doi:{10.1371/journal.pone.0082906}.
+
+\bibitem{shirreff_transmission_2011}
+Shirreff G, Pellis L, Laeyendecker O, Fraser C.
+\newblock Transmission Selects for {HIV-1} Strains of Intermediate Virulence: A
+  Modelling Approach.
+\newblock PLoS Computational Biology. 2011;7(10):e1002185.
+\newblock doi:{10.1371/journal.pcbi.1002185}.
+
+\bibitem{hollingsworth_hiv1_2008}
+Hollingsworth TD, Anderson RM, Fraser C.
+\newblock {HIV}-1 Transmission, by Stage of Infection.
+\newblock Journal of Infectious Diseases. 2008;198(5):687--693.
+\newblock doi:{10.1086/590501}.
+
+\bibitem{may_transmission_1988}
+May RM, Anderson RM.
+\newblock The Transmission Dynamics of Human Immunodeficiency Virus (HIV) [and
+  Discussion].
+\newblock Philosophical Transactions of the Royal Society of London Series B,
+  Biological Sciences. 1988;321(1207):565--607.
+
+\bibitem{herbeck_evolution_2016}
+Herbeck JT, Mittler JE, Gottlieb GS, Goodreau SM, Murphy JT, Cori A, et~al.
+\newblock Evolution of {HIV} virulence in response to widespread scale up of
+  antiretroviral therapy: a modeling study.
+\newblock Virus Evolution. 2016;2(2):vew028.
+\newblock doi:{10.1093/ve/vew028}.
+
+\end{thebibliography}
 \end{document}

### 2016-10-26 ms/S2_appendix.tex
tweaks

--- 2016-10-26 ms/plos_supp2.tex
@@
 
+\renewcommand\thefigure{S\arabic{figure}}
+
 %% END MACROS SECTION
@@
 \includegraphics[width=\textwidth]{../figures/fig_S2_2.pdf}
-\caption{{\bf Envelopes of progression trajectories under all models.}
+\caption{{\bf Envelopes of SPVL trajectories under all models.}
 This figure matches Fig 2 in main text, but displays the
@@
 
+%% \Lspvl\ doesn't get bolded so I got rid of log10
 \begin{figure}[!ht]
   \includegraphics[width=\textwidth]{../figures/fig_S2_3.pdf}
-\caption{{\bf Univariate distributions of transmission probabilities and \Lspvl\.}
+\caption{{\bf Univariate distributions of transmission probabilities and SPVL.}
 This figure matches Fig 3 in main text, but displays the
-univariate distributions for the transmission probability and 
-\Lspvl\ time at the virulence
+univariate distributions for the transmission probability and \Lspvl\ time at the virulence
 peak and at the epidemiological equilibrium,

### 2016-10-27 figures/figures.R
don't auto-make PNGs

@@ -15,6 +15,8 @@ library(GGally) ## need BMB version
 ## devtools::install_github("lionel-/ggstance")
 library(ggstance)
 
+do_png <- FALSE
+
 opts_chunk$set(echo=FALSE,error=FALSE)
 if (.Platform$OS.type=="windows") {
     windowsFonts(Times=windowsFont("Times"))
@@ -124,9 +126,11 @@ pdf(file="fig1.pdf",height=3.5,width=10)
 print(grid.arrange(g1.y, g2.y, g3.y, nrow=1))
 dev.off()
 
-png(file="fig1.png",height=3.5*600,width=10*600)
-print(grid.arrange(g1.y, g2.y, g3.y, nrow=1))
-dev.off()
+if (do_png) {
+    png(file="fig1.png",height=3.5*600,width=10*600)
+    print(grid.arrange(g1.y, g2.y, g3.y, nrow=1))
+    dev.off()
+}
 
 fig_objects <- c("g1.y","g2.y","g3.y")
 
@@ -180,7 +184,7 @@ gg_durtraj <- direct.label(gg_basetraj %+% ss_dur + nostrips +
 
 ## scale_x_continuous(limits=c(0,2500))
 ggsave(gg_durtraj,file="fig2.pdf",width=6,height=4)
-ggsave(gg_durtraj,file="fig2.png",width=6,height=4,dpi=400)
+if (do_png) ggsave(gg_durtraj,file="fig2.png",width=6,height=4,dpi=400)
 
 fig_objects <- c(fig_objects,"gg_durtraj","bottom.points2","top.points2")
 
@@ -198,14 +202,14 @@ gg_betatraj <- direct.label(gg_basetraj %+% ss_beta + nostrips +
 ## r fig2.1, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400} 
 
 ggsave(gg_betatraj,file="fig_S2_1.pdf",width=8,height=6)
-ggsave(gg_betatraj,file="fig_S2_1.png",width=8,height=6,dpi=400)
+if (do_png) ggsave(gg_betatraj,file="fig_S2_1.png",width=8,height=6,dpi=400)
 
 ## scale_x_continuous(limits=c(0,2500))
 
 ## r fig2.2, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400} 
 
[…]

### 2016-10-27 figures/figures.R
update units

@@ -302,8 +302,8 @@ mLw.epi$variable <- factor(mLw.epi$variable,
                            levels = c("eq_vir", "peak_vir", "eq_t", "peak_t"),
                            labels = c("Equilibrium~mean~log[10]~SPVL",
 													 					 "Peak~mean~log[10]~SPVL",
-													 					 "Equilibrium~mean~transmission~rate",
-													 					 "Peak~mean~transmission~rate"))
+													 					 "Equilibrium~mean~transmission~rate~(years^{-1})",
+													 					 "Peak~mean~transmission~rate~(years^{-1})"))
 
 gg_univ.epi <- ggplot(mLw.epi,aes(value,model,fill=model))+
 	geom_violinh(width=1)+
@@ -312,7 +312,7 @@ gg_univ.epi <- ggplot(mLw.epi,aes(value,model,fill=model))+
 	guides(fill=guide_legend(reverse=TRUE))+
 	labs(y="")+
 	geom_point(data=data.frame(model="random",
-                   variable=c("Equilibrium~mean~transmission~rate",
+                   variable=c("Equilibrium~mean~transmission~rate~\\(years^{-1}\\)",
 																											 "Equilibrium~mean~log[10]~SPVL"),
                                value=c(rval.epi.t, rval.epi.v)),pch=22,size=3,show.legend=FALSE) +
 	zero_x_margin

### 2016-10-27 figures/figures.R
forgot to commit this

@@ -302,8 +302,8 @@ mLw.epi$variable <- factor(mLw.epi$variable,
                            levels = c("eq_vir", "peak_vir", "eq_t", "peak_t"),
                            labels = c("Equilibrium~mean~log[10]~SPVL",
 													 					 "Peak~mean~log[10]~SPVL",
-													 					 "Equilibrium~mean~transmission~rate~(years^{-1})",
-													 					 "Peak~mean~transmission~rate~(years^{-1})"))
+													 					 "Equilibrium~mean~transmission~rate~(year^-1)",
+													 					 "Peak~mean~transmission~rate~(year^-1)"))
 
 gg_univ.epi <- ggplot(mLw.epi,aes(value,model,fill=model))+
 	geom_violinh(width=1)+
@@ -312,7 +312,7 @@ gg_univ.epi <- ggplot(mLw.epi,aes(value,model,fill=model))+
 	guides(fill=guide_legend(reverse=TRUE))+
 	labs(y="")+
 	geom_point(data=data.frame(model="random",
-                   variable=c("Equilibrium~mean~transmission~rate~\\(years^{-1}\\)",
+                   variable=c("Equilibrium~mean~transmission~rate~(year^-1)",
 																											 "Equilibrium~mean~log[10]~SPVL"),
                                value=c(rval.epi.t, rval.epi.v)),pch=22,size=3,show.legend=FALSE) +
 	zero_x_margin

### 2016-10-27 ms/plos_latex.tex
tweaks
more tweaks

@@
 therapy (ART) can be expected to change the trajectory of virulence
-evolution \cite{payne_impact_2014,roberts2015impact,herbeck2016evolution}.  While these
+evolution \cite{payne_impact_2014,roberts2015impact,herbeck_evolution_2016}.  While these
 efforts are well-intentioned, we caution that epidemiological and
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{Appendix S1: model details}
+\paragraph*{Appendix S1: Model Details}
 \label{S1_Appendix}
 
-\paragraph*{Appendix S2: dynamics of transmission and virulence}
+\paragraph*{Appendix S2: Dynamics of Transmission and Virulence}
 
@@
 
-\bibitem{herbeck2016evolution}
-Herbeck J, Mittler J, Gottlieb G, Goodreau S, Murphy J, Cori A, et~al.
-\newblock Evolution of {HIV} virulence in response to widespread scale up of
-  antiretroviral therapy: a modeling study.
-\newblock bioRxiv. 2016; p. 039560.
-
 \end{thebibliography}

### 2016-10-28 ms/cover_letter/plosHIVcover.tex
cleaning up

--- 2016-10-19 ms/plosHIVcover_resubm.tex
@@
 
-\newcommand{\revcomment}[1]{\emph{#1}}
-\newcommand{\response}[1]{#1}
-
 \begin{document}
@@
 
-\revcomment{The abstract really needs to be improved. It’s missing the main points
-of the paper (that different contact patterns change the predicted
-trajectories and optimums of virulence), and has unnecessary minor
-details (like the fact that parameters were sampled using a Latin
-hypercube).}
+We (Daniel Park and Ben Bolker) are submitting our manuscript, ``Effects of Epidemiological Structure on the Transient Evolution of HIV Virulence'', for consideration for publication in \emph{PLoS Computational Biology}.
+Using HIV as a model system, our study explores the transient, eco-evolutionary changes in pathogen characteristics over the course of an epidemic.
+Building on modeling frameworks and estimated parameters developed by Shirreff \emph{et al.} \cite{shirreff_transmission_2011} and Champredon \emph{et al.} \cite{champredon_hiv_2013}, we ask how predicted changes in virulence depend on way that contact patterns (e.g.
+the dynamics of sexual partnership change and contact outside of stable partnerships) are modeled.
 
-\response{We've followed the reviewer's suggestions in adding
-important points and removing unimportant details from the abstract}
+This study is important because researchers are actively developing --- and publishing in high-impact journals --- studies that predict the evolutionary consequences of public-health interventions \cite{payne_impact_2014,roberts2015impact,herbeck2016evolution}.
+Our work suggests that basic epidemiological structures that are neglected in these models can lead to two- to fourfold differences in predictions of outcomes such as transmission probability or expected time to progress to AIDS.
 
-\revcomment{The introduction is really lacking a cogent description of the study of
-the evolution of virulence, and the motivation of the paper would be
-totally unclear to someone not already very familiar with this field.
-The main points that are needed to spell out this issue are:
-
-[... detailed suggestions omitted]
-}
-
-\response{We thank the reviewer for pointing this out --- we've 
-been over this ground so often that we've come to take this basic
-framework for granted!  We've 
-}
-The paper claims to add considerations of population structure to the
-study of virulence but only focuses on the rules of partnership
-formation, while assuming all individuals have identical behavior
-patterns, and not on another major aspect of population structure, which
-is the heterogeneity in the number of partners (or rates of partnership
-change) among individuals. In the case of a static network, this would
-be considered the degree heterogeneity. Many studies have shown that
-sexual contact networks tend to me more on the scale-free spectrum, and
-so the assumption of uniformity of behavior seems weak. This study would
-be most impactful if on top of examining these switch patterns it
-examined the influence of contact heterogeneity.
-
-In Figure 3 it would be helpful to also include time to progression to
-AIDS as a figure panel
-
-Reviewer #2: Modeling studies have shown that HIV could evolve towards
-immediate virulence where the trade-off between host survival and
-transmission is optimized. The aim of this study is to investigate how
-different assumptions about sexual partnership dynamics affect the
-transient evolution of HIV virulence in a population. While this is a
-clear objective, the study is rather technical and, in my opinion, does
-not provide major insights or conclusions. My major concerns regarding
-the approach of the study is given below. Furthermore, there are several
-issues in terms of structure and the model description that should be
-considered by the authors.
-
-Approach:
-The authors argue that they compare more “realistic” models (based on
-ref. 14) to the relatively simple model by Shireff et al. (ref. 10). In
-my view, all these models are quite simplistic in that they do not take
-into account age-structure and heterogeneity in sexual behavior.
-Essentially, all these models are random mixing models where every
-individual can make contact with every other individual. The only
-difference to the random mixing model by Shireff et al. is that
-transmission can occur in ongoing partnerships and that some models
-allow for concurrency. Arguably much more important aspects that affect
-HIV epidemics and virulence evolution are the above-mentioned age- and
-risk-structure. Furthermore, the authors use a single-stage disease
-model and do not distinguish between differences in transmissibility
-between the acute (primary) and chronic (asymptomatic) stage of
-infections, which might influence HIV virulence evolution more than
-differences in the sexual partnership dynamics. Hence, I am not
-convinced that the authors use a “more realistic” model than the one by
-vShireff et al.
-
-Structure:
-- The references to the equations are not always correct.
-- Figure 1 is described in Materials and Methods before the model is
-fully explained. Consequently, the parameter r cannot be understood.
-- Tables and Figures often lack dimensions, i.e., is is unclear what the
-axes represent. Table 1 is more or less useless without providing
-dimensions for the various parameters.
-- The author summary seems to provide more information about the results
-than the abstract. However, the authors use SPVL and progression to AIDS
-as two different proxies for virulence which can be confusing.
-
-Model:
-The authors make use of a previously published model from ref. 14. This
-model is based on the pair-formation formalism initially developed by
-Dietz & Hadeler (1988, J Math Biol, PMID: 3351391) and later used by
-Kretzschmar et al. (1994 & 1998, Math Biosci, PMID: 7833594 & 9597826).
-I feel that the authors should refer to some of these original
-publications as well. Unfortunately, the description of the models in
-appendix S1 seems to be partially erroneous or incorrect. For example,
-the primes for the derivatives are often missing (e.g., equation 22, 24
-and 25) or it is unclear whether equations represent state variables or
-derivates thereof. Also, the reasoning behind the “instswitch” models is
-unclear to me. The authors say that “Once individuals leave a
-partnership, they enter temporary compartments…”. The variables (or
-derivatives?) X and Y_i in equation 19 consist of positive terms only,
-so how do individuals leave these compartments? In my understanding the
-standard pair model (equation 9) can easily account for instantaneous
-partnership formation by simply setting the pair formation rate rho to a
-high value.
- ------------------------------------------------------------------------
-
-*Reviewers who wish to reveal their identities to the authors and other
-reviewers should include their name here (optional). These names will
-not be published with the manuscript, should it be accepted.*
-
-Reviewer #1: (No Response)
-
-Reviewer #2: (No Response)
+%suggested reviewers: Christophe Fraser, Shirreff, Samuel Alizon, Sebastian Lion, Troy Day
 

### 2016-10-28 tests/modelTests.R
Delete modelTests.R

@@ -1,54 +1,0 @@
-##
-library("deSolve")
-library("testthat")
-source("../R/LHSparms3.R")
-source("../R/hivFuns.R")
-source("../R/hivModels.R")
-source("../R/simFuns.R")
-
-## slight asymmetry is due to enforcing row-sum=1 constraint
-mm <- makeMutMat(expand(ltab_mean))
-image(log10(mm))
-
-tvec <- 0:50
-fun1 <- gfun_epc(ltab_mean,mm=FALSE)
-fun2 <- gfun_epc(ltab_mean,mm=TRUE)
-y1 <- unlist(calc_yini(ltab_mean))
-
-grad1 <- fun1(t=0,yini=y1,ltab_mean)
-out1 <- c(7.18942772482107e-06, -8.55762401214908e-07, 2.5370567666425e-14, 
-          3.80162106809482e-10, -3.37110788855153e-07, -1.13305405627565e-05)
-expect_equal(head(grad1[[1]]),out1,tol=1e-4)
-grad2 <- fun2(t=0,yini=y1,ltab_mean)
-expect_equal(grad1,grad2)
-
-res1 <- ode(y=y1,
-            func=fun1,
-            parms=ltab_mean,
-            time=tvec)
-
-## basic plots
-
-## TESTS
-## * equality of models when EP=0
-## * equality of models when partnership formation is fast
-
-
-tvec2 <- seq(0,50,by=0.1)
-ltab2 <- ltab_mean
-ltab2$alphaDist["delta"] <- 0.1
-y2 <- unlist(calc_yini(ltab2))
-fun2 <- gfun(ltab2,FALSE)
-res2 <- ode(y=y2,
-            func=fun2,
-            parms=ltab2,
-            time=tvec2)
-dim(res2)  ## fast??
-if (FALSE) devtools::install_github("rstudio/profvis")
-library(profvis)
-profvis({
-    ode(y=y2,
-            func=fun2,
-            parms=ltab2,
-            time=tvec)
-},height="250px")

### 2016-12-13 ms/cover_letter/plosHIVcover_resubm2.tex
first pass at cover letter for second revision

\documentclass[10pt]{letter}

\usepackage{graphicx}
\usepackage{myltr2}
\usepackage{times}
\usepackage[utf8]{inputenc}
%\usepackage[sort&compress]{natbib}

\bibliographystyle{abbrv}

\makeatletter
% http://tex.stackexchange.com/questions/18033/using-bibtex-with-letter-class
\newenvironment{thebibliography}[1]
     {\list{\@biblabel{\@arabic\c@enumiv}}%
           {\settowidth\labelwidth{\@biblabel{#1}}%
            \leftmargin\labelwidth
            \advance\leftmargin\labelsep
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\newcommand\newblock{\hskip .11em\@plus.33em\@minus.07em}
\makeatother

\newcommand{\revcomment}[1]{\emph{#1}}
\newcommand{\response}[1]{#1}

\begin{document}

\date{\today}

\signature{\includegraphics[height=0.35in]{bbsig3.png}\\Benjamin Bolker}

\begin{letter}{
}

\opening{To the editor,}

\textbf{ADD GENERAL POINTS HERE}

\vskip10pt
\hrule

\textbf{Editor}

\revcomment{
In particular, we would encourage you to take into consideration the remarks of reviewer 1 on the presentation of your work. Furthermore, reviewer 3 raises important points regarding the limitation of the ODE equation approach. While you already discuss some of these points, it would be great if you put your study into the context of more realistic modeling approaches of HIV and other sexually transmitted diseases. It would also be insightful if you discussed to what extent your "heterogeneous" model covers age and sex structure.
}

\textbf{Reviewer \#1}

\revcomment{
The outstanding issue is that the paper remains quite challenging to
read. It is just not written as clearly as it could be. In many
paragraphs, the authors launch into discussion of complex points with
long, multi-clause sentences, instead of explaining their motivation
and results in simple, concise phrases. Too often the authors do not
make it clear WHY they did something. Reading the original papers by
Fraser’s group on this topic (PNAS 2007 and PLoS Comp Biol 2011)
provide good examples of clear writing.
}
%\response{}

\revcomment{
\textbf{Abstract:} The first few sentences of the are confusing,
especially "Changes in the fitness landscape generally select for
higher virulence early in an epidemic”. Do you really mean to suggest
the fitness landscape - which is the mapping between parasite genotype
and phenotype - is dynamically changing? And do you think that the
main driver for changes in virulence is simply the stage of the
epidemic? If so, then it would be better to complete this sentence
with a brief description of that idea and the expected trends in
virulence over time. The other important factor here is that evolution
of virulence is mostly likely to occur after a pathogen is introduced
into a new host. It may have been optimally adapted to the original
host, but differences between hosts mean it may no longer be
optimal. This seems very relevant in in the HIV case, as it is a
zoonosis from SIV.
}

%\response{}

\revcomment{
\textbf{Author summary:} In the first sentence, I think it would be clearer to say “infectivity” instead of “prevalence”. Obviously disease prevalence changes over the course of an epidemic .. the epidemic starts with one individual being infected, and then rapidly increases. This has nothing to do with evolution - it would occur for any strain with R0>1 as it spreads to others, even if no evolution could occur!
}

\revcomment{\textbf{Introduction:}
While the intro has been improved since the last version, it is still relatively confusing. It would really be improved by explicitly and concisely explaining the goals of this study, and by addressing the other concerns below.
}

\revcomment{
Second sentence - would be much clearer to say something like “… pathogens with higher reproduction numbers - that is, the number of secondary infections caused by a single infected host over the course of its infectious period - will tend to increase in prevalence relative to strains with lower reproductive ratios.” Your definition of “reproduction” was not very clear, and it’s not a good idea to call R0 a rate, as it does not have units of time.
}

\revcomment{
Confusing sentence “ …in studies of discordant couples (…), HIV virulence as measured by the rate of progression to AIDS was both heritable and covaried with the set-point viral load (..) and the probability of transmission”.
- You are combining results of multiple different studies here. I don’t think there is a study that compared rate of progression to AIDS with probability of transmission. I think studies have a) compared SPVL to rate of progression, b) showed that SPVL was heritable, c) compared SPVL to probability of transmission. I think b) and c) required studying serodiscordant couples but I don’t think most studies of a) did.
}
\revcomment{
The paragraphs starting at line 40 read like methods, not introduction
}

\revcomment{
The term “epidemiological structures” is used throughout to mean the dynamics of the contact process (partnership dynamics), but this is not a common phrase and is not at all descriptive of what you want it to mean.
}

\revcomment{\textbf{Methods:}
If you refer to parameters here, they must be defined here, not just in the SI (i.e. for $c_u$, $c_w$, etc)
}

\revcomment{
What is the rationale behind starting at SPVL of 3? All of your results about the general time course of the mean SPVL depend on this. If it is simply to compare to Shirreff, then should state this.
}

\revcomment{
Figure 1 is not well explained in the text or the caption. What is the point of these comparisons? Like where do these three values of r come from? Which one did Shirreff use? Same for other quantities.
}

\revcomment{
You can’t assume that your reader knows exactly what was done in Shirreff’s paper. Instead of always saying what you did relative to Shirreff, the comparison between your methods and theirs should be written in a way that is clear, even if the reader has never read Shirreff.
}

\revcomment{
The description of the different contact dynamics is confusing. Again, don’t assume readers have read Shirreff or Champredon - clearly explain what the models are. What you name the variables (e.g. $SI$, $II$) is not important, since you don’t list the equations in this section anyways. The paragraph starting at 167 is more clear, but makes no sense to have this AFTER the more detailed description of the differences between the models - should be first. Start simple, add details.
}

\revcomment{
Why did you assume a constant population size - that is, replace any dead individuals immediately with new susceptible ones? In this case, the limitation on susceptible that occurs near epidemic peak is greatly dampened, and is definitely going to change the infection dynamics and probably also the virulence trajectory. Since the goal of the paper is to add more realism by including more realistic partnership dynamics, then this seems like any easy place to avoid removing realism.
}

\revcomment{\textbf{Results:}
I would highly recommend that the figures and descriptions of the results refer mainly to SPVL, and augment with results on time to progression. It is just much simpler and more intuitive to think about SPVL. SPVL is the only quantity that is actually observed in most cases. Referring to time to progression is very confusing because LOWER time to progress = HIGHER virulence so just takes more effort to interpret.
}

\revcomment{
Throughout the paper you refer to the conclusion that the simplest and most complex models give the same results, and that models of intermediate complexity give different results. But none of the figures are really presenting this point, as someone looking at the figure probably does not remember the details of each model based only on its code-word, and the series in the figures do not seem to be organized in terms of model complexity.
}

\revcomment{
First paragraph: Don’t understand “As r decreases from 0.084 to 0.42" . Where did these r values come from? Is r changing over time? Or over different iterations? Why not just fit to the observed r value? And this change you’ve described is an INCREASE, not a decrease.
}

\revcomment{
Why are you changing the initial infectious density?
}

\revcomment{
Line 306: Why would the speed of virulence evolution affect the maximum attained virulence? Doesn’t make sense. Should just influence the time to peak.
}

\revcomment{
It is very hard as a reader to maintain any interest in the results of different models when there was so little description of them, and, there is just too much detail for one to store in one’s head throughout the paper. I would recommend having a diagram, or a table, that clearly compares each model, for what features it does and doesn’t have, so the reader can keep referring back to this to try to make sense of the findings.
}

\revcomment{
I don’t understand why Figure 5 shows the unscaled parameters, when in reality you had to scale the parameters. Does not make sense to show summary statistics for scenarios that you know are unrealistic (because they would lead to epidemics with r values that do not match the observed value)
}

\revcomment{
Figure captions are not publication level. Some general advice on figure captions, taken from another source: Figures should be able to be interpreted relatively independently of the text. Most readers will go straight to the figures before reading the text in detail, and so the captions should not assume the reader has internalized every detail of the text preceding the figure citation. Figure captions should walk the reader through what experiment (simulation) was done, what analysis (statistics) was done on the data, and what quantities are shown in each axis. 100-200 words is a normal length. The first sentence (might be called the title), is often bolded, and can be either a summary of what data (or model, etc) the figure is showing, or a statement of the main results. Try to be consistent. In general, figures are supposed to describe results, not conclusions. Next, describe each panel of the figure, referring to it by label (eg A)). For each panel, or at the end of the caption (if this info is the same for all panels), describe the values of any important quantities that were held constant in during the experiment that generated the figure, state any important assumptions, and state sampling size or time if relevant. Avoid using parameter symbols or abbreviations, but if they are used, provide definitions in the caption. These conventions are often ignored in technical mathematical biology papers (i.e. those in JTB), but are crucial for making theoretical papers palatable for interdisciplinary audiences.
}

\revcomment{
\textbf{Discussion:}
Would be helpful to start off by summarizing the main goals of the paper and then the main findings of the paper. Then, can get into limitations
}

\revcomment{
It would be very helpful to be able to give a intuitive or at least qualitative explanation for why these mixing patterns have such an influence on virulence.
}

\textbf{Reviewer \#2:}
\revcomment{
 The authors have appropriately dealt with my previous comments. Furthermore, I think that the addition of a model with heterogeneity in sexual activity is useful.
}

\closing{Sincerely,}

%\bibliography{virulence}

\end{letter}
\end{document}

### 2016-12-13 ms/ms.tex
name change for main ms file

--- 2016-10-27 ms/plos_latex.tex
@@
 % Please keep the abstract below 300 words
-% current is 282
+% current is 296
 \section*{Abstract}
-The evolutionary dynamics of parasite virulence change
-in important ways over the
-course of an emerging epidemic.
-Changes in the fitness landscape
-generally select for higher virulence early in an
-epidemic; however, quantitative outcomes
-may depend sensitively on epidemiological details and the structure of
-mathematical models used to portray them.  Fraser \emph{et al.} 
-have proposed a model for the eco-evolutionary
-dynamics of HIV that captures
+Early in an epidemic, high densities of susceptible hosts
+select for relatively high parasite virulence; later in the epidemic,
+lower susceptible densities select for lower virulence.
+Thus over the course of a typical epidemic the average virulence 
+of parasite strains increases initially,
+peaks partway through the epidemic, then declines again.
+However, precise quantitative outcomes, such as the peak virulence reached
+and its timing, may depend sensitively on epidemiological details. 
+Fraser \emph{et al.}
+proposed a model for the eco-evolutionary
+dynamics of HIV that incorporates
 the tradeoffs between transmission and virulence (mediated by
 set-point viral load, SPVL) and their heritability between
-hosts. However, these models use implicit
-representations of the transmission process that drastically simplify the
-partnership dynamics that previous research has found to be critical
-in driving epidemics of sexually transmitted diseases.  
+hosts. Their models use implicit equations to
+capture the effects of partnership dynamics that are at the core of 
+epidemics of sexually transmitted diseases. 
 Our models combine HIV virulence tradeoffs with a range of
-epidemiological structures, modeling partnership formation and
+contact models, modeling partnership formation and
 dissolution and allowing for individuals to transmit disease outside
 of partnerships. We assess summary statistics such as the peak virulence
-(corresponding to the minimum expected time of progression to AIDS) across all
+(corresponding to the minimum expected time of progression to AIDS) across
 models for a range of 
-partnership dynamic parameters 
-applicable to the HIV epidemic in sub-Saharan Africa.
+parameters applicable to the HIV epidemic in sub-Saharan Africa.
 Although virulence trajectories are broadly similar
-across model structures, the timing and magnitude of the 
-minimum expected time to progression vary
+across models, the timing and magnitude of the 
+virulence peak vary
 considerably.
-Models of intermediate
-complexity as used by Shirreff \emph{et al.}
-predicted lower slower progression/lower virulence (a minimum of 15 years to
-progress to AIDS) compared to both more realistic models
-and simple random-mixing models with no partnership structure
+Previously developed implicit models predicted 
+lower virulence and slower progression
+at the peak (a minimum of 15 years to
+progress to AIDS) compared both to more realistic models
+and to simple random-mixing models with no partnership structure
 at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
@@
 surprising outcome occurs because the dominance of extra-pair
-contact in the realistic model tends to swamp the effects of
+contact in the realistic model swamps the effects of
 partnership structure.
@@
 
-%% 162 words
-
-Pathogens such as HIV can evolve rapidly in response to changes in
-their environments; such changes include both increases in disease
-prevalence and disease virulence over the course of an epidemic, or
-decreases in both after treatment interventions. While researchers
-have successfully used computational models to explore these
-evolutionary dynamics, these models often neglect details such as the
-formation and dissolution of sexual partnerships; other research has
-shown that these processes can strongly affect epidemic outcomes. We
-built and compared models that used different methods to model both
+%% 199 words
+
+Pathogens such as HIV can evolve rapidly when the environment changes.
+Such changes include decreases in the probability that an infectious
+contact (such as a sneeze for a respiratory disease, or an unprotected
+sex act for a sexually transmitted disease) encounters an uninfected,
+susceptible person and thus has a chance to transmit the
+pathogen. Disease spread during an epidemic reduces the fraction of
+susceptible people in the population and thus changes the direction of
+parasite evolution.  While researchers have explored these
+evolutionary dynamics with simulations, their models have neglected
+important processes such as people entering and leaving sexual
+partnerships. We compared several evolutionary models that include
 partnership dynamics and sexual contact outside of stable
-partnerships. Models of intermediate complexity predicted much lower
+partnerships. Models of intermediate complexity predicted lower
 virulence over the course of the epidemic (a minimum of 15 years to
-progress to AIDS) compared to both more realistic models and simple
-random-mixing models with no partnership structure at all (both
-approx. 7.25 years to progress to AIDS); sexual contact outside of
-stable partnerships tended to wash out the effects of epidemiological
-structure. The large differences in evolutionary dynamics among
-different epidemiological models suggests that researchers trying to
-predict the evolution of pathogens should proceed with caution.
+progress to AIDS) than either more realistic models or simple models
+without any partnership structure at all (both approx. 7.25 years to
+progress to AIDS), because sexual contact outside of stable
+partnerships tended to wash out the effects of epidemiological
+structure. The large effect of epidemiological details on evolutionary
+dynamics suggests that researchers trying to predict the evolution of
+pathogens should proceed with caution.
 
@@
 this work was funded by NSERC Discovery Grant 386590-2010.
+Computational resources were provided by SHARCnet.
 
@@
 May RM, Anderson RM.
-\newblock The Transmission Dynamics of Human Immunodeficiency Virus (HIV) [and
-  Discussion].
+\newblock The Transmission Dynamics of Human Immunodeficiency Virus (HIV).
 \newblock Philosophical Transactions of the Royal Society of London Series B,

### 2016-12-13 ms/cover_letter/plosHIVcover_resubm2.tex
cover letter updates for abstract & author summary

@@
 
-%\response{}
+\response{We have modified the first few sentences to make them
+more concrete and (hopefully) more understandable. 
+Yes, we did mean that the stage of the epidemic is a strong
+driver of virulence evolution, and the one we focus on here.
+We have removed the potentially confusing term ``fitness landscape''.
+
+The point about initial virulence being measured immediately after
+a species jump is interesting, but since it is somewhat peripheral
+to our main exploration here, we have opted to include in the 
+Discussion rather than trying to squeeze it into the abstract
+\textbf{don't forget to do this}
+}
 
@@
 }
+
+\response{We were previously trying to pack too many ideas into a few sentences.
+We have made the text more concrete, clarifying that we do indeed mean to describe the fact that prevalence changes over the course of the epidemic, and that this changes the evolutionary environment for the pathogen.}
 

### 2016-12-20 ms/cover_letter/plosHIVcover_resubm2.tex
updating cover letter, tweaking figs; add rev 3 PDF

@@
 
+\response{comparing to Shirreff \ldots Figure 1 shows comparison \ldots picking high start virulence, won't see peak at all \ldots very much unknown \ldots implicit assumption that there's not an \emph{interaction} between starting SPVL and characteristics of models \ldots}
+
 \revcomment{
@@
 
+\textbf{Reviewer \#3:}
+\revcomment{
+p. 10 if not counteracted by increased mortality among hosts infected by these reproductive strains.
+}
+
+\response{(confusion between transmission rate and  ${\mathcal R}_0$ ---
+  clarify in text)}
+
+\revcomment{
+p.
+11 instead of calling this "epidemiological structure" it may be more appropriate to refer to it as "contact structure".
+The authors investigated the effect of adding structure to the pattern of sexual contacts.
+There is nothing intrinsically "epidemiological" about forming (stable and/or once-off) sexual relationships.
+}
+
+\revcomment{
+p. 11 Given the long time horizons of the simulations (500-4000
+years), this simplifying assumption [single-stage model of HIV
+progression] may introduce bias, but unfortunately the magnitude and importance thereof is unknown.
+In more realistic models, one would expect that the high-risk susceptible individuals get depleted at a higher rate than the low risk susceptibles, with the epidemiological effects of this risk gradient playing out over a small number of decades.
+If or how this dynamic affects the evolution of virulence is unclear but if the goal is explicitly to study this evolution, it is problematic that all models under study are unable to take these demographic effects into account, by design.
+}
+
+\revcomment{
+p. 12 As noted by reviewer 1, the decision to ignore temporal variation in
+infectiousness over the main stages of HIV infection is a serious
+limitation of this study. In particular, the epidemiological
+contribution of the spike in HIV viral load during acute and early HIV
+infection depends on the assumed contact pattern. In models where
+individuals are not "trapped" in serially monogamous relationships,
+recently infected individuals can make efficient use of this phase as
+they can transmit to new or additional sexual partners as soon as they
+are infected. In contrast, in models that only allow for stable
+partnerships, this spike in infectiousness will "go to waste" as the
+newly infected individuals are most likely still in the relationship
+with the partner that they acquired the infection from during these
+initial few months after HIV acquisition. For highly virulent viral
+strains (high SPVL and short survival) such window of opportunity is
+more important than for viral strains that allow the host to transmit
+for a decade or more.
+}
+
+\response{make sure we have a caveat/comment in text}
+
+\revcomment{
+p. 12 Showing that the model with only pair formation dynamics (~
+Shirreff et al.) produces qualitatively similar results to those from
+the Shirreff model, does not prove or suggest that the effect of
+adding additional extra-couple transmission is robust against the
+model's structure with respect to the stages of HIV infection and
+associated infectiousness levels.
+}
+
+\response{same thing again}
+
+\revcomment{
+p.
+21 The authors touch on two weaknesses of their analysis here: 1.
+The level of sufficient realism is questionable, despite great efforts to take into account various sources of variation (with several thousands of compartments as a result).
+2. The real-life implications of their main metrics (time to reach
+peak virulence, time to reach equilibrium virulence, progression time
+to AIDS at peak virulence) is unclear, because in all model scenarios
+(strain-specific) transmission and disease progression/mortality
+parameters stay constant over time. However, dramatic increases in the
+uptake of HIV treatment (ART) have been observed across the world, and
+trends in sexual risk behaviour may have changed both in the pre-ART
+and ART era.
+}
+
+\revcomment{
+p. 22  The model with heterogeneous levels of sexual activity (the
+"heterogeneity" model) is an extension of the pair formation + epc
+model. However, in principle, this extra layer of heterogeneity in
+sexual contact rate could have been added to the other 5 "base" models
+as well. In the current comparison it is not possible to assess the
+relative effect of adding extra-couple and uncouple transmission
+versus contact heterogeneity.
+}
+
+\revcomment{
+p. 23 
+This sentence [``These differences are practically as well as
+scientifically important''] seems to belong better in the discussion section. Here, the authors do not explain what the practical or scientific importance of the differences is.
+}
+
+\revcomment{
+p.
+27 The main finding of this analysis seems to be that allowing "unstructured" transmission events, be it extra-couple or uncoupled, influences the expected trajectory of SPVL evolution over rather long time scales.
+As Reviewer 1 notes, this is indeed a rather theoretical finding.
+In my option, two key questions remain unaddressed: 1.
+Why should we care more about the effect of this aspect of the assumed contact pattern than the effects of other aspects of the contact pattern and disease process (e.g.
+age-mixing, stage-varying infectiousness, ART status, depletion of high-risk susceptibles)?
+2. What are the practical implications of these findings for HIV
+epidemiology? For HIV prevention and treatment? Does it help us in any
+way to interpret empirical time-trends in SPVL or disease progression?
+}
+
 \closing{Sincerely,}

### 2016-12-20 figures/figures.R

@@ -17,7 +17,6 @@ library(GGally) ## need BMB version
 library(ggstance)
 
 do_png <- FALSE
-do_duration <- FALSE
 
 opts_chunk$set(echo=FALSE,error=FALSE)
 if (.Platform$OS.type=="windows") {
@@ -153,9 +152,11 @@ gg_basetraj <- ggplot(ss,aes(tvec,mean,ymin=lwr,ymax=upr))+
     geom_line(aes(colour=model,linetype=model),lwd=1)+
     geom_ribbon(aes(fill = model), alpha=0.3)+
     labs(x="time (years)")+
+    scale_x_continuous(expand=c(0,0))+
     theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
           axis.title.x = element_text(margin = margin(10,0,0,0)),
-		legend.position = "none") +
+		legend.position = "none",
+          panel.grid = element_blank()) +
 	facet_wrap(~cmodel, nrow = 2) +
 	zero_margin
 
@@ -175,9 +176,12 @@ nostrips <- theme(strip.background = element_blank(),
 
 ## r fig2, fig.width=6, fig.height = 4, echo = FALSE, cache = TRUE,dpi = 400}
 
-gg_virtraj <- direct.label(gg_basetraj + nostrips +
+gg_virtraj <-
+    suppressWarnings(direct.label(gg_basetraj + nostrips +
     labs(y=expression(population~mean~set-point~viral~load~(log[10]))),
-    method=list("top.points2","bumpup"))
+    method=list("top.points2","bumpup")))
+
+## FIXME: make minimal example for direct labels/linetype warning?
 
 ggsave(gg_virtraj,file="fig2.pdf",width=8,height=6)
 if (do_png) ggsave(gg_virtraj,file="fig2.png",width=8,height=6,dpi=400)
@@ -232,9 +236,9 @@ w <- with(mL,which(model=="random" & variable== "eq_vir"))
 rval <- mean(mL$value[w])
 mLw <- droplevels(mL[-w,])
 mLw$variable <- factor(mLw$variable,
-											 levels = c("peak_time", "peak_vir", "eq_vir", "rel_vir"),
-											 labels = c("peak~time~(years)",
-											 					 "maximum~mean~log[10]~SPVL",
+         levels = c("peak_time", "peak_vir", "eq_vir", "rel_vir"),
+         labels = c("peak~time~(years)",
+                    "maximum~mean~log[10]~SPVL",
[…]

### 2017-01-05 ms/cover_letter/plosHIVcover_resubm2.tex
working on resubm (text & cover ...)

@@
 \newcommand{\response}[1]{#1}
+\newcommand{\rzero}{\ensuremath{\mathcal R}_0}
 
@@
 
+\response{
+We were indeed thinking of reproduction \emph{rates} (i.e., $r$) rather than $\rzero$ here; we may be slightly more
+attuned to $r$ because of our eco-evolutionary work, but we agree that it's clearer to start out with an $\rzero$
+framing, which will be more familiar to technical readers (and clarifying the $r$-$\rzero$ distinction is really too much to try to pack into the first few sentences of the introduction).
+}
+
 \revcomment{
@@
 }
+
+\response{
+This is a good point. We have clarified the text here.
+}
+
 \revcomment{
 The paragraphs starting at line 40 read like methods, not introduction
+}
+
+\response{
+We think some of this material is useful to foreshadow what we're
+going to do in the body of the paper, but we have trimmed the
+technical details.
 }
@@
 }
+
+\response{We have changed this term to ``contact structure'' throughout.}
 

### 2017-01-06 figures/figures_hetero.R
added natural bd

@@ -2,19 +2,20 @@ library(ggplot2); theme_set(theme_bw())
 library(dplyr)
 library(tidyr)
 load("../simdata/ev_hetero.rda")
+load("../simdata/ev_bd.rda")
 
 zero_x_margin <-
 	theme(panel.spacing.x=grid::unit(0, "lines"))
 
 tvec <- c(1:1000)
-het <- seq(1, 7, 2)
+div <- seq(1, 7, 2)
 model <- c("full+epc", "full", "instswitch+epc", "instswitch")
 
-sort_matrix <- function(mat){
-	list(hetero = data.frame(tvec, mat[,het]), 
-			 homo = data.frame(tvec, mat[,-het])) %>%
+sort_matrix <- function(mat, name){
+	list(hetero = data.frame(tvec, mat[,div]), 
+			 homo = data.frame(tvec, mat[,-div])) %>%
 		lapply(function(x) setNames(x, c("tvec", model))) %>%
-		setNames(c("hetero", "homo")) %>%
+		setNames(name) %>%
 		bind_rows(.id = "type") %>%
 		as_data_frame %>%
 		gather(key, value, -type, -tvec) %>%
@@ -29,6 +30,14 @@ g_base <- ggplot(data = NULL, aes(tvec, value, col = type)) +
 	zero_x_margin +
 	theme(panel.grid = element_blank())
 
-g_base %+% (vir_matFull %>% sort_matrix)
+## homo vs hetero
 
-g_base %+% (I_matFull %>% sort_matrix)
+g_base %+% (vir_matFull %>% sort_matrix(c("hetero", "homo")))
+
+g_base %+% (I_matFull %>% sort_matrix(c("hetero", "homo")))
+
+## natural birth and death
+
+g_base %+% (vir_mat %>% sort_matrix(c("no_bd", "bd")))
+
+g_base %+% (I_mat %>% sort_matrix(c("no_bd", "bd")))

### 2017-01-06 ms/ms.tex

@@
 {\Large
-\textbf\newline{Effects of Epidemiological Structure on the Transient Evolution of HIV Virulence} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
+\textbf\newline{Effects of Contact Structure on the Transient Evolution of HIV Virulence} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
 }
@@
 progress to AIDS), because sexual contact outside of stable
-partnerships tended to wash out the effects of epidemiological
+partnerships tended to wash out the effects of contact
 structure. The large effect of epidemiological details on evolutionary
@@
 importance. In general, evolutionary theory suggests that
-disease strains that can reproduce more --- where reproduction
-is defined here as the amount of \emph{between-host} transmission, or
-the number of new hosts infected --- will increase in prevalence.
-Pathogens can increase their net reproduction rate either
+pathogens with higher reproduction numbers ($\rzero$) --- that is, the number of secondary infections caused by a single infected host over the course of its infectious period --- will tend to increase in prevalence relative to strains with lower reproductive ratios.
+Pathogens can increase their reproduction numbers either
 by increasing their transmission rate, 
@@
 The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
-the transmission and disease-induced mortality rate are both linked to the rate at which
+transmission and disease-induced mortality rates are both driven by the rate at which
 the pathogen exploits host resources for within-host reproduction, 
@@
 successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
-Fraser \etal\ have applied these ideas in a particularly
-interesting way 
-by showing that HIV appears to satisfy the prerequisites of
-the tradeoff theory: in studies of discordant couples (i.e. long-term
-sexual partnerships with one infected and one uninfected partner), HIV
-virulence as measured by the rate of progression to AIDS was 
-both heritable and covaried with the set-point viral load (SPVL: i.e., the
+
+Fraser \etal\ have applied these ideas to HIV,
+showing that the virus appears to satisfy the prerequisites of
+the tradeoff theory.
+The \emph{set-point viral load} (SPVL: i.e., the
 characteristic virus load measured in blood during the intermediate
-stage of infection) and the probability of transmission.
-Higher viral loads led to shorter progression times (higher virulence)
-and faster transmission, with a decelerating relationship
-between virulence and transmission as required by the tradeoff theory
-\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
+stage of infection) is a measurable proxy for the rate of HIV within-host
+reproduction. Higher viral loads are correlated with 
+shorter progression times (higher virulence).
+Studies of discordant couples --- stable
+sexual partnerships with one infected and one uninfected partner --- %
+have shown that SPVL is (1) correlated with faster transmission
+(the time to infection of the uninfected partner is shorter when
+their partner has a higher SPVL) and (2) heritable
+(when the uninfected member of
+the couple does become infected, their SPVL is similar to their
+partner's).
+Furthermore, the rate of transmission increases with
+increasing SPVL (and hence with decreasing progression time)
+but with a decreasing slope as progression time decreases,
+fulfilling the requirements of the tradeoff theory
+ \cite{Fraser+2007}.
+Subsequent studies
 \cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
@@
 However, despite these studies' attention to detail at the individual
-or physiological level, the epidemiological structures used in these
+or physiological level, the contact structures used in these
 models are relatively simple.
@@
 contact --- without representing the explicit dynamics of pair
-formation and dissolution or accounting for extra-partnership contact
+formation and dissolution or accounting for extra-partnership contact %
 --- or use an agent-based formulation with parameters that effectively
@@
 we explore the effects of incorporating \emph{explicit}
-epidemiological structure in eco-evolutionary models.
-
-We add complexity to the epidemiological model following the general
-approach of Champredon \etal\ \cite{champredon_hiv_2013},
-which is in turn based on work of Dietz and Hadeler
-\cite{dietz_epidemiological_1988}; individuals
-join and leave partnerships at a specified rate, and can have sexual
-contact both within and outside of established partnerships. In order
-to explore how virulence evolution depends on epidemiological
-structure, we consider a series of models with increasing levels of
-complexity. In order to avoid dependence of the results on a
-particular set of parameters --- as we explain below, finding matching
-sets of parameters across models with widely differing epidemiological
-structures is challenging --- we evaluate our models across a wide
-range of parameters, again following Champredon
-\etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
-design. For each model run, we compute a set of metrics (minimum
-progression time/peak virulence, timing of maximum virulence, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.
-
-As our primary goal is to explore how different epidemiological
-structures (i.e. partnership dynamics and contact structures) affect
-our conclusions about the evolution of virulence, our models use a
-simplified description of within-host dynamics and heritability
-derived from Shirreff \etal's multi-strain evolutionary model
-\cite{shirreff_transmission_2011}. Like Shirreff \etal, we use a
-simple susceptible-infected-susceptible demographic formulation;
-rather than modeling birth and death (or more specifically,
-recruitment into the sexually active population and death), we assume
-that whenever an individual dies from infection, another enters the
-susceptible compartment.
+contact structure in eco-evolutionary models.
+
+Because our main goal is to explore how conclusions about virulence
+evolution depend on the way in which contact structure
+is modeled, we consider a series of
+models with increasing levels of complexity in the contact
+structure, but simplify some of the other epidemiological
+processes (such as the within-host life history of HIV). 
+We evaluate our models
+across a wide range of parameters, using a Latin hypercube design; for
+each model run, we compute a set of metrics 
+that summarize the evolutionary trajectory of SPVL
+over the course of the epidemic.
 
@@
 
-We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} contact structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
 Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
@@
 In the particular example of HIV virulence eco-evolutionary dynamics
-and the complexity of epidemiological structures
+and the complexity of contact structures
 we reach the slightly ironic conclusion that the
@@
 we were making meaningful comparisons across a range of
-models with radically different epidemiological structures, and
+models with radically different contact structures, and
 hence involving different interpretations of the same quantitative
@@
 evolution \cite{payne_impact_2014,roberts2015impact,herbeck_evolution_2016}.  While these
-efforts are well-intentioned, we caution that epidemiological and
-other structural details that are currently omitted from these models
+efforts are well-intentioned, we caution that 
+structural details that are currently omitted from these models
 could significantly change their conclusions.
@@
 We would like to thank Christophe Fraser and
-David Champredon for access to simulation code;
+David Champredon for access to simulation code
+and Wim Delva and an anonymous reviewer for useful comments;
 this work was funded by NSERC Discovery Grant 386590-2010.

### 2017-01-07 R/hivModels_bd.R
added constant birth rate

@@ -1,0 +1,100 @@
+gfun_bd <- function(parameters, experimental=FALSE) {
+	px <- expand(parameters)
+	g <- function(t,yini,parameters) {
+		with(as.list(c(yini,px)), 
+				 { 
+				 	SI <- yini[3:(2+n.alpha)]
+				 	I <- yini[(3+n.alpha):(2+2*n.alpha)]
+				 	II <- matrix(yini[(3+2*n.alpha):(2+(2+n.alpha)*n.alpha)],
+				 							 n.alpha, n.alpha)
+				 	
+				 	II_adj <- II
+				 	diag(II_adj) <- 2 * diag(II_adj) 
+				 	
+				 	prop <- (c_u_ratio * I + c_e_ratio * (SI + colSums(II_adj)))/
+				 		(c_u_ratio * (S + sum(I)) +
+				 		 	c_e_ratio * (2 * SS + 2 * sum(SI) + sum(II_adj)))
+				 	
+				 	dS <- - rho * S * S/(S+sum(I)) +
+				 		2 * c_mean * SS - rho * S *sum( I)/(S+sum(I)) +
+				 		c_mean * sum(SI) + sum(lam * SI) - sum(prop * c_u) * S +
+				 		m * (1 - S) + 2 * m * SS + m * sum(SI)
+				 	dSS <- 0.5 * rho * S * S/(S+sum(I)) - c_mean * SS -
+				 		2 * sum(prop * c_e) * SS - 2 * m * SS
+				 	dSI <- rho * S * I/(S+sum(I)) - c_mean * SI -
+				 		lam * SI - Beta * SI +
+				 		(prop * 2 * c_e * SS) %*% p - sum(c_e * prop) * SI - 2 * m * SI
+				 	
+				 	infrate <- (Beta * SI) * p
+				 	infrate_adj <- infrate + t(infrate)
+				 	diag(infrate_adj) <- diag(infrate_adj)/2
+				 	
+				 	infrate_e <- outer(SI, c((c_e * prop) %*% p))
+				 	infrate_e_adj <- infrate_e + t(infrate_e)
+				 	infrate_e_save <- infrate_e_adj
+				 	diag(infrate_e_adj) <- diag(infrate_e_adj)/2
+				 	
+				 	frate.I <- rho * outer(I,I,"*")/(S+sum(I))
+				 	diag(frate.I) <- diag(frate.I)/2
+				 	
+				 	dI <- - rho * S * I/(S+sum(I)) +
[…]

### 2017-01-07 R/simulations/ev_bd.R

@@ -1,0 +1,65 @@
+library("deSolve")
+source("../simFuns.R")  ## for transform.list
+source("../hivFuns.R")
+source("../Param.R")
+source("../hivModels.R")
+source("../hivModels_bd.R")
+
+HIVpars.mean <- as.HIVvirparlist(apply(rbind(HIVpars_range, Het_range), 1, geom_mean))
+HIVpars.mean <- transform(HIVpars.mean, ini_V = 3, m = 1/50)
+
+I_mat <- matrix(NA, nrow = 1000, ncol = 8)
+vir_mat <- matrix(NA, nrow = 1000, ncol = 8)
+eq_vec <- rep(0, 8)
+peak_mat <- matrix(NA, nrow = 8, ncol = 2)
+val_vec <- rep(0, 8)
+
+tvec <- c(1:1000)
+
+fn <- "ev_bd.rda"
+
+HIVpars <- HIVpars.mean
+gfun1 <- gfun; gfun1_bd <- gfun_bd
+
+fun <- function(model, i){
+	yini <- calc_yini(HIVpars)
+	cat("little_r...")
+	little_r <- get_rval(model(HIVpars),yini, HIVpars, tvec = c(1:200))
+	
+	if(little_r < 0.04){
+		interval <- c(1,5)
+	}else{
+		interval <- c(0.01,1)
+	}
+	cat("adj...")
+	val_vec[i] <<- find_scale(0.04, model, yini, HIVpars, interval=interval, adjpar = "scale_all", tvec = c(1:200))
+	HIVpars_adj <- transform(HIVpars, scale_all = val_vec[i], hmax = 0.5)
+	cat("sim...\n")
+	r <- rk(unlist(yini), func = model(HIVpars_adj), parms = HIVpars_adj, times = tvec, hmax = 0.5)
+	
+	I_mat[,i] <<- r[,(ncol(r) - 1)]
+	vir_mat[,i] <<- r[,ncol(r)]
+	eq_vec[i] <<- vir_mat[length(tvec),i]
+	
+	peak_mat[i,1] <<- which.max(vir_mat[,i])
+	peak_mat[i,2] <<- max(vir_mat[,i])
+	return(NULL)
+}
+
+if(file.exists(fn)){
+	load(fn)
+}else{
+	for(i in 1:4){
+		if(i == 3){
+			HIVpars <- transform(HIVpars, rho_base = 5)
+		}
[…]

### 2017-01-09 ms/ms.tex
added vital dynamics and worked on figure 5

@@
 Finally, the sensitivity plot (\figurename~\ref{fig:plot_sens}) shows the effects 
-of each parameter on the summary statistics. In almost every case the
-effects of the parameters are monotonic; note that
-the plot shows the effects of the \emph{unscaled} parameters, i.e.
-before they have been calibrated to achieve a standard initial epidemic
-growth rate.
-Increases in the transmission rates ($\beta_P$, $\beta_D$)
-and durations ($D_P$, $D_D$) in the primary and disease stages generally
-decrease the equilibrium virulence, peak virulence, and peak time,
-although the random and pair-formation+epc 
-models have high, relatively
-constant values with respect to these parameters
-(because the patterns are so similar across this set of parameters,
-\figurename~\ref{fig:plot_sens} shows only $D_P$).
-
-The partnership dissolution rate ($c$), which essentially
-acts as a contact rate in the model,
-increases virulence and peak time in almost all
-cases, although the pair-formation+epc
-model is again relatively insensitive.
+of each parameter on the summary statistics. 
+The most notable difference can be observed by comparing the scaled parameters (e.g.
+$\beta_P$, $\beta_D$, $c$, $\rho$) with the unscaled parameters (e.g.
+$D_P$, $D_D$, $c_e/c_w$, $c_u/c_w$, $\kappa$, $\mu$);
+note that the effects of $\beta_D$ and $D_D$ are not shown in \figurename~\ref{fig:plot_sens} as their patterns are so similar to $\beta_P$ and $D_P$, respectively.
+
+One of the distinguishing features of the scaled parameters (in comparison to the unscaled parameters) is that the range of the parameters is compressed (relative to other models) for models without extra-pair contact as those models require parameters to be scaled by a large value in order to achieve a given $r$ of 0.04.
+Models with extra-pair contact show wide range parameters as they can display a wide range of dynamics depending on $c_e/c_w$ (as well as $c_u/c_w$ for models with uncoupled mixing) and thus require wide range of scaling factors to achieve the target $r$.
+Parameter range for random-mixing model is severely compressed because latin hypercube sampling does not affect the prevalence trajectory very much, and scaling factor does not vary much across parameters.
+
+For parameters involved in partnership turnover ($c$ and $\rho$), main distinction is between models with and without extra-pair contact.
+With extra-pair contact, we get gradual decrease in peak time, maximum \Lspvl, and equilibrium \Lspvl\ with increasing turnover rates; Otherwise we get increase in summary statistics.
+Although we have not verified this idea analytically, we speculate that increase in turnover rates diminishes the effect of epc (equivalently increasing relative importance of within-partnership contact), therefore selecting for lower \Lspvl.
+For models without epc, increased rates of partnership change decrease the level of structured mixing (mimicing epc models), resulting in selection for higher \Lspvl.
+
+We also note that the implicit model and the instantaneous partnership formation model show similar patterns in scaled parameters.
+In fact, we can see that the effect of partnership dissolution rate, $c$, on equilibrium \Lspvl\ of the two models are almost identical (in \figurename~\ref{fig:pairplot}, they have distinguishing features).
+Lastly, we note that increase in transmission rates ($\beta_P$ and $\beta_D$) has a decreasing effect on summary statistics for all models except for random mixing model.
+
+Surprisingly, once calibration
+is taken into account, the unscaled parameters have little to no effect overall.
+Increase in duration ($D_P$, $D_D$) in the primary and disease stages generally decrease the equilibrium virulence, peak virulence, and peak time,
+although the models with uncoupled mixing and random mixing model have high, relatively
+constant values with respect to these parameters.
 The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
-virulence in the instantaneous-switching+epc model, but not the pair-formation+epc
+summary statistics in the instantaneous-switching +epc model, but not the pair-formation+epc
 model (probably because the uncoupled individuals present in the pair-formation+epc
 model make extra-pair contact by coupled individuals less important).
-Surprisingly, once calibration
-is taken into account, the remaining parameters have little effect overall.
-The rate of partnership formation ($\rho$) 
-has little impact on the models with finite pair-formation times.
-The relative rate of uncoupled contact ($c_u/c_w$) slightly decreases the
-minimum and equilibrium progression time and delays the peak time in the
-pair-formation+epc model, but neither the uncoupled contact rate nor
+Similarly, increasing the ratio of uncoupled to within-pair contact, $c_u/c_w$, increases peak and equilibrium \Lspvl\ and delays peak time of the pair-formation+epc model but has almost no effect on the heterogeneous model.
+Neither the uncoupled contact rate nor
 the mean ($\mu$) or CV$^2$ of the number of non-cohabiting sexual

### 2017-01-09 R/hivFuns.R

@@ -31,10 +31,15 @@ expand.HIVvirparlist <- function(x, pair = TRUE, ...) {
           ##Risk distribution
           gamma.shape <- 1/kappa
           gamma.scale <- mu/gamma.shape
-          bin.risk <-  qgamma(seq(0.01, 0.99, length.out = n.risk+1), shape = gamma.shape, scale = gamma.scale)
-          
-          p.risk <- rep(1/n.risk, n.risk)
-          r.risk <- (bin.risk[-1] + bin.risk[-(n.risk+1)])/2
+          if(n.risk > 1){
+          	bin.risk <-  qgamma(seq(0.01, 0.99, length.out = n.risk+1), shape = gamma.shape, scale = gamma.scale)
+          	
+          	p.risk <- rep(1/n.risk, n.risk)
+          	r.risk <- (bin.risk[-1] + bin.risk[-(n.risk+1)])/2
+          }else{
+          	p.risk <- 1
+          	r.risk <- 1
+          }
           
           ##Heterogeneity
           alpha2 <- rep(alpha, n.risk)

### 2017-01-09 R/hivModels_bd.R

@@ -51,7 +51,7 @@ gfun_bd <- function(parameters, experimental=FALSE) {
 				 	
 				 	tot_V <- sum(SI * alpha) + sum(I * alpha) +
 				 		sum(sweep(II_adj, 2, alpha, "*"))
-				 	list(c(dS, dSS, dSI, dI, dII), tot = tot, tot_I = tot_I, mean_V = tot_V/tot_I)
+				 	list(c(dS, dSS, dSI, dI, dII), tot = tot, I = tot_I/tot, mean_V = tot_V/tot_I)
 				 })
 	}
 	return(g)
@@ -93,8 +93,26 @@ gfun2_bd <- function(parameters) {
 				 	
 				 	tot_V = sum(SI * alpha) + sum(I * alpha) + sum(sweep(II_adj, 2, alpha, "*"))
 				 	
-				 	list(c(dS, dSS, dSI, dI, dII), tot = tot, tot_I = tot_I, mean_V = tot_V/tot_I)
+				 	list(c(dS, dSS, dSI, dI, dII), tot = tot, I = tot_I/tot, mean_V = tot_V/tot_I)
 				 })
 	} 
 	return(g)
+}
+
+calc_yini_bd <- function(parameters){
+	with(c(expand(parameters)),{
+		x = (c_mean + 2 * m)/(c_mean + rho + 2 * m)
+		N = (1-x)/2
+		c = Ini_I
+		I_vec = c * I_initialize(parameters)
+		I_mat = 2 * outer(I_vec, I_vec)
+		diag(I_mat) = diag(I_mat)/2
+		yini <- list(
+			S  = (1 - c) * x,
+			SS = (1 - c)^2 * N,
+			SI = 2 * (1-c) * N * I_vec,
+			I = x * I_vec,
+			II = N * I_mat)
+		return(yini)
+	})
 }

### 2017-01-09 R/simulations/ev_bd.R

@@ -8,11 +8,11 @@ source("../hivModels_bd.R")
 HIVpars.mean <- as.HIVvirparlist(apply(rbind(HIVpars_range, Het_range), 1, geom_mean))
 HIVpars.mean <- transform(HIVpars.mean, ini_V = 3, m = 1/50)
 
-I_mat <- matrix(NA, nrow = 1000, ncol = 8)
-vir_mat <- matrix(NA, nrow = 1000, ncol = 8)
-eq_vec <- rep(0, 8)
-peak_mat <- matrix(NA, nrow = 8, ncol = 2)
-val_vec <- rep(0, 8)
+I_mat_bd <- matrix(NA, nrow = 1000, ncol = 8)
+vir_mat_bd <- matrix(NA, nrow = 1000, ncol = 8)
+eq_vec_bd <- rep(0, 8)
+peak_mat_bd <- matrix(NA, nrow = 8, ncol = 2)
+val_vec_bd <- rep(0, 8)
 
 tvec <- c(1:1000)
 
@@ -22,7 +22,13 @@ HIVpars <- HIVpars.mean
 gfun1 <- gfun; gfun1_bd <- gfun_bd
 
 fun <- function(model, i){
-	yini <- calc_yini(HIVpars)
+	if(i %% 2 == 0){
+		calc_y <- calc_yini_bd
+	}else{
+		calc_y <- calc_yini
+	}
+	
+	yini <- calc_y(HIVpars)
 	cat("little_r...")
 	little_r <- get_rval(model(HIVpars),yini, HIVpars, tvec = c(1:200))
 	
@@ -32,17 +38,17 @@ fun <- function(model, i){
 		interval <- c(0.01,1)
 	}
 	cat("adj...")
-	val_vec[i] <<- find_scale(0.04, model, yini, HIVpars, interval=interval, adjpar = "scale_all", tvec = c(1:200))
-	HIVpars_adj <- transform(HIVpars, scale_all = val_vec[i], hmax = 0.5)
+	val_vec_bd[i] <<- find_scale(0.04, model, yini, HIVpars, interval=interval, adjpar = "scale_all")
+	HIVpars_adj <- transform(HIVpars, scale_all = val_vec_bd[i])
 	cat("sim...\n")
[…]

### 2017-01-09 R/simulations/ev_hetero.R

@@ -12,7 +12,7 @@ HIVpars.mean <- transform(HIVpars.mean, ini_V = 3)
 
 param_list <- rep(list(HIVpars.mean), 4)
 
-param_list[[3]]$rho_base <- param_list[[4]]$rho_base <- 5
+param_list[[3]]$rho_base <- param_list[[4]]$rho_base <- 4
 
 param_list[[2]][c("c_u_ratio", "c_e_ratio")] <- 
 	param_list[[4]][c("c_u_ratio", "c_e_ratio")] <- 
@@ -34,7 +34,7 @@ fn <- "ev_hetero.rda"
 
 fun <- function(pars, model, i){
 	HIVpars <- pars
-	if(i == 3 | i == 4){
+	if(i > 4){
 		pair = FALSE
 	}else{
 		pair = TRUE
@@ -51,7 +51,7 @@ fun <- function(pars, model, i){
 	}
 	cat("adj...")
 	val_vecFull[i] <<- find_scale(0.04, model, yini, HIVpars, interval=interval,
-															adjpar = "scale_all", full = TRUE, hmax = 0.5, tvec = c(1:200))
+															adjpar = "scale_all", full = TRUE, hmax = 0.5, tvec = 1:400, verbose = TRUE)
 	HIVpars_adj <- transform(HIVpars, scale_all = val_vecFull[i])
 	pp_adj <- expand(HIVpars_adj)
 	cat("sim...\n")

### 2017-01-09 figures/figures.R

@@ -1,5 +1,3 @@
-## setup
-library(knitr)
 ## requires ggplot 2.2.0
 library(ggplot2); theme_set(theme_bw(base_size = 12,
                                        base_family = "Times"))
@@ -33,6 +31,8 @@ source("../R/Param.R")
 
 ## load data
 load("../simdata/combineResults.rda")
+load("../simdata/ev_hetero.rda")
+load("../simdata/ev_bd.rda")
 
 ## FIXME: maybe should go in hivFuns.R ?
 
@@ -253,23 +253,58 @@ maxvir
 
 gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
 	geom_violinh(width=1)+
-	theme(legend.position = "none",
-				panel.spacing.y = grid::unit(0.8, "lines")) +
 	facet_wrap(~variable,scale="free_x", labeller = label_parsed)+
 	guides(fill=guide_legend(reverse=TRUE))+
-	labs(y="")+
-	geom_point(data=data.frame(model="random",
-                variable="equilibrium~mean~log[10]~SPVL",
-                                   value=rval),
-                   pch=22,size=3,show.legend=FALSE) +
-  zero_x_margin +
+	labs(y="")
+	
+
+random_point <- geom_point(data=data.frame(model="random",
+																					 variable="equilibrium~mean~log[10]~SPVL",
+																					 value=rval),
+													 pch=22,size=3,show.legend=FALSE)
+
+gg_univ_fig <- gg_univ + 
+	random_point +
+	zero_x_margin +
 	theme(panel.grid.major.x = element_blank(),
-				panel.grid.minor.x = element_blank())
+				panel.grid.minor.x = element_blank(),
+				legend.position = "none",
+				panel.spacing.y = grid::unit(0.8, "lines"))
 
 ## r fig3, fig.width=8,fig.height=4.8, echo = FALSE, cache = TRUE,dpi = 600}
 
-ggsave(gg_univ,file="fig3.pdf",width=8,height=4.8)
-if (do_png) ggsave(gg_univ,file="fig3.png",width=8,height=4.8,dpi=600)
+ggsave(gg_univ_fig,file="fig3.pdf",width=8,height=4.8)
+if (do_png) ggsave(gg_univ_fig,file="fig3.png",width=8,height=4.8,dpi=600)
+
+hetero <- seq(1, 7, 2)
+
+compare_df <- data.frame(
[…]

### 2017-01-10 ms/ms.tex
working on text/cover letter

@@
 progression through the three main stages of HIV infection (primary,
-asymptomatic, and disease), and we use a simple exponentially
+asymptomatic, and disease), and we use an exponentially
 distributed infectious period instead of a more realistic
@@
 where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
-of infection are independent of the host's SPVL.
-Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:
+of infection are independent of the host's SPVL
+(Table~\ref{table:parmsTable} gives definitions, units, and values for all parameters).
+. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:
 
@@
 to have a similar initial epidemic growth rate $r$, the peak
-\Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly
-higher (4.6 \Lspvl\ vs. 4.3 \Lspvl, or 7\% higher: \figurename~\ref{fig:panel3}).
+\Lspvl\ occurs at the same time ($\approx 200$ years) but slightly
+higher (4.6 \Lspvl\ vs. 4.3 \Lspvl: \figurename~\ref{fig:panel3}).
 

### 2017-01-10 ms/ms.tex
worked on figure captions

--- 2017-01-09 ms/ms.tex
@@
 \caption{{\bf Baseline dynamics.}
-Time series of mean population \Lspvl.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
+Time series of mean population \Lspvl\ using mean parameters.
+Each initial condition was varied independently while rest were kept constant in order to test their effects on the transient dynamics of virulence evolution.
+Note that the initial condition has negligible effect on equilibrium virulence.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$. achieved quicker.
 (b) Effects of varying initial infectious density $I(0)$.
@@
 The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories.
-Panels b and c illustrate the sensitivity of virulence trajectories to initial conditions $I(0)$ and $\alpha(0)$, which we hold constant in our simulations.}
+Although the virulence dynamics are sensitive to each initial condition, we are still able to achieve similar qualitative dynamics dynamics across a range of parameters. 
+For fair comparison, we hold them constant in our simulations.}
 \label{fig:panel3}
@@
 \includegraphics[width=\textwidth]{../figures/fig2.pdf}
-\caption{{\bf Envelopes of virulence trajectories (expected
-time of progression to AIDS) under all models.}
-All models were run until $t=4000~\textrm{years}$; truncated series are shown here.}
+\caption{{\bf Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
+All models were run until $t=4000~\textrm{years}$ using 1000 parameter sets obtained from Latin Hypercube Sampling to illustrate the range of dynamics each model can display.
+Envelope contains middle 95\% trajectories and lines within the envelope show mean trajectories.
+Truncated series are shown here.}
 \label{fig:virtraj}
@@
 \caption{{\bf Univariate distributions of summary statistics.}
-The distribution of equilibrium expected progression
-time (lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
+For every simulation using LHS parameters, three summary statistics were obtained: peak time, maximum mean \Lspvl, and equilibrium mean \Lspvl.
+Minimum mean progression time was obtained by by applying the hill function to maximum mean \Lspvl\ of each run.
+The distribution of equilibrium mean \Lspvl\
+(lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
 \label{fig:unidist}
@@
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
-Dashed line in equilibrium vs. peak virulence plot shows 1:1 line. 100 values were sampled from each model to allow for clearer distinction between the models}
+Three summary statistics obtained from each simulation have been plotted against each other to visualize the relationship among the summary statistics.
+All models follow a similar trend but each have distinguishing features.
+Surprisingly, the implicit model, an approximation for instantaneous partnership formation model (instswitch), shows an almost identical trend with a model that has pair formation dynamics (pairform).
+Dashed line in equilibrium vs. peak virulence plot shows 1:1 line, where equilibrium and peak virulence are equal.
+100 values were sampled from each model to allow for clearer distinction between the models}
 \label{fig:pairplot}
@@
 \caption{{\bf Sensitivity plot.}
-For most parameters in the Latin hypercube sample and each summary statistic, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
-prior to adjusting the parameters to achieve the standard initial epidemic growth rate.}
+For most parameters in the Latin hypercube sample and each summary statistic, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the parameter value, some of which have adjusted to achieve the standard initial epidemic growth rate (see Methods).}
+Opposing trends in summary statistics with respect to partnership dissolution rate, $c$, (third column) between the model with and without extra-pair contact illustrates that similar model with different assumptions may yield opposing predictions.
 \label{fig:plot_sens}

### 2017-01-10 ms/ms.tex
edit figure captions

--- 2017-01-10 ms/ms.tex
@@
 \caption{{\bf Baseline dynamics.}
-Time series of mean population \Lspvl.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
+Time series of mean population \Lspvl\ using baseline parameters.
+Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution.
+Note that the initial condition has negligible effect on equilibrium virulence.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$. achieved quicker.
 (b) Effects of varying initial infectious density $I(0)$.
@@
 The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories.
-Panels b and c illustrate the sensitivity of virulence trajectories to initial conditions $I(0)$ and $\alpha(0)$, which we hold constant in our simulations.}
+Although virulence dynamics are sensitive to initial conditions, we are still able to achieve similar qualitative dynamics dynamics across a range of parameters. 
+For fair comparison, we hold them constant throughout our simulations.}
 \label{fig:panel3}
@@
 \caption{Parameter ranges/values.
+Parameters with fixed values are kept constant throughout simulations while other parameter values are taken from Latin hypercube samples using ranges specified in the table.
 Values of $c$ and $\rho$ are doubled from those given by Champredon \etal\ because we keep track of 	individuals in the model, while they keep track of couples.
@@
 \includegraphics[width=\textwidth]{../figures/fig2.pdf}
-\caption{{\bf Envelopes of virulence trajectories (expected
-time of progression to AIDS) under all models.}
-All models were run until $t=4000~\textrm{years}$; truncated series are shown here.}
+\caption{{\bf Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
+All models were run until $t=4000~\textrm{years}$ using parameters from 1000 Latin hypercube samples to illustrate the range of dynamics each model can exhibit.
+Envelopes contain mid 95\% trajectories, while lines within the envelopes show mean trajectories.
+Truncated series (up to year 800) are shown here.}
 \label{fig:virtraj}
@@
 \caption{{\bf Univariate distributions of summary statistics.}
-The distribution of equilibrium expected progression
-time (lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
+For every simulation, three summary statistics were obtained: peak time, maximum mean \Lspvl, and equilibrium mean \Lspvl. 
+Peak time is the time at which a virulence trajectory achieves its maximum value (maximum mean \Lspvl).
+Equilibrium mean \Lspvl\ is taken by taking the mean \Lspvl\ value at $t = 4000$.
+Minimum mean progression time is obtained by applying the hill function to maximum mean \Lspvl\ of each run.
+The distribution of equilibrium mean \Lspvl\
+(lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
 \label{fig:unidist}
@@
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
-Dashed line in equilibrium vs. peak virulence plot shows 1:1 line. 100 values were sampled from each model to allow for clearer distinction between the models}
+Three summary statistics obtained from each simulation have been plotted against each other to visualize the relationship among the summary statistics and to allow for clearer comparison between models.
+Surprisingly, the implicit model, an approximation for instantaneous partnership formation model (instswitch), shows an almost identical trend with a model that has pair formation dynamics (pairform).
+Dashed line in equilibrium vs. peak virulence plot shows 1:1 line, where equilibrium and peak virulence are equal.
+100 values were sampled from each model to allow for clearer distinction between the models}
 \label{fig:pairplot}
@@
 \caption{{\bf Sensitivity plot.}
-For most parameters in the Latin hypercube sample and each summary statistic, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the \emph{unscaled} parameter value, i.e.
-prior to adjusting the parameters to achieve the standard initial epidemic growth rate.}
+For most parameters in the Latin hypercube samples, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the parameter value, some of which have adjusted to achieve the standard initial epidemic growth rate (see Methods).}
+Opposing trends in summary statistics with respect to partnership dissolution rate, $c$, (third column) between the model with and without extra-pair contact illustrates that similar model with different assumptions may yield opposing predictions.
 \label{fig:plot_sens}

### 2017-01-10 ms/cover_letter/plosHIVcover_resubm2.tex
working on text/cover letter
working on ms and response

@@
 
+\response{We do give a full table of parameters in the main text (Table 1); 
+however, we did not refer to it until the Latin hypercube sampling section --- %
+we now refer to it near the beginning of the Methods section.}
+
 \revcomment{
@@
 
-\response{comparing to Shirreff \ldots Figure 1 shows comparison \ldots picking high start virulence, won't see peak at all \ldots very much unknown \ldots implicit assumption that there's not an \emph{interaction} between starting SPVL and characteristics of models \ldots}
+\response{Thanks to this comment, we realized that we did not say enough in the main text about our choice of initial conditions.
+We have added a subsection in the Methods section about initial conditions; we have also extended our discussion of our sensitivity tests of initial conditions (emphasizing that the peak SPVL is relatively insensitive to starting conditions).
+}
 

### 2017-01-10 ms/ms.tex
moved table to the right place (right after its first citation)
working on ms and response

@@
 
-The uncoupled and extra-couple transmission rates (i.e., the rates of
-transmission among people outside of a stable partnership, or between
-people inside of a stable partnership and people other than their
-partner) are scaled by
-multiplying the within-couple transmission rate $\beta$ by the contact
-ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1). Simplifying the model
-of HIV pathogenesis from three stages to a single stage could affect
-our conclusions about the evolution of virulence (e.g. Kretzschmar and
-Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
-and multiple stages of infectivity have interactive effects on
-$\rzero$). However, our simplified model produces results that are
-qualitatively similar to those of Shirreff \etal's
-\cite{shirreff_transmission_2011} model; when our model is calibrated
-to have a similar initial epidemic growth rate $r$, the peak
-\Lspvl\ occurs at the same time ($\approx 200$ years) but slightly
-higher (4.6 \Lspvl\ vs. 4.3 \Lspvl: \figurename~\ref{fig:panel3}).
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig1.pdf}
-\caption{{\bf Baseline dynamics.}
-Time series of mean population \Lspvl\ using baseline parameters.
-Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution.
-Note that the initial condition has negligible effect on equilibrium virulence.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$. achieved quicker.
-(b) Effects of varying initial infectious density $I(0)$.
-(c) Effects of varying initial mean virulence $\alpha(0)$.
-The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories.
-Although virulence dynamics are sensitive to initial conditions, we are still able to achieve similar qualitative dynamics dynamics across a range of parameters. 
-For fair comparison, we hold them constant throughout our simulations.}
-\label{fig:panel3}
-\end{figure}
-
-\subsection*{Mutation}
-
-Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL.
-We simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping rather
-than allowing for variation in phenotypes of a single genotype.
-The mutational process in our model is directly taken from Shirreff \etal.
-Over the course of infection, mutation occurs within the host.
-However, it is assumed that SPVL of the strain transmitted by an infected individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation.
-Instead, the mutational effect takes place in a single
-step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
-\begin{equation}
-\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
-\end{equation}
-We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e.
-a bin width of 0.1 \Lspvl\ for $\alpha$), but reducing $n$ to 21 (bin width = 0.25 \Lspvl) makes little difference; we use this coarser grid for all other simulations reported.
-
-We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
-Finally, the probabilities are normalized so that each row sums to 1:
-\begin{equation}
-M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
-\end{equation}
-where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$.
-Transmission rate and disease induced mortality rates are discretized as well:
-\begin{equation}
-\begin{aligned}
-\beta_i &= \beta(\alpha_i),\\
-\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
-\end{aligned}
-\end{equation}
-
-\subsection*{Contact structure and partnership dynamics}
-
-We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} contact structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
-Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
-non-instantaneous partnership formation; (2) zero vs.
-positive extra-partnership sexual contact and transmission; and (3) homogeneous vs.
-heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
-
-Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}.
-The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e.
-individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status.
-$S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals.
-$SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples.
-The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission.
-
-The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch'').
-The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
-Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.
-Although these models can also be implemented
-by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.
-
-The fifth and sixth models represent extreme simplifications of sexual partnership dynamics.
-One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
-It is actually a random mixing model that explicitly tracks only the total number of susceptible and infected individuals.
-However, to reflect the effect of partnership structure, it uses an adjusted transmission rate derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}.
-The second model of this pair (``random'') is a simple random-mixing model.
-
-Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero'').
-Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
-We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
-While this model lacks some
-important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
-As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
-
-The pairform+epc and heterogeneous models use the basic epidemiological framework of Champredon \etal \cite{champredon_hiv_2013}.
-Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$.
-Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
-Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
-Extra-couple and uncoupled transmission are modeled in the same way as in Champredon \etal's model.
-All the details have been adapted to a multi-strain scenario, so that we track (for example) a matrix
-$II_{ij}$ that records the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$. 
-The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix S1).
-
-\subsection*{Latin hypercube sampling}
-
-Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
-In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
-For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
-We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
-
-Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
-\begin{equation}
-z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1, 2, 3, \dots,N.
-\end{equation}
-Random permutations of these vectors form columns in a sample parameter matrix; each row contains a different parameter set that is used for one simulation run.
-
-Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
-Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
-The remaining parameters are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
-
 %Tables need to be placed after the first paragraph in which they are cited.
@@
 
+The uncoupled and extra-couple transmission rates (i.e., the rates of
+transmission among people outside of a stable partnership, or between
+people inside of a stable partnership and people other than their
+partner) are scaled by
+multiplying the within-couple transmission rate $\beta$ by the contact
+ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1). Simplifying the model
+of HIV pathogenesis from three stages to a single stage could affect
+our conclusions about the evolution of virulence (e.g. Kretzschmar and
+Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
+and multiple stages of infectivity have interactive effects on
+$\rzero$). However, our simplified model produces results that are
+qualitatively similar to those of Shirreff \etal's
+\cite{shirreff_transmission_2011} model; when our model is calibrated
+to have a similar initial epidemic growth rate $r$, the peak
+\Lspvl\ occurs at the same time ($\approx 200$ years) but slightly
+higher (4.6 \Lspvl\ vs. 4.3 \Lspvl: \figurename~\ref{fig:panel3}).
+
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig1.pdf}
+\caption{{\bf Baseline dynamics.}
+Time series of mean population \Lspvl\ using baseline parameters.
+Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution.
+Note that the initial condition has negligible effect on equilibrium virulence.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
+(b) Effects of varying initial infectious density $I(0)$.
+(c) Effects of varying initial mean virulence $\alpha(0)$.
+The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
+The eco-evolutionary dynamics are qualitatively similar
+across a range of initial conditions (panels b,c).}
+\label{fig:panel3}
+\end{figure}
+
+\subsection*{Mutation}
+
+Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL.
+We simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping rather
+than allowing for variation in phenotypes of a single genotype.
+The mutational process in our model is directly taken from Shirreff \etal.
+Over the course of infection, mutation occurs within the host.
+However, it is assumed that SPVL of the strain transmitted by an infected individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation.
+Instead, the mutational effect takes place in a single
+step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
+\begin{equation}
+\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
+\end{equation}
+We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e.
+a bin width of 0.1 \Lspvl\ for $\alpha$), but reducing $n$ to 21 (bin width = 0.25 \Lspvl) makes little difference; we use this coarser grid for all other simulations reported.
+
+We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
+Finally, the probabilities are normalized so that each row sums to 1:
+\begin{equation}
+M_{ij} = \frac{\Phi(\alpha_j + d/2;i) - \Phi(\alpha_j - d/2;i)}{\Phi(\tsub{\alpha}{max} + d/2;i) - \Phi(\tsub{\alpha}{min} - d/2;i)},
+\end{equation}
+where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$.
+Transmission rate and disease induced mortality rates are discretized as well:
+\begin{equation}
+\begin{aligned}
+\beta_i &= \beta(\alpha_i),\\
+\lambda_i &= \frac{1}{D_P + D_A (\alpha_i) + D_D}.
+\end{aligned}
+\end{equation}
+
+\subsection*{Contact structure and partnership dynamics}
+
+We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} contact structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
+non-instantaneous partnership formation; (2) zero vs.
+positive extra-partnership sexual contact and transmission; and (3) homogeneous vs.
+heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
+
+Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}.
+The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e.
+individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status.
+$S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals.
+$SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples.
+The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission.
+
+The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch'').
+The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
+Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.
+Although these models can also be implemented
+by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.
+
+The fifth and sixth models represent extreme simplifications of sexual partnership dynamics.
+One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
+It is actually a random mixing model that explicitly tracks only the total number of susceptible and infected individuals.
+However, to reflect the effect of partnership structure, it uses an adjusted transmission rate derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}.
+The second model of this pair (``random'') is a simple random-mixing model.
+
+Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero'').
+Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
+We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
+While this model lacks some
+important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
+As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
+
+The pairform+epc and heterogeneous models use the basic epidemiological framework of Champredon \etal \cite{champredon_hiv_2013}.
+Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$.
+Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
+Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
+Extra-couple and uncoupled transmission are modeled in the same way as in Champredon \etal's model.
+All the details have been adapted to a multi-strain scenario, so that we track (for example) a matrix
+$II_{ij}$ that records the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$. 
+The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix S1).
+
+\subsection*{Initial conditions}
+
+Choosing the initial conditions for the simulations is something of a
+challenge. In the most common theoretical modeling contexts
+researchers are primarily interested in equilibria (or other long-term
+dynamical attractors such as limit cycles) and are exploring models
+that have a single stable attractor, so initial conditions can be
+ignored as long as we run models for long enough.  In eco-evolutionary
+dynamics, however, the initial conditions do affect our
+conclusions. We have no empirical information that would justify a
+particular quantitative choice of the fraction infected and the mean
+and variance of the distribution of SPVL at the point when the
+pandemic strain of HIV-1 entered the human population; in any case,
+the level of realism of our model does not support such a detailed
+consideration of the early dynamics of HIV.  In most cases, we started
+with an initial \Lspvl\ of 3.0, and an initial fraction infected of
+$10^{-4}$, to match the parameters used by Shirreff \etal; we ran one
+small set of sensitivity analyses with baseline values of the
+epidemiological parameters (Table~1) to test the effects of the
+initial conditions. Appendix TODO provides
+further details on initial conditions, such
+as the initial distribution of SPVL around the mean and the
+distribution of initial infected density across single people
+and different partnership types.
+
+\subsection*{Latin hypercube sampling}
+
+Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
+In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
+For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
+We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
+
+Latin hypercube sampling is done as in Champredon \etal\ \cite{champredon_hiv_2013}. For each parameter, $z$, its range is divided into $N = 1000$ equal intervals on a log scale:
+\begin{equation}
+z_i = \exp\left(\log(\tsub{z}{min}) + [\log(\tsub{z}{max}) - \log(\tsub{z}{min})] \frac{i-1}{N-1}\right) \qquad i = 1, 2, 3, \dots,N.
+\end{equation}
+Random permutations of these vectors form columns in a sample parameter matrix; each row contains a different parameter set that is used for one simulation run.
+
+Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
+Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
+The remaining parameters are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
+
 One parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see Appendix S1);
@@
 
-Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases.
-Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
-Decreasing the initial virulence
-also leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).
+Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory.
+Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
+Decreasing the initial SPVL also leads to progressively later, higher peaks in SPVL (\figurename~\ref{fig:panel3}c).
+In this case the delay in the peak is more pronounced than for low $I(0)$, because the rate of SPVL increase is eventually limited by the mutation rate.
+The peak SPVL is actually larger for a lower starting SPVL, presumably because lower SPVL also allows for a longer epidemic phase.
+However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV is able to evolve most of the way to the SPVL that would be optimal in an indefinitely prolonged growth phase.
 

### 2017-01-11 ms/cover_letter/plosHIVcover_resubm2.tex
still working on ms/response

@@
 
+\response{We have added some text here.}
+
 \revcomment{
@@
 
+\response{
+We agree that this causes some difficulties. Earlier drafts
+actually focused on SPVL as the output measure, but we thought
+that non-technical readers would have a better feeling for the
+real-world significance of progression time; furthermore, because
+progression time (unlike log(SPVL) is on a ratio scale, it becomes
+reasonable to quote results in terms of proportional changes. 
+We are taking the reviewer's advice and switching back to SPVL.
+}
+
 \revcomment{
@@
 And this change you’ve described is an INCREASE, not a decrease.
+}
+
+\response{
+We have changed the wording, and corrected a typo (the first value should have been 0.84 --- sorry!) As we say in our responses above, we have added more information about the choice of $r$, $I(0)$, $\alpha(0)$ in our sensitivity analyses (Figure 1).
 }

### 2017-01-11 ms/ms.tex
still working on ms/response

@@
 (c) Effects of varying initial mean virulence $\alpha(0)$.
-The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
+The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
 The eco-evolutionary dynamics are qualitatively similar
@@
 consideration of the early dynamics of HIV.  In most cases, we started
-with an initial \Lspvl\ of 3.0, and an initial fraction infected of
-$10^{-4}$, to match the parameters used by Shirreff \etal; we ran one
-small set of sensitivity analyses with baseline values of the
+with an initial \Lspvl\ of 3.0, to match the value used by
+Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
+we show results for this value in \figurename~\ref{fig:panel3}.
+Since some parameters in our sampling process (see next section)
+lead to very low equilibrium prevalence, we switched to 
+$I(0)=10^{-4}$ for the remainder of our analysis.
+We ran one small set of sensitivity analyses with baseline values of the
 epidemiological parameters (Table~1) to test the effects of the
-initial conditions. Appendix TODO provides
+initial conditions. Appendix~S1 provides
 further details on initial conditions, such
@@
 Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
+As we change $r$ from 0.42 to 0.84 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
 However, our model produces lower peak virulence ($\approx 4.3$ vs.

### 2017-01-11 ms/cover_letter/plosHIVcover_resubm2.tex
initial values etc.

@@
 \newcommand{\rzero}{\ensuremath{\mathcal R}_0}
+\newcommand{\todo}[1]{{\color \red \bf #1}}
 
@@
 
-\textbf{ADD GENERAL POINTS HERE}
+\todo{ADD GENERAL POINTS HERE}
 
@@
 Discussion rather than trying to squeeze it into the abstract
-\textbf{don't forget to do this}
+\todo{don't forget to do this}
 }
@@
 \response{
-We have changed the wording, and corrected a typo (the first value should have been 0.84 --- sorry!) As we say in our responses above, we have added more information about the choice of $r$, $I(0)$, $\alpha(0)$ in our sensitivity analyses (Figure 1).
+We have changed the wording, and corrected a typo (the second value should have been 0.042 --- sorry!) As we say in our responses above, we have added more information about the choice of $r$, $I(0)$, $\alpha(0)$ in our sensitivity analyses (Figure 1).
 }
@@
 Why are you changing the initial infectious density?
+}
+
+\response{As part of a sensitivity analysis, to determine the effect of $I(0)$ on the results (since we see that the results are relatively insensitive, we fix a value of $I(0)$ thereafter)
 }

### 2017-01-11 ms/ms.tex
initial values etc.

@@
 we show results for this value in \figurename~\ref{fig:panel3}.
-Since some parameters in our sampling process (see next section)
+Because we calibrated parameters based on the initial epidemic growth rate
+(see next section), we set $I(0)$
+
 lead to very low equilibrium prevalence, we switched to 
@@
 Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As we change $r$ from 0.42 to 0.84 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
+As we change $r$ from 0.042 to 0.084 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
 However, our model produces lower peak virulence ($\approx 4.3$ vs.

### 2017-01-11 ms/cover_letter/plosHIVcover_resubm2.tex
still working ...

@@
 \usepackage{times}
+\usepackage{color}
 \usepackage[utf8]{inputenc}
@@
 \newcommand{\rzero}{\ensuremath{\mathcal R}_0}
-\newcommand{\todo}[1]{{\color \red \bf #1}}
+\newcommand{\todo}[1]{{\color{red} \bf #1}}
 
@@
 
+\response{
+  Because the total evolutionary change in SPVL over the course of the growth phase is
+equal to the time until the epidemic peak (years) times the 
+speed of virulence evolution (change in SPVL/year); the faster
+evolution proceeds, the closer SPVL can get to its growth-phase optimum (i.e., to the
+SPVL that maximizes $r$).
+}
+
 \revcomment{
@@
 
+\response{
+\todo{diagram of models}
+}
+
 \revcomment{
@@
 Does not make sense to show summary statistics for scenarios that you know are unrealistic (because they would lead to epidemics with r values that do not match the observed value)
+}
+
+\response{
+For clarity, we have changed Figure 5 to show the scaled parameters.
 }
@@
 
+\response{
+
+}
+
 \revcomment{
@@
 
+\response{
+Thank you!
+}
+
 \textbf{Reviewer \#3:}
@@
 \response{(confusion between transmission rate and  ${\mathcal R}_0$ ---
-  clarify in text)}
+  clarify in text)
+}
 
@@
 There is nothing intrinsically "epidemiological" about forming (stable and/or once-off) sexual relationships.
+}
+
+\response{
+  We have changed this wording to ``contact structure'' throughout
 }
@@
 versus contact heterogeneity.
+}
+
+\response{
+  added \ldots
 }

### 2017-01-11 ms/ms.tex
still working ...

@@
 
-\newcommand{\todo}[1]{\textbf{#1}}
+\usepackage{color}
+\newcommand{\todo}[1]{\color{red} \textbf{#1}}
 
@@
 Because we calibrated parameters based on the initial epidemic growth rate
-(see next section), we set $I(0)$
-
-lead to very low equilibrium prevalence, we switched to 
-$I(0)=10^{-4}$ for the remainder of our analysis.
+(see ``Simulation runs'' below), we set $I(0)$ to $10^{-4}$ to ensure
+that the exponential growth phase lasted long enough for reliable
+estimation of the initial growth rate.
+
 We ran one small set of sensitivity analyses with baseline values of the
@@
 to higher virulence).
-Although the speed of virulence evolution varies, leading to wide variation in the minimum expected progression time (means ranging from approximately 6 to 12 years), virulence peaks in all models between 200 and 300 years.
+Although the speed of virulence evolution varies, leading to wide variation in the peak SPVL (ranging from \todo{XX} to \todo{YY})), virulence peaks in all models between 200 and 300 years.
 

### 2017-01-12 ms/ms.tex
fixing progresion time to Lspvl (and trying to merge)

--- 2017-01-11 ms/ms.tex
@@
 of partnerships. We assess summary statistics such as the peak virulence
-(corresponding to the minimum expected time of progression to AIDS) across
+(corresponding to the maximum value of population mean \Lspvl\ achieved throughout the epidemic) across
 models for a range of 
@@
 lower virulence and slower progression
-at the peak (a minimum of 15 years to
-progress to AIDS) compared both to more realistic models
+at the peak (a maximum of 3.5 \Lspvl) compared both to more realistic models
 and to simple random-mixing models with no partnership structure
-at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
+at all (both with a maximum of $\approx$ 4.7 years to \Lspvl).
 In this range of models, the simplest random-mixing structure best
@@
 partnerships. Models of intermediate complexity predicted lower
-virulence over the course of the epidemic (a minimum of 15 years to
-progress to AIDS) than either more realistic models or simple models
-without any partnership structure at all (both approx. 7.25 years to
-progress to AIDS), because sexual contact outside of stable
+virulence over the course of the epidemic (a maximum of 3.5 \Lspvl) than either more realistic models or simple models
+without any partnership structure at all (both approx. 4.7 \Lspvl), because sexual contact outside of stable
 partnerships tended to wash out the effects of contact
@@
 (which we denote as $\alpha$), rather than the rate of progression to
-AIDS itself (hereafter ``virulence'' will refer either to the SPVL, or
-to the rate of progression to AIDS; these two quantities are
-deterministically linked in the model).  
+AIDS itself (hereafter ``virulence'' will refer to \Lspvl).  
 In contrast to Shirreff \etal, we use a
@@
 Note that the initial condition has negligible effect on equilibrium virulence.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$. achieved quicker.
 (b) Effects of varying initial infectious density $I(0)$.
 (c) Effects of varying initial mean virulence $\alpha(0)$.
-The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
-The eco-evolutionary dynamics are qualitatively similar
-across a range of initial conditions (panels b,c).}
+The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories.
+Although virulence dynamics are sensitive to initial conditions, we are still able to achieve similar qualitative dynamics dynamics across a range of parameters. 
+For fair comparison, we hold them constant throughout our simulations.}
 \label{fig:panel3}
@@
 The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix S1).
-
-\subsection*{Initial conditions}
-
-Choosing the initial conditions for the simulations is something of a
-challenge. In the most common theoretical modeling contexts
-researchers are primarily interested in equilibria (or other long-term
-dynamical attractors such as limit cycles) and are exploring models
-that have a single stable attractor, so initial conditions can be
-ignored as long as we run models for long enough.  In eco-evolutionary
-dynamics, however, the initial conditions do affect our
-conclusions. We have no empirical information that would justify a
-particular quantitative choice of the fraction infected and the mean
-and variance of the distribution of SPVL at the point when the
-pandemic strain of HIV-1 entered the human population; in any case,
-the level of realism of our model does not support such a detailed
-consideration of the early dynamics of HIV.  In most cases, we started
-with an initial \Lspvl\ of 3.0, to match the value used by
-Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
-we show results for this value in \figurename~\ref{fig:panel3}.
-Since some parameters in our sampling process (see next section)
-lead to very low equilibrium prevalence, we switched to 
-$I(0)=10^{-4}$ for the remainder of our analysis.
-We ran one small set of sensitivity analyses with baseline values of the
-epidemiological parameters (Table~1) to test the effects of the
-initial conditions. Appendix~S1 provides
-further details on initial conditions, such
-as the initial distribution of SPVL around the mean and the
-distribution of initial infected density across single people
-and different partnership types.
 
@@
 Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As we change $r$ from 0.42 to 0.84 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
+As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
 However, our model produces lower peak virulence ($\approx 4.3$ vs.
@@
 
-Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory.
-Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
-Decreasing the initial SPVL also leads to progressively later, higher peaks in SPVL (\figurename~\ref{fig:panel3}c).
-In this case the delay in the peak is more pronounced than for low $I(0)$, because the rate of SPVL increase is eventually limited by the mutation rate.
-The peak SPVL is actually larger for a lower starting SPVL, presumably because lower SPVL also allows for a longer epidemic phase.
-However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV is able to evolve most of the way to the SPVL that would be optimal in an indefinitely prolonged growth phase.
+Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases.
+Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
+Decreasing the initial virulence
+also leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).
 

### 2017-01-12 R/simulations/figure1.R

@@ -38,7 +38,7 @@ get_rval <- function(g, yini, pp, plot.it=FALSE,
     }
 }
 
-HIVpars.shirreff <- transform(HIVpars.shirreff, ini_V = 3)
+HIVpars.shirreff <- transform(HIVpars.shirreff, ini_V = 3, kappa = 0, mu = 1)
 
 tvec = seq(from =1, to =600, by = 0.1)
 

### 2017-01-12 ms/ms.tex
changing to lspvl...

@@
 
-Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of the expected time of progression
-to AIDS (\figurename~\ref{fig:virtraj}: lower progression time corresponds
-to higher virulence).
-Although the speed of virulence evolution varies, leading to wide variation in the peak SPVL (ranging from \todo{XX} to \todo{YY})), virulence peaks in all models between 200 and 300 years.
+Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of population mean \Lspvl\ (\figurename~\ref{fig:virtraj}: higher population mean \Lspvl\ corresponds
+to higher virulence). Although the speed of virulence evolution varies, leading to wide variation in the peak SPVL (ranging from 3 to 5.5), virulence peaks in all models between 200 and 300 years.
 
@@
 
-Our chosen summary statistics (peak time, minimum expected progression time, equilibrium expected
-progression time, and relative progression time) all vary considerably across models
+Our chosen summary statistics (peak time, maximum mean \Lspvl, equilibrium mean \Lspvl, and minimum mean progression time) all vary considerably across models
 (\figurename~\ref{fig:unidist}).
@@
 pair formation without extra-pair contact. Some parameter
-sets for these models lead to low equilibrium virulence ($\approx 18$ years
-to progression);
+sets for these models lead to low equilibrium virulence ($\approx 2.3$ \Lspvl);
 these same sets lead to correspondingly low
-peak virulence (16 years to progression) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
-At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
+peak virulence (\Lspvl\ of 3 \todo{This is same as the initial value...
+how did you decide on this value in terms of progression time before?
+Peak time is 0 in this case...}) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
+At the opposite extreme, parameter sets that produce high equilibrium virulence (4.7 \Lspvl)
 also produce late peaks ($> 200~\text{years}$) and
-high peak virulence (4 years to progression).
+high peak virulence (5.6 \Lspvl).
 The pair-formation without extra-pair contact and implicit models
@@
 the board that they never exceed their initial virulence, leading to a tail
-of peak times near zero.
+of peak times near zero \todo{and also implying that we can observe a peak virulence lower than 3 \Lspvl\ if we start with a lower initial condition}.
 
@@
 scientifically important. The random-mixing, pairform+epc,
-and heterogeneous models all predict rapid
-progression to AIDS at the virulence peak
-(median/95\% CI = 6.1 (5.7-6.3), 
-6.02 (5.04-7.7), 6.03  (4.8-9.2)). 
+and heterogeneous models all predict high population mean \Lspvl\ at the virulence peak
+(median/95\% CI = 4.7 (4.65-4.79), 
+4.72 (4.37-4.96), 4.72  (4.09-5.03)). 
 In contrast, 
-the implicit model predicts minimum progression times about twice as long:
-12.5 (9.6-15.6) years. The corresponding differences in 
+the implicit model predicts a much lower peak \Lspvl\ value:
+3.52 (3-4.02) years.
+\todo{I'm not sure if you want to leave the transmission rates here or not...}
+\todo{I guess we can keep the progression time here because we have it in our plot?}
+The corresponding differences in 
 within-couple transmission
@@
 plots showing univariate summaries
-of \Lspvl\ and transmission probability).
+of expected progression time to AIDS and transmission probability).
 
@@
 different models with similar univariate dynamical summaries. While the
-relationship between equilibrium progression time and peak time is
+relationship between equilibrium \Lspvl\ and peak time is
 similar for all model structures (top left panel), the other
@@
 among people outside of stable partnerships --- tends to drive
-faster virulence evolution, leading to higher peak virulence and 
+faster virulence evolution, leading to higher peak virulence and \todo{I'm guessing you still want to keep progression time here?}
 lower times to progression at the peak time.

### 2017-01-12 ms/ms.tex
address TODOs about SPVL/progression switch

@@
 Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory.
-Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
+Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer exponential-growth phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
 Decreasing the initial SPVL also leads to progressively later, higher peaks in SPVL (\figurename~\ref{fig:panel3}c).
@@
 
-Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of population mean \Lspvl\ (\figurename~\ref{fig:virtraj}: higher population mean \Lspvl\ corresponds
-to higher virulence). Although the speed of virulence evolution varies, leading to wide variation in the peak SPVL (ranging from 3 to 5.5), virulence peaks in all models between 200 and 300 years.
+Across the entire range of parameters covered by the LHS analysis, all
+of the classes of models we considered produce qualitatively similar
+virulence trajectories, which we quantify in terms of population mean
+\Lspvl\ (\figurename~\ref{fig:virtraj}: higher population mean
+\Lspvl\ corresponds to higher virulence). Although the speed of
+virulence evolution varies, leading to wide variation in the peak SPVL
+(ranging from 3 to 5.5) because HIV can evolve farther toward
+its exponential-phase optimum before the transition to the
+endemic phase, virulence peaks in all models between 200 and
+300 years.
 
@@
 
-Our chosen summary statistics (peak time, maximum mean \Lspvl, equilibrium mean \Lspvl, and minimum mean progression time) all vary considerably across models
+Our chosen summary statistics (peak time, maximum mean \Lspvl, equilibrium mean \Lspvl, and minimum mean progression time) 
+all vary considerably across models
 (\figurename~\ref{fig:unidist}).
@@
 instantaneous-switching with and without extra-pair contact, and
-pair formation without extra-pair contact. Some parameter
-sets for these models lead to low equilibrium virulence ($\approx 2.3$ \Lspvl);
-these same sets lead to correspondingly low
-peak virulence (\Lspvl\ of 3 \todo{This is same as the initial value...
-how did you decide on this value in terms of progression time before?
-Peak time is 0 in this case...}) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
+pair formation without extra-pair contact. 
+Some parameter
+sets for these models lead to low equilibrium virulence ($2.3$-$3$ \Lspvl).
+For these data sets, virulence may either increase from its initial value,
+reaching an early peak ($\approx 200$ years) between 3 and 4 \Lspvl
+and then declining to a lower equilibrium value, or in extreme cases
+virulence may decline immediately, leading to a peak virulence (as we have
+defined it ) equal to the starting value of $\alpha(0)=3$ \Lspvl at $t=0$.
+(\figurename~\ref{fig:pairplot}).
 At the opposite extreme, parameter sets that produce high equilibrium virulence (4.7 \Lspvl)
@@
 high peak virulence (5.6 \Lspvl).
-The pair-formation without extra-pair contact and implicit models
-occasionally have parameter sets that select for such low virulence across
-the board that they never exceed their initial virulence, leading to a tail
-of peak times near zero \todo{and also implying that we can observe a peak virulence lower than 3 \Lspvl\ if we start with a lower initial condition}.
 
@@
 and heterogeneous models all predict high population mean \Lspvl\ at the virulence peak
-(median/95\% CI = 4.7 (4.65-4.79), 
+(median (95\% CI) = 4.7 (4.65-4.79), 
 4.72 (4.37-4.96), 4.72  (4.09-5.03)). 
-In contrast, 
-the implicit model predicts a much lower peak \Lspvl\ value:
+In contrast, the implicit model predicts a much lower peak \Lspvl\ value:
 3.52 (3-4.02) years.
-\todo{I'm not sure if you want to leave the transmission rates here or not...}
-\todo{I guess we can keep the progression time here because we have it in our plot?}
-The corresponding differences in 
-within-couple transmission
-probability are even more extreme, about a fourfold difference:
+In terms of mean time to progression to AIDS,
+The random-mixing, pairform+epc, and heterogeneous models predict rapid
+progression to AIDS at the virulence peak
+(median/95\% CI = 6.1 (5.7-6.3), 6.02 (5.04-7.7), 6.03  (4.8-9.2)),
+while the implicit model predicts minimum progression times about twice as long
+(12.5 (9.6-15.6) years).
+The corresponding differences in  mean within-couple transmission
+probability at the peak are even more extreme, about a fourfold difference:
 0.249 (0.24-0.26), 0.252 (0.19-0.28), and 0.252 (0.15-0.28) per year for the 
@@
 among people outside of stable partnerships --- tends to drive
-faster virulence evolution, leading to higher peak virulence and \todo{I'm guessing you still want to keep progression time here?}
+faster virulence evolution, leading to higher peak virulence and 
 lower times to progression at the peak time.

### 2017-01-12 ms/cover_letter/plosHIVcover_resubm2.tex
still working ...

@@
 
+\response{We have added detail; we believe the presentation of our methods can stand on its own
+(although we still relegate minor technical details to the supplementary materials, so as not
+to overwhelm the reader). We have also changed presentation more generally so that we don't lead by
+saying ``as in Shireff et al \ldots'', which might make readers think that they need to know Shirreff's
+material, although we still try to be clear where we are following Shirref et al/Champredon et al and
+where we are breaking new ground.}
+
 \revcomment{
@@
 
+\response{
+We have moved the later paragraph up as suggested, and have rewritten the relevant paragraphs to clarify them.
+}
+
 \revcomment{
@@
 Since the goal of the paper is to add more realism by including more realistic partnership dynamics, then this seems like any easy place to avoid removing realism.
+}
+
+\response{
+We made this assumption following Shirreff et al, for simplicity.
+We now describe the SIS assumption explicitly in the main text; we also constructed additional models with vital dynamics, and report on how this assumption changes our results (not much).
 }

### 2017-01-12 ms/ms.tex
still working ...

@@
 \usepackage{color}
-\newcommand{\todo}[1]{\color{red} \textbf{#1}}
+\newcommand{\todo}[1]{{\color{red} \textbf{#1}}}
 
@@
 shorter progression times (higher virulence).
-Studies of discordant couples --- stable
+Studies of discordant partnerships --- stable
 sexual partnerships with one infected and one uninfected partner --- %
@@
 models that incorporate the average effects of within-couple sexual
-contact --- without representing the explicit dynamics of pair
+contact --- without representing the explicit dynamics of partnership
 formation and dissolution or accounting for extra-partnership contact %
@@
 
-As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
-explicitly track the evolution of mean $\log_{10}$ SPVL
-(which we denote as $\alpha$), rather than the rate of progression to
+Our models
+explicitly track the evolution of the distribution of $\log_{10}$ SPVL
+(which we denote as $\alpha$)
+rather than the rate of progression to
 AIDS itself (hereafter ``virulence'' will refer to \Lspvl).
-In contrast to Shirreff \etal, we use a
-single-stage disease model instead of accounting explicitly for
-progression through the three main stages of HIV infection (primary,
-asymptomatic, and disease), and we use an exponentially
-distributed infectious period instead of a more realistic
+We use a single-stage model of HIV that assumes constant
+infectivity over the course of an exponentially
+distributed infectious period; this contrasts
+with Shirreff \etal's previous model which
+explicitly tracked three stages of HIV infection (primary,
+asymptomatic, and AIDS) and used a more realistic
 Weibull-distributed infectious period; we show below that our results
-are not overly sensitive to this simplification. We account for
-varying transmission rates and durations of each disease stage by
-summing the durations of three stages (again based on Shirreff \etal's
-model) and taking the duration-weighted average of transmission rates
-of three stages. Thus the within-couple transmission rate, $\beta$,
+are not overly sensitive to this simplification. 
+Shirreff \etal\ allowed the transmission rate
+and infection duration to depend on virus load only during the
+asymptomatic stage. In order to adapt their parameterization to our
+single-stage model, we used their estimates to compute the SPVL-dependent transmission
+rate and duration during the asymptomatic stage and 
+then derived the overall duration of the infectious period
+as the sum of the three stage durations and the 
+average transmission rate as the duration-weighted 
+average of the three stage-specific transmission rates. 
+Thus the within-couple transmission rate, $\beta$,
 for our models is given by:
@@
 
-Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL.
-We simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping rather
-than allowing for variation in phenotypes of a single genotype.
-The mutational process in our model is directly taken from Shirreff \etal.
 Over the course of infection, mutation occurs within the host. 
-However, it is assumed that SPVL of the strain transmitted by an infected individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation.
-Instead, the mutational effect takes place in a single
+However, we follow Shirreff \etal\ in assuming that SPVL of the strain transmitted by an infected 
+individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation. 
+Instead, the mutational effect is modeled as occurring in a single
 step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
@@
 where $\Phi(x;i)$ is the Gaussian cumulative distribution function with mean $\alpha_i$ and variance of $\sigma_M^2$, and $d = (\tsub{\alpha}{max} - \tsub{\alpha}{min})/(n-1)$. 
-Transmission rate and disease induced mortality rates are discretized as well:
+Unlike Shirreff \etal, who allowed for variation in the expressed phenotype (SPVL) of each genotype,
+we use a one-to-one genotype-phenotype map. Thus there is a single value for transmission rate and
+for progression rate corresponding to each SPVL compartment in the model:
 \begin{equation}
@@
 
-Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}.
-The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e.
-individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status.
-$S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals.
-$SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples.
-The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission.
-
-The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch'').
+Our first four models consider explicit partnership dynamics \cite{champredon_hiv_2013}.
+The first assumes non-instantaneous partnership formation (i.e.
+individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status; single (uncoupled) susceptible individuals ($S$), single infected individuals ($I$), concordant negative (susceptible-susceptible) couples ($SS$), discordant (susceptible-infected) couples ($SI$), and concordant positive (infected-infected) couples ($II$).
+The rates of pair formation are based on the numbers of uncoupled susceptible and infected individuals and the pair-formation rate; partnerships can either dissolve into singletons or be transformed into other types of partnerships by infection of one partner.
+This model also includes extra-partnership contact with both uncoupled individuals and individuals in other partnerships (we denote it ``pairform+epc''), so that susceptible uncoupled individuals and susceptible partners in any type of partnership can be infected by infected uncoupled individuals or infected partners in any type of partnership.
+
+\todo{DP: check that my description is still correct}
+Specifically, single individuals ($S$ and $I$) acquire partners at a rate $\rho$, and partnerships dissolve at a rate $c$.
+Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
+Infected individuals in seropositive ($II$) partnerships can also infect any susceptible individual at rate $c_e$.
+Likewise, single infected individuals ($I$) can infect any susceptible individuals (single individuals $S$, or susceptible members of $SS$ or $SI$ partnerships) at a rate $c_u$ through uncoupled mixing.
+This parameterization follows Champredon \etal; we have adapted some of the details
+of their model to a multi-strain scenario, so that we track (for example) a matrix 
+$II_{ij}$ that records the number of concordant, HIV-positive partnerships in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$.
+
+Our second model (``pairform'') only considers within-couple transmission, in which case infection can only occur within a serodiscordant partnership; that is, we set $c_e$ and $c_u$ to zero.
+
+Our third and fourth models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch'').
 The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
@@
 
-Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero'').
+Lastly, we add a model of heterogeneity in sexual activity to the models.
 Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
+(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \todo{give ref in supp material} presents results on the effect of adding heterogeneity to other model variants.)
 While this model lacks some 
@@
 
-The pairform+epc and heterogeneous models use the basic epidemiological framework of Champredon \etal \cite{champredon_hiv_2013}.
-Individuals in single compartment acquire a partner at a rate $\rho$, and partnerships dissolve at a rate $c$.
-Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
-Likewise, a single infected individual can infect any susceptible individuals at a rate $c_u$ through uncoupled mixing.
-Extra-couple and uncoupled transmission are modeled in the same way as in Champredon \etal's model.
-All the details have been adapted to a multi-strain scenario, so that we track (for example) a matrix
-$II_{ij}$ that records the number of concordant, HIV-positive couples in which the two partners have \Lspvl\ of $\alpha_i$ and $\alpha_j$. 
-The second through fourth models (pairform, instswitch+epc, instswitch) are derived from the base model by simplifying epidemiological processes (partnership formation and uncoupled/extra-couple contact: see Appendix S1).
+For simplicity (and following
+Shirreff \etal), all of our base models use an SIS (susceptible-infected-susceptible) formulation, where 
+there is no natural mortality (and no explicit introduction of newly sexually active individuals into
+the susceptible pool) individuals
+who die from AIDS are immediately replaced by an individual in the single susceptible compartment.
+While admittedly unrealistic, this approach is reasonable given that (1) the natural mortality rate is low relative to
+the epidemiological dynamics and (2) the infectious period is long, so that the overall rates of
+disease-induced mortality and recruitment of newly sexually active individuals would roughly
+balance. We also built models with vital dynamics where individuals dying from AIDS
+are removed from the population, with constant recruitment rates and constant low
+\emph{per capita} natural mortality rates; this additional structure had minor effects on
+the results.
 
@@
 to coupled individuals, the resulting unstructured mixing overwhelms
-the effect of structured mixing within couples, leading to mixing
+the effect of structured mixing within partnerships, leading to mixing
 that is effectively close to random.
@@
 David Champredon for access to simulation code
-and Wim Delva and an anonymous reviewer for useful comments;
+and David Earn, Wim Delva, and an anonymous reviewer for useful comments;
 this work was funded by NSERC Discovery Grant 386590-2010.

### 2017-01-12 ms/cover_letter/plosHIVcover_resubm2.tex
reorder models in figures; keep working on ms/response

@@
 
+\response{
+This is a great idea; we have reordered the model presentation in the figures so they do reflect a complexity series (i.e., starting with ``random'' and ending with ``hetero''.
+}
+
 \revcomment{
@@
 p. 11 Given the long time horizons of the simulations (500-4000
-years), this simplifying assumption [single-stage model of HIV
-progression] may introduce bias, but unfortunately the magnitude and importance thereof is unknown.
+years), this simplifying assumption [SIS model] may introduce bias, but unfortunately the magnitude and importance thereof is unknown.
 In more realistic models, one would expect that the high-risk susceptible individuals get depleted at a higher rate than the low risk susceptibles, with the epidemiological effects of this risk gradient playing out over a small number of decades.
 If or how this dynamic affects the evolution of virulence is unclear but if the goal is explicitly to study this evolution, it is problematic that all models under study are unable to take these demographic effects into account, by design.
+}
+
+\response{
+
 }

### 2017-01-12 ms/ms.tex
reorder models in figures; keep working on ms/response

@@
 \caption{{\bf Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
+Models arranged in order of complexity of contact structure, from ``random'' (least complex) to ``hetero'' (most complex).
 All models were run until $t=4000~\textrm{years}$ using parameters from 1000 Latin hypercube samples to illustrate the range of dynamics each model can exhibit.
-Envelopes contain mid 95\% trajectories, while lines within the envelopes show mean trajectories.
+Envelopes contain the middle 95\% of trajectories (i.e.
+we select all points between the 0.025 and 0.975 quantiles for each model at each year), while lines within the envelopes show mean trajectories.
 Truncated series (up to year 800) are shown here.}
@@
 \caption{{\bf Univariate distributions of summary statistics.}
+Models arranged in order of complexity, as in \figurename~\ref{fig:virtraj}.
 For every simulation, three summary statistics were obtained: peak time, maximum mean \Lspvl, and equilibrium mean \Lspvl. 
-Peak time is the time at which a virulence trajectory achieves its maximum value (maximum mean \Lspvl).
+Peak time is the time at which the SPVL reaches its maximum for each simulation run (maximum mean \Lspvl).
 Equilibrium mean \Lspvl\ is taken by taking the mean \Lspvl\ value at $t = 4000$.
-Minimum mean progression time is obtained by applying the hill function to maximum mean \Lspvl\ of each run.
-The distribution of equilibrium mean \Lspvl\
-(lower left panel) for the random mixing model is very narrow, and has been replaced by a point in order to preserve the vertical axis scaling.}
+Minimum mean progression time is obtained by applying the Hill function to maximum mean \Lspvl\ of each run.
+Because the distribution of equilibrium mean \Lspvl\
+(lower left panel) for the random mixing model is very narrow, it
+has been replaced by a point in order to preserve the vertical axis scaling.}
 \label{fig:unidist}

### 2017-01-12 figures/figures.R

@@ -42,8 +42,8 @@ new_sum_labs <- c(bquote(atop("peak time", "(years)")),
 									bquote(atop("peak", "mean" ~ log[10] ~ "SPVL")),
 									bquote(atop("equilibrium", "mean" ~ log[10] ~ "SPVL")),
                   "peak:equilibrium ratio")
-m_order <- c("random", "heterogeneous","pairform+epc", "pairform",
-             "instswitch+epc","instswitch" , "implicit")
+m_order <- c("heterogeneous","pairform+epc", "pairform",
+             "instswitch+epc","instswitch" , "implicit", "random")
 fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
     if (atop) new_sum_labs <-
                   gsub("(.*) (.*)","atop(\\1,\\2)",new_sum_labs)

### 2017-01-13 ms/S1_appendix.tex
added vital dynamics and edited hetero

@@
 
-We extend the ``pairform+epc'' model by allowing for heterogeneity in sexual behaviour.
+We extend the explicit models by allowing for heterogeneity in sexual behaviour.
+We first extend the ``pairform+epc'' model and obtain rest of the models by adjusting the heterogeneous ``pairform+epc'' model.
 Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group.
@@
 For simplicity, we assume that the transmission rate per partnership is unaffected by sexual behaviour.
+
+We allow for hetereogeneity in other three explicit models as well by adjusting the parameters of the heterogeneous ``pairform+epc model''.
+Heterogeneous ``pairform model'' is easily obtained by removing the extra-pair and uncoupled contact structure (setting $c_e/c_w = c_u/c_w = 0$, equivalently) from the heterogeneous ``pairform+epc'' model.
+We approximate heterogeneous ``instswitch'' and ``instswitch+epc'' models by setting the baseline pair formation rate, $\tsub{\rho}{base}$, to a high value to let individuals to enter partnership almost instantaneously as soon as they achieve a single state.
+For ``instswitch'' model, both $c_e/c_w$ and $c_u/c_w$ were set to 0, whereas only $c_u/c_w$ was set to 0 for ``instswitch+epc'' model.
+The following sections describe the heterogeneous ``pairform+epc'' model:
 
@@
 
+\subsection*{Vital dynamics}
+
+We extend four explicit models by adding vital dynamics and removing the SIS assumption).
+We follow Champredon \etal's \cite{champredon_hiv_2013} and assume that all individuals suffer from per-capita death rate of $m$ and that new susceptibles are born at a constant rate.
+Since we remove SIS assumption, individuals that die from disease induced mortality are simply removed from the population.
+Therefore, changes in the population due to birth and (both natural and disease induced) death in ``pairform+epc'' model can be written as follows (note that infection and partnership formation/dissolution processes have been ommitted in the equations for clarity):
+
+$$
+\begin{aligned}
+S' &= m(1 - S) + \sum_k \lambda_k SI_k + 2 m SS + \sum_k m SI_k \\
+I_i' &= - \lambda_i I_i + \sum_k 2^{\lambda_{ik}} \lambda_k II_{ik} - m I_i + m SI_k + \sum_k 2^{\lambda_{ik}} m II_{ik} \\
+SS' &= - 2 m SS \\
+SI_i' &= - \lambda_i SI_i - 2 m SI_i \\
+II_{ij}' &= - (\lambda_i + \lambda_j) II_{ij} - 2 m II_{ij} \\
+\end{aligned}
+$$
+
+By adding partnership dynamics and infection process, we obtain a ``pairform+epc'' model with vital dynamics.
+Other three explicit models with vital dynamics is obtained by adjusting the paraemters as discussed in the previous section.
+
 \subsection*{Initial distribution of infected individuals}
 
-We follow Champredon \etal's result to calculate the initial distribution of infected individuals.
+We follow Champredon \etal's result to calculate the initial distribution of infected individuals \cite{champredon_hiv_2013}.
 For model 1 and 2, we have disease equilibrium state of $S^* = \frac{c}{c + \rho}$ and $SS^* = \frac{1-S^*}{2}$.
@@
 $D_i$ is taken from the Gaussian distribution with mean 3 and standard deviation 0.2 and is normalized so that $\sum_i D_i = 1$.
-Then, we have the following initial distribution of each states:
+Then, we have the following initial distribution of an each state:
 
@@
 \end{equation}
+Note that we use this exact approach to calculate the initial distribution of an each state of the models with vital dynamics after redefining $S^* = \frac{c + 2m}{c + \rho + 2m}$.
+
 Since model 3 and 4 do not have single states, $SS^*=1$ at the disease free equilibrium and the initial distribution becomes:

### 2017-01-13 figures/diagram_geom.tex
attempt at tikz pictures!

\tikzset{
	auto,
	compartment/.style={
		rectangle, minimum size=9mm, rounded corners=2mm,
		thick, draw=black!15, top color=white,bottom color=black!30
	},
	%
	bigcompartment/.style={
		rectangle, minimum size=24mm, rounded corners=2mm,
		thick, draw=black!15, top color=white,bottom color=black!20
	},
	%
	point/.style={
		circle, inner sep=2pt, fill=black!5
	},
	%
	mytextbox/.style={
		rectangle, text=black!50, thin, 
		draw=white, top color=white,bottom color=white, fill=white
	}
}
% http://tex.stackexchange.com/questions/40388/avoid-arrow-crossing-over-blocks-in-flow-chart
\tikzstyle{line} = [draw, -latex']

### 2017-01-13 figures/instswitch.tex
attempt at tikz pictures!

\newcommand{\instswitchflowvec}{ 
	\draw[->, very thick] (SS) to (SI); 
	\draw[->, very thick] (SI) to (II); 
}

\newcommand{\instswitchnoepcinf}{
	\draw[->, blue, dashed, very thick] (SI)
		to [bend right=45] (SItoII);
	\draw[->, blue, dashed, very thick] (SI)
		to [bend left=45] (SStoSI);

}

\newcommand{\instswitchboxes}{
	\node(SS)[bigcompartment]{SS};
	\node(SStoSI)[right=of SS]{};
	\node(SI)[bigcompartment, right=of SStoSI]{SI};
	\node(SItoII)[right=of SI]{};
	\node(II)[bigcompartment, right=of SItoII]{II};
}

\newcommand{\pairformboxes}{
        \node(S)[bigcompartment]{S};
        \node(StoI)[right= of S]{};
        \node(I)[bigcompartment,right= of StoI]{I};
	\node(SS)[bigcompartment,below left= of S]{SS};
	\node(SStoSI)[right=of SS]{};
	\node(SI)[bigcompartment, right=of SStoSI]{SI};
	\node(SItoII)[right=of SI]{};
	\node(II)[bigcompartment, right=of SItoII]{II};
}

### 2017-01-13 figures/partnerSIR.tex
attempt at tikz pictures!

\newcommand{\instswitchflowvec}{ 
	\draw[->, very thick] (SS) to (SI); 
	\draw[->, very thick] (SI) to (II); 
}

\newcommand{\instswitchnoepcinf}{
	\draw[->, blue, dashed, very thick] (SI)
		to [bend right=45] (SItoII);
	\draw[->, blue, dashed, very thick] (SI)
		to [bend left=45] (SStoSI);
}

\newcommand{\instswitchepcinf}{
	\draw[->, red, dashed, very thick] (II)
		to [bend right=45] (SItoII);
	\draw[->, red, dashed, very thick] (II)
		to [bend left=70] (SStoSI);
}

\newcommand{\pairformnoepcflowvec}{ 
	\draw[->, very thick] (SI) to (II); 
	\draw[<->, very thick, dotted] (I) to (II); 
	\draw[<->, very thick, dotted] (S) to (SS); 
	\draw[<->, very thick, dotted] (S) to (SI); 
	\draw[<->, very thick, dotted] (I) to (SI); 
}

\newcommand{\pairformepcflowvec}{ 
	\draw[->, very thick] (S) to (I); 
	\draw[->, very thick] (SS) to (SI); 
}

\newcommand{\pairformnoepcinf}{
	\draw[->, blue, dashed, very thick] (SI)
		to [bend right=45] (SItoII);
}

\newcommand{\pairformepcinf}{
	\draw[->, red, dashed, very thick] (I)
		to [bend right=45] (StoI);
	\draw[->, red, dashed, very thick] (SI)
		to  (StoI);
	\draw[->, red, dashed, very thick] (SI)
		to [bend right=45] (SStoSI);
	\draw[->, red, dashed, very thick] (II)
		to [bend left=60] (SStoSI);
	\draw[->, red, dashed, very thick] (I)
		to  (SItoII);
	\draw[->, red, dashed, very thick] (II)
		to [bend left=70] (SItoII);
%% these are the problematic ones
% does this help?
%   http://tex.stackexchange.com/questions/209942/curved-arrows-in-tikz
% http://tex.stackexchange.com/questions/40388/avoid-arrow-crossing-over-blocks-in-flow-chart
        %\path (k) |- ([xshift=-0.5cm,yshift=-0.5cm]j.south west) |- (h)
        % \path[line, purple, very thick] (II) -- ([xshift=0.5cm,yshift=0.5cm]I.north east) -- ([yshift=0.5cm]I.north west) -- (StoI);
         \draw[->, red, dashed, very thick] plot [smooth, tension=0.4] coordinates
            { (II.north) ([xshift=0.5cm,yshift=0.5cm]I.north east) ([yshift=0.5cm]I.north west) ([yshift=0.5cm]StoI) (StoI) };
	\draw[->, green, dashed, very thick] (I)
		to [bend right=45] (SStoSI);

}

\newcommand{\instswitchboxes}{
	\node(SS)[bigcompartment]{SS};
	\node(SStoSI)[right=of SS]{};
	\node(SI)[bigcompartment, right=of SStoSI]{SI};
	\node(SItoII)[right=of SI]{};
	\node(II)[bigcompartment, right=of SItoII]{II};
}

\newcommand{\pairformboxes}{
        \node(S)[bigcompartment]{S};
        \node(StoI)[right= of S]{};
        \node(I)[bigcompartment,right= of StoI]{I};
	\node(SS)[bigcompartment,below left= of S]{SS};
	\node(SStoSI)[right=of SS]{};
	\node(SI)[bigcompartment, right=of SStoSI]{SI};
	\node(SItoII)[right=of SI]{};
	\node(II)[bigcompartment, right=of SItoII]{II};
}

### 2017-01-13 ms/S2_appendix.tex
worked on todo

@@
 
+\clearpage
+
+\begin{figure}[!ht]
+  \includegraphics[width=\textwidth]{../figures/fig_S2_4.pdf}
+\caption{{\bf Effects of contact heterogeneity and vital dynamics.}
+This figure matches Fig 3 in main text, but displays the
+summary statistics of the modified explicit model (models with contact heterogeneity and vital dynamics).
+Summary statistics of an each model is represented by shapes and all models were simulated using the baseline parameters.
+}
+\label{fig:fig3aug}
+\end{figure}
+
 \end{document}

### 2017-01-13 ms/S2_appendix.tex
tweaks to figure caption

--- 2016-10-26 ms/S2_appendix.tex
@@
 
+\clearpage
+
+\begin{figure}[!ht]
+  \includegraphics[width=\textwidth]{../figures/fig_S2_4.pdf}
+\caption{{\bf Effects of contact heterogeneity and vital dynamics.}
+This figure matches Fig 3 in main text, but displays the
+summary statistics of the modified explicit models (models with contact heterogeneity and vital dynamics).
+For each model, its summary statistics are plotted on the univariate plot using different shapes.
+All models were simulated using the baseline parameters.
+}
+\label{fig:fig3aug}
+\end{figure}
+
 \end{document}

### 2017-01-13 figures/partnerSIR.tex
dinking around with figs

@@
 \newcommand{\pairformepcinf}{
-	\draw[->, red, dashed, very thick] (I)
+	\draw[->, purple, dashed, very thick] (I)
 		to [bend right=45] (StoI);
@@
 		to [bend left=60] (SStoSI);
-	\draw[->, red, dashed, very thick] (I)
+	\draw[->, purple, dashed, very thick] (I)
 		to  (SItoII);
@@
 % http://tex.stackexchange.com/questions/40388/avoid-arrow-crossing-over-blocks-in-flow-chart
-        %\path (k) |- ([xshift=-0.5cm,yshift=-0.5cm]j.south west) |- (h)
-        % \path[line, purple, very thick] (II) -- ([xshift=0.5cm,yshift=0.5cm]I.north east) -- ([yshift=0.5cm]I.north west) -- (StoI);
          \draw[->, red, dashed, very thick] plot [smooth, tension=0.4] coordinates
-            { (II.north) ([xshift=0.5cm,yshift=0.5cm]I.north east) ([yshift=0.5cm]I.north west) ([yshift=0.5cm]StoI) (StoI) };
-	\draw[->, green, dashed, very thick] (I)
-		to [bend right=45] (SStoSI);
+            { (II.north) ([xshift=0.2cm,yshift=0.2cm]I.north east) ([xshift=-0.5cm,yshift=0.2cm]I.north west) (StoI) } (StoI);
+            \draw[->, purple, dashed, very thick] plot [smooth] coordinates {
+             ([yshift=0.3cm]I.west) ([xshift=0.5cm,yshift=0.5cm]S.north east)
+             ([xshift=-0.2cm,yshift=0.2cm]S.north west) 
+             ([xshift=0.5cm,yshift=0.5cm]SS.north east)
+             (SStoSI) } (SStoSI);
 

### 2017-01-13 ms/ms.tex

@@
 
-\todo{DP: check that my description is still correct}
 Specifically, single individuals ($S$ and $I$) acquire partners at a rate $\rho$, and partnerships dissolve at a rate $c$.
@@
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
-(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \todo{give ref in supp material} presents results on the effect of adding heterogeneity to other model variants.)
+(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; Figure S4 in Appendix S2 presents results on the effect of adding heterogeneity to other model variants.)
 While this model lacks some 

### 2017-01-13 figures/partnerSIR.tex
fix arrowheads

@@
          \draw[->, red, dashed, very thick] plot [smooth, tension=0.4] coordinates
-            { (II.north) ([xshift=0.2cm,yshift=0.2cm]I.north east) ([xshift=-0.5cm,yshift=0.2cm]I.north west) (StoI) } (StoI);
+            { (II.north) ([xshift=0.2cm,yshift=0.2cm]I.north east) 
+              ([xshift=-0.5cm,yshift=0.2cm]I.north west) 
+              ([yshift=0.25cm]StoI) }
+              to (StoI) ;
+
             \draw[->, purple, dashed, very thick] plot [smooth] coordinates {
@@
              ([xshift=0.5cm,yshift=0.5cm]SS.north east)
-             (SStoSI) } (SStoSI);
+              %% hack to get arrows in the right direction
+               ([yshift=0.25cm]SStoSI) } to (SStoSI);
 

### 2017-01-13 ms/cover_letter/plosHIVcover_resubm2.tex
dinking around with figs
TODO, more work on ms/cover

@@
 \response{
+We have implemented vital dynamics in our models and briefly discuss the impacts of this assumption in the Results.
+Overall, implementing vital dynamics increases the peak SPVL and mean SPVL (both by about 0.1 log10-SPVL units). 
+Time to peak remains similar, although it increases from about 250 to 300 years for the ``instswitch'' model.
 

### 2017-01-13 ms/ms.tex
TODO, more work on ms/cover

@@
 and to simple random-mixing models with no partnership structure
-at all (both with a maximum of $\approx$ 4.7 years to \Lspvl).
+at all (both with a maximum of $\approx$ 4.7 \Lspvl).
 In this range of models, the simplest random-mixing structure best
@@
 
-%% 199 words
+%% 196 words
 
 Pathogens such as HIV can evolve rapidly when the environment changes.
-Such changes include decreases in the probability that an infectious
-contact (such as a sneeze for a respiratory disease, or an unprotected
-sex act for a sexually transmitted disease) encounters an uninfected,
-susceptible person and thus has a chance to transmit the
-pathogen. Disease spread during an epidemic reduces the fraction of
-susceptible people in the population and thus changes the direction of
-parasite evolution.  While researchers have explored these
-evolutionary dynamics with simulations, their models have neglected
+One important aspect of a pathogen's environment is the probability that an infectious
+contact (a sneeze for a respiratory disease, or an unprotected
+sex act for a sexually transmitted disease) encounters an uninfected
+person and thus has a chance to transmit the
+pathogen. As an epidemic grows the number of
+uninfected people shrinks, changing
+evolutionary pressures on the pathogen.
+While researchers have used models to explore
+pathogen evolution during epidemics, their models usually neglect
 important processes such as people entering and leaving sexual
-partnerships. We compared several evolutionary models that include
-partnership dynamics and sexual contact outside of stable
+partnerships. We compared several evolutionary models for HIV that include
+partnership dynamics as well as sexual contact outside of stable
 partnerships. Models of intermediate complexity predicted lower
-virulence over the course of the epidemic (a maximum of 3.5 \Lspvl) than either more realistic models or simple models
-without any partnership structure at all (both approx. 4.7 \Lspvl), because sexual contact outside of stable
-partnerships tended to wash out the effects of contact
-structure. The large effect of epidemiological details on evolutionary
-dynamics suggests that researchers trying to predict the evolution of
-pathogens should proceed with caution.
+virulence midway through the epidemic
+(a minimum of 15 years to
+progress to AIDS) than either more realistic models
+or simple models with no partnership structure
+(both with a minimum of $\approx$ 7.25 years to progress to AIDS), 
+because random sexual contacts tended to wash out the effects of stable partnerships.
+Researchers trying to predict the evolution of
+pathogens must try to understand the implications of their
+modeling choices; models of intermediate complexity may not
+produce intermediate conclusions.
 
@@
 However, despite these studies' attention to detail at the individual
-or physiological level, the contact structures used in these
+or physiological level, the population-level contact structures used in these
 models are relatively simple.
@@
 --- or use an agent-based formulation with parameters that effectively
-lead to random mixing among infected and uninfected individuals. Here
+lead to random mixing among infected and uninfected individuals
+\todo{may not be true any more, Herbeck may have fixed this in their revised versions}. Here
 we explore the effects of incorporating \emph{explicit}
@@
 rather than the rate of progression to
-AIDS itself (hereafter ``virulence'' will refer to \Lspvl).
+AIDS itself (hereafter we use ``virulence'' to denote \Lspvl).
 We use a single-stage model of HIV that assumes constant
 infectivity over the course of an exponentially
-distributed infectious period; this contrasts
+distributed infectious period. This assumption contrasts
 with Shirreff \etal's previous model which
@@
 (Table~\ref{table:parmsTable} gives definitions, units, and values for all parameters).
-. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:
+Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:
 
@@
 
-We construct an $n$ by $n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
+We construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
 Finally, the probabilities are normalized so that each row sums to 1:
@@
 
-Lastly, we add a model of heterogeneity in sexual activity to the models.
-Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
+Lastly, we incorporate heterogeneity in sexual activity into the models.
+Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
@@
 there is no natural mortality (and no explicit introduction of newly sexually active individuals into
-the susceptible pool) individuals
-who die from AIDS are immediately replaced by an individual in the single susceptible compartment.
+the susceptible pool). Individuals
+who die from AIDS are immediately replaced by an individual in the uncoupled-susceptible compartment.
 While admittedly unrealistic, this approach is reasonable given that (1) the natural mortality rate is low relative to
@@
 disease-induced mortality and recruitment of newly sexually active individuals would roughly
-balance. We also built models with vital dynamics where individuals dying from AIDS
+balance. To check the importance of this assumption,
+we also built models with vital dynamics where individuals dying from AIDS
 are removed from the population, with constant recruitment rates and constant low
-\emph{per capita} natural mortality rates; this additional structure had minor effects on
+\emph{per capita} natural mortality rates; this additional structure had only minor effects on
 the results.
@@
 partners has much systematic effect in the heterogeneous model.
+
+\todo{discussion of effect of heterogeneity and vital dynamics here (referring to supp material)}
 

### 2017-01-14 figures/partnerSIR.tex
diagrams

@@
 \newcommand{\instswitchflowvec}{ 
+	\draw[->, very thick] (SI) to (II); 
+}
+
+\newcommand{\instswitchepcflowvec}{
 	\draw[->, very thick] (SS) to (SI); 
-	\draw[->, very thick] (SI) to (II); 
+}
+
+\newcommand{\instswitchmixing}{
+	\draw[<->, dotted, very thick] (SS) to (pool);
+	\draw[<->, dotted, very thick] (SI) to (pool);
+	\draw[<->, dotted, very thick] (II) to (pool);
 }
@@
 		to [bend right=45] (SItoII);
-	\draw[->, blue, dashed, very thick] (SI)
-		to [bend left=45] (SStoSI);
 }
@@
 \newcommand{\instswitchepcinf}{
+	\draw[->, red, dashed, very thick] (SI)
+		to [bend left=45] (SStoSI);
 	\draw[->, red, dashed, very thick] (II)
-		to [bend right=45] (SItoII);
+		to [bend left=45] (SItoII);
 	\draw[->, red, dashed, very thick] (II)
@@
 	\node(II)[bigcompartment, right=of SItoII]{II};
+	\node(pool)[compartment, very thick, above = 10mm of SI, text width = 40mm, align = center, draw = black, top color = white, bottom color = white, dotted]{temporary mixing pool};
 }

### 2017-01-14 ms/ms.tex
comment hetero code; regen figures; ms text

@@
 \newcommand{\DM}{\textrm{DM}}
+\newcommand{\diagfig}[1]{\figurename~\ref{fig:diagrams}#1}
 
@@
 average of the three stage-specific transmission rates. 
-Thus the within-couple transmission rate, $\beta$,
+Thus the within-couple transmission rate, $\beta$
+(see ``Contact Structure'' below)
 for our models is given by:
@@
 
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/diagrams_panel.pdf}
+\caption{{\bf Model contact structures.} 
+Schematic representations of models with
+explicit contact structure. Top row (a,b) shows models with instantaneous
+partnership formation; bottom row (c,d)  shows those where individuals remain
+single for some period of time between partnerships. Left column (a,c) 
+shows models with within-partnership transmission only; in these models
+transmission can only occur within serodiscordant partnerships, and
+serodiscordant partnerships can only be formed by dissolution
+and reformation of partnerships. Right column (b,d) shows models
+that allow extra-pair and uncoupled transmission.
+\emph{Solid arrows} represent infection transitions; \emph{dotted arrows}
+represent partnership dissolution and formation; \emph{dashed arrows}
+show influences on infection rate, including both within-pair
+transmission (blue) and extra-pair/uncoupled transmission (red).
+In addition to these models, we also use random-mixing and implicit
+models with only $S$ and $I$ compartments and heterogeneous models
+where each compartment is subdivided by sexual activity rate
+(of both partners, in the case of partnership compartments).
+}
+\label{fig:diagrams}
+\end{figure}
+
 We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} contact structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
@@
 
-Our first four models consider explicit partnership dynamics \cite{champredon_hiv_2013}.
-The first assumes non-instantaneous partnership formation (i.e.
-individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status; single (uncoupled) susceptible individuals ($S$), single infected individuals ($I$), concordant negative (susceptible-susceptible) couples ($SS$), discordant (susceptible-infected) couples ($SI$), and concordant positive (infected-infected) couples ($II$).
+Our first four models (\figurename~\ref{fig:diagrams}) explicitly consider partnership dynamics \cite{champredon_hiv_2013}.
+The first (\diagfig{d}) assumes non-instantaneous partnership formation (i.e.
+individuals spend some time uncoupled, outside of partnerships) and consists of five states that are classified by infection status and partnership status; single (uncoupled) susceptible individuals ($S$), single infected individuals ($I$), concordant negative (susceptible-susceptible) couples ($SS$), discordant (susceptible-infected) couples ($SI$), and concordant positive (infected-infected) couples ($II$).
 The rates of pair formation are based on the numbers of uncoupled susceptible and infected individuals and the pair-formation rate; partnerships can either dissolve into singletons or be transformed into other types of partnerships by infection of one partner.
@@
 
-Our second model (``pairform'') only considers within-couple transmission, in which case infection can only occur within a serodiscordant partnership; that is, we set $c_e$ and $c_u$ to zero.
+Our second model (``pairform'', \diagfig{c}) only considers within-couple transmission, in which case infection can only occur within a serodiscordant partnership; that is, we set $c_e$ and $c_u$ to zero.
 
@@
 The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
-Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.
+Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'', \diagfig{b}) includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'', \diagfig{a}) only considers within-couple transmission.
 Although these models can also be implemented

### 2017-01-14 R/simulations/ev_hetero.R

@@ -10,19 +10,23 @@ source("../hivModels.R")
 HIVpars.mean <- as.HIVvirparlist(apply(rbind(HIVpars_range, Het_range), 1, geom_mean))
 HIVpars.mean <- transform(HIVpars.mean, ini_V = 3)
 
-param_list <- rep(list(HIVpars.mean), 4)
+param_list <- setNames(rep(list(HIVpars.mean), 4),
+                       c("pairform+epc","pairform",
+                         "instswitch+epc","instswitch"))
 
-param_list[[3]]$rho_base <- param_list[[4]]$rho_base <- 4
+## set large pair-formation rate for 'instswitch' models
+param_list[["instswitch+epc"]]$rho_base <-
+    param_list[["instswitch"]]$rho_base <- 4
 
-param_list[[2]][c("c_u_ratio", "c_e_ratio")] <- 
-	param_list[[4]][c("c_u_ratio", "c_e_ratio")] <- 
+## set no extrapair/no-uncoupled contact for 'noepc' models
+extra_un_params <- c("c_u_ratio", "c_e_ratio")
+param_list[["pairform"]][extra_un_params] <- 
+	param_list[["instswitch"]][extra_un_params] <- 
 	c(0, 0)
 
-param_list[[3]]$c_u_ratio <- 0
+param_list[["instswitch+epc"]]$c_u_ratio <- 0
 
-names(param_list) <- c("full+epc", "full", "instswitch+epc", "instswitch")
-
-tvec <- c(1:1000)
+tvec <- 1:1000
 
 I_matFull <- matrix(NA, nrow = length(tvec), ncol = 8)
 vir_matFull <- matrix(NA, nrow = length(tvec), ncol = 8)
@@ -32,51 +36,53 @@ val_vecFull <- rep(0, 8)
 
 fn <- "ev_hetero.rda"
 
-fun <- function(pars, model, i){
-	HIVpars <- pars
-	if(i > 4){
-		pair = FALSE
-	}else{
-		pair = TRUE
-	}
-	pp <- expand(HIVpars, pair = pair)
-	yini <- calc_yini_full(pp)
-	cat("finding r...")
-	little_r <- get_rval(model, yini, pp, hmax = 0.5, tvec = c(1:200))
-	
-	if(little_r < 0.04){
-		interval <- c(1, 10)
-	}else{
-		interval <- c(0.01, 1)
-	}
-	cat("adj...")
[…]

### 2017-01-14 figures/figures.R

@@ -1,15 +1,17 @@
 ## requires ggplot 2.2.0
 library(ggplot2); theme_set(theme_bw(base_size = 12,
-                                       base_family = "Times"))
+                                      base_family = "Times"))
+stopifnot(packageVersion("ggplot2")>="2.2.0")
 library(directlabels)
+## stopifnot(length(grep("package:dplyr",search()))==0)
 library(MASS)  ## must be before dplyr (masks 'select')
 library(tidyr)
 library(gridExtra)
 library(plyr)  ## for ldply() -- MUST be before dplyr!
 library(dplyr) ## for full_join()
-## library(magrittr) ## for ???
 library(reshape2)  ## for melt()
 library(GGally) ## need BMB version
+stopifnot(packageVersion("GGally")>="1.3.0")
 ## devtools::install_github("bbolker/GGally")
 ## devtools::install_github("lionel-/ggstance")
 library(ggstance)
@@ -66,8 +68,8 @@ shirreff_df_m <- subset(res_sum_df,run==1,select=c(run,Time, Mean.VL))
 shirreff_df_m$run <- "Shirreff"
 df_tot <- rbind(shirreff_df_m,d_littleR_m)
 tlab <- "time (years)"
-g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run,linetype=run))+
-    geom_line()+labs(x=tlab,
+g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
+    geom_line(aes(linetype=run))+labs(x=tlab,
             y=expression(population~mean~set-point~viral~load~(log[10])))+
 		scale_x_continuous(expand = c(0.005, 0), breaks = seq(0, 600, by = 100), 
 											 labels = c("0", "", "200", "", "400", "", "")) +
@@ -277,29 +279,32 @@ ggsave(gg_univ_fig,file="fig3.pdf",width=8,height=4.8)
 if (do_png) ggsave(gg_univ_fig,file="fig3.png",width=8,height=4.8,dpi=600)
 
 hetero <- seq(1, 7, 2)
-
 compare_df <- data.frame(
 	model = c(rep(m_order[2:5], each = 2), m_order[2:5]),
-	type = c(rep(c("base", "bd"), 4), rep("hetero", 4)),
+	type = c(rep(c("base", "bd"), 4),
+                 rep("hetero", 4)),
[…]

### 2017-01-15 figures/partnerSIR.tex
tweak diagrams

@@
 
-\newcommand{\instswitchboxes}{
-	\node(SS)[bigcompartment]{SS};
+\newcommand{\instswitchboxes}[1]{
+	\node(SS)[bigcompartment, bottom color = #1]{SS};
 	\node(SStoSI)[right=of SS]{};
-	\node(SI)[bigcompartment, right=of SStoSI]{SI};
+	\node(SI)[bigcompartment, right=of SStoSI, bottom color = #1]{SI};
 	\node(SItoII)[right=of SI]{};
-	\node(II)[bigcompartment, right=of SItoII]{II};
-	\node(pool)[compartment, very thick, above = 10mm of SI, text width = 40mm, align = center, draw = black, top color = white, bottom color = white, dotted]{temporary mixing pool};
+	\node(II)[bigcompartment, right=of SItoII, bottom color = #1]{II};
+	\node(pool)[compartment, very thick, above = 10mm of SI, text width = 40mm, align = center, draw = black, bottom color = white, dotted]{temporary mixing pool};
 }
 
-\newcommand{\pairformboxes}{
-        \node(S)[bigcompartment]{S};
+\newcommand{\pairformboxes}[1]{
+        \node(S)[bigcompartment, bottom color = #1]{S};
         \node(StoI)[right= of S]{};
-        \node(I)[bigcompartment,right= of StoI]{I};
-	\node(SS)[bigcompartment,below left= of S]{SS};
+        \node(I)[bigcompartment,right= of StoI, bottom color = #1]{I};
+	\node(SS)[bigcompartment,below left= of S, bottom color = #1]{SS};
 	\node(SStoSI)[right=of SS]{};
-	\node(SI)[bigcompartment, right=of SStoSI]{SI};
+	\node(SI)[bigcompartment, right=of SStoSI, bottom color = #1]{SI};
 	\node(SItoII)[right=of SI]{};
-	\node(II)[bigcompartment, right=of SItoII]{II};
+	\node(II)[bigcompartment, right=of SItoII, bottom color = #1]{II};
 }

### 2017-01-15 figures/partnerSIR.tex
tweaking colors

@@
+\definecolor{blue2}{RGB}{61, 22, 218}
+\definecolor{red2}{RGB}{203, 18, 18}
+
 \newcommand{\instswitchflowvec}{ 
@@
 \newcommand{\instswitchnoepcinf}{
-	\draw[->, blue, dashed, very thick] (SI)
+	\draw[->, blue2, dashed, very thick] (SI)
 		to [bend right=45] (SItoII);
@@
 \newcommand{\instswitchepcinf}{
-	\draw[->, red, dashed, very thick] (SI)
+	\draw[->, red2, dashed, very thick] (SI)
 		to [bend left=45] (SStoSI);
-	\draw[->, red, dashed, very thick] (II)
+	\draw[->, red2, dashed, very thick] (II)
 		to [bend left=45] (SItoII);
-	\draw[->, red, dashed, very thick] (II)
+	\draw[->, red2, dashed, very thick] (II)
 		to [bend left=70] (SStoSI);
@@
 \newcommand{\pairformnoepcinf}{
-	\draw[->, blue, dashed, very thick] (SI)
+	\draw[->, blue2, dashed, very thick] (SI)
 		to [bend right=45] (SItoII);
@@
 \newcommand{\pairformepcinf}{
-	\draw[->, purple, dashed, very thick] (I)
+	\draw[->, red2, dashed, very thick] (I)
 		to [bend right=45] (StoI);
-	\draw[->, red, dashed, very thick] (SI)
+	\draw[->, red2, dashed, very thick] (SI)
 		to  (StoI);
-	\draw[->, red, dashed, very thick] (SI)
+	\draw[->, red2, dashed, very thick] (SI)
 		to [bend right=45] (SStoSI);
-	\draw[->, red, dashed, very thick] (II)
+	\draw[->, red2, dashed, very thick] (II)
 		to [bend left=60] (SStoSI);
-	\draw[->, purple, dashed, very thick] (I)
+	\draw[->, red2, dashed, very thick] (I)
 		to  (SItoII);
-	\draw[->, red, dashed, very thick] (II)
+	\draw[->, red2, dashed, very thick] (II)
 		to [bend left=70] (SItoII);
@@
 % http://tex.stackexchange.com/questions/40388/avoid-arrow-crossing-over-blocks-in-flow-chart
-         \draw[->, red, dashed, very thick] plot [smooth, tension=0.4] coordinates
+         \draw[->, red2, dashed, very thick] plot [smooth, tension=0.4] coordinates
             { (II.north) ([xshift=0.2cm,yshift=0.2cm]I.north east) 
@@
 
-            \draw[->, purple, dashed, very thick] plot [smooth] coordinates {
+            \draw[->, red2, dashed, very thick] plot [smooth] coordinates {
              ([yshift=0.3cm]I.west) ([xshift=0.5cm,yshift=0.5cm]S.north east)

### 2017-01-15 ms/ms.tex
bring diagram subfigure structure into main ms

@@
 \doublespacing
+
+\usepackage{subcaption}
 
@@
 
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/diagrams_panel.pdf}
+\begin{figure}[t!p]
+
+\begin{subfigure}[b]{0.4\linewidth}
+	\centering
+	\raisebox{0.44\height}{
+	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f1}
+	}
+\caption{Instantaneous switching without extra-pair contact (``instswitch'')}
+\end{subfigure}
+\begin{subfigure}[b]{0.4\linewidth}
+	\centering
+	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f2}
+\caption{Instantaneous switching with extra-pair contact (``instswitch+epc'')}
+\end{subfigure}
+
+\begin{subfigure}[b]{0.4\linewidth}
+	\centering
+	\raisebox{0.33\height}{
+	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f3}
+	}
+\caption{Pair-formation model without extra-pair/uncoupled contact 
+(``pairform'')}
+\end{subfigure}
+\begin{subfigure}[b]{0.4\linewidth}
+	\centering
+	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f4}
+\caption{Pair-formation model with extra-pair/uncoupled contact
+(``pairform+epc'')}
+\end{subfigure}
+
 \caption{{\bf Model contact structures.} 

### 2017-01-15 ms/ms.tex
various ms/bib/TODO tweaks

@@
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
-(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; Figure S4 in Appendix S2 presents results on the effect of adding heterogeneity to other model variants.)
+(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix S2 presents results on the effect of adding heterogeneity to other model variants.)
 While this model lacks some 
@@
 
-These differences are practically as well as 
-scientifically important. The random-mixing, pairform+epc,
+The random-mixing, pairform+epc,
 and heterogeneous models all predict high population mean \Lspvl\ at the virulence peak

### 2017-01-15 ms/ms.tex
minor tweaks/fixes to ms

--- 2017-01-15 ms/ms.tex
@@
 some of our models, we scaled partnership change rate and
-dissolution rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =
+transmission rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =
 \gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$,
@@
 
-Although each disease strain's core characteristic is its SPVL, the
-SPVL has one-to-one correspondences (based on eq.~\ref{eq:hillfuns})
-with both the expected time to progression to AIDS and with the rate
-(probability per unit time) of HIV transmission. Because the time to
-progression (measured in years) is easier to interpret than
-SPVL (measured in \Lspvl\ units), we summarize the virulence
-trajectories for each model run in terms of time to progression
-rather than SPVL. Because the  time to progression is inversely
-related to SPVL (increasing SPVL decreases the time to progression),
-the time to progression is technically measuring inverse
-virulence rather than virulence (we did not think that 
-reporting virulence as the rate of progression to AIDS, in units
-of $\textrm{years}^{-1}$, would help interpretability).
 For each model we derive the following summary statistics:
-minimum expected time to progression;
-time at which this minimum occurs
+maximum population mean \Lspvl;
+time at which this maximum occurs
 (corresponding to peak virulence --- this is also the time at which the
-maximum rate of progression, maximum SPVL, and maximum transmission rate
-occur); equilibrium time to progression; 
-and the ratio of progression time at its minimum to the equilibrium
-value. Equilibrium progression time is calculated after 4000 years of simulated
+maximum rate of progression and maximum transmission rate
+occur); equilibrium \Lspvl; 
+and minimum expected time to progression. 
+While the rest of the summary statistics measure \Lspvl, we also report minimum expected time to progression for clearer interpretability.
+Minimum mean progression time is obtained by applying the Hill function (eq.~\ref{eq:hillfuns}) to maximum mean \Lspvl\ of each run.
+Equilibrium \Lspvl\ is calculated after 4000 years of simulated
 time. 
@@
 
-Knowing the minimum progression time, timing of the minimum progression time/peak virulence, and equilibrium progression time provide sufficient detail to identify the overall shape of the virulence trajectory.
+Knowing the peak \Lspvl, timing of the peak\Lspvl/peak virulence, and equilibrium \Lspvl\ provide sufficient detail to identify the overall shape of the virulence trajectory.
 In particular, knowing the timing of the peak virulence (how many years
@@
 Equilibrium mean \Lspvl\ is taken by taking the mean \Lspvl\ value at $t = 4000$.
-Minimum mean progression time is obtained by applying the Hill function to maximum mean \Lspvl\ of each run.
 Because the distribution of equilibrium mean \Lspvl\
@@
 \caption{{\bf Sensitivity plot.}
-For most parameters in the Latin hypercube samples, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the parameter value, some of which have adjusted to achieve the standard initial epidemic growth rate (see Methods).}
+For most parameters in the Latin hypercube samples, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the parameter value, some of which have adjusted to achieve the standard initial epidemic growth rate (see Methods).
 Opposing trends in summary statistics with respect to partnership dissolution rate, $c$, (third column) between the model with and without extra-pair contact illustrates that similar model with different assumptions may yield opposing predictions.
+}
 \label{fig:plot_sens}

### 2017-01-18 ms/S2_appendix.tex
very small comment and tweaks to supp file

@@
 \caption{{\bf Envelopes of transmission trajectories under all models.}
-This figure matches Fig 2 in main text, but displays the
-envelope of population-mean transmission probabilities rather than expected progression time to AIDS over time
+This figure matches Fig 3 in main text, but displays the
+envelope of population-mean transmission probabilities rather than mean \Lspvl\ over time
 for each model.
@@
 \includegraphics[width=\textwidth]{../figures/fig_S2_2.pdf}
-\caption{{\bf Envelopes of SPVL trajectories under all models.}
-This figure matches Fig 2 in main text, but displays the
-envelope of \Lspvl\ rather than expected progression time to AIDS over time
+\caption{{\bf Envelopes of expected progression time to AIDS trajectories under all models.}
+This figure matches Fig 3 in main text, but displays the
+envelope of expected progression time to AIDS rather than mean \Lspvl\ over time
 for each model.
 }
-\label{fig:virtraj}
+\label{fig:durtraj}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{../figures/fig_S2_3.pdf}
-\caption{{\bf Univariate distributions of transmission probabilities and SPVL.}
-This figure matches Fig 3 in main text, but displays the
-univariate distributions for the transmission probability and \Lspvl\ time at the virulence
+\caption{{\bf Univariate distributions of transmission probabilities and expected progression time to AIDS.}
+This figure matches Fig 4 in main text, but displays the
+univariate distributions for the transmission probability and expected progression time to AIDS time at the virulence
 peak and at the epidemiological equilibrium,
-rather than the distributions of progression time to AIDS.
+rather than the distributions of \Lspvl.
 }
-\label{fig:tranvirsum}
+\label{fig:trandursum}
 \end{figure}
@@
 \caption{{\bf Effects of contact heterogeneity and vital dynamics.}
-This figure matches Fig 3 in main text, but displays the
+This figure matches Fig 4 in main text, but displays the
 summary statistics of the modified explicit models (models with contact heterogeneity and vital dynamics).

### 2017-01-26 ms/ms.tex
very small comment and tweaks to supp file
knit pdf
working ...

@@
 \newcommand{\diagfig}[1]{\figurename~\ref{fig:diagrams}#1}
-
+\newcommand{\peryear}{year$^{-1}$}
 \usepackage{color}
@@
 Our models combine HIV virulence tradeoffs with a range of
-contact models, modeling partnership formation and
+contact models, explicitly modeling partnership formation and
 dissolution and allowing for individuals to transmit disease outside
@@
 or simple models with no partnership structure
-(both with a minimum of $\approx$ 7.25 years to progress to AIDS), 
+(both with a minimum of 7.25 years to progress to AIDS), 
 because random sexual contacts tended to wash out the effects of stable partnerships.
@@
 and multiple stages of infectivity have interactive effects on
-$\rzero$). However, our simplified model produces results that are
-qualitatively similar to those of Shirreff \etal's
-\cite{shirreff_transmission_2011} model; when our model is calibrated
-to have a similar initial epidemic growth rate $r$, the peak
-\Lspvl\ occurs at the same time ($\approx 200$ years) but slightly
-higher (4.6 \Lspvl\ vs. 4.3 \Lspvl: \figurename~\ref{fig:panel3}).
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig1.pdf}
-\caption{{\bf Baseline dynamics.}
-Time series of mean population \Lspvl\ using baseline parameters.
-Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution.
-Note that the initial condition has negligible effect on equilibrium virulence.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
-(b) Effects of varying initial infectious density $I(0)$.
-(c) Effects of varying initial mean virulence $\alpha(0)$.
-The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
-The eco-evolutionary dynamics are qualitatively similar
-across a range of initial conditions (panels b,c).}
-\label{fig:panel3}
-\end{figure}
+$\rzero$).
 
@@
 Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
-we show results for this value in \figurename~\ref{fig:panel3}.
-Because we calibrated parameters based on the initial epidemic growth rate
-(see ``Simulation runs'' below), we set $I(0)$ to $10^{-4}$ to ensure
+because we calibrated parameters based on the initial epidemic growth rate
+(see ``Simulation runs'' below), we set $I(0)$ to $10^{-4}$
+for most runs to ensure
 that the exponential growth phase lasted long enough for reliable
@@
 
-We ran one small set of sensitivity analyses with baseline values of the
-epidemiological parameters (Table~1) to test the effects of the
-initial conditions. Appendix~S1 provides
+Appendix~S1 provides
 further details on initial conditions, such
@@
 and different partnership types.
+
+\subsection*{Sensitivity analyses}
+
+We ran most of our models across a wide range of parameters, as described
+in the next section. In several cases, however, we inspected only a few
+cases, to qualitatively assess the sensitivity of the models to
+initial conditions or to model structure. 
+In particular, we tested model sensitivity 
+to the initial prevalence $I(0)$ and initial mean SPVL
+$\alpha(0)$, using baseline values of all parameters (Table~\ref{table:parmsTable}).
+Using baseline parameter values,
+we also ran all four basic model structure (\figurename~\ref{fig:diagrams})
+with vital dynamics and with heterogeneity in sexual contact,
+to assess the sensitivity of our results to these phenomena.
 
@@
 every run has the same initial exponential growth rate in 
-disease incidence.
+disease incidence. \todo{prevalence?}
 
@@
 and dissolution rate) or transmission rate fails this requirement for
-some of our models, we scaled partnership change rate and
+some of our models, we scaled both partnership change rate and
 transmission rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =
@@
 
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figures/fig1.pdf}
+\caption{{\bf Baseline dynamics.}
+Time series of mean population \Lspvl\ using baseline parameters.
+Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution.
+Note that the initial condition has negligible effect on equilibrium virulence.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
+(b) Effects of varying initial infectious density $I(0)$.
+(c) Effects of varying initial mean virulence $\alpha(0)$.
+The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
+The eco-evolutionary dynamics are qualitatively similar
+across a range of initial conditions (panels b,c).}
+\label{fig:panel3}
+\end{figure}
+
 Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As we change $r$ from 0.042 to 0.084 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
-However, our model produces lower peak virulence ($\approx 4.3$ vs.
-$\approx 4.6$ \Lspvl)
-and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
+As we calibrate the parameters to achieve initial growth rates $r$ from 0.042 \peryear\ to 0.084 \peryear\ (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
+For matching initial growth rates ($r=0.042$)
+the peak \Lspvl\ occurs at the same time ($\approx 200$ years)
+but the peak virulence is lower than 
+Shirreff's ($\approx 4.3$ vs. $\approx 4.6$ \Lspvl),
+as is equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl).
 
@@
 \caption{{\bf Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
-Models arranged in order of complexity of contact structure, from ``random'' (least complex) to ``hetero'' (most complex).
-All models were run until $t=4000~\textrm{years}$ using parameters from 1000 Latin hypercube samples to illustrate the range of dynamics each model can exhibit.
+Models arranged in order of complexity of contact structure, from 
+``hetero'' (most complex) to ``random'' (least complex).
+All models were run until $t=4000~\textrm{years}$ using parameters from 1000 Latin hypercube samples to illustrate the range of possible dynamics.
 Envelopes contain the middle 95\% of trajectories (i.e.
@@
 
-\todo{discussion of effect of heterogeneity and vital dynamics here (referring to supp material)}
+Finally, incorporating additional realism to the model, i.e.
+combining heterogeneity with all four basic contact structures or
+allowing for vital dynamics rather than assuming an SIS model,
+leads to only small differences in the conclusions stated so far
+(\figurename~??). Relative to our baseline SIS assumption, the
+effect of adding vital dynamics is to delay the virulence peak
+slightly and increase both the peak and equilibrium virulence.
+The changes are small, however: across all models, the
+maximum increase in time until
+the virulence peak is 40 years (for the instswitch+epc model),in 
+the peak \Lspvl\ is 0.24 units (instswitch), and in the equilibrium
+\Lspvl\ is 0.4 (pairform). The changes in the most realistic
+model (pairform+epc with heterogeneity)
 
@@
 \paragraph*{Appendix S2: Dynamics of Transmission and Virulence}
-
 \label{S2_Appendix}

### 2017-01-26 figures/figures.R

@@ -1,9 +1,9 @@
+
 ## requires ggplot 2.2.0
 library(ggplot2); theme_set(theme_bw(base_size = 12,
                                       base_family = "Times"))
 stopifnot(packageVersion("ggplot2")>="2.2.0")
 library(directlabels)
-## stopifnot(length(grep("package:dplyr",search()))==0)
 library(MASS)  ## must be before dplyr (masks 'select')
 library(tidyr)
 library(gridExtra)
@@ -16,7 +16,9 @@ stopifnot(packageVersion("GGally")>="1.3.0")
 ## devtools::install_github("lionel-/ggstance")
 library(ggstance)
 
+## set accordingly ...
 do_png <- FALSE
+do_pdf <- FALSE
 
 if (.Platform$OS.type=="windows") {
     windowsFonts(Times=windowsFont("Times"))
@@ -128,7 +130,7 @@ g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
 
 ## r fig1,fig.height=3.5,fig.width=10, echo = FALSE, dpi = 600}
 fig1 <- arrangeGrob(g1.y, g2.y, g3.y, nrow=1)
-ggsave("fig1.pdf", fig1, height = 3.5, width = 10)
+if (do_pdf) ggsave("fig1.pdf", fig1, height = 3.5, width = 10)
 
 if (do_png) {
     png(file="fig1.png",height=3.5*600,width=10*600)
@@ -184,7 +186,7 @@ gg_virtraj <-
 
 ## FIXME: make minimal example for direct labels/linetype warning?
 
-ggsave(gg_virtraj,file="fig2.pdf",width=8,height=6)
+if (do_pdf) ggsave(gg_virtraj,file="fig2.pdf",width=8,height=6)
 if (do_png) ggsave(gg_virtraj,file="fig2.png",width=8,height=6,dpi=400)
 
 fig_objects <- c(fig_objects,"gg_virtraj","bottom.points2","top.points2")
@@ -196,13 +198,13 @@ ss_beta <- transform(ss, mean = returnBeta(mean,HIVpars.skeleton),
                       upr = returnBeta(upr,HIVpars.skeleton))
 
 gg_betatraj <- direct.label(gg_basetraj %+% ss_beta + nostrips +
-           labs(y=expression(mean~transmission~rate~during~asymptomatic~stage~(year^-1))),
[…]

### 2017-01-26 ms/cover_letter/plosHIVcover_resubm2.tex
various ms/bib/TODO tweaks
still working ...

@@
 \response{
-\todo{diagram of models}
+This is a great idea.  We have added a figure that shows four diagrams (one for each of the central models), with a colour scheme matching the colours used to denote models in later figures.
 }
@@
 \response{
-
+We agree that our analysis is not by itself enough to assess 
+the eco-evolutionary dynamics of HIV in a setting that is realistic
+enough to apply in a public policy context. However, we still feel
+it represents a valuable contribution to the HIV literature, by
+providing guidance to other modelers on both how and why 
+eco-evolutionary conclusions can depend on model structure.
 }
@@
 \response{
-  added \ldots
+We extended our analysis to apply both heterogeneity and vital dynamics
+(separately) to all four of the base models (pairform/instswitch with
+and without extra-pair contact); we report our findings in the
+last paragraph of the Results section and in Figure S4 (Supplementary material 2).
 }
@@
 scientifically important''] seems to belong better in the discussion section. Here, the authors do not explain what the practical or scientific importance of the differences is.
+}
+
+\response{
+We removed this sentence.
 }
@@
 
+\response{
+
+}
+
 \closing{Sincerely,}
 
-%\bibliography{virulence}
+\bibliography{virulence}
 

### 2017-01-26 ms/ms.tex
still working ...

@@
 \begin{table}[h!]
-\caption{Parameter ranges/values.
+\caption{\textbf{Parameter ranges and values}.
 Parameters with fixed values are kept constant throughout simulations while other parameter values are taken from Latin hypercube samples using ranges specified in the table.
@@
 
-\caption{{\bf Model contact structures.} 
+\caption{\textbf{Model contact structures.} 
 Schematic representations of models with
@@
 \includegraphics[width=\textwidth]{../figures/fig1.pdf}
-\caption{{\bf Baseline dynamics.}
+\caption{\textbf{Baseline dynamics.}
 Time series of mean population \Lspvl\ using baseline parameters.
-Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution.
-Note that the initial condition has negligible effect on equilibrium virulence.
 (a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
@@
 The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
-The eco-evolutionary dynamics are qualitatively similar
-across a range of initial conditions (panels b,c).}
+The other panels (b,c) show that eco-evolutionary dynamics are qualitatively similar across a range of initial conditions.}
 \label{fig:panel3}
@@
 \includegraphics[width=\textwidth]{../figures/fig2.pdf}
-\caption{{\bf Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
-Models arranged in order of complexity of contact structure, from 
-``hetero'' (most complex) to ``random'' (least complex).
+\caption{\textbf{Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
+Panels are arranged in order of complexity of contact structure, from 
+``random'' (least complex) to ``hetero'' (most complex).
 All models were run until $t=4000~\textrm{years}$ using parameters from 1000 Latin hypercube samples to illustrate the range of possible dynamics.
@@
 \Lspvl\ is 0.4 (pairform). The changes in the most realistic
-model (pairform+epc with heterogeneity)
+model (pairform+epc) are considerably smaller: 
+an increase of 2 years vs. a decrease of 13 years in the
+time to the virulence peak for the
+models with vital dynamics and heterogeneity, respectively;
+an increase of 0.1 units vs. a decrease of 0.01 units in peak \Lspvl;
+and an increase of 0.145 units vs. a decrease of 0.03 units
+in equilibrium~\Lspvl. Thus, while we can never rule out the
+possibility of some higher-order interaction among epidemiological
+phenomena leading to significant changes in our conclusions,
+we are reasonably confident that the results reported here are
+robust to additional complexities.
 

### 2017-01-26 figures/figures.R

@@ -18,7 +18,7 @@ library(ggstance)
 
 ## set accordingly ...
 do_png <- FALSE
-do_pdf <- FALSE
+do_pdf <- TRUE
 
 if (.Platform$OS.type=="windows") {
     windowsFonts(Times=windowsFont("Times"))
@@ -48,6 +48,7 @@ new_sum_labs <- c(bquote(atop("peak time", "(years)")),
                   "peak:equilibrium ratio")
 m_order <- c("heterogeneous","pairform+epc", "pairform",
              "instswitch+epc","instswitch" , "implicit", "random")
+
 fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
     if (atop) new_sum_labs <-
                   gsub("(.*) (.*)","atop(\\1,\\2)",new_sum_labs)
@@ -148,7 +149,7 @@ ss$model <- factor(ss$model,levels=m_order)
 
 ss <- transform(ss,
                 cmodel=droplevels(factor(gsub("+epc","",model,fixed=TRUE),
-                                         levels=levels(model))))
+                                         levels=rev(levels(model)))))
 
 ## base plot for Figure 2 and related (duration, SPVL, beta trajectories)
 gg_basetraj <- ggplot(ss,aes(tvec,mean,ymin=lwr,ymax=upr))+
@@ -280,10 +281,12 @@ if (do_pdf) ggsave(gg_univ_fig,file="fig3.pdf",width=8,height=4.8)
 if (do_png) ggsave(gg_univ_fig,file="fig3.png",width=8,height=4.8,dpi=600)
 
 hetero <- seq(1, 7, 2)
+basemodels <- grep("(instswitch|pairform)",m_order,value=TRUE)
+nm <- length(basemodels)
 compare_df <- data.frame(
-	model = c(rep(m_order[2:5], each = 2), m_order[2:5]),
-	type = c(rep(c("base", "bd"), 4),
-                 rep("hetero", 4)),
+	model = c(rep(basemodels, each = 2), basemodels),
+	type = c(rep(c("base", "bd"), nm),
+                 rep("hetero", nm)),
 	eq_vir = c(eq_vec_bd, eq_vecFull[hetero])) %>%
 	bind_cols(rbind(peak_mat_bd, peak_matFull[hetero,]) %>% 
 	     as.data.frame %>% 
@@ -624,7 +627,7 @@ ggsens <- ggplot(mL3,aes(LHSval,sumval,colour=model))+
[…]

### 2017-01-27 figures/figures.R
still working on figures (reverse legends, panel spacing etc.)

@@ -446,8 +446,11 @@ tweak_legends_gg <- function(gg,pos=1.5) {
         for (j in 1:n) {
             inner <- getPlot(gg,i,j)
             if (i==1 & j==1) {
+                newguide <- guide_legend(override.aes = list(size=4),
+                                                 reverse=TRUE)
                 inner <- inner + ggplot2::theme(legend.position=c(pos,0.5)) +
-                    guides(colour = guide_legend(override.aes = list(size=6)))
+                    guides(colour = newguide,
+                           shape = guide_legend(reverse=TRUE))
             } else {
                 inner <- inner + ggplot2::theme(legend.position="none")
             }
@@ -456,6 +459,7 @@ tweak_legends_gg <- function(gg,pos=1.5) {
     }
     return(gg)
 }
+
 tweak_colours_gg <- function(gg) {
     n <- gg$nrow
     for (i in 1:n) {
@@ -481,10 +485,7 @@ sL2 <- sL2 %>%
 ggp1 <- ggpairs(sL2,
         mapping = ggplot2::aes(color = model,pch=model),
         columns=3:5,
-        ## axisLabels="show",
-        ## legends=TRUE,
         lower = list(continuous = wrap("points",alpha=0.6,size=2)),
-        ## alpha = 0.3,size=0.5)),
         diag = list(continuous = "blankDiag"),
         upper = list(continuous = "blank"),
         labeller=label_parsed,
@@ -492,11 +493,9 @@ ggp1 <- ggpairs(sL2,
                          "'peak time'~('years')",
                          "'peak mean SPVL'~(log[10])"),
         switch="both"
-        ## columnLabels = expression(equilibrium~mean~SPVL~(log[10]),
-        ## peak~time~(years),
-        ## peak~mean~SPVL~(log[10]))
         )+
-    ## theme_gray()+
+    ## blank strip boxes to make strip labels look like
+    ##  regular axis labels
     theme(strip.background = element_rect(fill = NA, colour=NA),
           strip.placement = "outside",
           panel.border = element_rect(fill = NA)
@@ -508,25 +507,28 @@ ggp1 <- ggpairs(sL2,
 ggp2 <- tweak_colours_gg(tweak_legends_gg(trim_gg(ggp1,
[…]

### 2017-01-27 figures/figures.R
work on figures

@@ -522,7 +522,9 @@ for(i in 1:2) {
 ## ``{r fig4,fig.width=7,fig.height=7, echo = FALSE, cache = TRUE,dpi = 600}
 
 if (do_pdf) pdf(file="fig4.pdf",width=7,height=7)
-theme_set(theme_bw()+theme(panel.spacing=grid::unit(0,"lines")))
+orig_theme <- theme_get()
+theme_set(orig_theme+
+          theme(panel.spacing=grid::unit(0,"lines")))
 print(ggp2)
 if (do_pdf) dev.off()
 
@@ -535,6 +537,7 @@ if (do_png) {
 
 fig_objects <- c(fig_objects,"ggp2")
 
+theme_set(orig_theme)
 ### Figure 5
 remove_runs <- unique(subset(mL3, sumvar == "peak_time" & is.na(sumval))[,"run"])
 mL3 <- subset(mL3, !(model == "heterogeneous" & run %in% remove_runs))

### 2017-01-27 figures/figures.R
update S2_4 with duration panel

@@ -225,7 +225,8 @@ if (do_png) ggsave(gg_durtraj,file="fig_S2_2.png",width=6,height=4,dpi=400)
 
 ### Figure 3
 
-sL <- transform(sum_list[["sum_mat"]], peak_dur = returnDur(peak_vir,HIVpars.skeleton))
+sL <- transform(sum_list[["sum_mat"]],
+                peak_dur = returnDur(peak_vir,HIVpars.skeleton))
 sL$model <- factor(sL$model, m_order)
 
 ## we get long tails because eq_vir has default value of 0.
@@ -239,12 +240,14 @@ mL <- na.omit(melt(sL,id.vars=c("model","run")))
 w <- with(mL,which(model=="random" & variable== "eq_vir"))
 rval <- mean(mL$value[w])
 mLw <- droplevels(mL[-w,])
+var_levels <- c("peak_time", "peak_vir", "eq_vir", "peak_dur")
+var_labels <- c("peak~time~(years)",
+                "peak~mean~log[10]~SPVL",
+                "equilibrium~mean~log[10]~SPVL",
+                "minimum~mean~progression~time~(years)")
 mLw$variable <- factor(mLw$variable,
-         levels = c("peak_time", "peak_vir", "eq_vir", "peak_dur"),
-         labels = c("peak~time~(years)",
-                    "peak~mean~log[10]~SPVL",
-										"equilibrium~mean~log[10]~SPVL",
-         					  "minimum~mean~progression~time~(years)"))
+         levels = var_levels, 
+         labels = var_labels)
 
 maxvir <- mL %>%
 	group_by(model) %>%
@@ -291,20 +294,18 @@ compare_df <- data.frame(
 	bind_cols(rbind(peak_mat_bd, peak_matFull[hetero,]) %>% 
 	     as.data.frame %>% 
              setNames(c("peak_time", "peak_vir"))) %>%
+    mutate(peak_dur = returnDur(peak_vir,HIVpars.skeleton)) %>%
 	gather(data, key, -model, -type) %>%
 	setNames(c("model", "type", "variable", "value"))
 
 ## FIXME: match columnLabels from above ...
 compare_df$variable <- factor(compare_df$variable,
-											 levels = c("peak_time", "peak_vir", "eq_vir"),
-											 labels = c("peak~time~(years)",
-											            "peak~mean~log[10]~SPVL",
-                                                                                                    "equilibrium~mean~log[10]~SPVL"))
+                              levels = var_levels,
[…]

### 2017-01-27 ms/S2_appendix.tex
update fig S4 caption, etc

@@
 \caption{{\bf Effects of contact heterogeneity and vital dynamics.}
-This figure matches Fig 4 in main text, but displays the
-summary statistics of the modified explicit models (models with contact heterogeneity and vital dynamics).
-For each model, its summary statistics are plotted on the univariate plot using different shapes.
-All models were simulated using the baseline parameters.
+This figure matches Fig 4 in main text, but adds results
+for model variants with contact heterogeneity and vital dynamics.
+The violin plots, copied from Fig 4 in the main text, show the
+distribution of outcomes across the Latin hypercube sample
+for SIS model without contact heterogeneity. The points show
+the outcome of various models using the baseline parameters
+from Table 1: (\emph{circle}) unmodified 
+models (SIS, homogeneous); (\emph{triangle}) models with vital dynamics; 
+(\emph{square}) models with contact heterogeneity.
 }

### 2017-01-27 ms/ms.tex
update fig S4 caption, etc

@@
 every run has the same initial exponential growth rate in 
-disease incidence. \todo{prevalence?}
+disease prevalence.
 
@@
 leads to only small differences in the conclusions stated so far
-(\figurename~??). Relative to our baseline SIS assumption, the
+(\figurename~S4). Relative to our baseline SIS assumption, the
 effect of adding vital dynamics is to delay the virulence peak
@@
 
-%In Herbeck \etal's \cite{herbeck_hiv_2014} network model of partnerships, the partnership duration is set to 1 day --- very unrealistic in epidemic terms, but perhaps
-%actually more true to real-world HIV epidemiological dynamics than a
-%model with realistic partnership durations that neglects extra-pair
-%contact \cite{herbeck2016evolution}. 
 Taking further steps to make the model even more realistic

### 2017-01-27 ms/cover_letter/plosHIVcover_resubm2.tex
update fig S4 caption, etc
work on cover letter

@@
 \response{
-
+We have thoroughly revised the captions.
 }
@@
 }
+\response{
+
+}
 
@@
 It would be very helpful to be able to give a intuitive or at least qualitative explanation for why these mixing patterns have such an influence on virulence.
+}
+\response{
+
 }
@@
 \response{
-
+We agree with the reviewers that this is a ``rather theoretical'' finding,
+but we think it is important and useful nonetheless.  As we mention
+in the discussion, the decision to examine the effects of contact
+structure does not mean we can feel free to ignore all other aspects
+of epidemiological realism; on the other hand, data for
+parameterization and time for building and checking models is
+limited, so we have to prioritize in some way. We would argue that
+both quantifying the effects of contact structure and understanding
+how they work is a useful, albeit incremental, step toward reliable
+models of HIV eco-evolutionary dynamics.
 }

### 2017-02-01 ms/cover_letter/plosHIVcover_resubm2.tex
working on cover letter

@@
 
-\todo{ADD GENERAL POINTS HERE}
+We are happy to have the chance to revise our manuscript
+(and thank you for your patience while we've taken a little
+longer than average to do so). Besides general clarification,
+some of the main changes are that we
+\begin{itemize}
+  \item added a figure illustrating the main model structures we use
+  \item switched our primary descriptor of viral evolution back to SPVL, as recommended by reviewer \#1
+  \item implemented vital dynamics and heterogeneity in conjunction
+with each of our four 'base' models (i.e., models with and 
+without instantaenous switching, with and without extra-pair contact); 
+the detailed results for these models are given in the supplementary materials, but
+we do comment on the results in the main text.
+\end{itemize}
+
+We look forward to your decision on the revised version.
 
@@
 \hrule
-
-\textbf{Editor}
-
-\revcomment{
-In particular, we would encourage you to take into consideration the remarks of reviewer 1 on the presentation of your work.
-Furthermore, reviewer 3 raises important points regarding the limitation of the ODE equation approach.
-While you already discuss some of these points, it would be great if you put your study into the context of more realistic modeling approaches of HIV and other sexually transmitted diseases.
-It would also be insightful if you discussed to what extent your "heterogeneous" model covers age and sex structure.
-}
 
@@
 }
-%\response{}
+\response{This is a very general comment, but we have gone
+through the entire manuscript and tried to clarify it.}
 
@@
 how they work is a useful, albeit incremental, step toward reliable
-models of HIV eco-evolutionary dynamics.
+models of HIV eco-evolutionary dynamic.
 }

### 2017-02-03 ms/ms.tex
add to discussion

@@
 
+How contact structures are modeled can strongly affect
+researchers' conclusions about the evolutionary dynamics of
+virulence. In particular, a relatively simple (strategic) 
+eco-evolutionary model of HIV can predict peak $\log_{10}$
+set-point viral loads (over the course of an epidemic)
+ranging from 3.5 to 4.8 depending on the specific model
+of sexual partnership behaviour used. This difference in \Lspvl\ is
+epidemiologically significant, corresponding to a twofold difference
+(12 vs. 6 years) in expected time to progression.
+
+Surprisingly, the biggest differences are not between the simplest
+and the most complex contact structures, but between the
+complex contact structures and models of \emph{intermediate}
+complexity. These intermediate-complexity models attempt,
+quite reasonably, to add at least some of the realism of
+human sexual behaviour, but leave out apparently insignificant
+details such as extra-pair contact. If partial complexity
+may lead to such mistakes, how can modelers do anything but always
+strive to build the most realistic models possible?
+
 All models must simplify the world.  Many constraints --- among them data

### 2017-02-07 ms/ms.tex
still working ...

@@
 One of the most difficult parts of model comparison is finding
-parameter sets that are commensurate with many different model
-structures. For the most part, our models are too complex to easily
-derive analytical correspondences among them. Given a numerical
+parameter sets that are commensurate across radically different model
+structures. For the most part, our models are too complex to
+derive analytical correspondences among the parameters
+for different models. Given a numerical
 criterion, such as $r$ (initial exponential growth rate) or $\rzero$ 
@@
 
-For parameters involved in partnership turnover ($c$ and $\rho$), main distinction is between models with and without extra-pair contact.
-With extra-pair contact, we get gradual decrease in peak time, maximum \Lspvl, and equilibrium \Lspvl\ with increasing turnover rates; Otherwise we get increase in summary statistics.
-Although we have not verified this idea analytically, we speculate that increase in turnover rates diminishes the effect of epc (equivalently increasing relative importance of within-partnership contact), therefore selecting for lower \Lspvl.
-For models without epc, increased rates of partnership change decrease the level of structured mixing (mimicing epc models), resulting in selection for higher \Lspvl.
+For parameters involved in partnership turnover ($c$ and $\rho$), the main distinction is between models with and without extra-pair contact.
+With extra-pair contact, we see a gradual decrease in peak time, maximum \Lspvl, and equilibrium \Lspvl\ with increasing turnover rates.
+Increases in the other parameters lead to increases in all
+three summary statistics.
+We interpret these patterns as evidence
+that the increase in turnover rates diminishes the effect of extra-pair contact (equivalently increasing relative importance of within-partnership contact), therefore selecting for lower \Lspvl.
+For models without extra-pair contact, increased rates of partnership change decrease the level of structured mixing (mimicing extra-pair contact models), resulting in selection for higher \Lspvl.
 
@@
 
-Surprisingly, the biggest differences are not between the simplest
-and the most complex contact structures, but between the
+The restriction of transmission within stable partnerships
+strongly limits eco-evolutionary dynamics by
+limiting the maximum speed of epidemic growth. An HIV genotype
+that optimizes SPVL to maximize the speed of spread in
+a homogeneous population will be sub-optimal in a context
+where infection can only spread beyond a partnership once
+it dissolves. This finding echoes a long line of
+studies that show that population structure leads to the
+evolution of ``prudent'' parasites, although most of these studies
+focus on equilibrium optima rather than eco-evolutionary dynamics
+\citep{herre_factors_1995,kerr_local_2006,webb_theoretical_2013,berngruber_spatial_2015}. 
+The more complex contact structures we modeled
+mitigate these constraints by allowing HIV to spread among uncoupled individuals (through finite pair-formation) and members of stable partnerships
+(through extra-pair contact), albeit at lower rates than within
+partnerships.
+
+Thus, we see the biggest differences not between the simplest
+and the most complex contact structures, which either
+ignore pair structure completely or allow for extra-pair
+contact that reduces its impact, but between the
 complex contact structures and models of \emph{intermediate}
@@
 quite reasonably, to add at least some of the realism of
-human sexual behaviour, but leave out apparently insignificant
-details such as extra-pair contact. If partial complexity
+human sexual behaviour, but err by neglecting the apparently insignificant
+detail of extra-pair contact. If partial complexity
 may lead to such mistakes, how can modelers do anything but always

### 2017-02-08 ms/cover_letter/plosHIVcover_resubm2.tex
add to discussion
still working ...
a few more changes

@@
 with each of our four 'base' models (i.e., models with and 
-without instantaenous switching, with and without extra-pair contact); 
+without instantaneous switching, with and without extra-pair contact); 
 the detailed results for these models are given in the supplementary materials, but
@@
 \response{
-
+We have added several paragraphs at the beginning of the discussion.
 }
@@
 \response{
+We discussed these issues some in the paragraph starting ``The most striking aspect of the univariate comparisons ...'' (p.
+15 of previous ms., p.
+18 of current version).
+The second paragraph of the Discussion now goes into more depth.
 
@@
 
-\response{(confusion between transmission rate and  ${\mathcal R}_0$ ---
-  clarify in text)
+\response{
+We have clarified this sentence by explicitly framing ``reproduce more'' as a net increase in lifetime fitness, i.e. increasing $\rzero$.
 }

### 2017-02-08 ms/ms.tex
a few more changes

@@
 
-As we discuss in detail below,
-existing models of HIV eco-evolutionary dynamics either use implicit
+Many existing models of HIV eco-evolutionary dynamics use implicit
 models that incorporate the average effects of within-couple sexual
 contact --- without representing the explicit dynamics of partnership
-formation and dissolution or accounting for extra-partnership contact %
---- or use an agent-based formulation with parameters that effectively
-lead to random mixing among infected and uninfected individuals
-\todo{may not be true any more, Herbeck may have fixed this in their revised versions}. Here
-we explore the effects of incorporating \emph{explicit}
-contact structure in eco-evolutionary models.
+formation and dissolution or accounting for extra-partnership contact;
+agent-based formulations are more realistic. Here we explore the
+effects of incorporating explicit contact structure in
+eco-evolutionary models.
 
@@
 focus on equilibrium optima rather than eco-evolutionary dynamics
-\citep{herre_factors_1995,kerr_local_2006,webb_theoretical_2013,berngruber_spatial_2015}. 
+\cite{Herre1995,Kerr+2006,webb_theoretical_2013,berngruber_spatial_2015}. 
 The more complex contact structures we modeled
@@
 
-Taking further steps to make the model even more realistic
-might add further structure,
-making the random-mixing model predictions less accurate. For
-example, our model forms partnerships randomly, and assumes that
-extra-pair contact is randomly mixing across the population;
-one could instead model extra-pair contact as arising from
+Taking further steps to make the model even more realistic might add
+further structure, making the random-mixing model predictions less
+accurate. For example, our model forms partnerships randomly, and
+assumes that extra-pair contact is randomly mixing across the
+population; one could instead model extra-pair contact as arising from
 multiple concurrent partnerships (some, such as contact with sex
@@
 formation (by age, ethnicity, or behaviour group). The effects of
-other realistic complications such as explicit modeling of two
-sexes (both in contact structure and differential transmission
+other realistic complications such as explicit modeling of two sexes
+(both in contact structure and differential transmission
 probabilities), temporal and spatial variation in epidemic processes,
 or presence of genetic variation in hosts are harder to predict.
+In contrast, the elevated viral load in the early stage of HIV infection,
+neglected in our model, will likely lead to higher maximum
+epidemic growth rates and allow more scope for transient viral
+evolution, although only if extra-pair contact is possible.
+As mentioned above, our compartmental model already requires tens
+of thousands of coupled differential equations, which will increase
+multiplicatively with additional model dimensions such as age,
+sex, or HIV stage. Thus, further model elaboration will best be 
+done with agent-based models.
 
@@
 
-\begin{thebibliography}{10}
-
-\bibitem{alizon_virulence_2009}
-Alizon S, Hurford A, Mideo N, van Baalen M.
-\newblock Virulence evolution and the trade-off hypothesis: history, current
-  state of affairs and the future.
-\newblock J Evol Biol. 2009;22:245--259.
-\newblock doi:{10.1111/j.1420-9101.2008.01658.x}.
-
-\bibitem{EbertBull2003}
-Ebert D, Bull JJ.
-\newblock Challenging the trade-off model for the evolution of virulence: is
-  virulence management feasible?
-\newblock Trends Microbiol. 2003;11(1):15--20.
-
-\bibitem{alizon_adaptive_2015}
-Alizon S, Michalakis Y.
-\newblock Adaptive virulence evolution: the good old fitness-based approach.
-\newblock Trends in Ecology \& Evolution. 2015;30(5):248--254.
-\newblock doi:{10.1016/j.tree.2015.02.009}.
-
-\bibitem{Dwyer+1990}
-Dwyer G, Levin SA, Buttel L.
-\newblock A simulation model of the population dynamics and evolution of
-  myxomatosis.
-\newblock Ecol Monog. 1990;60:423--447.
-
-\bibitem{mackinnon1999genetic}
-Mackinnon MJ, Read AF.
-\newblock Genetic relationships between parasite virulence and transmission in
-  the rodent malaria {{\em Plasmodium chabaudi}}.
-\newblock Evolution. 1999; p. 689--703.
-
-\bibitem{jensen2006empirical}
-Jensen KH, Little T, Skorping A, Ebert D.
-\newblock Empirical support for optimal virulence in a castrating parasite.
-\newblock PLoS Biol. 2006;4(7):e197.
-
-\bibitem{deroode2008virulence}
-De~Roode JC, Yates AJ, Altizer S.
-\newblock Virulence-transmission trade-offs and population divergence in
-  virulence in a naturally occurring butterfly parasite.
-\newblock Proceedings of the National Academy of Sciences.
-  2008;105(21):7489--7494.
-
-\bibitem{Fraser+2007}
-Fraser C, Hollingsworth TD, Chapman R, de~Wolf F, Hanage WP.
-\newblock Variation in {HIV}-1 set-point viral load: Epidemiological analysis
-  and an evolutionary hypothesis.
-\newblock PNAS. 2007;104:17441--17446.
-
-\bibitem{fraser_virulence_2014}
-Fraser C, Lythgoe K, Leventhal GE, Shirreff G, Hollingsworth TD, Alizon S,
-  et~al.
-\newblock Virulence and Pathogenesis of {HIV}-1 Infection: An Evolutionary
-  Perspective.
-\newblock Science. 2014;343(6177):1243727.
-\newblock doi:{10.1126/science.1243727}.
-
-\bibitem{shirreff_transmission_2011}
-Shirreff G, Pellis L, Laeyendecker O, Fraser C.
-\newblock Transmission Selects for {HIV-1} Strains of Intermediate Virulence: A
-  Modelling Approach.
-\newblock PLoS Computational Biology. 2011;7(10):e1002185.
-\newblock doi:{10.1371/journal.pcbi.1002185}.
-
-\bibitem{herbeck_hiv_2014}
-Herbeck JT, Mittler JE, Gottlieb GS, Mullins JI.
-\newblock An {HIV} Epidemic Model Based on Viral Load Dynamics: Value in
-  Assessing Empirical Trends in {HIV} Virulence and Community Viral Load.
-\newblock {PLoS} Comput Biol. 2014;10(6):e1003673.
-
-\bibitem{herbeck_evolution_2016}
-Herbeck JT, Mittler JE, Gottlieb GS, Goodreau SM, Murphy JT, Cori A, et~al.
-\newblock Evolution of {HIV} virulence in response to widespread scale up of
-  antiretroviral therapy: a modeling study.
-\newblock Virus Evolution. 2016;2(2):vew028.
-\newblock doi:{10.1093/ve/vew028}.
-
-\bibitem{day_virulence_2004}
-Day T, Proulx SR.
-\newblock A General Theory for the Evolutionary Dynamics of Virulence.
-\newblock The American Naturalist. 2004;163(4):E40--E63.
-\newblock doi:{10.1086/382548}.
-
-\bibitem{alizon_price_2009}
-Alizon S.
-\newblock The {Price} equation framework to study disease within-host
-  evolution.
-\newblock Journal of Evolutionary Biology. 2009;22(5):1123--1132.
-\newblock doi:{10.1111/j.1420-9101.2009.01726.x}.
-
-\bibitem{champredon_hiv_2013}
-Champredon D, Bellan S, Dushoff J.
-\newblock {HIV} Sexual Transmission Is Predominantly Driven by Single
-  Individuals Rather than Discordant Couples: A Model-Based Approach.
-\newblock PLoS ONE. 2013;8(12):e82906.
-\newblock doi:{10.1371/journal.pone.0082906}.
-
-\bibitem{dietz_epidemiological_1988}
-Dietz K, Hadeler KP.
-\newblock Epidemiological models for sexually transmitted diseases.
-\newblock Journal of Mathematical Biology. 1988;26(1):1--25.
-
-\bibitem{kretzschmar_effect_1998}
-Kretzschmar M, Dietz K.
-\newblock The effect of pair formation and variable infectivity on the spread
-  of an infection without recovery.
-\newblock Mathematical Biosciences. 1998;148(1):83--113.
-
-\bibitem{hollingsworth_hiv1_2008}
-Hollingsworth TD, Anderson RM, Fraser C.
-\newblock {HIV}-1 Transmission, by Stage of Infection.
-\newblock Journal of Infectious Diseases. 2008;198(5):687--693.
-\newblock doi:{10.1086/590501}.
-
-\bibitem{omori2015dynamics}
-Omori R, Chemaitelly H, Abu-Raddad LJ.
-\newblock Dynamics of non-cohabiting sex partnering in sub-{Saharan} {Africa}:
-  a modelling study with implications for {HIV} transmission.
-\newblock Sexually Transmitted Infections.
-  2015;doi:{10.1136/sextrans-2014-051925}.
-
-\bibitem{may_transmission_1988}
-May RM, Anderson RM.
-\newblock The Transmission Dynamics of Human Immunodeficiency Virus (HIV).
-\newblock Philosophical Transactions of the Royal Society of London Series B,
-  Biological Sciences. 1988;321(1207):565--607.
-
-\bibitem{delva_connecting_2016}
-Delva W, Leventhal GE, Helleringer S.
-\newblock Connecting the dots: network data and models in {HIV} epidemiology.
-\newblock AIDS. 2016;30(13):2009--2020.
-\newblock doi:{10.1097/QAD.0000000000001184}.
-
-\bibitem{blower_drugs_1991}
-Blower SM, Hartel D, Dowlatabadi H, Anderson RM, May RM.
-\newblock Drugs, Sex and {HIV}: A Mathematical Model for {New York City}.
-\newblock Philosophical Transactions of the Royal Society of London B:
-  Biological Sciences. 1991;331(1260):171--187.
-\newblock doi:{10.1098/rstb.1991.0006}.
-
-\bibitem{ma_estimating_2014}
-Ma J, Dushoff J, Bolker BM, Earn DJD.
-\newblock Estimating Initial Epidemic Growth Rates.
-\newblock Bulletin of Mathematical Biology. 2014;76(1):245--260.
-\newblock doi:{10.1007/s11538-013-9918-2}.
-
-\bibitem{wallinga_how_2007}
-Wallinga J, Lipsitch M.
-\newblock How generation intervals shape the relationship between growth rates
-  and reproductive numbers.
-\newblock Proceedings of the Royal Society of London B: Biological Sciences.
-  2007;274(1609):599--604.
-\newblock doi:{10.1098/rspb.2006.3754}.
-
-\bibitem{soetaert_solving_2010}
-Soetaert K, Petzoldt T, Setzer RW.
-\newblock Solving Differential Equations in {R}: Package {deSolve}.
-\newblock Journal of Statistical Software. 2010;33(9):1--25.
-\newblock doi:{10.18637/jss.v033.i09}.
-
-\bibitem{wawer2005rates}
-Wawer MJ, Gray RH, Sewankambo NK, Serwadda D, Li X, Laeyendecker O, et~al.
-\newblock Rates of {HIV}-1 transmission per coital act, by stage of {HIV}-1
-  infection, in {Rakai}, {Uganda}.
-\newblock Journal of Infectious Diseases. 2005;191(9):1403--1409.
-
-\bibitem{faria_early_2014}
-Faria NR, Rambaut A, Suchard MA, Baele G, Bedford T, Ward MJ, et~al.
-\newblock The early spread and epidemic ignition of {HIV}-1 in human
-  populations.
-\newblock Science (New York, NY). 2014;346(6205):56--61.
-\newblock doi:{10.1126/science.1256739}.
-
-\bibitem{elderd_uncertainty_2006}
-Elderd BD, Dukic VM, Dwyer G.
-\newblock Uncertainty in predictions of disease spread and public health
-  responses to bioterrorism and emerging diseases.
-\newblock Proceedings of the National Academy of Sciences. 2006;103(42):15693
-  --15697.
-\newblock doi:{10.1073/pnas.0600816103}.
-
-\bibitem{herbeck_is_2012}
-Herbeck JT, Müller V, Maust BS, Ledergerber B, Torti C, Di~Giambenedetto S,
-  et~al.
-\newblock Is the virulence of {HIV} changing? {A} meta-analysis of trends in
-  prognostic markers of {HIV} disease progression and transmission.
-\newblock AIDS (London, England). 2012;26(2):193--205.
-\newblock doi:{10.1097/QAD.0b013e32834db418}.
-
-\bibitem{payne_impact_2014}
-Payne R, Muenchhoff M, Mann J, Roberts HE, Matthews P, Adland E, et~al.
-\newblock Impact of {HLA}-driven {HIV} adaptation on virulence in populations
-  of high {HIV} seroprevalence.
-\newblock Proceedings of the National Academy of Sciences.
-  2014;111(50):E5393--E5400.
-\newblock doi:{10.1073/pnas.1413339111}.
-
-\bibitem{roberts2015impact}
-Roberts HE, Goulder PJ, McLean AR.
-\newblock The impact of antiretroviral therapy on population-level virulence
-  evolution of {HIV}-1.
-\newblock Journal of The Royal Society Interface. 2015;12(113):20150888.
-
-\end{thebibliography}
+\bibliography{virulence}
 \end{document}

### 2017-02-10 ms/ms.tex
comments

@@
 
+\todo{Define virulence? I think it might be a good idea to add it to author summary as well for non-technical readers...}
 The evolution of pathogen virulence has both 
@@
 reproduction. Higher viral loads are correlated with 
-shorter progression times (higher virulence).
+shorter progression times (higher virulence). \todo{We say the same thing in the next sentence.}
 Studies of discordant partnerships --- stable
@@
 Thus the within-couple transmission rate, $\beta$
-(see ``Contact Structure'' below)
+(see ``Contact Structure'' below),
 for our models is given by:
@@
 
+\todo{I changed $V_\alpha$ to $V(\alpha)$ because $\alpha$ isn't necessarily an integer...}
+
 \begin{equation}
 \begin{split}
-D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
-\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
-{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
+D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V(\alpha) ^{D_k} + D_{50}^{D_k}}, \\
+\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V(\alpha) ^ {\beta_k}}%
+{V(\alpha)^{\beta_k} + \beta_{50} ^{\beta_k}},
 \end{split}
@@
 \end{equation}
-where $V_{\alpha} = 10^\alpha$.
+where $V(\alpha) = 10^\alpha$.
 
@@
 Instead, the mutational effect is modeled as occurring in a single
-step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
+step at the time of transmission.
+
+First, the distribution of \Lspvl\ is discretized into a vector:
 \begin{equation}
@@
 
-We construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
+We then construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
 Finally, the probabilities are normalized so that each row sums to 1:
@@
 Lastly, we incorporate heterogeneity in sexual activity into the models.
-Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
+Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1) \todo{We should also mention that appendix S1 has full model details...
+It doesn't seem like we mention that anywhere in the txet.}.
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
@@
 We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
-of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$.
+of mean \Lspvl\ of 3 \todo{We mention that initial sd is 0.2 in the Appendix but we don't say it in the text.
+We can add it here or to the table...} and epidemic size of $10^{-4}$.
 After each run, initial exponential growth rate is calculated.
@@
 
-Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory.
+Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material) \todo{I'm not sure where in supplementary material you're referring to...
+Isn't this just figure 1?}, has little effect on the virulence trajectory.
 Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer exponential-growth phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
@@
 However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV is able to evolve most of the way to the SPVL that would be optimal in an indefinitely prolonged growth phase.
-
-Across the entire range of parameters covered by the LHS analysis, all
+\todo{The last sentence makes sense but it's a little bit awkward...
+I'm not sure why...}
+
+Across the entire range of parameters covered by the LHS analysis \todo{replace this with Latin hypercube samples? Seems like a better terminology than LHS analysis}, all
 of the classes of models we considered produce qualitatively similar
@@
 \Lspvl\ corresponds to higher virulence). Although the speed of
-virulence evolution varies, leading to wide variation in the peak SPVL
+virulence evolution varies, leading to wide variation in the peak SPVL \todo{log10 SPVL?}
 (ranging from 3 to 5.5) because HIV can evolve farther toward
@@
 Models with extra-pair contact show wide range parameters as they can display a wide range of dynamics depending on $c_e/c_w$ (as well as $c_u/c_w$ for models with uncoupled mixing) and thus require wide range of scaling factors to achieve the target $r$.
-Parameter range for random-mixing model is severely compressed because latin hypercube sampling does not affect the prevalence trajectory very much, and scaling factor does not vary much across parameters.
+Parameter range for random-mixing model is severely compressed because Latin hypercube sampling does not affect the prevalence trajectory very much, and scaling factor does not vary much across parameters.
 

### 2017-02-11 ms/cover_letter/plosHIVcover_resubm2.tex
almost last edits to ms ...

@@
 The point about initial virulence being measured immediately after
-a species jump is interesting, but since it is somewhat peripheral
-to our main exploration here, we have opted to include in the 
-Discussion rather than trying to squeeze it into the abstract
-\todo{don't forget to do this}
+a species jump is interesting, but seems peripheral
+to our goals for this paper.
 }
@@
 
-\response{make sure we have a caveat/comment in text}
+\response{We now discuss this issue (l. 500).}
 
@@
 
-\response{same thing again}
+\response{This is essentially the same point as above; we have tried to 
+temper our conclusions on lines 60 and 500.}
 

### 2017-02-11 ms/ms.tex
almost last edits to ms ...

--- 2017-02-08 ms/ms.tex
@@
 \newcommand{\peryear}{year$^{-1}$}
+\newcommand{\percap}{\emph{per capita}}
 \usepackage{color}
@@
 set-point viral load, SPVL) and their heritability between
-hosts. Their models use implicit equations to
+hosts. Their model used implicit equations to
 capture the effects of partnership dynamics that are at the core of 
@@
 important processes such as people entering and leaving sexual
-partnerships. We compared several evolutionary models for HIV that include
+partnerships. 
+We compared several evolutionary models for HIV that include
 partnership dynamics as well as sexual contact outside of stable
@@
 
-The evolution of pathogen virulence has both 
+The evolution of pathogen virulence 
+(the harm done to the pathogen's host) has both 
 theoretical and, potentially, practical
-importance. In general, evolutionary theory suggests that
-pathogens with higher reproduction numbers ($\rzero$) --- that is, the number of secondary infections caused by a single infected host over the course of its infectious period --- will tend to increase in prevalence relative to strains with lower reproductive ratios.
+importance. 
+Evolutionary theory suggests that
+pathogens with higher reproduction numbers ($\rzero$) --- the number of secondary infections caused by a single infected host over the course of its infectious period --- will increase in prevalence relative to strains with lower reproductive ratios.
 Pathogens can increase their reproduction numbers either
@@
 
-Fraser \etal\ have applied these ideas to HIV,
-showing that the virus appears to satisfy the prerequisites of
+Fraser \etal\ have showed that HIV appears to satisfy the prerequisites of
 the tradeoff theory.
@@
 stage of infection) is a measurable proxy for the rate of HIV within-host
-reproduction. Higher viral loads are correlated with 
-shorter progression times (higher virulence).
+reproduction.
+Higher viral loads are
+correlated with faster progression to AIDS (higher virulence).
 Studies of discordant partnerships --- stable
 sexual partnerships with one infected and one uninfected partner --- %
-have shown that SPVL is (1) correlated with faster transmission
-(the time to infection of the uninfected partner is shorter when
-their partner has a higher SPVL) and (2) heritable
-(when the uninfected member of
-the couple does become infected, their SPVL is similar to their
-partner's).
-Furthermore, the rate of transmission increases with
-increasing SPVL (and hence with decreasing progression time)
-but with a decreasing slope as progression time decreases,
+have shown that SPVL is (1) positively correlated with transmission
+(people with higher SPVL transmit HIV to their uninfected partners 
+sooner) and (2) heritable
+(when the uninfected partner does become infected, 
+their SPVL is similar to their originally infected partner's).
+Furthermore, the rate of increase in transmission 
+has a decreasing slope as progression time decreases,
 fulfilling the requirements of the tradeoff theory
@@
 virulence, peaking after approximately 100-200 years and then declining
-slightly to a long-stable virulence level.
+slightly to a long-term stable virulence level.
 
@@
 contact --- without representing the explicit dynamics of partnership
-formation and dissolution or accounting for extra-partnership contact;
-agent-based formulations are more realistic. Here we explore the
-effects of incorporating explicit contact structure in
+formation and dissolution or accounting for extra-pair contact;
+agent-based formulations are more realistic, but can make it difficult
+to tease apart the reasons behind particular epidemic phenomena. 
+Here we explore the effects of incorporating explicit contact structure in
 eco-evolutionary models.
@@
 asymptomatic, and AIDS) and used a more realistic
-Weibull-distributed infectious period; we show below that our results
-are not overly sensitive to this simplification. 
-Shirreff \etal\ allowed the transmission rate
-and infection duration to depend on virus load only during the
+Weibull-distributed infectious period.
+We show below that our result
+are not overly sensitive to this simplification,
+although it could conceivably affect
+our conclusions about the evolution of virulence (e.g., Kretzschmar and
+Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
+and multiple stages of infectivity have interactive effects on
+$\rzero$). 
+In Shirreff \etal's model,the transmission rate
+and infection duration depend on virus load only during the
 asymptomatic stage. In order to adapt their parameterization to our
-single-stage model, we used their estimates to compute the SPVL-dependent transmission
+single-stage model, we used their parameters to compute the SPVL-dependent transmission
 rate and duration during the asymptomatic stage and 
@@
 \begin{split}
-D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
-\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
-{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
+D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V(\alpha) ^{D_k} + D_{50}^{D_k}}, \\
+\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V(\alpha) ^ {\beta_k}}%
+{V(\alpha)^{\beta_k} + \beta_{50} ^{\beta_k}},
 \end{split}
@@
 
-The uncoupled and extra-couple transmission rates (i.e., the rates of
+In models that allow extra-pair contact,
+the uncoupled and extra-couple transmission rates (i.e., the rates of
 transmission among people outside of a stable partnership, or between
@@
 multiplying the within-couple transmission rate $\beta$ by the contact
-ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1). Simplifying the model
-of HIV pathogenesis from three stages to a single stage could affect
-our conclusions about the evolution of virulence (e.g. Kretzschmar and
-Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics
-and multiple stages of infectivity have interactive effects on
-$\rzero$).
+ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1).
 
@@
 
-We construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that the infector has \Lspvl\ of $\alpha_i$.
+We then construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that their infected partner has \Lspvl\ of $\alpha_i$.
 Finally, the probabilities are normalized so that each row sums to 1:
@@
 non-instantaneous partnership formation; (2) zero vs.
-positive extra-partnership sexual contact and transmission; and (3) homogeneous vs.
+positive extra-pair sexual contact and transmission; and (3) homogeneous vs.
 heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
@@
 The rates of pair formation are based on the numbers of uncoupled susceptible and infected individuals and the pair-formation rate; partnerships can either dissolve into singletons or be transformed into other types of partnerships by infection of one partner.
-This model also includes extra-partnership contact with both uncoupled individuals and individuals in other partnerships (we denote it ``pairform+epc''), so that susceptible uncoupled individuals and susceptible partners in any type of partnership can be infected by infected uncoupled individuals or infected partners in any type of partnership.
-
-Specifically, single individuals ($S$ and $I$) acquire partners at a rate $\rho$, and partnerships dissolve at a rate $c$.
+This model also includes extra-pair contact with both uncoupled individuals and individuals in other partnerships (we denote it ``pairform+epc''), so that susceptible uncoupled individuals and susceptible partners in any type of partnership can be infected by infected uncoupled individuals or infected partners in any type of partnership.
+
+Specifically, single individuals ($S$ and $I$) form partnerships at a \percap\ rate $\rho$, and partnerships dissolve at a rate $c$.
 Infected individuals in a discordant partnership infect their susceptible partner at a rate $\beta$ (within-couple transmission rate) and susceptible individuals outside the partnership at a rate $c_e$ (extra-couple transmission rate).
@@
 The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$.
-Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'', \diagfig{b}) includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'', \diagfig{a}) only considers within-couple transmission.
+Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'', \diagfig{b}) includes extra-pair contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'', \diagfig{a}) only considers within-couple transmission.
 Although these models can also be implemented
-by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.
+by setting the partnership formation rate of the explicit partnership models to a high value (we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently so that scaling the partnership formation rate during model calibration (see \emph{Simulation runs} below) does not affect the eco-evolutionary dynamics.
 
 The fifth and sixth models represent extreme simplifications of sexual partnership dynamics.
-One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
-It is actually a random mixing model that explicitly tracks only the total number of susceptible and infected individuals.
+The fifth (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}.
+It is a random-mixing model that explicitly tracks only the total number of susceptible and infected individuals.
 However, to reflect the effect of partnership structure, it uses an adjusted transmission rate derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}.
-The second model of this pair (``random'') is a simple random-mixing model.
-
-Lastly, we incorporate heterogeneity in sexual activity into the models.
-Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1).
+The sixth model (``random'') is a simple random-mixing model.
+
+Lastly, we incorporated heterogeneity in sexual activity into the models.
+Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1 for full model details).
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
-(In the main text we focus on the model with non-instantaneous pair formation, extra-partnership contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix S2 presents results on the effect of adding heterogeneity to other model variants.)
+(In the main text we focus on the model with non-instantaneous pair formation, extra-pair contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix~S2 presents results on the effect of adding heterogeneity to other model variants.)
 While this model lacks some 
@@
 are removed from the population, with constant recruitment rates and constant low
-\emph{per capita} natural mortality rates; this additional structure had only minor effects on
+\percap\ natural mortality rates; this additional structure had only minor effects on
 the results.
@@
 
-Choosing the initial conditions for the simulations is something of a
-challenge. In the most common theoretical modeling contexts
+Choosing the initial conditions for the simulations is challenging.
+In many modeling studies,
 researchers are primarily interested in equilibria (or other long-term
@@
 conclusions. We have no empirical information that would justify a
-particular quantitative choice of the fraction infected and the mean
+particular choice of the fraction infected and the mean
 and variance of the distribution of SPVL at the point when the
@@
 We ran most of our models across a wide range of parameters, as described
-in the next section. In several cases, however, we inspected only a few
-cases, to qualitatively assess the sensitivity of the models to
+in the next (\emph{Latin hypercube sampling}) 
+section. In several cases, however, we inspected only a few
+parameter sets, to qualitatively assess the sensitivity of the models to
 initial conditions or to model structure. 
@@
 
-Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
+Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this uncertainty led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values.
 In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$).
 For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well.
-We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
+We do not allow for uncertainty in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.
 
@@
 Table~\ref{table:parmsTable} gives the ranges of the model parameters used for LHS.
-Parameter ranges regarding contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those regarding infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
-The remaining parameters are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
-
-One parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see Appendix S1);
-it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  so we need to pick a reasonable range for it.
+Ranges of parameters controlling contact and partnership dynamics ($\rho$, $c$, and $c_e/c_w$) are taken from Champredon \etal\ \cite{champredon_hiv_2013}, whereas those controlling infection ($\beta_P$, $D_P$, $\beta_D$, and $D_D$) are taken from Hollingsworth \etal\ \cite{hollingsworth_hiv1_2008}.
+The remaining parameter values are taken from Shirreff \etal\ \cite{shirreff_transmission_2011}.
+
+One new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see Appendix S1).
+Since it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  we need to pick a reasonable range for it.
 Champredon \etal\ \cite{champredon_hiv_2013} assume that the effective within-couple contact rate and effective uncoupled contact rate have the same range of 0.05 - 0.25.
@@
 Therefore, we use 1/5-5 as the range for the parameter $c_u/c_w$.
-Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is not much known about the uncoupled transmission rate.
+Although this adds more uncertainty to the parameter $c_u$ --- Champredon \etal's range implies a 5-fold difference whereas ours gives a 25-fold difference --- we consider the wider range appropriate, as little is known about the uncoupled transmission rate.
 
@@
 we narrowed this range for $\textrm{CV}^2$ to 0.01-100.
-At the bottom end of the range, estimating that a group behaves
+At the bottom end of the range, an observation that a group behaves
 perfectly homogeneously ($CV=0$) is likely to be a sampling artifact;
@@
 
-One of the most difficult parts of model comparison is finding
+One of the hardest parts of model comparison is finding
 parameter sets that are commensurate across radically different model
@@
 some of our models, we scaled both partnership change rate and
-transmission rate by the same factor of $\gamma$: $\tsub{\beta}{adj} =
+transmission rate by the same factor $\gamma$: $\tsub{\beta}{adj} =
 \gamma \tsub{\beta}{base}$, $\tsub{c}{adj} = \gamma \tsub{c}{base}$,
 $\tsub{\rho}{adj} = \gamma \tsub{\rho}{base}$. Since transmission rate
-is adjusted by the scale of $\gamma$, uncoupled and extra-couple
+is scaled by $\gamma$, uncoupled and extra-couple
 transmission rates are adjusted as well. For the instantaneous-switching
@@
 We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
-of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$.
-After each run, initial exponential growth rate is calculated.
-Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
-For calibration purposes, we run each model for only 500 years
+of mean \Lspvl\ of 3.0, standard deviation of \Lspvl\ of 0.2, and epidemic size of $10^{-4}$.
+After each run, the initial exponential growth rate is calculated.
+Then the parameters are scaled as described above so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
+When calibrating, we run each model for only 500 years
 (full simulations are run for 4000 years), which is always long
@@
 (For the heterogeneous model, approximately
-10\% of the samples failed due to numerical instability.)
+10\% of the samples failed due to numerical instability;
+we discarded these runs.)
 
@@
 and minimum expected time to progression. 
-While the rest of the summary statistics measure \Lspvl, we also report minimum expected time to progression for clearer interpretability.
-Minimum mean progression time is obtained by applying the Hill function (eq.~\ref{eq:hillfuns}) to maximum mean \Lspvl\ of each run.
+Minimum expected progression time is obtained by applying the Hill function (eq.~\ref{eq:hillfuns}) to the maximum mean \Lspvl\ of each run.
 Equilibrium \Lspvl\ is calculated after 4000 years of simulated
 time. 
-Although most simulations reach equilibrium much earlier, we set our time horizon at a much later date as some simulation runs have slow rate of evolution depending on the parameter set and model assumptions.
-
-Knowing the peak \Lspvl, timing of the peak\Lspvl/peak virulence, and equilibrium \Lspvl\ provide sufficient detail to identify the overall shape of the virulence trajectory.
+Although most simulations reach equilibrium much earlier than 4000 years, we set this very long time horizon because a small subset of the simulation runs show very slow evolution rates.
+
+Knowing the peak \Lspvl, timing of the peak \Lspvl/peak virulence, and equilibrium \Lspvl\ provide sufficient detail to identify the overall shape of the virulence trajectory.
 In particular, knowing the timing of the peak virulence (how many years
@@
 guess whether the virulence of an emerging pathogen is likely (1)
-to have peaked early, possibly even before the pathogen is detected
+to peak early, possibly even before the pathogen is detected
 spreading in the population, and decline over the remaining course
@@
 Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
-As we calibrate the parameters to achieve initial growth rates $r$ from 0.042 \peryear\ to 0.084 \peryear\ (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
+As we calibrate the parameters to achieve initial epidemic growth rates $r$ ranging
+from 0.042 \peryear\ to 0.084 \peryear\ (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a).
 For matching initial growth rates ($r=0.042$)
@@
 
-Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory.
+Changing the initial infectious density ($I(0)$) has little effect on the virulence trajectory.
 Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer exponential-growth phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b).
@@
 The peak SPVL is actually larger for a lower starting SPVL, presumably because lower SPVL also allows for a longer epidemic phase.
-However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV is able to evolve most of the way to the SPVL that would be optimal in an indefinitely prolonged growth phase.
-
-Across the entire range of parameters covered by the LHS analysis, all
-of the classes of models we considered produce qualitatively similar
+However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV can evolve close to its
+optimal growth-phase SPVL.
+
+Across the entire range of parameters covered by the
+Latin hypercube samples, all
+of our models produce qualitatively similar
 virulence trajectories, which we quantify in terms of population mean
@@
 \Lspvl\ corresponds to higher virulence). Although the speed of
-virulence evolution varies, leading to wide variation in the peak SPVL
-(ranging from 3 to 5.5) because HIV can evolve farther toward
-its exponential-phase optimum before the transition to the
-endemic phase, virulence peaks in all models between 200 and
-300 years.
+virulence evolution varies, leading to wide variation in the peak
+\Lspvl\ (ranging from 3 to 5.5) because HIV can evolve farther toward
+its growth-phase optimum before the transition to the
+endemic phase, virulence peaks between 200 and
+300 years in all models.
 
@@
 Envelopes contain the middle 95\% of trajectories (i.e.
-we select all points between the 0.025 and 0.975 quantiles for each model at each year), while lines within the envelopes show mean trajectories.
+we select all points between the 0.025 and 0.975 quantiles for each model at each year), while center lines show mean trajectories.
 Truncated series (up to year 800) are shown here.}
@@
 virulence may decline immediately, leading to a peak virulence (as we have
-defined it ) equal to the starting value of $\alpha(0)=3$ \Lspvl at $t=0$.
+defined it) equal to the starting value of $\alpha(0)=3$ \Lspvl at $t=0$
 (\figurename~\ref{fig:pairplot}).
@@
 Models arranged in order of complexity, as in \figurename~\ref{fig:virtraj}.
-For every simulation, three summary statistics were obtained: peak time, maximum mean \Lspvl, and equilibrium mean \Lspvl. 
+For every simulation, three summary statistics are shown: peak time, maximum mean \Lspvl, and equilibrium mean \Lspvl.
 Peak time is the time at which the SPVL reaches its maximum for each simulation run (maximum mean \Lspvl).
@@
 Because the distribution of equilibrium mean \Lspvl\
-(lower left panel) for the random mixing model is very narrow, it
+(lower left panel) for the random-mixing model is very narrow, it
 has been replaced by a point in order to preserve the vertical axis scaling.}
@@
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
-Three summary statistics obtained from each simulation have been plotted against each other to visualize the relationship among the summary statistics and to allow for clearer comparison between models.
+The three summary statistics for each simulation are plotted against each other in order to visualize the relationship among the summary statistics and to help compare models.
 Surprisingly, the implicit model, an approximation for instantaneous partnership formation model (instswitch), shows an almost identical trend with a model that has pair formation dynamics (pairform).
-Dashed line in equilibrium vs. peak virulence plot shows 1:1 line, where equilibrium and peak virulence are equal.
-100 values were sampled from each model to allow for clearer distinction between the models}
+The \emph{dashed line} in the equilibrium vs. peak virulence plot (lower left) shows the 1:1 line, where equilibrium and peak virulence are equal.
+To avoid too much overplotting, only 10\% of the parameter sets
+(randomly sampled) are shown here.}
 \label{fig:pairplot}
@@
 The most striking aspect of the univariate comparisons in
-\figurename~\ref{fig:unidist}, (and the bivariate comparisons in
-\figurename~\ref{fig:pairplot}) is the similarity between the results of the
-least (random mixing) and the most complex (pair formation with
-extra-pair contact, pairform+epc with heterogeneity) models.
+\figurename~\ref{fig:unidist} (and the bivariate comparisons in
+\figurename~\ref{fig:pairplot}), is the similarity between the results of the
+least (random-mixing) and the most complex (pair formation with
+extra-pair contact and pairform+epc with heterogeneity) models.
 The random-mixing model has the lowest variability, because it is unaffected by uncertainty in pair formation and extra-pair contact parameters, but otherwise the virulence
@@
 3.52 (3-4.02) years.
-In terms of mean time to progression to AIDS,
 The random-mixing, pairform+epc, and heterogeneous models predict rapid
@@
 random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model (see Appendix S2 for 
-plots showing univariate summaries
-of expected progression time to AIDS and transmission probability).
+for the implicit model. (Appendix S2 
+presents plots showing univariate summaries
+of expected progression time to AIDS and transmission probability.)
 
@@
 \caption{{\bf Sensitivity plot.}
-For most parameters in the Latin hypercube samples, the figure shows the distribution (points) and trend (smooth line) of the summary statistic as a function of the parameter value, some of which have adjusted to achieve the standard initial epidemic growth rate (see Methods).
-Opposing trends in summary statistics with respect to partnership dissolution rate, $c$, (third column) between the model with and without extra-pair contact illustrates that similar model with different assumptions may yield opposing predictions.
+Distribution (points) and trend (smooth line) of summary statistics (y-axis,
+rows) as a function of parameter values (x-axis, columns), some of which have been scaled to calibrate the initial epidemic growth rate.
+Changing parameters may have differing effects for models with different
+contact structures: for example, partnership dissolution rate ($c$, third column) increases peak time and virulence in models without explicit pair formation but decreases them in more realistic models.
 }
@@
 
-The bivariate relationships (\figurename~\ref{fig:pairplot}) help distinguish the results of 
-different models with similar univariate dynamical summaries. While the
+Bivariate relationships (\figurename~\ref{fig:pairplot}) help distinguish the results of 
+different models with similar univariate 
+distributions of dynamical summaries. While the
 relationship between equilibrium \Lspvl\ and peak time is
@@
 relationships show more variation. In particular, the implicit
-and pair-formation (without extra-pair contact) are very similar
+and pair-formation (without extra-pair contact) models are very similar
 to each other, but distinct from the other models. We still do
@@
 the instantaneous-switching model without extra-pair contact,
-which most closely matches its derivation. However, we note
+which most closely matches its underlying assumptions. However, we note
 that the implicit model derivation is based on defining
@@
 $D_P$, $D_D$, $c_e/c_w$, $c_u/c_w$, $\kappa$, $\mu$);
-note that the effects of $\beta_D$ and $D_D$ are not shown in \figurename~\ref{fig:plot_sens} as their patterns are so similar to $\beta_P$ and $D_P$, respectively.
-
-One of the distinguishing features of the scaled parameters (in comparison to the unscaled parameters) is that the range of the parameters is compressed (relative to other models) for models without extra-pair contact as those models require parameters to be scaled by a large value in order to achieve a given $r$ of 0.04.
-Models with extra-pair contact show wide range parameters as they can display a wide range of dynamics depending on $c_e/c_w$ (as well as $c_u/c_w$ for models with uncoupled mixing) and thus require wide range of scaling factors to achieve the target $r$.
-Parameter range for random-mixing model is severely compressed because latin hypercube sampling does not affect the prevalence trajectory very much, and scaling factor does not vary much across parameters.
-
-For parameters involved in partnership turnover ($c$ and $\rho$), the main distinction is between models with and without extra-pair contact.
-With extra-pair contact, we see a gradual decrease in peak time, maximum \Lspvl, and equilibrium \Lspvl\ with increasing turnover rates.
+the effects of $\beta_D$ and $D_D$ are not shown in \figurename~\ref{fig:plot_sens} as 
+they show patterns almost identical to $\beta_P$ and $D_P$, respectively.
+
+For the scaled parameters, the parameter ranges 
+(horizontal axis) are compressed 
+for models without extra-pair contact because these models 
+require a large amount of parameter scaling in order 
+to achieve the specified initial epidemic growth rate ($r=0.04$).
+In contrast, models with extra-pair contact show a wide range of parameters as they can display a wide range of dynamics depending on $c_e/c_w$ (as well as $c_u/c_w$ for models with uncoupled mixing) and thus require a wide range of scaling factors to achieve the target growth rate.
+Parameter ranges for the random-mixing model (especially $c$) are severely compressed because this model has little flexibility.
+
+For parameters involved in partnership turnover ($c$ and $\rho$), 
+the figure again shows differences between models with and without extra-pair contact.
+Models with extra-pair contact show a gradual decrease in peak time, maximum \Lspvl, and equilibrium \Lspvl\ with increasing turnover rates.
 Increases in the other parameters lead to increases in all
 three summary statistics.
-We interpret these patterns as evidence
-that the increase in turnover rates diminishes the effect of extra-pair contact (equivalently increasing relative importance of within-partnership contact), therefore selecting for lower \Lspvl.
-For models without extra-pair contact, increased rates of partnership change decrease the level of structured mixing (mimicing extra-pair contact models), resulting in selection for higher \Lspvl.
-
-We also note that the implicit model and the instantaneous partnership formation model show similar patterns in scaled parameters.
-In fact, we can see that the effect of partnership dissolution rate, $c$, on equilibrium \Lspvl\ of the two models are almost identical (in \figurename~\ref{fig:pairplot}, they have distinguishing features).
-Lastly, we note that increase in transmission rates ($\beta_P$ and $\beta_D$) has a decreasing effect on summary statistics for all models except for random mixing model.
+In these models, increased turnover rates diminish the effect of extra-pair contact, thus selecting for lower \Lspvl.
+For models without extra-pair contact, increased turnover rates
+decrease the level of structured mixing (mimicking extra-pair contact models), resulting in selection for higher \Lspvl.
+
+The implicit model and the instantaneous partnership formation model show similar patterns in scaled parameters.
+In fact, the effect of partnership dissolution rate, $c$, on equilibrium \Lspvl\ is almost identical
+in these models (although they can be
+distinguished in \figurename~\ref{fig:pairplot}).
+Lastly, increasing in transmission rates ($\beta_P$ and $\beta_D$)
+causes the summary statistics to decrease in all models except the random-mixing model.
 
 Surprisingly, once calibration
-is taken into account, the unscaled parameters have little to no effect overall.
-Increase in duration ($D_P$, $D_D$) in the primary and disease stages generally decrease the equilibrium virulence, peak virulence, and peak time,
-although the models with uncoupled mixing and random mixing model have high, relatively
-constant values with respect to these parameters.
+is taken into account, the unscaled parameters have little effect overall.
+Increase in duration ($D_P$, $D_D$) in the primary and disease stages generally decreases the equilibrium virulence, peak virulence, and peak time,
+although the models with uncoupled mixing and random-mixing model have high, relatively constant values with respect to these parameters.
 The ratio of extra-pair to within-pair contact ($c_e/c_w$) affects
@@
 Neither the uncoupled contact rate nor
-the mean ($\mu$) or CV$^2$ of the number of non-cohabiting sexual
+the mean ($\mu$) or $\textrm{CV}^2$ of the number of non-cohabiting sexual
 partners has much systematic effect in the heterogeneous model.
@@
 researchers' conclusions about the evolutionary dynamics of
-virulence. In particular, a relatively simple (strategic) 
+virulence. In particular, a relatively simple, strategic
 eco-evolutionary model of HIV can predict peak $\log_{10}$
@@
 workers, of very short duration) and/or more structured partnership
-formation (by age, ethnicity, or behaviour group). The effects of
+formation (by age, ethnicity, or behaviour group).
+In contrast, the elevated viral load in the early stage of HIV infection,
+neglected in our model, will likely lead to higher maximum
+epidemic growth rates and allow more scope for transient viral
+evolution, although only if extra-pair contact is possible.
+The effects of
 other realistic complications such as explicit modeling of two sexes
@@
 or presence of genetic variation in hosts are harder to predict.
-In contrast, the elevated viral load in the early stage of HIV infection,
-neglected in our model, will likely lead to higher maximum
-epidemic growth rates and allow more scope for transient viral
-evolution, although only if extra-pair contact is possible.
 As mentioned above, our compartmental model already requires tens
@@
 but might vary over the course of an epidemic.
-It is not obvious whether models with equal parameters but
-different structures should be directly compared; calibration
+Models with equal parameters but
+different structures cannot be compared directly; calibration
 solves this problem.

### 2017-02-11 ms/S2_appendix.tex
using \section instead of \paragraph in S2

@@
 
-\paragraph*{Supporting Information S2: dynamics of transmission and virulence}
+\section*{Supporting Information S2: dynamics of transmission and virulence}
 
@@
 the figures from the main text showing time dynamics and summary
-statistics in terms of \Lspvl\ and per-year transmission probability
-rather than expected progression time to AIDS.
+statistics in terms of expected progression time to AIDS and per-year transmission probability
+rather than \Lspvl\.
 

### 2017-02-11 ms/ms.tex
add one ref; re-do bib hack

@@
 important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
-As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.
+As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016,blanquart_transmission-virulence_2016}.
 
@@
 % 
-
-\bibliography{virulence}
+\begin{thebibliography}{10}
+
+\bibitem{alizon_virulence_2009}
+Alizon S, Hurford A, Mideo N, van Baalen M.
+\newblock Virulence evolution and the trade-off hypothesis: history, current
+  state of affairs and the future.
+\newblock J Evol Biol. 2009;22:245--259.
+\newblock doi:{10.1111/j.1420-9101.2008.01658.x}.
+
+\bibitem{EbertBull2003}
+Ebert D, Bull JJ.
+\newblock Challenging the trade-off model for the evolution of virulence: is
+  virulence management feasible?
+\newblock Trends Microbiol. 2003;11(1):15--20.
+
+\bibitem{alizon_adaptive_2015}
+Alizon S, Michalakis Y.
+\newblock Adaptive virulence evolution: the good old fitness-based approach.
+\newblock Trends in Ecology \& Evolution. 2015;30(5):248--254.
+\newblock doi:{10.1016/j.tree.2015.02.009}.
+
+\bibitem{Dwyer+1990}
+Dwyer G, Levin SA, Buttel L.
+\newblock A simulation model of the population dynamics and evolution of
+  myxomatosis.
+\newblock Ecol Monog. 1990;60:423--447.
+
+\bibitem{mackinnon1999genetic}
+Mackinnon MJ, Read AF.
+\newblock Genetic relationships between parasite virulence and transmission in
+  the rodent malaria {{\em Plasmodium chabaudi}}.
+\newblock Evolution. 1999; p. 689--703.
+
+\bibitem{jensen2006empirical}
+Jensen KH, Little T, Skorping A, Ebert D.
+\newblock Empirical support for optimal virulence in a castrating parasite.
+\newblock PLoS Biol. 2006;4(7):e197.
+
+\bibitem{deroode2008virulence}
+De~Roode JC, Yates AJ, Altizer S.
+\newblock Virulence-transmission trade-offs and population divergence in
+  virulence in a naturally occurring butterfly parasite.
+\newblock Proceedings of the National Academy of Sciences.
+  2008;105(21):7489--7494.
+
+\bibitem{Fraser+2007}
+Fraser C, Hollingsworth TD, Chapman R, de~Wolf F, Hanage WP.
+\newblock Variation in {HIV}-1 set-point viral load: Epidemiological analysis
+  and an evolutionary hypothesis.
+\newblock PNAS. 2007;104:17441--17446.
+
+\bibitem{shirreff_transmission_2011}
+Shirreff G, Pellis L, Laeyendecker O, Fraser C.
+\newblock Transmission Selects for {HIV-1} Strains of Intermediate Virulence: A
+  Modelling Approach.
+\newblock PLoS Computational Biology. 2011;7(10):e1002185.
+\newblock doi:{10.1371/journal.pcbi.1002185}.
+
+\bibitem{herbeck_hiv_2014}
+Herbeck JT, Mittler JE, Gottlieb GS, Mullins JI.
+\newblock An {HIV} Epidemic Model Based on Viral Load Dynamics: Value in
+  Assessing Empirical Trends in {HIV} Virulence and Community Viral Load.
+\newblock {PLoS} Comput Biol. 2014;10(6):e1003673.
+
+\bibitem{herbeck_evolution_2016}
+Herbeck JT, Mittler JE, Gottlieb GS, Goodreau SM, Murphy JT, Cori A, et~al.
+\newblock Evolution of {HIV} virulence in response to widespread scale up of
+  antiretroviral therapy: a modeling study.
+\newblock Virus Evolution. 2016;2(2):vew028.
+\newblock doi:{10.1093/ve/vew028}.
+
+\bibitem{day_virulence_2004}
+Day T, Proulx SR.
+\newblock A General Theory for the Evolutionary Dynamics of Virulence.
+\newblock The American Naturalist. 2004;163(4):E40--E63.
+\newblock doi:{10.1086/382548}.
+
+\bibitem{alizon_price_2009}
+Alizon S.
+\newblock The {Price} equation framework to study disease within-host
+  evolution.
+\newblock Journal of Evolutionary Biology. 2009;22(5):1123--1132.
+\newblock doi:{10.1111/j.1420-9101.2009.01726.x}.
+
+\bibitem{champredon_hiv_2013}
+Champredon D, Bellan S, Dushoff J.
+\newblock {HIV} Sexual Transmission Is Predominantly Driven by Single
+  Individuals Rather than Discordant Couples: A Model-Based Approach.
+\newblock PLoS ONE. 2013;8(12):e82906.
+\newblock doi:{10.1371/journal.pone.0082906}.
+
+\bibitem{hollingsworth_hiv1_2008}
+Hollingsworth TD, Anderson RM, Fraser C.
+\newblock {HIV}-1 Transmission, by Stage of Infection.
+\newblock Journal of Infectious Diseases. 2008;198(5):687--693.
+\newblock doi:{10.1086/590501}.
+
+\bibitem{omori2015dynamics}
+Omori R, Chemaitelly H, Abu-Raddad LJ.
+\newblock Dynamics of non-cohabiting sex partnering in sub-{Saharan} {Africa}:
+  a modelling study with implications for {HIV} transmission.
+\newblock Sexually Transmitted Infections.
+  2015;doi:{10.1136/sextrans-2014-051925}.
+
+\bibitem{kretzschmar_effect_1998}
+Kretzschmar M, Dietz K.
+\newblock The effect of pair formation and variable infectivity on the spread
+  of an infection without recovery.
+\newblock Mathematical Biosciences. 1998;148(1):83--113.
+
+\bibitem{may_transmission_1988}
+May RM, Anderson RM.
+\newblock The Transmission Dynamics of Human Immunodeficiency Virus (HIV) [and
+  Discussion].
+\newblock Philosophical Transactions of the Royal Society of London Series B,
+  Biological Sciences. 1988;321(1207):565--607.
+
+\bibitem{delva_connecting_2016}
+Delva W, Leventhal GE, Helleringer S.
+\newblock Connecting the dots: network data and models in {HIV} epidemiology.
+\newblock AIDS. 2016;30(13):2009--2020.
+\newblock doi:{10.1097/QAD.0000000000001184}.
+
+\bibitem{blower_drugs_1991}
+Blower SM, Hartel D, Dowlatabadi H, Anderson RM, May RM.
+\newblock Drugs, Sex and {HIV}: A Mathematical Model for {New York City}.
+\newblock Philosophical Transactions of the Royal Society of London B:
+  Biological Sciences. 1991;331(1260):171--187.
+\newblock doi:{10.1098/rstb.1991.0006}.
+
+\bibitem{ma_estimating_2014}
+Ma J, Dushoff J, Bolker BM, Earn DJD.
+\newblock Estimating Initial Epidemic Growth Rates.
+\newblock Bulletin of Mathematical Biology. 2014;76(1):245--260.
+\newblock doi:{10.1007/s11538-013-9918-2}.
+
+\bibitem{wallinga_how_2007}
+Wallinga J, Lipsitch M.
+\newblock How generation intervals shape the relationship between growth rates
+  and reproductive numbers.
+\newblock Proceedings of the Royal Society of London B: Biological Sciences.
+  2007;274(1609):599--604.
+\newblock doi:{10.1098/rspb.2006.3754}.
+
+\bibitem{soetaert_solving_2010}
+Soetaert K, Petzoldt T, Setzer RW.
+\newblock Solving Differential Equations in {R}: Package {deSolve}.
+\newblock Journal of Statistical Software. 2010;33(9):1--25.
+\newblock doi:{10.18637/jss.v033.i09}.
+
+\bibitem{Herre1995}
+Herre EA.
+\newblock Factors affecting the evolution of virulence: Nematode parasites of
+  fig wasps as a case study.
+\newblock Parasitology. 1995;111:S179--S191.
+
+\bibitem{Kerr+2006}
+Kerr B, Neuhauser C, Bohannan BJM, Dean AM.
+\newblock Local migration promotes competitive restraint in a host-pathogen
+  `tragedy of the commons'.
+\newblock Nature. 2006;442(7098):75--78.
+
+\bibitem{webb_theoretical_2013}
+Webb SD, Keeling MJ, Boots M.
+\newblock A theoretical study of the role of spatial population structure in
+  the evolution of parasite virulence.
+\newblock Theoretical population biology. 2013;84:36--45.
+
+\bibitem{berngruber_spatial_2015}
+Berngruber TW, Lion S, Gandon S.
+\newblock Spatial {Structure}, {Transmission} {Modes} and the {Evolution} of
+  {Viral} {Exploitation} {Strategies}.
+\newblock PLOS Pathog. 2015;11(4):e1004810.
+\newblock doi:{10.1371/journal.ppat.1004810}.
+
+\bibitem{wawer2005rates}
+Wawer MJ, Gray RH, Sewankambo NK, Serwadda D, Li X, Laeyendecker O, et~al.
+\newblock Rates of {HIV}-1 transmission per coital act, by stage of {HIV}-1
+  infection, in {Rakai}, {Uganda}.
+\newblock Journal of Infectious Diseases. 2005;191(9):1403--1409.
+
+\bibitem{faria_early_2014}
+Faria NR, Rambaut A, Suchard MA, Baele G, Bedford T, Ward MJ, et~al.
+\newblock The early spread and epidemic ignition of {HIV}-1 in human
+  populations.
+\newblock Science (New York, NY). 2014;346(6205):56--61.
+\newblock doi:{10.1126/science.1256739}.
+
+\bibitem{elderd_uncertainty_2006}
+Elderd BD, Dukic VM, Dwyer G.
+\newblock Uncertainty in predictions of disease spread and public health
+  responses to bioterrorism and emerging diseases.
+\newblock Proceedings of the National Academy of Sciences. 2006;103(42):15693
+  --15697.
+\newblock doi:{10.1073/pnas.0600816103}.
+
+\bibitem{herbeck_is_2012}
+Herbeck JT, Müller V, Maust BS, Ledergerber B, Torti C, Di~Giambenedetto S,
+  et~al.
+\newblock Is the virulence of {HIV} changing? {A} meta-analysis of trends in
+  prognostic markers of {HIV} disease progression and transmission.
+\newblock AIDS (London, England). 2012;26(2):193--205.
+\newblock doi:{10.1097/QAD.0b013e32834db418}.
+
+\bibitem{payne_impact_2014}
+Payne R, Muenchhoff M, Mann J, Roberts HE, Matthews P, Adland E, et~al.
+\newblock Impact of {HLA}-driven {HIV} adaptation on virulence in populations
+  of high {HIV} seroprevalence.
+\newblock Proceedings of the National Academy of Sciences.
+  2014;111(50):E5393--E5400.
+\newblock doi:{10.1073/pnas.1413339111}.
+
+\bibitem{roberts2015impact}
+Roberts HE, Goulder PJ, McLean AR.
+\newblock The impact of antiretroviral therapy on population-level virulence
+  evolution of {HIV}-1.
+\newblock Journal of The Royal Society Interface. 2015;12(113):20150888.
+
+\end{thebibliography}
+
 \end{document}

### 2017-02-11 ms/cover_letter/plosHIVcover_resubm2.tex
stuff for resubmission

@@
 }
+
 \response{This is a very general comment, but we have gone
@@
 to overwhelm the reader). We have also changed presentation more generally so that we don't lead by
-saying ``as in Shireff et al \ldots'', which might make readers think that they need to know Shirreff's
-material, although we still try to be clear where we are following Shirref et al/Champredon et al and
+saying ``as in Shirreff et al \ldots'', which might make readers think that they need to know Shirreff's
+material, although we still try to be clear where we are following Shirreff et al/Champredon et al and
 where we are breaking new ground.}

### 2017-02-11 ms/ms.tex
stuff for resubmission

@@
 important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, it represents a first step toward assessing the effects of epidemiological complexity.
-As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016,blanquart_transmission-virulence_2016}.
+As even the models shown here push the limits of compartmental-based models (the heterogeneity model comprises 24530 coupled ordinary differential equations), adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016,blanquart_transmission_virulence_2016}.
 
@@
 
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here.
-% 
+% \bibliography{virulence}
+
 \begin{thebibliography}{10}
@@
 
+\bibitem{kretzschmar_effect_1998}
+Kretzschmar M, Dietz K.
+\newblock The effect of pair formation and variable infectivity on the spread
+  of an infection without recovery.
+\newblock Mathematical Biosciences. 1998;148(1):83--113.
+
 \bibitem{champredon_hiv_2013}
@@
 
-\bibitem{kretzschmar_effect_1998}
-Kretzschmar M, Dietz K.
-\newblock The effect of pair formation and variable infectivity on the spread
-  of an infection without recovery.
-\newblock Mathematical Biosciences. 1998;148(1):83--113.
-
 \bibitem{may_transmission_1988}
@@
 \newblock doi:{10.1097/QAD.0000000000001184}.
+
+\bibitem{blanquart_transmission_virulence_2016}
+Blanquart F, Grabowski MK, Herbeck J, Nalugoda F, Serwadda D, Eller MA, et~al.
+\newblock A transmission-virulence evolutionary trade-off explains attenuation
+  of {HIV}-1 in {Uganda}.
+\newblock eLife. 2016;5:e20492.
+\newblock doi:{10.7554/eLife.20492}.
 
@@
 
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here.
+%
+
 \end{document}

### 2017-03-14 figures/figures.R
dinking around with figures

@@ -57,7 +57,8 @@ fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
            labels=new_sum_labs)
 }
 
-### Figure 1
+fignum <- 2
+### Figure 2 (3-panel figure)
 
 for (i in c("","2","3")) 
     load(sprintf("../simdata/ev_LHS_resS%s.rda",i))
@@ -73,12 +74,17 @@ df_tot <- rbind(shirreff_df_m,d_littleR_m)
 tlab <- "time (years)"
 g1 <- ggplot(df_tot,aes(x=Time,y=Mean.VL,colour=run))+
     geom_line(aes(linetype=run))+labs(x=tlab,
-            y=expression(population~mean~set-point~viral~load~(log[10])))+
+            y=expression("pop. mean set-point viral load"~(log[10])))+
 		scale_x_continuous(expand = c(0.005, 0), breaks = seq(0, 600, by = 100), 
 											 labels = c("0", "", "200", "", "400", "", "")) +
-    theme(legend.position=c(0.63,0.45),
-    			legend.justification = c(0, 1),
-    			panel.grid = element_blank())
+    theme(legend.position=c(0.5,0.45),
+          ## plot.margin=grid::unit(c(1,0.5,0.5,0.5),"cm"),
+          legend.justification = c(0, 1),
+          legend.key.height=grid::unit(0.7,"line"),
+          legend.text=element_text(size=10),
+          ## *positive* hjust shifts to the *right* ???
+          axis.title.y = element_text(size=11.5,hjust=1),
+          panel.grid = element_blank())
 
 ## Compute difference between 
 
@@ -130,18 +136,25 @@ g3.y <- g3 + scale_y_continuous(limits = limits, breaks = breaks)+
     scale_linetype_discrete(name=bquote(alpha*"(0)"))
 
 ## r fig1,fig.height=3.5,fig.width=10, echo = FALSE, dpi = 600}
-fig1 <- arrangeGrob(g1.y, g2.y, g3.y, nrow=1)
-if (do_pdf) ggsave("fig1.pdf", fig1, height = 3.5, width = 10)
+fig_panel3 <- arrangeGrob(g1.y, g2.y, g3.y, nrow=1)
+figtweak <- 0.8
+figname <- paste0("fig",fignum)
+
+if (do_pdf) ggsave(paste0(figname,".pdf"), fig_panel3,
[…]

### 2017-03-18 ms/ms.tex
dinking around with figures
figure orders and acknowledgement

@@
 
+\begin{figure}[!ht]
+\caption{\textbf{Baseline dynamics.}
+Time series of mean population \Lspvl\ using baseline parameters.
+(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
+(b) Effects of varying initial infectious density $I(0)$.
+(c) Effects of varying initial mean virulence $\alpha(0)$.
+The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
+The other panels (b,c) show that eco-evolutionary dynamics are qualitatively similar across a range of initial conditions.}
+\label{fig:panel3}
+\end{figure}
+
 We then construct an $n \times n$ mutational matrix, $M$ --- which is multiplied with the transmission term ---  so that $M_{ij}$ is the probability that a newly infected individual will have \Lspvl\ of $\alpha_j$ given that their infected partner has \Lspvl\ of $\alpha_i$.
@@
 
+We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} contact structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
+Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
+non-instantaneous partnership formation; (2) zero vs.
+positive extra-pair sexual contact and transmission; and (3) homogeneous vs.
+heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
+
+Our first four models (\figurename~\ref{fig:diagrams}) explicitly consider partnership dynamics \cite{champredon_hiv_2013}.
+The first (\diagfig{d}) assumes non-instantaneous partnership formation (i.e.
+individuals spend some time uncoupled, outside of partnerships) and consists of five states that are classified by infection status and partnership status; single (uncoupled) susceptible individuals ($S$), single infected individuals ($I$), concordant negative (susceptible-susceptible) couples ($SS$), discordant (susceptible-infected) couples ($SI$), and concordant positive (infected-infected) couples ($II$).
+The rates of pair formation are based on the numbers of uncoupled susceptible and infected individuals and the pair-formation rate; partnerships can either dissolve into singletons or be transformed into other types of partnerships by infection of one partner.
+This model also includes extra-pair contact with both uncoupled individuals and individuals in other partnerships (we denote it ``pairform+epc''), so that susceptible uncoupled individuals and susceptible partners in any type of partnership can be infected by infected uncoupled individuals or infected partners in any type of partnership.
+
 \begin{figure}[t!p]
-
-\begin{subfigure}[b]{0.4\linewidth}
-	\centering
-	\raisebox{0.44\height}{
-	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f1}
-	}
-\caption{Instantaneous switching without extra-pair contact (``instswitch'')}
-\end{subfigure}
-\begin{subfigure}[b]{0.4\linewidth}
-	\centering
-	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f2}
-\caption{Instantaneous switching with extra-pair contact (``instswitch+epc'')}
-\end{subfigure}
-
-\begin{subfigure}[b]{0.4\linewidth}
-	\centering
-	\raisebox{0.33\height}{
-	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f3}
-	}
-\caption{Pair-formation model without extra-pair/uncoupled contact 
-(``pairform'')}
-\end{subfigure}
-\begin{subfigure}[b]{0.4\linewidth}
-	\centering
-	  \includegraphics[width=0.9\linewidth]{../figures/tikz-f4}
-\caption{Pair-formation model with extra-pair/uncoupled contact
-(``pairform+epc'')}
-\end{subfigure}
-
 \caption{\textbf{Model contact structures.} 
@@
 \end{figure}
-
-We developed seven multi-strain evolutionary models covering a gamut including Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} contact structures, each of which is based on different assumptions regarding contact structure and partnership dynamics.
-Specifically, we focus on the effects of the assumptions of (1) instantaneous vs.
-non-instantaneous partnership formation; (2) zero vs.
-positive extra-pair sexual contact and transmission; and (3) homogeneous vs.
-heterogeneous levels of sexual activity on the evolution of mean \Lspvl.
-
-Our first four models (\figurename~\ref{fig:diagrams}) explicitly consider partnership dynamics \cite{champredon_hiv_2013}.
-The first (\diagfig{d}) assumes non-instantaneous partnership formation (i.e.
-individuals spend some time uncoupled, outside of partnerships) and consists of five states that are classified by infection status and partnership status; single (uncoupled) susceptible individuals ($S$), single infected individuals ($I$), concordant negative (susceptible-susceptible) couples ($SS$), discordant (susceptible-infected) couples ($SI$), and concordant positive (infected-infected) couples ($II$).
-The rates of pair formation are based on the numbers of uncoupled susceptible and infected individuals and the pair-formation rate; partnerships can either dissolve into singletons or be transformed into other types of partnerships by infection of one partner.
-This model also includes extra-pair contact with both uncoupled individuals and individuals in other partnerships (we denote it ``pairform+epc''), so that susceptible uncoupled individuals and susceptible partners in any type of partnership can be infected by infected uncoupled individuals or infected partners in any type of partnership.
 
@@
 
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig1.pdf}
-\caption{\textbf{Baseline dynamics.}
-Time series of mean population \Lspvl\ using baseline parameters.
-(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
-(b) Effects of varying initial infectious density $I(0)$.
-(c) Effects of varying initial mean virulence $\alpha(0)$.
-The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
-The other panels (b,c) show that eco-evolutionary dynamics are qualitatively similar across a range of initial conditions.}
-\label{fig:panel3}
-\end{figure}
-
 Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig2.pdf}
 \caption{\textbf{Envelopes of virulence trajectories (population mean \Lspvl) under all models.}
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig3.pdf}
 \caption{{\bf Univariate distributions of summary statistics.}
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig4.pdf}
 \caption{{\bf Pairs plot: bivariate relationships among summary statistics for each model structure.}
@@
 
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figures/fig5.pdf}
-\caption{{\bf Sensitivity plot.}
-Distribution (points) and trend (smooth line) of summary statistics (y-axis,
-rows) as a function of parameter values (x-axis, columns), some of which have been scaled to calibrate the initial epidemic growth rate.
-Changing parameters may have differing effects for models with different
-contact structures: for example, partnership dissolution rate ($c$, third column) increases peak time and virulence in models without explicit pair formation but decreases them in more realistic models.
-}
-\label{fig:plot_sens}
-\end{figure}
-
 Bivariate relationships (\figurename~\ref{fig:pairplot}) help distinguish the results of 
@@
 they show patterns almost identical to $\beta_P$ and $D_P$, respectively.
+
+\begin{figure}[!ht]
+\caption{{\bf Sensitivity plot.}
+Distribution (points) and trend (smooth line) of summary statistics (y-axis,
+rows) as a function of parameter values (x-axis, columns), some of which have been scaled to calibrate the initial epidemic growth rate.
+Changing parameters may have differing effects for models with different
+contact structures: for example, partnership dissolution rate ($c$, third column) increases peak time and virulence in models without explicit pair formation but decreases them in more realistic models.
+}
+\label{fig:plot_sens}
+\end{figure}
 
@@
 David Champredon for access to simulation code
-and David Earn, Wim Delva, and an anonymous reviewer for useful comments;
-this work was funded by NSERC Discovery Grant 386590-2010.
+and David Earn for useful comments.
 Computational resources were provided by SHARCnet.

### 2017-03-22 figures/figures.R
working on figures

@@ -57,8 +57,8 @@ fixfac2 <- function(x,atop=FALSE,newlines=FALSE) {
            labels=new_sum_labs)
 }
 
-fignum <- 2
-### Figure 2 (3-panel figure)
+fignum <- 1
+### Figure 1 (3-panel figure)
 
 for (i in c("","2","3")) 
     load(sprintf("../simdata/ev_LHS_resS%s.rda",i))
@@ -152,7 +152,9 @@ if (do_png) {
 
 fig_objects <- c("g1.y","g2.y","g3.y")
 
-fignum <- fignum + 1
+
+## skip figure 2 (3-panel figure)
+fignum <- 3
 
 ### Figure 3 (SPVL trajectory envelopes)
 

### 2017-03-22 ms/sub_bib.R

@@ -1,0 +1,8 @@
+txt <- readLines("plos_latex.tex")
+bibtxt <- readLines("plos_latex.bbl")
+target <- grep("\\bibliography{virulence}",txt,fixed=TRUE)
+if (length(target)==0) stop("can't find bib line")
+out_txt <- c(txt[1:(target-1)],
+             bibtxt,
+             txt[target:length(txt)])
+writeLines("plos_latex_out.tex")

### 2017-03-22 poster/figure/tradeoff.R
cleanup; remove poster

@@ -1,16 +1,0 @@
-source("fig_setup.R")
-
-pars2 <- HIVpars.base
-pars2$alphaDist <- c(min=0,max=9,delta=0.25)
-pars2 <- expand(pars2)
-dd <-with(pars2,
-     melt(data.frame(alpha=alpha,"transmission probability"=Beta2,
-                     "duration (years)"=Duration2,check.names=FALSE),
-          id.vars="alpha"))
-gg_tradeoff <- ggplot(dd,aes(alpha,value))+facet_wrap(~variable,scale="free")+
-    geom_line(size=7)+
-    labs(x= Lspvl,y="")
-
-tikz(file="tradeoff.tikz",width=12,height=8,standAlone=TRUE)
-print(gg_tradeoff)
-dev.off()

### 2017-03-22 poster/figure/univ.R

@@ -1,32 +1,0 @@
-source("fig_setup.R")
-
-sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
-sL$model <- factor(sL$model, c("random","pairform+epc", "pairform", "instswitch+epc","instswitch" , "implicit"))
-mL <- melt(sL,id.vars=c("model","run"))
-## horrible hack (but doesn't help); subsample
-## w <- which(mL$model=="random")
-## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
-w <- with(mL,which(model=="random" & variable=="eq_vir"))
-rval <- mean(mL$value[w])
-mLw <- droplevels(mL[-w,])
-mLw$variable <- fixfac2(mLw$variable)
-
-gg_univ <- ggplot(mLw,aes(value,model,fill=model))+
-    geom_violinh(width=1)+
-    facet_wrap(~variable,scale="free_x")+
-    guides(fill=guide_legend(reverse=TRUE))+
-    labs(y="",x="")+
-    geom_point(data=data.frame(model="random",
-                               variable="equilibrium virulence",
-                               value=rval),
-               pch=22,size=8,show.legend=FALSE) +
-    theme(legend.position="none")+
-    zero_x_margin
-
-tikz(file="univ.tikz",w
-     ## width=12,height=10,
-     ## width=10,height=8,
-     width=9,height=9*5/6,
-     standAlone=TRUE)
-print(gg_univ)
-dev.off()

### 2017-03-22 poster/figure/virtraj.R

@@ -1,21 +1,0 @@
-source("fig_setup.R")
-
-ss <- subset(sum_list[["vir_mat"]],tvec<750)
-ss$model <- factor(ss$model, c("random",
-         "pairform+epc", "pairform","instswitch+epc","instswitch","implicit"))
-
-gg_virtraj <- ggplot(ss,
-                     aes(tvec,mean,ymin=lwr,ymax=upr))+
-    geom_line(aes(colour=model,linetype=model),lwd=1)+
-    geom_ribbon(aes(fill = model), alpha=0.3)+
-    labs(x="time (years)",y="population mean set-point viral load ($\\log_{10}$)") +
-    theme(axis.title.y = element_text(margin = margin(0,10,0,0)),
-          axis.title.x = element_text(margin = margin(10,0,0,0))) +
-    facet_wrap(~model, nrow = 3) +
-    zero_margin+
-    theme(legend.position="none")+
-    scale_x_continuous(breaks=c(0,600))
-
-tikz(file="virtraj.tikz",width=8,height=11,standAlone=TRUE)
-print(gg_virtraj)
-dev.off()

### 2017-03-22 ms/S2_appendix.tex
minor tweaks to ms

@@
 
-\renewcommand\thefigure{S\arabic{figure}}
+\renewcommand\thefigure{\Alph{figure}}
 
@@
 statistics in terms of expected progression time to AIDS and per-year transmission probability
-rather than \Lspvl\.
+rather than \Lspvl.
 

### 2017-03-22 ms/ms.tex
minor tweaks to ms

@@
 multiplying the within-couple transmission rate $\beta$ by the contact
-ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix S1).
+ratios $c_u/c_w$ and $c_e/c_w$ (see S1 Appendix).
 
@@
 Lastly, we incorporated heterogeneity in sexual activity into the models.
-Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix S1 for full model details).
+Individuals are divided into different risk groups based on their level of sexual activity; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see S1 Appendix for full model details).
 We assume random activity-weighted mixing between risk groups \cite{may_transmission_1988}.
-(In the main text we focus on the model with non-instantaneous pair formation, extra-pair contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~S4 in Appendix~S2 presents results on the effect of adding heterogeneity to other model variants.)
+(In the main text we focus on the model with non-instantaneous pair formation, extra-pair contact (``pairform+epc'') and heterogeneous sexual activity, which we denote as ``hetero''; \figurename~D in S2 Appendix presents results on the effect of adding heterogeneity to other model variants.)
 While this model lacks some 
@@
 with an initial \Lspvl\ of 3.0, to match the value used by
-Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
+Shirreff \etal \cite{shirreff_transmission_2011}. Shirreff \etal\ use an initial prevalence $I(0)=10^{-3}$;
 because we calibrated parameters based on the initial epidemic growth rate
@@
 
-Appendix~S1 provides
+S1 Appendix provides
 further details on initial conditions, such
@@
 In particular, we tested model sensitivity 
-to the initial prevalence $I(0)$ and initial mean SPVL
+to the initial prevalence, $I(0)$, and initial mean \Lspvl,
 $\alpha(0)$, using baseline values of all parameters (Table~\ref{table:parmsTable}).
@@
 
-One new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see Appendix S1).
+One new parameter in our model, the ratio of uncoupled to within-couple transmission $c_u/c_w$, is needed to more flexibly contrast uncoupled and extra-couple transmission dynamics within multi-strain models (see S1 Appendix).
 Since it appears neither in either Shirreff \etal\ nor Champredon \etal's models,  we need to pick a reasonable range for it.
@@
 After each run, the initial exponential growth rate is calculated.
-Then the parameters are scaled as described above so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
+Then the parameters are scaled as described above so that the initial exponential growth rate is scaled to 0.04 \peryear, a value that approximates the growth rates of Shirreff \etal's original models.
 When calibrating, we run each model for only 500 years
@@
 virulence may decline immediately, leading to a peak virulence (as we have
-defined it) equal to the starting value of $\alpha(0)=3$ \Lspvl at $t=0$
+defined it) equal to the starting value of $\alpha(0)=3$ \Lspvl\ at $t=0$
 (\figurename~\ref{fig:pairplot}).
@@
 random and pairform+epc models vs. 0.059 (0.02-0.13) per year
-for the implicit model. (Appendix S2 
-presents plots showing univariate summaries
+for the implicit model. (S2 Appendix presents plots showing univariate summaries
 of expected progression time to AIDS and transmission probability.)
@@
 leads to only small differences in the conclusions stated so far
-(\figurename~S4). Relative to our baseline SIS assumption, the
+(\figurename~D). Relative to our baseline SIS assumption, the
 effect of adding vital dynamics is to delay the virulence peak
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{Appendix S1: Model Details}
+\paragraph*{S1 Appendix: Model Details}
 \label{S1_Appendix}
 
-\paragraph*{Appendix S2: Dynamics of Transmission and Virulence}
+\paragraph*{S2 Appendix: Dynamics of Transmission and Virulence}
 \label{S2_Appendix}

### 2017-03-22 figures/figures.R

@@ -237,8 +237,8 @@ gg_durtraj <- direct.label(gg_basetraj %+% ss_dur + nostrips +
 													 	labs(y="expected progression to AIDS (years)"),
 													 method=list("bottom.points2","bumpup"))
 
-if (do_pdf) ggsave(gg_durtraj,file="fig_S2_2.pdf",width=6,height=4)
-if (do_png) ggsave(gg_durtraj,file="fig_S2_2.png",width=6,height=4,dpi=400)
+if (do_pdf) ggsave(gg_durtraj,file="fig_S2_2.pdf",width=8,height=6)
+if (do_png) ggsave(gg_durtraj,file="fig_S2_2.png",width=8,height=6,dpi=400)
 
 fignum <- fignum+1
 
